---
title: "posteriorHCA brms fit compression test"
date: today
date-format: "DD MMM YYYY"
title-block-banner: true
author: 
  - name: Chen Zhan
    orcid: 0000-0002-4794-8339
    email: chen.zhan@adelaide.edu.au
    affiliation:
      - id: sgc
        name: SAiGENCI, The University of Adelaide
format: 
  html:
    theme: minty
    df-print: paged
    code-line-numbers: true
    embed-resources: true
    code-fold: true
    code-fold-default: true
    toc: true
    toc-location: left
    toc-title: "Menu"
knitr:
  opts_chunk: 
    message: false
    warning: false
comments:
  hypothesis: 
    theme: clean
editor: source
abstract: |
  posteriorHCA brms fit compression test
---

# Library

```{r}
#| label: setup
.libPaths("/hpcfs/users/a1237163/R/x86_64-pc-linux-gnu-library/4.4")
dyn.load("/home/a1237163/miniconda3/lib/libglpk.so.40")
library(tidyverse)
library(tidySummarizedExperiment)
library(tidySingleCellExperiment)
library(tidybulk)
library(SingleCellExperiment)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(ggplot2)
library(scales)
library(tidytext)
library(glue)
library(targets)
library(here)
library(Seurat)
library(tidyseurat)
library(patchwork)
library(vissE)
library(tidyHeatmap)
library(circlize)
library(ggupset)
library(brms)
library(edgeR)
library(magrittr)
# set_cmdstan_path('/hpcfs/users/a1237163/R/.cmdstan/cmdstan-2.36.0')
```

# Pre-load and setting

```{r}

library(mclust)

source("/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/R/compress_mclust.R")

ct <- 
  'plasma' %>% 
  paste0('V1_',.)

target_pipeline_path = '/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/HPC_posterior/NEW_CELL_TYPE/'
target_obj_path = file.path(target_pipeline_path, ct, '_targets')

```


# Read brms fit file from pipeline
```{r}
set.seed(2603)

estimates_chunk_meta <- 
  tar_meta(
    starts_with('estimates_chunk_'),
    store = target_obj_path
  )


estimates_chunk <-
  tar_read_raw(
    estimates_chunk_meta$name %>% sample(1),
    store = target_obj_path
  )

fit_brms <- estimates_chunk$brms_fit[[1]]
  
```

```{r}
class(fit_brms)
print(object.size(fit_brms), units = "auto")
# summary(fit_brms)
```


## Inspect the Original Fit

```{r inspect_fit}
# Model summary
# summary(fit_brms)

# Check convergence
cat("\nConvergence diagnostics:\n")
cat("Max Rhat:", max(rhat(fit_brms), na.rm = TRUE), "\n")
cat("Min Bulk ESS:", min(neff_ratio(fit_brms) * 4000, na.rm = TRUE), "\n")

# Original posterior structure
original_draws <- as_draws_matrix(fit_brms)
cat("\nOriginal posterior dimensions:", dim(original_draws), "\n")
cat("Parameters:", colnames(original_draws)[1:10], "...\n")
```

# Compression
```{r}
# "EII" spherical, equal volume
# "EEI" diagonal, equal volume and shape

fit_brms_compressed <- compress_brmsfit(
  brmsfit = fit_brms,
  n_components = 3,      # Use more components for complex posterior
  # model_name = 'EII',
  remove_csvs = TRUE    # Keep CSVs (or TRUE to clean up)
)

```
```{r}
print(object.size(fit_brms_compressed$compressed), units = "auto")
```

Save compressed 
```{r}
compressed_file <- "brms_posterior_compressed.rds"
saveRDS(fit_brms_compressed$compressed, compressed_file, compress = "xz")


compressed_size <- file.size(compressed_file)
cat("\nCompressed size (with xz):", format(compressed_size, big.mark = ","), "bytes\n")
cat("That's", round(compressed_size / 1024, 2), "KB\n")
```

# Step 4: Regenerate Posterior Samples

Sample from the compressed posterior distribution.

```{r regenerate_samples}
# Generate same number of samples as original
n_draws_original <- nrow(original_draws)
cat("Generating", n_draws_original, "samples from compressed posterior...\n")

# Sample from compressed distribution
regenerated_samples <- sample_from_mclust(
  compressed_rds = compressed_file,
  n_draws = n_draws_original
)

cat("Regenerated samples dimensions:", dim(regenerated_samples), "\n")
cat("Parameters:", colnames(regenerated_samples)[1:10], "...\n")

param_names <- readRDS(compressed_file)$param_names

# Basic quality check: compare means and SDs
original_means <- colMeans(original_draws[, param_names])
regenerated_means <- colMeans(regenerated_samples)

original_sds <- apply(original_draws[, param_names], 2, sd)
regenerated_sds <- apply(regenerated_samples, 2, sd)

cat("\nQuality check - first 5 parameters:\n")
comparison_df <- data.frame(
  Parameter = param_names[1:5],
  Original_Mean = round(original_means[1:5], 3),
  Regenerated_Mean = round(regenerated_means[1:5], 3),
  Original_SD = round(original_sds[1:5], 3),
  Regenerated_SD = round(regenerated_sds[1:5], 3)
)
print(comparison_df)
```


# Visualization: Compare Original vs Regenerated Posteriors

With ~`r length(param_names)` parameters we cannot eyeball every marginal,
so the plots below summarise the comparison along three complementary axes:

1. **Parameter-wide calibration** (one dot per parameter) of posterior means
   and SDs.
2. **Marginal density overlays** for a random subset of parameters.
3. **Joint 2D density** for random parameter pairs, to check whether the
   compressed mixture preserves dependencies and not just marginals.
4. **Per-parameter Q-Q** to assess tail fidelity.

```{r viz_setup}
# Color palette + small helper, kept local to this section.
src_colors <- c(original = "steelblue", regenerated = "darkgreen")

# Pre-build the two aligned matrices we visualize repeatedly.
orig_mat  <- as.matrix(original_draws[, param_names, drop = FALSE])
regen_mat <- as.matrix(regenerated_samples[, param_names, drop = FALSE])

# Drop posterior classes (e.g. draws_matrix) so tidyverse can combine columns.
orig_mat <- unclass(orig_mat)
regen_mat <- unclass(regen_mat)
storage.mode(orig_mat) <- "double"
storage.mode(regen_mat) <- "double"
```

## 1. Mean and SD calibration across all parameters

A perfect compression sits exactly on the red identity line; deviations reveal
which parameters are biased or over/under-dispersed by the GMM approximation.

```{r calibration_plots, fig.width=11, fig.height=5}
calibration_df <- tibble::tibble(
  parameter        = param_names,
  original_mean    = original_means,
  regenerated_mean = regenerated_means,
  original_sd      = original_sds,
  regenerated_sd   = regenerated_sds
)

p_mean_calib <- ggplot(calibration_df,
                       aes(x = regenerated_mean, y = original_mean)) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_point(alpha = 0.4, color = src_colors[["original"]], size = 0.8) +
  labs(
    title    = "Posterior means",
    subtitle = paste(length(param_names), "parameters"),
    x        = "Regenerated posterior mean",
    y        = "Original posterior mean"
  ) +
  theme_minimal()

# Robust axis limits so extreme outliers do not collapse all points.
mean_xy <- c(calibration_df$original_mean, calibration_df$regenerated_mean)
mean_lim <- stats::quantile(mean_xy, probs = c(0.01, 0.99), na.rm = TRUE)

p_mean_calib <- p_mean_calib +
  coord_equal(xlim = mean_lim, ylim = mean_lim)

p_sd_calib <- ggplot(calibration_df,
                     aes(x = regenerated_sd, y = original_sd)) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_point(alpha = 0.4, color = "darkorange", size = 0.8) +
  labs(
    title    = "Posterior SDs",
    subtitle = paste(length(param_names), "parameters"),
    x        = "Regenerated posterior SD",
    y        = "Original posterior SD"
  ) +
  theme_minimal()

sd_xy <- c(calibration_df$original_sd, calibration_df$regenerated_sd)
sd_lim <- stats::quantile(sd_xy, probs = c(0.01, 0.99), na.rm = TRUE)
sd_lim[1] <- max(0, sd_lim[1])  # SD cannot be negative

p_sd_calib <- p_sd_calib +
  coord_equal(xlim = sd_lim, ylim = sd_lim)

p_mean_calib + p_sd_calib
```

```{r calibration_summary}
mean_err <- with(calibration_df, regenerated_mean - original_mean)
sd_rel   <- with(calibration_df,
                 (regenerated_sd - original_sd) /
                 pmax(original_sd, .Machine$double.eps))

calibration_summary <- tibble::tibble(
  metric = c(
    "RMSE(mean)",
    "MAE(mean)",
    "median |relative SD error|",
    "max |relative SD error|"
  ),
  value = c(
    sqrt(mean(mean_err^2)),
    mean(abs(mean_err)),
    median(abs(sd_rel)),
    max(abs(sd_rel))
  )
)
print(calibration_summary)
```

## 2. Marginal density overlays for a random subset

```{r marginal_overlays, fig.width=11, fig.height=8}
set.seed(42)
n_show <- 9
sampled_params <- sample(param_names, min(n_show, length(param_names)))

marginal_df <- dplyr::bind_rows(
  tibble::as_tibble(orig_mat[, sampled_params, drop = FALSE]) |>
    tidyr::pivot_longer(dplyr::everything(),
                        names_to = "parameter", values_to = "value") |>
    dplyr::mutate(source = "original"),
  tibble::as_tibble(regen_mat[, sampled_params, drop = FALSE]) |>
    tidyr::pivot_longer(dplyr::everything(),
                        names_to = "parameter", values_to = "value") |>
    dplyr::mutate(source = "regenerated")
) |>
  dplyr::mutate(parameter = factor(parameter, levels = sampled_params))

ggplot(marginal_df, aes(x = value, fill = source, color = source)) +
  geom_density(alpha = 0.35) +
  facet_wrap(~ parameter, scales = "free", ncol = 3) +
  scale_fill_manual(values = src_colors) +
  scale_color_manual(values = src_colors) +
  labs(
    title    = "Marginal posterior densities",
    subtitle = paste("Random subset of", length(sampled_params), "parameters"),
    x        = "Parameter value",
    y        = "Density"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## 3. Joint 2D densities for random parameter pairs

This checks whether the GMM preserves *dependencies* and not just marginals.
A diagonal-covariance mixture (e.g. `EEI`/`VVI`) cannot represent within-component
correlations, so off-diagonal "tilts" in original pairs may be lost — these plots
make any such loss visible.

```{r joint_density, fig.width=11, fig.height=10}
set.seed(7)
n_pairs  <- 4
pair_idx <- replicate(n_pairs, sample(seq_along(param_names), 2),
                      simplify = FALSE)

pair_plots <- lapply(pair_idx, function(idx) {
  pp <- param_names[idx]
  df <- dplyr::bind_rows(
    tibble::tibble(x = orig_mat[, pp[1]],  y = orig_mat[, pp[2]],
                   source = "original"),
    tibble::tibble(x = regen_mat[, pp[1]], y = regen_mat[, pp[2]],
                   source = "regenerated")
  )
  ggplot(df, aes(x = x, y = y, color = source)) +
    geom_point(alpha = 0.25, size = 0.4) +
    stat_density_2d(alpha = 0.7, linewidth = 0.4) +
    scale_color_manual(values = src_colors) +
    labs(x = pp[1], y = pp[2]) +
    theme_minimal() +
    theme(legend.position = "none",
          axis.title = element_text(size = 8))
})

patchwork::wrap_plots(pair_plots, ncol = 2) +
  patchwork::plot_annotation(
    title    = "Joint posteriors for random parameter pairs",
    subtitle = "Original (blue) vs regenerated (green)"
  )
```

```{r correlation_calibration, fig.width=6, fig.height=6}
# Quantitative companion to the joint plots: how well are pairwise correlations
# preserved? We sample a few hundred random pairs and compare correlations.
set.seed(11)
n_corr_pairs <- min(500, choose(length(param_names), 2))
corr_pair_idx <- replicate(
  n_corr_pairs,
  sample(seq_along(param_names), 2),
  simplify = FALSE
)

corr_df <- purrr::map_dfr(corr_pair_idx, function(idx) {
  pp <- param_names[idx]
  tibble::tibble(
    pair        = paste(pp, collapse = " ~ "),
    cor_orig    = cor(orig_mat[, pp[1]],  orig_mat[, pp[2]]),
    cor_regen   = cor(regen_mat[, pp[1]], regen_mat[, pp[2]])
  )
})

ggplot(corr_df, aes(x = cor_orig, y = cor_regen)) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_point(alpha = 0.4, color = "purple", size = 0.9) +
  coord_equal(xlim = c(-1, 1), ylim = c(-1, 1)) +
  labs(
    title    = "Pairwise correlations: original vs regenerated",
    subtitle = paste(nrow(corr_df), "random parameter pairs"),
    x        = "cor(original)",
    y        = "cor(regenerated)"
  ) +
  theme_minimal()
```

## 4. Per-parameter Q-Q plots

Tail behaviour matters for predictive intervals. The Q-Q points should hug
the red identity line if the GMM captures the full distribution (not just
mean/SD).

```{r qq_plots, fig.width=11, fig.height=7}
qq_params <- sampled_params[seq_len(min(6, length(sampled_params)))]
qq_probs  <- ppoints(200)

qq_df <- purrr::map_dfr(qq_params, function(p) {
  tibble::tibble(
    parameter = p,
    q_orig    = stats::quantile(orig_mat[, p],  probs = qq_probs, names = FALSE),
    q_regen   = stats::quantile(regen_mat[, p], probs = qq_probs, names = FALSE)
  )
}) |>
  dplyr::mutate(parameter = factor(parameter, levels = qq_params))

ggplot(qq_df, aes(x = q_orig, y = q_regen)) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_point(alpha = 0.6, color = "purple", size = 0.7) +
  facet_wrap(~ parameter, scales = "free", ncol = 3) +
  labs(
    title    = "Q-Q plots: original vs regenerated quantiles",
    subtitle = "Points on the red line indicate good distributional fidelity",
    x        = "Original quantiles",
    y        = "Regenerated quantiles"
  ) +
  theme_minimal()
```


