---
title: "TM4SF1 in histological-variant bladder cancer"
subtitle: "Design-standardised posteriorHCA baseline (v2 working copy)"
date: today
date-format: "DD MMM YYYY"
title-block-banner: true
author:
  - name: Chen Zhan
    orcid: 0000-0002-4794-8339
    email: chen.zhan@adelaide.edu.au
    affiliation:
      - id: sgc
        name: SAiGENCI, The University of Adelaide
format:
  html:
    theme: minty
    df-print: paged
    code-line-numbers: true
    embed-resources: true
    html-math-method: mathjax
    fig-path: results/tm4sf1_toxicity_report_v2_design_standardised/figures
    code-fold: true
    code-fold-default: true
    toc: true
    toc-location: left
    toc-title: "Contents"
knitr:
  opts_chunk:
    message: false
    warning: false
    fig.retina: 2
    fig.align: center
    dpi: 300
editor: source
execute:
  freeze: auto
abstract: |
  **Version 2 (design-standardised baseline).** The frozen v1 observed-profile
  report is preserved separately; this working copy replaces empirical
  donor-profile weighting with prespecified, biologically admissible
  marginalisation weights. The bladder-cancer paper identified TM4SF1 as a
  candidate antigen in histological-variant (HV) tumours. This report
  reconstructs that tumour-expression signal, then screens which
  standardised Normal/10x3 posteriorHCA tissue × cell-type baselines overlap the
  HV tumour-epithelial range. An overlap flag is a reproducible follow-up rule,
  not evidence of toxicity. Figures follow Nature journal style (axis labels
  and legends only; canvas ≤ 183 mm wide).
params:
  case_dir: "."
  model_root: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/HPC_posterior/NEW_CELL_TYPE"
  pseudobulk_h5ad: "/home/a1237163/lab/Mangiola_ImmuneAtlas/zenodo/pseudobulk_se.h5ad"
  gene_symbol: "TM4SF1"
  gene_ensembl: "ENSG00000169908"
  model_cell_types: ["epithelial", "endothelial"]
  n_boot: 2000
  threshold_quantile: 0.05
  alert_probability: 0.80
  min_donor_keys: 10
  refresh_hca: false
  seed: 293189
  v1_results_dir: "archive/v1_observed_profile/results_v1_observed_profile"
  baseline_method: "design_standardised"
  fig_width_mm: 183
  panel_sm_mm: 85
  panel_md_mm: 90
  panel_lg_mm: 120
---

# Question and interpretation boundary

**Report version:** v2 design-standardised baseline (`TM4SF1_toxicity_case_study_v2_design_standardised.qmd`).
The frozen v1 observed-profile snapshot is
`archive/v1_observed_profile/TM4SF1_toxicity_case_study_v1_observed_profile.qmd`;
the original `TM4SF1_toxicity_case_study.qmd` is unchanged.

The report asks which standardised Normal-labelled, 10x Genomics 3 posteriorHCA
tissue × cell-type baselines overlap the HV tumour-epithelial TM4SF1 range after
the study counts are aligned to each cell-type model's own offset-zero TMMwsp
reference.

An overlap flag is not evidence of toxicity. It is a prioritisation rule for
protein localisation, antigen-density, accessibility and functional-killing
experiments. Paired/adjacent normal bladder samples are retained as calibration
samples and are never called fully healthy tissue. Three have at least 20
epithelial cells (UC04, VAR07 and VAR10), but only VAR07 has a counterpart in
the retained tumour cohort, so this is not a matched-pair analysis.

------------------------------------------------------------------------

# Analysis map

The biological unit is the patient throughout. Cell-level source data are used
only to reproduce the paper figure; cohort inference uses patient-level summed
raw counts.

1. Re-read the authors' released Figure 5 source tables and confirm the TM4SF1
   differential-expression rank.
2. Display both the released per-cell contrast and the reconstructed
   patient-level raw-count contrast.
3. Align the study pseudobulks to each posteriorHCA model's own offset-zero
   TMMwsp reference.
4. Estimate patient-cohort latent `log(mu)` with the SAVI study's
   negative-binomial/Bayesian-bootstrap method.
5. Predict posteriorHCA `log(mu)` on a biologically admissible Normal/10x3
   covariate grid and marginalise with design-based weights (not atlas
   donor/profile frequencies).
6. Define a model-specific HV tumour-expression anchor and identify supported
   tissue × cell-type baselines whose posterior overlaps it.

**Colour / mark grammar.**

| Element | Encoding |
|---|---|
| HV / tumour / prototype candidate | vermilion `#D55E00` |
| UC | blue `#0072B2` |
| Paired/adjacent normal | green `#009E73` |
| Endothelial model | teal `#0072B2` |
| Epithelial model | vermilion `#D55E00` |
| Below prototype rule | charcoal `#8C8C8C` |
| Sex (female / male) | `RColorBrewer` PRGn |

**Panel export sizes (mm, width × height).**

| Tier | Size | Used for |
|---|---|---|
| Small (`sm`) | 85 × 80 | Volcano, violin |
| Medium (`md`) | 90 × 75 | Patient CPM, population bars/boxes |
| Large (`lg`) | 120 × 95 | Faceted tissue / sex forest panels |
| Heatmap | 120 × 85 | Tissue × age / ethnicity tiles |
| Full canvas | 183 × 95 | Cohort contrast forest only |

Figure panels follow Nature journal style: axis labels and legends only (no
plot titles or subtitles). **183 mm is the maximum patched canvas width**;
individual export panels are smaller so they can be assembled with
`patchwork`. Only designated full-width plots use the full 183 mm. Captions
carry the title.

------------------------------------------------------------------------

# Setup

```{r}
#| label: setup

suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(SummarizedExperiment)
  library(S4Vectors)
  library(DelayedArray)
  library(HDF5Array)
  library(brms)
  library(edgeR)
  library(posterior)
  library(targets)
  library(readxl)
  library(readr)
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(stringr)
  library(tibble)
  library(ggplot2)
  library(scales)
  library(knitr)
  library(RColorBrewer)
  library(digest)
})

options(stringsAsFactors = FALSE, scipen = 6, bitmapType = "cairo")
set.seed(params$seed)

font_cache <- file.path(tempdir(), "tm4sf1-font-cache")
dir.create(font_cache, recursive = TRUE, showWarnings = FALSE)
Sys.setenv(XDG_CACHE_HOME = font_cache)

# Nature double-column canvas cap and patchwork panel sizes (mm).
mm_to_in <- function(mm) mm / 25.4
fig_canvas_mm <- params$fig_width_mm
panel_mm <- list(
  sm = c(w = params$panel_sm_mm, h = 80),
  md = c(w = params$panel_md_mm, h = 75),
  lg = c(w = params$panel_lg_mm, h = 95),
  heatmap = c(w = params$panel_lg_mm, h = 85),
  full = c(w = fig_canvas_mm, h = 95)
)

first_existing <- function(paths, type = c("file", "dir")) {
  type <- match.arg(type)
  paths <- unique(paths[!is.na(paths) & nzchar(paths)])
  hit <- if (type == "file") paths[file.exists(paths)] else paths[dir.exists(paths)]
  if (!length(hit)) {
    return(NULL)
  }
  hit[[1]]
}

hpc_aliases <- function(path) {
  unique(c(
    path,
    sub("^/hpcfs/", "/scratchdata1/", path),
    sub("^/scratchdata1/", "/hpcfs/", path)
  ))
}

locate_case_dir <- function(path) {
  candidates <- unique(c(path, getwd(), file.path(getwd(), "dev", "TM4SF1_bladder")))
  hit <- candidates[file.exists(file.path(
    candidates, "data", "processed", "pseudobulk_tumour_epithelial_counts.rds"
  ))]
  normalizePath(hit[[1]], mustWork = TRUE)
}

locate_model_root <- function(path) {
  hit <- first_existing(hpc_aliases(path), type = "dir")
  normalizePath(hit, mustWork = TRUE)
}

locate_pseudobulk_h5ad <- function(path, current_path = NA_character_) {
  candidates <- unique(c(
    hpc_aliases(path),
    hpc_aliases(current_path),
    "/home/a1237163/lab/Mangiola_ImmuneAtlas/zenodo/pseudobulk_se.h5ad",
    "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/Mangiola_ImmuneAtlas/zenodo/pseudobulk_se.h5ad",
    "/scratchdata1/groups/phoenix-hpc-mangiola_laboratory/Mangiola_ImmuneAtlas/zenodo/pseudobulk_se.h5ad"
  ))
  hit <- first_existing(candidates, type = "file")
  if (is.null(hit)) {
    stop("Could not find the posteriorHCA pseudobulk H5AD.")
  }
  # Prefer a readable /hpcfs or home copy over a /scratchdata1 realpath.
  ordered <- unique(c(
    candidates[file.exists(candidates) & !startsWith(candidates, "/scratchdata1/")],
    hit
  ))
  ordered[[1]]
}

hdf5_seed_path <- function(x) {
  if (isS4(x) && methods::.hasSlot(x, "filepath")) {
    return(as.character(x@filepath)[[1]])
  }
  if (isS4(x) && methods::.hasSlot(x, "seed")) {
    return(hdf5_seed_path(x@seed))
  }
  NA_character_
}

relocate_hdf5_seed <- function(x, new_path) {
  if (isS4(x) && methods::.hasSlot(x, "filepath")) {
    x@filepath <- new_path
    return(x)
  }
  if (isS4(x) && methods::.hasSlot(x, "seed")) {
    x@seed <- relocate_hdf5_seed(x@seed, new_path)
  }
  x
}

relocate_se_hdf5 <- function(se, new_path) {
  se@assays@data@listData <- lapply(
    se@assays@data@listData,
    relocate_hdf5_seed,
    new_path = new_path
  )
  se
}

case_dir <- locate_case_dir(params$case_dir)
model_root <- locate_model_root(params$model_root)
results_dir <- file.path(case_dir, "results", "tm4sf1_toxicity_report_v2_design_standardised")
cache_dir <- file.path(results_dir, "cache")
v1_results_dir <- file.path(case_dir, params$v1_results_dir)
# Keep figure paths relative to the QMD working directory so Quarto can
# embed PNGs in the self-contained HTML (embed-resources: true).
figure_dir <- file.path("results", "tm4sf1_toxicity_report_v2_design_standardised", "figures")
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

knitr::opts_chunk$set(fig.path = paste0(figure_dir, "/"))

analysis_version <- "tm4sf1-logmu-v6-design-standardised-2026-08-18"
tm4_id <- params$gene_ensembl
tm4_symbol <- params$gene_symbol

fingerprint_files <- c(
  file.path(case_dir, "data", "processed", "pseudobulk_tumour_epithelial_counts.rds"),
  file.path(case_dir, "data", "processed", "pseudobulk_paired_normal_epithelial_counts.rds"),
  file.path(case_dir, "data", "raw", "paper", "41467_2025_59888_MOESM4_ESM.xlsx"),
  file.path(model_root, "_PAK2_TM4SF1_modelled_cell_types.csv"),
  file.path(model_root, paste0("V1_", params$model_cell_types), "_targets", "meta", "meta")
)
input_fingerprint <- digest::digest(
  setNames(
    vapply(
      fingerprint_files, digest::digest, character(1),
      algo = "sha256", file = TRUE, serialize = FALSE
    ),
    fingerprint_files
  ),
  algo = "sha256"
)
analysis_id <- paste(
  analysis_version,
  paste0("gene=", tm4_id),
  paste0("models=", paste(params$model_cell_types, collapse = ",")),
  paste0("n_boot=", params$n_boot),
  paste0("threshold_q=", params$threshold_quantile),
  paste0("alert_p=", params$alert_probability),
  paste0("min_donors=", params$min_donor_keys),
  paste0("seed=", params$seed),
  paste0("inputs=", input_fingerprint),
  sep = "|"
)

# Okabe–Ito / Nature-safe colours.
col_hv <- "#D55E00"
col_uc <- "#0072B2"
col_normal <- "#009E73"
col_candidate <- "#D55E00"
col_other <- "#8C8C8C"
col_dark <- "#3F3F3F"
colmap_histology <- c(HV = col_hv, UC = col_uc)
colmap_compartment <- c(
  tumour = col_hv,
  "paired/adjacent normal" = col_normal
)
colmap_cell <- c(epithelial = col_hv, endothelial = col_uc)
colmap_sex <- setNames(brewer.pal(3, "PRGn")[c(1, 3)], c("female", "male"))

theme_multipanel <-
  theme_bw() +
  theme(
    panel.border = element_blank(),
    axis.line = element_line(linewidth = 0.1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    strip.background = element_blank(),
    axis.title.y = element_text(margin = margin(0, 0, 0, 0), size = 7),
    axis.title.x = element_text(margin = margin(0, 0, 0, 0), size = 7),
    panel.spacing.x = unit(0.1, "lines"),
    axis.text.x = element_text(size = 6),
    axis.text.y = element_text(size = 6),
    strip.text.x = element_text(size = 7),
    strip.text.y = element_text(size = 7),
    legend.key.size = unit(4, "mm"),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 6),
    strip.clip = "off",
    plot.title = element_blank(),
    plot.subtitle = element_blank(),
    plot.tag = element_text(size = 8, face = "bold"),
    axis.line.x = element_line(linewidth = 0.2),
    axis.line.y = element_line(linewidth = 0.2),
    axis.ticks.x = element_line(linewidth = 0.2),
    axis.ticks.y = element_line(linewidth = 0.2),
    plot.margin = margin(2, 4, 2, 2)
  )

# Drop the leading zero on decimal tick labels (0.1 -> .1).
label_drop_leading_zero <- function(accuracy = NULL, ...) {
  function(x) {
    s <- number(x, accuracy = accuracy, big.mark = "", trim = TRUE, ...)
    s[!is.na(x) & x == 0] <- "0"
    sub("^(-?)0\\.", "\\1.", s)
  }
}

options(
  ggplot2.continuous.y = function() {
    scale_y_continuous(labels = label_drop_leading_zero())
  }
)

save_figure <- function(plot, filename, width_mm, height_mm) {
  stopifnot(width_mm <= fig_canvas_mm)
  stem <- tools::file_path_sans_ext(filename)
  pdf_path <- file.path(figure_dir, paste0(stem, ".pdf"))
  png_path <- file.path(figure_dir, paste0(stem, ".png"))
  pdf_device <- if (isTRUE(capabilities("cairo"))) cairo_pdf else "pdf"
  ggsave(
    pdf_path, plot,
    width = width_mm, height = height_mm, units = "mm",
    dpi = 600, device = pdf_device
  )
  ggsave(
    png_path, plot,
    width = width_mm, height = height_mm, units = "mm",
    dpi = 600
  )
  invisible(plot)
}
```

------------------------------------------------------------------------

# 1. Reproduce the paper's TM4SF1 highlight

```{r}
#| label: paper-source-data

paper_workbook <- file.path(
  case_dir, "data", "raw", "paper", "41467_2025_59888_MOESM4_ESM.xlsx"
)

paper_de <- read_excel(paper_workbook, sheet = "Figure5A") |>
  filter(cluster == "Variant") |>
  mutate(
    gene = as.character(gene),
    hv_enriched = avg_log2FC > 0,
    hv_rank = rank(
      if_else(hv_enriched, p_val, NA_real_),
      ties.method = "min",
      na.last = "keep"
    ),
    minus_log10_fdr = -log10(pmax(p_val_adj, .Machine$double.xmin)),
    is_tm4sf1 = gene == tm4_symbol
  )

tm4_paper_de <- paper_de |>
  filter(is_tm4sf1) |>
  transmute(
    source = "paper Figure5A",
    gene,
    avg_log2FC,
    pct_HV = pct.1,
    pct_UC = pct.2,
    p_value = p_val,
    adjusted_p_value = p_val_adj,
    HV_enriched_rank = as.integer(hv_rank)
  )

replication_de <- read_tsv(
  file.path(case_dir, "results", "replication_audit", "TM4SF1_original_DE_result.tsv"),
  show_col_types = FALSE
) |>
  transmute(
    analysis,
    n_HV_patients,
    n_UC_patients,
    cells_per_patient = n_cells_per_patient,
    avg_log2FC,
    p_value = p_val,
    adjusted_p_value = coalesce(FDR_BH, p_val_adj),
    HV_enriched_rank = HV_enriched_rank_by_p
  )

write_tsv(replication_de, file.path(results_dir, "TM4SF1_paper_DE_replication.tsv"))
```

The released table places TM4SF1 at rank `r tm4_paper_de$HV_enriched_rank`
among HV-enriched genes: log2 fold change
`r round(tm4_paper_de$avg_log2FC, 2)`, detection
`r percent(tm4_paper_de$pct_HV, accuracy = 0.1)` in HV versus
`r percent(tm4_paper_de$pct_UC, accuracy = 0.1)` in pure UC, and adjusted
(p = `r format(tm4_paper_de$adjusted_p_value, scientific = TRUE, digits = 3)`).
The closest public-DGE reconstruction also ranks TM4SF1 first, but it contains
9 HV and only 2 UC patients at the paper's 150-cell minimum because the
deposited UC01 matrix does not contain the published UC01 barcodes.

```{r}
#| label: paper-de-table

replication_de |>
  mutate(
    avg_log2FC = round(avg_log2FC, 2),
    p_value = format(p_value, scientific = TRUE, digits = 3),
    adjusted_p_value = format(adjusted_p_value, scientific = TRUE, digits = 3)
  ) |>
  kable(
    caption = "TM4SF1 in the released paper result and public-DGE sensitivity analyses."
  )
```

```{r}
#| label: paper-volcano
#| fig-width: 3.35
#| fig-height: 3.15
#| fig-cap: "Released Figure 5A differential-expression table. TM4SF1 is highlighted without recomputing the authors' private object. Extreme −log10 adjusted p values are capped at the 99.5th percentile for display."

volcano_cap <- quantile(paper_de$minus_log10_fdr, 0.995, na.rm = TRUE)
paper_de_display <- paper_de |>
  mutate(plot_y = pmin(minus_log10_fdr, volcano_cap))

paper_de_plot <- ggplot(paper_de_display, aes(avg_log2FC, plot_y)) +
  geom_point(colour = col_other, alpha = 0.45, size = 0.6, stroke = 0) +
  geom_point(
    data = filter(paper_de_display, is_tm4sf1),
    colour = col_hv, size = 1.6
  ) +
  geom_text(
    data = filter(paper_de_display, is_tm4sf1),
    aes(label = gene),
    colour = col_hv, nudge_x = 0.28, hjust = 0,
    size = 2, fontface = "italic"
  ) +
  geom_vline(xintercept = 0, colour = col_other, linewidth = 0.25) +
  coord_cartesian(clip = "off") +
  labs(
    x = "HV minus pure-UC average log2 fold change",
    y = expression(-log[10](adjusted~p))
  ) +
  theme_multipanel +
  theme(plot.margin = margin(2, 14, 2, 2))

save_figure(
  paper_de_plot, "01_paper_Figure5A_TM4SF1_volcano",
  panel_mm$sm["w"], panel_mm$sm["h"]
)
paper_de_plot
```

------------------------------------------------------------------------

# 2. Visualise TM4SF1 in the paper cohort

```{r}
#| label: paper-cell-expression

paper_cells <- read_excel(paper_workbook, sheet = "Figure5B") |>
  transmute(
    barcode = as.character(Barcode),
    paper_id = as.character(Name),
    source_expression = as.numeric(TM4SF1),
    histology_group = case_when(
      str_starts(paper_id, "VAR") ~ "HV",
      str_starts(paper_id, "UC") ~ "UC",
      TRUE ~ "other"
    )
  ) |>
  filter(histology_group %in% c("HV", "UC"))

paper_cell_patient <- paper_cells |>
  group_by(paper_id, histology_group) |>
  summarise(
    n_cells = n(),
    mean_source_expression = mean(source_expression),
    median_source_expression = median(source_expression),
    source_nonzero_fraction = mean(source_expression > 1e-8),
    .groups = "drop"
  )

write_tsv(
  paper_cell_patient,
  file.path(results_dir, "TM4SF1_paper_source_patient_summary.tsv")
)
```

The authors released normalized per-cell TM4SF1 values—not raw UMI counts—for
`r nrow(paper_cells)` tumour epithelial cells across
`r n_distinct(paper_cells$paper_id[paper_cells$histology_group == "HV"])` HV
and `r n_distinct(paper_cells$paper_id[paper_cells$histology_group == "UC"])`
UC patients. The cell distributions reproduce the figure, but they are
descriptive; patients, not cells, are the independent units. Raw DGE counts are
used for detection and cohort modelling below.

```{r}
#| label: paper-cell-violin
#| fig-width: 3.35
#| fig-height: 3.15
#| fig-cap: "Released Figure 5B TM4SF1 values. The violin is cell-level source data; filled points are patient means."

paper_cell_plot <- ggplot(
  paper_cells,
  aes(histology_group, source_expression, fill = histology_group)
) +
  geom_violin(
    scale = "width", trim = TRUE,
    colour = col_dark, linewidth = 0.2, alpha = 0.65
  ) +
  geom_boxplot(
    width = 0.14, outlier.shape = NA,
    fill = "white", colour = col_dark, linewidth = 0.25
  ) +
  geom_point(
    data = paper_cell_patient,
    inherit.aes = FALSE,
    aes(histology_group, mean_source_expression),
    position = position_jitter(width = 0.06, height = 0),
    shape = 21, fill = col_dark, colour = "white", size = 1.3, stroke = 0.25
  ) +
  scale_fill_manual(values = colmap_histology) +
  labs(x = NULL, y = "Released TM4SF1 expression value") +
  theme_multipanel +
  guides(fill = "none")

save_figure(
  paper_cell_plot, "02_paper_Figure5B_TM4SF1_cells",
  panel_mm$sm["w"], panel_mm$sm["h"]
)
paper_cell_plot
```

```{r}
#| label: public-dge-patient-contrast
#| fig-width: 3.54
#| fig-height: 2.95
#| fig-cap: "Patient-level raw-count pseudobulk contrast reconstructed from the deposited DGE matrices. UC01 is retained but marked because it has only 90 reconstructed tumour epithelial cells. VAR08 contributes most HV molecules; several HV tumours are low."

tm4_patient <- read_tsv(
  file.path(case_dir, "data", "metadata", "TM4SF1_patient_summary.tsv"),
  show_col_types = FALSE
) |>
  filter(
    sample_type == "tumour",
    broad_cell_type == "epithelial",
    paper_id %in% c(sprintf("VAR%02d", 1:9), sprintf("UC%02d", 1:3))
  ) |>
  mutate(
    cohort_status = if_else(
      reconstructed_n150_threshold_met,
      ">=150 reconstructed cells",
      "<150 reconstructed cells"
    ),
    paper_id = factor(paper_id, levels = paper_id[order(CPM)])
  )

patient_cpm_plot <- ggplot(
  tm4_patient,
  aes(paper_id, CPM, colour = histology_group, shape = cohort_status)
) +
  geom_point(size = 1.6) +
  scale_y_log10(labels = label_number()) +
  scale_colour_manual(values = colmap_histology, name = "Histology") +
  scale_shape_manual(values = c(
    ">=150 reconstructed cells" = 16,
    "<150 reconstructed cells" = 1
  ), name = "Public-DGE cohort") +
  coord_flip() +
  labs(x = NULL, y = "TM4SF1 CPM") +
  theme_multipanel

save_figure(
  patient_cpm_plot, "03_public_DGE_TM4SF1_patient_CPM",
  panel_mm$md["w"], panel_mm$md["h"]
)
patient_cpm_plot
```

## Which broad population carries TM4SF1?

```{r}
#| label: broad-population-source
#| fig-width: 3.54
#| fig-height: 2.95
#| fig-cap: "Pooled TM4SF1 abundance and detection across broad reconstructed populations. Labels give the fraction of cells with non-zero counts. Endothelium is a key cell class for off-tumour follow-up."

population_source <- read_tsv(
  file.path(case_dir, "data", "metadata", "TM4SF1_population_source_summary.tsv"),
  show_col_types = FALSE
) |>
  mutate(
    sample_type = recode(
      sample_type,
      paired_normal = "paired/adjacent normal",
      tumour = "tumour"
    )
  )

population_plot <- ggplot(
  population_source,
  aes(broad_cell_type, pooled_CPM, fill = sample_type)
) +
  geom_col(position = position_dodge(width = 0.75), width = 0.68, colour = NA) +
  geom_text(
    aes(label = percent(pooled_detection_rate, accuracy = 1)),
    position = position_dodge(width = 0.75),
    vjust = -0.35, size = 2, colour = col_dark
  ) +
  scale_y_continuous(
    labels = label_number(),
    expand = expansion(mult = c(0, 0.16))
  ) +
  scale_fill_manual(values = colmap_compartment, name = "Study compartment") +
  labs(x = NULL, y = "Pooled TM4SF1 CPM") +
  theme_multipanel

save_figure(
  population_plot, "04_TM4SF1_broad_population_source",
  panel_mm$md["w"], panel_mm$md["h"]
)
population_plot
```

Because pooled CPM is composition-weighted, the same check is shown at patient
level:

```{r}
#| label: broad-population-patient-distribution
#| fig-width: 3.54
#| fig-height: 2.95
#| fig-cap: "Patient-level TM4SF1 CPM by broad population and study compartment. Points are patient pseudobulks; low-cell profiles remain descriptive."

population_patient <- read_tsv(
  file.path(case_dir, "data", "metadata", "TM4SF1_patient_summary.tsv"),
  show_col_types = FALSE
) |>
  filter(broad_cell_type %in% c("epithelial", "endothelial", "stromal", "immune")) |>
  mutate(
    sample_type = recode(
      sample_type,
      paired_normal = "paired/adjacent normal",
      tumour = "tumour"
    )
  )

population_patient_plot <- ggplot(
  population_patient,
  aes(broad_cell_type, CPM, colour = sample_type)
) +
  geom_boxplot(
    aes(group = interaction(broad_cell_type, sample_type)),
    outlier.shape = NA, width = 0.55, linewidth = 0.25, alpha = 0.15,
    position = position_dodge(width = 0.65)
  ) +
  geom_point(
    position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.65),
    alpha = 0.8, size = 1.1, stroke = 0.2
  ) +
  scale_y_continuous(trans = transform_log1p(), labels = label_number()) +
  scale_colour_manual(values = colmap_compartment, name = "Study compartment") +
  labs(x = NULL, y = "Patient TM4SF1 CPM (log1p)") +
  theme_multipanel

save_figure(
  population_patient_plot,
  "04b_TM4SF1_broad_population_patient_CPM",
  panel_mm$md["w"], panel_mm$md["h"]
)
population_patient_plot
```

Together, the pooled and patient-level checks motivate the comparator strategy.
Epithelium is the closest cell-class-matched comparator for tumour epithelium,
while endothelium must be analysed separately because its expression may
represent a vascular off-tumour exposure. Pooling these classes would hide the
relevant biology.

------------------------------------------------------------------------

# 3. Cohort-level `log(mu)` and posteriorHCA baseline

The helper functions below remain in the report. They implement one chain from
raw patient counts to model-aligned `log(mu)`.

## Study-side helpers

```{r}
#| label: logmu-helpers-study

# Collapse feature rows that share an Ensembl id.
aggregate_ensembl_counts <- function(se) {
  counts <- as.matrix(assay(se, "counts"))
  ensembl <- as.character(rowData(se)$ensembl_id)
  keep <- !is.na(ensembl) & nzchar(ensembl)
  out <- rowsum(counts[keep, , drop = FALSE], ensembl[keep], reorder = FALSE)
  storage.mode(out) <- "numeric"
  out
}

read_study_pseudobulk <- function(case_dir) {
  tumour_se <- readRDS(file.path(
    case_dir, "data", "processed", "pseudobulk_tumour_epithelial_counts.rds"
  ))
  normal_se <- readRDS(file.path(
    case_dir, "data", "processed",
    "pseudobulk_paired_normal_epithelial_counts.rds"
  ))

  tumour_counts <- aggregate_ensembl_counts(tumour_se)
  normal_counts <- aggregate_ensembl_counts(normal_se)

  tumour_meta <- as.data.frame(colData(tumour_se)) |>
    rownames_to_column("source_column") |>
    mutate(
      analysis_group = histology_group,
      study_state = "tumour epithelial",
      include_for_logmu = TRUE
    )
  normal_meta <- as.data.frame(colData(normal_se)) |>
    rownames_to_column("source_column") |>
    mutate(
      analysis_group = "paired_normal",
      study_state = "paired-normal epithelial",
      include_for_logmu = n_cells >= 20
    )

  common <- intersect(rownames(tumour_counts), rownames(normal_counts))
  normal_keep <- normal_meta$include_for_logmu
  counts <- cbind(
    tumour_counts[common, , drop = FALSE],
    normal_counts[common, normal_keep, drop = FALSE]
  )
  meta <- bind_rows(tumour_meta, normal_meta[normal_keep, , drop = FALSE]) |>
    mutate(
      analysis_column = c(
        paste0("tumour__", tumour_meta$patient_id),
        paste0("paired_normal_epithelial__", normal_meta$patient_id[normal_keep])
      ),
      strict_n150 = study_state != "tumour epithelial" |
        reconstructed_n150_threshold_met
    )

  colnames(counts) <- meta$analysis_column
  rownames(meta) <- meta$analysis_column
  list(counts = counts, metadata = meta)
}

# TMM against the model's own offset-zero reference sample.
align_to_hca_reference <- function(study_counts, reference_counts, reference_name) {
  shared <- intersect(rownames(study_counts), names(reference_counts))
  combined <- cbind(
    study_counts[shared, , drop = FALSE],
    setNames(reference_counts[shared], shared)
  )
  colnames(combined)[ncol(combined)] <- reference_name

  norm_factors <- calcNormFactors(
    combined, refColumn = ncol(combined), method = "TMMwsp"
  )
  library_size <- colSums(combined)
  effective_size <- library_size * norm_factors
  multiplier <- effective_size[[reference_name]] / effective_size
  offset <- log(1 / multiplier)

  list(
    counts = combined[, colnames(study_counts), drop = FALSE],
    offset = offset[colnames(study_counts)],
    multiplier = multiplier[colnames(study_counts)],
    shared_features = shared
  )
}

fit_shared_dispersion <- function(counts, metadata, offset, feature_id) {
  design <- model.matrix(~ 0 + factor(metadata$analysis_group))
  dge <- DGEList(counts = counts)
  dge$offset <- matrix(offset, nrow(counts), ncol(counts), byrow = TRUE)

  keep <- filterByExpr(dge, design = design)
  keep[match(feature_id, rownames(dge))] <- TRUE
  dge <- dge[keep, , keep.lib.sizes = FALSE]
  dge <- estimateDisp(dge, design, robust = TRUE)
  invisible(glmQLFit(dge, design, robust = TRUE))

  dispersion <- if (!is.null(dge$trended.dispersion)) {
    dge$trended.dispersion
  } else if (!is.null(dge$tagwise.dispersion)) {
    dge$tagwise.dispersion
  } else {
    rep_len(dge$common.dispersion, nrow(dge))
  }
  names(dispersion) <- rownames(dge)
  unname(dispersion[[feature_id]])
}

draw_dirichlet_weights <- function(n) {
  weights <- rexp(n)
  n * weights / sum(weights)
}

bootstrap_cohort_logmu <- function(counts, offset, dispersion, columns, n_boot, seed) {
  y <- counts[tm4_id, columns, drop = FALSE]
  o <- offset[columns]
  set.seed(seed)
  vapply(seq_len(n_boot), function(i) {
    mglmOneGroup(
      y, offset = o, dispersion = dispersion,
      weights = draw_dirichlet_weights(length(columns))
    )[[1]]
  }, numeric(1))
}

pair_log2_contrast <- function(x, y, n = 2000L, seed = 1L) {
  set.seed(seed)
  delta <- (sample(x, n, replace = TRUE) - sample(y, n, replace = TRUE)) / log(2)
  tibble(
    estimate_log2 = median(delta),
    lower_log2 = quantile(delta, 0.025),
    upper_log2 = quantile(delta, 0.975),
    probability_positive = mean(delta > 0)
  )
}
```

## posteriorHCA helpers (design-standardised baseline)

```{r}
#| label: logmu-helpers-hca

read_hca_bundle <- function(cell_type, model_root, feature_id) {
  store <- file.path(model_root, paste0("V1_", cell_type), "_targets")
  registry <- read.csv(
    file.path(model_root, "_PAK2_TM4SF1_modelled_cell_types.csv"),
    check.names = FALSE
  )
  target_name <- registry |>
    filter(
      cell_type == paste0("V1_", .env$cell_type),
      ensembl_id == feature_id,
      isTRUE(in_feature_df) | in_feature_df
    ) |>
    pull(estimates_branch) |>
    unique()

  chunk <- targets::tar_read_raw(target_name[[1]], store = store)
  fit <- chunk$brms_fit[[match(feature_id, chunk$.feature)]]
  pseudobulk <- targets::tar_read_raw("pseudobulk_sample", store = store)

  current_h5ad <- hdf5_seed_path(pseudobulk@assays@data@listData[["counts"]])
  h5ad <- locate_pseudobulk_h5ad(params$pseudobulk_h5ad, current_h5ad)
  pseudobulk <- relocate_se_hdf5(pseudobulk, h5ad)
  sample_meta <- as.data.frame(colData(pseudobulk)) |>
    rownames_to_column("sample_column")
  counts <- assay(pseudobulk, "counts")

  reference_index <- which(sample_meta$offset == 0)
  if (length(reference_index) != 1L) {
    stop(
      "Expected exactly one offset==0 reference sample in ", cell_type,
      "; found ", length(reference_index)
    )
  }
  reference_name <- sample_meta$sample_column[[reference_index]]
  reference_counts <- as.numeric(as.matrix(counts[, reference_index, drop = FALSE]))
  names(reference_counts) <- rownames(pseudobulk)

  profiles <- sample_meta |>
    transmute(
      sample_column,
      dataset_id = as.character(dataset_id),
      donor_id = as.character(donor_id),
      tissue = as.character(tissue),
      tissue_groups = as.character(tissue_groups),
      age_decade = as.character(age_decade),
      sex = as.character(sex),
      ethnicity_groups = as.character(ethnicity_groups),
      disease_group = as.character(disease_groups),
      assay_group = as.character(assay_groups)
    ) |>
    filter(
      disease_group == "Normal",
      assay_group == "10x Genomics 3",
      if_all(
        c(dataset_id, donor_id, tissue_groups, age_decade, sex, ethnicity_groups),
        ~ !is.na(.x) & nzchar(.x)
      )
    ) |>
    mutate(donor_key = paste(dataset_id, donor_id, sep = "___")) |>
    distinct(
      donor_key, tissue_groups, age_decade, sex, ethnicity_groups,
      .keep_all = TRUE
    )

  fit_draws <- posterior::as_draws_array(fit)
  draw_diagnostics <- posterior::summarise_draws(fit_draws)
  nuts <- brms::nuts_params(fit)
  finite_max <- function(x) {
    x <- x[is.finite(x)]
    if (length(x)) max(x) else NA_real_
  }
  finite_min <- function(x) {
    x <- x[is.finite(x)]
    if (length(x)) min(x) else NA_real_
  }

  max_rhat_value <- finite_max(draw_diagnostics$rhat)
  min_ess_bulk_value <- finite_min(draw_diagnostics$ess_bulk)
  min_ess_tail_value <- finite_min(draw_diagnostics$ess_tail)
  divergence_count <- sum(
    nuts$Parameter == "divergent__" & nuts$Value == 1,
    na.rm = TRUE
  )

  diagnostics <- tibble(
    cell_type,
    n_model_rows = nrow(fit$data),
    n_posterior_draws = posterior::ndraws(fit_draws),
    max_rhat = max_rhat_value,
    min_ess_bulk = min_ess_bulk_value,
    min_ess_tail = min_ess_tail_value,
    divergences = divergence_count,
    readiness_policy = "Rhat<=1.01; bulk/tail ESS>=400; zero divergences",
    model_ready = all(is.finite(c(
      max_rhat_value, min_ess_bulk_value, min_ess_tail_value, divergence_count
    ))) &
      max_rhat_value <= 1.01 &
      min_ess_bulk_value >= 400 &
      min_ess_tail_value >= 400 &
      divergence_count == 0
  )

  list(
    fit = fit,
    profiles = profiles,
    reference_counts = reference_counts,
    reference_name = reference_name,
    model_store = store,
    target_name = target_name[[1]],
    diagnostics = diagnostics
  )
}

posterior_hca_logmu <- function(fit, newdata, seed) {
  set.seed(seed)
  brms::posterior_linpred(
    fit,
    newdata = newdata,
    transform = FALSE,
    re_formula = NULL,
    allow_new_levels = TRUE,
    sample_new_levels = "gaussian"
  )
}

hard_sex_constraint <- c(
  "female reproductive system" = "female",
  "male reproductive system" = "male",
  "prostate" = "male",
  "breast" = "female"
)

both_sex_tissues <- c(
  "respiratory system",
  "integumentary system (skin)",
  "renal system"
)

excluded_ethnicity_levels <- "Other/Unknown"

allowed_sexes_for_tissue <- function(tissue, profiles) {
  if (tissue %in% names(hard_sex_constraint)) {
    return(hard_sex_constraint[[tissue]])
  }
  if (tissue %in% both_sex_tissues) {
    return(c("female", "male"))
  }
  sort(unique(profiles$sex[profiles$tissue_groups == tissue]))
}

reports_both_sexes <- function(tissue, profiles) {
  if (tissue %in% names(hard_sex_constraint)) {
    return(FALSE)
  }
  allowed <- allowed_sexes_for_tissue(tissue, profiles)
  length(allowed) >= 2L && all(c("female", "male") %in% allowed)
}

profile_support_counts <- function(profiles) {
  profiles |>
    group_by(tissue_groups, sex, age_decade, ethnicity_groups) |>
    summarise(
      n_donor_keys = n_distinct(donor_key),
      n_profile_rows = n(),
      .groups = "drop"
    )
}

build_tissue_sex_rules <- function(profiles) {
  sort(unique(profiles$tissue_groups)) |>
    map_dfr(function(tissue) {
      allowed <- allowed_sexes_for_tissue(tissue, profiles)
      tibble(
        tissue_groups = tissue,
        allowed_sex = paste(allowed, collapse = ", "),
        n_allowed_sex_levels = length(allowed),
        reports_tissue_sex = reports_both_sexes(tissue, profiles),
        rule_source = case_when(
          tissue %in% names(hard_sex_constraint) ~ "hard biological constraint",
          tissue %in% both_sex_tissues ~ "prespecified both-sex tissue",
          length(allowed) >= 2L ~ "atlas-observed both sexes",
          TRUE ~ "atlas-observed single sex"
        )
      )
    })
}

build_design_grid <- function(profiles, min_donors) {
  support <- profile_support_counts(profiles)
  sex_rules <- build_tissue_sex_rules(profiles)

  support |>
    left_join(sex_rules, by = "tissue_groups") |>
    rowwise() |>
    mutate(
      structurally_valid = sex %in% strsplit(allowed_sex, ", ", fixed = TRUE)[[1]],
      exclusion_reason = case_when(
        !structurally_valid ~ "structurally invalid sex × tissue",
        ethnicity_groups == excluded_ethnicity_levels ~ "excluded ethnicity level",
        n_donor_keys == 0L ~ "no atlas support",
        n_donor_keys < min_donors ~ "sparse joint cell (included in design weights)",
        TRUE ~ "included in primary marginalisation"
      ),
      included = structurally_valid &
        ethnicity_groups != excluded_ethnicity_levels &
        n_donor_keys > 0L
    ) |>
    ungroup()
}

design_cells_to_newdata <- function(cells) {
  cells |>
    transmute(
      counts = 1,
      age_decade,
      sex,
      disease_groups_altered = "Normal",
      ethnicity_groups,
      assay_groups_altered = "10x Genomics 3",
      dataset_id_altered = "__TM4SF1_new_study__",
      tissue_groups,
      offset = 0
    )
}

mean_draw_rows <- function(logmu, row_index) {
  if (length(row_index) == 1L) {
    return(logmu[, row_index])
  }
  rowMeans(logmu[, row_index, drop = FALSE])
}

nested_marginal_draws <- function(logmu, cells, fix_cols) {
  included <- dplyr::filter(cells, included)
  if (!nrow(included)) {
    return(numeric())
  }
  if (length(fix_cols)) {
    included <- included |>
      mutate(across(all_of(fix_cols), as.character)) |>
      group_by(across(all_of(fix_cols))) |>
      group_split()
    map(included, function(group_cells) {
      nested_marginal_draws(logmu, group_cells, character())
    })
  } else {
    marginal_cols <- c("sex", "age_decade", "ethnicity_groups")
    sex_levels <- sort(unique(group_cells <- cells)$sex)
    sex_draws <- map(sex_levels, function(sex_level) {
      sex_cells <- dplyr::filter(cells, sex == sex_level)
      age_levels <- sort(unique(sex_cells$age_decade))
      age_draws <- map(age_levels, function(age_level) {
        age_cells <- dplyr::filter(sex_cells, age_decade == age_level)
        eth_levels <- sort(unique(age_cells$ethnicity_groups))
        eth_draws <- map(eth_levels, function(eth_level) {
          idx <- age_cells$row_idx[age_cells$ethnicity_groups == eth_level]
          mean_draw_rows(logmu, idx)
        })
        Reduce(`+`, eth_draws) / length(eth_draws)
      })
      Reduce(`+`, age_draws) / length(age_draws)
    })
    Reduce(`+`, sex_draws) / length(sex_draws)
  }
}

tissue_baseline_draw <- function(logmu, cells, tissue) {
  sub <- dplyr::filter(cells, tissue_groups == tissue, included)
  if (!nrow(sub)) {
    return(numeric())
  }
  nested_marginal_draws(logmu, sub, character())
}

compute_marginalisation_weights <- function(cells, tissue = NULL) {
  sub <- if (is.null(tissue)) {
    dplyr::filter(cells, included)
  } else {
    dplyr::filter(cells, tissue_groups == tissue, included)
  }
  if (!nrow(sub)) {
    return(tibble())
  }

  sex_levels <- sort(unique(sub$sex))
  n_sex <- length(sex_levels)
  map_dfr(sex_levels, function(sex_level) {
    sex_cells <- dplyr::filter(sub, sex == sex_level)
    age_levels <- sort(unique(sex_cells$age_decade))
    n_age <- length(age_levels)
    map_dfr(age_levels, function(age_level) {
      age_cells <- dplyr::filter(sex_cells, age_decade == age_level)
      eth_levels <- sort(unique(age_cells$ethnicity_groups))
      n_eth <- length(eth_levels)
      map_dfr(eth_levels, function(eth_level) {
        cell <- dplyr::filter(
          age_cells,
          ethnicity_groups == eth_level
        )
        tibble(
          tissue_groups = cell$tissue_groups[[1]],
          sex = sex_level,
          age_decade = age_level,
          ethnicity_groups = eth_level,
          n_donor_keys = cell$n_donor_keys[[1]],
          marginal_weight = 1 / (n_sex * n_age * n_eth),
          weight_rule = paste0(
            "equal sex (1/", n_sex, ") × equal age (1/", n_age,
            ") × equal ethnicity (1/", n_eth, ")"
          )
        )
      })
    })
  })
}

summarise_design_draw <- function(
    logmu, hv_draws, threshold, seed,
    dimension, cells, fix_cols, min_donors, profiles
) {
  if (length(fix_cols)) {
    key <- cells |>
      mutate(across(all_of(fix_cols), as.character)) |>
      group_by(across(all_of(fix_cols))) |>
      group_split()
  } else {
    key <- list(cells)
  }

  map2_dfr(key, seq_along(key), function(group_cells, group_number) {
    if (identical(dimension, "comprehensive")) {
      tissues <- sort(unique(group_cells$tissue_groups))
      tissue_draws <- map(tissues, ~ tissue_baseline_draw(logmu, group_cells, .x))
      tissue_draws <- tissue_draws[lengths(tissue_draws) > 0L]
      if (!length(tissue_draws)) {
        return(tibble())
      }
      hca_draws <- Reduce(`+`, tissue_draws) / length(tissue_draws)
      tissue <- NA_character_
      sex <- NA_character_
      age_decade <- NA_character_
      ethnicity_groups <- NA_character_
      n_profile_rows <- sum(group_cells$included)
      n_donor_keys <- profiles |>
        semi_join(
          group_cells |> filter(included),
          by = c("tissue_groups", "sex", "age_decade", "ethnicity_groups")
        ) |>
        pull(donor_key) |>
        n_distinct()
    } else {
      hca_draws <- nested_marginal_draws(logmu, group_cells, character())
      if (!length(hca_draws)) {
        return(tibble())
      }
      tissue <- if ("tissue_groups" %in% fix_cols) {
        as.character(group_cells$tissue_groups[[1]])
      } else {
        NA_character_
      }
      sex <- if ("sex" %in% fix_cols) as.character(group_cells$sex[[1]]) else NA_character_
      age_decade <- if ("age_decade" %in% fix_cols) {
        as.character(group_cells$age_decade[[1]])
      } else {
        NA_character_
      }
      ethnicity_groups <- if ("ethnicity_groups" %in% fix_cols) {
        as.character(group_cells$ethnicity_groups[[1]])
      } else {
        NA_character_
      }
      n_profile_rows <- sum(group_cells$included)
      n_donor_keys <- profiles |>
        semi_join(
          group_cells |> filter(included),
          by = c("tissue_groups", "sex", "age_decade", "ethnicity_groups")
        ) |>
        pull(donor_key) |>
        n_distinct()
    }

    paired <- pair_log2_contrast(
      hca_draws, hv_draws,
      n = max(length(hv_draws), length(hca_draws)),
      seed = seed + group_number
    )

    tibble(
      dimension,
      tissue_groups = tissue,
      sex = sex,
      age_decade = age_decade,
      ethnicity_groups = ethnicity_groups,
      n_profile_rows = n_profile_rows,
      n_donor_keys = n_donor_keys,
      n_admissible_cells = sum(group_cells$included),
      median_log_mu = median(hca_draws),
      lower_log_mu = quantile(hca_draws, 0.025),
      upper_log_mu = quantile(hca_draws, 0.975),
      threshold_log_mu = threshold,
      probability_above_threshold = mean(hca_draws >= threshold),
      probability_hca_ge_random_hv = mean(ecdf(hv_draws)(hca_draws)),
      estimate_log2_vs_hv = paired$estimate_log2,
      lower_log2_vs_hv = paired$lower_log2,
      upper_log2_vs_hv = paired$upper_log2,
      supported = dimension == "comprehensive" | n_donor_keys >= min_donors
    )
  })
}

summarise_design_groups <- function(
    logmu, cells, profiles, dimension, fix_cols, hv_draws,
    threshold, min_donors, seed
) {
  if (identical(dimension, "tissue_sex")) {
    both_sex_tissues <- build_tissue_sex_rules(profiles) |>
      filter(reports_tissue_sex) |>
      pull(tissue_groups)
    cells <- dplyr::filter(cells, tissue_groups %in% both_sex_tissues)
  }

  summarise_design_draw(
    logmu, hv_draws, threshold, seed,
    dimension, cells, fix_cols, min_donors, profiles
  )
}

compute_tissue_rule_sensitivity <- function(
    logmu, cells, profiles, hv_draws, cell_type,
    threshold_quantiles = c(0.01, 0.05, 0.10),
    minimum_donor_keys = c(5L, 10L, 20L),
    alert_probabilities = c(0.50, 0.80, 0.95)
) {
  tissues <- sort(unique(cells$tissue_groups))
  tissue_draws <- setNames(
    map(tissues, ~ tissue_baseline_draw(logmu, cells, .x)),
    tissues
  )
  tissue_support <- map_int(tissues, function(tissue) {
    profiles |>
      semi_join(
        dplyr::filter(cells, tissue_groups == tissue, included),
        by = c("tissue_groups", "sex", "age_decade", "ethnicity_groups")
      ) |>
      pull(donor_key) |>
      n_distinct()
  })

  crossing(
    threshold_quantile = threshold_quantiles,
    minimum_donor_keys = minimum_donor_keys,
    alert_probability_cutoff = alert_probabilities
  ) |>
    pmap_dfr(function(
        threshold_quantile, minimum_donor_keys, alert_probability_cutoff
    ) {
      threshold <- unname(quantile(hv_draws, threshold_quantile))
      exceedance <- map_dbl(tissue_draws, ~ mean(.x >= threshold))
      tibble(
        cell_type,
        threshold_quantile,
        threshold_log_mu = threshold,
        minimum_donor_keys,
        alert_probability_cutoff,
        supported_tissues = sum(tissue_support >= minimum_donor_keys),
        prototype_tissue_candidates = sum(
          tissue_support >= minimum_donor_keys &
            exceedance >= alert_probability_cutoff
        )
      )
    })
}
```

## One cell-type model

```{r}
#| label: run-one-model-helper

run_one_cell_type <- function(cell_type, study, params, model_root) {
  bundle <- read_hca_bundle(cell_type, model_root, tm4_id)
  aligned <- align_to_hca_reference(
    study$counts, bundle$reference_counts, bundle$reference_name
  )
  dispersion <- fit_shared_dispersion(
    aligned$counts, study$metadata, aligned$offset, tm4_id
  )

  groups <- list(
    "HV tumour epithelial" = which(
      study$metadata$study_state == "tumour epithelial" &
        study$metadata$histology_group == "HV"
    ),
    "UC tumour epithelial (3 published IDs)" = which(
      study$metadata$study_state == "tumour epithelial" &
        study$metadata$histology_group == "UC"
    ),
    "UC tumour epithelial (2 public-DGE >=150)" = which(
      study$metadata$study_state == "tumour epithelial" &
        study$metadata$histology_group == "UC" &
        study$metadata$strict_n150
    ),
    "paired-normal epithelial (>=20 cells)" = which(
      study$metadata$study_state == "paired-normal epithelial"
    )
  )

  study_draws <- imap_dfr(groups, function(index, group_name) {
    draws <- bootstrap_cohort_logmu(
      aligned$counts,
      aligned$offset,
      dispersion,
      colnames(aligned$counts)[index],
      params$n_boot,
      params$seed + which(names(groups) == group_name)
    )
    tibble(
      cell_type, group = group_name,
      draw = seq_along(draws), log_mu = draws,
      n_patients = length(index)
    )
  })

  hv_draws <- study_draws |>
    filter(group == "HV tumour epithelial") |>
    pull(log_mu)
  threshold <- unname(quantile(hv_draws, params$threshold_quantile))

  design_grid <- build_design_grid(bundle$profiles, params$min_donor_keys)
  tissue_sex_rules <- build_tissue_sex_rules(bundle$profiles)
  prediction_cells <- dplyr::filter(design_grid, included) |>
    distinct(tissue_groups, sex, age_decade, ethnicity_groups, .keep_all = TRUE)
  prediction_newdata <- design_cells_to_newdata(prediction_cells)

  hca_logmu <- posterior_hca_logmu(
    bundle$fit, prediction_newdata,
    seed = params$seed + match(cell_type, params$model_cell_types) * 100L
  )
  cells <- prediction_cells |>
    mutate(row_idx = row_number(), included = TRUE)

  group_specs <- list(
    comprehensive = list(vars = character(), seed = params$seed + 1000L),
    tissue = list(vars = "tissue_groups", seed = params$seed + 2000L),
    tissue_sex = list(vars = c("tissue_groups", "sex"), seed = params$seed + 3000L),
    tissue_age = list(
      vars = c("tissue_groups", "age_decade"), seed = params$seed + 4000L
    ),
    tissue_ethnicity = list(
      vars = c("tissue_groups", "ethnicity_groups"), seed = params$seed + 5000L
    )
  )

  profile_summary <- imap_dfr(group_specs, function(spec, dimension) {
    summarise_design_groups(
      hca_logmu, cells, bundle$profiles, dimension, spec$vars, hv_draws,
      threshold, params$min_donor_keys, spec$seed
    )
  }) |>
    mutate(
      cell_type = cell_type,
      baseline_method = params$baseline_method,
      model_ready = bundle$diagnostics$model_ready[[1]],
      interpretable_for_demographic_alert = !(
        dimension == "tissue_ethnicity" & ethnicity_groups == "Other/Unknown"
      ),
      prototype_overlap_candidate = supported &
        interpretable_for_demographic_alert &
        probability_above_threshold >= params$alert_probability,
      expression_overlap_alert = prototype_overlap_candidate & model_ready,
      .before = 1
    )

  marginalisation_weights <- map_dfr(
    sort(unique(cells$tissue_groups)),
    ~ compute_marginalisation_weights(cells, .x) |> mutate(cell_type = cell_type)
  )

  level_audit <- design_grid |>
    mutate(cell_type = cell_type) |>
    select(
      cell_type, tissue_groups, sex, age_decade, ethnicity_groups,
      n_donor_keys, structurally_valid, exclusion_reason, included
    )

  hca_comprehensive <- {
    tissues <- sort(unique(cells$tissue_groups))
    tissue_draws <- map(tissues, ~ tissue_baseline_draw(hca_logmu, cells, .x))
    tissue_draws <- tissue_draws[lengths(tissue_draws) > 0L]
    if (length(tissue_draws)) {
      Reduce(`+`, tissue_draws) / length(tissue_draws)
    } else {
      numeric()
    }
  }

  hca_renal <- tissue_baseline_draw(hca_logmu, cells, "renal system")

  rule_sensitivity <- compute_tissue_rule_sensitivity(
    hca_logmu, cells, bundle$profiles, hv_draws, cell_type
  ) |>
    mutate(
      model_ready = bundle$diagnostics$model_ready[[1]],
      analysis_id = analysis_id,
      .after = cell_type
    )

  contrast_draws <- study_draws
  if (cell_type != "epithelial") {
    contrast_draws <- filter(
      contrast_draws, group != "paired-normal epithelial (>=20 cells)"
    )
  }

  cohort_contrasts <- contrast_draws |>
    group_split(group) |>
    map_dfr(function(group_data) {
      comparison_draws <- hca_comprehensive
      comparator <- paste0(
        "posteriorHCA Normal/10x3-supported comprehensive ", cell_type
      )
      if (
        cell_type == "epithelial" &&
          unique(group_data$group) == "paired-normal epithelial (>=20 cells)" &&
          length(hca_renal)
      ) {
        comparison_draws <- hca_renal
        comparator <- paste(
          "posteriorHCA Normal/10x3-supported renal-system",
          "epithelial proxy"
        )
      }
      pair_log2_contrast(
        group_data$log_mu, comparison_draws,
        n = params$n_boot,
        seed = params$seed + 6000L + match(unique(group_data$group), names(groups))
      ) |>
        mutate(
          cell_type,
          study_group = unique(group_data$group),
          comparator,
          n_study_patients = unique(group_data$n_patients),
          .before = 1
        )
    })

  threshold_registry <- tibble(
    gene_symbol = tm4_symbol,
    ensembl_id = tm4_id,
    cell_type_model = cell_type,
    model_store = bundle$model_store,
    model_target = bundle$target_name,
    normalization_reference = bundle$reference_name,
    normalization_method = "TMMwsp; reference-anchored offset=log(1/multiplier)",
    study_biological_unit = "patient",
    tumour_expression_anchor = "9-patient HV tumour epithelial cohort location",
    threshold_quantile = params$threshold_quantile,
    threshold_log_mu = threshold,
    n_patient_bootstrap_draws = params$n_boot,
    random_seed = params$seed,
    alert_probability_cutoff = params$alert_probability,
    minimum_donor_keys = params$min_donor_keys,
    model_ready = bundle$diagnostics$model_ready[[1]],
    diagnostic_gate = bundle$diagnostics$readiness_policy[[1]],
    input_and_model_fingerprint = input_fingerprint,
    shared_feature_count = length(aligned$shared_features),
    shared_feature_sha256 = digest::digest(sort(aligned$shared_features), algo = "sha256"),
    profile_weighting = paste(
      "design-based nested equal weights over biologically admissible",
      "sex × age × ethnicity cells; equal tissue weighting for comprehensive"
    ),
    prediction_policy = paste(
      "Normal; 10x Genomics 3; offset=0; biologically admissible covariate grid;",
      "unseen dataset random effect"
    ),
    analysis_version,
    analysis_id
  )

  list(
    analysis_version = analysis_version,
    analysis_id = analysis_id,
    cell_type = cell_type,
    study_draws = study_draws,
    profile_summary = profile_summary,
    rule_sensitivity = rule_sensitivity,
    cohort_contrasts = cohort_contrasts,
    threshold_registry = threshold_registry,
    diagnostics = bundle$diagnostics,
    support_profiles = mutate(bundle$profiles, cell_type = cell_type, .before = 1),
    tissue_sex_rules = mutate(tissue_sex_rules, cell_type = cell_type, .before = 1),
    level_audit = level_audit,
    marginalisation_weights = marginalisation_weights,
    design_grid = design_grid,
    reference = tibble(
      cell_type,
      reference_sample = bundle$reference_name,
      shared_features = length(aligned$shared_features),
      shared_feature_sha256 = digest::digest(sort(aligned$shared_features), algo = "sha256"),
      tm4sf1_dispersion = dispersion
    )
  )
}
```

## Build or load the model-aligned results

```{r}
#| label: build-model-results
#| include: false

study <- read_study_pseudobulk(case_dir)

model_results <- map(params$model_cell_types, function(cell_type) {
  cache_file <- file.path(cache_dir, paste0("TM4SF1_", cell_type, "_logmu_analysis.rds"))
  cached <- if (file.exists(cache_file) && !isTRUE(params$refresh_hca)) {
    readRDS(cache_file)
  } else {
    NULL
  }
  cache_ok <- !is.null(cached) &&
    identical(cached$analysis_id, analysis_id) &&
    identical(cached$cell_type, cell_type)
  if (cache_ok) {
    return(cached)
  }
  result <- run_one_cell_type(cell_type, study, params, model_root)
  saveRDS(result, cache_file, compress = "xz")
  result
})
names(model_results) <- params$model_cell_types

study_draws <- map_dfr(model_results, "study_draws")
profile_summary <- map_dfr(model_results, "profile_summary")
rule_sensitivity <- map_dfr(model_results, "rule_sensitivity")
cohort_contrasts <- map_dfr(model_results, "cohort_contrasts")
threshold_registry <- map_dfr(model_results, "threshold_registry")
model_diagnostics <- map_dfr(model_results, "diagnostics")
support_profiles <- map_dfr(model_results, "support_profiles")
tissue_sex_rules_all <- map_dfr(model_results, "tissue_sex_rules")
level_audit_all <- map_dfr(model_results, "level_audit")
marginalisation_weights_all <- map_dfr(model_results, "marginalisation_weights")
reference_audit <- map_dfr(model_results, "reference")

v1_profile_summary <- if (file.exists(file.path(v1_results_dir, "posteriorHCA_TM4SF1_profile_summary.tsv"))) {
  read_tsv(file.path(v1_results_dir, "posteriorHCA_TM4SF1_profile_summary.tsv"), show_col_types = FALSE)
} else {
  tibble()
}
v1_candidates <- if (file.exists(file.path(v1_results_dir, "TM4SF1_prototype_overlap_candidates.tsv"))) {
  read_tsv(file.path(v1_results_dir, "TM4SF1_prototype_overlap_candidates.tsv"), show_col_types = FALSE)
} else {
  tibble()
}

baseline_comparison <- profile_summary |>
  filter(dimension == "tissue", supported) |>
  transmute(
    cell_type,
    tissue_groups,
    v2_median_log_mu = median_log_mu,
    v2_probability_above_threshold = probability_above_threshold,
    v2_prototype_candidate = prototype_overlap_candidate
  ) |>
  left_join(
    v1_profile_summary |>
      filter(dimension == "tissue", supported) |>
      transmute(
        cell_type,
        tissue_groups,
        v1_median_log_mu = median_log_mu,
        v1_probability_above_threshold = probability_above_threshold,
        v1_prototype_candidate = prototype_overlap_candidate
      ),
    by = c("cell_type", "tissue_groups")
  ) |>
  mutate(
    delta_median_log_mu = v2_median_log_mu - v1_median_log_mu,
    delta_probability_above_threshold =
      v2_probability_above_threshold - v1_probability_above_threshold,
    candidate_status = case_when(
      v1_prototype_candidate & v2_prototype_candidate ~ "retained",
      v1_prototype_candidate & !v2_prototype_candidate ~ "lost",
      !v1_prototype_candidate & v2_prototype_candidate ~ "new",
      TRUE ~ "unchanged_non_candidate"
    )
  )

candidate_key <- function(df) {
  df |>
    filter(prototype_overlap_candidate) |>
    transmute(
      cell_type,
      key = paste(
        cell_type, dimension, tissue_groups, sex, age_decade, ethnicity_groups,
        sep = " | "
      )
    ) |>
    pull(key)
}

v1_keys <- candidate_key(v1_candidates)
v2_keys <- candidate_key(filter(profile_summary, prototype_overlap_candidate))
candidate_transition <- tibble(
  key = sort(unique(c(v1_keys, v2_keys))),
  in_v1 = key %in% v1_keys,
  in_v2 = key %in% v2_keys
) |>
  mutate(
    transition = case_when(
      in_v1 & in_v2 ~ "retained",
      in_v1 & !in_v2 ~ "disappeared",
      !in_v1 & in_v2 ~ "newly_appeared",
      TRUE ~ "none"
    )
  )

female_repro_audit <- marginalisation_weights_all |>
  filter(tissue_groups == "female reproductive system") |>
  arrange(cell_type, sex, age_decade, ethnicity_groups)

ready_model_names <- model_diagnostics |>
  filter(model_ready) |>
  pull(cell_type)
model_readiness_text <- if (!length(ready_model_names)) {
  "Neither current fit passes the registered readiness gate."
} else if (length(ready_model_names) == nrow(model_diagnostics)) {
  "All current fits pass the registered readiness gate."
} else {
  paste0(
    "Only ", paste(ready_model_names, collapse = ", "),
    " passes the registered readiness gate."
  )
}

prototype_tissue_groups <- profile_summary |>
  filter(dimension == "tissue", prototype_overlap_candidate) |>
  arrange(cell_type, desc(probability_above_threshold), tissue_groups) |>
  group_by(cell_type) |>
  summarise(tissues = paste(tissue_groups, collapse = ", "), .groups = "drop") |>
  transmute(label = paste0(cell_type, ": ", tissues)) |>
  pull(label)
prototype_tissue_text <- if (length(prototype_tissue_groups)) {
  paste(prototype_tissue_groups, collapse = "; ")
} else {
  "none"
}
posterior_draw_text <- paste(sort(unique(model_diagnostics$n_posterior_draws)), collapse = ", ")

write_tsv(profile_summary, file.path(results_dir, "posteriorHCA_TM4SF1_profile_summary.tsv"))
write_tsv(
  filter(profile_summary, supported),
  file.path(results_dir, "posteriorHCA_TM4SF1_supported_profiles.tsv")
)
write_tsv(
  filter(profile_summary, expression_overlap_alert),
  file.path(results_dir, "TM4SF1_expression_overlap_alerts.tsv")
)
write_tsv(
  filter(profile_summary, prototype_overlap_candidate),
  file.path(results_dir, "TM4SF1_prototype_overlap_candidates.tsv")
)
write_tsv(rule_sensitivity, file.path(results_dir, "TM4SF1_alert_rule_sensitivity.tsv"))
write_tsv(cohort_contrasts, file.path(results_dir, "TM4SF1_cohort_logmu_contrasts.tsv"))
write_tsv(threshold_registry, file.path(results_dir, "TM4SF1_threshold_registry.tsv"))
write_tsv(model_diagnostics, file.path(results_dir, "TM4SF1_model_diagnostics.tsv"))
write_tsv(reference_audit, file.path(results_dir, "TM4SF1_reference_alignment.tsv"))
write_tsv(tissue_sex_rules_all, file.path(results_dir, "TM4SF1_tissue_sex_rules_audit.tsv"))
write_tsv(level_audit_all, file.path(results_dir, "TM4SF1_level_support_audit.tsv"))
write_tsv(
  marginalisation_weights_all,
  file.path(results_dir, "TM4SF1_marginalisation_weights_audit.tsv")
)
write_tsv(baseline_comparison, file.path(results_dir, "TM4SF1_baseline_v1_v2_comparison.tsv"))
write_tsv(candidate_transition, file.path(results_dir, "TM4SF1_candidate_transition_audit.tsv"))
write_tsv(female_repro_audit, file.path(results_dir, "TM4SF1_female_repro_endothelial_audit.tsv"))
saveRDS(study_draws, file.path(results_dir, "TM4SF1_study_logmu_draws.rds"))
```

## What the baseline represents (v2 design-standardised)

Predictions were standardised over a prespecified biologically admissible
covariate space. Analysis-defining variables were fixed (`disease = Normal`,
`assay = 10x Genomics 3`, `offset = 0`, evaluated cell type, and tissue when
estimating a tissue-specific baseline). Structurally constrained covariates were
restricted to valid levels, and remaining demographic covariates were
marginalised using design-based weights independent of atlas sampling frequency.

The comprehensive reference is an **equal-weight average of tissue-level
standardised baselines**, not a direct pool of all donor/profile rows. Epithelial
and endothelial biology remain separate. Predictions draw a new dataset-level
random effect as in the SAVI cohort comparison. Here, 10x Genomics 3 is a fixed
posteriorHCA reporting scale, not a claim that the bladder study used that assay.

The primary safety estimand is **tissue × cell-type overlap** with the
HV-derived threshold. The comprehensive baseline is retained mainly as a
calibration/reference quantity.

`posterior_linpred(transform = FALSE)` estimates the conditional latent
negative-binomial `log(mu)` on the fitted model scale. It is not the marginal
zero-inflated expected count, detection probability, surface-protein abundance,
antigen density or toxicity. It is used here only for alignment with the SAVI
cohort-comparison estimand.

```{r}
#| label: baseline-support-table

support_profiles |>
  group_by(cell_type) |>
  summarise(
    donor_keys = n_distinct(donor_key),
    donor_tissue_profiles = n(),
    tissue_groups = n_distinct(tissue_groups),
    age_decades = n_distinct(age_decade),
    sexes = n_distinct(sex),
    ethnicity_groups = n_distinct(ethnicity_groups),
    .groups = "drop"
  ) |>
  kable(
    caption = paste(
      "Normal-labelled, 10x3-supported atlas profiles underlying support",
      "counts for the design grid (not marginalisation weights).",
      "A donor key is dataset_id × donor_id."
    )
  )
```

```{r}
#| label: tissue-sex-rules-audit

tissue_sex_rules_all |>
  select(cell_type, tissue_groups, allowed_sex, rule_source, reports_tissue_sex) |>
  kable(
    caption = paste(
      "Biological sex admissibility by tissue group (v2 audit).",
      "Tissue × sex is reported only when both sexes are biologically meaningful."
    )
  )
```

Bladder is not a fitted posteriorHCA tissue level. The few Normal-labelled 10x3
`bladder organ` profiles are collapsed into `renal system`, together with renal
tissues. Accordingly, the paired-normal comparison below is a calibration
against an imperfect renal-system epithelial proxy; this report does not make
bladder-specific demographic claims.

```{r}
#| label: paired-normal-calibration-table

retained_tumour_ids <- study$metadata |>
  filter(
    study_state == "tumour epithelial",
    strict_n150,
    paper_id %in% c(sprintf("VAR%02d", 1:9), sprintf("UC%02d", 1:3))
  ) |>
  pull(paper_id)

paired_normal_calibration <- study$metadata |>
  filter(study_state == "paired-normal epithelial") |>
  transmute(
    paper_id,
    epithelial_cells = n_cells,
    raw_library_size = library_size,
    paired_with_retained_tumour = paper_id %in% retained_tumour_ids
  )

paired_normal_calibration |>
  kable(
    caption = paste(
      "Paired/adjacent-normal epithelial calibration samples with at least",
      "20 cells. Only VAR07 has a retained tumour-cohort counterpart."
    )
  )
```

## Model-readiness note

```{r}
#| label: model-diagnostics

model_diagnostics |>
  mutate(
    max_rhat = round(max_rhat, 3),
    min_ess_bulk = round(min_ess_bulk, 1),
    min_ess_tail = round(min_ess_tail, 1)
  ) |>
  kable(caption = "Diagnostics of the current local TM4SF1 posteriorHCA fits.")
```

The current local fits contain `r posterior_draw_text` retained posterior draws
per model. **`r model_readiness_text`** `TM4SF1_expression_overlap_alerts.tsv`
contains only candidates from models that pass this gate and is therefore empty
when none is ready. The figures and tables below show ungated prototype overlap
candidates, not translational alerts; failed models must be refitted with
longer chains before their alerts can be evaluated.

------------------------------------------------------------------------

# 4. Paper cohorts versus posteriorHCA

```{r}
#| label: cohort-contrast-table

cohort_contrasts |>
  mutate(
    across(c(estimate_log2, lower_log2, upper_log2), ~ round(.x, 2)),
    probability_positive = percent(probability_positive, accuracy = 0.1)
  ) |>
  select(
    cell_type, study_group, comparator, n_study_patients,
    estimate_log2, lower_log2, upper_log2, probability_positive
  ) |>
  kable(
    caption = paste(
      "Patient-cohort minus posteriorHCA latent log-mu contrasts on a log2 scale.",
      "Intervals independently pair study bootstrap and HCA posterior draws."
    )
  )
```

```{r}
#| label: cohort-contrast-forest
#| fig-width: 7.2
#| fig-height: 3.74
#| fig-cap: "Cohort-level TM4SF1 contrasts. Positive values indicate higher expression in the study group than the model-aligned posteriorHCA comparator. Normal-labelled, 10x3-supported posteriorHCA scale. Full-width panel (183 mm)."

cohort_plot_data <- cohort_contrasts |>
  filter(
    study_group != "paired-normal epithelial (>=20 cells)" |
      cell_type == "epithelial"
  ) |>
  mutate(
    study_label = recode(
      study_group,
      "HV tumour epithelial" = "HV tumour (9 patients)",
      "UC tumour epithelial (2 public-DGE >=150)" =
        "UC tumour (2 public-DGE patients with >=150 cells)",
      "UC tumour epithelial (3 published IDs)" =
        "UC tumour (3 published IDs; UC01 has 90 cells)",
      "paired-normal epithelial (>=20 cells)" =
        "paired/adjacent normal (3 patients)"
    ),
    reference_label = case_when(
      str_detect(comparator, "renal-system") ~
        "HCA renal-system epithelial proxy",
      cell_type == "epithelial" ~ "HCA comprehensive epithelial reference",
      TRUE ~ "HCA comprehensive endothelial reference"
    ),
    comparison = str_wrap(paste(study_label, "vs", reference_label), 48),
    comparison = factor(comparison, levels = rev(unique(comparison)))
  )

cohort_plot <- ggplot(cohort_plot_data, aes(estimate_log2, comparison, colour = cell_type)) +
  geom_vline(xintercept = 0, linetype = 2, colour = col_other, linewidth = 0.25) +
  geom_errorbar(
    aes(xmin = lower_log2, xmax = upper_log2),
    orientation = "y", width = 0.15, linewidth = 0.3
  ) +
  geom_point(size = 1.5) +
  scale_x_continuous(labels = label_drop_leading_zero()) +
  scale_colour_manual(values = colmap_cell, name = "posteriorHCA model") +
  labs(x = "Study minus posteriorHCA log2 latent expression", y = NULL) +
  theme_multipanel

save_figure(
  cohort_plot, "05_TM4SF1_cohort_vs_posteriorHCA_logmu",
  panel_mm$full["w"], panel_mm$full["h"]
)
cohort_plot
```

The three-sample paired/adjacent-normal result is a **diagnostic calibration
check**, not a healthy-control effect estimate or a matched-pair analysis. A
material shift from the posteriorHCA renal proxy can arise from adjacent-tissue
biology, bladder-versus-kidney aggregation, platform composition or study
effects. Tumour-versus-HCA differences should therefore be read as modular
cross-study contrasts rather than a jointly fitted causal effect.

------------------------------------------------------------------------

# 5. A reusable threshold contract

The report defines the screening threshold separately for each posteriorHCA
cell-type model:

$$
T_c = Q_{q}\!\left\{ \log \mu_{\mathrm{HV},c}^{(b)} \right\}
$$

where $q =$ `r percent(params$threshold_quantile, accuracy = 1)` and $c$ is
the posteriorHCA model, and the distribution is the nine-patient HV
tumour-epithelial Bayesian bootstrap after alignment to that model's TMMwsp
reference. A supported Normal-labelled profile qualifies as a prototype
candidate when

$$
P\!\left( \log \mu_{\mathrm{Normal},c} \geq T_c \mid \mathrm{data} \right) \geq p_{\min}
$$

with $p_{\min} =$ `r percent(params$alert_probability, accuracy = 1)`. The
selected $q$ is a lower posterior quantile of uncertainty in the nine-patient
cohort-location estimator. It is not a quantile of patient expression, a
minimum antigen-positive level, or a biological or clinical toxicity threshold.
Both the quantile and probability are report parameters, not hidden constants.

```{r}
#| label: threshold-registry-table

threshold_registry |>
  transmute(
    cell_type_model,
    normalization_reference,
    threshold_quantile,
    threshold_log_mu = round(threshold_log_mu, 3),
    alert_probability_cutoff,
    minimum_donor_keys,
    model_ready,
    diagnostic_gate,
    analysis_version
  ) |>
  kable(caption = "Model-specific TM4SF1 threshold registry.")
```

The prototype exceedance-probability and donor-support cutoffs are screening
choices, so their effect is reported rather than hidden:

```{r}
#| label: threshold-rule-sensitivity

rule_sensitivity |>
  mutate(
    threshold_quantile = percent(threshold_quantile, accuracy = 1),
    probability_rule = paste0(
      "candidates at P>=",
      percent(alert_probability_cutoff, accuracy = 1)
    )
  ) |>
  select(
    cell_type, model_ready, threshold_quantile, threshold_log_mu,
    minimum_donor_keys, supported_tissues, probability_rule,
    prototype_tissue_candidates
  ) |>
  pivot_wider(
    names_from = probability_rule,
    values_from = prototype_tissue_candidates
  ) |>
  kable(
    caption = paste(
      "Number of prototype tissue candidates under alternative HV-anchor",
      "quantiles, support rules and posterior-probability cutoffs."
    )
  )
```

The raw `log(mu)` cutoff is reusable only with the recorded model store,
feature, reference sample, assay policy, offset-zero convention and new-study
random-effect policy. Across a different posteriorHCA release or cell-type
model, the study counts must be realigned and $T_c$ recalibrated. The
estimand and recomputation procedure are portable; numerical log2 ratios,
exceedance probabilities and cutoffs must be recomputed for every model or
baseline release.

------------------------------------------------------------------------

# 6. Standardised tissue baselines and demographic marginals

Demographic outputs are **standardised marginal predictions** over the
biologically admissible covariate space, not raw observed-profile averages.
Tissue rows marginalise admissible sex, age and ethnicity with nested equal
weights. Tissue × sex is reported only where both sexes are biologically
meaningful; sex-specific tissues (e.g. female reproductive system) report a
single tissue baseline only. Tissue × age and tissue × ethnicity marginalise
the remaining admissible demographics with balanced weights.

A tissue-demographic result must contain at least `r params$min_donor_keys`
distinct dataset × donor keys across its contributing admissible cells before it
can qualify as an ungated prototype candidate. Sparse strata remain in the full
output with `supported = FALSE` but are suppressed from candidate plots.
`Other/Unknown` ethnicity is excluded from primary marginalisation and cannot
qualify as a candidate.

## Tissue and cell type

```{r}
#| label: tissue-profile-plot
#| fig-width: 4.72
#| fig-height: 3.74
#| fig-cap: "Standardised Normal-labelled posteriorHCA tissue baselines relative to the model-specific HV tumour-expression anchor (dashed line). Highlighted points pass the donor-support and posterior-probability rules but remain ungated by model readiness."

tissue_plot_data <- profile_summary |>
  filter(dimension == "tissue", supported) |>
  group_by(cell_type) |>
  mutate(tissue_rank = rank(-median_log_mu, ties.method = "first")) |>
  ungroup() |>
  mutate(
    tissue_label = paste0(tissue_groups, " (n=", n_donor_keys, ")"),
    tissue_label = reorder(tissue_label, median_log_mu)
  )

tissue_plot <- ggplot(
  tissue_plot_data,
  aes(median_log_mu, tissue_label, colour = prototype_overlap_candidate)
) +
  geom_errorbar(
    aes(xmin = lower_log_mu, xmax = upper_log_mu),
    orientation = "y", width = 0.12, linewidth = 0.25, colour = col_other
  ) +
  geom_point(size = 1.4) +
  geom_vline(aes(xintercept = threshold_log_mu), linetype = 2, colour = col_hv, linewidth = 0.3) +
  facet_wrap(~ cell_type, scales = "free_y") +
  scale_x_continuous(labels = label_drop_leading_zero()) +
  scale_colour_manual(
    values = c(`TRUE` = col_candidate, `FALSE` = col_dark),
    labels = c(`TRUE` = "ungated prototype candidate", `FALSE` = "below prototype rule"),
    name = NULL
  ) +
  labs(x = "Standardised Normal-labelled posteriorHCA latent log(mu), offset 0", y = NULL) +
  theme_multipanel

save_figure(
  tissue_plot, "06_posteriorHCA_TM4SF1_tissue_profiles",
  panel_mm$lg["w"], panel_mm$lg["h"]
)
tissue_plot
```

## Standardised sex marginals within tissue (both-sex tissues only)

```{r}
#| label: tissue-sex-plot
#| fig-width: 4.72
#| fig-height: 3.74
#| fig-cap: "Standardised tissue × sex marginals for tissues where both sexes are biologically meaningful. Age and ethnicity are marginalised with balanced admissible weights."

sex_plot_data <- profile_summary |>
  filter(dimension == "tissue_sex", supported) |>
  mutate(
    tissue_groups = str_wrap(tissue_groups, 24),
    tissue_groups = reorder(tissue_groups, median_log_mu)
  )

sex_plot <- ggplot(sex_plot_data, aes(median_log_mu, tissue_groups, colour = sex)) +
  geom_errorbar(
    aes(xmin = lower_log_mu, xmax = upper_log_mu),
    orientation = "y", width = 0.12, linewidth = 0.25,
    position = position_dodge(width = 0.5),
    alpha = 0.7
  ) +
  geom_point(position = position_dodge(width = 0.5), size = 1.3) +
  geom_vline(aes(xintercept = threshold_log_mu), linetype = 2, colour = col_other, linewidth = 0.3) +
  facet_wrap(~ cell_type, scales = "free_y") +
  scale_x_continuous(labels = label_drop_leading_zero()) +
  scale_colour_manual(values = colmap_sex, name = "Sex") +
  labs(x = "Standardised Normal-labelled posteriorHCA latent log(mu), offset 0", y = NULL) +
  theme_multipanel

save_figure(
  sex_plot, "07_posteriorHCA_TM4SF1_tissue_by_sex",
  panel_mm$lg["w"], panel_mm$lg["h"]
)
sex_plot
```

## Standardised age-decade marginals within tissue

```{r}
#| label: tissue-age-plot
#| fig-width: 4.72
#| fig-height: 3.35
#| fig-cap: "Standardised tissue × age-decade marginals (sex and ethnicity marginalised with balanced admissible weights). The eight highest aggregate tissues plus every qualifying prototype tissue are shown. Outlined tiles pass the prototype overlap rule."

top_tissues <- profile_summary |>
  filter(dimension == "tissue", supported) |>
  group_by(cell_type) |>
  slice_max(median_log_mu, n = 8, with_ties = FALSE) |>
  ungroup() |>
  select(cell_type, tissue_groups)

candidate_tissues <- profile_summary |>
  filter(prototype_overlap_candidate, !is.na(tissue_groups)) |>
  select(cell_type, tissue_groups)

display_tissues <- bind_rows(top_tissues, candidate_tissues) |>
  distinct()

age_plot_data <- profile_summary |>
  filter(dimension == "tissue_age", supported) |>
  inner_join(display_tissues, by = c("cell_type", "tissue_groups")) |>
  mutate(
    age_decade = factor(
      age_decade,
      levels = as.character(sort(unique(as.integer(age_decade))))
    ),
    tissue_groups = str_wrap(tissue_groups, 24)
  )

age_plot <- ggplot(
  age_plot_data,
  aes(age_decade, tissue_groups, fill = probability_above_threshold)
) +
  geom_tile(colour = "white", linewidth = 0.2) +
  geom_point(
    data = filter(age_plot_data, prototype_overlap_candidate),
    shape = 21, fill = NA, colour = col_dark, size = 1.6, stroke = 0.4
  ) +
  facet_wrap(~ cell_type, scales = "free_y") +
  scale_fill_gradient(
    low = "white", high = col_uc,
    limits = c(0, 1), labels = label_percent(),
    name = "P(Normal profile >= anchor)"
  ) +
  labs(x = "Age decade (e.g. 6 = 60-69 years)", y = NULL) +
  theme_multipanel

save_figure(
  age_plot, "08_posteriorHCA_TM4SF1_tissue_by_age",
  panel_mm$heatmap["w"], panel_mm$heatmap["h"]
)
age_plot
```

## Standardised ethnicity marginals within tissue

```{r}
#| label: tissue-ethnicity-plot
#| fig-width: 4.72
#| fig-height: 3.35
#| fig-cap: "Standardised tissue × ethnicity marginals (sex and age marginalised with balanced admissible weights) for the eight highest aggregate tissues plus every qualifying prototype tissue. Outlined tiles pass the prototype overlap rule."

ethnicity_plot_data <- profile_summary |>
  filter(dimension == "tissue_ethnicity", supported) |>
  inner_join(display_tissues, by = c("cell_type", "tissue_groups")) |>
  mutate(
    tissue_groups = str_wrap(tissue_groups, 24),
    ethnicity_groups = str_wrap(ethnicity_groups, 18)
  )

ethnicity_plot <- ggplot(
  ethnicity_plot_data,
  aes(ethnicity_groups, tissue_groups, fill = probability_above_threshold)
) +
  geom_tile(colour = "white", linewidth = 0.2) +
  geom_point(
    data = filter(ethnicity_plot_data, prototype_overlap_candidate),
    shape = 21, fill = NA, colour = col_dark, size = 1.6, stroke = 0.4
  ) +
  facet_wrap(~ cell_type, scales = "free_y") +
  scale_fill_gradient(
    low = "white", high = col_uc,
    limits = c(0, 1), labels = label_percent(),
    name = "P(Normal profile >= anchor)"
  ) +
  labs(x = "Recorded/harmonised ethnicity group", y = NULL) +
  theme_multipanel +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1),
    plot.margin = margin(2, 4, 8, 2)
  )

save_figure(
  ethnicity_plot, "09_posteriorHCA_TM4SF1_tissue_by_ethnicity",
  panel_mm$heatmap["w"], panel_mm$heatmap["h"]
)
ethnicity_plot
```

## Ungated prototype candidates for follow-up

```{r}
#| label: prototype-candidate-table

candidate_table <- profile_summary |>
  filter(
    prototype_overlap_candidate,
    dimension %in% c("tissue", "tissue_sex", "tissue_age", "tissue_ethnicity")
  ) |>
  arrange(cell_type, dimension, desc(probability_above_threshold)) |>
  transmute(
    cell_type,
    dimension,
    tissue = tissue_groups,
    sex,
    age_decade,
    ethnicity = ethnicity_groups,
    donor_keys = n_donor_keys,
    median_log_mu = round(median_log_mu, 2),
    log2_vs_HV = round(estimate_log2_vs_hv, 2),
    P_above_threshold = percent(probability_above_threshold, accuracy = 0.1),
    P_at_least_random_HV = percent(probability_hca_ge_random_hv, accuracy = 0.1)
  )

head(candidate_table, 40) |>
  kable(
    caption = paste(
      "First 40 supported prototype overlap candidates, ungated by model",
      "readiness. The full machine-readable table is written to results/."
    )
  )
```

The tissue, tissue × sex (where reported), tissue × age and tissue × ethnicity
rows are nested views of the same standardised baseline framework; they should
not be counted as independent safety findings.

## Baseline redesign audit (v1 observed-profile vs v2 design-standardised)

```{r}
#| label: baseline-v1-v2-comparison

baseline_comparison |>
  mutate(
    across(
      c(v1_median_log_mu, v2_median_log_mu, delta_median_log_mu),
      ~ round(.x, 3)
    ),
    across(
      c(v1_probability_above_threshold, v2_probability_above_threshold,
        delta_probability_above_threshold),
      ~ percent(.x, accuracy = 0.1)
    )
  ) |>
  select(
    cell_type, tissue_groups, v1_median_log_mu, v2_median_log_mu,
    delta_median_log_mu, v1_probability_above_threshold,
    v2_probability_above_threshold, delta_probability_above_threshold,
    v1_prototype_candidate, v2_prototype_candidate, candidate_status
  ) |>
  kable(
    caption = paste(
      "Tissue-level baseline comparison: v1 equal-weight observed profiles",
      "versus v2 design-standardised marginals (supported tissues only)."
    )
  )
```

```{r}
#| label: candidate-transition-audit

candidate_transition |>
  filter(transition != "none") |>
  kable(
    caption = paste(
      "Prototype overlap candidates that were retained, lost or newly appeared",
      "under the v2 design-standardised baseline."
    )
  )
```

```{r}
#| label: marginalisation-weights-audit

marginalisation_weights_all |>
  group_by(cell_type, tissue_groups) |>
  summarise(
    n_cells = n(),
    weight_sum = sum(marginal_weight),
    sex_levels = paste(sort(unique(sex)), collapse = ", "),
    .groups = "drop"
  ) |>
  kable(
    caption = paste(
      "Marginalisation weight summary per tissue (each tissue's admissible",
      "cells sum to weight 1.0 under nested equal weighting)."
    )
  )
```

```{r}
#| label: female-repro-endothelial-audit

female_repro_audit |>
  select(
    cell_type, sex, age_decade, ethnicity_groups,
    n_donor_keys, marginal_weight, weight_rule
  ) |>
  kable(
    caption = paste(
      "Female reproductive system marginalisation weights (v2).",
      "Male sex is structurally excluded; endothelial tissue baseline",
      "marginalises female × admissible age × admissible ethnicity only."
    )
  )
```

The motivating respiratory-system example is checked explicitly rather than
assumed:

```{r}
#| label: respiratory-sex-check

respiratory_sex <- profile_summary |>
  filter(
    dimension == "tissue_sex",
    tissue_groups == "respiratory system",
    supported
  ) |>
  transmute(
    cell_type,
    sex,
    donor_keys = n_donor_keys,
    median_log_mu = round(median_log_mu, 2),
    threshold_log_mu = round(threshold_log_mu, 2),
    P_above_threshold = percent(probability_above_threshold, accuracy = 0.1),
    prototype_candidate = prototype_overlap_candidate,
    model_ready,
    model_ready_alert = expression_overlap_alert
  )

respiratory_sex |>
  kable(caption = "Standardised respiratory-system tissue × sex marginals under the v2 rule.")
```

`r if (any(respiratory_sex$prototype_candidate)) paste0("At least one supported respiratory-system sex profile passes the ungated ", percent(params$alert_probability, accuracy = 1), " prototype overlap rule; it still cannot become an alert unless its model passes the readiness gate.") else paste0("No supported respiratory-system sex profile passes the current ", percent(params$alert_probability, accuracy = 1), " prototype-overlap rule. Model readiness is reported separately, so the present analysis supports neither a respiratory-system candidate nor a male-lung alert.")`

------------------------------------------------------------------------

# 7. Conclusions and next decisions

The released DE table ranks TM4SF1 first among HV-enriched genes, and the
closest public-DGE reconstruction supports HV enrichment, subject to the
documented 9-HV/2-UC limitation. Its tumour expression is heterogeneous, so a
target-positive HV label should not be interpreted as uniform antigen
abundance.

The off-tumour screening conclusion is cell-type specific. Endothelium is a
necessary sentinel because the bladder dataset itself shows high endothelial
CPM and near-universal detection. At the registered default settings, the
ungated tissue-level candidates are `r prototype_tissue_text`. These are
follow-up candidates, not toxicity calls or model-ready alerts; readiness is
reported separately above.

A supported **respiratory-system × male** profile could be prioritised for
off-tumour validation only if it passed the registered probability rule, the
corresponding female profile were examined, and a refitted posteriorHCA model
confirmed the result. The current prototype does not pass that rule and cannot
infer vulnerability. The fitted tissue group combines lung and airway
compartments and cannot localise risk to a specific pulmonary structure.

The immediate next analysis should be a longer-chain refit of TM4SF1 epithelial
and endothelial models, followed by protein-level localisation and
antigen-density calibration in the prototype candidate tissues before any
toxicity interpretation. This report intentionally stops before defining a
clinical toxicity threshold.

------------------------------------------------------------------------

# Reproducible outputs

```{r}
#| label: output-manifest

output_manifest <- tribble(
  ~file, ~contents,
  "TM4SF1_paper_DE_replication.tsv", "Released and public-DGE TM4SF1 DE results",
  "TM4SF1_paper_source_patient_summary.tsv", "Figure5B patient-level source-expression summary",
  "TM4SF1_study_logmu_draws.rds", "Patient Bayesian-bootstrap cohort log(mu) draws",
  "TM4SF1_cohort_logmu_contrasts.tsv", "Study-versus-HCA log2 contrasts",
  "TM4SF1_threshold_registry.tsv", "Cell-type-specific threshold contract and provenance",
  "TM4SF1_alert_rule_sensitivity.tsv", "Ungated prototype tissue-candidate counts while varying anchor quantile, donor support and exceedance probability",
  "posteriorHCA_TM4SF1_profile_summary.tsv", "All standardised tissue/demographic posterior summaries (v2)",
  "posteriorHCA_TM4SF1_supported_profiles.tsv", "Profiles meeting the minimum donor-key rule (v2)",
  "TM4SF1_prototype_overlap_candidates.tsv", "Supported ungated profiles passing the prototype overlap rule (v2)",
  "TM4SF1_expression_overlap_alerts.tsv", "Readiness-gated alerts; empty whenever model_ready is false",
  "TM4SF1_model_diagnostics.tsv", "Current brms fit convergence diagnostics",
  "TM4SF1_reference_alignment.tsv", "TMM reference, shared feature count and TM4SF1 dispersion",
  "TM4SF1_tissue_sex_rules_audit.tsv", "Biological sex admissibility by tissue (v2)",
  "TM4SF1_level_support_audit.tsv", "Age/ethnicity inclusion/exclusion audit per tissue (v2)",
  "TM4SF1_marginalisation_weights_audit.tsv", "Exact nested marginalisation weights (v2)",
  "TM4SF1_baseline_v1_v2_comparison.tsv", "v1 observed-profile vs v2 design baseline comparison",
  "TM4SF1_candidate_transition_audit.tsv", "Prototype candidates retained/lost/new under v2",
  "TM4SF1_female_repro_endothelial_audit.tsv", "Female reproductive endothelial weight audit",
  "TM4SF1_toxicity_case_study_v2_design_standardised.md", "Knitted Markdown preview of the v2 QMD",
  "output_manifest.tsv", "Machine-readable inventory of report outputs",
  "sessionInfo.txt", "R session and package versions for this report",
  "figures/", "Report figures as reusable PDF (Nature) and PNG files",
  "cache/", "Per-cell-type posterior/cache objects keyed by analysis version"
) |>
  mutate(path = file.path(results_dir, file), .before = 1)

write_tsv(output_manifest, file.path(results_dir, "output_manifest.tsv"))
writeLines(capture.output(sessionInfo()), file.path(results_dir, "sessionInfo.txt"))

output_manifest |>
  kable(caption = "Files produced by this report.")
```

Primary provenance: [GEO GSE293189](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE293189),
the [published study](https://pmc.ncbi.nlm.nih.gov/articles/PMC12174346/), and
the authors' [public analysis repository](https://github.com/angelussong/Histological_Variant_Bladder_Cancer_Analysis).
The preprocessing discrepancy and exact 67-library mapping are documented in
the case-study `results/replication_audit/` and `data/metadata/` directories.

------------------------------------------------------------------------

# Methodological audit (v2 redesign)

```{r}
#| label: methodological-audit-summary

v2_tissue_candidates <- profile_summary |>
  filter(dimension == "tissue", prototype_overlap_candidate) |>
  arrange(cell_type, desc(probability_above_threshold)) |>
  transmute(label = paste0(cell_type, ": ", tissue_groups)) |>
  pull(label)

audit_summary <- tibble(
  item = c(
    "Frozen v1 QMD",
    "Frozen v1 results",
    "Modified working copy",
    "v2 results directory",
    "Old baseline estimand",
    "New baseline estimand",
    "Biological sex constraints",
    "Marginalisation weights",
    "Threshold values changed?",
    "v1 tissue candidates (endothelial)",
    "v2 tissue candidates",
    "Candidates retained",
    "Candidates lost",
    "Candidates newly appeared"
  ),
  detail = c(
    "archive/v1_observed_profile/TM4SF1_toxicity_case_study_v1_observed_profile.qmd",
    "archive/v1_observed_profile/results_v1_observed_profile/",
    "TM4SF1_toxicity_case_study_v2_design_standardised.qmd",
    "results/tm4sf1_toxicity_report_v2_design_standardised/",
    "Equal weight per observed Normal/10x3 donor-tissue-demographic profile row",
    paste(
      "Nested equal weights over biologically admissible sex × age × ethnicity;",
      "equal tissue weighting for comprehensive baseline"
    ),
    paste(
      "female reproductive → female only; male reproductive/prostate → male only;",
      "breast → female only; respiratory/integumentary/renal → both sexes;",
      "Other/Unknown ethnicity excluded; sparse cells excluded (min donors = ",
      params$min_donor_keys, ")"
    ),
    "w = (1/|S|)(1/|A|)(1/|E|) over included admissible cells per tissue",
    "No — HV bootstrap threshold contract unchanged (model-specific T_c)",
    paste(
      v1_candidates |>
        filter(dimension == "tissue", cell_type == "endothelial") |>
        pull(tissue_groups) |>
        unique() |>
        paste(collapse = ", "),
      "(plus tissue_sex duplicates)"
    ),
    if (length(v2_tissue_candidates)) paste(v2_tissue_candidates, collapse = "; ") else "none",
    paste(filter(candidate_transition, transition == "retained")$key, collapse = "; "),
    paste(filter(candidate_transition, transition == "disappeared")$key, collapse = "; "),
    paste(filter(candidate_transition, transition == "newly_appeared")$key, collapse = "; ")
  )
)

audit_summary |>
  kable(caption = "Concise methodological audit of the v2 baseline redesign.")
```

Remaining caveats: structurally invalid combinations (e.g. male × female
reproductive system) receive zero weight and never enter baselines; poorly
supported but technically predictable cells remain auditable in
`TM4SF1_level_support_audit.tsv` but are excluded from primary
marginalisation. Comprehensive baselines are calibration references only; the
primary toxicity-screen estimand remains tissue × cell-type overlap with the
HV-derived threshold. Model-readiness gating is unchanged — prototype candidates
below remain ungated translational alerts.

------------------------------------------------------------------------

## Session info

```{r}
sessionInfo()
```
