---
title: "Cohort expression workflow (brms test)"
author: "Chen Zhan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cohort expression workflow (brms test)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  for_github_md: false
knit: >
    (function(x, ...){
        proj_root <- rprojroot::find_package_root_file() |> normalizePath()
        rmarkdown::render(
            x,
            output_file = "cohort-expression-brms-test.md",
            output_format = "github_document",
            output_dir = file.path(proj_root, "vignettes"),
            knit_root_dir = proj_root,
            params = list(for_github_md = TRUE),
            envir = new.env(parent = globalenv())
        )
        rmarkdown::render(
            x,
            output_format = "rmarkdown::html_vignette",
            params = list(for_github_md = FALSE),
            envir = new.env(parent = globalenv())
        )
    })
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)

if (isTRUE(params$for_github_md)) {
  root_dir <- knitr::opts_knit$get("root.dir")
  if (!is.null(root_dir)) {
    knitr::opts_knit$set(output.dir = file.path(root_dir, "vignettes"))
  }
  knitr::opts_chunk$set(fig.path = "cohort-expression-brms-test_files/figure-gfm/")
}

can_run <- requireNamespace("Seurat", quietly = TRUE) &&
  requireNamespace("org.Hs.eg.db", quietly = TRUE) &&
  requireNamespace("AnnotationDbi", quietly = TRUE) &&
  requireNamespace("brms", quietly = TRUE) &&
  requireNamespace("edgeR", quietly = TRUE) &&
  requireNamespace("limma", quietly = TRUE)

has_cmdstan <- requireNamespace("cmdstanr", quietly = TRUE) &&
  isTRUE(tryCatch({
    cmdstanr::cmdstan_version(error_on_NA = TRUE)
    TRUE
  }, error = function(e) FALSE))

can_fit_brms <- can_run && has_cmdstan

if (requireNamespace("devtools", quietly = TRUE)) {
  pkg_root <- if (file.exists("DESCRIPTION") && dir.exists("R")) {
    normalizePath(".")
  } else if (file.exists("../DESCRIPTION") && dir.exists("../R")) {
    normalizePath("..")
  } else {
    NULL
  }
  if (!is.null(pkg_root)) {
    devtools::load_all(pkg_root, quiet = TRUE)
  }
}
```

SAVI case study: *ADRB2* (`ENSG00000169252`) in disease-associated monocytes.

Same analysis as `vignette("cohort-expression-edger-native")`, with an extra
user-side Bayesian step: fit `brms` negative binomial (NB) and
zero-inflated NB (ZINB) models on the same formula / offset as edgeR, then
summarise SAVI `posterior_linpred` draws exactly as the HCA baseline path
does (`mean` → `log_mu`, `sd` → `se`).

```text
USER + HCA counts
  → TMMwsp
  → offset = log(E_user)
  → edgeR: glmQLFit → CategorySAVI → QLF SE / package Wald SE
  → brms NB / ZINB: counts ~ 1 + Category + offset(hca_offset)
  → linpred(newdata = SAVI, ndraws = 400) → log_mu, se
  → + log(E_HCA) for edgeR absolute scale
  → compare all estimates to HCA posterior
```

`brms` fitting chunks require `cmdstanr` with a working CmdStan
installation. Without that, the edgeR comparison still runs.

## Setup

```{r libraries, eval = can_run}
library(posteriorHCA)
library(Seurat)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(edgeR)
library(limma)
library(brms)
library(dplyr)
library(forcats)
library(tibble)

cell_type <- "monocytic"
gene <- "ENSG00000169252"
n_chains <- 2L
n_warmup <- 400L
n_iter <- 600L
n_linpred_draws <- 400L
```

## Counts

```{r load-savi-mono, eval = can_run}
data(savi_mono)

counts <- as.matrix(Seurat::GetAssayData(savi_mono, layer = "counts"))
ensembl <- AnnotationDbi::mapIds(
  org.Hs.eg.db,
  keys = rownames(counts),
  column = "ENSEMBL",
  keytype = "SYMBOL",
  multiVals = "first"
)
ok <- !is.na(ensembl) & !duplicated(ensembl)
counts <- counts[ok, , drop = FALSE]
rownames(counts) <- unname(ensembl[ok])

meta <- data.frame(
  Category = factor(savi_mono$Category[colnames(counts)]),
  Experiment = factor(savi_mono$Experiment[colnames(counts)]),
  row.names = colnames(counts)
)
table(meta$Category, meta$Experiment)
```

## HCA reference

```{r reference, eval = can_run}
ref <- load_reference_sample(cell_type)
ref_id <- ref$sample_id

counts <- merge_with_reference_sample(
  counts,
  reference = ref$counts,
  reference_name = ref_id
)
```

The reference is only for joint TMM. Drop it before fitting.

## TMMwsp scaling

```{r scale, eval = can_run}
nf <- calcNormFactors(counts, refColumn = match(ref_id, colnames(counts)), method = "TMMwsp")
eff <- colSums(counts) * nf

user <- setdiff(colnames(counts), ref_id)
log_E <- log(eff[user])
log_E_hca <- log(eff[[ref_id]])
hca_offset <- log(eff[user] / eff[[ref_id]])
```

## USER edgeR model

```{r fit-edger, eval = can_run}
y <- counts[, user, drop = FALSE]
meta <- meta[user, , drop = FALSE] |>
  mutate(Category = fct_relevel(Category, "CTRL"))

offset <- matrix(log_E, nrow = nrow(y), ncol = ncol(y), byrow = TRUE)
design <- model.matrix(~ 1 + Category, data = meta)
colnames(design)

fit <- glmQLFit(y, design = design, offset = offset, prior.count = 0, robust = TRUE)

contrast <- makeContrasts('(Intercept) + CategorySAVI', levels = design)
contrast
```

`CategorySAVI` is the SAVI effect under `~ 1 + Category` (CTRL reference).

## Estimate and QLF-implied SE

`glmQLFTest()` reports `logFC` on the log2 scale. Convert to natural log.
The SE below is implied by the one-dimensional QL F statistic, not a
coefficient-covariance SE.

```{r qlf, eval = can_run}
qlf <- glmQLFTest(fit, contrast = contrast)
tab <- qlf$table[gene, ]

estimate <- tab$logFC * log(2)
se_qlf <- abs(tab$logFC) / sqrt(tab$F) * log(2)

# Absolute estimand → place on HCA scale (SE unchanged).
log_mu <- estimate + log_E_hca
n_savi <- sum(meta$Category == "SAVI")
beta_start <- as.numeric(fit$coefficients[gene, ])

data.frame(gene, estimate, log_mu, se_qlf, n = n_savi)
```

## Package Wald SE (for comparison)

```{r wald, eval = can_run}
wald <- estimate_ql(
  counts = y,
  offset = log_E,
  metadata = meta,
  formula = ~ 1 + Category,
  contrast = "CategorySAVI"
)
wald <- wald[wald$gene == gene, ]
se_wald <- wald$se

data.frame(
  estimate_qlf = estimate,
  estimate_wald = wald$estimate,
  se_qlf,
  se_wald,
  se_ratio = se_qlf / se_wald
)
```

## USER brms models (NB and ZINB)

Same mean structure as edgeR: `~ 1 + Category` with posteriorHCA offset.

```{r brms-data, eval = can_fit_brms}
df_gene <- data.frame(
  counts = as.integer(y[gene, ]),
  Category = meta$Category,
  hca_offset = as.numeric(hca_offset),
  row.names = colnames(y)
) |>
  mutate(Category = fct_relevel(Category, "CTRL"))

make_inits <- function(zi = FALSE) {
  force(zi)
  function(chain_id = 1) {
    set.seed(1000L + as.integer(chain_id) + if (zi) 50L else 0L)
    out <- list(
      b = as.array(beta_start + stats::rnorm(length(beta_start), 0, 0.05)),
      shape = max(5 + stats::rnorm(1, 0, 0.1), 0.5)
    )
    if (isTRUE(zi)) {
      out$Intercept_zi <- -2 + stats::rnorm(1, 0, 0.1)
    }
    out
  }
}

inits <- make_inits(zi = FALSE)
inits_zi <- make_inits(zi = TRUE)

form_nb <- counts ~ 1 + Category + offset(hca_offset)
form_zinb <- brms::bf(counts ~ 1 + Category + offset(hca_offset), zi ~ 1)
```

```{r fit-brms-nb, eval = can_fit_brms}
fit_nb <- brms::brm(
  formula = form_nb,
  data = df_gene,
  family = brms::negbinomial(),
  chains = n_chains,
  warmup = n_warmup,
  iter = n_iter,
  refresh = 10,
  backend = "cmdstanr",
  seed = 1
)
fit_nb
```

```{r fit-brms-zinb, eval = can_fit_brms}
fit_zinb <- brms::brm(
  formula = form_zinb,
  data = df_gene,
  family = brms::zero_inflated_negbinomial(),
  chains = n_chains,
  warmup = n_warmup,
  iter = n_iter,
  refresh = 10,
  backend = "cmdstanr",
  seed = 2
)
fit_zinb
```

### SAVI linpred → log_mu / se

Mirror the HCA baseline path: `posterior_linpred` (latent log μ), then
`summarize_posterior_draws()` (`mean` / `sd`). Query SAVI at `hca_offset = 0`.

```{r brms-linpred, eval = can_fit_brms}
summarise_savi_linpred <- function(fit, newdata, ndraws) {
  draws_mat <- brms::posterior_linpred(
    fit,
    newdata = newdata,
    transform = FALSE,
    ndraws = ndraws
  )
  draws <- as.numeric(draws_mat)
  summarize_posterior_draws(draws)
}

newdata_brms <- list(Category = "SAVI", hca_offset = 0) |> as_tibble()

nb_savi <- summarise_savi_linpred(fit_nb, newdata_brms, n_linpred_draws)
zinb_savi <- summarise_savi_linpred(fit_zinb, newdata_brms, n_linpred_draws)

data.frame(
  method = c("brms NB", "brms ZINB"),
  log_mu = c(nb_savi$log_mu, zinb_savi$log_mu),
  se = c(nb_savi$se, zinb_savi$se),
  n_draws = c(nb_savi$n, zinb_savi$n)
)
```

```{r brms-skip-note, eval = !can_fit_brms, echo = FALSE, results = "asis"}
cat(
  "> **Note:** `cmdstanr` / CmdStan was not available, so NB and ZINB ",
  "`brms` fits were skipped. Install CmdStan via `cmdstanr` to run those ",
  "chunks locally.\n"
)
```

## HCA posterior

```{r hca, eval = can_run}
fit_hca <- load_expression_fit(cell_type = cell_type, gene_ensg = gene)
newdata <- build_newdata_grid(
  fit_hca,
  disease_groups = "Normal",
  tissue_groups = "blood",
  assay_groups = "10x Genomics 3"
)
draws <- expression_draws(fit_hca, newdata = newdata, quantity = "linpred", marginalise = "mean")
hca <- summarize_posterior_draws(draws, value = log_mu)
```

## Welch tests (all methods)

Same HCA baseline; cohort rows differ by estimator / SE.

```{r welch, eval = can_run}
welch_one <- function(mu1, se1) {
  welch_test_means(
    mu1 = mu1, se1 = se1,
    mu2 = hca$log_mu, se2 = hca$se,
    n1 = n_savi, n2 = hca$n
  )
}

cohort <- data.frame(
  group = c("edgeR QLF", "edgeR Wald"),
  n = n_savi,
  log_mu = c(log_mu, log_mu),
  se = c(se_qlf, se_wald),
  hca_log_mu = hca$log_mu,
  hca_se = hca$se,
  p_value = c(
    welch_one(log_mu, se_qlf)$p_value,
    welch_one(log_mu, se_wald)$p_value
  )
)

if (isTRUE(can_fit_brms)) {
  cohort <- rbind(
    cohort,
    data.frame(
      group = c("brms NB", "brms ZINB"),
      n = n_savi,
      log_mu = c(nb_savi$log_mu, zinb_savi$log_mu),
      se = c(nb_savi$se, zinb_savi$se),
      hca_log_mu = hca$log_mu,
      hca_se = hca$se,
      p_value = c(
        welch_one(nb_savi$log_mu, nb_savi$se)$p_value,
        welch_one(zinb_savi$log_mu, zinb_savi$se)$p_value
      )
    )
  )
}

cohort
```

## Plots

```{r plot-hca, eval = can_run, fig.asp = 0.7}
plot_hca_draws(
  draws,
  query_mu = cohort$log_mu,
  query_SE = cohort$se,
  query_label = as.character(cohort$group)
)
```

Point estimates and SEs side by side:

```{r plot-estimates, eval = can_run, fig.asp = 0.55}
cohort$group <- factor(cohort$group, levels = cohort$group)
ggplot2::ggplot(cohort, ggplot2::aes(x = log_mu, y = group)) +
  ggplot2::geom_vline(xintercept = hca$log_mu, linetype = 2, colour = "#4C78A8") +
  ggplot2::geom_errorbarh(
    ggplot2::aes(xmin = log_mu - se, xmax = log_mu + se),
    height = 0.2
  ) +
  ggplot2::geom_point(size = 2.5) +
  ggplot2::labs(
    title = paste(gene, "SAVI log(mu) estimators"),
    subtitle = "Dashed line: healthy HCA posterior mean",
    x = "log(mu)",
    y = NULL
  ) +
  ggplot2::theme_bw()
```

## Session info

```{r session-info}
sessionInfo()
```
