---
title: "Cohort expression workflow (wrappers)"
author: "Chen Zhan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cohort expression workflow (wrappers)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  for_github_md: false
knit: >
    (function(x, ...){
        proj_root <- rprojroot::find_package_root_file() |> normalizePath()
        rmarkdown::render(
            x,
            output_file = "cohort-expression-wrappers.md",
            output_format = "github_document",
            output_dir = file.path(proj_root, "vignettes"),
            knit_root_dir = proj_root,
            params = list(for_github_md = TRUE)
        )
        rmarkdown::render(
            x,
            output_format = "rmarkdown::html_vignette",
            params = list(for_github_md = FALSE)
        )
    })
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)

if (isTRUE(params$for_github_md)) {
  root_dir <- knitr::opts_knit$get("root.dir")
  if (!is.null(root_dir)) {
    knitr::opts_knit$set(output.dir = file.path(root_dir, "vignettes"))
  }
  knitr::opts_chunk$set(fig.path = "cohort-expression-wrappers_files/figure-gfm/")
}

has_seurat <- requireNamespace("Seurat", quietly = TRUE)
has_edb <- requireNamespace("org.Hs.eg.db", quietly = TRUE) &&
  requireNamespace("AnnotationDbi", quietly = TRUE)
has_brms <- requireNamespace("brms", quietly = TRUE)
can_run <- has_seurat && has_edb && has_brms

if (requireNamespace("devtools", quietly = TRUE)) {
  pkg_root <- if (file.exists("DESCRIPTION") && dir.exists("R")) {
    normalizePath(".")
  } else if (file.exists("../DESCRIPTION") && dir.exists("../R")) {
    normalizePath("..")
  } else {
    NULL
  }
  if (!is.null(pkg_root)) {
    devtools::load_all(pkg_root, quiet = TRUE)
  }
}
```

SAVI case study: *ADRB2* (`ENSG00000169252`) in disease-associated monocytes.

This vignette is the **high-level wrapper** path. Each wrapper only composes
core functions; it does not reimplement TMM, edgeR QL, brms draws, or Welch
mathematics. The transparent core path is
`vignette("cohort-expression-core", package = "posteriorHCA")`.

```text
HIGH-LEVEL WRAPPER                  CORE FUNCTIONS CALLED
──────────────────                  ─────────────────────

scale_to_hca_reference()
    ├── load_reference_sample()
    ├── merge_with_reference_sample()
    └── calculate_tmm_offset()

estimate_cohort_logmu()
    ├── extract counts / metadata
    ├── stats::model.matrix()
    └── estimate_logmu_ql()

expression_baseline_draws()
    ├── load_expression_fit()
    ├── build_newdata_grid()
    └── expression_draws()

compare_cohort_to_hca()
    ├── summarize_posterior_draws()
    └── welch_test_means()
```

## Setup

```{r libraries, eval = can_run}
library(posteriorHCA)
library(Seurat)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(purrr)

cell_type <- "monocytic"
gene_ensg <- "ENSG00000169252"
```

## Map symbols to Ensembl ids

Atlas models are keyed by ENSG. Map once with AnnotationDbi (outside
posteriorHCA), then rebuild a Seurat object that keeps `Category`.

```{r prepare-seurat, eval = can_run}
data(savi_mono)

counts <- as.matrix(Seurat::GetAssayData(savi_mono, layer = "counts"))
mapped <- AnnotationDbi::mapIds(
  org.Hs.eg.db,
  keys = rownames(counts),
  column = "ENSEMBL",
  keytype = "SYMBOL",
  multiVals = "first"
)
keep <- !is.na(mapped) & !duplicated(mapped)
counts <- counts[keep, , drop = FALSE]
rownames(counts) <- unname(mapped[keep])

meta <- savi_mono[[]][colnames(counts), , drop = FALSE]
savi_ensg <- Seurat::CreateSeuratObject(counts = counts, meta.data = meta)
```

## Wrapper workflow

`scale_to_hca_reference()` is a convenience wrapper around
`load_reference_sample()` (when `reference` is a cell-type string),
`merge_with_reference_sample()`, and `calculate_tmm_offset()`.

```{r scale, eval = can_run}
scaled <- scale_to_hca_reference(
  savi_ensg,
  reference = cell_type
)
```

`estimate_cohort_logmu()` is a convenience wrapper that constructs the design
with `model.matrix()` and delegates estimation to `estimate_logmu_ql()`.

```{r estimate, eval = can_run}
cohort_estimates <- estimate_cohort_logmu(
  scaled,
  formula = ~ 0 + Category,
  gene_ensg = gene_ensg
)
cohort_estimates
```

Intercept-only design (`~ 1`): subset to one `Category`, then call
`estimate_cohort_logmu()` with `formula = ~ 1`. The intercept is that cohort's
absolute log(μ). Offsets were already computed by `scale_to_hca_reference()`,
so the reference library is not needed in the subset.

```{r estimate-by-level, eval = can_run}
cohort_estimates_by_level <- map_dfr(
  unique(as.character(scaled$Category[scaled$sample_role == "user"])),
  function(category) {
    out <- estimate_cohort_logmu(
      subset(scaled, sample_role == "user" & Category == category),
      formula = ~ 1,
      gene_ensg = gene_ensg
    )
    out$group <- category
    out
  }
)
cohort_estimates_by_level
```

`expression_baseline_draws()` delegates model loading, query-grid construction
and posterior generation to `load_expression_fit()`, `build_newdata_grid()` and
`expression_draws()`.

```{r hca, eval = can_run}
hca_draws <- expression_baseline_draws(
  cell_type = cell_type,
  gene_ensg = gene_ensg,
  disease_groups = "Normal",
  tissue_groups = "blood",
  assay_groups = "10x Genomics 3"
)
```

`compare_cohort_to_hca()` summarizes HCA posterior draws with
`summarize_posterior_draws()` and delegates the actual statistical test to
`welch_test_means()`.

```{r compare, eval = can_run}
test_results <- compare_cohort_to_hca(
  cohort_estimates,
  hca_draws
)
test_results
```

## Plots

```{r plot-hca, eval = can_run, fig.asp = 0.7}
plot_hca_draws(
  draws = hca_draws,
  subtitle = "Normal, 10x Genomics 3 healthy baseline"
)
```

```{r plot-cohort, eval = can_run, fig.asp = 0.7}
plot_cohort_vs_hca(
  hca_draws = hca_draws,
  cohort_est = test_results,
  subtitle = "QL cohort estimates (wrapper path)",
  annotate = c("group", "p_value")
)
```

## Session info

```{r session-info}
sessionInfo()
```
