---
title: "Cohort expression workflow (core)"
author: "Chen Zhan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cohort expression workflow (core)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  for_github_md: false
knit: >
    (function(x, ...){
        proj_root <- rprojroot::find_package_root_file() |> normalizePath()
        rmarkdown::render(
            x,
            output_file = "cohort-expression-core.md",
            output_format = "github_document",
            output_dir = file.path(proj_root, "vignettes"),
            knit_root_dir = proj_root,
            params = list(for_github_md = TRUE)
        )
        rmarkdown::render(
            x,
            output_format = "rmarkdown::html_vignette",
            params = list(for_github_md = FALSE)
        )
    })
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)

if (isTRUE(params$for_github_md)) {
  root_dir <- knitr::opts_knit$get("root.dir")
  if (!is.null(root_dir)) {
    knitr::opts_knit$set(output.dir = file.path(root_dir, "vignettes"))
  }
  knitr::opts_chunk$set(fig.path = "cohort-expression-core_files/figure-gfm/")
}

has_seurat <- requireNamespace("Seurat", quietly = TRUE)
has_edb <- requireNamespace("org.Hs.eg.db", quietly = TRUE) &&
  requireNamespace("AnnotationDbi", quietly = TRUE)
has_brms <- requireNamespace("brms", quietly = TRUE)
can_run <- has_seurat && has_edb && has_brms

if (requireNamespace("devtools", quietly = TRUE)) {
  pkg_root <- if (file.exists("DESCRIPTION") && dir.exists("R")) {
    normalizePath(".")
  } else if (file.exists("../DESCRIPTION") && dir.exists("../R")) {
    normalizePath("..")
  } else {
    NULL
  }
  if (!is.null(pkg_root)) {
    devtools::load_all(pkg_root, quiet = TRUE)
  }
}
```

SAVI case study: *ADRB2* (`ENSG00000169252`) in disease-associated monocytes.

This vignette is the **core** path: each statistical step is written out
explicitly. It does **not** call the high-level wrappers
`estimate_cohort_logmu()`, `expression_baseline_draws()`, or
`compare_cohort_to_hca()`. The concise wrapper path is
`vignette("cohort-expression-wrappers", package = "posteriorHCA")`.

Workflow:

1. Scale user libraries to one HCA reference (TMM offsets)
2. Build a design with `model.matrix()` (`~ 0 + Category` or intercept-only `~ 1`)
3. Estimate user expression means with `estimate_logmu_ql()`
4. Load the HCA expression model and draw posteriors
5. Summarise draws and run a Welch test

## Setup

```{r libraries, eval = can_run}
library(posteriorHCA)
library(Seurat)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(purrr)

cell_type <- "monocytic"
gene_ensg <- "ENSG00000169252"
```

## Prepare counts with Ensembl gene ids

Atlas models are keyed by ENSG. Map gene symbols once with AnnotationDbi
before the posteriorHCA workflow.

```{r load-savi-mono, eval = can_run}
data(savi_mono)

user_counts <- as.matrix(Seurat::GetAssayData(savi_mono, layer = "counts"))
mapped <- AnnotationDbi::mapIds(
  org.Hs.eg.db,
  keys = rownames(user_counts),
  column = "ENSEMBL",
  keytype = "SYMBOL",
  multiVals = "first"
)
keep <- !is.na(mapped) & !duplicated(mapped)
user_counts <- user_counts[keep, , drop = FALSE]
rownames(user_counts) <- unname(mapped[keep])

sample_metadata <- data.frame(
  Category = factor(savi_mono$Category[colnames(user_counts)]),
  row.names = colnames(user_counts),
  stringsAsFactors = FALSE
)

dim(user_counts)
table(sample_metadata$Category)
```

## Scale to the HCA reference

The reference library is used only to compute TMM offsets. User estimation
below uses user libraries and their offsets only.

```{r scale, eval = can_run}
reference <- load_reference_sample(cell_type)

combined_counts <- merge_with_reference_sample(
  user_counts,
  reference = reference$counts,
  reference_name = reference$sample_id
)
scaling <- calculate_tmm_offset(
  combined_counts,
  reference_name = reference$sample_id
)

user_offset <- scaling$offset[colnames(user_counts)]
```

## Estimate user expression means

Cell-means design (`~ 0 + Category`):

```{r estimate, eval = can_run}
design_matrix <- model.matrix(~ 0 + Category, data = sample_metadata)
colnames(design_matrix) <- sub("^Category", "", colnames(design_matrix))

expression_estimates <- estimate_logmu_ql(
  user_counts,
  user_offset,
  design_matrix
)

expression_estimates <- expression_estimates[
  expression_estimates$gene == gene_ensg,
  ,
  drop = FALSE
]
expression_estimates
```

Intercept-only design (`~ 1`): subset to one `Category`, fit
`model.matrix(~ 1, ...)`, and treat the intercept as that cohort's absolute
log(μ).

```{r estimate-by-level, eval = can_run}
expression_estimates_by_level <- map_dfr(
  levels(sample_metadata$Category),
  function(category) {
    sample_ids <- rownames(sample_metadata)[sample_metadata$Category == category]
    design_one <- model.matrix(
      ~ 1,
      data = sample_metadata[sample_ids, , drop = FALSE]
    )
    out <- estimate_logmu_ql(
      user_counts[, sample_ids, drop = FALSE],
      user_offset[sample_ids],
      design_one
    )
    out <- out[out$gene == gene_ensg, , drop = FALSE]
    out$group <- category
    out
  }
)
expression_estimates_by_level
```

## Load HCA model and draw the healthy baseline

```{r hca, eval = can_run}
expression_fit <- load_expression_fit(
  cell_type = cell_type,
  gene_ensg = gene_ensg
)

newdata <- build_newdata_grid(
  expression_fit,
  disease_groups = "Normal",
  tissue_groups = "blood",
  assay_groups = "10x Genomics 3"
)

posterior_draws <- expression_draws(
  expression_fit,
  newdata = newdata,
  quantity = "linpred",
  collapse = "mean"
)
```

## Summarise posterior and Welch test

```{r welch, eval = can_run}
cohort_estimate <- expression_estimates[
  expression_estimates$group == "SAVI",
  ,
  drop = FALSE
]

posterior_summary <- summarize_posterior_draws(
  posterior_draws,
  value = cohort_estimate$log_mu
)

welch_test_means(
  cohort_estimate$log_mu,
  cohort_estimate$se,
  posterior_summary$log_mu,
  posterior_summary$se,
  n1 = cohort_estimate$n,
  n2 = posterior_summary$n
)
```

Compare every cohort the same way with ordinary R:

```{r welch-all, eval = can_run}
test_results <- map_dfr(
  expression_estimates$group,
  function(group) {
    cohort_estimate <- expression_estimates[
      expression_estimates$group == group,
      ,
      drop = FALSE
    ]
    posterior_summary <- summarize_posterior_draws(
      posterior_draws,
      value = cohort_estimate$log_mu
    )
    test <- welch_test_means(
      cohort_estimate$log_mu,
      cohort_estimate$se,
      posterior_summary$log_mu,
      posterior_summary$se,
      n1 = cohort_estimate$n,
      n2 = posterior_summary$n
    )
    data.frame(
      group = group,
      log_mu = test$mu1,
      se = test$se1,
      hca_log_mu = test$mu2,
      hca_se = test$se2,
      p_value = test$p_value,
      stringsAsFactors = FALSE
    )
  }
)
test_results
```

## Plots

```{r plot-hca, eval = can_run, fig.asp = 0.7}
plot_hca_draws(
  draws = posterior_draws,
  subtitle = "Normal, 10x Genomics 3 healthy baseline"
)
```

```{r plot-cohort, eval = can_run, fig.asp = 0.7}
plot_cohort_vs_hca(
  hca_draws = posterior_draws,
  cohort_est = test_results,
  subtitle = "QL cohort estimates",
  annotate = c("group", "p_value")
)
```

## Session info

```{r session-info}
sessionInfo()
```
