---
title: "Cohort expression workflow (native edgeR)"
author: "Chen Zhan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cohort expression workflow (native edgeR)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  for_github_md: false
knit: >
    (function(x, ...){
        proj_root <- rprojroot::find_package_root_file() |> normalizePath()
        rmarkdown::render(
            x,
            output_file = "cohort-expression-edger-native.md",
            output_format = "github_document",
            output_dir = file.path(proj_root, "vignettes"),
            knit_root_dir = proj_root,
            params = list(for_github_md = TRUE),
            envir = new.env(parent = globalenv())
        )
        rmarkdown::render(
            x,
            output_format = "rmarkdown::html_vignette",
            params = list(for_github_md = FALSE),
            envir = new.env(parent = globalenv())
        )
    })
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)

if (isTRUE(params$for_github_md)) {
  root_dir <- knitr::opts_knit$get("root.dir")
  if (!is.null(root_dir)) {
    knitr::opts_knit$set(output.dir = file.path(root_dir, "vignettes"))
  }
  knitr::opts_chunk$set(fig.path = "cohort-expression-edger-native_files/figure-gfm/")
}

can_run <- requireNamespace("Seurat", quietly = TRUE) &&
  requireNamespace("org.Hs.eg.db", quietly = TRUE) &&
  requireNamespace("AnnotationDbi", quietly = TRUE) &&
  requireNamespace("brms", quietly = TRUE) &&
  requireNamespace("edgeR", quietly = TRUE) &&
  requireNamespace("limma", quietly = TRUE)

if (requireNamespace("devtools", quietly = TRUE)) {
  pkg_root <- if (file.exists("DESCRIPTION") && dir.exists("R")) {
    normalizePath(".")
  } else if (file.exists("../DESCRIPTION") && dir.exists("../R")) {
    normalizePath("..")
  } else {
    NULL
  }
  if (!is.null(pkg_root)) {
    devtools::load_all(pkg_root, quiet = TRUE)
  }
}
```

SAVI case study: *ADRB2* (`ENSG00000169252`) in disease-associated monocytes.

Same analysis as `vignette("cohort-expression-core")`, but with the
edgeR / limma steps written out. Only `load_reference_sample()` and
`merge_with_reference_sample()` are used from posteriorHCA before the HCA
model query.

```text
USER + HCA counts
  → TMMwsp
  → offset = log(E_user)
  → glmQLFit → makeContrasts → glmQLFTest
  → estimate + QLF SE
  → + log(E_HCA)
  → compare to HCA posterior
```

## Setup

```{r libraries, eval = can_run}
library(posteriorHCA)
library(Seurat)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(edgeR)
library(limma)

cell_type <- "monocytic"
gene <- "ENSG00000169252"
ref_name <- "___hca_ref_sample___"
```

## Counts

```{r load-savi-mono, eval = can_run}
data(savi_mono, overwrite = TRUE)

counts <- as.matrix(Seurat::GetAssayData(savi_mono, layer = "counts"))
rn <- rownames(counts)

if (!all(grepl("^ENSG", rn))) {
  known <- rn %in% AnnotationDbi::keys(org.Hs.eg.db, keytype = "SYMBOL")
  counts <- counts[known, , drop = FALSE]
  ensembl <- AnnotationDbi::mapIds(
    org.Hs.eg.db,
    keys = rownames(counts),
    column = "ENSEMBL",
    keytype = "SYMBOL",
    multiVals = "first"
  )
  ok <- !is.na(ensembl) & !duplicated(ensembl)
  counts <- counts[ok, , drop = FALSE]
  rownames(counts) <- unname(ensembl[ok])
}

savi_mono <- Seurat::CreateSeuratObject(
  counts = counts,
  meta.data = savi_mono@meta.data
)
```

## HCA reference

```{r reference, eval = can_run}
ref <- load_reference_sample(cell_type)
counts <- merge_with_reference_sample(
  counts,
  reference = ref$counts,
  reference_name = ref_name
)
```

The reference is only for joint TMM. Drop it before fitting.

## TMMwsp scaling

```{r scale, eval = can_run}
nf <- calcNormFactors(
  counts,
  refColumn = match(ref_name, colnames(counts)),
  method = "TMMwsp"
)
eff <- colSums(counts) * nf

log_E <- log(eff[colnames(counts) != ref_name])
log_E_hca <- log(eff[[ref_name]])
```

## USER model

```{r fit, eval = can_run}
y <- counts[, colnames(counts) != ref_name, drop = FALSE]

offset <- matrix(log_E, nrow = nrow(y), ncol = ncol(y), byrow = TRUE)
design <- model.matrix(
  ~ 0 + Category,
  data = savi_mono@meta.data[colnames(y), , drop = FALSE]
)

fit <- glmQLFit(y, design = design, offset = offset, prior.count = 0, robust = TRUE)

contrast <- makeContrasts(CategorySAVI, levels = design)
```

`CategorySAVI` is the absolute SAVI group log mean under `~ 0 + Category`.

## Estimate and QLF-implied SE

`glmQLFTest()` reports `logFC` on the log2 scale. Convert to natural log.
The SE below is implied by the one-dimensional QL F statistic, not a
coefficient-covariance SE.

```{r qlf, eval = can_run}
qlf <- glmQLFTest(fit, contrast = contrast)

estimate <- qlf$table[gene, "logFC"] * log(2)
se_qlf <- abs(qlf$table[gene, "logFC"]) / sqrt(qlf$table[gene, "F"]) * log(2)

# Absolute estimand → place on HCA scale (SE unchanged).
log_mu <- estimate + log_E_hca
```

## HCA posterior

```{r hca, eval = can_run}
fit_hca <- load_expression_fit(cell_type = cell_type, gene_ensg = gene)
newdata <- build_newdata_grid(
  fit_hca,
  disease_groups = "Normal",
  tissue_groups = "blood",
  assay_groups = "10x Genomics 3"
)
draws <- expression_draws(
  fit_hca,
  newdata = newdata,
  quantity = "linpred",
  marginalise = "mean"
)
```

## Plot

```{r plot-hca, eval = can_run, fig.asp = 0.7}
plot_hca_draws(
  draws,
  query_mu = log_mu,
  query_SE = se_qlf,
  query_label = "SAVI"
)
```

## Session info

```{r session-info}
sessionInfo()
```
