---
title: "Cohort expression workflow"
author: "Chen Zhan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cohort expression workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  run_bootstrap: true
knit: >
    (function(x, ...){
        proj_root <- rprojroot::find_package_root_file() |> normalizePath()
        rmarkdown::render(
            x,
            output_format = "rmarkdown::html_vignette",
            params = list(run_bootstrap = TRUE)
        )
        rmarkdown::render(
            x,
            output_file = "cohort-expression-workflow.md",
            output_format = "github_document",
            output_dir = file.path(proj_root, "vignettes"),
            knit_root_dir = proj_root,
            params = list(run_bootstrap = TRUE)
        )
    })
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)

has_seurat <- requireNamespace("Seurat", quietly = TRUE)
has_edb <- requireNamespace("org.Hs.eg.db", quietly = TRUE) &&
  requireNamespace("AnnotationDbi", quietly = TRUE)
has_brms <- requireNamespace("brms", quietly = TRUE)
can_run <- has_seurat && has_edb && has_brms

if (requireNamespace("devtools", quietly = TRUE)) {
  pkg_root <- if (file.exists("DESCRIPTION") && dir.exists("R")) {
    normalizePath(".")
  } else if (file.exists("../DESCRIPTION") && dir.exists("../R")) {
    normalizePath("..")
  } else {
    NULL
  }
  if (!is.null(pkg_root)) {
    devtools::load_all(pkg_root, quiet = TRUE)
  }
}
```

This vignette walks through the **cohort vs healthy HCA baseline** workflow
using the SAVI case study and *ADRB2* in disease-associated monocytes. The
steps mirror `examples/savi_adrb2_workflow.R`:

1. Prepare pseudobulk counts (`savi_mono`)
2. Align cohort libraries to an HCA reference sample
3. Estimate cohort log(μ) with edgeR QL
4. Draw a healthy HCA posterior from pre-trained `brms` models
5. Test cohorts with a Welch *t*-test against the HCA draws
6. Optionally bootstrap cohort log(μ) and repeat the test
7. Plot healthy baseline and cohort comparisons

## Setup

```{r libraries, eval = can_run}
library(posteriorHCA)
library(Seurat)
library(ggplot2)
```

The workflow needs **Seurat**, **brms**, and gene-id mapping packages
(`org.Hs.eg.db`, `AnnotationDbi`). Chunks below run only when those are
available (`can_run = TRUE`).

## Prepare `savi_mono`

The SAVI study ([de Cevins *et al.*, 2023](https://doi.org/10.1016/j.xcrm.2023.101333))
includes pseudobulk RNA-seq from PBMC cell types. We focus on
**disease-associated monocytes** and the gene **ADRB2**.

The full pseudobulk object can be built from GEO accession
[GSE226598](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE226598)
(see the case-study `prepare_data.R` script). The code below shows how
`savi_mono` is defined:

```{r savi-mono-code, eval = FALSE}
savi_path <- "GSE226598_SAVI_pseudobulk_Sample_CellType.rds"
savi <- readRDS(savi_path)
savi_mono <- subset(savi, subset = CellType == "17. Disease-associated monocytes")
```

For this vignette we load the bundled example object:

```{r load-savi-mono, eval = can_run}
data(savi_mono)
dim(savi_mono)
table(savi_mono$Category)
```

Gene symbols are harmonised to Ensembl IDs so cohort counts and HCA models
use the same row names:

```{r harmonise, eval = can_run}
savi_mono <- harmonise_gene_ids(savi_mono, id_type = "symbol")
```

## Align to the HCA monocytic reference

`scale_to_hca_reference()` appends one HCA reference library, computes TMM
normalisation factors, and stores alignment metadata (`hca_offset`,
`sample_role`, etc.) on each sample.

```{r align, eval = can_run}
cell_type <- "monocytic"
ref <- get_reference_sample_ready(cell_type)
aligned <- scale_to_hca_reference(savi_mono, ref)

table(aligned$sample_role)
```

## Estimate cohort log(μ)

Cohort point estimates use edgeR quasi-likelihood standard errors at the
atlas offset-zero scale:

```{r estimate, eval = can_run}
est <- estimate_cohort_logmu(
  aligned,
  group = "Category",
  genes = "ADRB2"
)
est
```

## Healthy HCA baseline draws

Load the pre-trained expression model for monocytic *ADRB2* and draw from the
healthy (`Normal`) 10x Genomics 3 reference:

```{r hca-draws, eval = can_run}
fit <- load_expr_fit(cell_type = cell_type, gene = "ADRB2")

grid <- build_newdata_grid(
  fit,
  disease_groups = "Normal",
  assay_groups = "10x Genomics 3"
)

hca_res <- expr_draws(
  fit,
  newdata = grid,
  quantity = "linpred",
  collapse = "mean"
)

round(c(mean = mean(hca_res$draws), sd = sd(hca_res$draws)), 3)
```

`expr_predict()` is a convenience wrapper that also returns a density plot for
count-scale predictions:

```{r hca-predict, eval = can_run, fig.asp = 0.7}
hca_pred <- expr_predict(
  cell_type = cell_type,
  gene = "ADRB2",
  disease_groups = "Normal",
  assay_groups = "10x Genomics 3",
  quantity = "predict",
  collapse = "mean"
)
hca_pred$plot
```

## Test cohorts against the healthy baseline (QL)

`welch_t_test_cohort_hca()` compares each cohort log(μ) estimate to the HCA
posterior using a Welch *t*-test. The reference group is excluded by default.

```{r welch, eval = can_run}
test_results <- welch_t_test_cohort_hca(
  cohort_est = est,
  hca_draws = hca_res
)
test_results
```

## Bootstrap cohort log(μ) (optional)

For a non-parametric cohort uncertainty estimate, `bootstrap_cohort_logmu_batch()`
mirrors the QL workflow: one call bootstraps every cohort and returns a table
compatible with testing and plotting.

```{r bootstrap, eval = can_run && isTRUE(params$run_bootstrap)}
boot_est <- bootstrap_cohort_logmu_batch(
  aligned,
  group = "Category",
  genes = "ADRB2",
  cohort_est = est,
  hca_draws = hca_res,
  n_boot = 200L,
  seed = 42L
)
boot_est

boot_test <- welch_t_test_cohort_hca(
  cohort_est = boot_est,
  hca_draws = hca_res
)
boot_test
```

## Visualise results

`plot_hca_draws()` shows the healthy HCA posterior alone.
`plot_cohort_vs_hca()` overlays cohort estimates and accepts either Welch test
output or cohort estimate tables (QL or bootstrap).

```{r plot-hca, eval = can_run, fig.asp = 0.7}
plot_hca_draws(
  draws = hca_res,
  subtitle = "Normal, 10x Genomics 3 healthy baseline"
)
```

```{r plot-cohort-ql, eval = can_run, fig.asp = 0.7}
plot_cohort_vs_hca(
  hca_draws = hca_res,
  test_results = test_results,
  subtitle = "QL cohort estimates",
  annotate = c("group", "p_value", "direction")
)
```

```{r plot-cohort-boot, eval = can_run && isTRUE(params$run_bootstrap), fig.asp = 0.7}
plot_cohort_vs_hca(
  hca_draws = hca_res,
  test_results = boot_test,
  subtitle = "Bootstrap cohort estimates",
  annotate = c("group", "p_value", "direction", "method")
)
```

You can also pass `cohort_est = est` or `cohort_est = boot_est` directly to
`plot_cohort_vs_hca()` when test statistics are not needed on the plot.

## Session info

```{r session-info}
sessionInfo()
```
