---
title: "Cohort expression workflow (native edgeR)"
author: "Chen Zhan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cohort expression workflow (native edgeR)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  for_github_md: false
knit: >
    (function(x, ...){
        proj_root <- rprojroot::find_package_root_file() |> normalizePath()
        rmarkdown::render(
            x,
            output_file = "cohort-expression-edger-native.md",
            output_format = "github_document",
            output_dir = file.path(proj_root, "vignettes"),
            knit_root_dir = proj_root,
            params = list(for_github_md = TRUE),
            envir = new.env(parent = globalenv())
        )
        rmarkdown::render(
            x,
            output_format = "rmarkdown::html_vignette",
            params = list(for_github_md = FALSE),
            envir = new.env(parent = globalenv())
        )
    })
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)

if (isTRUE(params$for_github_md)) {
  root_dir <- knitr::opts_knit$get("root.dir")
  if (!is.null(root_dir)) {
    knitr::opts_knit$set(output.dir = file.path(root_dir, "vignettes"))
  }
  knitr::opts_chunk$set(fig.path = "cohort-expression-edger-native_files/figure-gfm/")
}

can_run <- requireNamespace("Seurat", quietly = TRUE) &&
  requireNamespace("org.Hs.eg.db", quietly = TRUE) &&
  requireNamespace("AnnotationDbi", quietly = TRUE) &&
  requireNamespace("brms", quietly = TRUE) &&
  requireNamespace("edgeR", quietly = TRUE) &&
  requireNamespace("limma", quietly = TRUE)

if (requireNamespace("devtools", quietly = TRUE)) {
  pkg_root <- if (file.exists("DESCRIPTION") && dir.exists("R")) {
    normalizePath(".")
  } else if (file.exists("../DESCRIPTION") && dir.exists("../R")) {
    normalizePath("..")
  } else {
    NULL
  }
  if (!is.null(pkg_root)) {
    devtools::load_all(pkg_root, quiet = TRUE)
  }
}
```

SAVI case study: *ADRB2* (`ENSG00000169252`) in disease-associated monocytes.

Same analysis as `vignette("cohort-expression-core")`, with edgeR / limma
written out. Before the HCA query, only `load_reference_sample()` and
`merge_with_reference_sample()` come from posteriorHCA.

```text
USER + HCA counts → TMMwsp → offset = log(E_user)
  → glmQLFit → makeContrasts → glmQLFTest
  → estimate + QLF SE → + log(E_HCA) → HCA posterior
```

## Setup

```{r libraries, eval = can_run}
library(posteriorHCA)
library(Seurat)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(edgeR)
library(limma)

cell_type <- "monocytic"
gene <- "ENSG00000169252"
ref_name <- "___hca_ref_sample___"
```

## Map gene name to Ensembl ids

```{r load-savi-mono, eval = can_run}
data(savi_mono, overwrite = TRUE)

ensembl <- mapIds(
  org.Hs.eg.db,
  keys = rownames(savi_mono),
  column = "ENSEMBL",
  keytype = "SYMBOL",
  multiVals = "first"
)
ok <- !is.na(ensembl) & !duplicated(ensembl)
savi_mono <- savi_mono[ok, , drop = FALSE]
rownames(savi_mono) <- ensembl[ok] |> unname()
  
```

## Joint TMMwsp with one HCA library

```{r scale, eval = can_run}
ref <- load_reference_sample(cell_type)
counts <- merge_with_reference_sample(
  GetAssayData(savi_mono, layer = "counts"),
  reference = ref$counts,
  reference_name = ref_name
)

eff <- colSums(counts) * calcNormFactors(
  counts,
  refColumn = match(ref_name, colnames(counts)),
  method = "TMMwsp"
)

# USER libraries only → store on savi_mono; HCA size kept for the later shift
savi_mono$eff <- unname(eff[match(colnames(savi_mono), names(eff))])
savi_mono$log_E <- log(savi_mono$eff)
log_E_hca <- log(eff[[ref_name]])
```

The HCA column is only for TMM. Fitting uses `savi_mono` alone.

## Fit

```{r fit, eval = can_run}
design <- model.matrix(~ 0 + Category, data = savi_mono[[]])

fit <- glmQLFit(
  GetAssayData(savi_mono, layer = "counts"),
  design = design,
  offset = savi_mono$log_E,
  prior.count = 0,
  robust = TRUE
)

contrast <- makeContrasts(CategorySAVI, levels = design)
```

`CategorySAVI` is the absolute SAVI group log mean under `~ 0 + Category`.

## Estimate and QLF-implied SE

`logFC` is log2; convert to natural log. The SE is implied by the
one-dimensional QL F statistic (not a coefficient-covariance SE).

```{r qlf, eval = can_run}
qlf <- glmQLFTest(fit, contrast = contrast)

estimate <- qlf$table[gene, "logFC"] * log(2)
se_qlf <- abs(qlf$table[gene, "logFC"]) / sqrt(qlf$table[gene, "F"]) * log(2)

# Absolute estimand → HCA scale (SE unchanged)
log_mu <- estimate + log_E_hca
```

## HCA posterior

```{r hca, eval = can_run}
fit_hca <- load_expression_fit(cell_type = cell_type, gene_ensg = gene)
draws <- expression_draws(
  fit_hca,
  newdata = build_newdata_grid(fit_hca, tissue_groups = "blood" )
)
hca <- summarize_posterior_draws(draws)
```

## Welch test

```{r welch, eval = can_run}
test_res <-
  welch_test_means(
    mu1 = log_mu,
    se1 = se_qlf,
    n1 = sum(savi_mono$Category == "SAVI"),
    mu2 = hca$log_mu,
    se2 = hca$se,
    n2 = hca$n
  )
```

## Plot

```{r plot-hca, eval = can_run, fig.asp = 0.7}
plot_hca_draws(
  draws,
  query_mu = log_mu,
  query_SE = se_qlf,
  query_label = paste0("SAVI", '\n', 'p-value:', round(test_res$p_value, digits = 2))
)
```

## Session info

```{r session-info}
sessionInfo()
```
