---
title: "Cohort expression workflow (native edgeR)"
author: "Chen Zhan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cohort expression workflow (native edgeR)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  for_github_md: false
knit: >
    (function(x, ...){
        proj_root <- rprojroot::find_package_root_file() |> normalizePath()
        rmarkdown::render(
            x,
            output_file = "cohort-expression-edger-native.md",
            output_format = "github_document",
            output_dir = file.path(proj_root, "vignettes"),
            knit_root_dir = proj_root,
            params = list(for_github_md = TRUE)
        )
        rmarkdown::render(
            x,
            output_format = "rmarkdown::html_vignette",
            params = list(for_github_md = FALSE)
        )
    })
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)

if (isTRUE(params$for_github_md)) {
  root_dir <- knitr::opts_knit$get("root.dir")
  if (!is.null(root_dir)) {
    knitr::opts_knit$set(output.dir = file.path(root_dir, "vignettes"))
  }
  knitr::opts_chunk$set(fig.path = "cohort-expression-edger-native_files/figure-gfm/")
}

can_run <- requireNamespace("Seurat", quietly = TRUE) &&
  requireNamespace("org.Hs.eg.db", quietly = TRUE) &&
  requireNamespace("AnnotationDbi", quietly = TRUE) &&
  requireNamespace("brms", quietly = TRUE) &&
  requireNamespace("edgeR", quietly = TRUE) &&
  requireNamespace("limma", quietly = TRUE)

if (requireNamespace("devtools", quietly = TRUE)) {
  pkg_root <- if (file.exists("DESCRIPTION") && dir.exists("R")) {
    normalizePath(".")
  } else if (file.exists("../DESCRIPTION") && dir.exists("../R")) {
    normalizePath("..")
  } else {
    NULL
  }
  if (!is.null(pkg_root)) {
    devtools::load_all(pkg_root, quiet = TRUE)
  }
}
```

SAVI case study: *ADRB2* (`ENSG00000169252`) in disease-associated monocytes.

Same analysis as `vignette("cohort-expression-core")`, but with the
edgeR / limma steps written out. Only `load_reference_sample()` and
`merge_with_reference_sample()` are used from posteriorHCA before the HCA
model query.

```text
USER + HCA counts
  → TMMwsp
  → offset = log(E_user)
  → glmQLFit → makeContrasts → glmQLFTest
  → estimate + QLF SE
  → + log(E_HCA)
  → compare to HCA posterior
```

## Setup

```{r libraries, eval = can_run}
library(posteriorHCA)
library(Seurat)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(edgeR)
library(limma)

cell_type <- "monocytic"
gene <- "ENSG00000169252"
```

## Counts

```{r load-savi-mono, eval = can_run}
data(savi_mono)

counts <- as.matrix(Seurat::GetAssayData(savi_mono, layer = "counts"))
ensembl <- AnnotationDbi::mapIds(
  org.Hs.eg.db,
  keys = rownames(counts),
  column = "ENSEMBL",
  keytype = "SYMBOL",
  multiVals = "first"
)
ok <- !is.na(ensembl) & !duplicated(ensembl)
counts <- counts[ok, , drop = FALSE]
rownames(counts) <- unname(ensembl[ok])

meta <- data.frame(
  Category = factor(savi_mono$Category[colnames(counts)]),
  Experiment = factor(savi_mono$Experiment[colnames(counts)]),
  row.names = colnames(counts)
)
table(meta$Category, meta$Experiment)
```

## HCA reference

```{r reference, eval = can_run}
ref <- load_reference_sample(cell_type)
ref_id <- ref$sample_id

counts <- merge_with_reference_sample(
  counts,
  reference = ref$counts,
  reference_name = ref_id
)
```

The reference is only for joint TMM. Drop it before fitting.

## TMMwsp scaling

```{r scale, eval = can_run}
nf <- calcNormFactors(counts, refColumn = match(ref_id, colnames(counts)), method = "TMMwsp")
eff <- colSums(counts) * nf

user <- setdiff(colnames(counts), ref_id)
log_E <- log(eff[user])
log_E_hca <- log(eff[[ref_id]])
```

## USER model

```{r fit, eval = can_run}
y <- counts[, user, drop = FALSE]
meta <- meta[user, , drop = FALSE]

offset <- matrix(log_E, nrow = nrow(y), ncol = ncol(y), byrow = TRUE)
design <- model.matrix(~ 0 + Category, data = meta)
colnames(design)

fit <- glmQLFit(y, design = design, offset = offset, prior.count = 0, robust = TRUE)

contrast <- makeContrasts(CategorySAVI, levels = design)
contrast
```

`CategorySAVI` is the absolute SAVI group log mean under `~ 0 + Category`.

## Estimate and QLF-implied SE

`glmQLFTest()` reports `logFC` on the log2 scale. Convert to natural log.
The SE below is implied by the one-dimensional QL F statistic, not a
coefficient-covariance SE.

```{r qlf, eval = can_run}
qlf <- glmQLFTest(fit, contrast = contrast)
tab <- qlf$table[gene, ]

estimate <- tab$logFC * log(2)
se_qlf <- abs(tab$logFC) / sqrt(tab$F) * log(2)

# Absolute estimand → place on HCA scale (SE unchanged).
log_mu <- estimate + log_E_hca
n_savi <- sum(meta$Category == "SAVI")

data.frame(gene, estimate, log_mu, se_qlf, n = n_savi)
```

## Package Wald SE (for comparison)

```{r wald, eval = can_run}
wald <- estimate_ql(
  counts = y,
  offset = log_E,
  metadata = meta,
  formula = ~ 0 + Category,
  contrast = "CategorySAVI"
)
wald <- wald[wald$gene == gene, ]
se_wald <- wald$se

data.frame(
  estimate_qlf = estimate,
  estimate_wald = wald$estimate,
  se_qlf,
  se_wald,
  se_ratio = se_qlf / se_wald
)
```

## HCA posterior

```{r hca, eval = can_run}
fit_hca <- load_expression_fit(cell_type = cell_type, gene_ensg = gene)
newdata <- build_newdata_grid(
  fit_hca,
  disease_groups = "Normal",
  tissue_groups = "blood",
  assay_groups = "10x Genomics 3"
)
draws <- expression_draws(fit_hca, newdata = newdata, quantity = "linpred", collapse = "mean")
hca <- summarize_posterior_draws(draws, value = log_mu)
```

## Welch tests

Same `log_mu`, two SEs.

```{r welch, eval = can_run}
welch_qlf <- welch_test_means(
  mu1 = log_mu, se1 = se_qlf,
  mu2 = hca$log_mu, se2 = hca$se,
  n1 = n_savi, n2 = hca$n
)
welch_wald <- welch_test_means(
  mu1 = log_mu, se1 = se_wald,
  mu2 = hca$log_mu, se2 = hca$se,
  n1 = n_savi, n2 = hca$n
)

cohort <- data.frame(
  group = c("QLF SE", "Wald SE"),
  n = n_savi,
  log_mu = log_mu,
  se = c(se_qlf, se_wald),
  hca_log_mu = hca$log_mu,
  hca_se = hca$se,
  p_value = c(welch_qlf$p_value, welch_wald$p_value)
)
cohort
```

## Plots

```{r plot-hca, eval = can_run, fig.asp = 0.7}
plot_hca_draws(draws, subtitle = "Normal, 10x Genomics 3")
```

```{r plot-cohort, eval = can_run, fig.asp = 0.7}
plot_cohort_vs_hca(
  hca_draws = draws,
  cohort_est = cohort,
  subtitle = "SAVI mean: QLF SE vs Wald SE",
  annotate = c("group", "p_value")
)
```

## Session info

```{r session-info}
sessionInfo()
```
