---
title: "Cohort expression workflow (wrappers)"
author: "Chen Zhan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cohort expression workflow (wrappers)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  for_github_md: false
knit: >
    (function(x, ...){
        proj_root <- rprojroot::find_package_root_file() |> normalizePath()
        rmarkdown::render(
            x,
            output_file = "cohort-expression-wrappers.md",
            output_format = "github_document",
            output_dir = file.path(proj_root, "vignettes"),
            knit_root_dir = proj_root,
            params = list(for_github_md = TRUE)
        )
        rmarkdown::render(
            x,
            output_format = "rmarkdown::html_vignette",
            params = list(for_github_md = FALSE)
        )
    })
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)

if (isTRUE(params$for_github_md)) {
  root_dir <- knitr::opts_knit$get("root.dir")
  if (!is.null(root_dir)) {
    knitr::opts_knit$set(output.dir = file.path(root_dir, "vignettes"))
  }
  knitr::opts_chunk$set(fig.path = "cohort-expression-wrappers_files/figure-gfm/")
}

has_seurat <- requireNamespace("Seurat", quietly = TRUE)
has_edb <- requireNamespace("org.Hs.eg.db", quietly = TRUE) &&
  requireNamespace("AnnotationDbi", quietly = TRUE)
has_brms <- requireNamespace("brms", quietly = TRUE)
can_run <- has_seurat && has_edb && has_brms

if (requireNamespace("devtools", quietly = TRUE)) {
  pkg_root <- if (file.exists("DESCRIPTION") && dir.exists("R")) {
    normalizePath(".")
  } else if (file.exists("../DESCRIPTION") && dir.exists("../R")) {
    normalizePath("..")
  } else {
    NULL
  }
  if (!is.null(pkg_root)) {
    devtools::load_all(pkg_root, quiet = TRUE)
  }
}
```

**Wrapper** path for the SAVI case study (*ADRB2* in disease-associated
monocytes, PBMC / blood). Same marked steps as the core vignette, with
Seurat / formula / gene-id handling. See
`vignette("cohort-expression-core", package = "posteriorHCA")` for the
matrix-level cores.

Steps:

0. Align to one HCA reference library (`scale_to_hca_reference`)  
1. Estimate cohort log(μ) (`estimate_cohort_logmu`)  
2. Load the healthy expression fit (`load_expr_fit`)  
3. Draw the HCA baseline (`expr_predict`)  
4–5. Welch test vs baseline (`welch_t_test_cohort_hca`)  

## Setup

```{r libraries, eval = can_run}
library(posteriorHCA)
library(Seurat)
library(dplyr)

cell_type <- "monocytic"
```

## Prepare `savi_mono`

The SAVI study ([de Cevins *et al.*, 2023](https://doi.org/10.1016/j.xcrm.2023.101333))
is PBMC scRNA-seq pseudobulked by sample and cell type. We subset
**disease-associated monocytes** (17 samples: CTRL, SAVI, SAVI_treated).

```{r savi-mono-code, eval = FALSE}
savi_path <- "GSE226598_SAVI_pseudobulk_Sample_CellType.rds"
savi <- readRDS(savi_path)
savi_mono <- subset(savi, subset = CellType == "17. Disease-associated monocytes")
```

```{r load-savi-mono, eval = can_run}
data(savi_mono)
savi_mono <- harmonise_gene_ids(savi_mono, id_type = "auto")
dim(savi_mono)
table(savi_mono$Category)
```

## Step 0. Align to the HCA monocytic reference

Calls `merge_with_reference_sample()` + `calculate_tmm_offset()`, then rebuilds
the Seurat object with `hca_offset` and `sample_role`.

```{r step0-align, eval = can_run}
aligned <- scale_to_hca_reference(savi_mono, cell_type)

table(aligned$sample_role)
aligned$hca_reference_name[1]
aligned$hca_offset[aligned$hca_reference_name[1]]
```

## Step 1. Estimate cohort log(μ)

`estimate_cohort_logmu()` builds the design from a formula, fits all genes
via `estimate_logmu_ql()`, then filters to `genes`. Dispersion still uses the
full matrix. Prefer a cell-means design such as `~ 0 + Category`.

```{r step1-estimate, eval = can_run}
est <- estimate_cohort_logmu(
  aligned,
  formula = ~ 0 + Category,
  genes = "ADRB2"
)
est
```

### Optional: single-cohort `~ 1`

Subset to one `Category`, fit with `formula = ~ 1`, and treat the intercept as
that cohort's absolute log(μ).

```{r step1-by-level, eval = can_run}
est_by_level <- purrr::map_dfr(
  levels(aligned$Category),
  .f = function(x) {
    estimate_cohort_logmu(
      subset(aligned, Category == x),
      formula = ~ 1,
      genes = "ADRB2"
    ) %>%
      dplyr::mutate(group = x)
  }
)
est_by_level
```

## Step 2. Load the healthy expression fit

```{r step2-fit, eval = can_run}
fit <- load_expr_fit(cell_type = cell_type, gene = "ADRB2")
```

## Step 3. Draw the healthy HCA baseline

`expr_predict()` wraps grid construction and `expr_draws()`.
`quantity = "linpred"` is log(μ) on the natural-log scale.

```{r step3-predict, eval = can_run}
hca_pred <- expr_predict(
  fit = fit,
  disease_groups = "Normal",
  tissue_groups = "blood",
  assay_groups = "10x Genomics 3",
  quantity = "linpred",
  collapse = "mean"
)

round(c(mean = mean(hca_pred$draws), sd = sd(hca_pred$draws)), 3)
```

## Steps 4–5. Welch test vs the healthy baseline

`welch_t_test_cohort_hca()` loops cohorts (skips atlas `reference` by default)
and combines baseline summary + `welch_test_means()`.

```{r step45-welch, eval = can_run}
test_results <- welch_t_test_cohort_hca(
  cohort_est = est,
  hca_draws = hca_pred
)
test_results
```

## Plots

```{r plot-hca, eval = can_run, fig.asp = 0.7}
plot_hca_draws(
  draws = hca_pred,
  subtitle = "Normal, 10x Genomics 3 healthy baseline"
)
```

```{r plot-cohort, eval = can_run, fig.asp = 0.7}
plot_cohort_vs_hca(
  hca_draws = hca_pred,
  test_results = test_results,
  subtitle = "QL cohort estimates (wrapper path)",
  annotate = c("group", "p_value", "direction")
)
```

## Session info

```{r session-info}
sessionInfo()
```
