---
title: "Cohort expression workflow (core)"
author: "Chen Zhan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cohort expression workflow (core)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  for_github_md: false
knit: >
    (function(x, ...){
        proj_root <- rprojroot::find_package_root_file() |> normalizePath()
        rmarkdown::render(
            x,
            output_file = "cohort-expression-core.md",
            output_format = "github_document",
            output_dir = file.path(proj_root, "vignettes"),
            knit_root_dir = proj_root,
            params = list(for_github_md = TRUE)
        )
        rmarkdown::render(
            x,
            output_format = "rmarkdown::html_vignette",
            params = list(for_github_md = FALSE)
        )
    })
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)

if (isTRUE(params$for_github_md)) {
  root_dir <- knitr::opts_knit$get("root.dir")
  if (!is.null(root_dir)) {
    knitr::opts_knit$set(output.dir = file.path(root_dir, "vignettes"))
  }
  knitr::opts_chunk$set(fig.path = "cohort-expression-core_files/figure-gfm/")
}

has_seurat <- requireNamespace("Seurat", quietly = TRUE)
has_edb <- requireNamespace("org.Hs.eg.db", quietly = TRUE) &&
  requireNamespace("AnnotationDbi", quietly = TRUE)
has_brms <- requireNamespace("brms", quietly = TRUE)
can_run <- has_seurat && has_edb && has_brms

if (requireNamespace("devtools", quietly = TRUE)) {
  pkg_root <- if (file.exists("DESCRIPTION") && dir.exists("R")) {
    normalizePath(".")
  } else if (file.exists("../DESCRIPTION") && dir.exists("../R")) {
    normalizePath("..")
  } else {
    NULL
  }
  if (!is.null(pkg_root)) {
    devtools::load_all(pkg_root, quiet = TRUE)
  }
}
```

SAVI case study: *ADRB2* (`ENSG00000169252`) in disease-associated monocytes.

posteriorHCA compares a **user-supplied mean estimate and SE** to an HCA
posterior mean and SE via `welch_test_means()`. The optional edgeR helper
`estimate_ql()` fits a covariate-aware NB QL model and extracts a
user-defined fitted expression quantity (`estimate` + `se`). Any other
valid statistical model that produces a comparable estimate and SE can be
used instead.

This vignette writes the steps explicitly. The concise one-hot wrapper
path is `vignette("cohort-expression-wrappers", package = "posteriorHCA")`.

Workflow:

1. Jointly TMM-scale user libraries with one HCA reference
2. Fit the user cohort with edgeR QL (`formula`), then extract a target
   expression estimand (`contrast`) and SE
3. If that estimand is absolute, place it on the HCA scale with `log(E_HCA)`
4. Load the HCA expression model and summarise the posterior
5. Run `welch_test_means()`

## Setup

```{r libraries, eval = can_run}
library(posteriorHCA)
library(Seurat)
library(AnnotationDbi)
library(org.Hs.eg.db)

cell_type <- "monocytic"
gene_ensg <- "ENSG00000169252"
```

## Prepare counts with Ensembl gene ids

```{r load-savi-mono, eval = can_run}
data(savi_mono)

user_counts <- as.matrix(Seurat::GetAssayData(savi_mono, layer = "counts"))
mapped <- AnnotationDbi::mapIds(
  org.Hs.eg.db,
  keys = rownames(user_counts),
  column = "ENSEMBL",
  keytype = "SYMBOL",
  multiVals = "first"
)
keep <- !is.na(mapped) & !duplicated(mapped)
user_counts <- user_counts[keep, , drop = FALSE]
rownames(user_counts) <- unname(mapped[keep])

sample_metadata <- data.frame(
  Category = factor(savi_mono$Category[colnames(user_counts)]),
  Experiment = factor(savi_mono$Experiment[colnames(user_counts)]),
  row.names = colnames(user_counts),
  stringsAsFactors = FALSE
)

dim(user_counts)
table(sample_metadata$Category, sample_metadata$Experiment)
```

## Scale to the HCA reference

Joint TMM / TMMwsp puts user and HCA effective library sizes on one scale.
edgeR is then fitted on **user libraries only** with offset
`log(effective library size)`.

```{r scale, eval = can_run}
reference <- load_reference_sample(cell_type)

combined_counts <- merge_with_reference_sample(
  user_counts,
  reference = reference$counts,
  reference_name = reference$sample_id
)
scaling <- calculate_tmm_scaling(
  combined_counts,
  reference_name = reference$sample_id
)

user_offset <- scaling$log_effective_library_size[colnames(user_counts)]
```

## Optional edgeR helper: estimate after accounting for covariates

`estimate_ql()` is an **estimation** helper, not differential-expression
testing. The `formula` controls which effects enter the NB QL model (for
example Category and Experiment). The `contrast` selects which fitted
expression quantity to extract:

```text
estimate = c' beta
SE       = sqrt(c' Var(beta) c)
```

posteriorHCA does not interpret the contrast and does not invent a unique
"batch-adjusted mean"; the user owns that estimand. Return value:

```text
gene, contrast, estimate, se, df, dispersion
```

`estimate` is natural-log edgeR scale (`log(mu) = X beta + offset`), not
DE-table log2 `logFC`, and not automatically named `log_mu`.

### Simple group mean

Under `~ 0 + Category`, `CategorySAVI` is an absolute normalized group
mean *under that parameterisation*.

```{r estimate-simple, eval = can_run}
savi_est <- estimate_ql(
  counts = user_counts,
  offset = user_offset,
  metadata = sample_metadata,
  formula = ~ 0 + Category,
  contrast = "CategorySAVI"
)
savi_est <- savi_est[savi_est$gene == gene_ensg, , drop = FALSE]
savi_est
```

Because this contrast is an absolute mean, add the fixed HCA constant for
comparison on the matched reference scale. SE is unchanged.

```{r hca-scale-simple, eval = can_run}
user_log_mu <- savi_est$estimate + scaling$hca_log_effective_library_size
user_se <- savi_est$se
n_savi <- sum(sample_metadata$Category == "SAVI")
c(log_mu = user_log_mu, se = user_se, n = n_savi)
```

### Accounting for Experiment

Fit Category while including Experiment in the model. Different contrasts
select different fitted conditions of the **same** model — neither is the
unique "correct" adjusted mean.

```{r estimate-adjusted, eval = can_run}
exp2 <- paste0("Experiment", levels(sample_metadata$Experiment)[[2]])

# Absolute estimands: SAVI at reference Experiment vs SAVI at EXP2
batch_fit <- estimate_ql(
  counts = user_counts,
  offset = user_offset,
  metadata = sample_metadata,
  formula = ~ 0 + Category + Experiment,
  contrast = c(
    "CategorySAVI",
    paste("CategorySAVI +", exp2)
  )
)
batch_fit <- batch_fit[batch_fit$gene == gene_ensg, , drop = FALSE]
batch_fit
```

These are absolute fitted quantities, so add the fixed HCA constant for
comparison (SE unchanged), the same as in the simple group-mean case.

```{r hca-scale-adjusted, eval = can_run}
batch_fit$log_mu <- batch_fit$estimate + scaling$hca_log_effective_library_size
batch_fit[, c("gene", "contrast", "estimate", "log_mu", "se")]
```

A relative contrast does **not** get the HCA shift: if both group means
were shifted by the same `log(E_HCA)`, that constant would cancel in the
difference. Use `estimate` / `se` as returned.

```{r estimate-relative, eval = can_run}
hca_const <- scaling$hca_log_effective_library_size

rel_fit <- estimate_ql(
  counts = user_counts,
  offset = user_offset,
  metadata = sample_metadata,
  formula = ~ 0 + Category + Experiment,
  contrast = c(
    "CategorySAVI",
    "CategoryCTRL",
    "CategorySAVI - CategoryCTRL"
  )
)
rel_fit <- rel_fit[rel_fit$gene == gene_ensg, , drop = FALSE]
rel_fit

est <- setNames(rel_fit$estimate, rel_fit$contrast)
# Absolute means may be HCA-shifted; their difference is unchanged.
c(
  relative_contrast = unname(est[["CategorySAVI - CategoryCTRL"]]),
  log_mu_CategorySAVI = est[["CategorySAVI"]] + hca_const,
  log_mu_CategoryCTRL = est[["CategoryCTRL"]] + hca_const,
  difference_of_hca_shifted_absolutes =
    (est[["CategorySAVI"]] + hca_const) - (est[["CategoryCTRL"]] + hca_const)
)
```
## HCA posterior

```{r hca, eval = can_run}
expression_fit <- load_expression_fit(
  cell_type = cell_type,
  gene_ensg = gene_ensg
)

newdata <- build_newdata_grid(
  expression_fit,
  disease_groups = "Normal",
  tissue_groups = "blood",
  assay_groups = "10x Genomics 3"
)

posterior_draws <- expression_draws(
  expression_fit,
  newdata = newdata,
  quantity = "linpred",
  collapse = "mean"
)
```

## Comparison

`welch_test_means()` only needs two means and two SEs on the same scale.
Here we compare the simple SAVI group mean (HCA-shifted) to the HCA
baseline. Estimates from outside posteriorHCA can be passed the same way.

```{r welch, eval = can_run}
posterior_summary <- summarize_posterior_draws(
  posterior_draws,
  value = user_log_mu
)

welch_test_means(
  mu1 = user_log_mu,
  se1 = user_se,
  mu2 = posterior_summary$log_mu,
  se2 = posterior_summary$se,
  n1 = n_savi,
  n2 = posterior_summary$n
)
```

```{r welch-table, eval = can_run}
test <- welch_test_means(
  mu1 = user_log_mu,
  se1 = user_se,
  mu2 = posterior_summary$log_mu,
  se2 = posterior_summary$se,
  n1 = n_savi,
  n2 = posterior_summary$n
)
test_results <- data.frame(
  group = "SAVI",
  n = n_savi,
  log_mu = test$mu1,
  se = test$se1,
  hca_log_mu = test$mu2,
  hca_se = test$se2,
  p_value = test$p_value,
  stringsAsFactors = FALSE
)
test_results
```

## Plots

```{r plot-hca, eval = can_run, fig.asp = 0.7}
plot_hca_draws(
  draws = posterior_draws,
  subtitle = "Normal, 10x Genomics 3 healthy baseline"
)
```

```{r plot-cohort, eval = can_run, fig.asp = 0.7}
plot_cohort_vs_hca(
  hca_draws = posterior_draws,
  cohort_est = test_results,
  subtitle = "SAVI Category mean via edgeR formula + contrast",
  annotate = c("group", "p_value")
)
```

## Session info

```{r session-info}
sessionInfo()
```
