---
title: "Cohort expression workflow (core)"
author: "Chen Zhan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cohort expression workflow (core)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  for_github_md: false
knit: >
    (function(x, ...){
        proj_root <- rprojroot::find_package_root_file() |> normalizePath()
        rmarkdown::render(
            x,
            output_file = "cohort-expression-core.md",
            output_format = "github_document",
            output_dir = file.path(proj_root, "vignettes"),
            knit_root_dir = proj_root,
            params = list(for_github_md = TRUE)
        )
        rmarkdown::render(
            x,
            output_format = "rmarkdown::html_vignette",
            params = list(for_github_md = FALSE)
        )
    })
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)

if (isTRUE(params$for_github_md)) {
  root_dir <- knitr::opts_knit$get("root.dir")
  if (!is.null(root_dir)) {
    knitr::opts_knit$set(output.dir = file.path(root_dir, "vignettes"))
  }
  knitr::opts_chunk$set(fig.path = "cohort-expression-core_files/figure-gfm/")
}

has_seurat <- requireNamespace("Seurat", quietly = TRUE)
has_edb <- requireNamespace("org.Hs.eg.db", quietly = TRUE) &&
  requireNamespace("AnnotationDbi", quietly = TRUE)
has_brms <- requireNamespace("brms", quietly = TRUE)
can_run <- has_seurat && has_edb && has_brms

if (requireNamespace("devtools", quietly = TRUE)) {
  pkg_root <- if (file.exists("DESCRIPTION") && dir.exists("R")) {
    normalizePath(".")
  } else if (file.exists("../DESCRIPTION") && dir.exists("../R")) {
    normalizePath("..")
  } else {
    NULL
  }
  if (!is.null(pkg_root)) {
    devtools::load_all(pkg_root, quiet = TRUE)
  }
}
```

Matrix-level **core** path for the SAVI case study (*ADRB2* in disease-associated
monocytes, PBMC / blood). See
`vignette("cohort-expression-wrappers", package = "posteriorHCA")` for the
Seurat / formula wrapper path.

Steps:

0. Align to one HCA reference library (`merge_with_reference_sample` + `calculate_tmm_offset`)  
1. Estimate cohort log(μ) (`design_from_formula` + `estimate_logmu_ql`)  
2. Load the healthy expression fit (`load_expr_fit`)  
3. Draw the HCA baseline (`build_newdata_grid` + `expr_draws`)  
4. Summarise the baseline at the cohort estimate (`summarize_posterior_draws`)  
5. Welch test (`welch_test_means`)  

## Setup

```{r libraries, eval = can_run}
library(posteriorHCA)
library(Seurat)

cell_type <- "monocytic"
```

## Prepare `savi_mono`

```{r load-savi-mono, eval = can_run}
data(savi_mono)
savi_mono <- harmonise_gene_ids(savi_mono, id_type = "auto")
dim(savi_mono)
table(savi_mono$Category)
```

## Step 0. Align to the HCA monocytic reference

Reference sits at offset 0 (`offset = log(1 / multiplier)`).

```{r step0-align, eval = can_run}
ref <- load_reference_sample(cell_type)
user_mat <- as.matrix(Seurat::GetAssayData(savi_mono, layer = "counts"))

combined <- merge_with_reference_sample(
  user_mat,
  reference = ref$counts,
  reference_name = ref$sample_id
)
scaling <- calculate_tmm_offset(
  combined,
  reference_name = ref$sample_id,
  method = "TMMwsp"
)

c(
  n_genes = nrow(combined),
  n_samples = ncol(combined),
  reference_offset = unname(scaling$offset[[ref$sample_id]])
)
```

## Step 1. Estimate cohort log(μ)

`estimate_logmu_ql()` fits **all genes** once (`glmQLFit` with
`prior.count = 0`). Filter afterwards for the gene of interest.

```{r step1-estimate, eval = can_run}
meta_core <- data.frame(
  Category = c(
    as.character(savi_mono$Category[colnames(user_mat)]),
    "reference"
  ),
  row.names = colnames(combined),
  stringsAsFactors = FALSE
)
meta_core$Category <- factor(meta_core$Category)

design <- design_from_formula(~ 0 + Category, meta_core)
est_all <- estimate_logmu_ql(
  counts = combined,
  offset = scaling$offset,
  design = design,
  cell_type = cell_type
)

gene_ensg <- resolve_gene_one("ADRB2", cell_type = cell_type)
est <- est_all[est_all$gene == gene_ensg, , drop = FALSE]
est$gene_symbol <- "ADRB2"
est
```

### Optional: single-cohort `~ 1`

Subset to one `Category`, fit with `design_from_formula(~ 1, ...)`, and treat
the intercept as that cohort's absolute log(μ).

```{r step1-by-level, eval = can_run}
cats <- setdiff(levels(meta_core$Category), "reference")

est_by_level <- do.call(rbind, lapply(cats, function(x) {
  j <- which(meta_core$Category == x)
  design1 <- design_from_formula(~ 1, meta_core[j, , drop = FALSE])
  out <- estimate_logmu_ql(
    counts = combined[, j, drop = FALSE],
    offset = scaling$offset[j],
    design = design1,
    cell_type = cell_type
  )
  out <- out[out$gene == gene_ensg, , drop = FALSE]
  out$gene_symbol <- "ADRB2"
  out$group <- x
  out
}))
est_by_level
```

## Step 2. Load the healthy expression fit

```{r step2-fit, eval = can_run}
fit <- load_expr_fit(cell_type = cell_type, gene = "ADRB2")
```

## Step 3. Draw the healthy HCA baseline

Covariates fixed to Normal / blood / 10x Genomics 3.
`quantity = "linpred"` is log(μ) on the natural-log scale.

```{r step3-draws, eval = can_run}
grid <- build_newdata_grid(
  fit,
  disease_groups = "Normal",
  tissue_groups = "blood",
  assay_groups = "10x Genomics 3"
)

hca_res <- expr_draws(
  fit,
  newdata = grid,
  quantity = "linpred",
  collapse = "mean"
)

round(c(mean = mean(hca_res$draws), sd = sd(hca_res$draws)), 3)
```

## Step 4. Summarise the baseline at the cohort estimate

```{r step4-summarise, eval = can_run}
cohort <- cohort_estimate_at(est, group = "SAVI")
baseline <- summarize_posterior_draws(hca_res, value = cohort$mu)
baseline
```

## Step 5. Welch test

```{r step5-welch, eval = can_run}
welch_test_means(
  cohort$mu, cohort$se,
  baseline$mean, baseline$sd,
  n1 = cohort$n, n2 = baseline$n
)
```

Compare all non-reference cohorts with the same helpers:

```{r step5-welch-all, eval = can_run}
test_results <- welch_t_test_cohort_hca(
  cohort_est = est,
  hca_draws = hca_res
)
test_results
```

## Plots

```{r plot-hca, eval = can_run, fig.asp = 0.7}
plot_hca_draws(
  draws = hca_res,
  subtitle = "Normal, 10x Genomics 3 healthy baseline"
)
```

```{r plot-cohort, eval = can_run, fig.asp = 0.7}
plot_cohort_vs_hca(
  hca_draws = hca_res,
  test_results = test_results,
  subtitle = "QL cohort estimates (core path)",
  annotate = c("group", "p_value", "direction")
)
```

## Session info

```{r session-info}
sessionInfo()
```
