---
title: "Cohort expression workflow"
author: "Chen Zhan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cohort expression workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  for_github_md: false
knit: >
    (function(x, ...){
        # RStudio Knit opens whatever this function *returns*. Render HTML
        # last (or return its path explicitly) so the viewer shows HTML; the
        # github_document pass only writes the companion .md for GitHub.
        proj_root <- rprojroot::find_package_root_file() |> normalizePath()
        rmarkdown::render(
            x,
            output_file = "cohort-expression-workflow.md",
            output_format = "github_document",
            output_dir = file.path(proj_root, "vignettes"),
            knit_root_dir = proj_root,
            params = list(for_github_md = TRUE)
        )
        rmarkdown::render(
            x,
            output_format = "rmarkdown::html_vignette",
            params = list(for_github_md = FALSE)
        )
    })
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)

# Relative figure paths for github_document output (see rmarkdown#2473).
if (isTRUE(params$for_github_md)) {
  root_dir <- knitr::opts_knit$get("root.dir")
  if (!is.null(root_dir)) {
    knitr::opts_knit$set(
      output.dir = file.path(root_dir, "vignettes")
    )
  }
  knitr::opts_chunk$set(
    fig.path = "cohort-expression-workflow_files/figure-gfm/"
  )
}

has_seurat <- requireNamespace("Seurat", quietly = TRUE)
has_edb <- requireNamespace("org.Hs.eg.db", quietly = TRUE) &&
  requireNamespace("AnnotationDbi", quietly = TRUE)
has_brms <- requireNamespace("brms", quietly = TRUE)
can_run <- has_seurat && has_edb && has_brms

if (requireNamespace("devtools", quietly = TRUE)) {
  pkg_root <- if (file.exists("DESCRIPTION") && dir.exists("R")) {
    normalizePath(".")
  } else if (file.exists("../DESCRIPTION") && dir.exists("../R")) {
    normalizePath("..")
  } else {
    NULL
  }
  if (!is.null(pkg_root)) {
    devtools::load_all(pkg_root, quiet = TRUE)
  }
}
```

This vignette follows `examples/savi_adrb2_workflow.R` for the SAVI case study
(*ADRB2* in disease-associated monocytes, PBMC / blood).

At each modelling step we show **two equivalent paths**:

- **Core / manual** — matrix-level helpers you can wire yourself  
- **Wrapper** — same cores, with Seurat / SE / gene-id / metadata handling

Pipeline:

1. Prepare pseudobulk counts (`savi_mono`) and harmonise gene ids  
2. Align to the HCA monocytic reference  
   - core: `merge_with_reference_sample()` + `calculate_tmm_offset()`  
   - wrapper: `scale_to_hca_reference()`  
3. Estimate cohort log(μ)  
   - core: `design_from_formula()` + `estimate_logmu_ql()` (all genes)  
   - wrapper: `estimate_cohort_logmu()` (filter genes)  
   - demo: single-cohort `~ 1` (intercept as cohort log(μ))  
4. Healthy HCA posterior (Normal, blood, 10x Genomics 3)  
   - manual: `load_expr_fit()` → `build_newdata_grid()` → `expr_draws()`  
   - wrapper: `expr_predict()`  
5. Welch test vs HCA  
   - wrapper: `welch_t_test_cohort_hca()`  
   - manual: `cohort_estimate_at()` + `summarize_posterior_draws()` + `welch_test_means()`  
6. Plot results  

## Setup

```{r libraries, eval = can_run}
library(posteriorHCA)
library(Seurat)
library(dplyr)
library(ggplot2)

cell_type <- "monocytic"
```

## 1. Prepare `savi_mono`

The SAVI study ([de Cevins *et al.*, 2023](https://doi.org/10.1016/j.xcrm.2023.101333))
is PBMC scRNA-seq pseudobulked by sample and cell type. We subset
**disease-associated monocytes** (17 samples: CTRL, SAVI, SAVI_treated).

```{r savi-mono-code, eval = FALSE}
savi_path <- "GSE226598_SAVI_pseudobulk_Sample_CellType.rds"
savi <- readRDS(savi_path)
savi_mono <- subset(savi, subset = CellType == "17. Disease-associated monocytes")
```

```{r load-savi-mono, eval = can_run}
data(savi_mono)
dim(savi_mono)
table(savi_mono$Category)
```

Harmonise gene symbols to Ensembl IDs so cohort counts and HCA models share
row names:

```{r harmonise, eval = can_run}
savi_mono <- harmonise_gene_ids(savi_mono, id_type = "auto")
```

## 2. Align to the HCA monocytic reference - step 0

Reference sits at offset 0 (same convention as atlas brms training:
`offset = log(1 / multiplier)`).

### 2a. Core functions (matrix)

```{r align-core, eval = can_run}
ref <- load_reference_sample(cell_type)
user_mat <- as.matrix(Seurat::GetAssayData(savi_mono, layer = "counts"))

combined <- merge_with_reference_sample(
  user_mat,
  reference = ref$counts,
  reference_name = ref$sample_id
)
scaling <- calculate_tmm_offset(
  combined,
  reference_name = ref$sample_id,
  method = "TMMwsp"
)

c(
  n_genes = nrow(combined),
  n_samples = ncol(combined),
  reference_offset = unname(scaling$offset[[ref$sample_id]])
)
```

### 2b. Wrapper `scale_to_hca_reference()`

```{r align-wrapper, eval = can_run}
aligned <- scale_to_hca_reference(savi_mono, cell_type)

table(aligned$sample_role)
aligned$hca_reference_name[1]
```

```{r align-check, eval = can_run}
all.equal(
  unname(scaling$offset[colnames(user_mat)]),
  unname(aligned$hca_offset[colnames(user_mat)]),
  tolerance = 1e-10
)
```

## 3. Estimate cohort log(μ) - step 1

Absolute log(μ) comes from QL coefficients with `prior.count = 0` on the
TMM offset scale. Prefer a cell-means design such as `~ 0 + Category`.

### 3a. Core `design_from_formula()` + `estimate_logmu_ql()`

`estimate_logmu_ql()` fits **all genes** once and returns log(μ) + SE for
every gene × design column. Filter afterwards if you only need a subset.

```{r estimate-core, eval = can_run}
meta_core <- data.frame(
  Category = c(
    as.character(savi_mono$Category[colnames(user_mat)]),
    "reference"
  ),
  row.names = colnames(combined),
  stringsAsFactors = FALSE
)
meta_core$Category <- factor(meta_core$Category)

design <- design_from_formula(~ 0 + Category, meta_core)
est_all <- estimate_logmu_ql(
  counts = combined,
  offset = scaling$offset,
  design = design,
  cell_type = cell_type
)
gene_ensg <- resolve_gene_one("ADRB2", cell_type = cell_type)
est_core <- est_all[est_all$gene == gene_ensg, , drop = FALSE]
est_core$gene_symbol <- "ADRB2"
est_core
```

### 3b. Wrapper `estimate_cohort_logmu()`

Same cores, plus container I/O and gene-id resolution. `genes` only filters
the returned table (dispersion still uses the full matrix).

```{r estimate-wrapper, eval = can_run}
est <- estimate_cohort_logmu(
  aligned,
  formula = ~ 0 + Category,
  genes = "ADRB2"
)
est
```

```{r estimate-check, eval = can_run}
all.equal(
  est_core$log_mu[match(est$group, est_core$group)],
  est$log_mu,
  tolerance = 1e-8
)
```

### 3c. Demo: single-cohort `~ 1` (intercept = cohort log(μ))

Alternative to `~ 0 + Category` on the full object. Subset to one `Category`,
fit NB/QL with `formula = ~ 1`, and treat the intercept as that cohort's
absolute log(μ). Useful when you want per-cohort fits without a multi-level
design matrix.

```{r estimate-by-level, eval = can_run}
est_by_level <- purrr::map_dfr(
  levels(aligned$Category),
  .f = function(x) {
    estimate_cohort_logmu(
      subset(aligned, Category == x),
      formula = ~ 1,
      genes = "ADRB2"
    ) %>%
      dplyr::mutate(group = x)
  }
)
est_by_level
```

## 4. Healthy HCA baseline draws

Covariates fixed to healthy blood 10x Genomics 3 (matches SAVI tissue).
`quantity = "linpred"` is log(μ) on the natural-log scale.

### 4a. Manual coding

```{r hca-draws, eval = can_run}
fit <- load_expr_fit(cell_type = cell_type, gene = "ADRB2") # step 2

grid <- build_newdata_grid(
  fit,
  disease_groups = "Normal",
  tissue_groups = "blood",
  assay_groups = "10x Genomics 3"
)

hca_res <- expr_draws(  # step 3
  fit,
  newdata = grid,
  quantity = "linpred",
  collapse = "mean"
)

round(c(mean = mean(hca_res$draws), sd = sd(hca_res$draws)), 3)
```

### 4b. Wrapper `expr_predict()`

```{r hca-predict, eval = can_run}
hca_pred <- expr_predict(
  fit = fit,
  disease_groups = "Normal",
  tissue_groups = "blood",
  assay_groups = "10x Genomics 3",
  quantity = "linpred",
  collapse = "mean"
)
```

## 5. Test cohorts against the healthy baseline

### 5a. Wrapper `welch_t_test_cohort_hca()`

```{r welch-wrapper, eval = can_run}
test_results <- welch_t_test_cohort_hca(
  cohort_est = est,
  hca_draws = hca_pred
)
test_results
```

### 5b. Manual helpers (one cohort)

```{r welch-manual, eval = can_run}
cohort <- cohort_estimate_at(est, group = "SAVI")
baseline <- summarize_posterior_draws(hca_res, value = cohort$mu) # step 4

welch_test_means( # step 5
  cohort$mu, cohort$se,
  baseline$mean, baseline$sd,
  n1 = cohort$n, n2 = baseline$n
)
```

## 6. Visualise results

```{r plot-hca, eval = can_run, fig.asp = 0.7}
plot_hca_draws(
  draws = hca_res,
  subtitle = "Normal, 10x Genomics 3 healthy baseline"
)
```

```{r plot-cohort, eval = can_run, fig.asp = 0.7}
plot_cohort_vs_hca(
  hca_draws = hca_res,
  test_results = test_results,
  subtitle = "QL cohort estimates",
  annotate = c("group", "p_value", "direction")
)
```

```{r plot-hca-pred, eval = can_run, fig.asp = 0.7}
plot_hca_draws(
  draws = hca_pred,
  title = "ADRB2 predicted count posterior (healthy HCA)"
)
```

## Session info

```{r session-info}
sessionInfo()
```
