---
title: "Cohort expression workflow (native edgeR)"
author: "Chen Zhan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cohort expression workflow (native edgeR)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  for_github_md: false
knit: >
    (function(x, ...){
        proj_root <- rprojroot::find_package_root_file() |> normalizePath()
        rmarkdown::render(
            x,
            output_file = "cohort-expression-edger-native.md",
            output_format = "github_document",
            output_dir = file.path(proj_root, "vignettes"),
            knit_root_dir = proj_root,
            params = list(for_github_md = TRUE)
        )
        rmarkdown::render(
            x,
            output_format = "rmarkdown::html_vignette",
            params = list(for_github_md = FALSE)
        )
    })
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)

if (isTRUE(params$for_github_md)) {
  root_dir <- knitr::opts_knit$get("root.dir")
  if (!is.null(root_dir)) {
    knitr::opts_knit$set(output.dir = file.path(root_dir, "vignettes"))
  }
  knitr::opts_chunk$set(fig.path = "cohort-expression-edger-native_files/figure-gfm/")
}

has_seurat <- requireNamespace("Seurat", quietly = TRUE)
has_edb <- requireNamespace("org.Hs.eg.db", quietly = TRUE) &&
  requireNamespace("AnnotationDbi", quietly = TRUE)
has_brms <- requireNamespace("brms", quietly = TRUE)
has_edger <- requireNamespace("edgeR", quietly = TRUE)
has_limma <- requireNamespace("limma", quietly = TRUE)
can_run <- has_seurat && has_edb && has_brms && has_edger && has_limma

if (requireNamespace("devtools", quietly = TRUE)) {
  pkg_root <- if (file.exists("DESCRIPTION") && dir.exists("R")) {
    normalizePath(".")
  } else if (file.exists("../DESCRIPTION") && dir.exists("../R")) {
    normalizePath("..")
  } else {
    NULL
  }
  if (!is.null(pkg_root)) {
    devtools::load_all(pkg_root, quiet = TRUE)
  }
}
```

SAVI case study: *ADRB2* (`ENSG00000169252`) in disease-associated monocytes.

This report mirrors
`vignette("cohort-expression-core", package = "posteriorHCA")`, but
**exposes the edgeR/limma steps inline** instead of calling posteriorHCA
helpers for scaling, offsets, fitting, contrast estimation, or SE
extraction.

Allowed posteriorHCA helpers **before** the HCA expression query:

```text
load_reference_sample()
merge_with_reference_sample()
```

Everything else on the USER side is written with ordinary `edgeR`,
`limma`, and base R. The purpose is diagnostic:

> If we remove posteriorHCA's edgeR helper abstractions, what is the
> simplest transparent sequence of native edgeR/limma calls that
> reproduces our scaling, USER model fitting, contrast estimate, and the
> proposed `glmQLFTest()`-derived uncertainty?

Workflow:

1. Load / merge one HCA reference library with USER counts
2. Joint TMMwsp scaling (explicit `calcNormFactors`)
3. USER-only regression with `offset = log(E_user)`
4. `model.matrix()` → `glmQLFit()` → `makeContrasts()` → `glmQLFTest()`
5. Natural-log estimate + QLF-implied SE; shift absolute estimand by
   `log(E_HCA)`
6. HCA posterior query and `welch_test_means()` (same as the core report)

## Setup

```{r libraries, eval = can_run}
library(posteriorHCA)
library(Seurat)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(edgeR)
library(limma)

cell_type <- "monocytic"
gene_ensg <- "ENSG00000169252"
```

## Prepare counts with Ensembl gene ids

Same USER dataset and metadata as the core-function report.

```{r load-savi-mono, eval = can_run}
data(savi_mono)

user_counts <- as.matrix(Seurat::GetAssayData(savi_mono, layer = "counts"))
mapped <- AnnotationDbi::mapIds(
  org.Hs.eg.db,
  keys = rownames(user_counts),
  column = "ENSEMBL",
  keytype = "SYMBOL",
  multiVals = "first"
)
keep <- !is.na(mapped) & !duplicated(mapped)
user_counts <- user_counts[keep, , drop = FALSE]
rownames(user_counts) <- unname(mapped[keep])

sample_metadata <- data.frame(
  Category = factor(savi_mono$Category[colnames(user_counts)]),
  Experiment = factor(savi_mono$Experiment[colnames(user_counts)]),
  row.names = colnames(user_counts),
  stringsAsFactors = FALSE
)

dim(user_counts)
table(sample_metadata$Category, sample_metadata$Experiment)
```

## Reference preparation (posteriorHCA)

```{r reference, eval = can_run}
reference <- load_reference_sample(cell_type)
reference$sample_id
length(reference$counts)

counts_joint <- merge_with_reference_sample(
  user_counts,
  reference = reference$counts,
  reference_name = reference$sample_id
)

reference_name <- reference$sample_id
dim(counts_joint)
setdiff(colnames(counts_joint), colnames(user_counts))
```

The HCA reference is added only so joint TMM can put USER and HCA libraries
on a common scale. It must not enter the USER regression.

## Explicit joint TMMwsp scaling

Same default method as `calculate_tmm_scaling()` (`method = "TMMwsp"`).

```{r scale-native, eval = can_run}
ref_col <- match(reference_name, colnames(counts_joint))

norm_factors <- edgeR::calcNormFactors(
  counts_joint,
  refColumn = ref_col,
  method = "TMMwsp"
)

library_size <- colSums(counts_joint)
effective_size <- library_size * norm_factors

E_hca <- unname(effective_size[[reference_name]])
user_names <- setdiff(colnames(counts_joint), reference_name)
E_user <- effective_size[user_names]
log_E_user <- log(E_user)
log_E_hca <- log(E_hca)

data.frame(
  sample = c(user_names, reference_name),
  library_size = c(library_size[user_names], library_size[[reference_name]]),
  norm_factor = c(norm_factors[user_names], norm_factors[[reference_name]]),
  effective_size = c(E_user, E_hca),
  role = c(rep("user", length(user_names)), "hca_reference"),
  row.names = NULL
)
c(E_hca = E_hca, log_E_hca = log_E_hca)
```

## USER regression excludes the HCA reference

```{r user-only, eval = can_run}
counts_fit <- counts_joint[, user_names, drop = FALSE]
metadata_fit <- sample_metadata[user_names, , drop = FALSE]
stopifnot(identical(colnames(counts_fit), rownames(metadata_fit)))

offset_mat <- matrix(
  log_E_user[colnames(counts_fit)],
  nrow = nrow(counts_fit),
  ncol = ncol(counts_fit),
  byrow = TRUE
)
dim(offset_mat)
range(offset_mat[1, ] - log_E_user[colnames(counts_fit)])
```

Offset is `log(E_user)`. No `scaleOffset()`, and USER offsets are not
centred around `E_hca`.

## Design matrix

Primary design matches the core report's SAVI group-mean estimand used in
Welch comparison: one mean per `Category`.

```{r design, eval = can_run}
design <- model.matrix(~ 0 + Category, data = metadata_fit)
colnames(design)
head(design)
```

## Fit with `glmQLFit()` (edgeR v4 QL)

The current non-legacy edgeR v4 QL workflow estimates the NB dispersion
internally in `glmQLFit()`, while gene-specific variability is represented
through moderated quasi-dispersions (`s2.post`). No separate
`estimateDisp()` call. `prior.count = 0` matches the package helper on
Bioconductor release edgeR.

```{r glmqlfit, eval = can_run}
ql_args <- list(
  y = counts_fit,
  design = design,
  offset = offset_mat,
  robust = TRUE
)
if (utils::packageVersion("edgeR") < "4.99.0") {
  ql_args$prior.count <- 0
}
fit <- do.call(edgeR::glmQLFit, ql_args)

c(
  n_genes = nrow(fit),
  n_coef = ncol(fit$coefficients),
  nb_dispersion = unname(as.numeric(fit$dispersion)[[1]]),
  average_ql_dispersion = unname(as.numeric(fit$average.ql.dispersion)[[1]])
)
```

## Contrast: absolute SAVI group mean

```{r contrast, eval = can_run}
contrast <- limma::makeContrasts(
  CategorySAVI,
  levels = design
)
stopifnot(ncol(contrast) == 1L)
contrast
```

Interpretation: under `~ 0 + Category`, `CategorySAVI` is the absolute
normalized log mean for the SAVI group (not a difference, and not a
marginal mean over Experiment).

## `glmQLFTest()` → estimate + QLF-implied SE

```{r glmqlftest, eval = can_run}
qlf <- edgeR::glmQLFTest(fit, contrast = contrast)

# logFC is log2; convert to natural log used by posteriorHCA.
estimate_ln <- qlf$table$logFC * log(2)

# QLF-implied effective SE from the one-dimensional moderated QL F statistic.
# Not an exact coefficient-covariance SE; compatibility with Bayesian
# posterior SD is being evaluated separately.
se_ln <- abs(qlf$table$logFC) / sqrt(qlf$table$F) * log(2)

user_result <- data.frame(
  gene = rownames(qlf$table),
  estimate = estimate_ln,
  se = se_ln,
  F = qlf$table$F,
  PValue = qlf$table$PValue,
  stringsAsFactors = FALSE
)

user_gene <- user_result[user_result$gene == gene_ensg, , drop = FALSE]
user_gene
```

## Shift absolute estimand to the HCA scale

Valid here because `CategorySAVI` is an **absolute** estimand. Do not add
`log(E_hca)` to a relative difference contrast.

```{r hca-scale, eval = can_run}
user_gene$log_mu <- user_gene$estimate + log_E_hca
user_gene$mu <- exp(user_gene$log_mu)
user_gene$se_log_mu <- user_gene$se

n_savi <- sum(metadata_fit$Category == "SAVI")
user_log_mu <- user_gene$log_mu
user_se <- user_gene$se_log_mu

c(log_mu = user_log_mu, se = user_se, n = n_savi)
user_gene[, c("gene", "estimate", "log_mu", "mu", "se", "se_log_mu")]
```


## Diagnostic comparison with the core-function helpers

The block below calls posteriorHCA helpers **only** for side-by-side
diagnostics. The native path above does not depend on them.

```{r compare-core, eval = can_run}
scaling_pkg <- calculate_tmm_scaling(
  counts_joint,
  reference_name = reference_name
)

core_est <- estimate_ql(
  counts = counts_fit,
  offset = log_E_user[colnames(counts_fit)],
  metadata = metadata_fit,
  formula = ~ 0 + Category,
  contrast = "CategorySAVI"
)
core_est <- core_est[core_est$gene == gene_ensg, , drop = FALSE]

scaling_check <- data.frame(
  quantity = c("E_hca", "mean_E_user", "log_E_hca"),
  native = c(E_hca, mean(E_user), log_E_hca),
  package = c(
    scaling_pkg$hca_effective_library_size,
    mean(scaling_pkg$effective_size[user_names]),
    scaling_pkg$hca_log_effective_library_size
  )
)
scaling_check$diff <- scaling_check$native - scaling_check$package
scaling_check

comparison <- data.frame(
  gene = gene_ensg,
  estimate_native_qlf = user_gene$estimate,
  estimate_core_wald = core_est$estimate,
  estimate_diff = user_gene$estimate - core_est$estimate,
  log_mu_native = user_gene$log_mu,
  log_mu_core = core_est$estimate + scaling_pkg$hca_log_effective_library_size,
  SE_QLF = user_gene$se,
  SE_Wald_core = core_est$se,
  SE_QLF_over_SE_Wald = user_gene$se / core_est$se,
  stringsAsFactors = FALSE
)
comparison
```

Scaling and the natural-log point estimate should agree (same counts,
reference, TMMwsp, design, contrast, and `prior.count = 0`). The SE ratio
is diagnostic: QLF-implied SE versus the package Wald / `c'Vc` SE need not
be equal.

## HCA posterior

Same query as the core-function report (healthy blood, 10x Genomics 3).
The USER estimand is the SAVI Category mean under `~ 0 + Category` (no
Experiment conditioning). The HCA query is a healthy baseline for the same
cell type / gene; it is not conditioned on USER Experiment levels.

```{r hca, eval = can_run}
expression_fit <- load_expression_fit(
  cell_type = cell_type,
  gene_ensg = gene_ensg
)

newdata <- build_newdata_grid(
  expression_fit,
  disease_groups = "Normal",
  tissue_groups = "blood",
  assay_groups = "10x Genomics 3"
)

posterior_draws <- expression_draws(
  expression_fit,
  newdata = newdata,
  quantity = "linpred",
  collapse = "mean"
)
```

## Comparison

Welch tests of the same SAVI `log_mu` against the HCA posterior, using
either the QLF-implied SE (native path) or the Wald / `c'Vc` SE (core
helper). Point estimates match; only the SE changes.

```{r welch, eval = can_run}
posterior_summary <- summarize_posterior_draws(
  posterior_draws,
  value = user_log_mu
)

user_log_mu_core <- core_est$estimate + log_E_hca
user_se_core <- core_est$se

test_qlf <- welch_test_means(
  mu1 = user_log_mu,
  se1 = user_se,
  mu2 = posterior_summary$log_mu,
  se2 = posterior_summary$se,
  n1 = n_savi,
  n2 = posterior_summary$n
)
test_wald <- welch_test_means(
  mu1 = user_log_mu_core,
  se1 = user_se_core,
  mu2 = posterior_summary$log_mu,
  se2 = posterior_summary$se,
  n1 = n_savi,
  n2 = posterior_summary$n
)

test_results <- data.frame(
  group = c("SAVI (QLF SE)", "SAVI (Wald SE)"),
  n = n_savi,
  log_mu = c(test_qlf$mu1, test_wald$mu1),
  se = c(test_qlf$se1, test_wald$se1),
  hca_log_mu = c(test_qlf$mu2, test_wald$mu2),
  hca_se = c(test_qlf$se2, test_wald$se2),
  p_value = c(test_qlf$p_value, test_wald$p_value),
  stringsAsFactors = FALSE
)
test_results
```

## Plots

```{r plot-hca, eval = can_run, fig.asp = 0.7}
plot_hca_draws(
  draws = posterior_draws,
  subtitle = "Normal, 10x Genomics 3 healthy baseline"
)
```

```{r plot-cohort, eval = can_run, fig.asp = 0.7}
plot_cohort_vs_hca(
  hca_draws = posterior_draws,
  cohort_est = test_results,
  subtitle = "SAVI mean: QLF-implied SE vs Wald / c'Vc SE",
  annotate = c("group", "p_value")
)
```

## Session info

```{r session-info}
sessionInfo()
```
