---
title: "Cohort expression workflow"
author: "Chen Zhan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cohort expression workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  for_github_md: false
knit: >
    (function(x, ...){
        proj_root <- rprojroot::find_package_root_file() |> normalizePath()
        rmarkdown::render(
            x,
            output_format = "rmarkdown::html_vignette",
            params = list(for_github_md = FALSE)
        )
        rmarkdown::render(
            x,
            output_file = "cohort-expression-workflow.md",
            output_format = "github_document",
            output_dir = file.path(proj_root, "vignettes"),
            knit_root_dir = proj_root,
            params = list(for_github_md = TRUE)
        )
    })
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)

# Relative figure paths for github_document output (see rmarkdown#2473).
if (isTRUE(params$for_github_md)) {
  root_dir <- knitr::opts_knit$get("root.dir")
  if (!is.null(root_dir)) {
    knitr::opts_knit$set(
      output.dir = file.path(root_dir, "vignettes")
    )
  }
  knitr::opts_chunk$set(
    fig.path = "cohort-expression-workflow_files/figure-gfm/"
  )
}

has_seurat <- requireNamespace("Seurat", quietly = TRUE)
has_edb <- requireNamespace("org.Hs.eg.db", quietly = TRUE) &&
  requireNamespace("AnnotationDbi", quietly = TRUE)
has_brms <- requireNamespace("brms", quietly = TRUE)
can_run <- has_seurat && has_edb && has_brms

if (requireNamespace("devtools", quietly = TRUE)) {
  pkg_root <- if (file.exists("DESCRIPTION") && dir.exists("R")) {
    normalizePath(".")
  } else if (file.exists("../DESCRIPTION") && dir.exists("../R")) {
    normalizePath("..")
  } else {
    NULL
  }
  if (!is.null(pkg_root)) {
    devtools::load_all(pkg_root, quiet = TRUE)
  }
}
```

This vignette follows `examples/savi_adrb2_workflow.R` for the SAVI case study
(*ADRB2* in disease-associated monocytes, PBMC / blood):

1. Prepare pseudobulk counts (`savi_mono`)
2. Align to the HCA monocytic reference
3. Estimate cohort log(μ) with edgeR QL
4. Draw healthy HCA posterior (Normal, blood, 10x Genomics 3)
5. Welch test vs HCA (wrapper + manual helpers)
6. Plot results

## Setup

```{r libraries, eval = can_run}
library(posteriorHCA)
library(Seurat)
library(tidyseurat)
library(dplyr)
library(purrr)
library(ggplot2)

cell_type <- "monocytic"
```

## 1. Prepare `savi_mono`

The SAVI study ([de Cevins *et al.*, 2023](https://doi.org/10.1016/j.xcrm.2023.101333))
is PBMC scRNA-seq pseudobulked by sample and cell type. We subset
**disease-associated monocytes** (17 samples: CTRL, SAVI, SAVI_treated).

```{r savi-mono-code, eval = FALSE}
savi_path <- "GSE226598_SAVI_pseudobulk_Sample_CellType.rds"
savi <- readRDS(savi_path)
savi_mono <- subset(savi, subset = CellType == "17. Disease-associated monocytes")
```

```{r load-savi-mono, eval = can_run}
data(savi_mono)
dim(savi_mono)
table(savi_mono$Category)
```

Harmonise gene symbols to Ensembl IDs so cohort counts and HCA models share
row names:

```{r harmonise, eval = can_run}
savi_mono <- harmonise_gene_ids(savi_mono, id_type = "auto")
```

## 2. Align to the HCA monocytic reference

`scale_to_hca_reference()` appends one HCA library, TMM-aligns all samples,
and stores `hca_offset` and `sample_role` on each column.

```{r align, eval = can_run}
aligned <- scale_to_hca_reference(savi_mono, cell_type)

table(aligned$sample_role)
```

## 3. Estimate cohort log(μ)

Primary model: group means with `~ 0 + Category` (includes atlas reference).

```{r estimate, eval = can_run}
est <- estimate_cohort_logmu(
  aligned,
  formula = ~ 0 + Category,
  genes = "ADRB2"
)
est
```

Alternative: estimate each `Category` level separately with `~ 1` (as in the
reference script):

```{r estimate-by-level, eval = can_run}
est_by_level <- purrr::map_dfr(
  aligned$Category %>% levels(),
  .f = function(x) {
    estimate_cohort_logmu(
      aligned %>% dplyr::filter(Category == x),
      formula = ~ 1,
      genes = "ADRB2"
    ) %>%
      dplyr::mutate(group = x)
  }
)
est_by_level
```

## 4. Healthy HCA baseline draws

Load the cached brms fit and draw from Normal / blood / 10x Genomics 3.

```{r hca-draws, eval = can_run}
fit <- load_expr_fit(cell_type = cell_type, gene = "ADRB2")

grid <- build_newdata_grid(
  fit,
  disease_groups = "Normal",
  tissue_groups = "blood",
  assay_groups = "10x Genomics 3"
)

hca_res <- expr_draws(
  fit,
  newdata = grid,
  quantity = "linpred",
  collapse = "mean"
)

round(c(mean = mean(hca_res$draws), sd = sd(hca_res$draws)), 3)
```

Same fit via `expr_predict()` (convenience wrapper used in the Welch wrapper
below):

```{r hca-predict, eval = can_run}
hca_pred <- expr_predict(
  fit = fit,
  disease_groups = "Normal",
  tissue_groups = "blood",
  assay_groups = "10x Genomics 3",
  quantity = "linpred",
  collapse = "mean"
)
```

## 5. Test cohorts against the healthy baseline

### Wrapper

```{r welch, eval = can_run}
test_results <- welch_t_test_cohort_hca(
  cohort_est = est,
  hca_draws = hca_pred
)
test_results
```

### Manual workflow (one cohort)

```{r welch-manual, eval = can_run}
cohort <- cohort_estimate_at(est, group = "SAVI")
baseline <- summarize_posterior_draws(hca_res, value = cohort$mu)

welch_test_means(
  cohort$mu, cohort$se,
  baseline$mean, baseline$sd,
  n1 = cohort$n, n2 = baseline$n
)
```

## 6. Visualise results

```{r plot-hca, eval = can_run, fig.asp = 0.7}
plot_hca_draws(
  draws = hca_res,
  subtitle = "Normal, 10x Genomics 3 healthy baseline"
)
```

```{r plot-cohort, eval = can_run, fig.asp = 0.7}
plot_cohort_vs_hca(
  hca_draws = hca_res,
  test_results = test_results,
  subtitle = "QL cohort estimates",
  annotate = c("group", "p_value", "direction")
)
```

```{r plot-hca-pred, eval = can_run, fig.asp = 0.7}
plot_hca_draws(
  draws = hca_pred,
  title = "ADRB2 predicted count posterior (healthy HCA)"
)
```

## Session info

```{r session-info}
sessionInfo()
```
