---
title: "Composition workflow"
author: "Chen Zhan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Composition workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)

can_run <- requireNamespace("sccomp", quietly = TRUE) &&
  requireNamespace("ggplot2", quietly = TRUE)

if (requireNamespace("devtools", quietly = TRUE)) {
  pkg_root <- if (file.exists("DESCRIPTION") && dir.exists("R")) {
    normalizePath(".")
  } else if (file.exists("../DESCRIPTION") && dir.exists("../R")) {
    normalizePath("..")
  } else {
    NULL
  }
  if (!is.null(pkg_root)) {
    devtools::load_all(pkg_root, quiet = TRUE)
  }
}
```

This vignette walks through the **composition vs healthy HCA baseline**
workflow, mirroring the modular gene-expression design:

1. Load the healthy sccomp model
2. Draw posterior predictive proportions
3. Test observed proportions (empirical confidence)
4. Plot healthy draws and observed overlays

## Setup

```{r libraries, eval = can_run}
library(posteriorHCA)
library(ggplot2)

data(example_proportions)
example_proportions
```

## Load the healthy sccomp reference

```{r load-fit, eval = can_run}
fit <- load_sccomp_fit()
```

## Draw healthy reference proportions

Unknown covariates can be left as `NA`; sccomp marginalises over them.

```{r draws, eval = can_run}
comp <- composition_draws(
  fit,
  sex = "male",
  age_decade = "4",
  ethnicity_groups = "European",
  assay_groups = "10x Genomics 3",
  tissue_groups = "blood"
)
head(comp$draws)
```

## Test observed proportions

```{r test, eval = can_run}
test_results <- composition_test(
  proportions = example_proportions,
  draws = comp
)
test_results
```

For a single sample:

```{r test-sample, eval = can_run}
composition_test_sample(
  proportions = example_proportions,
  draws = comp,
  sample_id = "ID1"
)
```

## Plot

```{r plot-draws, eval = can_run, fig.asp = 0.7}
plot_composition_draws(
  comp,
  cell_types = unique(example_proportions$cell_type),
  subtitle = "Healthy HCA baseline"
)
```

```{r plot-vs-hca, eval = can_run, fig.asp = 0.7}
plot_composition_vs_hca(
  draws = comp,
  test_results = test_results,
  subtitle = "Observed proportions vs healthy baseline"
)
```

## Convenience wrapper

`composition_posterior_test()` still bundles load → draws → test → plot:

```{r wrapper, eval = can_run, fig.asp = 0.7}
result <- composition_posterior_test(
  proportions = example_proportions,
  sex = "male",
  age_decade = "4",
  ethnicity_groups = "European",
  assay_groups = "10x Genomics 3",
  tissue_groups = "blood",
  fit = fit
)
result$result_table
result$plot
```

## Session info

```{r session-info}
sessionInfo()
```
