---
title: "posteriorHCA"
output: "html_vignette"
params:
  demo_metadata: true
vignette: >
  %\VignetteIndexEntry{posteriorHCA}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
knit: >
    (function(x, ...){
        proj_root <- rprojroot::find_package_root_file() |> normalizePath()
        rmarkdown::render(
            x,
            output_format = "html_document",
            params = list(
                demo_metadata = TRUE
            )
        )
        rmarkdown::render(
            x,
            output_file = "README.md",
            output_format = "github_document",
            output_dir = proj_root,
            knit_root_dir = proj_root,
            params = list(
                demo_metadata = FALSE
            )
        )
    })
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
root_dir <- knitr::opts_knit$get("root.dir")
if (!is.null(root_dir)){
    # This hack fixes the relative image paths. 
    # See https://github.com/rstudio/rmarkdown/issues/2473
    knitr::opts_knit$set(
        output.dir = root_dir
    )
}
proj_root <- rprojroot::find_package_root_file() |> normalizePath() 
# Utility function for figures to force them to have the correct path
find_figure <- function(names){
    rprojroot::find_package_root_file() |>
        file.path("man", "figures", names) 
}
```

<!-- badges: start -->

[![Lifecycle:experimental](https://lifecycle.r-lib.org/articles/figures/lifecycle-experimental.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)

<!-- badges: end -->

`PosteriorHCA` applies **Bayesian inference** to model the comprehensive **Human Cell Atlas (HCA)**, allowing researchers to probabilistically analyze both **cellular composition** and **gene expression** patterns across human samples.

This package provides two main analytical capabilities:

1.  **Cellular Composition Analysis**: Uses the `sccomp` package to model variations in cell type proportions across:
    -   **Cell Types**
    -   **Sex**
    -   **Age**
    -   **Ethnicity**
    -   **Disease conditions**
    -   **Tissue Groups**

2.  **Gene Expression Prediction**: Uses `brms` models to predict gene expression levels for specific cell types and genes based on sample metadata including:
    -   **Cell Types**
    -   **Sex**
    -   **Age Decade**
    -   **Disease Groups**
    -   **Ethnicity Groups**
    -   **Assay Groups**
    -   **Tissue Groups**

The probabilistic landscapes are **pre-trained**, meaning they represent **posterior distributions** estimated from millions of cells. `PosteriorHCA` makes these **queryable**, allowing researchers to:

-   **Test hypotheses** against precomputed reference baselines.
-   **Compare new samples** to expected cellular composition distributions.
-   **Predict gene expression** levels for specific cell types and genes.
-   **Assess how cell proportions and gene expression vary** across human conditions.

## **Installation**

To install the latest development version of `PosteriorHCA` from GitHub, run:

```{r eval=FALSE}
# Install devtools if not already installed
install.packages("devtools")

# Install PosteriorHCA from GitHub
devtools::install_github("MangiolaLaboratory/posteriorHCA")
```

Once installed, load the package:

```{r}
library(posteriorHCA)
```

## Usage Examples

### **1. Cellular Composition Analysis**

#### Load relevant packages and example data

The package includes a small example dataset for demonstration:

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)

data(example_proportions)
print(example_proportions)
```

#### Run Composition Posterior Testing

The `composition_posterior_test()` function allows testing new cellular composition data against the probabilistic model:

```{r eval = instantiate::stan_cmdstan_exists(), fig.width= 12, fig.height= 10 }
result <- composition_posterior_test(
    proportions = example_proportions,
    sex = "male",
    age_decade = '4',
    disease_groups = "Normal",
    ethnicity_groups = "European",
    tissue_groups = "blood", 
    load_model_to_global_env = T
)
```

#### About `load_model_to_global_env`

In the `composition_posterior_test` function, we added arg - `load_model_to_global_env` to determine if the model file should be loaded into global env or local env (scope within the function running) in R session. By default, it is set to be TRUE - `load_model_to_global_env = T`, and this is particularly useful when you wish to repeatedly run the function. Loading model file into global env prevents re-loading the model file, which usually will cost you minutes to run.

#### Inspect Composition Results

**Posterior Distribution Table:**

```{r eval = instantiate::stan_cmdstan_exists()}
print(result$result_table)
```

**Density Plot of Posterior Predictions:**

```{r eval = instantiate::stan_cmdstan_exists(), fig.width= 12, fig.height= 10 }
print(result$plot)
```

### **2. Gene Expression Prediction**

The `expr_predict()` function allows predicting gene expression levels for specific cell types and genes based on sample metadata:

```{r eval = requireNamespace("brms", quietly = TRUE), fig.width= 10, fig.height= 6 }
# Predict gene expression for a specific cell type and gene
expr_result <- expr_predict(
  cell_type = 'cd4 naive',
  gene_ensg = 'ENSG00000000419',
  age_decade = '7',
  sex = 'female',
  disease_groups = 'Normal',
  ethnicity_groups = 'European',
  assay_groups = '10x Genomics 3',
  tissue_groups = 'blood'
)
```

#### Inspect Gene Expression Results

**Summary Statistics:**

```{r eval = requireNamespace("brms", quietly = TRUE)}
print(expr_result$summary)
```

**Density Plot of Predicted Expression:**

```{r eval = requireNamespace("brms", quietly = TRUE), fig.width= 10, fig.height= 6 }
print(expr_result$plot)
```

**Predicted Values:**

```{r eval = requireNamespace("brms", quietly = TRUE)}
head(expr_result$pred)
```

## **Input Arguments and Valid Values**

Both `composition_posterior_test()` and `expr_predict()` functions accept various metadata inputs corresponding to observed sample characteristics. These include **sex, age decade, disease condition, ethnicity, and tissue type, etc**. For composition analysis, users can also provide observed cell type proportions. However, if some metadata or proportions are unknown, users can leave these fields empty, and the functions will still generate meaningful posterior estimates based on the comprehensive reference.

If users choose to specify metadata, it must be consistent with the **Human Cell Atlas (HCA) annotations** to ensure accurate comparisons. Below is a comprehensive list of valid arguments:

-   **Cell Types:** "b memory", "b naive", "cd14 mono", "cd16 mono", "cd4 fh em", "cd4 naive", "cd4 tcm", "cd4 th1 em", "cd4 th1/th17 em", "cd4 th17 em", "cd4 th2 em", "cd8 naive", "cd8 tcm", "cd8 tem", "cdc", "granulocyte", "ilc", "macrophage", "mait", "mast", "nk", "nkt", "pdc", "plasma", "tgd", "treg"

-   **Sex:** "female", "male", "unknown"

-   **Age Decades:** 1, 2, 3, 4, 5, 6, 7, 8, 9, 10

-   **Disease Groups:** "Normal", "COVID-19 related_blood", "Metabolic and Other Disorders: Renal and Urinary Disorders_renal system", "Systemic Lupus Erythematosus_blood", "Respiratory Conditions: General Respiratory Disorders_blood", "COVID-19 related_nasal, oral, and pharyngeal regions", "COVID-19 related_respiratory system", "Glioblastoma_cerebral lobes and cortical areas", "Cancer: Hematologic Cancer_bone marrow", "Lung Adenocarcinoma_respiratory system", "Infectious and Immune-related Diseases: Autoimmune and Immune-Related Disorders_small intestine", "Infectious and Immune-related Diseases: Respiratory Infections_blood", "Metabolic and Other Disorders: Gastrointestinal Disorders_large intestine", "Cancer: Renal Cancer_renal system", "Infectious and Immune-related Diseases: Autoimmune and Immune-Related Disorders_respiratory system", "Infectious and Immune-related Diseases: Respiratory Infections_nasal, oral, and pharyngeal regions", "Metabolic and Other Disorders: Renal and Urinary Disorders_prostate", "Metabolic and Other Disorders_sensory-related structures", "Cancer: Gastrointestinal Cancer_large intestine", "Other Diseases_brainstem and cerebellar structures", "Metabolic and Other Disorders: Gastrointestinal Disorders_oesophagus", "Cancer: Lung Cancer_respiratory system", "Lung Adenocarcinoma_cerebral lobes and cortical areas", "Metabolic and Other Disorders: Gastrointestinal Disorders_stomach", "Lung Adenocarcinoma_liver", "Cancer: Lung Cancer_liver", "Respiratory Conditions: Restrictive Lung Diseases_respiratory system", "Cancer_integumentary system (skin)", "Infectious and Immune-related Diseases: Autoimmune and Immune-Related Disorders_large intestine", "Other Diseases_nasal, oral, and pharyngeal regions", "Cancer: Lung Cancer_lymphatic system", "Neurodegenerative and Neurological Disorders: Neurodegenerative Disorders_cerebral lobes and cortical areas", "Cancer: Breast Cancer_respiratory system", "Metabolic and Other Disorders: Metabolic Disorders_blood", "Cancer: Gastrointestinal Cancer_liver", "Cancer: Renal Cancer_blood", "Glioblastoma_brainstem and cerebellar structures", "Lung Adenocarcinoma_lymphatic system", "Cancer: Lung Cancer_integumentary system (skin)", "Respiratory Conditions: Obstructive Lung Diseases_respiratory system", "Metabolic and Other Disorders_female reproductive system", "Metabolic and Other Disorders: Gastrointestinal Disorders_liver", "Cancer: Lung Cancer_endocrine system", "Lung Adenocarcinoma_endocrine system", "COVID-19 related_trachea", "Neurodegenerative and Neurological Disorders_cerebral lobes and cortical areas", "Other Diseases_renal system", "Cancer: Renal Cancer_endocrine system", "Cancer: Renal Cancer_lymphatic system"

-   **Ethnicity Groups:** "African", "Other/Unknown", "European", "East Asian", "Hispanic/Latin American", "South Asian", "Native American & Pacific Islander"

-   **Tissue Groups:** "respiratory system", "blood", "renal system", "small intestine", "nasal, oral, and pharyngeal regions", "cerebral lobes and cortical areas", "bone marrow", "breast", "cardiovascular system", "trachea", "endocrine system", "liver", "large intestine", "lymphatic system", "prostate", "sensory-related structures", "oesophagus", "spleen", "brainstem and cerebellar structures", "stomach", "female reproductive system", "thymus", "adipose tissue", "integumentary system (skin)", "digestive system (general)", "gallbladder", "pancreas"

# Session Info

```{r}
sessionInfo()
```
