---
title: "PAM50 toy model — Wu breast cancer epithelial"
date: today
date-format: "DD MMM YYYY"
title-block-banner: true
author:
  - name: Chen Zhan
    orcid: 0000-0002-4794-8339
    email: chen.zhan@adelaide.edu.au
    affiliation:
      - id: sgc
        name: SAiGENCI, The University of Adelaide
format:
  html:
    theme: minty
    df-print: paged
    code-line-numbers: true
    embed-resources: true
    code-fold: true
    code-fold-default: true
    toc: true
    toc-location: left
    toc-title: "Menu"
knitr:
  opts_chunk:
    message: false
    warning: false
editor: source
abstract: |
  Fit and compress brms models for PAM50 genes on breast epithelial reference data,
  merge Wu et al. cancer pseudobulk with reference samples, and compare tumour counts
  to healthy posterior predictions.
params:
  run_fits: false
  cmdstan_path: "/hpcfs/users/a1237163/R/.cmdstan/cmdstan-2.36.0"
  se_breast_rds: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/HPC_posterior/Age_Clock/TAR_missing_tissue/se_list_missing.rds"
  se_cancer_rds: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/dev/wu_breast_cancer/data/Wu_etal_2021_BRCA_se_cancer_epithelial.rds"
  estimates_rds: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/dev/wu_breast_cancer/data/pam50_estimates_chunk.rds"
  n_components: 5
  corr_threshold: 0.1
---

# Setup

```{r}
#| label: setup
.libPaths("/hpcfs/users/a1237163/R/x86_64-pc-linux-gnu-library/4.5")

library(tibble)
library(tidyr)
library(dplyr)
library(tidySummarizedExperiment)
library(SummarizedExperiment)
library(S4Vectors)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(tidybulk)
library(glue)
library(purrr)
library(furrr)
library(future)
library(lobstr)
library(posterior)
library(progressr)
library(brms)
library(poco)
library(tidyverse)
library(ggplot2)
library(forcats)
library(pheatmap)
library(RColorBrewer)

if (requireNamespace("cmdstanr", quietly = TRUE)) {
  cmdstanr::set_cmdstan_path(params$cmdstan_path)
}
```

# Breast reference and PAM50 biomarkers

```{r}
#| label: load-breast
se_breast <-
  readRDS(params$se_breast_rds)[["breast"]]

se_breast %>%
  rowData() %>%
  as.data.frame() %>%
  tibble::rownames_to_column(".feature") %>%
  dplyr::select(.feature) %>%
  mutate(
    symb = AnnotationDbi::mapIds(
      org.Hs.eg.db::org.Hs.eg.db,
      keys = .feature,
      column = "SYMBOL",
      keytype = "ENSEMBL",
      multiVals = "first"
    )
  ) -> ab

pam50_modules_tbl <- tibble(
  module = c(
    "luminal",
    "her2",
    "basal",
    "proliferation",
    "other_pam50"
  ),
  gene = list(
    c(
      "ESR1", "PGR", "FOXA1", "BCL2", "BAG1",
      "NAT1", "MLPH", "SLC39A6", "GPR160", "MAPT"
    ),
    c("ERBB2", "GRB7", "FGFR4"),
    c(
      "KRT5", "KRT14", "KRT17", "EGFR",
      "FOXC1", "CDH3", "MIA", "PHGDH"
    ),
    c(
      "ANLN", "BIRC5", "CCNB1", "CCNE1", "CDC20", "CDC6",
      "CDCA1", "CENPF", "CEP55", "EXO1", "KIF2C", "KNTC2",
      "MELK", "MKI67", "MYBL2", "ORC6L", "PTTG1", "RRM2",
      "TYMS", "UBE2C", "UBE2T"
    ),
    c(
      "ACTR3B", "BLVRA", "CXXC5", "MDM2", "MMP11",
      "MYC", "SFRP1", "TMEM45B"
    )
  )
) %>%
  unnest(gene)

# Literature mapping: Wu clinical subtypes (ER+, HER2+, TNBC) to PAM50 intrinsic classes
# and expected deviation vs healthy reference (Parker et al. 2009, J Clin Oncol;
# see also PAM50 / Prosigna module structure, BMC Med Genomics 2015).
subtype_module_expected <- tibble::tribble(
  ~subtype, ~module, ~pam50_intrinsic, ~expected_deviation, ~rationale,
  "ER+", "luminal", "Luminal A/B", "higher_than_healthy",
  "ER+ tumours: luminal lineage genes (ESR1, PGR, FOXA1) enriched vs other classes",
  "ER+", "her2", "Luminal A/B", "lower_than_healthy",
  "HER2 module low in luminal/ER+ disease",
  "ER+", "basal", "Luminal A/B", "lower_than_healthy",
  "Basal module low in luminal/ER+ disease",
  "ER+", "proliferation", "Luminal A/B", "consistent_with_healthy",
  "Proliferation variable in luminal A vs B; not a single direction",
  "HER2+", "her2", "HER2-enriched", "higher_than_healthy",
  "HER2-enriched: ERBB2/GRB7 module high",
  "HER2+", "luminal", "HER2-enriched", "lower_than_healthy",
  "HER2-enriched: typically ER-low / luminal gene low",
  "HER2+", "basal", "HER2-enriched", "lower_than_healthy",
  "Basal module low in HER2-enriched",
  "HER2+", "proliferation", "HER2-enriched", "higher_than_healthy",
  "HER2-enriched often proliferation-high",
  "TNBC", "basal", "Basal-like", "higher_than_healthy",
  "TNBC ~ basal-like: KRT5/14, EGFR, FOXC1 high",
  "TNBC", "luminal", "Basal-like", "lower_than_healthy",
  "Luminal module low in basal-like/TNBC",
  "TNBC", "her2", "Basal-like", "lower_than_healthy",
  "HER2 module low in basal-like/TNBC",
  "TNBC", "proliferation", "Basal-like", "higher_than_healthy",
  "Basal-like tumours often proliferation-high"
)

tibble(
  pam50_genes = nrow(pam50_modules_tbl),
  mapped_in_reference = sum(pam50_modules_tbl$gene %in% ab$symb)
)

subtype_module_expected
```

```{r}
#| label: biomarkers
ab %>%
  filter(symb %in% pam50_modules_tbl$gene) %>%
  pull(.feature) -> biomarkers

length(biomarkers)
```

# Fit and compress PAM50 models (reference epithelial)

Set `params$run_fits: true` in the YAML header (or when rendering) to re-run the heavy loop. By default the document loads a saved `pam50_estimates_chunk.rds`.

```{r}
#| label: fit-pam50
#| eval: !expr params$run_fits


handlers(global = TRUE)
handlers("progress")

options(future.globals.maxSize = 5 * 1024^3)

n_components <- params$n_components
corr_threshold <- params$corr_threshold

fit_one_biomarker <- function(ensg) {
  estimates_chunk <-
    tibble(
      .feature = ensg,
      symb = ab$symb[ab$.feature == ensg],
      cell_type = "epithelial"
    ) %>%
    mutate(
      se = map(
        ensg,
        ~ se_breast %>%
          filter(cell_type_unified_ensemble == "epithelial") %>%
          .[.x, , drop = FALSE]
      ),
      brms_fit = map(se, ~ {
        data <-
          .x %>%
          as_tibble() %>%
          mutate(counts = as.integer(counts)) %>%
          droplevels()

        n_NAs <- sum(is.na(data$counts))
        if (n_NAs > 0) {
          warning(glue("You have {n_NAs} NAs in counts. They have been filtered out"))
          data <- data %>% filter(!is.na(counts)) %>% droplevels()
        }

        colnames(data) <- stringr::str_replace_all(colnames(data), "_+", "_")

        formula_chr_counts <-
          "counts ~ 1 + offset(offset) + age_decade * sex + disease_groups_altered + ethnicity_groups + assay_groups_altered + (1 | dataset_id_altered)"
        formula_chr_shape <-
          "shape ~ 1 + disease_groups_altered + assay_groups_altered + ethnicity_groups "

        if (nrow(distinct(data, disease_groups_altered)) == 1) {
          formula_chr_counts <- str_remove_all(formula_chr_counts, fixed("+ disease_groups_altered"))
          formula_chr_shape <- str_remove_all(formula_chr_shape, fixed("+ disease_groups_altered"))
        }
        if (nrow(distinct(data, ethnicity_groups)) == 1) {
          formula_chr_counts <- str_remove_all(formula_chr_counts, fixed("+ ethnicity_groups"))
          formula_chr_shape <- str_remove_all(formula_chr_shape, fixed("+ ethnicity_groups"))
        }
        if (nrow(distinct(data, assay_groups_altered)) == 1) {
          formula_chr_counts <- str_remove_all(formula_chr_counts, fixed("+ assay_groups_altered"))
          formula_chr_shape <- str_remove_all(formula_chr_shape, fixed("+ assay_groups_altered"))
        }
        if (nrow(distinct(data, dataset_id_altered)) == 1) {
          formula_chr_counts <- str_remove_all(formula_chr_counts, fixed("+ (1 | dataset_id_altered)"))
        }
        if (nrow(distinct(data, sex)) == 1) {
          formula_chr_counts <- str_remove_all(formula_chr_counts, fixed("* sex"))
        }

        formula <- bf(
          as.formula(formula_chr_counts),
          as.formula(formula_chr_shape)
        )

        if (stringr::str_trim(formula_chr_shape) != "shape ~ 1") {
          prior <- c(
            prior(student_t(3, i, 1.5), class = Intercept),
            prior(student_t(3, 0, 1), class = Intercept, dpar = shape),
            prior(student_t(3, 0, 5), class = b),
            prior(student_t(3, 0, 2), class = b, dpar = shape)
          ) %>%
            substitute(env = list(i = mean(log1p(data$counts / exp(data$offset))))) %>%
            eval()
        } else {
          prior <- c(
            prior(student_t(3, i, 1.5), class = Intercept),
            prior(student_t(3, 0, 1), class = Intercept, dpar = shape),
            prior(student_t(3, 0, 5), class = b)
          ) %>%
            substitute(env = list(i = mean(log1p(data$counts / exp(data$offset))))) %>%
            eval()
        }

        chains <- 2L
        bterms <- brmsterms(
          formula = brms:::validate_formula(
            formula,
            data = data,
            family = zero_inflated_negbinomial(),
            autocor = NULL,
            sparse = NULL,
            cov_ranef = NULL
          )
        )
        bframe <- brms:::brmsframe(bterms, data)
        sdata <- brms:::.standata(
          bframe,
          data = data,
          prior = prior,
          data2 = NULL,
          stanvars = NULL,
          threads = NULL
        )

        inits <- lapply(seq_len(chains), function(i) {
          list(
            b = rnorm(sdata$Kc, 0, 5),
            Intercept = rnorm(1, mean(log1p(data$counts / exp(data$offset))), 1.5),
            b_shape = rnorm(sdata$Kc_shape, 0, 2),
            Intercept_shape = rnorm(1, 0, 1)
          )
        })

        res <- brm(
          formula = formula,
          data = data,
          family = zero_inflated_negbinomial(),
          prior = prior,
          chains = chains,
          cores = 2,
          warmup = 400,
          refresh = 10,
          backend = "cmdstanr",
          init = inits,
          iter = 600
        ) %>%
          poco::strip_brmsfit_envs()

        res
      })
    )

  fit_brms <- estimates_chunk$brms_fit[[1]]

  posterior_cor <- poco::posterior_correlation(
    fit_brms,
    type = "correlation",
    method = "pearson"
  )
  original_draws <- posterior::as_draws_matrix(fit_brms)

  partition <- poco::partition_parameters_clusters(
    posterior_cor,
    threshold = corr_threshold,
    min_size = 2L,
    simple_output = FALSE
  )

  layout_tbl <- tibble(
    n_components = n_components,
    corr_threshold = corr_threshold,
    n_blocks = length(partition$blocks),
    block_size_max = if (length(partition$blocks)) max(lengths(partition$blocks)) else 0L,
    block_size_med = if (length(partition$blocks)) as.integer(stats::median(lengths(partition$blocks))) else 0L,
    n_remainder = length(partition$remainder)
  )

  compress_tic <- Sys.time()
  compressed_blockwise <- tryCatch(
    poco::compress_brmsfit(
      brmsfit = fit_brms,
      method = "mclust",
      n_components = n_components,
      partition = partition,
      cluster_BPPARAM = 1L,
      verbose = FALSE
    ),
    error = function(e) {
      warning(
        "compress_brmsfit failed: ",
        conditionMessage(e),
        call. = FALSE,
        immediate. = TRUE
      )
      NULL
    }
  )
  time_used <- as.numeric(difftime(Sys.time(), compress_tic, units = "secs"))

  fid_blockwise <- if (is.null(compressed_blockwise)) {
    NULL
  } else {
    tryCatch(
      poco::evaluate_compression(
        compressed_blockwise$compressed,
        original_draws,
        seed = 1L
      ),
      error = function(e) {
        warning("evaluate_compression failed: ", conditionMessage(e), call. = FALSE, immediate. = TRUE)
        NULL
      }
    )
  }

  if (is.null(compressed_blockwise)) {
    layout_tbl <- layout_tbl %>%
      mutate(
        compression_ok = FALSE,
        time_used = NA_real_,
        compression_rate = NA_real_,
        energy_pct = NA_real_,
        c2st_pct = NA_real_,
        reproduction_pct = NA_real_
      )
  } else {
    estimates_chunk$brms_fit_compressed <- list(compressed_blockwise)
    layout_tbl <- layout_tbl %>%
      mutate(
        compression_ok = TRUE,
        time_used = time_used,
        compression_rate =
          as.numeric(lobstr::obj_size(compressed_blockwise)) /
          as.numeric(lobstr::obj_size(fit_brms))
      )
  }

  if (is.null(fid_blockwise)) {
    layout_tbl <- layout_tbl %>%
      mutate(
        energy_pct = NA_real_,
        c2st_pct = NA_real_,
        reproduction_pct = NA_real_
      )
  } else {
    en <- fid_blockwise$metrics$energy$reproduction_pct
    c2 <- fid_blockwise$metrics$c2st$reproduction_pct
    layout_tbl <- layout_tbl %>%
      mutate(
        energy_pct = if (is.null(en)) NA_real_ else en,
        c2st_pct = if (is.null(c2)) NA_real_ else c2,
        reproduction_pct = fid_blockwise$reproduction_pct
      )
  }

  estimates_chunk %>% tibble::add_column(layout_tbl)
}

with_progress({
  p <- progressor(along = biomarkers)
  pam50_estimates_chunk <- purrr::map_dfr(biomarkers, function(ensg) {
    p(sprintf("Processing %s", ensg))
    fit_one_biomarker(ensg)
  })
})

saveRDS(pam50_estimates_chunk, params$estimates_rds)
```

```{r}
#| label: load-estimates
pam50_estimates_chunk <- readRDS(params$estimates_rds)

pam50_estimates_chunk %>%
  dplyr::select(.feature, symb, compression_ok, compression_rate, reproduction_pct) %>%
  head(10)
```

# Merge cancer pseudobulk with reference epithelial

```{r}
#| label: merge-se


se_cancer_epithelial <- readRDS(params$se_cancer_rds)

se_breast_epithelial_ref_sample <-
  se_breast %>%
  filter(cell_type_unified_ensemble == "epithelial") %>%
  filter(offset == 0)

se_cancer_epithelial_ensembl <- {
  sym <- rownames(se_cancer_epithelial)
  ensg <- AnnotationDbi::mapIds(
    org.Hs.eg.db::org.Hs.eg.db,
    keys = sym,
    column = "ENSEMBL",
    keytype = "SYMBOL",
    multiVals = "first"
  )
  keep <- !is.na(ensg) & !duplicated(ensg)
  ensg <- ensg[keep]

  counts_mat <- SummarizedExperiment::assay(se_cancer_epithelial, "counts")[keep, , drop = FALSE]
  rownames(counts_mat) <- ensg
  colnames(counts_mat) <- colnames(se_cancer_epithelial)

  cd <- as.data.frame(SummarizedExperiment::colData(se_cancer_epithelial))
  cd$sample_role <- "cancer_pseudobulk"
  cd$cell_type_unified_ensemble <- "epithelial"
  cd$offset <- 0

  SummarizedExperiment::SummarizedExperiment(
    assays = list(counts = counts_mat),
    colData = S4Vectors::DataFrame(cd, row.names = colnames(counts_mat)),
    rowData = S4Vectors::DataFrame(
      gene_symbol = sym[keep],
      row.names = ensg
    )
  )
}

colData(se_breast_epithelial_ref_sample)$sample_role <- "reference"

merge_summarized_experiments <- function(...) {
  ses <- list(...)
  assay_names <- Reduce(intersect, lapply(ses, SummarizedExperiment::assayNames))
  if (length(assay_names) == 0) stop("No shared assays between SummarizedExperiment objects.")

  genes <- Reduce(intersect, lapply(ses, rownames))
  if (length(genes) == 0) stop("No shared rownames between SummarizedExperiment objects.")

  ses <- lapply(ses, function(se) se[genes, , drop = FALSE])

  cds <- lapply(ses, function(se) as.data.frame(SummarizedExperiment::colData(se)))
  all_nms <- Reduce(union, lapply(cds, names))
  cds <- lapply(cds, function(cd) {
    missing <- setdiff(all_nms, names(cd))
    for (nm in missing) cd[[nm]] <- NA
    cd[, all_nms, drop = FALSE]
  })

  assays <- stats::setNames(
    lapply(assay_names, function(an) {
      do.call(cbind, lapply(ses, function(se) SummarizedExperiment::assay(se, an)))
    }),
    assay_names
  )

  SummarizedExperiment::SummarizedExperiment(
    assays = assays,
    colData = S4Vectors::DataFrame(
      do.call(rbind, cds),
      row.names = unlist(lapply(ses, colnames), use.names = FALSE)
    ),
    rowData = SummarizedExperiment::rowData(ses[[1]])
  )
}

se_epithelial <- merge_summarized_experiments(
  se_cancer_epithelial_ensembl,
  se_breast_epithelial_ref_sample
)

rowData(se_epithelial)$.abundant <-
  rowData(se_breast_epithelial_ref_sample)[
    rownames(se_epithelial),
    ".abundant_epithelial"
  ]

tibble(
  n_features = nrow(se_epithelial),
  n_samples = ncol(se_epithelial),
  n_cancer = sum(se_epithelial$sample_role == "cancer_pseudobulk"),
  n_reference = sum(se_epithelial$sample_role == "reference")
)
```

```{r}
#| label: scale-abundance
ref_samples <-
  se_epithelial %>%
  colData() %>%
  as.data.frame() %>%
  filter(sample_role == "reference") %>%
  rownames()

se_epithelial <-
  se_epithelial %>%
  tidybulk::scale_abundance(
    method = "TMMwsp",
    reference_sample = ref_samples
  ) %>%
  mutate(offset = log(1 / multiplier))
```

# Compare samples to healthy posterior predictive baseline

Posterior predictive comparison for all PAM50 biomarker genes (47) x 21 samples (20 tumour pseudobulks + 1 TMM reference).

```{r}
#| label: gene-deviation-fn
newdata_for_sample <- function(se, sample_id) {
  cd <- SummarizedExperiment::colData(se)[sample_id, , drop = FALSE] |>
    as.data.frame()

  tibble(
    age_decade = if ("age_decade" %in% names(cd)) cd$age_decade else NA,
    sex = if ("sex" %in% names(cd)) cd$sex else NA,
    disease_groups_altered = if ("disease_groups_altered" %in% names(cd)) {
      cd$disease_groups_altered
    } else {
      'Normal'
    },
    ethnicity_groups = if ("ethnicity_groups" %in% names(cd)) {
      cd$ethnicity_groups
    } else {
      NA
    },
    assay_groups_altered = if ("assay_groups_altered" %in% names(cd)) {
      cd$assay_groups_altered
    } else {
      "10x Genomics 5"
    },
    offset = cd$offset
  )
}

gene_deviation_from_fit <- function(
  fit,
  newdata,
  obs,
  cred_mass = 0.95,
  alpha = 0.025
) {
  y_rep <- brms::posterior_predict(
    fit,
    newdata = newdata,
    summary = FALSE,
    re_formula = NA
  ) |>
    as.numeric()

  hpd <- if (requireNamespace("HDInterval", quietly = TRUE)) {
    HDInterval::hdi(y_rep, credMass = cred_mass)
  } else {
    probs <- c((1 - cred_mass) / 2, 1 - (1 - cred_mass) / 2)
    stats::quantile(y_rep, probs = probs, names = FALSE)
  }

  p_above <- mean(y_rep >= obs)
  p_below <- mean(y_rep <= obs)

  tibble(
    observed_count = obs,
    healthy_median = median(y_rep),
    healthy_mean = mean(y_rep),
    healthy_hpd_lo = hpd[1],
    healthy_hpd_hi = hpd[2],
    obs_percentile = mean(y_rep <= obs),
    p_above_healthy = p_above,
    p_below_healthy = p_below,
    posterior_tail_p = 2 * min(p_above, p_below),
    log2fc_vs_median = log2((obs + 1) / (median(y_rep) + 1)),
    direction = case_when(
      p_above < alpha ~ "higher_than_healthy",
      p_below < alpha ~ "lower_than_healthy",
      TRUE ~ "consistent_with_healthy"
    )
  )
}
```

```{r}
#| label: comparison-samples
cd_epithelial <-
  se_epithelial %>%
  colData() %>%
  as.data.frame() %>%
  tibble::rownames_to_column("sample")

comparison_samples_tbl <- cd_epithelial %>%
  filter(sample_role == "cancer_pseudobulk") %>%
  bind_rows(cd_epithelial %>% filter(sample == ref_samples[1])) %>%
  distinct(sample, .keep_all = TRUE) %>%
  mutate(sample_idx = match(sample, colnames(se_epithelial)))

tibble(
  n_genes = nrow(pam50_estimates_chunk),
  n_comparison_samples = nrow(comparison_samples_tbl),
  n_tumor = sum(comparison_samples_tbl$sample_role == "cancer_pseudobulk"),
  n_reference = sum(comparison_samples_tbl$sample_role == "reference")
)
```

```{r}
#| label: predict-all
#| cache: true

handlers("progress")

with_progress({
  p <- progressor(
    steps = nrow(pam50_estimates_chunk) * nrow(comparison_samples_tbl)
  )

  gene_deviation_all <- purrr::imap_dfr(
    seq_len(nrow(pam50_estimates_chunk)),
    function(i, idx) {
      chunk_row <- pam50_estimates_chunk[i, ]
      feature_id <- chunk_row$.feature
      symb <- chunk_row$symb
      fit <- poco::reconstruct_brmsfit(chunk_row$brms_fit_compressed[[1]])

      purrr::pmap_dfr(
        comparison_samples_tbl,
        function(sample, sample_idx, sample_role, ...) {
          p()

          newdata <- newdata_for_sample(se_epithelial, sample_idx)
          obs <-
            SummarizedExperiment::assay(
              se_epithelial[feature_id, sample_idx],
              "counts"
            ) |>
            as.numeric()

          gene_deviation_from_fit(fit, newdata, obs) %>%
            mutate(
              .feature = feature_id,
              symb = symb,
              sample = as.character(sample),
              sample_role = as.character(sample_role),
              .before = observed_count
            )
        }
      )
    }
  )
})

sample_annotations <-
  se_epithelial %>%
  colData() %>%
  as.data.frame() %>%
  tibble::rownames_to_column("sample") %>%
  dplyr::distinct(sample, .keep_all = TRUE) %>%
  dplyr::select(
    sample,
    subtype,
    dplyr::any_of("orig.ident")
  )

gene_deviation_all <- gene_deviation_all %>%
  tibble::as_tibble() %>%
  dplyr::select(tidyselect::where(\(x) !is.list(x)))

if (!"sample_role" %in% colnames(gene_deviation_all)) {
  gene_deviation_all <- gene_deviation_all %>%
    dplyr::left_join(
      comparison_samples_tbl %>%
        dplyr::select(sample, sample_role),
      by = "sample"
    )
}

gene_deviation_all <- gene_deviation_all %>%
  dplyr::left_join(pam50_modules_tbl, by = c("symb" = "gene")) %>%
  dplyr::left_join(sample_annotations, by = "sample") %>%
  dplyr::relocate(module, .feature, symb, sample, sample_role, subtype)

gene_deviation_all
```

```{r}
#| label: deviation-summary
gene_deviation_tumor <-
  gene_deviation_all %>%
  dplyr::filter(sample_role == "cancer_pseudobulk") %>%
  left_join(subtype_module_expected, by = c("subtype", "module")) %>%
  mutate(
    matches_literature = dplyr::case_when(
      is.na(expected_deviation) ~ NA,
      expected_deviation == direction ~ TRUE,
      TRUE ~ FALSE
    )
  )

gene_deviation_all %>%
  dplyr::count(sample_role, direction) %>%
  pivot_wider(names_from = direction, values_from = n, values_fill = 0)

gene_deviation_tumor %>%
  dplyr::filter(!is.na(expected_deviation)) %>%
  dplyr::count(subtype, module, expected_deviation, direction) %>%
  tidyr::pivot_wider(
    names_from = direction,
    values_from = n,
    values_fill = 0
  )

gene_deviation_tumor %>%
  dplyr::filter(!is.na(expected_deviation)) %>%
  group_by(subtype, module, pam50_intrinsic, expected_deviation) %>%
  summarise(
    n_genes = dplyr::n(),
    n_match = sum(matches_literature, na.rm = TRUE),
    pct_match = 100 * n_match / n_genes,
    .groups = "drop"
  ) %>%
  arrange(subtype, module)
```

# Clustered heatmaps vs literature mapping

Tumour samples are annotated with Wu **subtype** (ER+, HER2+, TNBC); genes with PAM50 **module**. Rows and columns are clustered on `log2fc_vs_median` to see whether unsupervised structure aligns with [Parker et al. (2009)](https://pubmed.ncbi.nlm.nih.gov/19204204/) intrinsic-subtype biology.

```{r}
#| label: heatmap-matrix
module_levels <- c(
  "luminal",
  "her2",
  "basal",
  "proliferation",
  "other_pam50"
)

heatmap_df <- gene_deviation_tumor %>%
  mutate(
    module = factor(module, levels = module_levels),
    subtype = factor(subtype, levels = c("ER+", "HER2+", "TNBC"))
  )

mat_log2fc <- heatmap_df %>%
  dplyr::select(symb, sample, log2fc_vs_median) %>%
  tidyr::pivot_wider(names_from = sample, values_from = log2fc_vs_median) %>%
  tibble::column_to_rownames("symb") %>%
  as.matrix()

annotation_col <- heatmap_df %>%
  dplyr::distinct(sample, subtype) %>%
  tibble::column_to_rownames("sample") %>%
  dplyr::select(subtype)

annotation_row <- pam50_modules_tbl %>%
  dplyr::distinct(gene, module) %>%
  dplyr::filter(gene %in% rownames(mat_log2fc)) %>%
  tibble::column_to_rownames("gene") %>%
  dplyr::select(module)

ann_colors <- list(
  subtype = c(
    "ER+" = "#2ca25f",
    "HER2+" = "#756bb1",
    "TNBC" = "#de2d26"
  ),
  module = c(
    luminal = "#3182bd",
    her2 = "#e6550d",
    basal = "#31a354",
    proliferation = "#756bb1",
    other_pam50 = "#969696"
  )
)

score_mat <- heatmap_df %>%
  mutate(
    deviation_score = dplyr::case_when(
      direction == "higher_than_healthy" ~ 1L,
      direction == "lower_than_healthy" ~ -1L,
      TRUE ~ 0L
    )
  ) %>%
  dplyr::select(symb, sample, deviation_score) %>%
  tidyr::pivot_wider(names_from = sample, values_from = deviation_score) %>%
  tibble::column_to_rownames("symb") %>%
  as.matrix()
```

```{r}
#| label: plot-clustered-log2fc
#| fig-width: 11
#| fig-height: 9
#| fig-align: center
pheatmap(
  mat_log2fc,
  annotation_col = annotation_col,
  annotation_row = annotation_row,
  annotation_colors = ann_colors,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  clustering_method = "ward.D2",
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  color = colorRampPalette(rev(RColorBrewer::brewer.pal(11, "RdBu")))(100),
  breaks = seq(
    -max(abs(mat_log2fc), na.rm = TRUE),
    max(abs(mat_log2fc), na.rm = TRUE),
    length.out = 101
  ),
  border_color = NA,
  fontsize_row = 7,
  fontsize_col = 8,
  main = "log2 FC vs healthy predictive median\n(clustered; subtype x module annotations)"
)
```

```{r}
#| label: plot-clustered-direction
#| fig-width: 11
#| fig-height: 9
#| fig-align: center
pheatmap(
  score_mat,
  annotation_col = annotation_col,
  annotation_row = annotation_row,
  annotation_colors = ann_colors,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  clustering_method = "ward.D2",
  color = colorRampPalette(c("#2166ac", "#f7f7f7", "#b2182b"))(3),
  breaks = c(-1.5, -0.5, 0.5, 1.5),
  legend_breaks = c(-1, 0, 1),
  legend_labels = c("lower", "consistent", "higher"),
  border_color = NA,
  fontsize_row = 7,
  fontsize_col = 8,
  main = "Deviation direction vs healthy baseline\n(clustered; subtype x module annotations)"
)
```

```{r}
#| label: plot-literature-agreement
#| fig-width: 9
#| fig-height: 5
agreement_by_subtype_module <- gene_deviation_tumor %>%
  dplyr::filter(!is.na(expected_deviation)) %>%
  group_by(subtype, module, expected_deviation) %>%
  summarise(
    pct_match = 100 * mean(matches_literature, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    module = factor(module, levels = module_levels),
    subtype = factor(subtype, levels = c("ER+", "HER2+", "TNBC"))
  )

ggplot(
  agreement_by_subtype_module,
  aes(x = module, y = subtype, fill = pct_match)
) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.0f%%", pct_match)), size = 3.5) +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "white",
    high = "#b2182b",
    midpoint = 50,
    limits = c(0, 100),
    name = "% genes matching\nliterature expectation"
  ) +
  labs(
    title = "Literature concordance (Parker PAM50 mapping)",
    subtitle = "Per subtype x module: % genes with expected deviation direction",
    x = "PAM50 module",
    y = "Wu tumour subtype"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

```{r}
#| label: plot-ordered-heatmap
#| fig-width: 11
#| fig-height: 9
#| fig-align: center
sample_order <- heatmap_df %>%
  dplyr::distinct(sample, subtype) %>%
  dplyr::arrange(subtype, sample) %>%
  pull(sample)

gene_order <- pam50_modules_tbl %>%
  dplyr::filter(gene %in% rownames(mat_log2fc)) %>%
  dplyr::mutate(module = factor(module, levels = module_levels)) %>%
  dplyr::arrange(module, gene) %>%
  pull(gene)

pheatmap(
  mat_log2fc[gene_order, sample_order, drop = FALSE],
  annotation_col = annotation_col[sample_order, , drop = FALSE],
  annotation_row = annotation_row[gene_order, , drop = FALSE],
  annotation_colors = ann_colors,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  color = colorRampPalette(rev(RColorBrewer::brewer.pal(11, "RdBu")))(100),
  breaks = seq(
    -max(abs(mat_log2fc), na.rm = TRUE),
    max(abs(mat_log2fc), na.rm = TRUE),
    length.out = 101
  ),
  border_color = NA,
  fontsize_row = 7,
  fontsize_col = 8,
  main = "log2 FC (literature order: modules x subtypes)\nno clustering — reference layout for mapping"
)
```

## Example: ERBB2 density plot

```{r}
#| label: plot-example
#| fig-width: 7
#| fig-height: 4
example_row <- gene_deviation_all %>%
  dplyr::filter(symb == "ERBB2", sample_role == "cancer_pseudobulk") %>%
  dplyr::slice(1)

i <- match(example_row$.feature, pam50_estimates_chunk$.feature)
j <- match(example_row$sample, colnames(se_epithelial))

fit_compressed <- pam50_estimates_chunk$brms_fit_compressed[[i]] %>%
  poco::reconstruct_brmsfit()

newdata <- newdata_for_sample(se_epithelial, j)
obs <- example_row$observed_count

y_rep <- brms::posterior_predict(
  fit_compressed,
  newdata = newdata,
  summary = FALSE,
  re_formula = NA
) |>
  as.numeric()

tibble(draw = y_rep) %>%
  ggplot(aes(x = draw)) +
  geom_density(fill = "steelblue", alpha = 0.4) +
  geom_vline(xintercept = obs, colour = "firebrick", linewidth = 1) +
  geom_vline(
    xintercept = c(example_row$healthy_hpd_lo, example_row$healthy_hpd_hi),
    linetype = "dashed",
    colour = "grey40"
  ) +
  labs(
    title = glue("{example_row$symb} — {example_row$sample}"),
    subtitle = glue(
      "{example_row$direction} (p_above = {round(example_row$p_above_healthy, 4)})"
    ),
    x = "Posterior predictive count (healthy reference)",
    y = "Density"
  ) +
  theme_minimal()
```
