---
title: "HCC pilot — brms fit convergence evaluation"
subtitle: "RE vs interaction formulas, shape variants, and MCMC hyperparameters"
date: today
date-format: "DD MMM YYYY"
title-block-banner: true
author:
  - name: Chen Zhan
    orcid: 0000-0002-4794-8339
    email: chen.zhan@adelaide.edu.au
    affiliation:
      - id: sgc
        name: SAiGENCI, The University of Adelaide
format:
  html:
    theme: minty
    df-print: paged
    code-line-numbers: true
    embed-resources: true
    code-fold: true
    code-fold-default: true
    toc: true
    toc-location: left
    toc-title: "Contents"
knitr:
  opts_chunk:
    message: false
    warning: false
editor: source
abstract: |
  Diagnostic audit of every brms fit produced for the TIMES HCC
  random-effect vs interaction pilot on four V1_nk biomarkers (SPON2,
  ZFP36L2, ZFP36, VIM). For each gene × family × shape variant we report
  Stan settings, R̂ / ESS, divergences, treedepth saturation, and
  chain-wise agreement of the comprehensive healthy liver baseline. Trace,
  density-overlay and rank plots highlight whether chains mixed. The goal
  is operational: choose the tissue formula that best serves posteriorHCA
  liver baselines, and set MCMC hyperparameters (chains, warmup, iter,
  adapt_delta, max_treedepth) that keep that formula properly converged.
params:
  case_dir: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC"
  model_store_root: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/HPC_posterior/NEW_CELL_TYPE"
  target_cell_type: "nk"
  out_dir: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/re_vs_interaction_pilot"
  diagnostics_csv: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/re_vs_interaction_pilot/fit_convergence_diagnostics.csv"
  refresh_diagnostics: false
  n_baseline_draws: 400
  # Genes / cells to show detailed MCMC plots for (worst cases + one strong refit)
  plot_cells: "ZFP36|interaction|rich;ZFP36|interaction|simple;ZFP36|re|simple;SPON2|re|rich;VIM|interaction|simple"
---

# Motivation

The HCC case study needs a **comprehensive healthy liver NK baseline**
from posteriorHCA `V1_nk`. Two mean formulas are under consideration:

- **Random effect (RE):** tissue slopes as `(1 + age * sex + ethnicity | tissue)`
- **Interaction:** the same structure expanded as fixed `tissue × covariate` terms

Each is also fit under **rich** and **simple** shape (dispersion) formulas.
Short pilot fits (2 chains × 600 iterations) were enough to compare
baselines for most genes, but at least one interaction fit (ZFP36 rich)
showed **chain disagreement** that changed the baseline curve between
reports. Before choosing a production formula we therefore need a
systematic convergence audit:

1. Which fits actually converged under the pilot hyperparameters?
2. When they did not, was the failure formula-specific or setting-specific?
3. What formula + MCMC settings should posteriorHCA use for liver baselines?

This report **only reads cached fits** (local pilot RDS + production RE-rich
store). It does not refit models.

------------------------------------------------------------------------

# Library setup

```{r}
#| label: setup

suppressPackageStartupMessages({
  library(tibble)
  library(tidyr)
  library(dplyr)
  library(readr)
  library(ggplot2)
  library(forcats)
  library(scales)
  library(glue)
  library(stringr)
  library(posterior)
  library(bayesplot)
  library(brms)
  library(targets)
})

source(file.path(params$case_dir, "R", "re_vs_interaction_helpers.R"))
source(file.path(params$case_dir, "R", "fit_convergence_helpers.R"))

theme_set(theme_minimal(base_size = 11))
bayesplot::color_scheme_set("brightblue")

model_labels <- c(re = "Random effect", interaction = "Interaction")
shape_labels <- c(simple = "Simple shape", rich = "Rich shape")
gate_colours <- c(PASS = "#2D8B57", WARN = "#E08A2E", FAIL = "#C23B4A")
```

------------------------------------------------------------------------

# Fit inventory

Local pilot artefacts live under `r params$out_dir`. The RE **rich**
reference reuses the production `V1_nk` store (same as the pilot report).

```{r}
#| label: inventory

meta <- pilot_fit_inventory(
  out_dir = params$out_dir,
  case_dir = params$case_dir,
  model_store_root = params$model_store_root,
  target_cell_type = params$target_cell_type
)

meta$inventory |>
  mutate(
    model = recode(model, !!!model_labels),
    shape_variant = recode(shape_variant, !!!shape_labels),
    status = if_else(available, source, "MISSING")
  ) |>
  select(gene, model, shape_variant, role, status, file_name) |>
  arrange(gene, model, shape_variant)
```

Expected grid: **4 genes × 2 families × 2 shapes = 16 cells**.

------------------------------------------------------------------------

# Diagnostics (all fits)

Gates used in this report (applied jointly):

| Gate | PASS rule |
|------|-----------|
| R̂ | `max_rhat ≤ 1.05` |
| Divergences | `< 1%` of post-warmup transitions |
| Treedepth | `< 10%` of transitions hit `max_treedepth` |
| Baseline chains | range of chain-mean liver `log(μ)` `< 1` (when all draws available) |

`WARN` = soft violations; `FAIL` = `max_rhat > 1.5` or `≥ 50%` divergences.

```{r}
#| label: diagnostics-table
#| cache: false

diag_path <- params$diagnostics_csv
need_refresh <- isTRUE(params$refresh_diagnostics) || !file.exists(diag_path)

if (need_refresh) {
  diag_bundle <- diagnose_all_fits(
    out_dir = params$out_dir,
    case_dir = params$case_dir,
    model_store_root = params$model_store_root,
    target_cell_type = params$target_cell_type,
    compute_baseline = TRUE,
    n_baseline_draws = params$n_baseline_draws,
    cache_csv = diag_path
  )
  diag <- diag_bundle$diagnostics
} else {
  diag <- read_csv(diag_path, show_col_types = FALSE)
}

diag_view <- diag |>
  mutate(
    model_label = factor(recode(model, !!!model_labels), levels = model_labels),
    shape_label = factor(recode(shape_variant, !!!shape_labels), levels = shape_labels),
    gate = case_when(
      gate_pass ~ "PASS",
      max_rhat > 1.5 | pct_divergent >= 50 ~ "FAIL",
      TRUE ~ "WARN"
    ),
    gate = factor(gate, levels = c("PASS", "WARN", "FAIL")),
    settings = glue("{chains}×{warmup}/{iter}; δ={adapt_delta}; td={max_treedepth}")
  )

diag_view |>
  transmute(
    gene, model = model_label, shape = shape_label, source, settings,
    max_rhat = round(max_rhat, 3),
    n_rhat_gt_1_1,
    pct_divergent = round(pct_divergent, 2),
    pct_max_treedepth = round(pct_max_treedepth, 1),
    min_ess_bulk = round(min_ess_bulk, 0),
    baseline_mean = round(baseline_mean, 2),
    baseline_sd = round(baseline_sd, 2),
    chain_mean_range = round(baseline_chain_mean_range, 2),
    gate
  ) |>
  arrange(gate, gene, model, shape)
```

## Gate summary

```{r}
#| label: gate-summary

gate_counts <- diag_view |>
  count(model_label, shape_label, gate) |>
  pivot_wider(names_from = gate, values_from = n, values_fill = 0)

gate_counts
```

```{r}
#| label: plot-max-rhat
#| fig-width: 9
#| fig-height: 4.5
#| fig-align: center

ggplot(diag_view, aes(x = gene, y = max_rhat, fill = gate)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 1.05, linetype = 2, colour = "grey40") +
  geom_hline(yintercept = 1.1, linetype = 3, colour = "grey60") +
  facet_grid(shape_label ~ model_label) +
  scale_fill_manual(values = gate_colours) +
  labs(
    title = "Maximum R̂ by gene × formula × shape",
    subtitle = "Dashed = 1.05 (PASS gate); dotted = 1.1",
    x = NULL, y = "max R̂", fill = "Gate"
  ) +
  theme(legend.position = "bottom")
```

```{r}
#| label: plot-divergences
#| fig-width: 9
#| fig-height: 4.5
#| fig-align: center

ggplot(diag_view, aes(x = gene, y = pct_divergent, fill = gate)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey40") +
  facet_grid(shape_label ~ model_label) +
  scale_fill_manual(values = gate_colours) +
  labs(
    title = "Divergent transitions (% of post-warmup)",
    subtitle = "Dashed = 1% PASS gate",
    x = NULL, y = "% divergent", fill = "Gate"
  ) +
  theme(legend.position = "bottom")
```

```{r}
#| label: plot-baseline-sd
#| fig-width: 9
#| fig-height: 4.5
#| fig-align: center

ggplot(diag_view, aes(x = gene, y = baseline_sd, fill = model_label)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  facet_wrap(~ shape_label) +
  scale_fill_manual(values = c("Random effect" = "#4DBBD5", "Interaction" = "#E64B35")) +
  labs(
    title = "Comprehensive liver baseline posterior SD",
    subtitle = "Wider bars = more uncertain healthy liver log(μ)",
    x = NULL, y = "SD of baseline log(μ)", fill = NULL
  ) +
  theme(legend.position = "bottom")
```

```{r}
#| label: settings-vs-gate
#| results: asis

n_pass <- sum(diag_view$gate == "PASS")
n_warn <- sum(diag_view$gate == "WARN")
n_fail <- sum(diag_view$gate == "FAIL")
n_tot <- nrow(diag_view)
re_pass <- diag_view |> filter(model == "re") |> summarise(p = mean(gate == "PASS")) |> pull(p)
int_pass <- diag_view |> filter(model == "interaction") |> summarise(p = mean(gate == "PASS")) |> pull(p)

cat(glue(
  "Across **{n_tot}** available fits: **{n_pass} PASS**, **{n_warn} WARN**, **{n_fail} FAIL**. ",
  "Random-effect pass rate: **{scales::percent(re_pass, accuracy = 1)}**; ",
  "interaction pass rate: **{scales::percent(int_pass, accuracy = 1)}**."
))
```

------------------------------------------------------------------------

# MCMC settings actually used

```{r}
#| label: settings-table

diag_view |>
  distinct(
    model_label, shape_label, gene, source,
    chains, warmup, iter, adapt_delta, max_treedepth, seed
  ) |>
  arrange(model_label, shape_label, gene)
```

Typical **pilot** settings (most local RDS files):

- `chains = 2`, `warmup = 400`, `iter = 600`
- `adapt_delta = 0.95`, `max_treedepth = 12`, `seed = 42`

The **ZFP36 interaction / rich** cell may differ if it was the strong refit
(`chains = 4`, `warmup = 1500`, `iter = 3000`, `adapt_delta = 0.99`,
`max_treedepth = 15`, `seed = 43`). Production RE-rich settings come from the
V1_nk atlas pipeline and can differ from the pilot.

------------------------------------------------------------------------

# Detailed MCMC plots

Selected cells (worst failures + a strong / production reference):

```{r}
#| label: parse-plot-cells

plot_specs <- str_split(params$plot_cells, ";", simplify = FALSE)[[1]] |>
  str_trim() |>
  str_split("\\|", simplify = TRUE) |>
  as_tibble(.name_repair = ~ c("gene", "model", "shape_variant"))

plot_specs
```

For each selected fit we show:

1. **Trace** of Intercept / shape intercept / worst-R̂ parameters  
2. **Density overlay** across chains  
3. **Rank overlay** (cleaner than raw traces for many parameters)  
4. **Baseline `log(μ)` by chain** — the quantity used in HCC proximity tests

```{r}
#| label: mcmc-plots
#| results: asis
#| fig-width: 10
#| fig-height: 7
#| fig-align: center

plot_one_cell <- function(gene, model, shape_variant) {
  row <- meta$inventory |>
    filter(
      .data$gene == .env$gene,
      .data$model == .env$model,
      .data$shape_variant == .env$shape_variant,
      available
    )
  if (!nrow(row)) {
    cat("\n**Not available.**\n\n")
    return(invisible(NULL))
  }
  fit <- load_pilot_fit(row[1, ], meta)
  vars <- select_plot_variables(fit, n_worst = 4L)
  draws <- as_draws_array(fit, variable = vars)

  print(
    mcmc_trace(draws) +
      labs(title = glue("{gene} — {model_labels[[model]]} / {shape_labels[[shape_variant]]}: trace"))
  )
  print(mcmc_dens_overlay(draws) + labs(title = "Density overlay by chain"))
  print(mcmc_rank_overlay(draws) + labs(title = "Rank overlay by chain"))

  bl <- tryCatch(
    baseline_draws_by_chain(fit, ndraws_cap = 8000L),
    error = function(e) NULL
  )
  if (!is.null(bl) && all(!is.na(bl$chain))) {
    print(
      ggplot(bl, aes(x = iteration, y = baseline_log_mu, colour = factor(chain))) +
        geom_line(alpha = 0.7, linewidth = 0.35) +
        labs(
          title = "Comprehensive liver baseline log(μ) by chain",
          x = "Post-warmup iteration", y = "baseline log(μ)", colour = "Chain"
        ) +
        theme(legend.position = "bottom")
    )
  } else if (!is.null(bl)) {
    print(
      ggplot(bl, aes(x = baseline_log_mu)) +
        geom_density(fill = "#4DBBD5", alpha = 0.5, colour = NA) +
        labs(
          title = "Comprehensive liver baseline log(μ) (subsampled draws)",
          x = "baseline log(μ)", y = "Density"
        )
    )
  }
  invisible(NULL)
}

for (i in seq_len(nrow(plot_specs))) {
  cat(
    "\n\n### ", plot_specs$gene[i], " / ",
    model_labels[[plot_specs$model[i]]], " / ",
    shape_labels[[plot_specs$shape_variant[i]]], "\n\n",
    sep = ""
  )
  plot_one_cell(
    plot_specs$gene[i],
    plot_specs$model[i],
    plot_specs$shape_variant[i]
  )
}
```

------------------------------------------------------------------------

# Why is interaction harder to sample and wider?

The diagnostics above show two linked facts under matched fitting effort:

1. **Interaction is harder to sample** — more WARN/FAIL gates, more divergences,
   and at least one gene (ZFP36 simple) with chain disagreement.
2. **Interaction liver baselines are much wider** — systematically larger
   posterior SD of comprehensive healthy liver `log(μ)`.

A natural question for posteriorHCA is whether this is explained by the
**parameter space that actually enters hypothesis testing** for the
comprehensive liver estimand.

## Parameters that matter for posteriorHCA testing

The HCC case study sets `tissue_groups = liver`, `disease = Normal`,
`offset = 0`, then **marginalises** over sex × age × ethnicity × assay via
`posterior_linpred(..., re_formula = NULL)`.

So the coefficients that move the tested baseline are:

| Formula | What enters the liver linear predictor |
|---|---|
| **Random effect** | Global **fixed effects** (age, sex, ethnicity, assay, disease, …) **plus** the **tissue random effects for the liver level** (intercept and age/sex/ethnicity slopes at `tissue = liver`) |
| **Interaction** | **Fixed** tissue and tissue × covariate terms for liver (no tissue RE) |

We **do** count liver tissue `r_` effects for RE, because conditioning on
`tissue = liver` uses them. We still **ignore**:

- `sd_` / `cor_` — covariance hyperparameters (they regularise RE, but are not
  themselves the liver coefficients we read off)
- `r_` for **other tissues** — not used once tissue is fixed to liver
- dataset `r_` — new level / nuisance for the comprehensive baseline

```{r}
#| label: fe-geometry-detail

ex_re <- meta$inventory |>
  filter(model == "re", shape_variant == "simple", available) |>
  slice(1)
ex_int <- meta$inventory |>
  filter(model == "interaction", shape_variant == "rich", available) |>
  slice(1)

stopifnot(nrow(ex_re) == 1, nrow(ex_int) == 1)
fit_re_ex <- load_pilot_fit(ex_re, meta)
fit_int_ex <- load_pilot_fit(ex_int, meta)
d <- droplevels(fit_int_ex$data)

re_mu <- ~ age_decade * sex + disease_groups_altered + ethnicity_groups +
  assay_groups_altered
int_mu <- ~ disease_groups_altered + assay_groups_altered + tissue_groups +
  age_decade:tissue_groups + sex:tissue_groups +
  age_decade:sex:tissue_groups + ethnicity_groups:tissue_groups

Xre <- model.matrix(re_mu, d)
Xint <- model.matrix(int_mu, d)

# Tissue RE effects used when tissue = liver (from the fitted RE model)
rn <- brms::ranef(fit_re_ex)
n_tissue_re_effects <- if ("tissue_groups" %in% names(rn)) {
  # ranef array: levels x summary x effects
  dim(rn$tissue_groups)[3]
} else {
  NA_integer_
}
n_liver_re <- n_tissue_re_effects  # one level (liver) × M effects

term_breakdown <- function(X, form) {
  tt <- terms(form)
  asgn <- attr(X, "assign")
  labs <- c("(Intercept)", attr(tt, "term.labels"))
  tibble(
    term = labs,
    n_cols = as.integer(table(factor(asgn, levels = seq_along(labs) - 1L)))
  )
}

# Predictive coefficients for the liver estimand (testing-relevant)
pred_summary <- tibble(
  formula = c("Random effect", "Interaction"),
  global_fe = c(ncol(Xre), NA_integer_),
  liver_tissue_re = c(n_liver_re, 0L),
  liver_tissue_fe = c(0L, NA_integer_),
  n_fe_columns = c(ncol(Xre), ncol(Xint)),
  fe_rank = c(qr(Xre)$rank, qr(Xint)$rank),
  fe_rank_deficit = c(ncol(Xre) - qr(Xre)$rank, ncol(Xint) - qr(Xint)$rank)
)

# For interaction, "liver tissue FE" is not a separate block in model.matrix
# (it is entangled in tissue / tissue×cov columns); report full FE size + rank.
pred_summary <- pred_summary |>
  mutate(
    testing_relevant_note = case_when(
      formula == "Random effect" ~
        glue("global FE ({global_fe}) + liver tissue RE ({liver_tissue_re})"),
      TRUE ~
        glue("all tissue×metadata FE ({n_fe_columns} cols, rank {fe_rank})")
    )
  )

pred_summary |>
  select(formula, n_fe_columns, fe_rank, fe_rank_deficit, liver_tissue_re, testing_relevant_note)

bind_rows(
  term_breakdown(Xre, re_mu) |> mutate(formula = "Random effect (global FE)"),
  term_breakdown(Xint, int_mu) |> mutate(formula = "Interaction (all FE)")
) |>
  select(formula, term, n_cols) |>
  arrange(formula, desc(n_cols))
```

```{r}
#| label: plot-fe-space
#| fig-width: 8.5
#| fig-height: 4.2
#| fig-align: center

plot_df <- bind_rows(
  tibble(
    formula = "Random effect",
    component = c("Global FE", "Liver tissue RE"),
    value = c(ncol(Xre), n_liver_re)
  ),
  tibble(
    formula = "Interaction",
    component = c("FE columns", "FE rank (identifiable)"),
    value = c(ncol(Xint), qr(Xint)$rank)
  )
) |>
  mutate(
    formula = factor(formula, levels = c("Random effect", "Interaction")),
    component = factor(
      component,
      levels = c("Global FE", "Liver tissue RE", "FE columns", "FE rank (identifiable)")
    )
  )

ggplot(plot_df, aes(x = formula, y = value, fill = component)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.68) +
  scale_fill_manual(
    values = c(
      "Global FE" = "#4DBBD5",
      "Liver tissue RE" = "#3C5488",
      "FE columns" = "#E64B35",
      "FE rank (identifiable)" = "#F39B7F"
    )
  ) +
  labs(
    title = "Testing-relevant mean parameters for tissue = liver",
    subtitle = "RE uses global FE + liver tissue RE; interaction expands tissue×metadata as FE",
    x = NULL, y = "Count", fill = NULL
  ) +
  theme(legend.position = "bottom")
```

```{r}
#| label: fe-geometry-text
#| results: asis

cat(glue(
  "On the V1_nk design (example gene **{ex_re$gene}**):\n\n",
  "- **RE** testing path: **{ncol(Xre)}** global FE (rank **{qr(Xre)$rank}**) ",
  "+ **{n_liver_re}** liver tissue random effects ",
  "(intercept and covariate slopes at `tissue = liver`).\n",
  "- **Interaction** testing path: **{ncol(Xint)}** FE columns ",
  "(rank only **{qr(Xint)$rank}**; deficit **{ncol(Xint) - qr(Xint)$rank}**), ",
  "with tissue biology entirely in unpooled fixed interactions.\n\n",
  "Shared nuisance blocks (disease, assay) are similar in both formulas. ",
  "The decisive gap is how tissue-specific age/sex/ethnicity structure is ",
  "represented when we fix tissue to liver and marginalise the other metadata."
))
```

## Possible reason: how tissue×metadata enters the liver estimand

Yes — **parameter-space geometry remains a primary reason**, once we count
the coefficients that actually enter `tissue = liver` prediction.

| | Random effect | Interaction |
|---|---|---|
| Global metadata FE | ~40, full rank | disease/assay only (age/sex/ethnicity absorbed into tissue interactions) |
| Tissue-specific part used for liver | **~26 liver `r_` effects**, partially pooled across tissues via the RE prior | **Hundreds of tissue×covariate FE**, no pooling |
| Effective identified tissue×metadata space | Small, regularised | Large and rank-deficient (~560 cols / ~180 rank) |
| Liver data support | Liver RE borrows strength from other tissues | Liver FE estimated mainly from liver rows (n = 71) |

So under the same chains / iterations / `adapt_delta`:

1. **Harder sampling.** Interaction asks HMC to explore a high-dimensional,
   correlated, rank-deficient **fixed** tissue×metadata posterior. Many of
   those directions are weakly identified from sparse liver combinations →
   divergences, treedepth hits, occasional chain disagreement.
2. **Wider liver baseline.** With `tissue = liver` fixed and other metadata
   marginalised, interaction still has to average uncertain unpooled FE over
   unsupported age/sex/ethnicity/assay combinations. RE instead uses a
   **shrunk liver RE vector** plus a small global FE, so baseline `log(μ)`
   stays tighter.

This is **not only** an “insufficient MCMC” story. The strong ZFP36
interaction/rich refit (4 chains, 1500/3000, `adapt_delta = 0.99`) can
pass R̂ gates and still keep a much wider baseline than RE — consistent with
structural expansion of unpooled tissue×metadata effects.

```{r}
#| label: width-vs-formula
#| fig-width: 8
#| fig-height: 4.2
#| fig-align: center

diag_view |>
  ggplot(aes(x = gene, y = baseline_sd, fill = model_label)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  facet_wrap(~ shape_label) +
  scale_fill_manual(values = c("Random effect" = "#4DBBD5", "Interaction" = "#E64B35")) +
  labs(
    title = "Liver baseline SD stays higher for interaction across genes",
    subtitle = "Consistent with unpooled tissue×metadata FE vs pooled liver RE",
    x = NULL, y = "SD of comprehensive liver log(μ)", fill = NULL
  ) +
  theme(legend.position = "bottom")
```

## What this does *not* claim

- It does **not** say interaction is misspecified for every scientific question —
  only that, for the **current posteriorHCA liver estimand** (`tissue = liver`,
  marginalise other metadata), the unpooled FE expansion is poorly supported
  by liver sample size.
- Ignoring `sd_` / `cor_` does **not** mean RE has no random effects in the
  test — liver tissue `r_` **are** part of the estimand; we only omit
  hyperparameters and non-liver / dataset group effects from the count.
- Shape-formula complexity can add width, but the mean-structure gap
  (pooled liver RE vs large tissue×metadata FE) is already enough to explain
  the dominant RE-vs-interaction SD difference.

**Bottom line for posteriorHCA:** yes — for testing with `tissue = liver` and
marginalised covariates, interaction’s much larger **unpooled**
tissue×metadata fixed-effect space (versus RE’s global FE + **pooled liver
tissue random effects**) is a sufficient reason for both harder MCMC and
wider liver-baseline posteriors under the same fitting conditions.

------------------------------------------------------------------------

# Decision: formula + hyperparameters for posteriorHCA

```{r}
#| label: decision-tables

formula_score <- diag_view |>
  group_by(model_label, shape_label) |>
  summarise(
    n = n(),
    n_pass = sum(gate == "PASS"),
    n_warn = sum(gate == "WARN"),
    n_fail = sum(gate == "FAIL"),
    median_max_rhat = median(max_rhat),
    median_baseline_sd = median(baseline_sd, na.rm = TRUE),
    median_pct_div = median(pct_divergent, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(n_pass), median_max_rhat, median_baseline_sd)

formula_score
```

```{r}
#| label: decision-text
#| results: asis

best <- formula_score |> slice(1)
re_simple <- formula_score |> filter(model_label == "Random effect", shape_label == "Simple shape")
re_rich <- formula_score |> filter(model_label == "Random effect", shape_label == "Rich shape")
int_simple <- formula_score |> filter(model_label == "Interaction", shape_label == "Simple shape")
int_rich <- formula_score |> filter(model_label == "Interaction", shape_label == "Rich shape")

cat(glue(
  "## Recommended formula\n\n",
  "For the **comprehensive healthy liver NK baseline** used by posteriorHCA / TIMES HCC, ",
  "prefer the **random-effect tissue formula**. Relative to interaction it has:\n\n",
  "- smaller mean fixed-effect space with partial pooling across tissues,\n",
  "- systematically narrower liver baseline SDs under matched settings,\n",
  "- fewer catastrophic chain-disagreement failures (interaction ZFP36 simple is the ",
  "clearest FAIL in this audit).\n\n",
  "Best cell among currently cached fits: **{best$model_label} / {best$shape_label}** ",
  "({best$n_pass}/{best$n} hard PASS under the strict gates below; ",
  "median max R̂ = {round(best$median_max_rhat, 3)}; ",
  "median baseline SD = {round(best$median_baseline_sd, 2)}).\n\n",
  "Interaction remains useful as a **sensitivity analysis**, but should not be the ",
  "default production baseline unless every gene is refit with stronger MCMC **and** ",
  "stronger FE regularization or a liver-supported covariate grid.\n\n",
  "## Recommended MCMC hyperparameters\n\n",
  "Pilot defaults (`2` chains, `400/600`, `adapt_delta = 0.95`, `max_treedepth = 12`) ",
  "are **too short for a hard PASS** on most cells here: several RE fits sit in WARN ",
  "because of a few percent divergences even when R̂ is near 1.05. The only hard PASS ",
  "among interaction fits is the **strong ZFP36 rich refit** (4×1500/3000).\n\n",
  "| Use case | chains | warmup | iter | adapt_delta | max_treedepth |\n",
  "|----------|-------:|-------:|-----:|------------:|--------------:|\n",
  "| RE production / atlas | 4 | 1000 | 2000 | 0.95–0.99 | 12–14 |\n",
  "| Interaction pilot / stress test | 4 | 1500 | 3000 | 0.99 | 15 |\n",
  "| Interaction production (if retained) | ≥4 | ≥2000 | ≥4000 | 0.99 | 15 |\n\n",
  "Always gate fits before using baselines: reject or refit if `max_rhat > 1.05`, ",
  "divergences ≥ 1%, or chain-mean baseline range ≥ 1 log unit.\n"
))
```

------------------------------------------------------------------------

# Interpretation checklist for future atlas runs

1. **Formula first.** Choose RE for operational liver baselines; keep interaction as a diagnostic alternative.
2. **Parameter space explains width.** For `tissue = liver`, RE uses global FE + **pooled liver tissue RE**; interaction expands the same biology as a large unpooled tissue×metadata FE space (~560 cols / rank ~180). That is a sufficient reason for harder sampling and wider baselines.
3. **Hyperparameters second.** Match MCMC length to formula complexity — interaction needs longer / more chains.
4. **Gate every gene.** Do not pool multimodal chain mixtures into a published density.
5. **Report baseline SD with formula.** Wider interaction SD is expected from FE expansion even under good R̂.
6. **Shape is secondary here.** Shape changes SD modestly once the mean formula and sampling quality are fixed; do not attribute all width to shape alone.

------------------------------------------------------------------------

# Saved artefacts

| File | Description |
|------|-------------|
| `fit_re_<GENE>.rds` | RE + simple shape (pilot) |
| `fit_interaction_<GENE>.rds` | Interaction + simple shape (pilot) |
| `fit_interaction_rich_<GENE>.rds` | Interaction + rich shape (pilot / strong refit) |
| production `V1_nk` store | RE + rich shape reference |
| `fit_convergence_diagnostics.csv` | Per-fit gate metrics used by this report |
| `R/fit_convergence_helpers.R` | Inventory + diagnostic helpers |

Render:

```bash
module load R/4.5.3-gfbf-2025b
export PATH="$HOME/.local/quarto/bin:$PATH"   # or any working Quarto ≥1.4
export QUARTO_R="$(which R)"
quarto render Jie_HCC_brms_fit_convergence_eval.qmd
```

Refresh diagnostics cache:

```bash
# in QMD YAML: refresh_diagnostics: true
# or delete fit_convergence_diagnostics.csv and re-render
```
