---
title: "TIMES HCC case study - personalised posteriorHCA baseline"
subtitle: "Patient-specific healthy liver NK predictive and heteroscedastic log-mean tests"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor: source
abstract: |
  This companion analysis replaces the comprehensive cohort-level baseline with
  a personalised healthy liver NK baseline for each HCC patient-compartment
  pseudobulk. Known age decade, sex, liver tissue, and the aligned TMMwsp offset
  are retained. Unknown ethnicity and the unseen GeoMx assay level are
  marginalised using their empirical joint distribution among healthy liver NK
  atlas samples. A new GeoMx dataset effect is drawn from the fitted hierarchical
  distribution. Because each comparison is one observed count against a
  posterior distribution, the primary test uses `posterior_predict()`.
  A secondary `posterior_linpred()` analysis estimates a patient log mean and
  NB sampling SE, then applies a heteroscedastic normal (Welch-style) test
  against the matched healthy posterior mean and SD.
  `posterior_epred()` summarises the personalised healthy expected count and
  effect size.
params:
  case_dir: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC"
  model_store_root: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/HPC_posterior/NEW_CELL_TYPE"
  target_cell_type: "nk"
  baseline_tissue_group: "liver"
  expression_csv: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/codeocean_GeoMx_all_gene_expression_matrix.csv"
  roi_metadata_csv: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/codeocean_GeoMx_all_roi_metadata.csv"
  normalization_csv: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/Jie_HCC_IF_TC_TMMwsp_offsets.csv"
  patient_metadata_csv: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/Jie_HCC_personalised_baseline_patient_metadata.csv"
  test_results_csv: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/Jie_HCC_personalised_baseline_gene_patient_tests.csv"
  patient_scores_csv: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/Jie_HCC_personalised_baseline_patient_scores.csv"
  performance_csv: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/Jie_HCC_personalised_baseline_performance.csv"
  linpred_test_results_csv: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/Jie_HCC_personalised_baseline_linpred_welch_tests.csv"
  linpred_patient_scores_csv: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/Jie_HCC_personalised_baseline_linpred_welch_patient_scores.csv"
  predictive_draws_rds: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/Jie_HCC_personalised_baseline_predictive_draws.rds"
---

# Analysis question

The comprehensive-baseline report estimates one healthy grand mean per gene and
one mean per HCC cohort. This report instead asks a patient-level question:

> Is one observed HCC patient-compartment gene count surprising under the
> posterior predictive distribution of a metadata-matched healthy liver NK
> sample at the same library-size offset?

The healthy counterfactual keeps only information known for that patient and
represented by the atlas model:

| Model term | Personalised treatment |
|---|---|
| `age_decade` | Patient age converted with `ceiling(age / 10)` |
| `sex` | Patient sex, lower-case atlas level |
| `tissue_groups` | Fixed to `liver` |
| `disease_groups_altered` | Fixed to `Normal` |
| `offset` | Patient-specific TMMwsp offset from the SAVI-aligned normalization |
| `ethnicity_groups` | Unknown; empirically marginalised over healthy liver NK |
| `assay_groups_altered` | GeoMx is unseen by the model; empirically marginalised |
| `dataset_id_altered` | New shared GeoMx dataset level drawn hierarchically |

The assay term cannot be set to `GeoMx_DSP`: it is a fixed effect and GeoMx was
not a fitted level. Treating it as a known atlas assay would invent information.

## Which posterior quantity answers which question

For a single observed count, the functions answer different questions:

| Function | Quantity | Role here |
|---|---|---|
| `posterior_linpred(transform = FALSE)` | Latent log count-component mean, `log(mu)` | Matched healthy mean and SD for a secondary heteroscedastic log-mean test |
| `posterior_epred()` | Expected observed count, including zero inflation but excluding observation noise | Effect-size summary |
| `posterior_predict()` | Replicated observed count, including parameter, random-effect, zero-inflation and NB sampling uncertainty | **Primary predictive test** |

For each gene-sample pair, the report computes lower and upper posterior
predictive tail probabilities, a two-sided predictive tail probability, the
central 95% predictive interval, and a mid-rank predictive percentile. The
percentile is transformed to a standard-normal score so that absolute values
can be averaged across genes for an outcome-free patient deviation score.
Benjamini-Hochberg values are exploratory because only four atlas-covered TIMES
genes and five patient-compartment observations are available.

The log-mean analysis aligns with the `log(mu)` quantity used in the
comprehensive cohort report. For a positive patient count `y`, offset `o`, and
NB shape `r`, the patient estimate is `log(y) - o` and its delta-method NB SE
is `sqrt(1 / y + 1 / r)`. It is compared with the mean and SD of matched
healthy `posterior_linpred()` draws using a heteroscedastic normal statistic.
This is the appropriate one-observation analogue of a Welch comparison, but
not a literal `t.test()`: there is no within-patient replicate sample from
which to estimate a classical Welch degrees of freedom. The reported p-values
therefore use the normal reference distribution and remain exploratory.

------------------------------------------------------------------------

# Setup

```{r}
#| label: setup

suppressPackageStartupMessages({
  library(tibble)
  library(tidyr)
  library(dplyr)
  library(readr)
  library(purrr)
  library(ggplot2)
  library(forcats)
  library(stringr)
  library(targets)
  library(brms)
  library(posterior)
})

theme_set(theme_minimal(base_size = 11))

set.seed(20250625)

data_dir <- file.path(params$case_dir, "data")
processed_dir <- file.path(data_dir, "processed")
dir.create(processed_dir, recursive = TRUE, showWarnings = FALSE)

model_path <- file.path(
  params$model_store_root,
  paste0("V1_", params$target_cell_type)
)
model_store <- file.path(model_path, "_targets")

if (!dir.exists(model_store)) {
  stop("Missing targets store: ", model_store)
}
```

# Biomarker and model coverage

```{r}
#| label: biomarker-coverage

times_biomarkers <- readr::read_csv(
  file.path(data_dir, "times_biomarkers.csv"),
  show_col_types = FALSE
)

model_meta <- targets::tar_meta(
  starts_with("estimates_chunk_"),
  store = model_store
) |>
  dplyr::filter(.data$type == "branch") |>
  dplyr::mutate(
    .feature = stringr::str_extract(
      .data$warnings,
      "(?<=Gene:___)ENSG\\d+(?=___)"
    )
  )

biomarker_model_status <- times_biomarkers |>
  dplyr::mutate(
    in_v1_nk_store = .data$.feature %in% model_meta$.feature
  ) |>
  dplyr::select(gene, .feature, role, in_v1_nk_store)

atlas_biomarkers <- biomarker_model_status |>
  dplyr::filter(.data$in_v1_nk_store)

biomarker_model_status
```

`HLA-DRB1` is retained in the direct within-HCC recurrence analysis but cannot
be tested against a personalised posteriorHCA baseline because its `V1_nk`
model branch is absent. The personalised test therefore uses SPON2, VIM, ZFP36,
and ZFP36L2.

# Recover patient metadata and observed counts

```{r}
#| label: patient-data

roi_metadata <- readr::read_csv(
  params$roi_metadata_csv,
  show_col_types = FALSE
) |>
  dplyr::filter(
    .data$included_in_author_nk_analysis,
    .data$compartment %in% c("IF", "TC")
  ) |>
  dplyr::mutate(
    sample_id = paste(.data$patient_id, .data$compartment, sep = "_")
  )

expression_df <- utils::read.csv(
  params$expression_csv,
  check.names = FALSE,
  stringsAsFactors = FALSE
)

missing_genes <- setdiff(atlas_biomarkers$gene, expression_df$gene)
if (length(missing_genes)) {
  stop("Expression matrix is missing: ", paste(missing_genes, collapse = ", "))
}

expression_mat <- as.matrix(
  expression_df[
    match(atlas_biomarkers$gene, expression_df$gene),
    setdiff(names(expression_df), "gene"),
    drop = FALSE
  ]
)
mode(expression_mat) <- "numeric"
rownames(expression_mat) <- atlas_biomarkers$gene

roi_columns_by_sample <- split(
  roi_metadata$author_column,
  roi_metadata$sample_id
)

patient_counts <- do.call(
  cbind,
  purrr::map(names(roi_columns_by_sample), function(sample_id) {
    columns <- roi_columns_by_sample[[sample_id]]
    values <- round(expression_mat[, columns, drop = FALSE])
    n_observed <- rowSums(!is.na(values))
    counts <- rowSums(values, na.rm = TRUE)
    counts[n_observed == 0L] <- NA_real_
    counts
  })
)
colnames(patient_counts) <- names(roi_columns_by_sample)
rownames(patient_counts) <- rownames(expression_mat)

complete_samples <- colnames(patient_counts)[
  colSums(!is.na(patient_counts)) == nrow(patient_counts)
]
patient_counts <- patient_counts[, complete_samples, drop = FALSE]

normalization <- readr::read_csv(
  params$normalization_csv,
  show_col_types = FALSE
) |>
  dplyr::filter(.data$sample_role == "hcc_patient_pseudobulk") |>
  dplyr::select(
    sample_id, multiplier, offset, library_size, reference_sample
  )

patient_metadata <- roi_metadata |>
  dplyr::filter(.data$sample_id %in% complete_samples) |>
  dplyr::group_by(
    .data$sample_id, .data$patient_id, .data$compartment,
    .data$recurrence, .data$age, .data$gender, .data$relapse,
    .data$death, .data$platform
  ) |>
  dplyr::summarise(
    n_rois = dplyr::n_distinct(.data$author_column),
    roi_contexts = paste(sort(unique(.data$cell_context)), collapse = "; "),
    .groups = "drop"
  ) |>
  dplyr::left_join(normalization, by = "sample_id") |>
  dplyr::mutate(
    age_decade = as.character(ceiling(.data$age / 10)),
    sex = stringr::str_to_lower(.data$gender),
    tissue_groups = params$baseline_tissue_group,
    disease_groups_altered = "Normal",
    ethnicity_groups = NA_character_,
    assay_groups_altered = NA_character_,
    dataset_id_altered = "__Jie_HCC_GeoMx_new_dataset__"
  ) |>
  dplyr::arrange(.data$compartment, .data$recurrence, .data$patient_id)

if (anyNA(patient_metadata$offset)) {
  stop("At least one selected patient-compartment lacks an aligned offset.")
}

observed_counts <- as.data.frame(as.table(patient_counts)) |>
  tibble::as_tibble() |>
  dplyr::rename(
    gene = "Var1",
    sample_id = "Var2",
    observed_count = "Freq"
  ) |>
  dplyr::mutate(
    gene = as.character(.data$gene),
    sample_id = as.character(.data$sample_id)
  ) |>
  dplyr::left_join(patient_metadata, by = "sample_id") |>
  dplyr::left_join(
    dplyr::select(atlas_biomarkers, gene, .feature),
    by = "gene"
  ) |>
  dplyr::mutate(
    observed_count = as.numeric(.data$observed_count),
    observed_count_offset_zero = .data$observed_count / exp(.data$offset)
  )

readr::write_csv(patient_metadata, params$patient_metadata_csv)

patient_metadata |>
  dplyr::select(
    sample_id, patient_id, compartment, recurrence, age, age_decade,
    sex, relapse, death, n_rois, roi_contexts, platform, multiplier, offset
  ) |>
  dplyr::mutate(
    dplyr::across(dplyr::where(is.numeric), ~round(.x, 4))
  )
```

Only age and sex are patient-specific covariates in the fitted gene model.
Recurrence, death, compartment, and ROI context are preserved for evaluation
and interpretation but are not supplied to the healthy atlas model.

# Personalised posterior prediction

## Helper functions

```{r}
#| label: prediction-helpers

fit_levels <- function(fit, variable) {
  if (is.null(fit$data) || !variable %in% names(fit$data)) {
    return(character())
  }
  values <- fit$data[[variable]]
  levels_out <- if (is.factor(values)) {
    levels(values)
  } else {
    sort(unique(as.character(values)))
  }
  levels_out[!is.na(levels_out) & nzchar(levels_out)]
}

known_or_missing <- function(value, fit, variable) {
  value <- as.character(value[[1]])
  if (is.na(value) || !nzchar(value)) {
    return(NA_character_)
  }
  if (!value %in% fit_levels(fit, variable)) {
    return(NA_character_)
  }
  value
}

read_fit_for_feature <- function(feature_id) {
  target_name <- model_meta |>
    dplyr::filter(.data$.feature == feature_id) |>
    dplyr::slice(1) |>
    dplyr::pull(.data$name)

  if (!length(target_name)) {
    stop("No target found for feature ", feature_id)
  }

  targets::tar_read_raw(
    target_name[[1]],
    store = model_store
  )$brms_fit[[1]]
}

# Preserve the observed joint distribution of unknown fixed effects among
# healthy liver NK atlas samples. Duplicate rows are deliberate empirical
# weights; unsupported ethnicity-assay combinations are not invented.
healthy_unknown_reference <- function(fit) {
  fit_data <- tibble::as_tibble(fit$data)
  reference <- fit_data |>
    dplyr::filter(
      as.character(.data$disease_groups_altered) == "Normal",
      as.character(.data$tissue_groups) == params$baseline_tissue_group
    ) |>
    dplyr::transmute(
      ethnicity_groups = as.character(.data$ethnicity_groups),
      assay_groups_altered = as.character(.data$assay_groups_altered)
    ) |>
    tidyr::drop_na() |>
    dplyr::filter(
      nzchar(.data$ethnicity_groups),
      nzchar(.data$assay_groups_altered)
    )

  if (!nrow(reference)) {
    stop("No complete healthy liver rows available for marginalisation.")
  }
  reference
}

build_personalised_grid <- function(fit, sample_metadata) {
  unknown_reference <- healthy_unknown_reference(fit)

  sample_rows <- sample_metadata |>
    dplyr::mutate(
      age_decade = purrr::map_chr(
        .data$age_decade,
        ~known_or_missing(.x, fit, "age_decade")
      ),
      sex = purrr::map_chr(
        .data$sex,
        ~known_or_missing(.x, fit, "sex")
      )
    )

  if (anyNA(sample_rows$age_decade) || anyNA(sample_rows$sex)) {
    stop("Observed age decade or sex is absent from the fitted model levels.")
  }

  grid <- purrr::map_dfr(seq_len(nrow(sample_rows)), function(i) {
    sample_row <- sample_rows[i, , drop = FALSE]

    unknown_reference |>
      dplyr::mutate(
        sample_id = sample_row$sample_id,
        patient_id = sample_row$patient_id,
        compartment = sample_row$compartment,
        recurrence = sample_row$recurrence,
        age_decade = sample_row$age_decade,
        sex = sample_row$sex,
        disease_groups_altered = "Normal",
        tissue_groups = params$baseline_tissue_group,
        offset = sample_row$offset,
        dataset_id_altered = "__Jie_HCC_GeoMx_new_dataset__",
        unknown_row = dplyr::row_number()
      )
  })

  list(
    grid = grid,
    n_unknown_rows = nrow(unknown_reference),
    unknown_reference = unknown_reference
  )
}

extract_mixture_draws <- function(
    draw_matrix,
    selected_unknown_rows,
    n_unknown_rows
) {
  n_draws <- nrow(draw_matrix)
  n_samples <- ncol(selected_unknown_rows)

  sample_index <- matrix(
    rep(seq_len(n_samples), each = n_draws),
    nrow = n_draws,
    ncol = n_samples
  )
  column_index <- (
    sample_index - 1L
  ) * n_unknown_rows + selected_unknown_rows

  matrix(
    draw_matrix[cbind(
      rep(seq_len(n_draws), times = n_samples),
      as.vector(column_index)
    )],
    nrow = n_draws,
    ncol = n_samples
  )
}

predict_one_gene <- function(gene, feature_id, sample_metadata, seed) {
  fit <- read_fit_for_feature(feature_id)
  grid_info <- build_personalised_grid(fit, sample_metadata)
  grid <- grid_info$grid

  prediction_newdata <- grid |>
    dplyr::select(
      age_decade, sex, disease_groups_altered, ethnicity_groups,
      assay_groups_altered, dataset_id_altered, tissue_groups, offset
    )

  n_available <- posterior::ndraws(posterior::as_draws_df(fit))
  draw_ids <- seq_len(n_available)

  set.seed(seed)
  predictive_matrix <- brms::posterior_predict(
    fit,
    newdata = prediction_newdata,
    draw_ids = draw_ids,
    re_formula = NULL,
    allow_new_levels = TRUE,
    sample_new_levels = "gaussian"
  )

  set.seed(seed + 1L)
  expected_matrix <- brms::posterior_epred(
    fit,
    newdata = prediction_newdata,
    draw_ids = draw_ids,
    re_formula = NULL,
    allow_new_levels = TRUE,
    sample_new_levels = "gaussian"
  )

  # `transform = FALSE` returns the log-NB count mean, including the offset.
  set.seed(seed + 2L)
  linear_predictor_matrix <- brms::posterior_linpred(
    fit,
    newdata = prediction_newdata,
    draw_ids = draw_ids,
    re_formula = NULL,
    allow_new_levels = TRUE,
    sample_new_levels = "gaussian",
    transform = FALSE
  )

  # NB shape is needed to propagate one-count uncertainty on the latent scale.
  set.seed(seed + 3L)
  shape_matrix <- brms::posterior_linpred(
    fit,
    newdata = prediction_newdata,
    draw_ids = draw_ids,
    re_formula = NULL,
    allow_new_levels = TRUE,
    sample_new_levels = "gaussian",
    dpar = "shape",
    transform = TRUE
  )

  set.seed(seed + 4L)
  selected_unknown_rows <- matrix(
    sample.int(
      grid_info$n_unknown_rows,
      n_available * nrow(sample_metadata),
      replace = TRUE
    ),
    nrow = n_available,
    ncol = nrow(sample_metadata)
  )

  predictive_draws <- extract_mixture_draws(
    predictive_matrix,
    selected_unknown_rows,
    grid_info$n_unknown_rows
  )
  expected_draws <- extract_mixture_draws(
    expected_matrix,
    selected_unknown_rows,
    grid_info$n_unknown_rows
  )
  linear_predictor_draws <- extract_mixture_draws(
    linear_predictor_matrix,
    selected_unknown_rows,
    grid_info$n_unknown_rows
  )
  shape_draws <- extract_mixture_draws(
    shape_matrix,
    selected_unknown_rows,
    grid_info$n_unknown_rows
  )

  colnames(predictive_draws) <- sample_metadata$sample_id
  colnames(expected_draws) <- sample_metadata$sample_id
  colnames(linear_predictor_draws) <- sample_metadata$sample_id
  colnames(shape_draws) <- sample_metadata$sample_id

  purrr::map_dfr(seq_len(nrow(sample_metadata)), function(i) {
    tibble::tibble(
      gene = gene,
      .feature = feature_id,
      sample_id = sample_metadata$sample_id[[i]],
      draw = seq_len(n_available),
      predicted_count = predictive_draws[, i],
      expected_count = expected_draws[, i],
      healthy_linpred_log_mu = linear_predictor_draws[, i] -
        sample_metadata$offset[[i]],
      healthy_nb_shape = shape_draws[, i],
      unknown_reference_row = selected_unknown_rows[, i],
      n_unknown_reference_rows = grid_info$n_unknown_rows
    )
  })
}
```

The same new dataset tag is supplied for every patient row in a gene-level
prediction call. This preserves the fact that all samples came from one GeoMx
dataset while drawing that unseen random intercept from the fitted Gaussian
population. For each posterior draw and patient, one healthy-liver
ethnicity-assay row is sampled. This is a finite-mixture predictive draw; taking
the row mean would incorrectly average away uncertainty for a single patient.

## Generate personalised draws

```{r}
#| label: generate-personalised-draws

predictive_draws <- purrr::pmap_dfr(
  list(
    atlas_biomarkers$gene,
    atlas_biomarkers$.feature,
    20250625L + seq_len(nrow(atlas_biomarkers))
  ),
  function(gene, feature_id, seed) {
    predict_one_gene(
      gene = gene,
      feature_id = feature_id,
      sample_metadata = patient_metadata,
      seed = seed
    )
  }
)

saveRDS(predictive_draws, params$predictive_draws_rds)

predictive_draws |>
  dplyr::count(.data$gene, .data$sample_id, name = "n_draws") |>
  dplyr::arrange(.data$gene, .data$sample_id)
```

# `posterior_predict()`: count-scale test results

```{r}
#| label: predictive-tests

predictive_tests <- predictive_draws |>
  dplyr::left_join(
    observed_counts |>
      dplyr::select(
        gene, .feature, sample_id, patient_id, compartment, recurrence,
        age, age_decade, sex, relapse, death, n_rois, roi_contexts,
        platform, multiplier, offset, library_size, observed_count,
        observed_count_offset_zero
      ),
    by = c("gene", ".feature", "sample_id")
  ) |>
  dplyr::group_by(
    .data$gene, .data$.feature, .data$sample_id, .data$patient_id,
    .data$compartment, .data$recurrence, .data$age, .data$age_decade,
    .data$sex, .data$relapse, .data$death, .data$n_rois,
    .data$roi_contexts, .data$platform, .data$multiplier, .data$offset,
    .data$library_size, .data$observed_count,
    .data$observed_count_offset_zero
  ) |>
  dplyr::summarise(
    n_draws = dplyr::n(),
    n_unknown_reference_rows = dplyr::first(
      .data$n_unknown_reference_rows
    ),
    healthy_predictive_median = stats::median(.data$predicted_count),
    healthy_predictive_lo = stats::quantile(.data$predicted_count, 0.025),
    healthy_predictive_hi = stats::quantile(.data$predicted_count, 0.975),
    healthy_epred_median = stats::median(.data$expected_count),
    healthy_epred_lo = stats::quantile(.data$expected_count, 0.025),
    healthy_epred_hi = stats::quantile(.data$expected_count, 0.975),
    predictive_p_lower = (
      1 + sum(.data$predicted_count <= .data$observed_count)
    ) / (dplyr::n() + 1),
    predictive_p_upper = (
      1 + sum(.data$predicted_count >= .data$observed_count)
    ) / (dplyr::n() + 1),
    predictive_percentile = (
      0.5 +
        sum(.data$predicted_count < .data$observed_count) +
        0.5 * sum(.data$predicted_count == .data$observed_count)
    ) / (dplyr::n() + 1),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    predictive_tail_p = pmin(
      1,
      2 * pmin(.data$predictive_p_lower, .data$predictive_p_upper)
    ),
    predictive_z = stats::qnorm(
      pmin(
        1 - 0.5 / (.data$n_draws + 1),
        pmax(
          0.5 / (.data$n_draws + 1),
          .data$predictive_percentile
        )
      )
    ),
    absolute_predictive_z = abs(.data$predictive_z),
    log2fc_vs_healthy_epred = log2(
      (.data$observed_count + 0.5) /
        (.data$healthy_epred_median + 0.5)
    ),
    direction = dplyr::case_when(
      .data$predictive_p_lower <= 0.025 ~ "below_personalised_healthy",
      .data$predictive_p_upper <= 0.025 ~ "above_personalised_healthy",
      TRUE ~ "consistent_with_personalised_healthy"
    ),
    q_value_global = stats::p.adjust(
      .data$predictive_tail_p,
      method = "BH"
    )
  ) |>
  dplyr::group_by(.data$sample_id) |>
  dplyr::mutate(
    q_value_within_sample = stats::p.adjust(
      .data$predictive_tail_p,
      method = "BH"
    )
  ) |>
  dplyr::ungroup() |>
  dplyr::arrange(
    .data$compartment, .data$recurrence, .data$patient_id, .data$gene
  )

readr::write_csv(predictive_tests, params$test_results_csv)

predictive_tests |>
  dplyr::select(
    gene, sample_id, recurrence, age_decade, sex, offset, observed_count,
    healthy_predictive_median, healthy_predictive_lo,
    healthy_predictive_hi, healthy_epred_median,
    log2fc_vs_healthy_epred, predictive_percentile, predictive_tail_p,
    q_value_global, direction
  ) |>
  dplyr::mutate(
    dplyr::across(dplyr::where(is.numeric), ~round(.x, 4))
  )
```

The two-sided value is a posterior predictive tail probability with a
finite-draw correction, not a Welch or edgeR cohort test. A result outside the
central predictive mass means that the observed count is difficult to generate
from the personalised healthy counterfactual after propagating all modeled
sources of uncertainty.

# `posterior_linpred()`: patient log-mean test results

```{r}
#| label: linpred-welch-tests

if (any(predictive_tests$observed_count <= 0)) {
  stop("The NB delta-method log-mean SE requires positive observed counts.")
}

# The healthy distribution is the matched posterior_linpred distribution at
# offset zero. A single patient has one NB count, so its log(mu) SE is obtained
# by the delta method: Var(log(y)) approximately 1 / y + 1 / shape.
linpred_tests <- predictive_draws |>
  dplyr::select(
    gene, .feature, sample_id, draw, healthy_linpred_log_mu,
    healthy_nb_shape, n_unknown_reference_rows
  ) |>
  dplyr::left_join(
    predictive_tests |>
      dplyr::select(
        gene, .feature, sample_id, patient_id, compartment, recurrence,
        age, age_decade, sex, relapse, death, n_rois, roi_contexts, platform,
        multiplier, offset, library_size, observed_count
      ),
    by = c("gene", ".feature", "sample_id")
  ) |>
  dplyr::group_by(
    .data$gene, .data$.feature, .data$sample_id, .data$patient_id,
    .data$compartment, .data$recurrence, .data$age, .data$age_decade,
    .data$sex, .data$relapse, .data$death, .data$n_rois,
    .data$roi_contexts, .data$platform, .data$multiplier, .data$offset,
    .data$library_size, .data$observed_count
  ) |>
  dplyr::summarise(
    n_draws = dplyr::n(),
    n_unknown_reference_rows = dplyr::first(
      .data$n_unknown_reference_rows
    ),
    healthy_log_mu_mean = mean(.data$healthy_linpred_log_mu),
    healthy_log_mu_sd = stats::sd(.data$healthy_linpred_log_mu),
    healthy_log_mu_lo = stats::quantile(.data$healthy_linpred_log_mu, 0.025),
    healthy_log_mu_hi = stats::quantile(.data$healthy_linpred_log_mu, 0.975),
    healthy_nb_shape_median = stats::median(.data$healthy_nb_shape),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    patient_log_mu = log(.data$observed_count) - .data$offset,
    patient_log_mu_se = sqrt(
      1 / .data$observed_count + 1 / .data$healthy_nb_shape_median
    ),
    delta_log_mu_vs_healthy = (
      .data$patient_log_mu - .data$healthy_log_mu_mean
    ),
    delta_log_mu_se = sqrt(
      .data$patient_log_mu_se^2 + .data$healthy_log_mu_sd^2
    ),
    welch_style_z = (
      .data$delta_log_mu_vs_healthy / .data$delta_log_mu_se
    ),
    welch_style_p = 2 * stats::pnorm(-abs(.data$welch_style_z)),
    direction = dplyr::case_when(
      .data$welch_style_z <= -1.96 ~ "below_personalised_healthy",
      .data$welch_style_z >= 1.96 ~ "above_personalised_healthy",
      TRUE ~ "not_significant"
    ),
    q_value_global = stats::p.adjust(.data$welch_style_p, method = "BH")
  ) |>
  dplyr::group_by(.data$sample_id) |>
  dplyr::mutate(
    q_value_within_sample = stats::p.adjust(
      .data$welch_style_p,
      method = "BH"
    )
  ) |>
  dplyr::ungroup() |>
  dplyr::arrange(
    .data$compartment, .data$recurrence, .data$patient_id, .data$gene
  )

linpred_patient_scores <- linpred_tests |>
  dplyr::group_by(
    .data$sample_id, .data$patient_id, .data$compartment,
    .data$recurrence, .data$age, .data$sex, .data$death
  ) |>
  dplyr::summarise(
    n_genes = dplyr::n(),
    mean_absolute_welch_z = mean(abs(.data$welch_style_z)),
    mean_signed_welch_z = mean(.data$welch_style_z),
    n_p_lt_0_05 = sum(.data$welch_style_p < 0.05),
    n_global_q_lt_0_05 = sum(.data$q_value_global < 0.05),
    .groups = "drop"
  ) |>
  dplyr::arrange(.data$compartment, .data$recurrence, .data$patient_id)

readr::write_csv(linpred_tests, params$linpred_test_results_csv)
readr::write_csv(linpred_patient_scores, params$linpred_patient_scores_csv)

linpred_tests |>
  dplyr::select(
    gene, sample_id, recurrence, observed_count, patient_log_mu,
    patient_log_mu_se, healthy_log_mu_mean, healthy_log_mu_sd,
    delta_log_mu_vs_healthy, delta_log_mu_se, welch_style_z,
    welch_style_p, q_value_global, direction
  ) |>
  dplyr::mutate(
    dplyr::across(dplyr::where(is.numeric), ~round(.x, 4))
  )
```

This is a patient-level heteroscedastic log-mean test. The patient SE comes
from the NB sampling model for one positive count; the healthy SD is the
spread of the patient's matched `posterior_linpred()` distribution, including
model, covariate-marginalisation, and new-dataset uncertainty. It is
Welch-style because the two sources have unequal variance, but it is not a
literal Welch t-test with independent patient replicates. The normal-reference
p-values and BH values are therefore exploratory.

# Outcome-free patient deviation score

```{r}
#| label: patient-scores

patient_scores <- predictive_tests |>
  dplyr::group_by(
    .data$sample_id, .data$patient_id, .data$compartment,
    .data$recurrence, .data$age, .data$sex, .data$death
  ) |>
  dplyr::summarise(
    n_genes = dplyr::n(),
    mean_absolute_predictive_z = mean(.data$absolute_predictive_z),
    mean_signed_predictive_z = mean(.data$predictive_z),
    mean_abs_log2fc_vs_epred = mean(abs(.data$log2fc_vs_healthy_epred)),
    mean_tail_surprisal = mean(-log10(.data$predictive_tail_p)),
    n_outside_95 = sum(
      .data$direction != "consistent_with_personalised_healthy"
    ),
    n_global_q_lt_0_05 = sum(.data$q_value_global < 0.05),
    .groups = "drop"
  ) |>
  dplyr::arrange(.data$compartment, .data$recurrence, .data$patient_id)

readr::write_csv(patient_scores, params$patient_scores_csv)

patient_scores |>
  dplyr::mutate(
    dplyr::across(dplyr::where(is.numeric), ~round(.x, 4))
  )
```

The primary patient score is the mean absolute predictive z-score across the
four genes. It is outcome-free: recurrence labels are not used to construct the
personalised null, gene-level tail probabilities, or score.

```{r}
#| label: performance

auc_rank <- function(score, recurrence) {
  recurrence <- as.character(recurrence)
  positives <- score[recurrence == "REC"]
  negatives <- score[recurrence == "non-REC"]
  if (!length(positives) || !length(negatives)) {
    return(NA_real_)
  }
  mean(outer(positives, negatives, ">")) +
    0.5 * mean(outer(positives, negatives, "=="))
}

combined_patient_scores <- patient_scores |>
  dplyr::group_by(.data$patient_id, .data$recurrence) |>
  dplyr::summarise(
    mean_absolute_predictive_z = mean(.data$mean_absolute_predictive_z),
    n_compartments = dplyr::n(),
    .groups = "drop"
  )

combined_linpred_patient_scores <- linpred_patient_scores |>
  dplyr::group_by(.data$patient_id, .data$recurrence) |>
  dplyr::summarise(
    mean_absolute_welch_z = mean(
      .data$mean_absolute_welch_z
    ),
    n_compartments = dplyr::n(),
    .groups = "drop"
  )

predictive_performance <- dplyr::bind_rows(
  patient_scores |>
    dplyr::group_by(.data$compartment) |>
    dplyr::summarise(
      n_patients = dplyr::n_distinct(.data$patient_id),
      n_rec = sum(.data$recurrence == "REC"),
      auc = auc_rank(
        .data$mean_absolute_predictive_z,
        .data$recurrence
      ),
      .groups = "drop"
    ) |>
    dplyr::transmute(
      scope = as.character(.data$compartment),
      n_patients, n_rec, auc
    ),
  combined_patient_scores |>
    dplyr::summarise(
      scope = "combined_available_IF_TC",
      n_patients = dplyr::n(),
      n_rec = sum(.data$recurrence == "REC"),
      auc = auc_rank(
        .data$mean_absolute_predictive_z,
        .data$recurrence
      )
  )
)

linpred_performance <- dplyr::bind_rows(
  linpred_patient_scores |>
    dplyr::group_by(.data$compartment) |>
    dplyr::summarise(
      n_patients = dplyr::n_distinct(.data$patient_id),
      n_rec = sum(.data$recurrence == "REC"),
      auc = auc_rank(
        .data$mean_absolute_welch_z,
        .data$recurrence
      ),
      .groups = "drop"
    ) |>
    dplyr::transmute(
      scope = as.character(.data$compartment),
      n_patients, n_rec, auc
    ),
  combined_linpred_patient_scores |>
    dplyr::summarise(
      scope = "combined_available_IF_TC",
      n_patients = dplyr::n(),
      n_rec = sum(.data$recurrence == "REC"),
      auc = auc_rank(
        .data$mean_absolute_welch_z,
        .data$recurrence
      )
    )
)

performance <- dplyr::bind_rows(
  predictive_performance |>
    dplyr::mutate(analysis = "posterior_predict"),
  linpred_performance |>
    dplyr::mutate(analysis = "posterior_linpred_welch_style")
) |>
  dplyr::select(.data$analysis, dplyr::everything())

readr::write_csv(performance, params$performance_csv)
performance
```

# Visual summaries

## `posterior_predict()`: observed-count predictive checks

```{r}
#| label: plot-predictive-intervals
#| fig-width: 9
#| fig-height: 7
#| fig-align: center

plot_tests <- predictive_tests |>
  dplyr::mutate(
    sample_label = paste0(.data$sample_id, " (", .data$recurrence, ")"),
    sample_label = factor(
      .data$sample_label,
      levels = rev(unique(.data$sample_label))
    )
  )

ggplot2::ggplot(
  plot_tests,
  ggplot2::aes(
    y = .data$sample_label,
    colour = .data$recurrence
  )
) +
  ggplot2::geom_errorbar(
    ggplot2::aes(
      xmin = log1p(.data$healthy_predictive_lo),
      xmax = log1p(.data$healthy_predictive_hi)
    ),
    orientation = "y",
    width = 0.18,
    linewidth = 0.7,
    colour = "grey55"
  ) +
  ggplot2::geom_point(
    ggplot2::aes(x = log1p(.data$healthy_predictive_median)),
    shape = 3,
    size = 2.5,
    stroke = 0.8
  ) +
  ggplot2::geom_point(
    ggplot2::aes(x = log1p(.data$observed_count)),
    size = 2.8
  ) +
  ggplot2::facet_wrap(~ gene, scales = "free_x", ncol = 2) +
  ggplot2::scale_colour_manual(
    values = c(`non-REC` = "#2878B5", REC = "#C23B4A")
  ) +
  ggplot2::labs(
    title = "Observed counts against personalised healthy predictions",
    subtitle = "Points = observed; plus signs = predictive median; bars = central 95% posterior predictive interval",
    x = "log(1 + count), evaluated at each patient's TMMwsp offset",
    y = NULL,
    colour = "Outcome"
  ) +
  ggplot2::theme(legend.position = "bottom")
```

```{r}
#| label: plot-predictive-z
#| fig-width: 8
#| fig-height: 5
#| fig-align: center

ggplot2::ggplot(
  plot_tests,
  ggplot2::aes(
    x = .data$sample_label,
    y = forcats::fct_rev(factor(.data$gene)),
    fill = .data$predictive_z
  )
) +
  ggplot2::geom_tile(colour = "white", linewidth = 0.5) +
  ggplot2::geom_text(
    ggplot2::aes(label = sprintf("p=%.3f", .data$predictive_tail_p)),
    size = 3
  ) +
  ggplot2::scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    name = "Predictive z"
  ) +
  ggplot2::labs(
    title = "Direction and strength of personalised healthy deviation",
    subtitle = "Negative values are below healthy; positive values are above healthy",
    x = NULL,
    y = NULL
  ) +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    axis.text.x = ggplot2::element_text(angle = 25, hjust = 1)
  )
```

## `posterior_linpred()`: patient and matched healthy density overlap

```{r}
#| label: plot-linpred-density-overlap
#| fig-width: 12
#| fig-height: 10
#| fig-align: center

set.seed(20250630)
patient_density <- linpred_tests |>
  dplyr::select(
    gene, sample_id, patient_id, compartment, recurrence,
    patient_log_mu, patient_log_mu_se
  ) |>
  tidyr::uncount(2000L, .id = "density_draw") |>
  dplyr::mutate(
    log_mu = stats::rnorm(
      dplyr::n(),
      mean = .data$patient_log_mu,
      sd = .data$patient_log_mu_se
    ),
    distribution = .data$recurrence
  )

healthy_density <- predictive_draws |>
  dplyr::select(gene, sample_id, healthy_linpred_log_mu) |>
  dplyr::left_join(
    linpred_tests |>
      dplyr::select(gene, sample_id, patient_id, compartment, recurrence),
    by = c("gene", "sample_id")
  ) |>
  dplyr::transmute(
    gene, sample_id, patient_id, compartment, recurrence,
    log_mu = .data$healthy_linpred_log_mu,
    distribution = "Healthy baseline"
  )

density_means <- dplyr::bind_rows(
  linpred_tests |>
    dplyr::transmute(
      gene, sample_id, patient_id, compartment, recurrence,
      mean_log_mu = .data$patient_log_mu,
      distribution = .data$recurrence
    ),
  linpred_tests |>
    dplyr::transmute(
      gene, sample_id, patient_id, compartment, recurrence,
      mean_log_mu = .data$healthy_log_mu_mean,
      distribution = "Healthy baseline"
    )
) |>
  dplyr::mutate(
    sample_label = paste0(.data$sample_id, " (", .data$recurrence, ")")
  )

density_data <- dplyr::bind_rows(patient_density, healthy_density) |>
  dplyr::mutate(
    sample_label = paste0(.data$sample_id, " (", .data$recurrence, ")")
  )

ggplot2::ggplot(
  density_data,
  ggplot2::aes(
    x = .data$log_mu,
    colour = .data$distribution,
    fill = .data$distribution
  )
) +
  ggplot2::geom_density(alpha = 0.12, linewidth = 0.7, adjust = 1.15) +
  ggplot2::geom_point(
    data = density_means,
    ggplot2::aes(
      x = .data$mean_log_mu,
      y = 0,
      colour = .data$distribution,
      fill = .data$distribution
    ),
    inherit.aes = FALSE,
    shape = 21,
    size = 2.5,
    stroke = 0.7
  ) +
  ggplot2::facet_grid(sample_label ~ gene) +
  ggplot2::scale_colour_manual(
    values = c(
      "Healthy baseline" = "grey35",
      "non-REC" = "#2878B5",
      REC = "#C23B4A"
    )
  ) +
  ggplot2::scale_fill_manual(
    values = c(
      "Healthy baseline" = "grey70",
      "non-REC" = "#2878B5",
      REC = "#C23B4A"
    )
  ) +
  ggplot2::labs(
    title = "Patient log-mean estimates against matched healthy NK baselines",
    subtitle = "Lines = patient and that patient's healthy reference; dots on the x-axis = means",
    x = "Offset-zero log(mu)",
    y = "Density",
    colour = "Distribution",
    fill = "Distribution"
  ) +
  ggplot2::theme(legend.position = "bottom")
```

Each panel contains exactly one HCC patient-compartment and the healthy
`posterior_linpred()` distribution matched to that same patient. No baseline
draws or patient estimates are pooled across individuals. Blue and red curves
denote non-REC and REC only as an outcome annotation; coloured curves are
normal approximations centred on NB-derived patient log means with their
delta-method SEs.

## Outcome-free predictive patient scores

```{r}
#| label: plot-patient-scores
#| fig-width: 8
#| fig-height: 4.5
#| fig-align: center

ggplot2::ggplot(
  patient_scores,
  ggplot2::aes(
    x = .data$compartment,
    y = .data$mean_absolute_predictive_z,
    colour = .data$recurrence,
    label = .data$patient_id
  )
) +
  ggplot2::geom_point(size = 3) +
  ggplot2::geom_text(
    nudge_x = 0.08,
    hjust = 0,
    size = 3.2,
    show.legend = FALSE
  ) +
  ggplot2::scale_colour_manual(
    values = c(`non-REC` = "#2878B5", REC = "#C23B4A")
  ) +
  ggplot2::labs(
    title = "Outcome-free personalised deviation scores",
    subtitle = "Mean absolute posterior predictive z-score across four atlas-covered TIMES genes",
    x = NULL,
    y = "Mean absolute predictive z",
    colour = "Outcome"
  ) +
  ggplot2::theme(legend.position = "bottom")
```

# Interpretation

```{r}
#| label: interpretation-values

n_tests <- nrow(predictive_tests)
n_outside <- sum(
  predictive_tests$direction != "consistent_with_personalised_healthy"
)
n_global_q <- sum(predictive_tests$q_value_global < 0.05)
n_below <- sum(
  predictive_tests$direction == "below_personalised_healthy"
)
n_above <- sum(
  predictive_tests$direction == "above_personalised_healthy"
)
n_below_epred <- sum(predictive_tests$log2fc_vs_healthy_epred < 0)

n_linpred_significant <- sum(linpred_tests$welch_style_p < 0.05)
n_linpred_global_q <- sum(linpred_tests$q_value_global < 0.05)
n_linpred_below <- sum(
  linpred_tests$direction == "below_personalised_healthy"
)
n_linpred_above <- sum(
  linpred_tests$direction == "above_personalised_healthy"
)
n_linpred_below_rec <- sum(
  linpred_tests$direction == "below_personalised_healthy" &
    linpred_tests$recurrence == "REC"
)
n_linpred_below_nonrec <- sum(
  linpred_tests$direction == "below_personalised_healthy" &
    linpred_tests$recurrence == "non-REC"
)

rec_scores <- patient_scores |>
  dplyr::filter(.data$recurrence == "REC") |>
  dplyr::pull(.data$mean_absolute_predictive_z)
nonrec_scores <- patient_scores |>
  dplyr::filter(.data$recurrence == "non-REC") |>
  dplyr::pull(.data$mean_absolute_predictive_z)

linpred_rec_scores <- linpred_patient_scores |>
  dplyr::filter(.data$recurrence == "REC") |>
  dplyr::pull(.data$mean_absolute_welch_z)
linpred_nonrec_scores <- linpred_patient_scores |>
  dplyr::filter(.data$recurrence == "non-REC") |>
  dplyr::pull(.data$mean_absolute_welch_z)

outside_predictive_interval <- predictive_tests |>
  dplyr::filter(
    .data$direction != "consistent_with_personalised_healthy"
  ) |>
  dplyr::select(
    gene, sample_id, recurrence, observed_count,
    healthy_predictive_lo, healthy_predictive_hi,
    predictive_tail_p, q_value_global, direction
  )
```

Across `r n_tests` gene-by-patient-compartment tests, `r n_outside` observations
fall outside the central 95% personalised posterior predictive distribution:
`r n_below` below and `r n_above` above healthy expectation. After global
Benjamini-Hochberg adjustment, `r n_global_q` remain below 0.05.

```{r}
#| label: outside-predictive-interval

outside_predictive_interval |>
  dplyr::mutate(
    dplyr::across(dplyr::where(is.numeric), ~round(.x, 4))
  )
```

All `r n_below_epred` observed counts are below their personalised healthy
`posterior_epred` median. The stronger result is comparative rather than
gene-wise significance: the REC patient A29 has mean absolute predictive
z-scores of `r round(patient_scores$mean_absolute_predictive_z[patient_scores$sample_id == "A29_IF"], 3)`
in IF and
`r round(patient_scores$mean_absolute_predictive_z[patient_scores$sample_id == "A29_TC"], 3)`
in TC. The available non-REC scores range from
`r round(min(nonrec_scores), 3)` to `r round(max(nonrec_scores), 3)`.

The `posterior_linpred` log-mean test reports `r n_linpred_significant` of
`r n_tests` comparisons with an unadjusted normal-reference p-value below
0.05 (`r n_linpred_global_q` after global BH adjustment):
`r n_linpred_below` below and `r n_linpred_above` above the personalised
healthy mean. `r n_linpred_below_rec` below-healthy calls are in the REC
patient, versus `r n_linpred_below_nonrec` in non-REC patients. A29's mean
absolute Welch-style z-score is
`r round(linpred_patient_scores$mean_absolute_welch_z[linpred_patient_scores$sample_id == "A29_IF"], 3)`
in IF and
`r round(linpred_patient_scores$mean_absolute_welch_z[linpred_patient_scores$sample_id == "A29_TC"], 3)`
in TC, versus `r round(min(linpred_nonrec_scores), 3)` to
`r round(max(linpred_nonrec_scores), 3)` in non-REC samples. As in the
predictive analysis, the outcome-free patient ordering is more stable than any
single-gene threshold.

The patient score is a standardized predictive-deviation index, not a
recurrence probability. Its recurrence discrimination is descriptive:

```{r}
#| label: performance-display

performance |>
  dplyr::mutate(auc = round(.data$auc, 3))
```

These results test whether posteriorHCA can create a patient-specific healthy
counterfactual, not whether it has equal power to the original spatial TIMES
model. The current evaluation has three patients, one REC patient, four
atlas-covered genes, author-processed GeoMx values, and an assay platform absent
from model training. Age, sex, tissue, offset, disease state, unknown
fixed-effect uncertainty, dataset heterogeneity, and observation noise are
handled explicitly; the remaining modality mismatch cannot be removed without
healthy GeoMx liver NK training data or a fitted cross-platform calibration
term.

The difference from the comprehensive cohort analysis is intentional. That
analysis compares estimated latent log means and can provide sharper evidence
about cohort displacement. Here, `posterior_linpred()` supplies the matched
healthy log-mean distribution, while the patient log-mean SE is derived from
its one-count NB sampling model. The primary personalised analysis instead
asks whether one raw count is unusual for a new healthy observation, so
`posterior_predict()` additionally represents zero-inflation and count
generation. The log-mean test complements that calibrated predictive check;
it must not be presented as a standard replicated-sample Welch t-test.

# Output files

```{r}
#| label: output-files

output_paths <- c(
  patient_metadata_csv = params$patient_metadata_csv,
  test_results_csv = params$test_results_csv,
  patient_scores_csv = params$patient_scores_csv,
  performance_csv = params$performance_csv,
  linpred_test_results_csv = params$linpred_test_results_csv,
  linpred_patient_scores_csv = params$linpred_patient_scores_csv,
  predictive_draws_rds = params$predictive_draws_rds
)

tibble::tibble(
  output = names(output_paths),
  path = unname(output_paths),
  exists = file.exists(output_paths)
)
```
