---
title: "TIMES HCC posteriorHCA proxy - planning report"
subtitle: "Healthy liver NK distance as an outcome-free recurrence score"
date: today
date-format: "DD MMM YYYY"
title-block-banner: true
author:
  - name: Chen Zhan
    orcid: 0000-0002-4794-8339
    email: chen.zhan@adelaide.edu.au
    affiliation:
      - id: sgc
        name: SAiGENCI, The University of Adelaide
format:
  html:
    theme: minty
    df-print: paged
    code-line-numbers: true
    embed-resources: true
    code-fold: true
    code-fold-default: true
    toc: true
    toc-location: left
    toc-title: "Contents"
knitr:
  opts_chunk:
    message: false
    warning: false
editor: source
abstract: |
  Planning document for a posteriorHCA case study of the TIMES hepatocellular
  carcinoma spatial multi-omics paper (Jia *et al.*, *Nature* 2025). The case
  study tests whether a universal healthy liver NK reference can act as an
  outcome-free proxy for recurrence: non-REC HCC should remain closer to healthy
  NK expression, whereas REC HCC should be more atlas-discordant in invasive
  front (IF) and tumour centre (TC). Direct REC/non-REC edgeR contrasts are a
  positive control, and supplied TIMES validation scores are a descriptive
  benchmark. External validation is required before claiming equal power.
params:
  case_dir: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC"
  target_cell_type: "nk"
  model_store_root: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/HPC_posterior/NEW_CELL_TYPE"
  pseudobulk_h5ad: "/home/a1237163/lab/Mangiola_ImmuneAtlas/zenodo/pseudobulk_se.h5ad"
  paper_doi: "10.1038/s41586-025-08668-x"
---

# Introduction

Jia *et al.* (2025) developed the **Tumour Immune Microenvironment Spatial (TIMES)**
score to predict HCC recurrence after resection. The score integrates spatial
expression of five biomarkers — **SPON2, ZFP36L2, ZFP36, VIM, HLA-DRB1** — in
multiplex immunohistochemistry (mIHC) tiles at the **invasive front (IF)** and
**tumour centre (TC)**. The paper emphasises that predictive power arises from
**spatial integration**, not bulk expression alone (TCGA bulk TC showed no significant
REC vs non-REC separation for these genes).

**Primary publication:** Jia, G., *et al.* (2025). Spatial immune scoring system
predicts hepatocellular carcinoma recurrence. *Nature* **640**, 1029–1040.
<https://doi.org/10.1038/s41586-025-08668-x>

**Companion SAVI case study:** `dev/SAVI/SAVI_case_study_universal_baseline.qmd`
and `dev/SAVI/mu_test_report.qmd` provide the normalization and healthy-baseline
machinery. The inferential target differs: SAVI has a disease-versus-healthy
claim, while this study asks whether healthy-reference distance can proxy a
recurrence claim within HCC.

**Scope of this document:**

| In scope | Out of scope |
|---|---|
| Define RQs, hierarchy and analysis plan | TIMES XGBoost score reproduction |
| Catalogue TIMES biomarkers + Ensembl mapping | mIHC tile-level spatial modelling |
| Download registry + `prepare_data.R` pseudobulk pipeline | Full survival/classification performance reproduction |
| Code Ocean processed GeoMx extraction + validation | Controlled-access raw-data analysis until access is granted |
| IF/TC patient-pseudobulk recurrence contrasts | Treating pseudobulk as equivalent to the spatial TIMES score |
| Primary healthy-distance recurrence proxy | Formal equivalence claims from the current small cohort |
| Data access documentation | sccomp composition (unless NK abundance needed later) |

------------------------------------------------------------------------

# Research questions

Patient pseudobulks are split into **IF-REC**, **IF-non-REC**, **TC-REC** and
**TC-non-REC**.

## RQ1 — Primary posteriorHCA proxy

> Is REC HCC farther from the comprehensive healthy liver NK reference than
> non-REC HCC in IF and TC?

**Score:** For each atlas-covered gene,
`D = abs(cohort log(mu) - healthy mean log(mu)) / healthy grand-mean SD`.
Average `D` across genes for compartment and patient scores. Recurrence labels
are not used to construct the score.

**Success pattern:** REC has larger `D` than non-REC across genes and
compartments; the combined posterior probability
`P(D_REC > D_non-REC)` exceeds 0.95; patient ordering survives
leave-one-gene-out analysis.

## RQ2 — Positive control against paper direction

> Does the available GeoMx subset retain the paper's `non-REC > REC` expression
> direction within IF and TC?

**Test:** Patient-level edgeR QL contrasts, `non-REC - REC`, using TMMwsp
offsets. This HCC-only fit retains all five TIMES genes, including `HLA-DRB1`.

**Ground truth:** Paper CodeOcean DE results specify `non-REC > REC` in both IF
and TC for all five biomarkers.

## RQ3 — Discrimination and benchmark

> Does the outcome-free posteriorHCA proxy rank REC above non-REC patients, and
> how does its descriptive AUC compare with supplied TIMES validation scores?

**Evaluation:** IF, TC and combined patient AUC; exact rank p-value;
leave-one-gene-out robustness; descriptive comparison to the CodeOcean
TIMES-score validation subset. Different sample sizes and modalities preclude a
formal equivalence or non-inferiority claim.

## RQ4 — Spatial-omics platform robustness

> Do **Visium TC spot pseudobulk** (all cell types in TC) and **GeoMx NK-enriched
> TC ROIs** yield **concordant** comparisons to the healthy NK baseline for the
> same patients where both platforms exist?

**Note:** Atlas reference is **healthy liver NK cells**; Visium TC pseudobulk is
**mixed-cell**. Interpretation treats Visium as "TC tissue programme" and GeoMx as
"NK-enriched ROI". This differs from SAVI, where the study material was PBMC and
the appropriate reference tissue was blood.

------------------------------------------------------------------------

# Study design overview

```{mermaid}
flowchart TB
  subgraph paper [Jia et al. 2025 TIMES]
    V[10x Visium 17 pts] --> TC[TC compartment spots]
    G[GeoMx DSP 8 tissues] --> NK[NK-enriched ROIs in TC]
    TC --> PB1[PatientBOARD TC pseudobulk]
    NK --> PB2[patient x TC pseudobulk]
  end
  subgraph atlas [posteriorHCA Immune Atlas]
    NKref[V1_nk brms models] --> BL[Universal healthy baseline]
  end
  PB1 --> PROXY[Primary: healthy-distance proxy]
  PB2 --> PROXY
  BL --> PROXY
  PROXY --> EVAL[Evaluate REC farther than non-REC]
  PB1 --> REC[Positive control: direct REC contrast]
  PB2 --> REC
  REC --> GT[Compare with paper direction]
```

------------------------------------------------------------------------

# TIMES biomarkers

## Selection path (from paper)

1. **GeoMx DSP** whole-transcriptome atlas on **CD3⁻CD57⁺ NK-enriched** ROIs across
   **AS, IF, TC** in 8 discovery HCC tissues → **241 DEGs** (REC vs non-REC).
2. Filter to **23 genes** also NK-enriched in independent scRNA-seq (Sun *et al.*,
   *Cell* 2021, ref. 28).
3. GLM on **IF + TC** counts (AS excluded as normal tissue) → **5 top predictors**:
   SPON2, ZFP36L2, ZFP36, VIM, HLA-DRB1 (Supplementary Table 12).
4. **XGBoost** on mIHC protein quantification in **61 patients** (IF + TC tiles) →
   TIMES score (AUC > 0.87 per biomarker on validation split).

## Biomarker table

```{r}
#| label: setup-lib
.libPaths(c(
  "/scratchdata1/users/a1237163/local/RLibs",
  "/home/a1237163/miniconda3/envs/R_env/lib/R/library"
))
```

```{r}
#| label: biomarkers
library(tibble)
library(dplyr)
library(readr)
library(AnnotationDbi)
library(org.Hs.eg.db)

data_dir <- file.path(params$case_dir, "data")
biomarkers <- read_csv(file.path(data_dir, "times_biomarkers.csv"), show_col_types = FALSE) |>
  mutate(
    .feature = AnnotationDbi::mapIds(
      org.Hs.eg.db::org.Hs.eg.db,
      keys = gene,
      column = "ENSEMBL",
      keytype = "SYMBOL",
      multiVals = "first"
    ),
    hla_note = if_else(gene == "HLA-DRB1", "MHC II; may map to multiple Ensembl paralogs — verify in count matrix", NA_character_)
  )

biomarkers
```

```{r}
#| label: biomarker-summary
tibble(
  n_times_biomarkers = nrow(biomarkers),
  mapped_to_ensembl = sum(!is.na(biomarkers$.feature)),
  primary_gene = "SPON2",
  atlas_cell_type = params$target_cell_type
)
```

### Gene-level notes

| Gene | Symbol | Role in TIMES / NK biology |
|------|--------|----------------------------|
| SPON2 | Spondin-2 | Top predictor; NK migration to IF; IFN-γ; validated in *Spon2* NK-KO mice |
| ZFP36L2 | ZFP36L2 | RNA-binding; stress-response programme in NK DEG set |
| ZFP36 | TTP | mRNA destabilisation; inflammatory response |
| VIM | Vimentin | Mesenchymal/activation marker in spatial mIHC |
| HLA-DRB1 | HLA-DRB1 | Antigen presentation; NK-enriched spatial ROIs |

------------------------------------------------------------------------

# Tissue compartments and pseudobulk unit

Paper defines three compartments on FFPE sections:

| Code | Name | Cancer association | Use in this case study |
|------|------|--------------------|------------------------|
| TC | Tumour centre | Core tumour | Co-primary recurrence compartment |
| IF | Invasive front | Tumour–stroma interface; central to TIMES biology | Co-primary recurrence compartment |
| AS | Adjacent stroma | Peri-tumour tissue | Not used in the primary recurrence tests |

**Primary pseudobulk unit:** one column per **`patient_id × compartment`** for
IF and TC, summing raw counts or rounded processed ROI values:

- **Visium:** all spots annotated as IF or TC within each patient section.
- **GeoMx:** all NK-enriched ROIs within IF or TC.

For the author Code Ocean processed GeoMx data, we keep the **whole ROI matrix**
with `compartment = AS / IF / TC` and `cell_context` metadata. IF and TC are
analysed separately for recurrence. Healthy liver NK comparisons are secondary
context for the same four recurrence-by-compartment cohorts.

Metadata retained: `patient_id`, `recurrence` (REC / non-REC), `compartment`,
`platform`, `cell_context`, `cohort = discovery`.

------------------------------------------------------------------------

# Data sources and download status

```{r}
#| label: data-registry
registry <- readRDS(file.path(data_dir, "data_registry.rds"))
manifest <- readRDS(file.path(data_dir, "download_manifest.rds"))

registry
```

```{r}
#| label: download-manifest
manifest
```

### Access notes (verified Jun 2025)

| Deposit | Content | Access |
|---------|---------|--------|
| **OMIX005738** | GeoMx WTA/CTA xlsx + ROI labels | Controlled; files **Unavailable** pending Chinese human genetic resource registration |
| **HRA006579** | 10x Visium raw/fastq | Controlled-access GSA-Human; request via DAC |
| **CNP0000650** | In-house scRNA-seq | CNGBdb project; controlled |
| **OMIX005736** | LC-MS/MS proteomics | Controlled (proteomics validation; optional) |
| **Code Ocean** | Author capsule `3972922/tree/v1`; processed analysis inputs | Exported locally under `dev/Jie_HCC/codeocean/` |
| **Nature supplementary** | Tables 1–38, spot/compartment annotations | Manual (Springer static host blocked from HPC DNS at planning time) |

**Action required for raw data:** Request CNCB DAC access (submitter: Cheng Sun,
`charless@ustc.edu.cn`; BioProject **PRJCA023**). Place approved files under
`data/raw/` subfolders listed in `prepare_data.R`, then re-run the script.

**Code Ocean status:** The refreshed author capsule now includes processed RData
inputs under `codeocean/data/Input_1_Spatial_transcriptomics_analysis/`. These are
not raw Visium FASTQs or raw Space Ranger matrices, but they do contain the
processed GeoMx DSP ROI expression matrix used by the paper's spatial
transcriptomics analyses, including AS, IF and TC compartments.

**Visium status:** I do not see extractable Visium count matrices, Space Ranger
outputs, spot coordinates, or spot-level annotations in the Code Ocean capsule.
The Visium branch therefore remains dependent on `HRA006579` / author-provided
spot annotation files.

------------------------------------------------------------------------

# Data preparation pipeline

Run the raw-data preparation scaffold from project root with conda **`R_env`**:

```bash
conda activate R_env
cd /home/a1237163/lab/chen/posteriorHCA
Rscript dev/Jie_HCC/data/prepare_data.R
```

```{r}
#| label: run-prepare-data
#| eval: false
system2("Rscript", file.path(params$case_dir, "data", "prepare_data.R"))
```

Run the Code Ocean processed-data extractor:

```bash
conda activate R_env
cd /home/a1237163/lab/chen/posteriorHCA
Rscript dev/Jie_HCC/data/extract_codeocean_processed_data.R
```

```{r}
#| label: run-codeocean-extract
#| eval: false
system2("Rscript", file.path(params$case_dir, "data", "extract_codeocean_processed_data.R"))
```

### Expected directory layout after download

```
dev/Jie_HCC/data/
├── raw/
│   ├── omix005738/
│   │   ├── OMIX005738-01_DSP_WTA.xlsx
│   │   └── OMIX005738-03_DSP_labels.xlsx
│   └── hra006579/
│       ├── visium_spaceranger/<sample>/filtered_feature_bc_matrix/
│       └── spot_compartment_annotations.csv   # from supplementary / authors
├── processed/
│   ├── codeocean_GeoMx_all_gene_expression_matrix.csv
│   ├── codeocean_GeoMx_all_roi_metadata.csv
│   ├── codeocean_GeoMx_all_segments_long_all_genes.csv
│   ├── codeocean_GeoMx_all_patient_compartment_summary_all_genes.csv
│   ├── codeocean_TIMES_GeoMx_TC_segments.csv
│   ├── codeocean_TIMES_GeoMx_TC_patient_summary.csv
│   ├── codeocean_TIMES_GeoMx_TC_validation.csv
│   ├── codeocean_GeoMx_TC_gene_expression_matrix.csv
│   ├── codeocean_GeoMx_TC_roi_metadata.csv
│   ├── codeocean_GeoMx_TC_segments_long_all_genes.csv
│   ├── codeocean_GeoMx_TC_patient_summary_all_genes.csv
│   ├── TIMES_CodeOcean_GeoMx_TC_processed.rds
│   └── TIMES_GeoMx_TC_pseudobulk.rds          # after raw files present
├── times_biomarkers.rds
├── data_registry.rds
├── download_manifest.rds
└── processing_plan.rds
```

### Code Ocean processed GeoMx extraction

The author Code Ocean file
`codeocean/data/Input_1_Spatial_transcriptomics_analysis/transcriptome_differential_analysis_input_1.RData`
contains:

- `data_gene_df_ed`: processed GeoMx gene-by-ROI matrix (`18,677 × 93`).
- `data_meta_df_ed`: ROI metadata with sample id, label, age, sex, relapse and
  death fields.
- `NKatStroma_indx`, `NKatBorder_indx`, `NKatTumor_indx`: author indices for
  CD57-rich NK-enriched ROIs in AS, IF and TC.

`data/extract_codeocean_processed_data.R` extracts the **complete GeoMx ROI gene
expression matrix** (`18,677 genes × 93 ROIs`), classifies ROIs into **AS / IF /
TC** from the author labels, joins recurrence metadata, and writes both
matrix-shaped and tidy summaries under `data/processed/`. The previous TC
NK-enriched subset and the five TIMES biomarkers are retained as validation and
analysis convenience subsets, not as the only extracted data.

Main full-matrix outputs:

- `codeocean_GeoMx_all_gene_expression_matrix.csv`: genes × all GeoMx ROIs.
- `codeocean_GeoMx_all_roi_metadata.csv`: all ROI metadata with `compartment =
  AS / IF / TC`.
- `codeocean_GeoMx_all_segments_long_all_genes.csv`: tidy gene × ROI values for
  all compartments.
- `codeocean_GeoMx_all_patient_compartment_summary_all_genes.csv`: per-patient,
  per-compartment summaries for all genes.
- `TIMES_CodeOcean_GeoMx_TC_processed.rds`: compact R object containing all of
  the above plus TC and TIMES subsets.

TC convenience outputs:

- `codeocean_GeoMx_TC_gene_expression_matrix.csv`: genes × author NK-enriched TC ROIs.
- `codeocean_GeoMx_TC_roi_metadata.csv`
- `codeocean_GeoMx_TC_segments_long_all_genes.csv`
- `codeocean_GeoMx_TC_patient_summary_all_genes.csv`

TIMES-specific outputs:

- `codeocean_TIMES_GeoMx_TC_segments.csv`
- `codeocean_TIMES_GeoMx_TC_patient_summary.csv`
- `codeocean_TIMES_GeoMx_TC_validation.csv`

```{r}
#| label: codeocean-full-matrix-summary
matrix_path <- file.path(data_dir, "processed", "codeocean_GeoMx_all_gene_expression_matrix.csv")
roi_meta_path <- file.path(data_dir, "processed", "codeocean_GeoMx_all_roi_metadata.csv")
if (file.exists(matrix_path) && file.exists(roi_meta_path)) {
  full_matrix <- read_csv(matrix_path, show_col_types = FALSE)
  roi_meta <- read_csv(roi_meta_path, show_col_types = FALSE)
  expression_values <- as.matrix(full_matrix[, -1, drop = FALSE])
  roi_columns_with_values <- colnames(expression_values)[colSums(!is.na(expression_values)) > 0]
  tibble(
    n_genes = nrow(full_matrix),
    n_rois = ncol(full_matrix) - 1,
    n_patients = n_distinct(roi_meta$patient_id),
    n_patients_with_any_gene_value = n_distinct(
      roi_meta$patient_id[roi_meta$author_column %in% roi_columns_with_values]
    ),
    compartment_counts = paste(names(table(roi_meta$compartment)), table(roi_meta$compartment), sep = "=", collapse = "; ")
  )
} else {
  tibble(note = "Run data/extract_codeocean_processed_data.R after exporting Code Ocean data.")
}
```

```{r}
#| label: codeocean-validation
validation_path <- file.path(data_dir, "processed", "codeocean_TIMES_GeoMx_TC_validation.csv")
if (file.exists(validation_path)) {
  read_csv(validation_path, show_col_types = FALSE)
} else {
  tibble(note = "Run data/extract_codeocean_processed_data.R after exporting Code Ocean data.")
}
```

```{r}
#| label: codeocean-patient-summary
patient_summary_path <- file.path(data_dir, "processed", "codeocean_TIMES_GeoMx_TC_patient_summary.csv")
if (file.exists(patient_summary_path)) {
  read_csv(patient_summary_path, show_col_types = FALSE) |>
    dplyr::filter(n_non_missing > 0) |>
    dplyr::select(gene, .feature, patient_id, recurrence, n_non_missing, value_sum, value_mean)
} else {
  tibble(note = "No Code Ocean patient summary extracted yet.")
}
```

**Validation note:** the complete Code Ocean GeoMx matrix has **18,677 genes ×
93 ROIs** with **26 AS**, **26 IF** and **41 TC** ROIs. The author NK-enriched TC
subset has **18,677 genes × 19 TC ROIs**, and all **6** patients have some
gene-level TC values. However, the five TIMES biomarkers are sparse in this
processed matrix: only **7 of 19** NK-enriched TC ROIs contain non-missing values
for those genes, corresponding to **2 of 6** patients (`A24`, non-REC; `A29`,
REC). Therefore, the full matrix is suitable for extracting additional genes and
checking broader GeoMx AS/IF/TC signal, while TIMES-only recurrence testing
remains underpowered. The values are treated as author-processed GeoMx
transcript-read measurements, not raw Visium counts.

### Processing plan snapshot

```{r}
#| label: processing-plan
plan_path <- file.path(data_dir, "processing_plan.rds")
if (file.exists(plan_path)) {
  readRDS(plan_path)
} else {
  tibble(note = "Run prepare_data.R to create processing_plan.rds")
}
```

------------------------------------------------------------------------

# posteriorHCA integration workflow

Following `SAVI_case_study_universal_baseline.qmd`:

1. **Build IF/TC patient pseudobulk** from the complete CodeOcean GeoMx matrix,
   restricting to author-defined NK-enriched ROIs and rounding values as in the
   paper's DESeq2 capsule.
2. **Map gene symbols → Ensembl**; intersect with `V1_nk` `targets` store.
3. **Merge** with healthy reference pseudobulk (`offset == 0`) from Zenodo h5ad-backed
   `pseudobulk_sample`, then run `tidybulk::scale_abundance(method = "TMMwsp",
   reference_sample = reference)` and set `offset = log(1 / multiplier)` exactly
   as in SAVI.
4. **Four-cohort estimates:** extract cohort **NB μ** at offset 0 and QL Wald SE
   for IF-REC, IF-non-REC, TC-REC and TC-non-REC.
5. **Healthy baseline:** load compressed `brms` fits (`estimates_chunk`) or precomputed
   `gene_baseline_draws_universal.rds`; **`posterior_linpred(transform = FALSE)`** with
   universal marginalisation grid.
6. **Primary proxy:** standardize absolute distance from the healthy grand mean,
   average across genes, propagate posterior and QL uncertainty, and test
   `REC distance > non-REC distance`.
7. **Patient evaluation:** compute IF, TC and combined outcome-free scores, AUC,
   exact rank tests and leave-one-gene-out sensitivity.
8. **Positive control:** test one-sided `non-REC > REC` QL contrasts separately
   in IF and TC for all five genes.
9. **Benchmark:** compare descriptively with supplied TIMES validation-score
   AUC, without claiming formal equivalence.

```{r}
#| label: model-store-preview
library(targets)
library(stringr)

model_path <- file.path(params$model_store_root, paste0("V1_", params$target_cell_type))

if (dir.exists(file.path(model_path, "_targets"))) {
  model_meta <- tar_meta(
    starts_with("estimates_chunk_"),
    store = file.path(model_path, "_targets")
  ) |>
    filter(type == "branch") |>
    mutate(
      .feature = str_extract(warnings, "(?<=Gene:___)ENSG\\d+(?=___)")
    )

  tibble(
    atlas_cell_type = params$target_cell_type,
    n_genes_in_store = n_distinct(model_meta$.feature),
    n_biomarkers_in_store = sum(biomarkers$.feature %in% model_meta$.feature, na.rm = TRUE),
    n_biomarkers_mapped = sum(!is.na(biomarkers$.feature))
  )
} else {
  tibble(note = glue("Model store not found at {model_path}"))
}
```

------------------------------------------------------------------------

# Expected outcomes and falsifiable predictions

| Prediction | If true | If false |
|------------|---------|----------|
| REC is farther from healthy than non-REC | Positive proxy margins across IF/TC and combined posterior probability >0.95 | Proxy does not recover recurrence ordering |
| Non-REC is baseline-compatible | Mean distance below one healthy SD | non-REC is equally atlas-discordant |
| Proxy is not driven by one biomarker | Leave-one-gene-out patient ordering remains intact | Ordering collapses when one gene is removed |
| Available GeoMx retains paper biology | Positive non-REC vs REC effects in IF and TC | Positive control directions disagree |
| Proxy has competitive descriptive discrimination | Patient AUC approaches supplied TIMES benchmark | Healthy reference adds little recurrence information |

------------------------------------------------------------------------

# Timeline and deliverables

| Step | Deliverable | Status |
|------|-------------|--------|
| 1 | Planning qmd (this document) | Done |
| 2 | `prepare_data.R` + biomarker/registry RDS | Done |
| 3 | Code Ocean processed GeoMx input export | Done |
| 4 | Code Ocean IF/TC biomarker extraction + validation | Done — sparse, 5 patient-compartment samples |
| 5 | CNCB raw Visium / GeoMx download | **Blocked** — controlled access |
| 6 | Raw-count IF/TC pseudobulk RDS | Pending raw data |
| 7 | `Jie_HCC_case_study_universal_baseline.qmd` (analysis) | Done |
| 8 | Healthy-proximity proxy, positive control and benchmark | Done |

------------------------------------------------------------------------

# References

- Jia, G., *et al.* (2025). Spatial immune scoring system predicts hepatocellular
  carcinoma recurrence. *Nature* **640**, 1029–1040. doi: [`10.1038/s41586-025-08668-x`](https://doi.org/10.1038/s41586-025-08668-x)
- Sun, Y., *et al.* (2021). Single-cell landscape of the ecosystem in early-relapse
  hepatocellular carcinoma. *Cell* **184**, 404–421.e16. (scRNA NK validation, ref. 28)
- de Cevins, C., *et al.* (2023). SAVI monocyte signature — template case study in
  `dev/SAVI/`
