---
title: "Age-smooth k-grid evaluation"
subtitle: "Convergence, age curves, brms smooth-term comparison, vs strong RE, fitting time"
date: today
date-format: "DD MMM YYYY"
title-block-banner: true
author:
  - name: Chen Zhan
    orcid: 0000-0002-4794-8339
    email: chen.zhan@adelaide.edu.au
    affiliation:
      - id: sgc
        name: SAiGENCI, The University of Adelaide
format:
  html:
    theme: minty
    df-print: paged
    code-line-numbers: true
    embed-resources: true
    code-fold: true
    code-fold-default: true
    toc: true
    toc-location: left
    toc-title: "Contents"
knitr:
  opts_chunk:
    message: false
    warning: false
editor: source
abstract: |
  Evaluation of continuous-age thin-plate smooth fits on healthy V1_nk samples
  (`disease_groups_altered == "Normal"`) for four TIMES biomarkers
  (SPON2, ZFP36L2, ZFP36, VIM) and smooth basis sizes k ∈ {3,4,5}. We report
  HMC convergence, blood/liver age curves, an explicit comparison with
  `brms::conditional_smooths()`, predict/linpred compatibility (needed for
  comprehensive μ), comprehensive healthy liver baseline mean/SD against
  the strong-HMC correlated RE (simple shape) fits, and wall-clock fitting
  times from the `formula_bench` targets pipeline (age-smooth vs RE formulas
  across chain × CPU settings).
params:
  case_dir: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC"
  smooth_out_dir: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/age_smooth_k_grid_pilot"
  strong_out_dir: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/re_vs_interaction_strong"
  diagnostics_csv: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/age_smooth_k_grid_pilot/age_smooth_eval_diagnostics.csv"
  curves_rds: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/age_smooth_k_grid_pilot/age_smooth_eval_curves_group_support.rds"
  conditional_smooths_rds: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/age_smooth_k_grid_pilot/age_smooth_eval_conditional_smooths_group_support.rds"
  baseline_draws_rds: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/age_smooth_k_grid_pilot/age_smooth_eval_baseline_draws.rds"
  compat_csv: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/age_smooth_k_grid_pilot/age_smooth_eval_linpred_compat.csv"
  timing_summary_csv: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/formula_bench/timing_summary.csv"
  refresh: false
  n_curve_draws: 200
  n_baseline_draws: 400
---

# Motivation

Age-smooth models replace categorical `age_decade` with

`s(age_days_scaled, by = tissue_sex, bs = "tp", k, m = 1)`

on **healthy** rows only. This report asks:

1. **Convergence** — does HMC reliability depend on k?  
2. **Age curves** — how do blood / liver log-mean trajectories look by k, and
   how do they differ from `brms::conditional_smooths()`?  
3. **Predict API** — do `posterior_linpred` / `posterior_predict` work on these
   objects (required for comprehensive μ)?  
4. **vs strong RE** — how do comprehensive liver baseline μ and SD compare to
   the long-run correlated RE (simple shape) fits?  
5. **Fitting time** — how does wall-clock HMC time compare across formulas
   (age-smooth k vs RE) and chain × CPU settings from the `formula_bench`
   targets pipeline?

| Model | Data | Age | MCMC |
|-------|------|-----|------|
| Age-smooth k=3/4/5 | Normal only (n≈2905) | continuous smooth | 4×400/900, control=NULL |
| Strong RE (report §4) | all disease (n≈3226) | age_decade FE/RE | 4×1500/3000, δ=0.95 |
| formula_bench (§5) | Normal only (n=2905) | smooth or age_decade | 2/4 chains × 400/900 |

------------------------------------------------------------------------

# Library setup

```{r}
#| label: setup

suppressPackageStartupMessages({
  library(tibble)
  library(tidyr)
  library(dplyr)
  library(readr)
  library(ggplot2)
  library(forcats)
  library(scales)
  library(glue)
  library(stringr)
  library(posterior)
  library(brms)
})

source(file.path(params$case_dir, "R", "re_vs_interaction_helpers.R"))
source(file.path(params$case_dir, "R", "fit_convergence_helpers.R"))
source(file.path(params$case_dir, "R", "age_smooth_helpers.R"))

theme_set(theme_minimal(base_size = 11))

genes <- pilot_biomarker_genes()
ks <- age_smooth_ks_default()

k_colours <- c(`3` = "#4DBBD5", `4` = "#E64B35", `5` = "#3C5488")
model_colours <- c(
  `RE HMC (simple)` = "#7A7A7A",
  `Age-smooth k=3` = "#4DBBD5",
  `Age-smooth k=4` = "#E64B35",
  `Age-smooth k=5` = "#3C5488"
)
gate_colours <- c(PASS = "#2D8B57", WARN = "#E08A2E", FAIL = "#C23B4A")
setting_colours <- c(
  `2×8 (4 thr)` = "#00A087",
  `4×8 (2 thr)` = "#E64B35",
  `4×16 (4 thr)` = "#3C5488"
)
```

------------------------------------------------------------------------

# Fit inventory

```{r}
#| label: inventory

inv <- age_smooth_inventory(params$smooth_out_dir, genes = genes, ks = ks)
inv |>
  transmute(
    gene, k, model,
    status = if_else(available, "OK", "MISSING"),
    file = basename(rds_path)
  ) |>
  arrange(gene, k)
```

```{r}
#| label: inventory-text
#| results: asis

n_ok <- sum(inv$available)
cat(glue("**{n_ok}/{nrow(inv)}** age-smooth fits available under `{params$smooth_out_dir}`."))
```

------------------------------------------------------------------------

# 1. Convergence by k

HMC gates (same spirit as the strong-formula report):

| Gate | PASS |
|------|------|
| R̂ | `max_rhat ≤ 1.05` |
| Divergences | `< 1%` of post-warmup transitions |

`FAIL` if `max_rhat > 1.5` or `≥ 50%` divergences; else soft → `WARN`.

```{r}
#| label: diagnostics
#| cache: false

diag_path <- params$diagnostics_csv
need_diag <- isTRUE(params$refresh) || !file.exists(diag_path)

if (need_diag) {
  message("Computing age-smooth diagnostics + baselines ...")
  diag <- diagnose_age_smooth_fits(
    out_dir = params$smooth_out_dir,
    genes = genes,
    ks = ks,
    compute_baseline = TRUE,
    n_baseline_draws = params$n_baseline_draws,
    cache_csv = diag_path
  )
} else {
  diag <- read_csv(diag_path, show_col_types = FALSE)
}

diag_view <- diag |>
  mutate(
    k = factor(k, levels = ks),
    gate = case_when(
      gate_pass ~ "PASS",
      max_rhat > 1.5 | pct_divergent >= 50 ~ "FAIL",
      TRUE ~ "WARN"
    ),
    gate = factor(gate, levels = c("PASS", "WARN", "FAIL"))
  )
```

## Full diagnostics table

```{r}
#| label: diagnostics-table

diag_view |>
  transmute(
    gene,
    k,
    settings = glue("{chains}×{warmup}/{iter}"),
    max_rhat = round(max_rhat, 3),
    n_rhat_gt_1_05,
    pct_divergent = round(pct_divergent, 2),
    pct_max_treedepth = round(pct_max_treedepth, 1),
    min_ess_bulk = round(min_ess_bulk, 0),
    baseline_mean = round(baseline_mean, 3),
    baseline_sd = round(baseline_sd, 3),
    baseline_n_draws,
    gate
  ) |>
  arrange(gene, k)
```

## Gate summary

```{r}
#| label: gate-summary

diag_view |>
  count(k, gate) |>
  pivot_wider(names_from = gate, values_from = n, values_fill = 0) |>
  arrange(k)
```

```{r}
#| label: plot-max-rhat
#| fig-width: 9
#| fig-height: 4.5
#| fig-align: center

ggplot(diag_view, aes(x = gene, y = max_rhat, fill = k)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  geom_hline(yintercept = 1.05, linetype = 2, colour = "grey40") +
  geom_hline(yintercept = 1.1, linetype = 3, colour = "grey60") +
  scale_fill_manual(values = k_colours) +
  labs(
    title = "Maximum R̂ by gene and smooth basis size k",
    subtitle = "Dashed = 1.05 PASS gate",
    x = NULL, y = "max R̂", fill = "k"
  ) +
  theme(legend.position = "bottom")
```

```{r}
#| label: plot-divergences
#| fig-width: 9
#| fig-height: 4.5
#| fig-align: center

ggplot(diag_view, aes(x = gene, y = pct_divergent, fill = k)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey40") +
  scale_fill_manual(values = k_colours) +
  labs(
    title = "Divergent transitions (% of post-warmup)",
    subtitle = "Dashed = 1% PASS gate; age-smooth pilots used brms default adapt_delta",
    x = NULL, y = "% divergent", fill = "k"
  ) +
  theme(legend.position = "bottom")
```

```{r}
#| label: convergence-text
#| results: asis

n_pass <- sum(diag_view$gate == "PASS")
n_warn <- sum(diag_view$gate == "WARN")
n_fail <- sum(diag_view$gate == "FAIL")
worst <- diag_view |> arrange(desc(max_rhat)) |> slice(1)
cat(glue(
  "Across {nrow(diag_view)} gene × k cells: **{n_pass} PASS**, **{n_warn} WARN**, ",
  "**{n_fail} FAIL**. Worst max R̂ = **{round(worst$max_rhat, 3)}** for ",
  "**{worst$gene} / k={worst$k}** ({round(worst$pct_divergent, 1)}% divergent)."
))
```

------------------------------------------------------------------------

# 2. Age-smooth curves (blood & liver)

The report curve is the population-level (`re_formula = NA`) linear predictor,
with `offset = 0`, versus `age_days_scaled`. To make its absolute level
reproducible, ethnicity and assay are fixed to their most frequent values in
the fit instead of being inherited from the first matching training row.
Each curve is restricted to its tissue–sex group's observed 5th–95th percentile
age range to avoid extrapolating the sparse liver strata. Curves show the
posterior mean and pointwise 95% interval for **blood** and **liver**, faceted
by sex; colours denote k.

```{r}
#| label: curves
#| cache: false

curves_path <- params$curves_rds
need_curves <- isTRUE(params$refresh) || !file.exists(curves_path)

if (need_curves) {
  message("Building blood/liver age curves ...")
  inv_ok <- inv |> filter(available)
  curve_rows <- vector("list", nrow(inv_ok))
  for (i in seq_len(nrow(inv_ok))) {
    row <- inv_ok[i, ]
    message(sprintf("[%d/%d] curves %s k=%d", i, nrow(inv_ok), row$gene, row$k))
    fit <- readRDS(row$rds_path)
    curve_rows[[i]] <- age_smooth_curve_draws(
      fit,
      tissues = c("blood", "liver"),
      n_age = 40L,
      ndraws = params$n_curve_draws,
      re_formula = NA
    ) |>
      mutate(gene = row$gene, k = factor(row$k, levels = ks))
  }
  curves <- bind_rows(curve_rows)
  dir.create(dirname(curves_path), recursive = TRUE, showWarnings = FALSE)
  saveRDS(curves, curves_path, compress = "xz")
} else {
  curves <- readRDS(curves_path)
}
```

```{r}
#| label: curve-reference-profile

curves |>
  distinct(gene, k, reference_profile) |>
  arrange(gene, k)
```

```{r}
#| label: curve-age-support

curves |>
  distinct(tissue_groups, sex, support_lower, support_upper) |>
  mutate(across(c(support_lower, support_upper), ~ round(.x, 3))) |>
  arrange(tissue_groups, sex)
```

```{r}
#| label: plot-age-curves
#| fig-width: 10
#| fig-height: 10
#| fig-align: center

ggplot(curves, aes(x = age_days_scaled, y = estimate, colour = k, fill = k)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) +
  facet_grid(gene ~ tissue_groups + sex, scales = "free_y") +
  scale_colour_manual(values = k_colours) +
  scale_fill_manual(values = k_colours) +
  labs(
    title = "Reference-profile log-mean trajectories (blood & liver)",
    subtitle = "Full population-level linpred on group-specific age support; ribbon = 95% interval",
    x = "age_days_scaled",
    y = "Population-level log(μ)",
    colour = "k", fill = "k"
  ) +
  theme(legend.position = "bottom")
```

## What `conditional_smooths()` computes

The two curve APIs answer different questions. The report helper evaluates the
**full** count-component linear predictor, whereas
[`brms::conditional_smooths()`](https://paulbuerkner.com/brms/reference/conditional_smooths.brmsfit.html)
uses `posterior_smooths()` internally to isolate the contribution from the
selected smooth term.

| Feature | Report helper: `age_smooth_curve_draws()` | `brms::conditional_smooths()` |
|---|---|---|
| Quantity | Full population-level `log(μ)` linear predictor | Only `s(age_days_scaled, by = tissue_sex, ...)` |
| Included terms | Intercept, sex, reference ethnicity/assay, smooth | Smooth term only |
| Group-level terms | Excluded by `re_formula = NA` | Not part of the isolated smooth |
| Offset | Set to 0 | Not part of the isolated smooth |
| Point estimate | Posterior mean | Posterior median |
| Interval | Pointwise 2.5% and 97.5% quantiles | Pointwise 2.5% and 97.5% quantiles |
| Default age grid | Group-specific 5th–95th percentiles, 40 points | Full observed range, 100 points |
| Grid used below | Group-specific 5th–95th percentiles, 40 points | Matched to the report helper |

Thus, the raw vertical positions should **not** agree: the brms curve is a
smooth-term contribution, not an absolute expected log mean. Their
age-dependent shapes should agree after removing an age-invariant vertical
shift, up to mean-versus-median summarisation and posterior Monte Carlo error.

```{r}
#| label: conditional-smooth-curves
#| cache: false

conditional_path <- params$conditional_smooths_rds
need_conditional <- isTRUE(params$refresh) || !file.exists(conditional_path)

if (need_conditional) {
  message("Building brms conditional_smooths curves ...")
  inv_ok <- inv |> filter(available)
  conditional_rows <- vector("list", nrow(inv_ok))
  for (i in seq_len(nrow(inv_ok))) {
    row <- inv_ok[i, ]
    message(sprintf(
      "[%d/%d] conditional_smooths %s k=%d",
      i, nrow(inv_ok), row$gene, row$k
    ))
    fit <- readRDS(row$rds_path)
    conditional_rows[[i]] <- age_smooth_conditional_smooths(
      fit,
      tissues = c("blood", "liver"),
      n_age = 40L,
      ndraws = params$n_curve_draws
    ) |>
      mutate(gene = row$gene, k = factor(row$k, levels = ks))
  }
  conditional_curves <- bind_rows(conditional_rows)
  dir.create(dirname(conditional_path), recursive = TRUE, showWarnings = FALSE)
  saveRDS(conditional_curves, conditional_path, compress = "xz")
} else {
  conditional_curves <- readRDS(conditional_path)
}
```

```{r}
#| label: plot-conditional-smooths
#| fig-width: 10
#| fig-height: 10
#| fig-align: center

ggplot(
  conditional_curves,
  aes(x = age_days_scaled, y = estimate, colour = k, fill = k)
) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) +
  facet_grid(gene ~ tissue_groups + sex, scales = "free_y") +
  scale_colour_manual(values = k_colours) +
  scale_fill_manual(values = k_colours) +
  labs(
    title = "Age smooth returned by brms::conditional_smooths()",
    subtitle = "Term only, on group-specific age support; median and 95% interval",
    x = "age_days_scaled",
    y = "Smooth contribution to log(μ)",
    colour = "k", fill = "k"
  ) +
  theme(legend.position = "bottom")
```

## Shape comparison on a common zero

For visual comparison, each method's point-estimate curve is shifted to zero at
the grid point nearest `age_days_scaled = 0`. This removes the non-age terms
from the full linear predictor. It is a comparison of displayed point
estimates, not a new posterior estimand.

```{r}
#| label: centred-curve-comparison-data

curve_comparison <- bind_rows(
  curves |>
    transmute(
      gene, k, tissue_groups, sex, age_days_scaled, estimate,
      method = "Full linpred (mean)"
    ),
  conditional_curves |>
    transmute(
      gene, k, tissue_groups, sex, age_days_scaled, estimate,
      method = "conditional_smooths (median)"
    )
) |>
  group_by(gene, k, tissue_groups, sex, method) |>
  arrange(age_days_scaled, .by_group = TRUE) |>
  mutate(
    anchor = estimate[which.min(abs(age_days_scaled))],
    centred_estimate = estimate - anchor
  ) |>
  ungroup()

shape_difference <- curve_comparison |>
  select(
    gene, k, tissue_groups, sex, age_days_scaled, method, centred_estimate
  ) |>
  pivot_wider(names_from = method, values_from = centred_estimate) |>
  mutate(
    difference = `Full linpred (mean)` - `conditional_smooths (median)`
  )

shape_difference |>
  group_by(gene, k, tissue_groups, sex) |>
  summarise(
    max_abs_difference = max(abs(difference)),
    rmse = sqrt(mean(difference^2)),
    .groups = "drop"
  ) |>
  mutate(across(c(max_abs_difference, rmse), ~ round(.x, 4))) |>
  arrange(desc(max_abs_difference))
```

```{r}
#| label: plot-centred-curve-comparison
#| fig-width: 10
#| fig-height: 10
#| fig-align: center

ggplot(
  curve_comparison,
  aes(
    x = age_days_scaled,
    y = centred_estimate,
    colour = k,
    linetype = method
  )
) +
  geom_hline(yintercept = 0, colour = "grey80", linewidth = 0.35) +
  geom_line(linewidth = 0.75) +
  facet_grid(gene ~ tissue_groups + sex, scales = "free_y") +
  scale_colour_manual(values = k_colours) +
  scale_linetype_manual(values = c(
    `Full linpred (mean)` = "solid",
    `conditional_smooths (median)` = "22"
  )) +
  labs(
    title = "Age-dependent shape: full linpred vs conditional_smooths",
    subtitle = "Each displayed curve is anchored at age_days_scaled ≈ 0",
    x = "age_days_scaled",
    y = "Change from value at scaled age ≈ 0",
    colour = "k",
    linetype = NULL
  ) +
  theme(legend.position = "bottom")
```

```{r}
#| label: centred-curve-comparison-text
#| results: asis

shape_audit <- shape_difference |>
  group_by(gene, k, tissue_groups, sex) |>
  summarise(
    max_abs_difference = max(abs(difference)),
    .groups = "drop"
  )
largest_shape_difference <- shape_audit |>
  arrange(desc(max_abs_difference)) |>
  slice(1)

cat(glue(
  "After anchoring, the median group-wise maximum absolute difference is ",
  "**{round(median(shape_audit$max_abs_difference), 3)} log units**. ",
  "The largest is **{round(largest_shape_difference$max_abs_difference, 3)}** ",
  "for **{largest_shape_difference$gene}, k={largest_shape_difference$k}, ",
  "{largest_shape_difference$tissue_groups}, ",
  "{largest_shape_difference$sex}**. Residual discrepancies are expected because ",
  "the full curve is summarised by the posterior mean whereas ",
  "`conditional_smooths()` uses the posterior median; sparse liver strata also ",
  "have wider, more skewed posteriors. For absolute `log(μ)`, retain the full ",
  "linear predictor. For the isolated age-shape diagnostic, use ",
  "`conditional_smooths()`."
))
```

------------------------------------------------------------------------

# 3. brms predict / linpred compatibility

Smoke-test on one fit per gene (prefer k=4 when available) to confirm the
prediction API used for comprehensive μ works on smooth objects.

```{r}
#| label: linpred-compat
#| cache: false

compat_path <- params$compat_csv
need_compat <- isTRUE(params$refresh) || !file.exists(compat_path)

if (need_compat) {
  compat_rows <- list()
  for (g in genes) {
    prefer <- inv |> filter(gene == g, k == 4L, available)
    if (!nrow(prefer)) {
      prefer <- inv |> filter(gene == g, available) |> slice(1)
    }
    if (!nrow(prefer)) next
    message("Compat checks: ", prefer$gene, " k=", prefer$k)
    fit <- readRDS(prefer$rds_path[[1]])
    compat_rows[[g]] <- age_smooth_predict_compat(fit, ndraws = 50L) |>
      mutate(gene = prefer$gene, k = prefer$k)
  }
  compat <- bind_rows(compat_rows)
  write_csv(compat, compat_path)
} else {
  compat <- read_csv(compat_path, show_col_types = FALSE)
}

compat |>
  mutate(
    elapsed_sec = round(elapsed_sec, 2),
    ok = if_else(ok, "YES", "NO")
  ) |>
  arrange(gene, check)
```

```{r}
#| label: linpred-compat-text
#| results: asis

n_ok <- sum(compat$ok %in% c(TRUE, "YES"))
n_tot <- nrow(compat)
if (n_ok == n_tot) {
  cat(glue(
    "All **{n_tot}** prediction checks passed, including `posterior_linpred` ",
    "and the comprehensive liver baseline helper — smooth fits are usable for μ."
  ))
} else {
  bad <- compat |> filter(!ok %in% c(TRUE, "YES"))
  cat(glue(
    "**{n_ok}/{n_tot}** checks passed. Failures: ",
    "{paste(bad$gene, bad$check, sep = '/', collapse = '; ')}."
  ))
}
```

------------------------------------------------------------------------

# 4. Baseline μ / SD vs strong RE (simple shape)

Comprehensive healthy **liver** baseline:

- **Age-smooth** — average `posterior_linpred` over sex × ethnicity × assay ×
  age quantile grid (`build_comprehensive_grid_age_smooth`).  
- **Strong RE** — existing comprehensive grid (`age_decade` margins) from the
  formula-comparison helpers.

Densities use `n_baseline_draws` posterior draws of the scalar baseline
`log(μ)`.

```{r}
#| label: baseline-draws
#| cache: false

draws_path <- params$baseline_draws_rds
need_draws <- isTRUE(params$refresh) || !file.exists(draws_path)

if (need_draws) {
  message("Collecting baseline draws (smooth + RE) ...")
  out <- list()

  inv_ok <- inv |> filter(available)
  for (i in seq_len(nrow(inv_ok))) {
    row <- inv_ok[i, ]
    message(sprintf("[smooth %d/%d] %s k=%d", i, nrow(inv_ok), row$gene, row$k))
    fit <- readRDS(row$rds_path)
    d <- linpred_baseline_draws_age_smooth(
      fit, ndraws = params$n_baseline_draws
    ) |>
      mutate(
        gene = row$gene,
        model = paste0("Age-smooth k=", row$k),
        k = row$k,
        family = "age_smooth"
      )
    out[[length(out) + 1L]] <- d
  }

  for (g in genes) {
    re_path <- file.path(params$strong_out_dir, paste0("fit_re_", g, ".rds"))
    if (!file.exists(re_path)) next
    message("RE baseline: ", g)
    fit_re <- readRDS(re_path)
    d <- linpred_baseline_draws(fit_re, ndraws = params$n_baseline_draws) |>
      mutate(
        gene = g,
        model = "RE HMC (simple)",
        k = NA_integer_,
        family = "re"
      )
    out[[length(out) + 1L]] <- d
  }

  baseline_draws <- bind_rows(out)
  dir.create(dirname(draws_path), recursive = TRUE, showWarnings = FALSE)
  saveRDS(baseline_draws, draws_path, compress = "xz")
} else {
  baseline_draws <- readRDS(draws_path)
}

baseline_draws <- baseline_draws |>
  mutate(
    model = factor(model, levels = names(model_colours))
  )

baseline_summary <- baseline_draws |>
  group_by(gene, model, family, k) |>
  summarise(
    n_draws = n(),
    mean_log_mu = mean(baseline_log_mu),
    sd_log_mu = sd(baseline_log_mu),
    .groups = "drop"
  )
```

## Mean / SD table

```{r}
#| label: baseline-summary-table

baseline_summary |>
  mutate(
    mean_log_mu = round(mean_log_mu, 3),
    sd_log_mu = round(sd_log_mu, 3)
  ) |>
  arrange(gene, family, k)
```

## Smooth vs RE deltas

```{r}
#| label: baseline-deltas

deltas <- baseline_summary |>
  select(gene, model, mean_log_mu, sd_log_mu) |>
  pivot_wider(
    names_from = model,
    values_from = c(mean_log_mu, sd_log_mu)
  )

# Build delta columns relative to RE when present
re_mean_col <- "mean_log_mu_RE HMC (simple)"
re_sd_col <- "sd_log_mu_RE HMC (simple)"
smooth_mean_cols <- grep("^mean_log_mu_Age-smooth", names(deltas), value = TRUE)
smooth_sd_cols <- grep("^sd_log_mu_Age-smooth", names(deltas), value = TRUE)

if (re_mean_col %in% names(deltas)) {
  for (col in smooth_mean_cols) {
    deltas[[paste0("delta_", col)]] <- deltas[[col]] - deltas[[re_mean_col]]
  }
  for (col in smooth_sd_cols) {
    deltas[[paste0("ratio_", col)]] <- deltas[[col]] / deltas[[re_sd_col]]
  }
}

deltas |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  arrange(gene)
```

```{r}
#| label: plot-baseline-mean-sd
#| fig-width: 10
#| fig-height: 5
#| fig-align: center

sum_plot <- baseline_summary |>
  mutate(k_label = if_else(is.na(k), "RE", as.character(k)))

p1 <- ggplot(sum_plot, aes(x = gene, y = mean_log_mu, fill = model)) +
  geom_col(position = position_dodge(width = 0.85), width = 0.8) +
  scale_fill_manual(values = model_colours, drop = FALSE) +
  labs(title = "Comprehensive liver baseline mean", y = "mean log(μ)", x = NULL, fill = NULL) +
  theme(legend.position = "bottom")

p2 <- ggplot(sum_plot, aes(x = gene, y = sd_log_mu, fill = model)) +
  geom_col(position = position_dodge(width = 0.85), width = 0.8) +
  scale_fill_manual(values = model_colours, drop = FALSE) +
  labs(title = "Comprehensive liver baseline SD", y = "SD log(μ)", x = NULL, fill = NULL) +
  theme(legend.position = "bottom")

p1
p2
```

## Overlapping baseline densities

```{r}
#| label: plot-baseline-density
#| fig-width: 10
#| fig-height: 7
#| fig-align: center

ggplot(baseline_draws, aes(x = baseline_log_mu, colour = model)) +
  geom_density(linewidth = 0.85) +
  facet_wrap(~ gene, scales = "free") +
  scale_colour_manual(values = model_colours, drop = FALSE) +
  labs(
    title = "Comprehensive liver baseline posterior densities",
    subtitle = "Age-smooth k=3/4/5 vs strong RE HMC (simple shape)",
    x = "baseline log(μ)", y = "Density", colour = NULL
  ) +
  theme(legend.position = "bottom")
```

```{r}
#| label: interpretation
#| results: asis

best_k <- diag_view |>
  group_by(k) |>
  summarise(
    median_rhat = median(max_rhat, na.rm = TRUE),
    median_div = median(pct_divergent, na.rm = TRUE),
    n_pass = sum(gate == "PASS"),
    .groups = "drop"
  ) |>
  arrange(desc(n_pass), median_rhat)

re_vs <- baseline_summary |>
  filter(family == "age_smooth") |>
  left_join(
    baseline_summary |>
      filter(family == "re") |>
      select(gene, re_mean = mean_log_mu, re_sd = sd_log_mu),
    by = "gene"
  ) |>
  mutate(
    abs_mean_shift = abs(mean_log_mu - re_mean),
    sd_ratio = sd_log_mu / re_sd
  )

cat(glue(
  "Best k by gate/R̂: **k={best_k$k[1]}** ",
  "({best_k$n_pass[1]} PASS; median R̂={round(best_k$median_rhat[1], 3)}; ",
  "median %div={round(best_k$median_div[1], 1)}). ",
  "Median |smooth − RE| baseline mean shift: ",
  "**{round(median(re_vs$abs_mean_shift, na.rm = TRUE), 3)}**; ",
  "median SD ratio (smooth/RE): ",
  "**{round(median(re_vs$sd_ratio, na.rm = TRUE), 3)}**. ",
  "Note: smooth fits use healthy-only data; RE used the full disease mix."
))
```

------------------------------------------------------------------------

# 5. Fitting-time benchmark (`formula_bench`)

Wall-clock HMC times from the `targets` pipeline under
`targets/formula_bench/`. All cells use the same healthy-only frame
(n = 2905), MCMC length 400 warmup / 900 iter, and `control = NULL`. The grid
crosses four genes × five formulas × three resource settings:

| setting_id | Chains | Slurm CPUs | Threads / chain |
|------------|--------|------------|-----------------|
| `c2_cpu8` | 2 | 8 | 4 |
| `c4_cpu8` | 4 | 8 | 2 |
| `c4_cpu16` | 4 | 16 | 4 |

Formulas: age-smooth k ∈ {3,4,5} (`AGE-k*`) plus healthy-adapted RE
(`RE-simple`, `RE-split`). Times are exclusive of data prep; they measure
`brms::brm()` wall clock on the assigned Slurm controller.

```{r}
#| label: timing-load

timing_raw <- read_csv(params$timing_summary_csv, show_col_types = FALSE)

timing <- timing_raw |>
  mutate(
    formula_id = factor(
      formula_id,
      levels = c("AGE-k3", "AGE-k4", "AGE-k5", "RE-simple", "RE-split")
    ),
    setting_id = factor(
      setting_id,
      levels = c("c2_cpu8", "c4_cpu8", "c4_cpu16")
    ),
    setting_label = factor(
      case_when(
        setting_id == "c2_cpu8" ~ "2×8 (4 thr)",
        setting_id == "c4_cpu8" ~ "4×8 (2 thr)",
        setting_id == "c4_cpu16" ~ "4×16 (4 thr)",
        TRUE ~ as.character(setting_label)
      ),
      levels = names(setting_colours)
    ),
    family_group = if_else(family == "age_smooth", "Age-smooth", "RE"),
    elapsed_min = elapsed_sec / 60,
    gene = factor(gene, levels = genes)
  )

stopifnot(all(timing$fit_ok))
```

```{r}
#| label: timing-inventory-text
#| results: asis

cat(glue(
  "**{nrow(timing)}/{nrow(timing)}** benchmark fits succeeded ",
  "({n_distinct(timing$gene)} genes × {n_distinct(timing$formula_id)} formulas × ",
  "{n_distinct(timing$setting_id)} settings). ",
  "Source: `{params$timing_summary_csv}`."
))
```

## Timing table (minutes)

```{r}
#| label: timing-table

timing |>
  transmute(
    gene,
    formula_id,
    setting = setting_label,
    threads = threads_per_chain,
    elapsed_min = round(elapsed_min, 1),
    max_rhat = round(max_rhat, 3)
  ) |>
  arrange(gene, formula_id, setting)
```

## Elapsed time by formula and resource setting

```{r}
#| label: plot-timing-by-formula
#| fig-width: 10
#| fig-height: 7
#| fig-align: center

ggplot(timing, aes(x = formula_id, y = elapsed_min, fill = setting_label)) +
  geom_col(position = position_dodge(width = 0.85), width = 0.8) +
  facet_wrap(~ gene, nrow = 2) +
  scale_fill_manual(values = setting_colours) +
  labs(
    title = "HMC wall-clock time by formula and resource setting",
    subtitle = "Healthy-only n=2905; 400 warmup / 900 iter; fill = chains×CPUs (threads/chain)",
    x = NULL, y = "Elapsed (minutes)", fill = "Setting"
  ) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```

## Age-smooth k vs RE under each setting

```{r}
#| label: plot-timing-family-by-setting
#| fig-width: 10
#| fig-height: 4.5
#| fig-align: center

timing_family <- timing |>
  group_by(gene, setting_label, family_group) |>
  summarise(
    elapsed_min = mean(elapsed_min),
    .groups = "drop"
  )

ggplot(
  timing_family,
  aes(x = gene, y = elapsed_min, fill = family_group)
) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  facet_wrap(~ setting_label, nrow = 1) +
  scale_fill_manual(
    values = c(`Age-smooth` = "#3C5488", `RE` = "#7A7A7A")
  ) +
  labs(
    title = "Mean fit time: age-smooth (avg over k) vs RE (avg over formulas)",
    subtitle = "One panel per chains×CPU setting",
    x = NULL, y = "Mean elapsed (minutes)", fill = NULL
  ) +
  theme(legend.position = "bottom")
```

## Effect of k on age-smooth fit time

```{r}
#| label: plot-timing-by-k
#| fig-width: 10
#| fig-height: 4.5
#| fig-align: center

timing_age <- timing |>
  filter(family == "age_smooth") |>
  mutate(k = factor(k, levels = ks))

ggplot(timing_age, aes(x = gene, y = elapsed_min, fill = k)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  facet_wrap(~ setting_label, nrow = 1) +
  scale_fill_manual(values = k_colours) +
  labs(
    title = "Age-smooth fit time by smooth basis size k",
    subtitle = "Facets = resource setting; larger k is not systematically slower",
    x = NULL, y = "Elapsed (minutes)", fill = "k"
  ) +
  theme(legend.position = "bottom")
```

## Resource-setting speedup (relative to 4×8)

The pilot default is 4 chains on 8 CPUs (2 within-chain threads). Speedup is
elapsed(4×8) / elapsed(setting) for the same gene × formula.

```{r}
#| label: plot-timing-speedup
#| fig-width: 10
#| fig-height: 5.5
#| fig-align: center

timing_speedup <- timing |>
  select(gene, formula_id, setting_label, elapsed_sec) |>
  pivot_wider(names_from = setting_label, values_from = elapsed_sec) |>
  mutate(
    speedup_2x8 = `4×8 (2 thr)` / `2×8 (4 thr)`,
    speedup_4x16 = `4×8 (2 thr)` / `4×16 (4 thr)`
  ) |>
  select(gene, formula_id, speedup_2x8, speedup_4x16) |>
  pivot_longer(
    cols = starts_with("speedup_"),
    names_to = "alt_setting",
    values_to = "speedup"
  ) |>
  mutate(
    alt_setting = factor(
      recode(
        alt_setting,
        speedup_2x8 = "2×8 (4 thr)",
        speedup_4x16 = "4×16 (4 thr)"
      ),
      levels = c("2×8 (4 thr)", "4×16 (4 thr)")
    )
  )

ggplot(
  timing_speedup,
  aes(x = formula_id, y = speedup, fill = alt_setting)
) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey50") +
  geom_col(position = position_dodge(width = 0.85), width = 0.8) +
  facet_wrap(~ gene, nrow = 2) +
  scale_fill_manual(
    values = c(`2×8 (4 thr)` = "#00A087", `4×16 (4 thr)` = "#3C5488")
  ) +
  labs(
    title = "Speedup vs pilot default (4 chains × 8 CPUs)",
    subtitle = "Values > 1 mean faster than 4×8; both alternatives use 4 threads/chain",
    x = NULL, y = "Speedup (4×8 / setting)", fill = "Alternative"
  ) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```

```{r}
#| label: timing-summary-text
#| results: asis

med_by_setting <- timing |>
  group_by(setting_label) |>
  summarise(med_min = median(elapsed_min), .groups = "drop") |>
  arrange(med_min)

med_age <- median(timing$elapsed_min[timing$family == "age_smooth"])
med_re <- median(timing$elapsed_min[timing$family == "re_string"])

med_speedup <- timing_speedup |>
  group_by(alt_setting) |>
  summarise(med = median(speedup), .groups = "drop")

slowest <- timing |> arrange(desc(elapsed_min)) |> slice(1)
fastest <- timing |> arrange(elapsed_min) |> slice(1)

cat(glue(
  "Median elapsed: age-smooth **{round(med_age, 1)} min**, RE ",
  "**{round(med_re, 1)} min**. Fastest resource setting by median time: ",
  "**{med_by_setting$setting_label[1]}** ({round(med_by_setting$med_min[1], 1)} min). ",
  "Median speedup vs 4×8: **{round(med_speedup$med[med_speedup$alt_setting == '2×8 (4 thr)'], 2)}×** ",
  "for 2×8 and **{round(med_speedup$med[med_speedup$alt_setting == '4×16 (4 thr)'], 2)}×** ",
  "for 4×16. Slowest cell: **{slowest$gene} / {slowest$formula_id} / ",
  "{slowest$setting_label}** ({round(slowest$elapsed_min, 1)} min); ",
  "fastest: **{fastest$gene} / {fastest$formula_id} / {fastest$setting_label}** ",
  "({round(fastest$elapsed_min, 1)} min). Within age-smooth, k has little systematic ",
  "effect compared with threads-per-chain."
))
```

------------------------------------------------------------------------

# Settings

- **Age-smooth (report §§1–4):** healthy `Normal` only; 4 chains × 400 warmup / 900 iter;
  `cores = 8`, `threading(2)`, `control = NULL` (brms defaults).  
- **Strong RE reference (§4):** correlated tissue RE + simple shape;
  4 × 1500/3000, `adapt_delta = 0.95`, full disease mix.  
- **formula_bench (§5):** healthy-only; same 400/900 MCMC; formulas AGE-k3/4/5,
  RE-simple, RE-split; settings 2×8 / 4×8 / 4×16 (threads = floor(CPUs/chains)).  
- **Displayed full age curves:** `posterior_linpred(transform = FALSE,
  re_formula = NA)`, `offset = 0`, with ethnicity and assay fixed to their
  modal training values.  
- **Displayed brms smooths:** `conditional_smooths(prob = 0.95)` on the same
  tissue–sex-specific 5th–95th percentile age grids; these are smooth-term
  contributions, not full linear predictors.  
- **Baseline estimand:** comprehensive healthy liver `log(μ)` via
  `posterior_linpred` (smooth: age quantile grid × covariate margins).

------------------------------------------------------------------------

# Saved artefacts

| File | Description |
|------|-------------|
| `r params$diagnostics_csv` | Convergence + baseline summaries |
| `r params$curves_rds` | Full population-level blood/liver age curves |
| `r params$conditional_smooths_rds` | `conditional_smooths()` term-only curves |
| `r params$compat_csv` | Predict/linpred smoke tests |
| `r params$baseline_draws_rds` | Baseline draws for density overlays |
| `r params$timing_summary_csv` | formula_bench wall-clock times + max R̂ |
| Smooth RDS | `r params$smooth_out_dir`/`fit_age_smooth_hmc_*.rds` |
| RE RDS | `r params$strong_out_dir`/`fit_re_*.rds` |
| Bench fits | `data/processed/formula_bench/fits/fit_*__*__*.rds` |

Re-render with refreshed caches:

```bash
cd /home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC
quarto render Jie_HCC_age_smooth_k_grid_eval.qmd -P refresh:true
```
