---
title: "HCC formula comparison — HMC + Pathfinder"
subtitle: "Production, strong HMC, and Pathfinder RE variants on four TIMES biomarkers"
date: today
date-format: "DD MMM YYYY"
title-block-banner: true
author:
  - name: Chen Zhan
    orcid: 0000-0002-4794-8339
    email: chen.zhan@adelaide.edu.au
    affiliation:
      - id: sgc
        name: SAiGENCI, The University of Adelaide
format:
  html:
    theme: minty
    df-print: paged
    code-line-numbers: true
    embed-resources: true
    code-fold: true
    code-fold-default: true
    toc: true
    toc-location: left
    toc-title: "Contents"
knitr:
  opts_chunk:
    message: false
    warning: false
editor: source
abstract: |
  Gene-wise comparison of six V1_nk tissue-formula / inference variants for the
  TIMES HCC biomarkers (SPON2, ZFP36L2, ZFP36, VIM): production HMC (RE rich),
  three strong-HMC simple-shape refits (correlated RE, interaction, uncorrelated
  RE), and two CPU Pathfinder dense fits (original RE and split-ethnicity RE).
  For each gene × variant we report convergence-style diagnostics and the
  comprehensive healthy liver baseline mean and SD of log(μ). Pathfinder cells
  use a draw-count target matched to strong HMC (6000). For non-production strong
  HMC fits we also trace how max R̂, baseline μ, and baseline SD change as
  post-warmup draws accumulate, and report a minimum-draw stability threshold.
  Rendering reads cached fits only; set `refresh_diagnostics: true` after new
  fits land (also refreshes the HMC draw-stability cache).
params:
  case_dir: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC"
  model_store_root: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/HPC_posterior/NEW_CELL_TYPE"
  target_cell_type: "nk"
  strong_out_dir: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/re_vs_interaction_strong"
  diagnostics_csv: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/re_vs_interaction_strong/formula_comparison_diagnostics.csv"
  draws_rds: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/re_vs_interaction_strong/formula_comparison_baseline_draws.rds"
  hmc_stability_rds: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/re_vs_interaction_strong/hmc_draw_stability.rds"
  refresh_diagnostics: false
  # NULL = all post-warmup draws for baseline mean/SD and density overlays
  # (must match the draw-stability endpoint).
  n_baseline_draws: null
---

# Motivation

We compare **six** formula / inference settings on the same four TIMES genes
and the same comprehensive healthy liver baseline estimand:

1. **Convergence / reliability** — HMC R̂, divergences, treedepth, ESS; for
   Pathfinder, draw count and finite baseline summaries (no multi-chain HMC gates)  
2. **Baseline mean** — posterior mean of comprehensive liver `log(μ)`  
3. **Baseline SD** — posterior uncertainty of that estimand  

| Label | Counts formula | Shape | Inference |
|-------|----------------|-------|-----------|
| Production | correlated RE `(\| tissue)` | rich | HMC (atlas) |
| RE HMC | correlated RE `(\| tissue)` | simple | HMC 4×1500/3000 |
| Interaction HMC | fixed tissue × covariate | simple | HMC 4×1500/3000 |
| RE uncorrelated HMC | uncorrelated RE `(\|\| tissue)` | simple | HMC 4×1500/3000 |
| Pathfinder RE | correlated RE (same as RE HMC) | simple | Pathfinder CPU dense |
| Pathfinder RE-split | `(1+age×sex\|tissue) + (0+ethnicity\|tissue)` | simple | Pathfinder CPU dense |

Pathfinder settings follow the successful pilot (CPU, dense, not OpenCL):
`num_paths = 50`, `single_path_draws = 120` → **6000** draws targeted to match
strong HMC post-warmup draws (`4 × 1500`).

------------------------------------------------------------------------

# Library setup

```{r}
#| label: setup

suppressPackageStartupMessages({
  library(tibble)
  library(tidyr)
  library(dplyr)
  library(readr)
  library(ggplot2)
  library(forcats)
  library(scales)
  library(glue)
  library(stringr)
  library(posterior)
  library(brms)
  library(targets)
})

source(file.path(params$case_dir, "R", "re_vs_interaction_helpers.R"))
source(file.path(params$case_dir, "R", "fit_convergence_helpers.R"))

theme_set(theme_minimal(base_size = 11))

formula_levels <- c(
  "production", "re", "interaction", "re_uncorrelated",
  "pathfinder_re", "pathfinder_re_split"
)
formula_labels <- c(
  production = "Production (RE rich)",
  re = "RE HMC (simple)",
  interaction = "Interaction HMC (simple)",
  re_uncorrelated = "RE uncorrelated HMC (simple)",
  pathfinder_re = "Pathfinder RE (simple)",
  pathfinder_re_split = "Pathfinder RE-split (simple)"
)
formula_colours <- c(
  production = "#7A7A7A",
  re = "#4DBBD5",
  interaction = "#E64B35",
  re_uncorrelated = "#3C5488",
  pathfinder_re = "#00A087",
  pathfinder_re_split = "#91D1C2"
)
gate_colours <- c(PASS = "#2D8B57", WARN = "#E08A2E", FAIL = "#C23B4A")
fill_vals <- setNames(unname(formula_colours), unname(formula_labels))
```

------------------------------------------------------------------------

# Fit inventory

Expected grid: **4 genes × 6 formulas = 24 cells**.

```{r}
#| label: inventory

meta <- strong_formula_inventory(
  strong_out_dir = params$strong_out_dir,
  case_dir = params$case_dir,
  model_store_root = params$model_store_root,
  target_cell_type = params$target_cell_type
)

meta$inventory |>
  mutate(
    formula = factor(recode(model, !!!formula_labels), levels = unname(formula_labels)),
    status = if_else(available, source, "MISSING")
  ) |>
  select(gene, formula, role, shape_variant, status, file_name) |>
  arrange(gene, formula)
```

```{r}
#| label: inventory-check
#| results: asis

n_ok <- sum(meta$inventory$available)
n_tot <- nrow(meta$inventory)
n_pf <- meta$inventory |>
  filter(str_detect(model, "^pathfinder"), available) |>
  nrow()
if (n_ok < n_tot) {
  miss <- meta$inventory |> filter(!available)
  cat(glue(
    "**{n_ok}/{n_tot}** fits available ({n_pf} Pathfinder). Missing: ",
    "{paste(miss$gene, miss$model, sep = '/', collapse = ', ')}."
  ))
} else {
  cat(glue("All **{n_tot}** fits are available, including **{n_pf}** Pathfinder cells."))
}
```

------------------------------------------------------------------------

# Diagnostics

**HMC gates** (production + strong HMC cells):

| Gate | PASS rule |
|------|-----------|
| R̂ | `max_rhat ≤ 1.05` |
| Divergences | `< 1%` of post-warmup transitions |
| Treedepth | `< 10%` hit `max_treedepth` |
| Baseline chains | range of chain-mean liver `log(μ)` `< 1` |

**Pathfinder gates** (approximate; not multi-chain HMC): finite baseline mean/SD and
`n_draws ≥ 100`. R̂ / divergence / treedepth columns are shown when available but
are **not** used to FAIL Pathfinder cells.

`FAIL` (HMC only) = `max_rhat > 1.5` or `≥ 50%` divergences; else soft → `WARN`.

**Draw consistency.** Baseline mean/SD use **all** post-warmup draws with a
fixed linpred seed (`20260722`; `sample_new_levels = "gaussian"` is otherwise
RNG). Max R̂ is over **all** parameters via the same quantity as
`posterior::rhat(fit)`. Draw-stability endpoints for the 12 strong-HMC cells
are checked against this table below.

```{r}
#| label: diagnostics
#| cache: false

diag_path <- params$diagnostics_csv
draws_path <- params$draws_rds

# Refresh if forced, missing, or inventory has models absent from the cache
# (e.g. Pathfinder fits landed after the last diagnostics run).
need_refresh <- isTRUE(params$refresh_diagnostics) || !file.exists(diag_path)
if (!need_refresh && file.exists(diag_path)) {
  old <- read_csv(diag_path, show_col_types = FALSE)
  avail_models <- meta$inventory |>
    filter(available) |>
    distinct(gene, model)
  cached <- old |> distinct(gene, model)
  need_refresh <- nrow(
    avail_models |> anti_join(cached, by = c("gene", "model"))
  ) > 0
}

if (need_refresh) {
  message("Refreshing diagnostics (includes Pathfinder if present) ...")
  diag_bundle <- diagnose_strong_formula_fits(
    strong_out_dir = params$strong_out_dir,
    case_dir = params$case_dir,
    model_store_root = params$model_store_root,
    target_cell_type = params$target_cell_type,
    compute_baseline = TRUE,
    n_baseline_draws = params$n_baseline_draws,
    cache_csv = diag_path
  )
  diag <- diag_bundle$diagnostics
  # Force draws refresh whenever diagnostics refresh
  if (file.exists(draws_path)) file.remove(draws_path)
} else {
  diag <- read_csv(diag_path, show_col_types = FALSE)
}

diag_view <- diag |>
  mutate(
    formula = factor(
      recode(model, !!!formula_labels),
      levels = unname(formula_labels)
    ),
    is_pathfinder = coalesce(is_pathfinder, str_detect(model, "^pathfinder")),
    engine = if_else(is_pathfinder, "Pathfinder", "HMC"),
    gate = case_when(
      gate_pass ~ "PASS",
      is_pathfinder ~ "WARN",
      max_rhat > 1.5 | pct_divergent >= 50 ~ "FAIL",
      TRUE ~ "WARN"
    ),
    gate = factor(gate, levels = c("PASS", "WARN", "FAIL")),
    settings = if_else(
      is_pathfinder,
      glue("pathfinder; n_draws={coalesce(n_draws, post_warmup_per_chain)}"),
      glue("{chains}×{warmup}/{iter}; δ={adapt_delta}; td={max_treedepth}")
    )
  )
```

## Full table (convergence + baseline)

```{r}
#| label: diagnostics-table

diag_view |>
  transmute(
    gene,
    formula,
    engine,
    source,
    settings,
    max_rhat = round(max_rhat, 3),
    n_rhat_gt_1_1,
    pct_divergent = round(pct_divergent, 2),
    pct_max_treedepth = round(pct_max_treedepth, 1),
    min_ess_bulk = round(min_ess_bulk, 0),
    n_draws,
    baseline_n_draws,
    baseline_mean = round(baseline_mean, 3),
    baseline_sd = round(baseline_sd, 3),
    chain_mean_range = round(baseline_chain_mean_range, 3),
    gate
  ) |>
  arrange(gene, formula)
```

## Gate summary by formula

```{r}
#| label: gate-summary

diag_view |>
  count(formula, engine, gate) |>
  pivot_wider(names_from = gate, values_from = n, values_fill = 0) |>
  arrange(formula)
```

------------------------------------------------------------------------

# 1. Convergence

```{r}
#| label: plot-max-rhat
#| fig-width: 10
#| fig-height: 5
#| fig-align: center

diag_hmc <- diag_view |> filter(!is_pathfinder)

ggplot(diag_hmc, aes(x = gene, y = max_rhat, fill = formula)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  geom_hline(yintercept = 1.05, linetype = 2, colour = "grey40") +
  geom_hline(yintercept = 1.1, linetype = 3, colour = "grey60") +
  scale_fill_manual(values = fill_vals, drop = FALSE) +
  labs(
    title = "Maximum R̂ (HMC cells only)",
    subtitle = "Pathfinder omitted — single-path draws are not multi-chain HMC",
    x = NULL, y = "max R̂", fill = NULL
  ) +
  theme(legend.position = "bottom")
```

```{r}
#| label: plot-divergences
#| fig-width: 10
#| fig-height: 5
#| fig-align: center

ggplot(diag_hmc, aes(x = gene, y = pct_divergent, fill = formula)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey40") +
  scale_fill_manual(values = fill_vals, drop = FALSE) +
  labs(
    title = "Divergent transitions (% of post-warmup, HMC only)",
    subtitle = "Dashed = 1% PASS gate",
    x = NULL, y = "% divergent", fill = NULL
  ) +
  theme(legend.position = "bottom")
```

```{r}
#| label: plot-ess-draws
#| fig-width: 10
#| fig-height: 5
#| fig-align: center

# HMC: min bulk ESS; Pathfinder: reported n_draws (approximation quality proxy)
ess_plot <- diag_view |>
  mutate(
    metric = if_else(is_pathfinder, as.numeric(n_draws), as.numeric(min_ess_bulk)),
    metric_label = if_else(is_pathfinder, "n_draws (Pathfinder)", "min ESS bulk (HMC)")
  )

ggplot(ess_plot, aes(x = gene, y = metric, fill = formula)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  facet_wrap(~ metric_label, scales = "free_y") +
  scale_fill_manual(values = fill_vals, drop = FALSE) +
  labs(
    title = "Effective sample size (HMC) vs Pathfinder draw count",
    subtitle = "Pathfinder target was 6000 draws (50 paths × 120); actual n_draws may differ",
    x = NULL, y = NULL, fill = NULL
  ) +
  theme(legend.position = "bottom")
```

```{r}
#| label: convergence-text
#| results: asis

gate_by_formula <- diag_view |>
  group_by(formula, engine) |>
  summarise(
    n = n(),
    n_pass = sum(gate == "PASS"),
    n_warn = sum(gate == "WARN"),
    n_fail = sum(gate == "FAIL"),
    median_rhat = median(max_rhat, na.rm = TRUE),
    median_n_draws = median(n_draws, na.rm = TRUE),
    .groups = "drop"
  )

worst_hmc <- diag_view |>
  filter(!is_pathfinder) |>
  arrange(desc(max_rhat)) |>
  slice(1)

cat(glue(
  "Among HMC cells, worst max R̂ is **{round(worst_hmc$max_rhat, 3)}** for ",
  "**{worst_hmc$gene} / {worst_hmc$formula}** (gate = {worst_hmc$gate}).\n\n"
))
knitr::kable(gate_by_formula |> mutate(across(where(is.numeric), ~ round(.x, 3))))
```

------------------------------------------------------------------------

# 2. Baseline mean (comprehensive liver log μ)

```{r}
#| label: plot-baseline-mean
#| fig-width: 10
#| fig-height: 5
#| fig-align: center

ggplot(diag_view, aes(x = gene, y = baseline_mean, fill = formula)) +
  geom_col(position = position_dodge(width = 0.85), width = 0.8) +
  scale_fill_manual(values = fill_vals, drop = FALSE) +
  labs(
    title = "Comprehensive healthy liver baseline mean",
    subtitle = "Posterior mean of log(μ); includes Pathfinder RE / RE-split",
    x = NULL, y = "mean baseline log(μ)", fill = NULL
  ) +
  theme(legend.position = "bottom")
```

```{r}
#| label: baseline-mean-vs-production

mean_vs_prod <- diag_view |>
  select(gene, model, baseline_mean) |>
  pivot_wider(names_from = model, values_from = baseline_mean) |>
  mutate(
    across(
      any_of(c(
        "re", "interaction", "re_uncorrelated",
        "pathfinder_re", "pathfinder_re_split"
      )),
      ~ .x - production,
      .names = "delta_{.col}"
    )
  )

mean_vs_prod |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  arrange(gene)
```

### Pathfinder vs matched HMC RE (same counts formula)

```{r}
#| label: pathfinder-vs-hmc-re-mean

pf_vs_re <- diag_view |>
  filter(model %in% c("re", "pathfinder_re", "pathfinder_re_split")) |>
  select(gene, model, baseline_mean, baseline_sd, gate) |>
  pivot_wider(
    names_from = model,
    values_from = c(baseline_mean, baseline_sd, gate)
  ) |>
  mutate(
    delta_pf_re = baseline_mean_pathfinder_re - baseline_mean_re,
    delta_pf_split = baseline_mean_pathfinder_re_split - baseline_mean_re
  )

pf_vs_re |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  arrange(gene)
```

------------------------------------------------------------------------

# 3. Baseline SD

```{r}
#| label: plot-baseline-sd
#| fig-width: 10
#| fig-height: 5
#| fig-align: center

ggplot(diag_view, aes(x = gene, y = baseline_sd, fill = formula)) +
  geom_col(position = position_dodge(width = 0.85), width = 0.8) +
  scale_fill_manual(values = fill_vals, drop = FALSE) +
  labs(
    title = "Comprehensive healthy liver baseline SD",
    subtitle = "Lower = tighter healthy envelope for HCC proximity scores",
    x = NULL, y = "SD of baseline log(μ)", fill = NULL
  ) +
  theme(legend.position = "bottom")
```

```{r}
#| label: baseline-sd-vs-production

sd_vs_prod <- diag_view |>
  select(gene, model, baseline_sd) |>
  pivot_wider(names_from = model, values_from = baseline_sd) |>
  mutate(
    across(
      any_of(c(
        "re", "interaction", "re_uncorrelated",
        "pathfinder_re", "pathfinder_re_split"
      )),
      ~ .x / production,
      .names = "ratio_{.col}"
    )
  )

sd_vs_prod |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  arrange(gene)
```

```{r}
#| label: sd-text
#| results: asis

narrowest <- diag_view |>
  group_by(gene) |>
  slice_min(baseline_sd, n = 1, with_ties = FALSE) |>
  ungroup()

cat("Narrowest baseline SD per gene:\n\n")
knitr::kable(
  narrowest |>
    transmute(gene, formula, engine, baseline_sd = round(baseline_sd, 3), gate)
)
```

------------------------------------------------------------------------

# 4. HMC draw stability (non-production)

Production atlas fits are short (typically 2 × 200 post-warmup) and Pathfinder
is not multi-chain HMC, so this section uses only the three **strong HMC**
variants: RE, interaction, and RE-uncorrelated (4 genes × 3 = 12 fits).

**Method.** For each fit, take the first $k$ post-warmup iterations from every
chain ($k$ on a grid up to 1500 → $n_{\mathrm{draws}} = 4k$). At each $k$:

1. **Convergence** — max R̂ over **all** parameters (same as the main diagnostics table)  
2. **Baseline μ** — mean of comprehensive liver `log(μ)` over those draws  
3. **Baseline SD** — SD of the same estimand  

The $k = 1500$ endpoint therefore matches the main-table baseline and max R̂ for
these 12 cells.  

**Stability rule** (relative to the full 6000-draw endpoint):

| Metric | Stable when… |
|--------|----------------|
| Baseline μ | within `max(0.05, 2% × |μ_full|)` of μ_full and stays there |
| Baseline SD | within `max(0.05, 5% × SD_full)` of SD_full and stays there |
| max R̂ | within 0.02 of R̂_full **and** ≤ 1.05, and stays there |

The reported **minimum draws** for a fit is the first $n$ where **all three**
criteria hold thereafter (`n_draws_all_stable`).

```{r}
#| label: hmc-draw-stability
#| cache: false

stab_path <- params$hmc_stability_rds
need_stab <- isTRUE(params$refresh_diagnostics) || !file.exists(stab_path)

if (need_stab) {
  message("Computing HMC cumulative draw-stability curves ...")
  stab <- collect_hmc_draw_stability(
    strong_out_dir = params$strong_out_dir,
    case_dir = params$case_dir,
    model_store_root = params$model_store_root,
    target_cell_type = params$target_cell_type,
    cache_rds = stab_path,
    force = TRUE
  )
} else {
  stab <- readRDS(stab_path)
}

stab_curves <- stab$curves |>
  mutate(
    formula = factor(
      recode(model, !!!formula_labels),
      levels = unname(formula_labels)
    )
  ) |>
  group_by(gene, model) |>
  mutate(
    mean_final = dplyr::last(baseline_mean),
    sd_final = dplyr::last(baseline_sd),
    rhat_final = dplyr::last(max_rhat),
    mean_delta = baseline_mean - mean_final,
    sd_ratio = baseline_sd / sd_final,
    rhat_delta = max_rhat - rhat_final
  ) |>
  ungroup()

stab_thresholds <- stab$thresholds |>
  mutate(
    formula = factor(
      recode(model, !!!formula_labels),
      levels = unname(formula_labels)
    )
  )
```

## Stability thresholds (minimum draws)

```{r}
#| label: hmc-stability-threshold-table

stab_thresholds |>
  transmute(
    gene,
    formula,
    n_draws_full,
    max_rhat_full = round(max_rhat_full, 3),
    baseline_mean_full = round(baseline_mean_full, 3),
    baseline_sd_full = round(baseline_sd_full, 3),
    n_draws_rhat_stable,
    n_draws_mean_stable,
    n_draws_sd_stable,
    n_draws_all_stable
  ) |>
  arrange(gene, formula)
```

```{r}
#| label: hmc-stability-threshold-text
#| results: asis

ok <- stab_thresholds |> filter(!is.na(n_draws_all_stable))
n_miss <- nrow(stab_thresholds) - nrow(ok)
if (nrow(ok)) {
  med_all <- median(ok$n_draws_all_stable)
  max_row <- ok |> slice_max(n_draws_all_stable, n = 1, with_ties = FALSE)
  cat(glue(
    "**{nrow(ok)}/{nrow(stab_thresholds)}** strong-HMC fits reach the joint ",
    "stability criterion (median **{med_all}** draws). ",
    "Highest requirement: **{max_row$n_draws_all_stable}** for ",
    "**{max_row$gene} / {max_row$formula}** ",
    "(R̂@{max_row$n_draws_rhat_stable}, μ@{max_row$n_draws_mean_stable}, ",
    "SD@{max_row$n_draws_sd_stable})."
  ))
  if (n_miss > 0) {
    miss <- stab_thresholds |> filter(is.na(n_draws_all_stable))
    cat(glue(
      "\n\n**{n_miss}** fits never satisfy the joint rule — typically because ",
      "full-sample max R̂ stays above 1.05: ",
      "{paste(miss$gene, miss$model, sep = '/', collapse = ', ')}. ",
      "A median like {med_all} must **not** become a universal budget; ",
      "shortening runs is only safe gene-by-gene after R̂ / diagnostics are fixed."
    ))
  }
} else {
  cat("No joint stability thresholds found under the current tolerances.")
}
```

### Diagnostics ↔ stability endpoint check

```{r}
#| label: hmc-stability-reconcile

reconcile <- stab_thresholds |>
  select(
    gene, model,
    stab_max_rhat = max_rhat_full,
    stab_mean = baseline_mean_full,
    stab_sd = baseline_sd_full,
    stab_n = n_draws_full
  ) |>
  left_join(
    diag_view |>
      filter(model %in% c("re", "interaction", "re_uncorrelated")) |>
      select(
        gene, model,
        diag_max_rhat = max_rhat,
        diag_mean = baseline_mean,
        diag_sd = baseline_sd,
        diag_n = baseline_n_draws
      ),
    by = c("gene", "model")
  ) |>
  mutate(
    delta_rhat = stab_max_rhat - diag_max_rhat,
    delta_mean = stab_mean - diag_mean,
    delta_sd = stab_sd - diag_sd,
    ok = abs(delta_rhat) < 1e-6 &
      abs(delta_mean) < 1e-6 &
      abs(delta_sd) < 1e-6 &
      stab_n == diag_n
  )

reconcile |>
  mutate(across(where(is.numeric), ~ round(.x, 6))) |>
  arrange(gene, model)
```

```{r}
#| label: hmc-stability-reconcile-text
#| results: asis

n_ok <- sum(reconcile$ok, na.rm = TRUE)
n_tot <- nrow(reconcile)
if (n_ok == n_tot) {
  cat(glue(
    "All **{n_tot}** strong-HMC cells match between the main diagnostics table ",
    "and the draw-stability full-draw endpoint (max R̂, baseline μ, SD, n draws)."
  ))
} else {
  bad <- reconcile |> filter(!ok | is.na(ok))
  cat(glue(
    "**{n_ok}/{n_tot}** cells match. Mismatches remain for: ",
    "{paste(bad$gene, bad$model, sep = '/', collapse = ', ')}. ",
    "Re-run with `refresh_diagnostics: true` after helper changes."
  ))
}
```

## Trace: max R̂ vs draws

```{r}
#| label: plot-hmc-rhat-vs-draws
#| fig-width: 10
#| fig-height: 7
#| fig-align: center

ggplot(stab_curves, aes(x = n_draws, y = max_rhat, colour = formula)) +
  geom_hline(yintercept = 1.05, linetype = 2, colour = "grey40") +
  geom_hline(yintercept = 1.1, linetype = 3, colour = "grey60") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.4) +
  facet_wrap(~ gene, scales = "free_y") +
  scale_colour_manual(values = fill_vals, drop = TRUE) +
  scale_x_continuous(labels = label_comma()) +
  labs(
    title = "Cumulative max R̂ as post-warmup draws increase",
    subtitle = "Strong HMC only (RE / interaction / RE-uncorrelated); dashed = 1.05 PASS gate",
    x = "Cumulative draws (4 chains × first k iterations)",
    y = "max R̂",
    colour = NULL
  ) +
  theme(legend.position = "bottom")
```

## Trace: baseline μ vs draws

```{r}
#| label: plot-hmc-mean-vs-draws
#| fig-width: 10
#| fig-height: 7
#| fig-align: center

ggplot(stab_curves, aes(x = n_draws, y = baseline_mean, colour = formula)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.4) +
  facet_wrap(~ gene, scales = "free_y") +
  scale_colour_manual(values = fill_vals, drop = TRUE) +
  scale_x_continuous(labels = label_comma()) +
  labs(
    title = "Cumulative baseline mean as draws increase",
    subtitle = "Comprehensive liver log(μ); curves should flatten toward the full-sample mean",
    x = "Cumulative draws",
    y = "mean baseline log(μ)",
    colour = NULL
  ) +
  theme(legend.position = "bottom")
```

```{r}
#| label: plot-hmc-mean-delta-vs-draws
#| fig-width: 10
#| fig-height: 7
#| fig-align: center

ggplot(stab_curves, aes(x = n_draws, y = mean_delta, colour = formula)) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey50") +
  geom_hline(yintercept = c(-0.05, 0.05), linetype = 3, colour = "grey70") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.4) +
  facet_wrap(~ gene, scales = "free_y") +
  scale_colour_manual(values = fill_vals, drop = TRUE) +
  scale_x_continuous(labels = label_comma()) +
  labs(
    title = "Baseline mean − full-sample mean",
    subtitle = "Dotted bands ≈ ±0.05 absolute tolerance used in the stability rule",
    x = "Cumulative draws",
    y = "Δ mean log(μ)",
    colour = NULL
  ) +
  theme(legend.position = "bottom")
```

## Trace: baseline SD vs draws

```{r}
#| label: plot-hmc-sd-vs-draws
#| fig-width: 10
#| fig-height: 7
#| fig-align: center

ggplot(stab_curves, aes(x = n_draws, y = baseline_sd, colour = formula)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.4) +
  facet_wrap(~ gene, scales = "free_y") +
  scale_colour_manual(values = fill_vals, drop = TRUE) +
  scale_x_continuous(labels = label_comma()) +
  labs(
    title = "Cumulative baseline SD as draws increase",
    subtitle = "Early draws often understate SD; look for flattening",
    x = "Cumulative draws",
    y = "SD of baseline log(μ)",
    colour = NULL
  ) +
  theme(legend.position = "bottom")
```

```{r}
#| label: plot-hmc-sd-ratio-vs-draws
#| fig-width: 10
#| fig-height: 7
#| fig-align: center

ggplot(stab_curves, aes(x = n_draws, y = sd_ratio, colour = formula)) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey50") +
  geom_hline(yintercept = c(0.95, 1.05), linetype = 3, colour = "grey70") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.4) +
  facet_wrap(~ gene, scales = "free_y") +
  scale_colour_manual(values = fill_vals, drop = TRUE) +
  scale_x_continuous(labels = label_comma()) +
  labs(
    title = "Baseline SD / full-sample SD",
    subtitle = "Dotted = ±5% relative band used in the stability rule",
    x = "Cumulative draws",
    y = "SD ratio",
    colour = NULL
  ) +
  theme(legend.position = "bottom")
```

```{r}
#| label: plot-hmc-stability-bars
#| fig-width: 10
#| fig-height: 5
#| fig-align: center

thr_long <- stab_thresholds |>
  select(
    gene, formula,
    Rhat = n_draws_rhat_stable,
    Mean = n_draws_mean_stable,
    SD = n_draws_sd_stable,
    All = n_draws_all_stable
  ) |>
  pivot_longer(c(Rhat, Mean, SD, All), names_to = "metric", values_to = "n_draws") |>
  mutate(metric = factor(metric, levels = c("Rhat", "Mean", "SD", "All")))

ggplot(thr_long, aes(x = gene, y = n_draws, fill = formula)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  facet_wrap(~ metric, nrow = 1) +
  scale_fill_manual(values = fill_vals, drop = TRUE) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title = "Minimum cumulative draws to stability",
    subtitle = "All = max of R̂ / mean / SD thresholds (NA if criterion never holds)",
    x = NULL, y = "n draws", fill = NULL
  ) +
  theme(legend.position = "bottom")
```

------------------------------------------------------------------------

# Baseline posterior densities

```{r}
#| label: baseline-draws
#| cache: false

need_draws <- isTRUE(params$refresh_diagnostics) ||
  need_refresh ||
  !file.exists(draws_path)

if (need_draws) {
  message("Collecting baseline draws (includes Pathfinder) ...")
  draws_df <- collect_strong_baseline_draws(
    strong_out_dir = params$strong_out_dir,
    case_dir = params$case_dir,
    model_store_root = params$model_store_root,
    target_cell_type = params$target_cell_type,
    n_draws = params$n_baseline_draws,
    cache_rds = draws_path
  )
} else {
  draws_df <- readRDS(draws_path)
}

draws_labelled <- draws_df |>
  mutate(
    formula = factor(
      recode(model, !!!formula_labels),
      levels = unname(formula_labels)
    ),
    is_pathfinder = str_detect(model, "^pathfinder")
  )
```

```{r}
#| label: plot-baseline-density
#| fig-width: 10
#| fig-height: 7
#| fig-align: center

ggplot(draws_labelled, aes(x = baseline_log_mu, colour = formula)) +
  geom_density(linewidth = 0.8) +
  facet_wrap(~ gene, scales = "free") +
  scale_colour_manual(values = fill_vals, drop = FALSE) +
  labs(
    title = "Comprehensive liver baseline posterior by formula",
    subtitle = "All post-warmup draws (same estimand as diagnostics / stability endpoint)",
    x = "baseline log(μ)", y = "Density", colour = NULL
  ) +
  theme(legend.position = "bottom")
```

```{r}
#| label: plot-baseline-density-re-family
#| fig-width: 10
#| fig-height: 7
#| fig-align: center

re_family <- draws_labelled |>
  filter(model %in% c("production", "re", "pathfinder_re", "pathfinder_re_split"))

ggplot(re_family, aes(x = baseline_log_mu, colour = formula)) +
  geom_density(linewidth = 0.85) +
  facet_wrap(~ gene, scales = "free") +
  scale_colour_manual(values = fill_vals, drop = TRUE) +
  labs(
    title = "RE-family only: production vs HMC RE vs Pathfinder",
    subtitle = "Isolates Pathfinder approximation error relative to strong HMC RE",
    x = "baseline log(μ)", y = "Density", colour = NULL
  ) +
  theme(legend.position = "bottom")
```

------------------------------------------------------------------------

# Settings used

```{r}
#| label: settings-table

diag_view |>
  distinct(
    formula, engine, gene, source, algorithm,
    chains, warmup, iter, n_draws, adapt_delta, max_treedepth, seed
  ) |>
  arrange(formula, gene)
```

- **Production HMC:** atlas `V1_nk` defaults (typically 2 × 400/600).  
- **Strong HMC:** 4 chains × 1500 warmup / 3000 iter, `adapt_delta = 0.95`,
  `max_treedepth = 12`, 4 × 8 threads.  
- **Pathfinder:** CPU dense, `opencl = NULL`, `sparse = FALSE`, `num_paths = 50`,
  `single_path_draws = 120` (target 6000 draws), `threads = 8`,
  `psis_resample = FALSE`.

------------------------------------------------------------------------

# Interpretation

```{r}
#| label: interpretation
#| results: asis

n_pass <- sum(diag_view$gate == "PASS")
n_warn <- sum(diag_view$gate == "WARN")
n_fail <- sum(diag_view$gate == "FAIL")
n_pf <- sum(diag_view$is_pathfinder)

mean_sd <- diag_view |>
  group_by(formula) |>
  summarise(mean_sd = mean(baseline_sd, na.rm = TRUE), .groups = "drop") |>
  arrange(mean_sd)

best_sd <- mean_sd$formula[1]
worst_sd <- mean_sd$formula[nrow(mean_sd)]

pf_shift <- diag_view |>
  filter(model %in% c("re", "pathfinder_re")) |>
  select(gene, model, baseline_mean) |>
  pivot_wider(names_from = model, values_from = baseline_mean) |>
  mutate(abs_shift = abs(pathfinder_re - re))

stab_ok <- stab_thresholds |> filter(!is.na(n_draws_all_stable))
stab_note <- if (nrow(stab_ok)) {
  glue(
    " Strong-HMC joint draw-stability median = **{median(stab_ok$n_draws_all_stable)}** ",
    "(of {nrow(stab_ok)}/{nrow(stab_thresholds)} fits)."
  )
} else {
  " Strong-HMC joint draw-stability thresholds were not identified under current tolerances."
}

cat(glue(
  "Across {nrow(diag_view)} gene × formula cells ({n_pf} Pathfinder): ",
  "**{n_pass} PASS**, **{n_warn} WARN**, **{n_fail} FAIL**. ",
  "Tightest mean baseline SD: **{best_sd}** ({round(mean_sd$mean_sd[1], 3)}); ",
  "widest: **{worst_sd}** ({round(mean_sd$mean_sd[nrow(mean_sd)], 3)}). ",
  "Median |Pathfinder RE − HMC RE| baseline mean shift: ",
  "**{round(median(pf_shift$abs_shift, na.rm = TRUE), 3)}** log units.",
  "{stab_note}"
))
```

**How to read Pathfinder in this report**

1. **Convergence** — do not treat Pathfinder R̂ like HMC; use the Pathfinder-specific
   gate and compare baselines to the strong HMC RE reference.  
2. **Baseline mean** — Pathfinder is useful if it tracks HMC RE closely (small
   `delta_pf_re`); large shifts mean the approximation is not yet a drop-in.  
3. **Baseline SD** — Pathfinder can understate or overstate uncertainty; prefer
   HMC SD for proximity scores unless Pathfinder SD is validated gene-wise.  
4. **RE-split** — ask whether splitting ethnicity slopes buys speed/stability
   without moving the liver baseline relative to Pathfinder original RE.

**How to read HMC draw stability**

1. Curves use **prefix draws** (first $k$ iterations × all chains), not random
   subsets — early R̂ can look worse than the full run.  
2. Prefer the **joint** threshold (`n_draws_all_stable`) over single metrics;
   SD often stabilises later than the mean.  
3. **Do not treat the median threshold as a universal budget.** Fits that never
   clear R̂ ≤ 1.05, or genes whose baseline mean drifts slowly (e.g. high
   `n_draws_mean_stable`), need the full strong-HMC length (or a longer one)
   until diagnostics are fixed; shortening is only safe gene-by-gene.  
4. Endpoint values are reconciled with the main diagnostics table (same all-draw
   baseline and all-parameter max R̂).

------------------------------------------------------------------------

# Saved artefacts

| File | Description |
|------|-------------|
| `r params$strong_out_dir`/`fit_*.rds` | Strong HMC fits |
| `r params$strong_out_dir`/`fit_pathfinder_*.rds` | Pathfinder RE / RE-split |
| `r params$diagnostics_csv` | Cached diagnostics (auto-refreshes when new models appear) |
| `r params$draws_rds` | Cached baseline draws |
| `r params$hmc_stability_rds` | Cached HMC cumulative draw-stability curves + thresholds |
| Production | `V1_nk` `estimates_chunk_*` |
| Pathfinder store | `/hpcfs/.../re_vs_interaction_pathfinder/_targets` |
