---
title: "HCC case study — frozen posteriorHCA baseline transfer to GeoMx"
subtitle: "Source-feature-excluded liver-NK reference retrieval on an unseen spatial platform"
date: today
date-format: "DD MMM YYYY"
title-block-banner: true
author:
  - name: Chen Zhan
    orcid: 0000-0002-4794-8339
    email: chen.zhan@adelaide.edu.au
    affiliation:
      - id: sgc
        name: SAiGENCI, The University of Adelaide
format:
  html:
    theme: minty
    df-print: paged
    code-line-numbers: true
    embed-resources: true
    code-fold: true
    code-fold-default: true
    toc: true
    toc-location: left
    toc-title: "Contents"
knitr:
  opts_chunk:
    message: false
    warning: false
editor: source
abstract: |
  This report tests one question: can a frozen posteriorHCA healthy-liver NK
  baseline be transported to GeoMx, a spatial assay that was not represented
  when posteriorHCA was fitted? The external observations are within-patient
  tumour-centre contrasts between protein-defined NK-selected and cold-tumour
  regions from six HCC patients. The validation excludes genes selected by the
  source study, the TIMES biomarkers and the transcripts used to define the
  regions. A successful result requires the posteriorHCA liver-NK reference to
  match the GeoMx contrast better than six alternative cell references and to
  score highly under exhaustive reallocations of the spatial labels. A
  dataset-balanced, unmatched empirical HCA centroid is included only as a
  descriptive sensitivity analysis. The estimand is
  relative cross-gene structure; this report does not calibrate posteriorHCA
  counts to GeoMx Q3-normalized expression.
params:
  case_dir: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC"
  model_store_root: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/HPC_posterior/NEW_CELL_TYPE"
  output_dir: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/spatial_transfer_validation"
  baseline_tissue_group: "liver"
  target_cell_type: "nk"
  candidate_cell_types:
    - "nk"
    - "cd8.tem"
    - "mait"
    - "tgd"
    - "monocytic"
    - "macrophage"
    - "epithelial"
  n_posterior_draws: 400
  pass_maximum_rhat: 1.05
  pass_maximum_divergence_percent: 1
  fail_maximum_rhat: 1.5
  fail_minimum_divergence_percent: 50
  joint_null_draws: 100000
  seed: 20260812
  refresh_references: false
---

# Analysis question and claim boundary

The analysis asks whether a healthy cell-state reference learned from atlas
pseudobulk data retains its biological structure on an unseen spatial platform:

> Do genes predicted by frozen posteriorHCA models to be relatively enriched
> in healthy liver NK cells also increase in independently labelled
> NK-selected GeoMx tumour-centre regions, and is liver NK the best-matching
> reference among seven candidate cell types?

The outcome is **reference retrieval**, not equality of measurement scales.
Both sides of the comparison are contrasts:

$$
D_{gp} =
\operatorname{median}_{r \in \mathrm{NK\ selected\ TC}}
\{\log_2 Q3_{gpr}\}
-
\operatorname{median}_{r \in \mathrm{cold\ TC}}
\{\log_2 Q3_{gpr}\},
$$

and

$$
H_{gc} = \log\mu_{gc}
- \frac{1}{7}\sum_{c'=1}^{7}\log\mu_{gc'}.
$$

Here, $D_{gp}$ is the GeoMx contrast for gene $g$ and patient $p$;
$H_{gc}$ is the posteriorHCA specificity of candidate cell type $c$ from its
comprehensive healthy-liver grand-mean $\log\mu$. No GeoMx assay coefficient
is added to or estimated inside posteriorHCA.

The analysis can support transport of **relative cell-reference structure**.
It cannot establish absolute GeoMx calibration, predictive-interval coverage,
pure-cell expression or recurrence prediction.

Because the model stores are frozen, convergence diagnostics are audited but
not repaired here. The operational retrieval excludes catastrophic `FAIL`
fits and visibly retains intermediate `WARN` fits. It can be promoted from an
exploratory deployment result to strong inferential evidence only if a useful
gene universe also passes the strict diagnostic criteria in all seven models.

------------------------------------------------------------------------

# Setup

```{r}
#| label: setup

suppressPackageStartupMessages({
  library(tibble)
  library(tidyr)
  library(dplyr)
  library(readr)
  library(purrr)
  library(stringr)
  library(forcats)
  library(ggplot2)
  library(scales)
  library(targets)
  library(brms)
  library(posterior)
  library(AnnotationDbi)
  library(org.Hs.eg.db)
})

theme_set(theme_minimal(base_size = 11))
set.seed(params$seed)

candidate_cells <- unlist(params$candidate_cell_types)
target_cell <- params$target_cell_type
output_dir <- params$output_dir
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

core_rdata <- file.path(
  params$case_dir,
  "codeocean/data/Input_1_Spatial_transcriptomics_analysis",
  "transcriptome_differential_analysis_input_1.RData"
)

source_de_rdata <- file.path(
  params$case_dir,
  "codeocean/data/Input_1_Spatial_transcriptomics_analysis",
  "differential_gene_expression.RData"
)

cell_labels <- c(
  nk = "NK",
  `cd8.tem` = "CD8 TEM",
  mait = "MAIT",
  tgd = "gamma-delta T",
  monocytic = "Monocytic",
  macrophage = "Macrophage",
  epithelial = "Epithelial"
)

method_colours <- c(
  posteriorHCA = "#2878B5",
  `Naive empirical HCA` = "#E08A2E"
)
```

------------------------------------------------------------------------

# Helper functions

All data preparation, reference construction and validation helpers used by
the report are defined below. The report does not source external analysis
scripts.

## Data and contrast helpers

```{r}
#| label: data-helpers

load_rdata_environment <- function(path) {
  environment <- new.env(parent = emptyenv())
  load(path, envir = environment)
  environment
}

row_median <- function(x) {
  apply(x, 1, stats::median)
}

make_patient_contrast <- function(
    patient,
    genes,
    metadata,
    log_expression
) {
  patient_regions <- metadata |>
    filter(.data$patient_id == patient)

  nk_regions <- patient_regions$roi_key[
    patient_regions$region_group == "NK-selected TC"
  ]
  cold_regions <- patient_regions$roi_key[
    patient_regions$region_group == "Cold TC"
  ]

  tibble(
    gene = genes,
    patient_id = patient,
    panel = dplyr::first(patient_regions$panel),
    delta_q3 =
      row_median(log_expression[genes, nk_regions, drop = FALSE]) -
      row_median(log_expression[genes, cold_regions, drop = FALSE])
  )
}
```

## posteriorHCA reference helpers

These helpers follow the same comprehensive healthy-liver baseline used in the
universal baseline report: equal-weight grand mean of $\log\mu$ over the
factor grid, with a new `dataset_id` level drawn from the hierarchical prior.

```{r}
#| label: reference-helpers

model_store <- function(cell_type) {
  file.path(
    params$model_store_root,
    paste0("V1_", cell_type),
    "_targets"
  )
}

read_model_inventory <- function(cell_type) {
  store <- model_store(cell_type)

  targets::tar_meta(
    starts_with("estimates_chunk_"),
    store = store
  ) |>
    filter(.data$type == "branch") |>
    transmute(
      cell_type = cell_type,
      .feature = str_extract(
        .data$warnings,
        "(?<=Gene:___)ENSG\\d+(?=___)"
      ),
      target_name = .data$name,
      store = store
    ) |>
    filter(!is.na(.data$.feature))
}

read_model_fit <- function(target_name, store) {
  targets::tar_read_raw(target_name, store = store)$brms_fit[[1]]
}

fit_levels <- function(fit, var) {
  if (!var %in% names(fit$data)) {
    return(character(0))
  }

  values <- fit$data[[var]]
  lv <- if (is.factor(values)) {
    base::levels(values)
  } else {
    sort(unique(as.character(values)))
  }

  lv[!is.na(lv) & nzchar(lv)]
}

build_comprehensive_grid <- function(
    fit,
    tissue_group = params$baseline_tissue_group,
    marginalize_over = c(
      "sex",
      "age_decade",
      "ethnicity_groups",
      "assay_groups_altered"
    ),
    new_level_cols = "dataset_id_altered",
    new_level_tag = "__comprehensive_healthy_baseline__"
) {
  level_list <- lapply(
    intersect(marginalize_over, names(fit$data)),
    function(col) fit_levels(fit, col)
  )
  names(level_list) <- intersect(marginalize_over, names(fit$data))
  level_list <- level_list[lengths(level_list) > 0]

  grid <- if (length(level_list)) {
    tidyr::expand_grid(!!!level_list)
  } else {
    tibble(.rows = 1)
  }

  grid$disease_groups_altered <- "Normal"
  grid$tissue_groups <- tissue_group
  grid$offset <- 0

  for (col in new_level_cols) {
    grid[[col]] <- new_level_tag
  }

  grid
}

fit_diagnostics <- function(fit) {
  draw_diagnostics <- posterior::summarise_draws(
    posterior::as_draws_array(fit),
    "rhat",
    "ess_bulk"
  )

  divergent_draws <- brms::nuts_params(fit) |>
    filter(.data$Parameter == "divergent__")

  tibble(
    maximum_rhat = max(draw_diagnostics$rhat, na.rm = TRUE),
    minimum_ess_bulk = min(draw_diagnostics$ess_bulk, na.rm = TRUE),
    post_warmup_transitions = nrow(divergent_draws),
    divergent_transitions = sum(divergent_draws$Value == 1),
    divergent_percent = mean(divergent_draws$Value == 1) * 100
  )
}

posterior_liver_draws <- function(fit, n_draws, seed) {
  nd_grid <- build_comprehensive_grid(fit)
  n_use <- min(n_draws, posterior::ndraws(posterior::as_draws_df(fit)))

  set.seed(seed)
  log_mu <- brms::posterior_linpred(
    fit,
    newdata = nd_grid,
    transform = FALSE,
    re_formula = NULL,
    allow_new_levels = TRUE,
    sample_new_levels = "gaussian",
    ndraws = n_use
  )

  tibble(
    draw = seq_len(nrow(log_mu)),
    reference_log_mu = rowMeans(log_mu),
    n_grid_rows = nrow(nd_grid)
  )
}

empirical_liver_centroid <- function(fit) {
  healthy_liver <- fit$data |>
    filter(
      .data$disease_groups_altered == "Normal",
      .data$tissue_groups == params$baseline_tissue_group
    ) |>
    mutate(
      normalized_log_expression = log1p(.data$counts / exp(.data$offset))
    )

  dataset_means <- healthy_liver |>
    group_by(.data$dataset_id_altered) |>
    summarise(
      dataset_mean = mean(.data$normalized_log_expression),
      n_samples = n(),
      .groups = "drop"
    )

  tibble(
    empirical_log_expression = mean(dataset_means$dataset_mean),
    n_empirical_datasets = nrow(dataset_means),
    n_empirical_samples = sum(dataset_means$n_samples)
  )
}

compute_reference_job <- function(job, cache_dir) {
  cache_path <- file.path(
    cache_dir,
    paste0(job$job_id, "_", job$cell_type, "_", job$.feature, ".rds")
  )

  if (!isTRUE(params$refresh_references) && file.exists(cache_path)) {
    return(readRDS(cache_path))
  }

  fit <- read_model_fit(job$target_name, job$store)

  result <- list(
    posterior_draws = posterior_liver_draws(
      fit = fit,
      n_draws = params$n_posterior_draws,
      seed = params$seed + job$job_id
    ) |>
      mutate(
        gene = job$gene,
        .feature = job$.feature,
        cell_type = job$cell_type,
        .before = 1
      ),
    diagnostics = bind_cols(
      tibble(
        gene = job$gene,
        .feature = job$.feature,
        cell_type = job$cell_type
      ),
      fit_diagnostics(fit),
      empirical_liver_centroid(fit)
    )
  )

  saveRDS(result, cache_path, compress = "gzip")
  result
}
```

## Retrieval and label-null helpers

```{r}
#| label: retrieval-helpers

calculate_reference_correlations <- function(
    patient_contrasts,
    reference_values,
    method
) {
  patient_contrasts |>
    inner_join(
      reference_values,
      by = c("gene", ".feature")
    ) |>
    group_by(
      .data$patient_id,
      .data$panel,
      .data$cell_type
    ) |>
    summarise(
      correlation = cor(
        .data$delta_q3,
        .data$reference_value,
        method = "spearman"
      ),
      n_genes = n(),
      .groups = "drop"
    ) |>
    mutate(method = method, .before = 1)
}

summarise_retrieval <- function(correlations, target_cell = "nk") {
  correlations |>
    group_by(
      .data$method,
      .data$patient_id,
      .data$panel
    ) |>
    group_modify(function(data, key) {
      target_correlation <- data$correlation[
        data$cell_type == target_cell
      ][[1]]

      best_other <- data |>
        filter(.data$cell_type != target_cell) |>
        dplyr::slice_max(
          order_by = .data$correlation,
          n = 1,
          with_ties = FALSE
        )

      tibble(
        n_genes = dplyr::first(data$n_genes),
        nk_correlation = target_correlation,
        best_other_cell = best_other$cell_type,
        best_other_correlation = best_other$correlation,
        nk_margin = target_correlation - best_other$correlation,
        nk_rank = rank(
          -data$correlation,
          ties.method = "min"
        )[data$cell_type == target_cell][[1]]
      )
    }) |>
    ungroup()
}

reference_matrix_from_long <- function(reference_values) {
  wide <- reference_values |>
    dplyr::select(gene, cell_type, reference_value) |>
    pivot_wider(
      names_from = .data$cell_type,
      values_from = .data$reference_value
    ) |>
    arrange(.data$gene)

  matrix <- as.matrix(wide[, candidate_cells, drop = FALSE])
  rownames(matrix) <- wide$gene
  matrix
}

retrieval_margin <- function(
    delta,
    reference_matrix,
    target_cell = "nk"
) {
  genes <- intersect(names(delta), rownames(reference_matrix))

  correlations <- apply(
    reference_matrix[genes, , drop = FALSE],
    2,
    function(reference) {
      cor(delta[genes], reference, method = "spearman")
    }
  )

  other_correlations <- correlations[
    names(correlations) != target_cell
  ]
  best_other <- names(which.max(other_correlations))

  tibble(
    nk_correlation = unname(correlations[target_cell]),
    best_other_cell = best_other,
    best_other_correlation = unname(other_correlations[best_other]),
    nk_margin = unname(
      correlations[target_cell] - other_correlations[best_other]
    )
  )
}

exact_patient_permutations <- function(
    patient,
    genes,
    metadata,
    log_expression,
    reference_matrix,
    method
) {
  patient_regions <- metadata |>
    filter(.data$patient_id == patient)

  all_regions <- patient_regions$roi_key
  observed_nk <- patient_regions$roi_key[
    patient_regions$region_group == "NK-selected TC"
  ]
  assignments <- combn(
    all_regions,
    length(observed_nk),
    simplify = FALSE
  )

  map_dfr(seq_along(assignments), function(permutation_id) {
    nk_regions <- assignments[[permutation_id]]
    cold_regions <- setdiff(all_regions, nk_regions)

    delta <-
      row_median(log_expression[genes, nk_regions, drop = FALSE]) -
      row_median(log_expression[genes, cold_regions, drop = FALSE])
    names(delta) <- genes

    retrieval_margin(delta, reference_matrix) |>
      mutate(
        method = method,
        patient_id = patient,
        panel = dplyr::first(patient_regions$panel),
        permutation_id = permutation_id,
        observed_labels = setequal(nk_regions, observed_nk),
        .before = 1
      )
  })
}

summarise_exact_test <- function(permutation_results) {
  permutation_results |>
    group_by(
      .data$method,
      .data$patient_id,
      .data$panel
    ) |>
    summarise(
      n_assignments = n(),
      observed_margin = .data$nk_margin[.data$observed_labels][[1]],
      observed_percentile = mean(
        .data$nk_margin <= observed_margin
      ),
      upper_tail_fraction = mean(
        .data$nk_margin >= observed_margin
      ),
      .groups = "drop"
    )
}

simulate_joint_null <- function(
    permutation_results,
    n_draws,
    seed
) {
  methods <- unique(permutation_results$method)

  map_dfr(seq_along(methods), function(method_id) {
    method <- methods[[method_id]]
    method_results <- permutation_results |>
      filter(.data$method == method)
    patient_margins <- split(
      method_results$nk_margin,
      method_results$patient_id
    )

    set.seed(seed + method_id)
    sampled_margins <- vapply(
      patient_margins,
      sample,
      numeric(n_draws),
      size = n_draws,
      replace = TRUE
    )

    tibble(
      method = method,
      joint_draw = seq_len(n_draws),
      cohort_mean_margin = rowMeans(sampled_margins)
    )
  })
}

summarise_joint_null <- function(permutation_results, joint_null) {
  observed <- permutation_results |>
    filter(.data$observed_labels) |>
    group_by(.data$method) |>
    summarise(
      observed_cohort_mean_margin = mean(.data$nk_margin),
      .groups = "drop"
    )

  joint_null |>
    inner_join(observed, by = "method") |>
    group_by(.data$method, .data$observed_cohort_mean_margin) |>
    summarise(
      n_joint_draws = n(),
      observed_percentile = mean(
        .data$cohort_mean_margin <= .data$observed_cohort_mean_margin
      ),
      monte_carlo_upper_tail =
        (sum(
          .data$cohort_mean_margin >= .data$observed_cohort_mean_margin
        ) + 1) / (n() + 1),
      .groups = "drop"
    )
}
```

------------------------------------------------------------------------

# Frozen design and HCC data

The external spatial data and author-defined region labels come from
Jia *et al.* (2025), *Spatial immune scoring system predicts hepatocellular
carcinoma recurrence*, *Nature* **640**, 1029–1040
([doi:10.1038/s41586-025-08668-x](https://doi.org/10.1038/s41586-025-08668-x)).
Only files present in the audited local Code Ocean export are used; unavailable
raw GeoMx data and supplementary annotations are not reconstructed.

The primary biological unit is the patient; individual GeoMx regions are
spatial sampling units. Only matched tumour-centre comparisons are used, so
each patient acts as their own spatial control. Recurrence is retained only as
source metadata and is not used in this report.

```{r}
#| label: load-geomx-data

core <- load_rdata_environment(core_rdata)
source_de <- load_rdata_environment(source_de_rdata)

geomx_expression <- as.matrix(core$data_gene_df_ed)
mode(geomx_expression) <- "numeric"

geomx_metadata <- as.data.frame(
  t(core$data_meta_df_ed),
  stringsAsFactors = FALSE
) |>
  rownames_to_column("roi_key") |>
  mutate(
    author_index = as.integer(str_remove(.data$roi_key, "^\\.\\.\\.")),
    patient_id = .data$sampleid,
    panel = if_else(
      str_detect(.data$sampleid_raw, "_2_14Dec"),
      "low coverage",
      "high coverage"
    ),
    region_group = case_when(
      .data$author_index %in% core$NKatTumor_indx ~ "NK-selected TC",
      str_detect(
        .data$label,
        regex("^cold tumor", ignore_case = TRUE)
      ) ~ "Cold TC",
      TRUE ~ NA_character_
    ),
    region_detail = case_when(
      .data$region_group == "Cold TC" ~ "cold tumour",
      str_detect(
        .data$label,
        regex("CD3 rich too", ignore_case = TRUE)
      ) ~ "CD57/CD3 overlap",
      .data$region_group == "NK-selected TC" ~ "CD57-rich",
      TRUE ~ NA_character_
    )
  )

primary_patients <- c("A24", "A29", "A10", "A9", "A3", "A39")

primary_regions <- geomx_metadata |>
  filter(
    .data$patient_id %in% primary_patients,
    !is.na(.data$region_group)
  ) |>
  mutate(
    patient_id = factor(.data$patient_id, levels = primary_patients),
    region_group = factor(
      .data$region_group,
      levels = c("Cold TC", "NK-selected TC")
    )
  ) |>
  arrange(.data$patient_id, .data$region_group, .data$author_index)

region_design <- primary_regions |>
  count(
    .data$patient_id,
    .data$panel,
    .data$region_group,
    .data$region_detail,
    name = "n_regions",
    .drop = FALSE
  ) |>
  filter(.data$n_regions > 0)

write_csv(
  primary_regions,
  file.path(output_dir, "primary_tc_regions.csv")
)

region_design
```

The six patients contribute `r nrow(primary_regions)` regions:
`r sum(primary_regions$region_group == "NK-selected TC")` NK-selected TC and
`r sum(primary_regions$region_group == "Cold TC")` cold TC regions. Of the
NK-selected regions, `r sum(primary_regions$region_detail == "CD57-rich")`
are labelled CD57-rich and
`r sum(primary_regions$region_detail == "CD57/CD3 overlap")` are labelled
CD57/CD3 overlap. The overlap is therefore part of what “NK-selected” means in
this primary source-defined experiment and is reported explicitly.

------------------------------------------------------------------------

# Source-excluded common gene universe

The gene universe is frozen without using the observed retrieval result. It
starts from genes measured in every one of the 93 GeoMx segments, excludes all
241 source spatial genes and the transcripts used to define the CD57/CD3
regions, and then retains only unambiguous Ensembl mappings available in all
seven posteriorHCA model stores.

```{r}
#| label: model-inventories

model_inventories <- map_dfr(
  candidate_cells,
  read_model_inventory
)

common_model_features <- model_inventories |>
  group_by(.data$.feature) |>
  summarise(n_cell_types = n_distinct(.data$cell_type), .groups = "drop") |>
  filter(.data$n_cell_types == length(candidate_cells)) |>
  pull(.data$.feature)
```

```{r}
#| label: gene-universe

complete_geomx_genes <- rownames(geomx_expression)[
  rowSums(is.na(geomx_expression)) == 0
]

source_genes <- bind_rows(
  source_de$fd_df,
  source_de$fd_uf,
  source_de$fu_df,
  source_de$fu_uf
) |>
  distinct(.data$gene) |>
  pull(.data$gene)

times_genes <- c("SPON2", "ZFP36L2", "ZFP36", "VIM", "HLA-DRB1")
selection_genes <- c("B3GAT1", "CD3D", "CD3E", "CD3G")

source_excluded_symbols <- setdiff(
  complete_geomx_genes,
  c(source_genes, times_genes, selection_genes)
)

gene_mapping <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = source_excluded_symbols,
  columns = "ENSEMBL",
  keytype = "SYMBOL"
) |>
  as_tibble() |>
  filter(
    !is.na(.data$ENSEMBL),
    .data$ENSEMBL %in% common_model_features
  ) |>
  distinct(.data$SYMBOL, .data$ENSEMBL) |>
  add_count(.data$SYMBOL, name = "n_model_mappings") |>
  filter(.data$n_model_mappings == 1) |>
  transmute(
    gene = .data$SYMBOL,
    .feature = .data$ENSEMBL
  ) |>
  arrange(.data$gene)

gene_universe_audit <- tibble(
  stage = c(
    "All GeoMx matrix rows",
    "Complete in all 93 GeoMx segments",
    "After source, TIMES and region-selection exclusion",
    "Available in all seven posteriorHCA stores"
  ),
  n_genes = c(
    nrow(geomx_expression),
    length(complete_geomx_genes),
    length(source_excluded_symbols),
    nrow(gene_mapping)
  )
)

write_csv(
  gene_mapping,
  file.path(output_dir, "source_excluded_gene_universe_before_model_qc.csv")
)

gene_universe_audit
```

The five TIMES genes are not used. None is present in the all-segment common
universe, and they were selected using the same HCC source data. Among the 241
source genes, only `BIRC5` overlaps the 1,788-gene common GeoMx universe, but
the full source list is excluded before model intersection.

------------------------------------------------------------------------

# HCC within-patient spatial contrasts

```{r}
#| label: hcc-contrasts

log2_q3 <- log2(geomx_expression)

hcc_contrasts_before_model_qc <- map_dfr(
  primary_patients,
  make_patient_contrast,
  genes = gene_mapping$gene,
  metadata = primary_regions,
  log_expression = log2_q3
) |>
  left_join(gene_mapping, by = "gene")

write_csv(
  hcc_contrasts_before_model_qc,
  file.path(output_dir, "hcc_tc_contrasts_before_model_qc.csv")
)

hcc_contrasts_before_model_qc |>
  group_by(.data$patient_id, .data$panel) |>
  summarise(
    n_genes = n(),
    median_delta_q3 = median(.data$delta_q3),
    .groups = "drop"
  )
```

Positive `delta_q3` means that a gene is higher in the patient's NK-selected
TC regions than in their cold TC regions. The analysis compares the ranking of
these gene-wise changes with each frozen healthy cell reference; it never
compares Q3 values directly with posteriorHCA counts.

------------------------------------------------------------------------

# Frozen healthy-liver references

## Comprehensive healthy-liver baseline

Each gene-by-cell reference is the same comprehensive healthy-liver estimand
used in the universal baseline report. Disease is fixed to Normal, tissue to
liver, and offset to zero. Sex, age decade, ethnicity and assay are expanded
to the full factorial of levels present in that fitted model; every grid cell
receives equal weight. The scalar baseline draw is the mean of $\log\mu$ over
that grid (`mean(log mu)`, not `log(mean mu)`). Dataset identity is a new
shared level sampled from the hierarchical prior
(`allow_new_levels = TRUE`, `sample_new_levels = "gaussian"`,
`re_formula = NULL`), so dataset uncertainty enters the baseline posterior
rather than being pinned at the population mean or averaged over observed
dataset IDs. Retrieval still uses Spearman correlation across genes within
each cell reference, so a cell-model-specific additive intercept does not
change that cell's correlation or the NK retrieval margin.

## Model evaluation and caching

For each gene and cell type, `posterior_linpred()` returns posterior draws of
the comprehensive healthy-liver grand-mean $\log\mu$. The empirical benchmark
uses the same fitted model's training rows but does no modelling: it
normalizes counts by their fitted offset, averages within atlas dataset and
then gives each dataset equal weight. Each gene-by-cell job is written to
disk immediately so an interrupted run can resume.

```{r}
#| label: compute-reference-cache

reference_jobs <- tidyr::crossing(
  gene_mapping,
  cell_type = candidate_cells
) |>
  left_join(
    model_inventories,
    by = c(".feature", "cell_type")
  ) |>
  mutate(job_id = row_number())

reference_job_cache_dir <- file.path(
  output_dir,
  "frozen_reference_jobs"
)
dir.create(reference_job_cache_dir, recursive = TRUE, showWarnings = FALSE)

n_jobs <- nrow(reference_jobs)
results <- vector("list", n_jobs)
progress <- txtProgressBar(min = 0, max = n_jobs, style = 3)

for (i in seq_len(n_jobs)) {
  job <- reference_jobs[i, ]
  results[[i]] <- compute_reference_job(job, reference_job_cache_dir)
  setTxtProgressBar(progress, i)
  if (i == 1L || i == n_jobs || i %% 25L == 0L) {
    message(
      sprintf(
        "\n[%d/%d] %s | %s | %s",
        i,
        n_jobs,
        job$cell_type,
        job$gene,
        job$.feature
      )
    )
  }
}

close(progress)

reference_results <- list(
  posterior_draws = map_dfr(results, "posterior_draws"),
  diagnostics = map_dfr(results, "diagnostics")
)
```

## Model-diagnostic triage and final analysis genes

```{r}
#| label: model-quality-gate

reference_diagnostics <- reference_results$diagnostics |>
  mutate(
    diagnostic_status = case_when(
      .data$maximum_rhat > params$fail_maximum_rhat |
        .data$divergent_percent >= params$fail_minimum_divergence_percent ~
        "FAIL",
      .data$maximum_rhat <= params$pass_maximum_rhat &
        .data$divergent_percent < params$pass_maximum_divergence_percent ~
        "PASS",
      TRUE ~ "WARN"
    ),
    diagnostic_status = factor(
      .data$diagnostic_status,
      levels = c("PASS", "WARN", "FAIL")
    )
  )

gene_model_quality <- reference_diagnostics |>
  group_by(.data$gene, .data$.feature) |>
  summarise(
    n_models = n(),
    all_models_pass = all(.data$diagnostic_status == "PASS"),
    any_model_failed = any(.data$diagnostic_status == "FAIL"),
    n_models_warn = sum(.data$diagnostic_status == "WARN"),
    maximum_rhat = max(.data$maximum_rhat),
    minimum_ess_bulk = min(.data$minimum_ess_bulk),
    maximum_divergent_percent = max(.data$divergent_percent),
    divergent_transitions = sum(.data$divergent_transitions),
    .groups = "drop"
  )

strict_gene_mapping <- gene_mapping |>
  inner_join(
    gene_model_quality |>
      filter(.data$all_models_pass),
    by = c("gene", ".feature")
  ) |>
  arrange(.data$gene)

analysis_gene_mapping <- gene_mapping |>
  inner_join(
    gene_model_quality |>
      filter(!.data$any_model_failed),
    by = c("gene", ".feature")
  ) |>
  arrange(.data$gene)

hcc_contrasts <- hcc_contrasts_before_model_qc |>
  semi_join(
    analysis_gene_mapping,
    by = c("gene", ".feature")
  )

write_csv(
  reference_diagnostics,
  file.path(output_dir, "posteriorHCA_reference_model_diagnostics.csv")
)
write_csv(
  analysis_gene_mapping,
  file.path(output_dir, "final_analysis_gene_universe.csv")
)
write_csv(
  strict_gene_mapping,
  file.path(output_dir, "strict_pass_gene_universe.csv")
)
write_csv(
  hcc_contrasts,
  file.path(output_dir, "hcc_tc_contrasts_final.csv")
)

tibble(
  item = c(
    "Candidate genes before model diagnostics",
    "Genes with PASS in all seven cell models",
    "Genes without a FAIL among seven cell models",
    "Models evaluated",
    "Models classified PASS",
    "Models classified WARN",
    "Models classified FAIL"
  ),
  n = c(
    nrow(gene_mapping),
    nrow(strict_gene_mapping),
    nrow(analysis_gene_mapping),
    nrow(reference_diagnostics),
    sum(reference_diagnostics$diagnostic_status == "PASS"),
    sum(reference_diagnostics$diagnostic_status == "WARN"),
    sum(reference_diagnostics$diagnostic_status == "FAIL")
  )
)
```

The frozen models were fitted before this case study and are not refitted here.
Diagnostics therefore use the same three-level convention as the project's
convergence report. `PASS` requires maximum R-hat no greater than
`r params$pass_maximum_rhat` and fewer than
`r params$pass_maximum_divergence_percent`% divergent transitions. `FAIL`
requires maximum R-hat greater than `r params$fail_maximum_rhat` or at least
`r params$fail_minimum_divergence_percent`% divergent transitions; intermediate
fits are retained as visible `WARN` results. A gene is excluded only if at
least one of its seven cell models is `FAIL`. This preserves the frozen-model
constraint without presenting soft warnings as clean convergence. The
all-seven-`PASS` universe is saved separately as a claim-strength audit; the
broader no-`FAIL` universe is an operational, explicitly exploratory
evaluation of the deployed frozen stores.

------------------------------------------------------------------------

# Cell-reference retrieval

## Construct posteriorHCA and naive empirical cell specificities

```{r}
#| label: construct-reference-values

posterior_reference_draws <- reference_results$posterior_draws |>
  semi_join(
    analysis_gene_mapping,
    by = c("gene", ".feature")
  ) |>
  group_by(.data$gene, .data$.feature, .data$draw) |>
  mutate(
    reference_value =
      .data$reference_log_mu - mean(.data$reference_log_mu)
  ) |>
  ungroup()

posterior_reference <- posterior_reference_draws |>
  group_by(.data$gene, .data$.feature, .data$cell_type) |>
  summarise(
    reference_value = median(.data$reference_value),
    .groups = "drop"
  )

empirical_reference <- reference_diagnostics |>
  semi_join(
    analysis_gene_mapping,
    by = c("gene", ".feature")
  ) |>
  group_by(.data$gene, .data$.feature) |>
  mutate(
    reference_value =
      .data$empirical_log_expression -
      mean(.data$empirical_log_expression)
  ) |>
  ungroup() |>
  dplyr::select(
    gene,
    .feature,
    cell_type,
    reference_value
  )
```

## Primary retrieval metrics

For each patient and candidate cell, Spearman correlation compares the ranking
of GeoMx spatial changes with the ranking of reference specificity. The primary
specificity statistic is

$$
M_p = \rho_{p,\mathrm{NK}} -
\max_{c \neq \mathrm{NK}} \rho_{pc}.
$$

A positive margin means that liver NK is the best of the seven frozen
posteriorHCA cell references.

```{r}
#| label: primary-retrieval

retrieval_correlations <- bind_rows(
  calculate_reference_correlations(
    hcc_contrasts,
    posterior_reference,
    method = "posteriorHCA"
  ),
  calculate_reference_correlations(
    hcc_contrasts,
    empirical_reference,
    method = "Naive empirical HCA"
  )
) |>
  mutate(
    cell_label = unname(cell_labels[.data$cell_type])
  )

retrieval_metrics <- summarise_retrieval(
  retrieval_correlations,
  target_cell = target_cell
)

posterior_draw_correlations <- hcc_contrasts |>
  inner_join(
    posterior_reference_draws |>
      dplyr::select(
        gene,
        .feature,
        draw,
        cell_type,
        reference_value
      ),
    by = c("gene", ".feature")
  ) |>
  group_by(
    .data$draw,
    .data$patient_id,
    .data$panel,
    .data$cell_type
  ) |>
  summarise(
    correlation = cor(
      .data$delta_q3,
      .data$reference_value,
      method = "spearman"
    ),
    n_genes = n(),
    .groups = "drop"
  ) |>
  mutate(method = "posteriorHCA", .before = 1)

posterior_draw_metrics <- posterior_draw_correlations |>
  group_by(
    .data$method,
    .data$draw,
    .data$patient_id,
    .data$panel
  ) |>
  group_modify(function(data, key) {
    target_correlation <- data$correlation[
      data$cell_type == target_cell
    ][[1]]
    best_other_correlation <- max(
      data$correlation[data$cell_type != target_cell]
    )

    tibble(
      nk_correlation = target_correlation,
      nk_margin = target_correlation - best_other_correlation
    )
  }) |>
  ungroup()

posterior_uncertainty <- posterior_draw_metrics |>
  group_by(.data$patient_id, .data$panel) |>
  summarise(
    nk_correlation_lo = quantile(.data$nk_correlation, 0.025),
    nk_correlation_hi = quantile(.data$nk_correlation, 0.975),
    nk_margin_lo = quantile(.data$nk_margin, 0.025),
    nk_margin_hi = quantile(.data$nk_margin, 0.975),
    posterior_probability_margin_positive = mean(.data$nk_margin > 0),
    .groups = "drop"
  ) |>
  mutate(method = "posteriorHCA", .before = 1)

retrieval_metrics <- retrieval_metrics |>
  left_join(
    posterior_uncertainty,
    by = c("method", "patient_id", "panel")
  )

write_csv(
  retrieval_correlations,
  file.path(output_dir, "reference_correlations.csv")
)
write_csv(
  retrieval_metrics,
  file.path(output_dir, "retrieval_metrics.csv")
)
```

------------------------------------------------------------------------

# Spatial-label compatibility null

Within each patient, the analysis enumerates every allocation with the same
number of NK-selected and cold regions as observed. It then repeats the entire
gene-wise contrast and reference-retrieval calculation. This sensitivity asks
whether the protein-defined spatial labels are unusually compatible with the
frozen reference; it does not treat individual regions as independent patients
and does not assume that the source regions were randomized.

```{r}
#| label: exact-label-null

posterior_reference_matrix <- reference_matrix_from_long(
  posterior_reference
)
empirical_reference_matrix <- reference_matrix_from_long(
  empirical_reference
)

reference_matrices <- list(
  posteriorHCA = posterior_reference_matrix,
  `Naive empirical HCA` = empirical_reference_matrix
)

exact_permutations <- imap_dfr(
  reference_matrices,
  function(reference_matrix, method) {
    map_dfr(
      primary_patients,
      exact_patient_permutations,
      genes = analysis_gene_mapping$gene,
      metadata = primary_regions,
      log_expression = log2_q3,
      reference_matrix = reference_matrix,
      method = method
    )
  }
)

exact_test_summary <- summarise_exact_test(exact_permutations)

joint_null <- simulate_joint_null(
  exact_permutations,
  n_draws = params$joint_null_draws,
  seed = params$seed
)

joint_null_summary <- summarise_joint_null(
  exact_permutations,
  joint_null
)

write_csv(
  exact_permutations,
  file.path(output_dir, "exact_region_label_permutations.csv")
)
write_csv(
  exact_test_summary,
  file.path(output_dir, "exact_region_label_test_summary.csv")
)
write_csv(
  joint_null_summary,
  file.path(output_dir, "joint_region_label_null_summary.csv")
)

joint_null_summary
```

For a 3-versus-3 patient there are only 20 exact assignments, so the smallest
attainable individual upper-tail fraction is 0.05. These fractions are
compatibility summaries, not randomization-test P values: the available files
do not establish that all regions were randomized or freely exchangeable.
The cohort statistic is the pre-specified mean NK margin across six patients.
Its joint null independently samples from each patient's exhaustive allocation
set; the reported Monte Carlo tail fraction uses `r params$joint_null_draws`
draws and preserves the patient as the biological unit.

------------------------------------------------------------------------

# Results

## Patient-level result table

```{r}
#| label: result-table

result_table <- retrieval_metrics |>
  dplyr::select(
    method,
    patient_id,
    panel,
    n_genes,
    nk_correlation,
    best_other_cell,
    best_other_correlation,
    nk_margin,
    nk_rank,
    nk_margin_lo,
    nk_margin_hi,
    posterior_probability_margin_positive
  ) |>
  left_join(
    exact_test_summary,
    by = c("method", "patient_id", "panel")
  ) |>
  mutate(
    best_other_cell = unname(cell_labels[.data$best_other_cell]),
    across(
      where(is.numeric),
      ~ round(.x, 3)
    )
  ) |>
  arrange(.data$panel, .data$patient_id, .data$method)

write_csv(
  result_table,
  file.path(output_dir, "spatial_transfer_result_table.csv")
)

result_table
```

## Which cell reference is retrieved?

```{r}
#| label: correlation-heatmap
#| fig-width: 9
#| fig-height: 5.5
#| fig-align: center

correlation_heatmap <- retrieval_correlations |>
  filter(.data$method == "posteriorHCA") |>
  mutate(
    patient_id = fct_rev(factor(.data$patient_id, levels = primary_patients)),
    cell_label = factor(
      .data$cell_label,
      levels = unname(cell_labels[candidate_cells])
    )
  ) |>
  ggplot(aes(x = .data$cell_label, y = .data$patient_id)) +
  geom_tile(aes(fill = .data$correlation), colour = "white") +
  geom_text(
    aes(label = number(.data$correlation, accuracy = 0.01)),
    size = 3
  ) +
  scale_fill_gradient2(
    low = "#C23B4A",
    mid = "white",
    high = "#2878B5",
    midpoint = 0,
    name = "Spearman\nrho"
  ) +
  labs(
    title = "Frozen posteriorHCA reference retrieval on GeoMx",
    subtitle = "Each row compares one within-patient NK-selected-minus-cold TC contrast with seven healthy-liver cell references",
    x = NULL,
    y = "HCC patient"
  ) +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1),
    panel.grid = element_blank(),
    legend.position = "right"
  )

correlation_heatmap

ggsave(
  file.path(output_dir, "posteriorHCA_reference_correlation_heatmap.pdf"),
  correlation_heatmap,
  width = 9,
  height = 5.5
)
```

The transfer claim requires NK to be the best-matching reference, not merely
positively correlated. A positive NK margin means that NK outranks CD8 TEM,
MAIT, gamma-delta T, monocytic, macrophage and epithelial references.

## Descriptive sensitivity: posteriorHCA versus a naive empirical centroid

```{r}
#| label: retrieval-margin-plot
#| fig-width: 8.5
#| fig-height: 5
#| fig-align: center

margin_plot_data <- retrieval_metrics |>
  mutate(
    patient_id = factor(.data$patient_id, levels = primary_patients),
    method = factor(
      .data$method,
      levels = c("Naive empirical HCA", "posteriorHCA")
    )
  )

margin_plot <- margin_plot_data |>
  ggplot(aes(x = .data$patient_id, y = .data$nk_margin)) +
  geom_hline(yintercept = 0, colour = "grey55", linewidth = 0.5) +
  geom_line(aes(group = .data$patient_id), colour = "grey70") +
  geom_point(
    aes(colour = .data$method, shape = .data$method),
    size = 3
  ) +
  geom_errorbar(
    data = margin_plot_data |>
      filter(.data$method == "posteriorHCA"),
    aes(ymin = .data$nk_margin_lo, ymax = .data$nk_margin_hi),
    width = 0.08,
    colour = method_colours[["posteriorHCA"]]
  ) +
  facet_grid(cols = vars(.data$panel), scales = "free_x", space = "free_x") +
  scale_colour_manual(values = method_colours) +
  scale_shape_manual(
    values = c(`Naive empirical HCA` = 16, posteriorHCA = 18)
  ) +
  labs(
    title = "Specificity of liver-NK retrieval",
    subtitle = "Positive values indicate that NK outperforms every alternative cell reference",
    x = "HCC patient",
    y = "NK retrieval margin",
    colour = NULL,
    shape = NULL
  ) +
  theme(legend.position = "bottom")

margin_plot

ggsave(
  file.path(output_dir, "posteriorHCA_vs_empirical_retrieval_margin.pdf"),
  margin_plot,
  width = 8.5,
  height = 5
)
```

The empirical comparator is intentionally naive. It uses offset-normalized
healthy-liver observations and equal dataset weighting, but it does not use
the comprehensive factor-grid grand mean or the new-dataset hierarchical draw
used by posteriorHCA. Any difference can therefore reflect covariate mix,
dataset handling or the log transformation as well as hierarchical modelling.
It is shown only to contextualize the frozen reference and cannot establish
posteriorHCA superiority.

## Where does the observed labelling fall among alternative allocations?

```{r}
#| label: exact-null-plot
#| fig-width: 9
#| fig-height: 5.5
#| fig-align: center

exact_null_plot <- exact_permutations |>
  filter(.data$method == "posteriorHCA") |>
  mutate(
    patient_id = factor(.data$patient_id, levels = primary_patients)
  ) |>
  ggplot(aes(x = .data$patient_id, y = .data$nk_margin)) +
  geom_hline(yintercept = 0, colour = "grey55", linewidth = 0.5) +
  geom_jitter(
    data = function(data) filter(data, !.data$observed_labels),
    width = 0.13,
    height = 0,
    colour = "grey72",
    alpha = 0.75,
    size = 1.8
  ) +
  geom_point(
    data = function(data) filter(data, .data$observed_labels),
    colour = "#2878B5",
    shape = 18,
    size = 4
  ) +
  facet_grid(cols = vars(.data$panel), scales = "free_x", space = "free_x") +
  labs(
    title = "Within-patient spatial-label compatibility null",
    subtitle = "Blue diamonds are the protein-defined labels; grey points are every size-matched reassignment",
    x = "HCC patient",
    y = "posteriorHCA NK retrieval margin"
  )

exact_null_plot

ggsave(
  file.path(output_dir, "posteriorHCA_exact_spatial_label_null.pdf"),
  exact_null_plot,
  width = 9,
  height = 5.5
)
```

------------------------------------------------------------------------

# Interpretation

```{r}
#| label: interpretation-summary

posterior_results <- retrieval_metrics |>
  filter(.data$method == "posteriorHCA")

panel_results <- posterior_results |>
  group_by(.data$panel) |>
  summarise(
    n_patients = n(),
    median_nk_correlation = median(.data$nk_correlation),
    median_nk_margin = median(.data$nk_margin),
    patients_with_nk_rank_1 = sum(.data$nk_rank == 1),
    .groups = "drop"
  )

evidence_summary <- tibble(
  evidence = c(
    "Patients with positive posteriorHCA NK correlation",
    "Patients with NK ranked first",
    "Patients with positive NK retrieval margin",
    "Coverage panels with positive median NK margin",
    "Patients with observed-label percentile at least 0.90",
    "Cohort joint-null upper-tail fraction"
  ),
  observed = c(
    paste0(sum(posterior_results$nk_correlation > 0), " / ", nrow(posterior_results)),
    paste0(sum(posterior_results$nk_rank == 1), " / ", nrow(posterior_results)),
    paste0(sum(posterior_results$nk_margin > 0), " / ", nrow(posterior_results)),
    paste0(sum(panel_results$median_nk_margin > 0), " / ", nrow(panel_results)),
    paste0(
      sum(
        exact_test_summary$method == "posteriorHCA" &
          exact_test_summary$observed_percentile >= 0.90
      ),
      " / ",
      sum(exact_test_summary$method == "posteriorHCA")
    ),
    format(
      joint_null_summary$monte_carlo_upper_tail[
        joint_null_summary$method == "posteriorHCA"
      ],
      digits = 3
    )
  )
)

evidence_summary
```

Evidence for cross-platform transport is strongest when the following pattern
appears together:

1. the liver-NK correlation is positive;
2. NK ranks first among all seven cell references, giving a positive margin;
3. the result is present in both GeoMx coverage panels;
4. the observed spatial labels lie near the top of their exhaustive
   compatibility null; and
5. the pattern remains after source-gene, TIMES-gene and segmentation-marker
   exclusion.

The posterior-draw intervals and positive-margin probability are conditional
on the frozen fitted models and the fixed HCC contrasts. They represent
reference-model sensitivity, not total uncertainty and not GeoMx calibration.
The naive empirical centroid is descriptive context only; this experiment does
not use it to claim superiority of hierarchical modelling.

Even a strong result remains a structural validation. GeoMx Q3 values are not
raw counts, the regions are mixed-cell spatial samples, and no healthy GeoMx or
paired cross-platform bridge exists. Consequently, this report does not call
the baseline absolutely calibrated to GeoMx.

------------------------------------------------------------------------

# Claim–evidence map

| Claim | Required evidence in this report | Status rule |
|---|---|---|
| Frozen posteriorHCA structure transfers to unseen GeoMx | Positive NK correlation and margin, both coverage panels, source-excluded genes and label-null support | Strongly supported only if all evidence classes agree and a useful all-seven-PASS gene universe exists; otherwise exploratory |
| Transfer is NK-specific | NK outranks CD8 TEM, MAIT, gamma-delta T, monocytic, macrophage and epithelial references | Supported patient by patient by a positive NK margin |
| posteriorHCA improves on a direct HCA centroid | A covariate- and estimand-matched direct reference comparison | Not tested; the naive empirical sensitivity is not a fair superiority benchmark |
| posteriorHCA is absolutely calibrated to GeoMx | Healthy or paired cross-platform bridge and valid GeoMx observation model | Not testable with the available data |

------------------------------------------------------------------------

# Reviewer-facing self-review

| Dimension | Question | Current safeguard |
|---|---|---|
| Contribution | Does the experiment test something unavailable from a direct REC/non-REC comparison? | Yes: correct frozen cell-reference retrieval on an unseen assay |
| Writing clarity | Are the two contrasts, biological unit and platform boundary explicit? | Yes: equations and region-level design are stated before results |
| Experimental strength | Is posteriorHCA compared with hard alternative references? | Yes: six alternative cell types; the unmatched empirical centroid is descriptive only |
| Evaluation completeness | Could source selection or arbitrary region assignment explain the result? | Source features are excluded and every within-patient label allocation is enumerated |
| Method soundness | Are Q3 values treated as raw counts or calibrated posteriorHCA observations? | No: only within-patient log-Q3 ranks are compared with latent reference structure |

------------------------------------------------------------------------

# Saved artefacts

```{r}
#| label: output-files

output_files <- tibble(
  file = list.files(
    output_dir,
    full.names = TRUE,
    recursive = TRUE,
    include.dirs = FALSE
  ),
  size_mb = file.info(file)$size / 1024^2
) |>
  mutate(size_mb = round(.data$size_mb, 2)) |>
  arrange(.data$file)

output_files
```

```{r}
#| label: session-info

sessionInfo()
```
