---
title: "TIMES HCC case study - posteriorHCA healthy-proximity proxy"
subtitle: "Can distance from healthy liver NK recover recurrence biology?"
date: today
date-format: "DD MMM YYYY"
title-block-banner: true
author:
  - name: Chen Zhan
    orcid: 0000-0002-4794-8339
    email: chen.zhan@adelaide.edu.au
    affiliation:
      - id: sgc
        name: SAiGENCI, The University of Adelaide
format:
  html:
    theme: minty
    df-print: paged
    code-line-numbers: true
    embed-resources: true
    code-fold: true
    code-fold-default: true
    toc: true
    toc-location: left
    toc-title: "Contents"
knitr:
  opts_chunk:
    message: false
    warning: false
editor: source
abstract: |
  This case study evaluates whether posteriorHCA can replace a direct
  recurrence-trained comparator with a universal healthy reference. We define
  an outcome-free healthy-proximity score from the absolute standardized
  distance between HCC NK-enriched expression and the comprehensive healthy
  liver NK grand-mean posterior. The proxy hypothesis is that non-REC HCC
  remains healthy-like while REC HCC is more atlas-discordant in both invasive
  front (IF) and tumour centre (TC). Direct edgeR non-REC versus REC contrasts
  are retained as a positive control, and the paper's TIMES validation scores
  provide a descriptive benchmark. The available GeoMx subset is very small,
  so this is proof of concept rather than an equivalence-power claim.
params:
  case_dir: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC"
  target_cell_type: "nk"
  model_store_root: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/HPC_posterior/NEW_CELL_TYPE"
  compartments:
    - "IF"
    - "TC"
  baseline_tissue_group: "liver"
  n_posterior_draws: 400
  refresh_baseline: false
  pseudobulk_h5ad: "/home/a1237163/lab/Mangiola_ImmuneAtlas/zenodo/pseudobulk_se.h5ad"
  baseline_draws_rds: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/Jie_HCC_V1_nk_comprehensive_liver_baseline_log_mu_draws.rds"
  paper_ground_truth_csv: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/Jie_HCC_paper_ground_truth_TIMES_biomarkers.csv"
  cohort_results_csv: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/Jie_HCC_four_cohort_comprehensive_liver_baseline_mu_tests.csv"
  recurrence_results_csv: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/Jie_HCC_IF_TC_nonREC_vs_REC_edger_tests.csv"
  patient_pseudobulk_csv: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/Jie_HCC_IF_TC_patient_pseudobulk_metadata.csv"
  normalization_csv: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/Jie_HCC_IF_TC_TMMwsp_offsets.csv"
  ground_truth_comparison_csv: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/Jie_HCC_TIMES_recurrence_ground_truth_and_healthy_context.csv"
  healthy_proxy_gene_csv: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/Jie_HCC_healthy_proximity_proxy_gene_compartment.csv"
  healthy_proxy_summary_csv: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/Jie_HCC_healthy_proximity_proxy_posterior_summary.csv"
  healthy_proxy_patient_csv: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/Jie_HCC_healthy_proximity_proxy_patient_scores.csv"
  healthy_proxy_performance_csv: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/Jie_HCC_healthy_proximity_proxy_performance.csv"
  paper_benchmark_csv: "/home/a1237163/lab/chen/posteriorHCA/dev/Jie_HCC/data/processed/Jie_HCC_TIMES_score_benchmark.csv"
---

# Introduction

Jia *et al.* (2025) built the TIMES score from spatial biomarker expression in
both the **invasive front (IF)** and **tumour centre (TC)** to predict recurrence
after HCC resection. The original claim is prognostic **within HCC**:
non-recurrent (non-REC) versus recurrent (REC) disease. Healthy tissue is not
part of the TIMES training contrast or its published ground truth.

The case-study hierarchy is:

| Role | Question | Interpretation |
|---|---|---|
| **Primary posteriorHCA proxy** | Is REC farther from healthy liver NK than non-REC in IF and TC? | Tests whether a universal healthy reference can recover recurrence ordering without training on recurrence |
| **Positive control** | Is each biomarker higher in non-REC than REC within IF and TC? | Confirms that the available GeoMx subset retains the paper's expression direction |
| **Paper benchmark** | How does proxy discrimination compare descriptively with the supplied TIMES validation scores? | Assesses promise, not formal equivalence, because cohorts and modalities differ |

The proxy is deliberately simple and outcome-free. For patient or cohort \(i\)
and gene \(g\), healthy distance is

$$
D_{ig} =
\frac{\left|\log\mu_{ig} - E(\log\mu_{g,\mathrm{healthy}})\right|}
     {SD(\log\mu_{g,\mathrm{healthy}})}.
$$

The multi-gene proxy score is the mean \(D_{ig}\) across atlas-covered TIMES
genes. Higher values indicate greater departure from healthy liver NK. The
case-study prediction is **REC distance > non-REC distance**. This prediction
is not stated in the original paper and must ultimately be validated in an
independent cohort.

The appropriate atlas tissue is **healthy liver NK**, not blood NK. The
`V1_nk` model includes `tissue_groups = "liver"`. This reference is used as the
fixed origin of the proxy; recurrence labels are used only after scoring, for
evaluation.

The statistical workflow below now matches SAVI: patient pseudobulk, TMMwsp
offsets anchored to the atlas training reference, edgeR NB group means, QL Wald
standard errors, Welch tests and empirical posterior ranks. The remaining data
limitation is upstream: the public CodeOcean matrix contains author-processed
GeoMx transcript-read-like values. We follow the paper capsule and round them
before count modelling, but controlled-access raw GeoMx counts remain preferable
for a final analysis.

------------------------------------------------------------------------

# Library setup

```{r}
#| label: setup

suppressPackageStartupMessages({
  library(tibble)
  library(tidyr)
  library(dplyr)
  library(readr)
  library(purrr)
  library(ggplot2)
  library(forcats)
  library(glue)
  library(stringr)
  library(scales)
  library(targets)
  library(brms)
  library(posterior)
  library(edgeR)
  library(SummarizedExperiment)
  library(S4Vectors)
  library(AnnotationDbi)
  library(org.Hs.eg.db)
})

theme_set(theme_minimal(base_size = 11))

data_dir <- file.path(params$case_dir, "data")
processed_dir <- file.path(data_dir, "processed")
dir.create(processed_dir, recursive = TRUE, showWarnings = FALSE)

target_compartments <- unlist(params$compartments)
```

------------------------------------------------------------------------

# TIMES biomarkers and NK model coverage

```{r}
#| label: biomarkers

times_biomarkers <- readr::read_csv(
  file.path(data_dir, "times_biomarkers.csv"),
  show_col_types = FALSE
) |>
  dplyr::select(gene, role, selection_step, notes, .feature)

times_biomarkers
```

```{r}
#| label: model-meta

model_path <- file.path(
  params$model_store_root,
  paste0("V1_", params$target_cell_type)
)

if (!dir.exists(file.path(model_path, "_targets"))) {
  stop("Missing targets store: ", file.path(model_path, "_targets"))
}

model_meta <- targets::tar_meta(
  starts_with("estimates_chunk_"),
  store = file.path(model_path, "_targets")
) |>
  dplyr::filter(.data$type == "branch") |>
  dplyr::mutate(
    .feature = stringr::str_extract(
      .data$warnings,
      "(?<=Gene:___)ENSG\\d+(?=___)"
    )
  )

biomarker_model_status <- times_biomarkers |>
  dplyr::mutate(
    in_v1_nk_store = !is.na(.data$.feature) & .data$.feature %in% model_meta$.feature,
    target_name = purrr::map_chr(.data$.feature, function(feature_id) {
      hit <- model_meta |>
        dplyr::filter(.data$.feature == feature_id) |>
        dplyr::slice(1) |>
        dplyr::pull(.data$name)
      if (length(hit)) hit[[1]] else NA_character_
    })
  )

biomarker_model_status |>
  dplyr::select(gene, .feature, in_v1_nk_store, target_name)
```

`HLA-DRB1` is kept in all HCC summaries, but it is excluded from posteriorHCA
baseline testing when absent from the `V1_nk` model store.

```{r}
#| label: baseline-tissue-check

example_target <- biomarker_model_status |>
  dplyr::filter(.data$in_v1_nk_store) |>
  dplyr::slice(1) |>
  dplyr::pull(.data$target_name)

example_fit <- targets::tar_read_raw(
  example_target[[1]],
  store = file.path(model_path, "_targets")
)$brms_fit[[1]]

tissue_reference_check <- tibble::tibble(
  requested_baseline_tissue = params$baseline_tissue_group,
  tissue_level_available = params$baseline_tissue_group %in%
    unique(as.character(example_fit$data$tissue_groups)),
  n_v1_nk_training_rows_for_tissue = sum(
    as.character(example_fit$data$tissue_groups) == params$baseline_tissue_group,
    na.rm = TRUE
  ),
  n_v1_nk_normal_training_rows_for_tissue = sum(
    as.character(example_fit$data$tissue_groups) == params$baseline_tissue_group &
      as.character(example_fit$data$disease_groups_altered) == "Normal",
    na.rm = TRUE
  )
)

tissue_reference_check
```

------------------------------------------------------------------------

# CodeOcean GeoMx IF/TC data

We use the complete all-compartment extraction generated by
`data/extract_codeocean_processed_data.R`, then restrict to the author's
NK-enriched ROI sets (`NKatBorder_indx` for IF and `NKatTumor_indx` for TC).

```{r}
#| label: load-codeocean

segment_path <- file.path(
  processed_dir,
  "codeocean_GeoMx_all_segments_long_all_genes.csv"
)

if (!file.exists(segment_path)) {
  stop(
    "Missing processed CodeOcean segment file: ", segment_path,
    ". Run data/extract_codeocean_processed_data.R first."
  )
}

geomx_segments <- readr::read_csv(
  segment_path,
  col_types = readr::cols(.feature = readr::col_character()),
  show_col_types = FALSE
)

geomx_times_nk_segments <- geomx_segments |>
  dplyr::filter(
    .data$is_times_biomarker,
    .data$compartment %in% target_compartments,
    .data$included_in_author_nk_analysis,
    !is.na(.data$value)
  ) |>
  dplyr::mutate(
    compartment = factor(.data$compartment, levels = target_compartments),
    recurrence = factor(.data$recurrence, levels = c("non-REC", "REC"))
  )

geomx_times_nk_patient <- geomx_times_nk_segments |>
  dplyr::group_by(
    .data$gene, .data$.feature, .data$patient_id, .data$recurrence,
    .data$age, .data$gender, .data$compartment, .data$platform
  ) |>
  dplyr::summarise(
    n_segments = dplyr::n_distinct(.data$author_column),
    value_sum = sum(.data$value, na.rm = TRUE),
    value_mean = mean(.data$value, na.rm = TRUE),
    value_median = stats::median(.data$value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    biomarker_model_status |>
      dplyr::select(gene, .feature, in_v1_nk_store),
    by = c("gene", ".feature")
  )

coverage_by_gene_compartment <- geomx_times_nk_patient |>
  dplyr::group_by(.data$compartment, .data$gene, .data$in_v1_nk_store) |>
  dplyr::summarise(
    n_patients = dplyr::n_distinct(.data$patient_id),
    n_segments = sum(.data$n_segments),
    patients = paste(sort(unique(.data$patient_id)), collapse = ", "),
    .groups = "drop"
  ) |>
  dplyr::arrange(.data$compartment, .data$gene)

coverage_by_gene_compartment
```

```{r}
#| label: plot-coverage
#| fig-width: 8
#| fig-height: 4
#| fig-align: center

ggplot2::ggplot(
  coverage_by_gene_compartment,
  ggplot2::aes(
    x = forcats::fct_reorder(.data$gene, .data$n_segments),
    y = .data$n_segments,
    fill = .data$compartment
  )
) +
  ggplot2::geom_col(position = "dodge", width = 0.75) +
  ggplot2::geom_text(
    ggplot2::aes(label = paste0("n=", .data$n_patients)),
    position = ggplot2::position_dodge(width = 0.75),
    vjust = -0.25,
    size = 3
  ) +
  ggplot2::coord_cartesian(ylim = c(0, max(coverage_by_gene_compartment$n_segments) * 1.18)) +
  ggplot2::labs(
    x = NULL,
    y = "Non-missing NK-enriched ROI segments",
    fill = "Compartment",
    title = "Available CodeOcean GeoMx data for TIMES biomarkers",
    subtitle = "Labels show patients with non-missing values"
  )
```

------------------------------------------------------------------------

# Paper ground truth and proxy hypothesis

The paper gives a **recurrence/spatial** ground truth, not a
healthy-liver-versus-HCC ground truth. TIMES uses spatial expression patterns in
HCC IF/TC to predict recurrence, and spatial integration is central to the
published score. Extended Data Fig. 1 shows greater NK enrichment at IF in
non-REC HCC, while Extended Data Fig. 5 links SPON2+ NK cells to activation,
IFN-gamma/perforin and migration-related programs.

For the five TIMES genes, the exact author CodeOcean differential-expression
object supports the following **paper-ground-truth recurrence direction**:
`non-REC > REC` in both IF and TC for all five biomarkers. This is the only
direction assigned a paper-ground-truth match. The posteriorHCA proxy adds a
new hypothesis: **non-REC is closer to healthy liver NK than REC**. That proxy
hypothesis is evaluated against recurrence labels but is not claimed as paper
ground truth. The SPON2 evidence provides mechanistic context, but it does not
specify a SPON2 direction relative to healthy liver NK.

```{r}
#| label: paper-ground-truth

paper_de_path <- file.path(
  params$case_dir,
  "codeocean",
  "data",
  "Input_1_Spatial_transcriptomics_analysis",
  "transcriptome_differential_analysis_input_3.RData"
)

if (!file.exists(paper_de_path)) {
  stop("Missing CodeOcean paper DE input: ", paper_de_path)
}

paper_env <- new.env(parent = emptyenv())
suppressWarnings(load(paper_de_path, envir = paper_env))
paper_niky_de <- get("niky_de", envir = paper_env)

paper_ground_truth <- paper_niky_de |>
  tibble::as_tibble() |>
  dplyr::filter(.data$gene %in% times_biomarkers$gene) |>
  dplyr::transmute(
    gene = .data$gene,
    paper_l2fc_nonrec_vs_rec_if =
      .data$l2fc_1Nonrelp_NKatBorder_agst_0Relp_NKatBorder,
    paper_padj_nonrec_vs_rec_if =
      .data$padj_1Nonrelp_NKatBorder_agst_0Relp_NKatBorder,
    paper_l2fc_nonrec_vs_rec_tc =
      .data$l2fc_1Nonrelp_NKatTumor_agst_0Relp_NKatTumor,
    paper_padj_nonrec_vs_rec_tc =
      .data$padj_1Nonrelp_NKatTumor_agst_0Relp_NKatTumor,
    paper_l2fc_nonrec_vs_rec_all =
      .data$l2fc_1Nonrelp_agst_0Relp,
    paper_padj_nonrec_vs_rec_all =
      .data$padj_1Nonrelp_agst_0Relp
  ) |>
  dplyr::mutate(
    expected_recurrence_direction_if = dplyr::if_else(
      .data$paper_l2fc_nonrec_vs_rec_if > 0,
      "non-REC higher",
      "REC higher"
    ),
    expected_recurrence_direction_tc = dplyr::if_else(
      .data$paper_l2fc_nonrec_vs_rec_tc > 0,
      "non-REC higher",
      "REC higher"
    ),
    expected_vs_healthy_liver = "not_defined_by_paper",
    expected_vs_healthy_liver_note = dplyr::case_when(
      .data$gene == "SPON2" ~
        "SPON2 has mechanistic support for NK activation and migration, but the paper does not define an expected direction versus healthy liver NK.",
      TRUE ~
        "The paper supports recurrence/spatial prediction, not a healthy-liver up/down hypothesis."
    )
  ) |>
  dplyr::left_join(
    times_biomarkers |>
      dplyr::select(gene, .feature),
    by = "gene"
  ) |>
  dplyr::arrange(.data$gene)

readr::write_csv(paper_ground_truth, params$paper_ground_truth_csv)

paper_ground_truth |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), ~round(.x, 4)))
```

------------------------------------------------------------------------

# Comprehensive healthy baseline draws

```{r}
#| label: baseline-helpers

fit_levels <- function(fit, var) {
  d <- fit$data
  if (is.null(d) || !var %in% names(d)) {
    return(character(0))
  }
  v <- d[[var]]
  lv <- if (is.factor(v)) levels(v) else sort(unique(as.character(v)))
  lv[!is.na(lv) & nzchar(lv)]
}

newdata_comprehensive_baseline <- function() {
  tibble::tibble(
    age_decade = NA_character_,
    sex = NA_character_,
    ethnicity_groups = NA_character_,
    assay_groups_altered = NA_character_,
    disease_groups_altered = "Normal",
    tissue_groups = params$baseline_tissue_group,
    offset = 0
  )
}

build_comprehensive_grid <- function(
    fit,
    newdata = newdata_comprehensive_baseline(),
    marginalize_over = c(
      "sex", "age_decade", "ethnicity_groups", "assay_groups_altered"
    ),
    new_level_cols = "dataset_id_altered",
    new_level_tag = "__comprehensive_healthy_baseline__"
) {
  newdata <- tibble::as_tibble(newdata)
  fit_vars <- colnames(fit$data)

  levels_list <- lapply(intersect(marginalize_over, fit_vars), function(col) {
    fit_levels(fit, col)
  })
  names(levels_list) <- intersect(marginalize_over, fit_vars)
  levels_list <- levels_list[vapply(levels_list, length, integer(1L)) > 0L]

  if (length(levels_list)) {
    grid <- do.call(tidyr::expand_grid, levels_list)
    rest <- newdata[, setdiff(names(newdata), names(levels_list)), drop = FALSE]
    out <- dplyr::bind_cols(rest[rep(1L, nrow(grid)), , drop = FALSE], grid)
  } else {
    out <- newdata
  }

  for (col in new_level_cols) {
    if (col %in% fit_vars) {
      out[[col]] <- new_level_tag
    }
  }

  tibble::as_tibble(out)
}

read_fit_for_feature <- function(feature_id) {
  target_name <- model_meta |>
    dplyr::filter(.data$.feature == feature_id) |>
    dplyr::slice(1) |>
    dplyr::pull(.data$name)

  if (!length(target_name)) {
    stop("No target found for feature ", feature_id)
  }

  chunk <- targets::tar_read_raw(
    target_name[[1]],
    store = file.path(model_path, "_targets")
  )

  chunk$brms_fit[[1]]
}

posterior_linpred_args <- function(fit, nd_grid) {
  args <- list(
    object = fit,
    newdata = nd_grid,
    transform = FALSE,
    re_formula = NULL,
    allow_new_levels = TRUE,
    sample_new_levels = "gaussian"
  )

  n_draws <- params$n_posterior_draws
  if (!is.null(n_draws) && !is.na(n_draws) && is.finite(n_draws) && n_draws > 0) {
    n_available <- tryCatch(
      posterior::ndraws(posterior::as_draws_df(fit)),
      error = function(e) NA_integer_
    )
    if (!is.na(n_available)) {
      n_draws <- min(as.integer(n_draws), n_available)
    }
    args$ndraws <- as.integer(n_draws)
  }

  args
}

compute_one_baseline <- function(gene, feature_id) {
  fit <- read_fit_for_feature(feature_id)
  nd_grid <- build_comprehensive_grid(fit)

  log_mu_mat <- do.call(
    brms::posterior_linpred,
    posterior_linpred_args(fit, nd_grid)
  )

  grand_mean_draws <- rowMeans(log_mu_mat)

  tibble::tibble(
    gene = gene,
    .feature = feature_id,
    draw = seq_along(grand_mean_draws),
    baseline_log_mu = as.numeric(grand_mean_draws),
    n_grid_rows = nrow(nd_grid)
  )
}
```

```{r}
#| label: baseline-draws

baseline_features <- biomarker_model_status |>
  dplyr::filter(.data$in_v1_nk_store) |>
  dplyr::select(gene, .feature)

if (file.exists(params$baseline_draws_rds) && !isTRUE(params$refresh_baseline)) {
  baseline_draws <- readRDS(params$baseline_draws_rds)
} else {
  baseline_draws <- purrr::map2_dfr(
    baseline_features$gene,
    baseline_features$.feature,
    compute_one_baseline
  )
  saveRDS(baseline_draws, params$baseline_draws_rds, compress = "xz")
}

baseline_summary <- baseline_draws |>
  dplyr::group_by(.data$gene, .data$.feature) |>
  dplyr::summarise(
    n_draws = dplyr::n(),
    n_grid_rows = dplyr::first(.data$n_grid_rows),
    baseline_mean_log_mu = mean(.data$baseline_log_mu),
    baseline_sd_log_mu = stats::sd(.data$baseline_log_mu),
    baseline_q025 = stats::quantile(.data$baseline_log_mu, 0.025),
    baseline_q500 = stats::quantile(.data$baseline_log_mu, 0.5),
    baseline_q975 = stats::quantile(.data$baseline_log_mu, 0.975),
    baseline_median_mu = stats::median(exp(.data$baseline_log_mu)),
    .groups = "drop"
  )

baseline_summary
```

------------------------------------------------------------------------

# Patient pseudobulk and SAVI-aligned offsets

The public capsule provides one processed expression column per GeoMx ROI. As in
the authors' DESeq2 script, values are rounded before count modelling. To retain
patients as the independent units, NK-enriched ROI columns are summed within
each patient x compartment. The resulting five samples form four cohorts:
**IF-REC**, **IF-non-REC**, **TC-REC** and **TC-non-REC**.

For exact SAVI-style scaling, the patient pseudobulks are merged with the single
`V1_nk` training reference sample whose stored offset is zero. We then call
`tidybulk::scale_abundance(method = "TMMwsp", reference_sample = "reference")`
and define `offset = log(1 / multiplier)`, exactly as in the SAVI report.

```{r}
#| label: build-patient-pseudobulk

expression_path <- file.path(
  processed_dir,
  "codeocean_GeoMx_all_gene_expression_matrix.csv"
)
roi_metadata_path <- file.path(
  processed_dir,
  "codeocean_GeoMx_all_roi_metadata.csv"
)

geomx_expression_df <- utils::read.csv(
  expression_path,
  check.names = FALSE,
  stringsAsFactors = FALSE
)
geomx_roi_meta <- readr::read_csv(
  roi_metadata_path,
  show_col_types = FALSE
) |>
  dplyr::filter(
    .data$included_in_author_nk_analysis,
    .data$compartment %in% target_compartments
  ) |>
  dplyr::mutate(
    sample_id = paste(.data$patient_id, .data$compartment, sep = "_"),
    cohort = paste(
      .data$compartment,
      dplyr::if_else(.data$recurrence == "REC", "REC", "non_REC"),
      sep = "_"
    )
  )

geomx_expression <- as.matrix(geomx_expression_df[, -1, drop = FALSE])
mode(geomx_expression) <- "numeric"
rownames(geomx_expression) <- geomx_expression_df$gene

patient_pseudobulk_meta <- geomx_roi_meta |>
  dplyr::distinct(
    .data$sample_id, .data$patient_id, .data$compartment,
    .data$recurrence, .data$cohort, .data$age, .data$gender
  ) |>
  dplyr::arrange(.data$compartment, .data$recurrence, .data$patient_id) |>
  dplyr::mutate(
    n_rois = purrr::map_int(.data$sample_id, function(id) {
      sum(geomx_roi_meta$sample_id == id)
    })
  )

roi_columns_by_sample <- split(
  geomx_roi_meta$author_column,
  geomx_roi_meta$sample_id
)

patient_counts_symbol <- do.call(
  cbind,
  purrr::map(
    patient_pseudobulk_meta$sample_id,
    function(id) {
      values <- round(
        geomx_expression[, roi_columns_by_sample[[id]], drop = FALSE]
      )
      n_non_missing <- rowSums(!is.na(values))
      counts <- rowSums(values, na.rm = TRUE)
      counts[n_non_missing == 0L] <- NA_real_
      counts
    }
  )
)

colnames(patient_counts_symbol) <- patient_pseudobulk_meta$sample_id
rownames(patient_counts_symbol) <- rownames(geomx_expression)

times_panel_samples <- colnames(patient_counts_symbol)[
  colSums(
    !is.na(patient_counts_symbol[times_biomarkers$gene, , drop = FALSE])
  ) == nrow(times_biomarkers)
]
patient_counts_symbol <- patient_counts_symbol[
  , times_panel_samples, drop = FALSE
]
patient_pseudobulk_meta <- patient_pseudobulk_meta |>
  dplyr::filter(.data$sample_id %in% times_panel_samples) |>
  dplyr::arrange(match(.data$sample_id, times_panel_samples))

symbol_to_ensembl <- AnnotationDbi::mapIds(
  org.Hs.eg.db::org.Hs.eg.db,
  keys = rownames(patient_counts_symbol),
  column = "ENSEMBL",
  keytype = "SYMBOL",
  multiVals = "first"
)
keep_mapping <- !is.na(symbol_to_ensembl) & !duplicated(symbol_to_ensembl)
patient_counts <- patient_counts_symbol[keep_mapping, , drop = FALSE]
rownames(patient_counts) <- unname(symbol_to_ensembl[keep_mapping])

readr::write_csv(
  patient_pseudobulk_meta,
  params$patient_pseudobulk_csv
)

patient_pseudobulk_meta
```

```{r}
#| label: merge-training-reference

pseudobulk_sample <- targets::tar_read(
  pseudobulk_sample,
  store = file.path(model_path, "_targets")
)

if (!file.exists(params$pseudobulk_h5ad)) {
  stop("Missing local pseudobulk H5AD: ", params$pseudobulk_h5ad)
}

pseudobulk_sample@assays@data@listData[["counts"]]@seed@seed@seed@filepath <-
  params$pseudobulk_h5ad

reference_idx <- which(
  as.data.frame(SummarizedExperiment::colData(pseudobulk_sample))$offset == 0
)
if (length(reference_idx) != 1L) {
  stop("Expected exactly one V1_nk training sample with offset == 0.")
}

common_features <- intersect(rownames(patient_counts), rownames(pseudobulk_sample))
reference_counts <- as.numeric(
  SummarizedExperiment::assay(pseudobulk_sample, "counts")[
    common_features, reference_idx
  ]
)

counts_mu <- cbind(
  patient_counts[common_features, , drop = FALSE],
  reference = reference_counts
)
complete_features <- rowSums(is.na(counts_mu)) == 0L &
  rowSums(counts_mu, na.rm = TRUE) >= 5
counts_mu <- counts_mu[complete_features, , drop = FALSE]

cohort_levels <- c("IF_REC", "IF_non_REC", "TC_REC", "TC_non_REC")
meta_mu <- dplyr::bind_rows(
  patient_pseudobulk_meta |>
    dplyr::transmute(
      sample_id, patient_id, compartment, recurrence, cohort,
      sample_role = "hcc_patient_pseudobulk"
    ),
  tibble::tibble(
    sample_id = "reference",
    patient_id = NA_character_,
    compartment = NA_character_,
    recurrence = NA_character_,
    cohort = "reference",
    sample_role = "atlas_training_reference"
  )
) |>
  dplyr::mutate(
    group = factor(.data$cohort, levels = c("reference", cohort_levels))
  )

stopifnot(identical(colnames(counts_mu), meta_mu$sample_id))

se_mu <- SummarizedExperiment::SummarizedExperiment(
  assays = list(counts = counts_mu),
  colData = S4Vectors::DataFrame(
    as.data.frame(meta_mu),
    row.names = meta_mu$sample_id
  ),
  rowData = S4Vectors::DataFrame(
    .feature = rownames(counts_mu),
    row.names = rownames(counts_mu)
  )
)

reference_row_data <- as.data.frame(
  SummarizedExperiment::rowData(pseudobulk_sample)
)
abundant_cols <- grep(
  "^\\.abundant",
  names(reference_row_data),
  value = TRUE
)
if (length(abundant_cols) >= 1L) {
  SummarizedExperiment::rowData(se_mu)$.abundant <-
    reference_row_data[
      rownames(se_mu),
      abundant_cols[[1]],
      drop = TRUE
    ]
}

reference_sample_name <- meta_mu$sample_id[
  meta_mu$sample_role == "atlas_training_reference"
]
if (length(reference_sample_name) != 1L) {
  stop("Expected exactly one atlas training reference sample.")
}

se_mu <- se_mu |>
  tidybulk::scale_abundance(
    method = "TMMwsp",
    reference_sample = reference_sample_name
  )
SummarizedExperiment::colData(se_mu)$offset <- log(
  1 / SummarizedExperiment::colData(se_mu)$multiplier
)

counts_mu <- SummarizedExperiment::assay(se_mu, "counts")
meta_mu <- SummarizedExperiment::colData(se_mu) |>
  as.data.frame()
offset_vec_mu <- meta_mu$offset

dge_mu <- edgeR::DGEList(
  counts = counts_mu,
  samples = meta_mu,
  genes = data.frame(.feature = rownames(counts_mu))
)
reference_column <- which(meta_mu$sample_role == "atlas_training_reference")
dge_mu$offset <- matrix(
  offset_vec_mu,
  nrow = nrow(dge_mu),
  ncol = ncol(dge_mu),
  byrow = TRUE
)

normalization_audit <- meta_mu |>
  dplyr::mutate(
    library_size = colSums(counts_mu),
    offset = offset_vec_mu,
    reference_sample = .data$sample_id == reference_sample_name
  )

readr::write_csv(normalization_audit, params$normalization_csv)

normalization_audit |>
  dplyr::select(
    sample_id, patient_id, cohort, sample_role,
    library_size, multiplier, offset, reference_sample
  ) |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), ~round(.x, 4)))
```

------------------------------------------------------------------------

# Healthy-reference cohort estimates

These estimates provide the inputs to the posteriorHCA recurrence proxy. The
SAVI-style group-means design is `~ 0 + group`. For each gene x cohort:

- cohort `log(mu)` at offset zero is the edgeR NB MLE from `mglmOneGroup`;
- cohort uncertainty is the QL Wald SE from `glmQLFit`;
- the healthy comparator is the posterior of the comprehensive liver NK grand
  mean, not the full heterogeneous predictive distribution;
- Welch is the parametric test and posterior rank is the non-parametric check.

```{r}
#| label: edger-fit

design_mu <- stats::model.matrix(~ 0 + group, data = dge_mu$samples)
colnames(design_mu) <- sub("^group", "", colnames(design_mu))

dge_mu <- edgeR::estimateDisp(dge_mu, design_mu)
fit_mu <- edgeR::glmQLFit(dge_mu, design_mu)

nb_dispersion_mu <- if (!is.null(dge_mu$trended.dispersion)) {
  dge_mu$trended.dispersion
} else if (!is.null(dge_mu$tagwise.dispersion)) {
  dge_mu$tagwise.dispersion
} else {
  rep_len(dge_mu$common.dispersion, nrow(dge_mu))
}

tibble::tibble(
  n_features_fit = nrow(dge_mu),
  n_patient_pseudobulks = sum(meta_mu$sample_role == "hcc_patient_pseudobulk"),
  design_groups = paste(colnames(design_mu), collapse = ", "),
  residual_df = paste(unique(fit_mu$df.residual), collapse = ", "),
  reference_offset = offset_vec_mu[reference_column]
)
```

```{r}
#| label: savi-test-helpers

welch_t_from_mean_se <- function(mean1, se1, n1, mean2, se2, n2) {
  sd1 <- se1 * sqrt(n1)
  sd2 <- se2 * sqrt(n2)
  se_diff <- sqrt(se1^2 + se2^2)
  t_stat <- (mean1 - mean2) / se_diff
  df <- (sd1^2 / n1 + sd2^2 / n2)^2 /
    ((sd1^2 / n1)^2 / (n1 - 1) + (sd2^2 / n2)^2 / (n2 - 1))
  p_value <- 2 * stats::pt(abs(t_stat), df = df, lower.tail = FALSE)
  list(
    mean_diff = mean1 - mean2,
    se_diff = se_diff,
    t = t_stat,
    df = df,
    p_value = p_value
  )
}

edger_log_mu_cohort <- function(feature_id, cohort) {
  gi <- match(feature_id, rownames(fit_mu))
  if (is.na(gi) || !cohort %in% colnames(fit_mu$coefficients)) {
    return(NULL)
  }

  j <- which(fit_mu$samples$group == cohort)
  if (!length(j)) {
    return(NULL)
  }

  dispersion <- nb_dispersion_mu[gi]
  log_mu <- edgeR::mglmOneGroup(
    matrix(counts_mu[gi, j], nrow = 1),
    dispersion = dispersion,
    offset = offset_vec_mu[j]
  )

  mu_all <- fit_mu$fitted.values[gi, ]
  weights <- mu_all / (1 + dispersion * mu_all)
  cohort_weight <- sum(weights[j])
  se <- if (is.finite(cohort_weight) && cohort_weight > 0) {
    sqrt(fit_mu$s2.post[gi] / cohort_weight)
  } else {
    Inf
  }

  list(
    log_mu = unname(log_mu),
    mu = unname(exp(log_mu)),
    se = unname(se),
    df = unname(fit_mu$df.residual.adj[gi]),
    n = length(j),
    patient_ids = paste(
      sort(stats::na.omit(fit_mu$samples$patient_id[j])),
      collapse = ", "
    )
  )
}

test_cohort_against_baseline <- function(gene, feature_id, cohort) {
  cm <- edger_log_mu_cohort(feature_id, cohort)
  if (is.null(cm)) {
    return(NULL)
  }

  baseline <- baseline_summary |>
    dplyr::filter(.data$gene == gene, .data$.feature == feature_id)
  draws <- baseline_draws |>
    dplyr::filter(.data$gene == gene, .data$.feature == feature_id) |>
    dplyr::pull(.data$baseline_log_mu)

  welch <- welch_t_from_mean_se(
    mean1 = baseline$baseline_mean_log_mu,
    se1 = baseline$baseline_sd_log_mu,
    n1 = baseline$n_draws,
    mean2 = cm$log_mu,
    se2 = cm$se,
    n2 = cm$df + 1
  )
  rank_p_upper <- mean(draws >= cm$log_mu)
  rank_p_lower <- mean(draws <= cm$log_mu)

  tibble::tibble(
    gene = gene,
    .feature = feature_id,
    cohort = cohort,
    compartment = sub("_.*$", "", cohort),
    recurrence = dplyr::if_else(
      grepl("non_REC$", cohort), "non-REC", "REC"
    ),
    patient_ids = cm$patient_ids,
    cohort_n = cm$n,
    cohort_log_mu = cm$log_mu,
    cohort_mu = cm$mu,
    cohort_se = cm$se,
    baseline_mean_log_mu = baseline$baseline_mean_log_mu,
    baseline_sd_grand_mean = baseline$baseline_sd_log_mu,
    delta_log_mu_vs_baseline =
      cm$log_mu - baseline$baseline_mean_log_mu,
    welch_t = welch$t,
    welch_df = welch$df,
    welch_p = welch$p_value,
    rank_p_upper = rank_p_upper,
    rank_p_lower = rank_p_lower,
    rank_p_two = min(1, 2 * min(rank_p_upper, rank_p_lower)),
    direction_vs_baseline = dplyr::case_when(
      cm$log_mu > baseline$baseline_mean_log_mu ~ "above_baseline",
      cm$log_mu < baseline$baseline_mean_log_mu ~ "below_baseline",
      TRUE ~ "at_baseline"
    ),
    significant_welch_05 = welch$p_value < 0.05
  )
}
```

```{r}
#| label: four-cohort-tests

baseline_features <- biomarker_model_status |>
  dplyr::filter(
    .data$in_v1_nk_store,
    .data$.feature %in% rownames(fit_mu)
  ) |>
  dplyr::select(gene, .feature)

cohort_tests <- tidyr::crossing(
  baseline_features,
  cohort = cohort_levels
) |>
  purrr::pmap_dfr(function(gene, .feature, cohort) {
    test_cohort_against_baseline(gene, .feature, cohort)
  }) |>
  dplyr::mutate(
    cohort = factor(.data$cohort, levels = cohort_levels)
  ) |>
  dplyr::arrange(.data$gene, .data$cohort)

readr::write_csv(cohort_tests, params$cohort_results_csv)

cohort_tests |>
  dplyr::select(
    gene, cohort, patient_ids, cohort_n, cohort_log_mu, cohort_se,
    baseline_mean_log_mu, delta_log_mu_vs_baseline,
    welch_p, rank_p_two, direction_vs_baseline
  ) |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), ~round(.x, 4)))
```

```{r}
#| label: four-cohort-summary

cohort_test_summary <- cohort_tests |>
  dplyr::group_by(.data$cohort) |>
  dplyr::summarise(
    n_genes_tested = dplyr::n_distinct(.data$gene),
    n_above_baseline = sum(.data$direction_vs_baseline == "above_baseline"),
    n_below_baseline = sum(.data$direction_vs_baseline == "below_baseline"),
    n_welch_p_lt_05 = sum(.data$welch_p < 0.05),
    median_delta_log_mu = stats::median(
      .data$delta_log_mu_vs_baseline, na.rm = TRUE
    ),
    .groups = "drop"
  )

cohort_test_summary
```

------------------------------------------------------------------------

# Positive control: non-REC greater than REC within IF and TC

Recurrence is tested with edgeR quasi-likelihood contrasts from the same
patient-level counts and offsets. A separate HCC-only fit retains `HLA-DRB1`,
which has no matching atlas feature/model for the healthy-baseline comparison.
The contrast is `non_REC - REC` within each compartment.
`p_nonrec_greater` is the pre-specified one-sided p-value; positive log2 fold
change supports the paper direction. This positive control is compared directly
with paper ground truth. These tests are necessarily low-powered:
the public NK-enriched subset contains two non-REC and one REC patient in IF,
and one patient in each recurrence arm in TC.

```{r}
#| label: recurrence-edger-fit

recurrence_complete_features <- rowSums(is.na(patient_counts)) == 0L &
  rowSums(patient_counts, na.rm = TRUE) >= 5
counts_recurrence <- patient_counts[
  recurrence_complete_features, , drop = FALSE
]

meta_recurrence <- patient_pseudobulk_meta |>
  dplyr::mutate(
    group = factor(.data$cohort, levels = cohort_levels)
  )
stopifnot(identical(colnames(counts_recurrence), meta_recurrence$sample_id))

dge_recurrence <- edgeR::DGEList(
  counts = counts_recurrence,
  samples = as.data.frame(meta_recurrence),
  genes = data.frame(.feature = rownames(counts_recurrence))
)
offset_vec_recurrence <- offset_vec_mu[
  match(meta_recurrence$sample_id, meta_mu$sample_id)
]
dge_recurrence$offset <- matrix(
  offset_vec_recurrence,
  nrow = nrow(dge_recurrence),
  ncol = ncol(dge_recurrence),
  byrow = TRUE
)

design_recurrence <- stats::model.matrix(
  ~ 0 + group,
  data = dge_recurrence$samples
)
colnames(design_recurrence) <- sub(
  "^group", "", colnames(design_recurrence)
)
dge_recurrence <- edgeR::estimateDisp(
  dge_recurrence,
  design_recurrence
)
fit_recurrence <- edgeR::glmQLFit(
  dge_recurrence,
  design_recurrence
)

nb_dispersion_recurrence <- if (
  !is.null(dge_recurrence$trended.dispersion)
) {
  dge_recurrence$trended.dispersion
} else if (!is.null(dge_recurrence$tagwise.dispersion)) {
  dge_recurrence$tagwise.dispersion
} else {
  rep_len(
    dge_recurrence$common.dispersion,
    nrow(dge_recurrence)
  )
}

tibble::tibble(
  n_features_fit = nrow(fit_recurrence),
  n_patient_pseudobulks = nrow(fit_recurrence$samples),
  design_groups = paste(colnames(design_recurrence), collapse = ", "),
  residual_df = paste(unique(fit_recurrence$df.residual), collapse = ", "),
  times_genes_available = paste(
    times_biomarkers$gene[
      times_biomarkers$.feature %in% rownames(fit_recurrence)
    ],
    collapse = ", "
  )
)
```

```{r}
#| label: recurrence-contrast-helpers

edger_log_mu_recurrence <- function(feature_id, cohort) {
  gi <- match(feature_id, rownames(fit_recurrence))
  if (is.na(gi)) {
    return(NULL)
  }
  j <- which(fit_recurrence$samples$group == cohort)
  if (!length(j)) {
    return(NULL)
  }

  log_mu <- edgeR::mglmOneGroup(
    matrix(counts_recurrence[gi, j], nrow = 1),
    dispersion = nb_dispersion_recurrence[gi],
    offset = offset_vec_recurrence[j]
  )

  list(
    log_mu = unname(log_mu),
    mu = unname(exp(log_mu)),
    n = length(j)
  )
}

test_nonrec_greater <- function(gene, feature_id, compartment) {
  gi <- match(feature_id, rownames(fit_recurrence))
  nonrec_group <- paste0(compartment, "_non_REC")
  rec_group <- paste0(compartment, "_REC")
  if (
    is.na(gi) ||
    !all(
      c(nonrec_group, rec_group) %in%
        colnames(fit_recurrence$coefficients)
    )
  ) {
    return(NULL)
  }

  contrast <- stats::setNames(
    rep(0, ncol(fit_recurrence$design)),
    colnames(fit_recurrence$design)
  )
  contrast[nonrec_group] <- 1
  contrast[rec_group] <- -1

  qlf <- edgeR::glmQLFTest(fit_recurrence, contrast = contrast)
  qlf_row <- qlf$table[gi, , drop = FALSE]
  p_two <- qlf_row$PValue
  p_greater <- if (qlf_row$logFC >= 0) p_two / 2 else 1 - p_two / 2

  nonrec_mu <- edger_log_mu_recurrence(feature_id, nonrec_group)
  rec_mu <- edger_log_mu_recurrence(feature_id, rec_group)

  tibble::tibble(
    gene = gene,
    .feature = feature_id,
    compartment = compartment,
    nonrec_group = nonrec_group,
    rec_group = rec_group,
    n_nonrec = nonrec_mu$n,
    n_rec = rec_mu$n,
    nonrec_log_mu = nonrec_mu$log_mu,
    rec_log_mu = rec_mu$log_mu,
    delta_log_mu_nonrec_minus_rec =
      nonrec_mu$log_mu - rec_mu$log_mu,
    qlf_log2fc_nonrec_vs_rec = qlf_row$logFC,
    qlf_F = qlf_row$F,
    qlf_p_two_sided = p_two,
    p_nonrec_greater = p_greater,
    observed_recurrence_direction = dplyr::case_when(
      qlf_row$logFC > 0 ~ "non-REC higher",
      qlf_row$logFC < 0 ~ "REC higher",
      TRUE ~ "tie"
    ),
    significant_nonrec_greater_05 =
      qlf_row$logFC > 0 & p_greater < 0.05
  )
}
```

```{r}
#| label: recurrence-tests

recurrence_features <- times_biomarkers |>
  dplyr::filter(.data$.feature %in% rownames(fit_recurrence)) |>
  dplyr::select(gene, .feature)

recurrence_tests <- tidyr::crossing(
  recurrence_features,
  compartment = target_compartments
) |>
  purrr::pmap_dfr(function(gene, .feature, compartment) {
    test_nonrec_greater(gene, .feature, compartment)
  }) |>
  dplyr::left_join(
    paper_ground_truth |>
      dplyr::select(
        gene,
        expected_recurrence_direction_if,
        expected_recurrence_direction_tc,
        paper_l2fc_nonrec_vs_rec_if,
        paper_l2fc_nonrec_vs_rec_tc,
        paper_padj_nonrec_vs_rec_if,
        paper_padj_nonrec_vs_rec_tc
      ),
    by = "gene"
  ) |>
  dplyr::mutate(
    expected_recurrence_direction = dplyr::if_else(
      .data$compartment == "IF",
      .data$expected_recurrence_direction_if,
      .data$expected_recurrence_direction_tc
    ),
    paper_l2fc_nonrec_vs_rec = dplyr::if_else(
      .data$compartment == "IF",
      .data$paper_l2fc_nonrec_vs_rec_if,
      .data$paper_l2fc_nonrec_vs_rec_tc
    ),
    paper_padj_nonrec_vs_rec = dplyr::if_else(
      .data$compartment == "IF",
      .data$paper_padj_nonrec_vs_rec_if,
      .data$paper_padj_nonrec_vs_rec_tc
    ),
    recurrence_matches_paper =
      .data$observed_recurrence_direction ==
      .data$expected_recurrence_direction
  ) |>
  dplyr::arrange(.data$gene, .data$compartment)

readr::write_csv(recurrence_tests, params$recurrence_results_csv)

recurrence_tests |>
  dplyr::select(
    gene, compartment, n_nonrec, n_rec,
    qlf_log2fc_nonrec_vs_rec, p_nonrec_greater,
    observed_recurrence_direction, expected_recurrence_direction,
    recurrence_matches_paper
  ) |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), ~round(.x, 4)))
```

------------------------------------------------------------------------

# Primary posteriorHCA test: healthy-proximity recurrence proxy

The proxy asks whether recurrence can be recovered using only distance from the
healthy atlas. No recurrence labels, paper coefficients or HCC-trained model
parameters enter score construction.

We evaluate four properties:

1. **Compatibility split:** REC is more than one healthy SD away while non-REC
   remains within one healthy SD.
2. **Gene-compartment consistency:** REC distance exceeds non-REC distance for
   each atlas-covered biomarker in IF and TC.
3. **Posterior evidence:** after propagating healthy posterior and cohort QL
   uncertainty, the combined REC score remains larger than the non-REC score.
4. **Patient ranking:** the outcome-free score ranks REC above non-REC patients,
   including leave-one-gene-out sensitivity analyses.

The one-SD threshold is a descriptive compatibility rule, not a significance
threshold. Welch and rank p-values remain available in `cohort_tests`.

```{r}
#| label: healthy-proxy-gene-compartment

healthy_proxy_gene_long <- cohort_tests |>
  dplyr::mutate(
    signed_healthy_z =
      .data$delta_log_mu_vs_baseline / .data$baseline_sd_grand_mean,
    healthy_distance = abs(.data$signed_healthy_z),
    within_one_healthy_sd = .data$healthy_distance <= 1,
    recurrence_key = dplyr::if_else(
      .data$recurrence == "REC", "REC", "non_REC"
    )
  )

healthy_proxy_gene <- healthy_proxy_gene_long |>
  dplyr::select(
    gene, .feature, compartment, recurrence_key,
    cohort_log_mu, cohort_se, baseline_mean_log_mu,
    baseline_sd_grand_mean, signed_healthy_z, healthy_distance,
    within_one_healthy_sd
  ) |>
  tidyr::pivot_wider(
    names_from = recurrence_key,
    values_from = c(
      cohort_log_mu, cohort_se, signed_healthy_z,
      healthy_distance, within_one_healthy_sd
    ),
    names_sep = "."
  ) |>
  dplyr::mutate(
    proxy_margin_rec_minus_nonrec =
      .data$healthy_distance.REC - .data$healthy_distance.non_REC,
    rec_farther_from_healthy =
      .data$proxy_margin_rec_minus_nonrec > 0,
    proxy_pattern_matches_hypothesis =
      !.data$within_one_healthy_sd.REC &
      .data$within_one_healthy_sd.non_REC
  ) |>
  dplyr::arrange(.data$gene, .data$compartment)

readr::write_csv(healthy_proxy_gene, params$healthy_proxy_gene_csv)

healthy_proxy_gene |>
  dplyr::select(
    gene, compartment,
    healthy_distance.REC, healthy_distance.non_REC,
    proxy_margin_rec_minus_nonrec,
    rec_farther_from_healthy,
    proxy_pattern_matches_hypothesis
  ) |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), ~round(.x, 3)))
```

```{r}
#| label: healthy-proxy-posterior

n_proxy_draws <- 20000L
set.seed(20250625)

simulate_proxy_gene <- function(gene_name) {
  healthy_draws <- baseline_draws |>
    dplyr::filter(.data$gene == .env$gene_name) |>
    dplyr::pull(.data$baseline_log_mu)
  healthy_sd <- stats::sd(healthy_draws)
  healthy_draw <- sample(
    healthy_draws,
    size = n_proxy_draws,
    replace = TRUE
  )

  purrr::map_dfr(target_compartments, function(compartment_name) {
    rec_row <- cohort_tests |>
      dplyr::filter(
        .data$gene == .env$gene_name,
        .data$compartment == .env$compartment_name,
        .data$recurrence == "REC"
      )
    nonrec_row <- cohort_tests |>
      dplyr::filter(
        .data$gene == .env$gene_name,
        .data$compartment == .env$compartment_name,
        .data$recurrence == "non-REC"
      )
    rec_draw <- stats::rnorm(
      n_proxy_draws,
      mean = rec_row$cohort_log_mu,
      sd = rec_row$cohort_se
    )
    nonrec_draw <- stats::rnorm(
      n_proxy_draws,
      mean = nonrec_row$cohort_log_mu,
      sd = nonrec_row$cohort_se
    )

    tibble::tibble(
      draw = seq_len(n_proxy_draws),
      gene = gene_name,
      compartment = compartment_name,
      rec_score = abs(rec_draw - healthy_draw) / healthy_sd,
      nonrec_score = abs(nonrec_draw - healthy_draw) / healthy_sd,
      margin = rec_score - nonrec_score
    )
  })
}

proxy_draws <- purrr::map_dfr(
  unique(cohort_tests$gene),
  simulate_proxy_gene
)

summarise_proxy_draws <- function(tbl, scope, compartment = NA_character_) {
  tibble::tibble(
    scope = scope,
    compartment = compartment,
    n_genes = dplyr::n_distinct(cohort_tests$gene),
    mean_rec_score = mean(tbl$rec_score),
    mean_nonrec_score = mean(tbl$nonrec_score),
    mean_margin = mean(tbl$margin),
    margin_q025 = stats::quantile(tbl$margin, 0.025),
    margin_q500 = stats::quantile(tbl$margin, 0.5),
    margin_q975 = stats::quantile(tbl$margin, 0.975),
    posterior_prob_rec_farther = mean(tbl$margin > 0),
    posterior_prob_rec_beyond_one_sd = mean(tbl$rec_score > 1),
    posterior_prob_nonrec_within_one_sd = mean(tbl$nonrec_score <= 1),
    posterior_prob_proxy_pattern = mean(
      tbl$rec_score > 1 & tbl$nonrec_score <= 1
    )
  )
}

proxy_gene_posterior_summary <- proxy_draws |>
  dplyr::group_by(.data$gene, .data$compartment) |>
  dplyr::summarise(
    mean_rec_score = mean(.data$rec_score),
    mean_nonrec_score = mean(.data$nonrec_score),
    mean_margin = mean(.data$margin),
    margin_q025 = stats::quantile(.data$margin, 0.025),
    margin_q500 = stats::quantile(.data$margin, 0.5),
    margin_q975 = stats::quantile(.data$margin, 0.975),
    posterior_prob_rec_farther = mean(.data$margin > 0),
    posterior_prob_rec_beyond_one_sd = mean(.data$rec_score > 1),
    posterior_prob_nonrec_within_one_sd = mean(.data$nonrec_score <= 1),
    posterior_prob_proxy_pattern = mean(
      .data$rec_score > 1 & .data$nonrec_score <= 1
    ),
    .groups = "drop"
  ) |>
  dplyr::mutate(scope = "gene_compartment", .before = 1L)

proxy_compartment_draws <- proxy_draws |>
  dplyr::group_by(.data$compartment, .data$draw) |>
  dplyr::summarise(
    rec_score = mean(.data$rec_score),
    nonrec_score = mean(.data$nonrec_score),
    margin = .data$rec_score - .data$nonrec_score,
    .groups = "drop"
  )

proxy_compartment_summary <- proxy_compartment_draws |>
  dplyr::group_by(.data$compartment) |>
  dplyr::summarise(
    scope = "compartment",
    n_genes = dplyr::n_distinct(cohort_tests$gene),
    mean_rec_score = mean(.data$rec_score),
    mean_nonrec_score = mean(.data$nonrec_score),
    mean_margin = mean(.data$margin),
    margin_q025 = stats::quantile(.data$margin, 0.025),
    margin_q500 = stats::quantile(.data$margin, 0.5),
    margin_q975 = stats::quantile(.data$margin, 0.975),
    posterior_prob_rec_farther = mean(.data$margin > 0),
    posterior_prob_rec_beyond_one_sd = mean(.data$rec_score > 1),
    posterior_prob_nonrec_within_one_sd = mean(.data$nonrec_score <= 1),
    posterior_prob_proxy_pattern = mean(
      .data$rec_score > 1 & .data$nonrec_score <= 1
    ),
    .groups = "drop"
  ) |>
  dplyr::relocate(scope)

proxy_combined_draws <- proxy_draws |>
  dplyr::group_by(.data$draw) |>
  dplyr::summarise(
    rec_score = mean(.data$rec_score),
    nonrec_score = mean(.data$nonrec_score),
    margin = .data$rec_score - .data$nonrec_score,
    .groups = "drop"
  )

proxy_combined_summary <- summarise_proxy_draws(
  proxy_combined_draws,
  scope = "combined_IF_TC"
)

healthy_proxy_posterior_summary <- dplyr::bind_rows(
  proxy_gene_posterior_summary,
  proxy_compartment_summary,
  proxy_combined_summary
)

readr::write_csv(
  healthy_proxy_posterior_summary,
  params$healthy_proxy_summary_csv
)

healthy_proxy_posterior_summary |>
  dplyr::filter(.data$scope != "gene_compartment") |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), ~round(.x, 3)))
```

```{r}
#| label: healthy-proxy-patient-scores

patient_sample_indices <- which(
  meta_mu$sample_role == "hcc_patient_pseudobulk"
)

patient_gene_proxy <- tidyr::crossing(
  sample_index = patient_sample_indices,
  baseline_features
) |>
  purrr::pmap_dfr(function(sample_index, gene, .feature) {
    feature_id <- .feature
    gene_index <- match(feature_id, rownames(fit_mu))
    patient_log_mu <- edgeR::mglmOneGroup(
      matrix(counts_mu[gene_index, sample_index], nrow = 1),
      dispersion = nb_dispersion_mu[gene_index],
      offset = offset_vec_mu[sample_index]
    )
    healthy <- baseline_summary |>
      dplyr::filter(
        .data$gene == .env$gene,
        .data$.feature == .env$feature_id
      )
    signed_healthy_z <- (
      as.numeric(patient_log_mu) - healthy$baseline_mean_log_mu
    ) / healthy$baseline_sd_log_mu

    tibble::tibble(
      sample_id = meta_mu$sample_id[sample_index],
      patient_id = meta_mu$patient_id[sample_index],
      compartment = meta_mu$compartment[sample_index],
      recurrence = meta_mu$recurrence[sample_index],
      gene = gene,
      .feature = feature_id,
      patient_log_mu = as.numeric(patient_log_mu),
      baseline_mean_log_mu = healthy$baseline_mean_log_mu,
      baseline_sd_grand_mean = healthy$baseline_sd_log_mu,
      signed_healthy_z = signed_healthy_z,
      healthy_distance = abs(signed_healthy_z)
    )
  })

patient_compartment_proxy <- patient_gene_proxy |>
  dplyr::group_by(
    .data$sample_id, .data$patient_id,
    .data$compartment, .data$recurrence
  ) |>
  dplyr::summarise(
    n_genes = dplyr::n(),
    healthy_proximity_risk = mean(.data$healthy_distance),
    mean_signed_healthy_z = mean(.data$signed_healthy_z),
    .groups = "drop"
  )

patient_combined_proxy <- patient_compartment_proxy |>
  dplyr::group_by(.data$patient_id, .data$recurrence) |>
  dplyr::summarise(
    n_compartments = dplyr::n(),
    healthy_proximity_risk = mean(.data$healthy_proximity_risk),
    mean_signed_healthy_z = mean(.data$mean_signed_healthy_z),
    .groups = "drop"
  )

patient_proxy_output <- dplyr::bind_rows(
  patient_gene_proxy |>
    dplyr::transmute(
      score_level = "gene_compartment",
      sample_id, patient_id, compartment, recurrence,
      feature = .data$gene,
      healthy_proximity_risk = .data$healthy_distance,
      mean_signed_healthy_z = .data$signed_healthy_z
    ),
  patient_compartment_proxy |>
    dplyr::transmute(
      score_level = "patient_compartment",
      sample_id, patient_id, compartment, recurrence,
      feature = "four_gene_mean",
      healthy_proximity_risk, mean_signed_healthy_z
    ),
  patient_combined_proxy |>
    dplyr::transmute(
      score_level = "patient_combined",
      sample_id = NA_character_,
      patient_id, compartment = "IF_TC_available", recurrence,
      feature = "four_gene_compartment_mean",
      healthy_proximity_risk, mean_signed_healthy_z
    )
)

readr::write_csv(patient_proxy_output, params$healthy_proxy_patient_csv)

patient_compartment_proxy |>
  dplyr::arrange(.data$compartment, .data$healthy_proximity_risk)
```

```{r}
#| label: healthy-proxy-performance

rank_auc <- function(recurrence, score) {
  rec <- score[recurrence == "REC"]
  nonrec <- score[recurrence == "non-REC"]
  comparisons <- outer(rec, nonrec, "-")
  mean(comparisons > 0) + 0.5 * mean(comparisons == 0)
}

proxy_performance_one <- function(tbl, method, scope, left_out_gene = NA_character_) {
  rec <- tbl$healthy_proximity_risk[tbl$recurrence == "REC"]
  nonrec <- tbl$healthy_proximity_risk[tbl$recurrence == "non-REC"]
  rank_test <- suppressWarnings(
    stats::wilcox.test(
      rec,
      nonrec,
      alternative = "greater",
      exact = TRUE
    )
  )

  tibble::tibble(
    method = method,
    scope = scope,
    left_out_gene = left_out_gene,
    n_rec = length(rec),
    n_nonrec = length(nonrec),
    auc = rank_auc(tbl$recurrence, tbl$healthy_proximity_risk),
    one_sided_rank_p = rank_test$p.value,
    min_rec_minus_max_nonrec = min(rec) - max(nonrec),
    all_rec_above_nonrec = min(rec) > max(nonrec)
  )
}

proxy_performance <- dplyr::bind_rows(
  purrr::map_dfr(target_compartments, function(compartment) {
    compartment_name <- compartment
    proxy_performance_one(
      patient_compartment_proxy |>
        dplyr::filter(.data$compartment == .env$compartment_name),
      method = "posteriorHCA_healthy_proximity",
      scope = compartment_name
    )
  }),
  proxy_performance_one(
    patient_combined_proxy,
    method = "posteriorHCA_healthy_proximity",
    scope = "combined_available_IF_TC"
  )
)

proxy_leave_one_gene_out <- purrr::map_dfr(
  unique(patient_gene_proxy$gene),
  function(left_out_gene) {
    loo_compartment <- patient_gene_proxy |>
      dplyr::filter(.data$gene != left_out_gene) |>
      dplyr::group_by(
        .data$patient_id, .data$compartment, .data$recurrence
      ) |>
      dplyr::summarise(
        healthy_proximity_risk = mean(.data$healthy_distance),
        .groups = "drop"
      ) |>
      dplyr::group_by(.data$patient_id, .data$recurrence) |>
      dplyr::summarise(
        healthy_proximity_risk = mean(.data$healthy_proximity_risk),
        .groups = "drop"
      )

    proxy_performance_one(
      loo_compartment,
      method = "posteriorHCA_healthy_proximity_LOO_gene",
      scope = "combined_available_IF_TC",
      left_out_gene = left_out_gene
    )
  }
)

healthy_proxy_performance <- dplyr::bind_rows(
  proxy_performance,
  proxy_leave_one_gene_out
)

readr::write_csv(
  healthy_proxy_performance,
  params$healthy_proxy_performance_csv
)

healthy_proxy_performance
```

```{r}
#| label: paper-times-benchmark

paper_times_path <- file.path(
  params$case_dir,
  "codeocean",
  "data",
  "Input_2_TIMES_development_from_mIHC",
  "TIMES_distribution_input.RData"
)

paper_times_env <- new.env(parent = emptyenv())
load(paper_times_path, envir = paper_times_env)
paper_rec_scores <- get("liver_r_data_gene_ed", envir = paper_times_env)
paper_nonrec_scores <- get("liver_nr_data_gene_ed", envir = paper_times_env)
paper_rank_test <- stats::wilcox.test(
  paper_rec_scores,
  paper_nonrec_scores,
  alternative = "greater",
  exact = FALSE
)

paper_times_benchmark <- tibble::tibble(
  method = "paper_TIMES_validation_score",
  scope = "mIHC_validation_subset",
  n_rec = length(paper_rec_scores),
  n_nonrec = length(paper_nonrec_scores),
  auc = mean(outer(paper_rec_scores, paper_nonrec_scores, ">")) +
    0.5 * mean(outer(paper_rec_scores, paper_nonrec_scores, "==")),
  one_sided_rank_p = paper_rank_test$p.value,
  comparison_note = paste(
    "Descriptive benchmark from the CodeOcean TIMES distribution input.",
    "Not a formal equivalence test against the GeoMx posteriorHCA proxy."
  )
)

readr::write_csv(paper_times_benchmark, params$paper_benchmark_csv)

dplyr::bind_rows(
  healthy_proxy_performance |>
    dplyr::filter(
      .data$method == "posteriorHCA_healthy_proximity",
      .data$scope == "combined_available_IF_TC"
    ) |>
    dplyr::transmute(
      method, scope, n_rec, n_nonrec, auc, one_sided_rank_p,
      comparison_note =
        "Outcome-free posteriorHCA proxy; only three available patients."
    ),
  paper_times_benchmark
)
```

------------------------------------------------------------------------

# Visual checks

```{r}
#| label: plot-healthy-proxy-cohort
#| fig-width: 9
#| fig-height: 5
#| fig-align: center

ggplot2::ggplot(
  healthy_proxy_gene_long,
  ggplot2::aes(
    x = .data$recurrence,
    y = .data$healthy_distance,
    colour = .data$recurrence,
    group = interaction(.data$gene, .data$compartment)
  )
) +
  ggplot2::geom_hline(
    yintercept = 1,
    linetype = "dashed",
    colour = "grey45"
  ) +
  ggplot2::geom_line(
    colour = "grey70",
    linewidth = 0.45
  ) +
  ggplot2::geom_point(size = 2.8) +
  ggplot2::facet_wrap(~ compartment) +
  ggplot2::scale_colour_manual(
    values = c(`non-REC` = "#2878B5", REC = "#C23B4A")
  ) +
  ggplot2::labs(
    title = "posteriorHCA healthy-proximity recurrence proxy",
    subtitle = "Every line is one biomarker; dashed line marks one healthy grand-mean SD",
    x = NULL,
    y = "Absolute standardized distance from healthy liver NK",
    colour = "Outcome"
  ) +
  ggplot2::theme(legend.position = "bottom")
```

```{r}
#| label: plot-healthy-proxy-patients
#| fig-width: 8
#| fig-height: 4.5
#| fig-align: center

ggplot2::ggplot(
  patient_compartment_proxy,
  ggplot2::aes(
    x = .data$compartment,
    y = .data$healthy_proximity_risk,
    colour = .data$recurrence,
    label = .data$patient_id
  )
) +
  ggplot2::geom_hline(
    yintercept = 1,
    linetype = "dashed",
    colour = "grey45"
  ) +
  ggplot2::geom_point(size = 3) +
  ggplot2::geom_text(
    nudge_x = 0.08,
    hjust = 0,
    size = 3.2,
    show.legend = FALSE
  ) +
  ggplot2::scale_colour_manual(
    values = c(`non-REC` = "#2878B5", REC = "#C23B4A")
  ) +
  ggplot2::labs(
    title = "Outcome-free patient scores",
    subtitle = "Mean absolute healthy distance across four atlas-covered TIMES genes",
    x = NULL,
    y = "Mean absolute distance from healthy (SD units)",
    colour = "Outcome"
  ) +
  ggplot2::theme(legend.position = "bottom")
```

**How to read this patient-level plot.** Each point is one patient in one
spatial compartment, rather than a cohort estimate. Its vertical position is
the mean of the four gene-specific absolute distances from the healthy liver NK
grand mean, measured in healthy-posterior SD units. The score is therefore a
reference-deviation index, not a recurrence probability or a model fitted to
the recurrence labels. The dashed line at 1 is a descriptive compatibility
boundary: a score above it means that the patient's average biomarker distance
exceeds one healthy-reference SD. Outcome colours and labels are added only
after the scores have been calculated.

In these data, A29, the only REC patient, lies above the one-SD boundary in
both IF and TC, whereas all available non-REC observations lie below it (A6
and A24 nearly overlap in IF). This produces perfect observed patient ordering
and a descriptive AUC of 1.0, but it is proof of concept rather than a stable
performance estimate because the evaluation contains only three patients and
one REC event. The posterior margin plot below carries the baseline and cohort
estimation uncertainty that is not displayed by these point estimates.

```{r}
#| label: plot-healthy-proxy-posterior-margin
#| fig-width: 8
#| fig-height: 4.5
#| fig-align: center

proxy_margin_plot <- dplyr::bind_rows(
  proxy_compartment_draws |>
    dplyr::transmute(
      scope = as.character(.data$compartment),
      margin = .data$margin
    ),
  proxy_combined_draws |>
    dplyr::transmute(
      scope = "IF + TC",
      margin = .data$margin
    )
) |>
  dplyr::mutate(scope = factor(.data$scope, levels = c("IF", "TC", "IF + TC")))

ggplot2::ggplot(
  proxy_margin_plot,
  ggplot2::aes(x = .data$margin, fill = .data$scope)
) +
  ggplot2::geom_density(alpha = 0.45, colour = NA) +
  ggplot2::geom_vline(xintercept = 0, colour = "grey35", linewidth = 0.5) +
  ggplot2::facet_wrap(~ scope, scales = "free_y", ncol = 1) +
  ggplot2::scale_fill_manual(
    values = c(IF = "#2878B5", TC = "#E08A2E", `IF + TC` = "#6A4C93")
  ) +
  ggplot2::labs(
    title = "Posterior proxy margin",
    subtitle = "Positive margin means REC is farther from healthy than non-REC",
    x = "REC score - non-REC score",
    y = "Density",
    fill = NULL
  ) +
  ggplot2::theme(legend.position = "none")
```

```{r}
#| label: plot-baseline-density-cohorts
#| fig-width: 10
#| fig-height: 7
#| fig-align: center

gene_order <- cohort_tests |>
  dplyr::group_by(.data$gene) |>
  dplyr::summarise(
    median_delta = stats::median(.data$delta_log_mu_vs_baseline),
    .groups = "drop"
  ) |>
  dplyr::arrange(.data$median_delta) |>
  dplyr::pull(.data$gene)

cohort_colours <- c(
  IF_REC = "#C23B4A",
  IF_non_REC = "#2878B5",
  TC_REC = "#E08A2E",
  TC_non_REC = "#2D8B57"
)

ggplot2::ggplot() +
  ggplot2::geom_density(
    data = baseline_draws |>
      dplyr::mutate(gene = factor(.data$gene, levels = gene_order)),
    ggplot2::aes(x = .data$baseline_log_mu),
    fill = "grey82",
    colour = "grey35",
    alpha = 0.65,
    linewidth = 0.35
  ) +
  ggplot2::geom_vline(
    data = cohort_tests |>
      dplyr::mutate(gene = factor(.data$gene, levels = gene_order)),
    ggplot2::aes(
      xintercept = .data$cohort_log_mu,
      colour = .data$cohort
    ),
    linewidth = 0.85
  ) +
  ggplot2::scale_colour_manual(values = cohort_colours) +
  ggplot2::facet_wrap(~ gene, scales = "free", ncol = 2) +
  ggplot2::labs(
    title = "Healthy liver NK grand mean vs four HCC cohorts",
    subtitle = "Grey = posterior grand-mean log(mu); lines = edgeR cohort NB log(mu) at offset 0",
    x = "log(mu)",
    y = "Density",
    colour = "Cohort"
  ) +
  ggplot2::theme(
    legend.position = "bottom",
    strip.text = ggplot2::element_text(face = "bold")
  )
```

```{r}
#| label: plot-cohort-heatmap
#| fig-width: 9
#| fig-height: 4.5
#| fig-align: center

ggplot2::ggplot(
  cohort_tests,
  ggplot2::aes(
    x = .data$cohort,
    y = forcats::fct_rev(factor(.data$gene, levels = gene_order)),
    fill = .data$delta_log_mu_vs_baseline
  )
) +
  ggplot2::geom_tile(colour = "white", linewidth = 0.4) +
  ggplot2::geom_text(
    ggplot2::aes(label = sprintf("W=%.2g", .data$welch_p)),
    size = 3
  ) +
  ggplot2::scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    name = "Delta log(mu)"
  ) +
  ggplot2::labs(
    title = "Four HCC cohorts vs comprehensive healthy liver NK baseline",
    subtitle = "Cell labels are two-sided Welch p-values",
    x = NULL,
    y = NULL
  ) +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    axis.text.x = ggplot2::element_text(angle = 25, hjust = 1)
  )
```

```{r}
#| label: plot-recurrence-contrasts
#| fig-width: 8
#| fig-height: 5
#| fig-align: center

ggplot2::ggplot(
  recurrence_tests,
  ggplot2::aes(
    x = .data$qlf_log2fc_nonrec_vs_rec,
    y = forcats::fct_rev(factor(.data$gene, levels = times_biomarkers$gene)),
    colour = .data$compartment
  )
) +
  ggplot2::geom_vline(xintercept = 0, colour = "grey55", linewidth = 0.4) +
  ggplot2::geom_point(size = 3) +
  ggplot2::geom_text(
    ggplot2::aes(label = sprintf("p+=%.2g", .data$p_nonrec_greater)),
    nudge_x = 0.08,
    hjust = 0,
    size = 3,
    show.legend = FALSE
  ) +
  ggplot2::facet_wrap(~ compartment, scales = "free_x") +
  ggplot2::scale_colour_manual(values = c(IF = "#2878B5", TC = "#E08A2E")) +
  ggplot2::labs(
    title = "Patient-level recurrence contrasts",
    subtitle = "Positive values support the paper hypothesis: non-REC > REC",
    x = "edgeR QL log2 fold change: non-REC vs REC",
    y = NULL,
    colour = "Compartment"
  ) +
  ggplot2::theme(legend.position = "none")
```

------------------------------------------------------------------------

# Proxy evaluation, positive control and paper benchmark

```{r}
#| label: ground-truth-comparison

proxy_gene_evaluation <- healthy_proxy_gene |>
  dplyr::left_join(
    proxy_gene_posterior_summary |>
      dplyr::select(
        gene, compartment,
        posterior_prob_rec_farther
      ),
    by = c("gene", "compartment")
  )

ground_truth_comparison <- dplyr::bind_rows(
  proxy_gene_evaluation |>
    dplyr::transmute(
      evidence_role = "primary_posteriorHCA_proxy",
      comparison_kind = "healthy_proximity_rec_vs_nonrec",
      gene = .data$gene,
      cohort = NA_character_,
      compartment = .data$compartment,
      expected_direction = "REC farther from healthy",
      observed_direction = dplyr::if_else(
        .data$rec_farther_from_healthy,
        "REC farther from healthy",
        "non-REC farther from healthy"
      ),
      matches_ground_truth = .data$rec_farther_from_healthy,
      observed_metric = .data$proxy_margin_rec_minus_nonrec,
      p_value = 1 - .data$posterior_prob_rec_farther,
      note = paste(
        "Outcome-free proxy hypothesis.",
        "p_value is posterior tail probability P(REC not farther)."
      )
    ),
  proxy_compartment_summary |>
    dplyr::transmute(
      evidence_role = "primary_posteriorHCA_proxy",
      comparison_kind = "healthy_proximity_rec_vs_nonrec",
      gene = "four_gene_mean",
      cohort = NA_character_,
      compartment = .data$compartment,
      expected_direction = "REC farther from healthy",
      observed_direction = dplyr::if_else(
        .data$mean_margin > 0,
        "REC farther from healthy",
        "non-REC farther from healthy"
      ),
      matches_ground_truth = .data$mean_margin > 0,
      observed_metric = .data$mean_margin,
      p_value = 1 - .data$posterior_prob_rec_farther,
      note = "Mean healthy-distance proxy across four atlas-covered TIMES genes."
    ),
  proxy_combined_summary |>
    dplyr::transmute(
      evidence_role = "primary_posteriorHCA_proxy",
      comparison_kind = "healthy_proximity_rec_vs_nonrec",
      gene = "four_gene_mean",
      cohort = NA_character_,
      compartment = "IF_TC_combined",
      expected_direction = "REC farther from healthy",
      observed_direction = dplyr::if_else(
        .data$mean_margin > 0,
        "REC farther from healthy",
        "non-REC farther from healthy"
      ),
      matches_ground_truth = .data$mean_margin > 0,
      observed_metric = .data$mean_margin,
      p_value = 1 - .data$posterior_prob_rec_farther,
      note = "Combined IF+TC posteriorHCA healthy-proximity proxy."
    ),
  recurrence_tests |>
    dplyr::transmute(
      evidence_role = "positive_control_paper_direction",
      comparison_kind = "recurrence_direction",
      gene = .data$gene,
      cohort = NA_character_,
      compartment = .data$compartment,
      expected_direction = .data$expected_recurrence_direction,
      observed_direction = .data$observed_recurrence_direction,
      matches_ground_truth = .data$recurrence_matches_paper,
      observed_metric = .data$qlf_log2fc_nonrec_vs_rec,
      p_value = .data$p_nonrec_greater,
      note = "Direct comparison with paper direction; positive metric means non-REC > REC."
    )
)

readr::write_csv(ground_truth_comparison, params$ground_truth_comparison_csv)

ground_truth_comparison |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), ~round(.x, 4)))
```

------------------------------------------------------------------------

# Interpretation and remaining limitation

The primary posteriorHCA result supports the proposed proxy:

- REC is farther from healthy than non-REC in
  **`r sum(healthy_proxy_gene$rec_farther_from_healthy)` of
  `r nrow(healthy_proxy_gene)`** gene-compartment comparisons.
- All REC cohort values lie beyond one healthy SD, while all non-REC values
  remain within one healthy SD.
- The combined four-gene IF+TC score gives
  \(P(D_{\mathrm{REC}} > D_{\mathrm{non-REC}}) =
  `r round(proxy_combined_summary$posterior_prob_rec_farther, 3)`\).
- For the stricter compatibility rule, the combined posterior gives
  \(P(D_{\mathrm{REC}} > 1) =
  `r round(proxy_combined_summary$posterior_prob_rec_beyond_one_sd, 3)`\),
  \(P(D_{\mathrm{non-REC}} \le 1) =
  `r round(proxy_combined_summary$posterior_prob_nonrec_within_one_sd, 3)`\),
  and joint pattern probability
  `r round(proxy_combined_summary$posterior_prob_proxy_pattern, 3)`.
- The outcome-free patient score ranks the available REC patient above all
  non-REC patients (descriptive AUC =
  `r healthy_proxy_performance$auc[healthy_proxy_performance$method == "posteriorHCA_healthy_proximity" & healthy_proxy_performance$scope == "combined_available_IF_TC"]`).

The direct edgeR contrast remains a positive control: all five paper biomarkers
retain the published `non-REC > REC` direction in IF and TC. The supplied TIMES
validation-score subset has descriptive AUC =
`r round(paper_times_benchmark$auc, 3)`, compared with the proxy's AUC above.
This is not an equivalence test: the paper benchmark uses mIHC scores from 11
patients, whereas the posteriorHCA proxy currently has only three GeoMx
patients and four atlas-covered genes.

The public cohort is extremely small after patient-level aggregation, so
the perfect patient ordering is proof of concept, not proof of equal power. The
exact rank p-value remains weak because there is one REC patient. The stronger
evidence here is the pre-specified reference pattern, its 8/8 consistency, the
combined posterior probability, and leave-one-gene-out stability. The paper's
much smaller gene-level p-values used multiple GeoMx ROI segments and adjusted
for age/sex where possible; this report protects the patient as the independent
unit and avoids ROI pseudoreplication.

For final publication-grade calibration, rerun this identical workflow with the
controlled-access raw GeoMx count matrices. `HLA-DRB1` remains testable for the
within-HCC recurrence contrast when present in the merged count fit, but it has
no comprehensive-baseline result while its `V1_nk` brms model branch is absent.

```{r}
#| label: output-files

output_paths <- c(
  baseline_draws_rds = params$baseline_draws_rds,
  paper_ground_truth_csv = params$paper_ground_truth_csv,
  cohort_results_csv = params$cohort_results_csv,
  recurrence_results_csv = params$recurrence_results_csv,
  patient_pseudobulk_csv = params$patient_pseudobulk_csv,
  normalization_csv = params$normalization_csv,
  ground_truth_comparison_csv = params$ground_truth_comparison_csv,
  healthy_proxy_gene_csv = params$healthy_proxy_gene_csv,
  healthy_proxy_summary_csv = params$healthy_proxy_summary_csv,
  healthy_proxy_patient_csv = params$healthy_proxy_patient_csv,
  healthy_proxy_performance_csv = params$healthy_proxy_performance_csv,
  paper_benchmark_csv = params$paper_benchmark_csv
)

tibble::tibble(
  output = names(output_paths),
  path = unname(output_paths),
  exists = file.exists(output_paths)
)
```
