---
title: "posteriorHCA Figure 2 — SAVI manuscript figure (v1)"
subtitle: "External population-reference audit of a published STING-activation signature"
date: today
date-format: "DD MMM YYYY"
title-block-banner: true
author:
  - name: Chen Zhan
    orcid: 0000-0002-4794-8339
    email: chen.zhan@adelaide.edu.au
    affiliation:
      - id: sgc
        name: SAiGENCI, The University of Adelaide
format:
  html:
    theme: minty
    df-print: paged
    code-line-numbers: true
    embed-resources: true
    code-fold: true
    code-fold-default: true
    toc: true
    toc-location: left
    toc-title: "Contents"
knitr:
  opts_chunk:
    message: false
    warning: false
    fig.retina: 2
editor: source
abstract: |
  Fresh v1 manuscript-figure report for the SAVI case (Nature Methods
  double-column canvas, max 183 mm × 220 mm; no panel titles/subtitles).
  Panel a freezes the observed cohort design with documented age imputation
  for GEO non-numeric ages, and regenerates healthy-blood support from the
  production `V1_monocytic` model frame. Panel b compares observed SAVI
  composition (CellType → Immune Atlas L3 → paper compartments) to a
  comprehensive healthy sccomp baseline. Panel c places the 16 covered STING
  signature genes against a comprehensive healthy log-μ baseline, an *ADRB2*
  personalised predictive check, and *ADRB2* UMAPs (all donors vs without
  high-depth `P6_STING_ht`).
params:
  case_dir: "/home/a1237163/lab/chen/posteriorHCA/dev/SAVI"
  fig_dir: "/home/a1237163/lab/chen/posteriorHCA/dev/SAVI/figures"
  data_dir: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/dev/SAVI/data"
  model_store_root: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/HPC_posterior/NEW_CELL_TYPE"
  target_cell_type: "monocytic"
  fig_width_mm: 183
  fig_height_mm: 220
  panel_a_height_mm: 58
  savi_seurat_gz: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/dev/SAVI/data/GSE226598_SeuratObject_SAVI.rds.gz"
  sccomp_estimates_rds: "/home/a1237163/lab/Mangiola_ImmuneAtlas/taskforce_shared_folder/sccomp_on_cellNexus_1_0_12/estimates_age_decade___L3___disease_TRUE___immune_only_TRUE.rds"
  composition_predict_rds: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/dev/SAVI/data/SAVI_composition_predict_res_comprehensive_10x3.rds"
  composition_predict_personalised_rds: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/dev/SAVI/data/SAVI_composition_predict_res_personalised_untreated_10x3.rds"
  panel_b_counts_rds: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/dev/SAVI/data/SAVI_panel_b_celltype_counts.rds"
  panel_b_savi_fit_rds: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/dev/SAVI/data/SAVI_panel_b_sccomp_shared15_untreated_control_hmc.rds"
  panel_b_savi_group_only_fit_rds: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/dev/SAVI/data/SAVI_panel_b_sccomp_shared15_untreated_control_group_only_v1.rds"
  run_sccomp_panel_b: false
  panel_b_prediction_draws: 4000
  panel_b_rope_pp: 1
  panel_b_pair_draws: 20000
  estimates_rds: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/dev/SAVI/data/SAVI_estimates_chunk.rds"
  panel_c_se_rds: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/dev/SAVI/data/SAVI_panel_c_se_monocytic.rds"
  panel_c_mu_density_rds: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/dev/SAVI/data/SAVI_panel_c_comprehensive_mu_density_v1.rds"
  panel_c_adrb2_personalised_rds: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/dev/SAVI/data/SAVI_panel_c_adrb2_personalised_predict_v1.rds"
  run_panel_c_rebuild: false
  savi_sc_celltype: "17. Disease-associated monocytes"
  adrb2_outlier_sample: "P6_STING_ht"
---

# Scope

**This report (v1).** Build the manuscript multipanel from audited tables.
**Panels a–c are in progress** (a–b complete; c1–c3 gene-level panels). Helper
functions and `theme_multipanel` live in this file.

**Colour / mark grammar.**

| Element | Legend (full) | Plot text (short) | Encoding |
|---|---|---|---|
| Study healthy control | Control | CTRL | green `#009E73`, circle |
| Pre-therapy SAVI | Untreated SAVI | Untreated | vermilion `#D55E00`, diamond |
| Post–JAK-inhibitor SAVI | JAK-inhibitor treated | Treated | purple `#7A5195`, square |
| Sex | female / male | — | `RColorBrewer` PRGn (`colmap_sex`) |
| Comprehensive HCA | — | — | charcoal grey |
| Metadata-conditioned HCA | — | — | teal |

Figure panels follow Nature Methods style: axis labels and legends only
(no plot titles or subtitles); canvas ≤ 183 mm wide × 220 mm high.
Continuous **y-axis** tick labels drop the leading zero on decimals
(`0.1` → `.1`) via `label_drop_leading_zero()`.

------------------------------------------------------------------------

# Setup

```{r}
#| label: setup

suppressPackageStartupMessages({
  library(tibble)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(ggplot2)
  library(patchwork)
  library(targets)
  library(SummarizedExperiment)
  library(scales)
  library(RColorBrewer)
  library(sccomp)
  library(edgeR)
  library(brms)
  library(forcats)
  library(Seurat)
  library(SeuratObject)
})

options(bitmapType = "cairo")
dir.create(params$fig_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(params$data_dir, recursive = TRUE, showWarnings = FALSE)

model_path <- file.path(
  params$model_store_root,
  paste0("V1_", params$target_cell_type)
)

mm_to_in <- function(mm) mm / 25.4
fig_w_in <- mm_to_in(params$fig_width_mm)
fig_h_in <- mm_to_in(params$fig_height_mm)
panel_a_h_in <- mm_to_in(params$panel_a_height_mm)

# Arcsine–square-root transform for composition proportions (same as posteriorHCA).
arcsine_sqrt_trans <- scales::trans_new(
  "arcsine_sqrt",
  transform = function(x) asin(sqrt(pmax(pmin(x, 1), 0))),
  inverse = function(x) sin(x)^2
)

read_geo_rds_gz <- function(path) {
  con <- gzcon(gzfile(path))
  on.exit(close(con), add = TRUE)
  readRDS(con)
}

# Figure grammar colours.
col_ctrl <- "#009E73"
col_untreated <- "#D55E00"
col_treated <- "#7A5195"
col_hca <- "#8C8C8C"
col_meta <- "#2A9D8F"
col_dark <- "#3F3F3F"
col_mid <- "#777777"
col_light <- "#D9D9D9"

colmap_sex <- setNames(
  RColorBrewer::brewer.pal(3, "PRGn")[c(1, 3)],
  c("female", "male")
)

state_levels <- c("control", "untreated", "treated")
state_labels <- c(
  control = "Control",
  untreated = "Untreated SAVI",
  treated = "JAK-inhibitor treated"
)
state_labels_short <- c(
  control = "CTRL",
  untreated = "Untreated",
  treated = "Treated"
)
state_colours <- c(
  control = col_ctrl,
  untreated = col_untreated,
  treated = col_treated
)
state_shapes <- c(control = 21, untreated = 23, treated = 22)

theme_multipanel <-
  theme_bw() +
  theme(
    panel.border = element_blank(),
    axis.line = element_line(linewidth = 0.1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    strip.background = element_blank(),
    axis.title.y = element_text(margin = margin(0, 0, 0, 0), size = 7),
    axis.title.x = element_text(margin = margin(0, 0, 0, 0), size = 7),
    panel.spacing.x = unit(0.1, "lines"),
    axis.text.x = element_text(size = 6),
    axis.text.y = element_text(size = 6),
    strip.text.x = element_text(size = 7),
    strip.text.y = element_text(size = 7),
    legend.key.size = unit(4, "mm"),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 6),
    strip.clip = "off",
    plot.title = element_blank(),
    plot.subtitle = element_blank(),
    plot.tag = element_text(size = 8, face = "bold"),
    axis.line.x = element_line(linewidth = 0.2),
    axis.line.y = element_line(linewidth = 0.2),
    axis.ticks.x = element_line(linewidth = 0.2),
    axis.ticks.y = element_line(linewidth = 0.2),
    plot.margin = margin(2, 4, 2, 2)
  )

# Y-axis labels: drop the leading zero on decimals (0.1 -> .1; -0.1 -> -.1).
label_drop_leading_zero <- function(accuracy = NULL, ...) {
  function(x) {
    s <- scales::number(
      x,
      accuracy = accuracy,
      big.mark = "",
      trim = TRUE,
      ...
    )
    s[!is.na(x) & x == 0] <- "0"
    sub("^(-?)0\\.", "\\1.", s)
  }
}

# Default continuous y scale for plots that do not set labels explicitly.
options(
  ggplot2.continuous.y = function() {
    ggplot2::scale_y_continuous(labels = label_drop_leading_zero())
  }
)
```

------------------------------------------------------------------------

# Panel a — Observed cohort design and HCA training support

**Purpose.** Show what posteriorHCA can condition on: library ages (including
documented imputations), sequencing batches / chemistry, repeated measures,
family links, and unique healthy-blood donor age support in the frozen
`V1_monocytic` frame.

**Age imputation (same rule as the SAVI case-study reports).** GEO lists
non-numeric ages for three libraries. We fill numeric years so every library
can sit on the age axis and feed decade-based models:

| Sample(s) | GEO label | Imputed `age_years` | Rationale |
|-----------|-----------|---------------------|-----------|
| **C1**, **C2** | Adult | **35** | Mean of adult controls with numeric ages (**C8** 36 y, **C9** 35 y). |
| **P5_STING** | Ped | **8** | Central paediatric value among other SAVI / control ages in the series. |

All other donors keep GEO-reported ages. `age_imputed = TRUE` only for C1, C2,
and P5.

## Freeze the 17-library source metadata table

Ages, sex and STING batch from GEO GSE226598; subject / family / treatment
annotations from de Cevins *et al.* (2023) Fig. 1a. Chemistry: STING1 = 10x 3′
v2; STING2 / STING3 = 10x 3′ v3.

```{r}
#| label: panel-a-savi-libraries

savi_libraries <- tibble::tribble(
  ~library_id,          ~subject_id, ~state,      ~sex,     ~age_years, ~age_imputed, ~sting_batch, ~chemistry,   ~family_role,              ~tx_label,
  "C7",                 "C7",        "control",   "female",         5,        FALSE, "STING1",     "10x 3′ v2",  NA_character_,             NA_character_,
  "C8_aunt_P1_STING",   "C8",        "control",   "female",        36,        FALSE, "STING1",     "10x 3′ v2",  "paternal aunt of P1",     NA_character_,
  "C9_mother_P1_STING", "C9",        "control",   "female",        35,        FALSE, "STING1",     "10x 3′ v2",  "mother of P1",            NA_character_,
  "P1_STING_ht",        "P1",        "untreated", "female",         4,        FALSE, "STING1",     "10x 3′ v2",  "proband",                 NA_character_,
  "P1_STING_ht_T",      "P1",        "treated",   "female",         8,        FALSE, "STING1",     "10x 3′ v2",  "proband",                 "R52",
  "P1_STING_ht_T2",     "P1",        "treated",   "female",         8,        FALSE, "STING1",     "10x 3′ v2",  "proband",                 "T2",
  "P2_STING_ht",        "P2",        "untreated", "male",          31,        FALSE, "STING1",     "10x 3′ v2",  "uncle of P1",             NA_character_,
  "P2_STING_ht_T",      "P2",        "treated",   "male",          34,        FALSE, "STING1",     "10x 3′ v2",  "uncle of P1",             "R1",
  "C10",                "C10",       "control",   "male",          10,        FALSE, "STING2",     "10x 3′ v3",  NA_character_,             NA_character_,
  "C11",                "C11",       "control",   "male",          16,        FALSE, "STING2",     "10x 3′ v3",  NA_character_,             NA_character_,
  "P4_STING_ht",        "P4",        "untreated", "male",           7,        FALSE, "STING2",     "10x 3′ v3",  NA_character_,             NA_character_,
  "P4_STING_ht_T",      "P4",        "treated",   "male",           9,        FALSE, "STING2",     "10x 3′ v3",  NA_character_,             "R13",
  "P6_STING_ht",        "P6",        "untreated", "male",           8,        FALSE, "STING2",     "10x 3′ v3",  NA_character_,             NA_character_,
  "P6_STING_ht_T",      "P6",        "treated",   "male",          10,        FALSE, "STING2",     "10x 3′ v3",  NA_character_,             "R13",
  "C1",                 "C1",        "control",   "male",          35,         TRUE, "STING3",     "10x 3′ v3",  NA_character_,             NA_character_,
  "C2",                 "C2",        "control",   "male",          35,         TRUE, "STING3",     "10x 3′ v3",  NA_character_,             NA_character_,
  "P5_STING",           "P5",        "untreated", "male",           8,         TRUE, "STING3",     "10x 3′ v3",  NA_character_,             NA_character_
) |>
  dplyr::mutate(
    state = factor(state, levels = state_levels),
    sex = factor(sex, levels = c("female", "male")),
    batch_row = factor(
      paste0(sting_batch, " · ", dplyr::recode(chemistry, "10x 3′ v2" = "v2", "10x 3′ v3" = "v3")),
      levels = c("STING1 · v2", "STING2 · v3", "STING3 · v3")
    ),
    display_id = dplyr::case_when(
      subject_id %in% c("C1", "C2") ~ subject_id,
      subject_id == "P5" ~ "P5",
      !is.na(tx_label) ~ tx_label,
      state == "untreated" ~ subject_id,
      TRUE ~ NA_character_
    )
  )

savi_libraries
```

## Lane offsets and subject connectors

Vertical jitter within a batch row keeps co-located libraries readable. Label
offsets are hand-tuned for the STING2 7–10 cluster, the P1 R52/T2 stack, and
imputed C1/C2 at age 35.

```{r}
#| label: panel-a-lane-offsets

assign_lanes <- function(df, lane_step = 0.32) {
  df |>
    dplyr::group_by(batch_row, age_years) |>
    dplyr::arrange(state, subject_id, library_id, .by_group = TRUE) |>
    dplyr::mutate(
      lane = (dplyr::row_number() - (dplyr::n() + 1) / 2) * lane_step,
      y_plot = as.numeric(batch_row) + lane
    ) |>
    dplyr::ungroup()
}

label_offset_tbl <- tibble::tribble(
  ~library_id,       ~label_dx, ~label_dy,
  "P1_STING_ht",           0.0,      0.40,
  "P1_STING_ht_T",        -2.0,      0.48,
  "P1_STING_ht_T2",        2.0,     -0.48,
  "P2_STING_ht",           0.0,      0.40,
  "P2_STING_ht_T",         0.0,      0.40,
  "P4_STING_ht",          -0.9,      0.48,
  "P6_STING_ht",          -0.9,     -0.48,
  "P4_STING_ht_T",         0.9,      0.48,
  "P6_STING_ht_T",         0.9,     -0.48,
  "P5_STING",              0.0,      0.40,
  "C1",                   -1.2,      0.40,
  "C2",                    1.2,     -0.40
)

savi_plot <- savi_libraries |>
  assign_lanes() |>
  dplyr::left_join(label_offset_tbl, by = "library_id") |>
  dplyr::mutate(
    label_dx = dplyr::coalesce(label_dx, 0),
    label_dy = dplyr::coalesce(label_dy, 0.40)
  )

savi_connectors <- savi_plot |>
  dplyr::group_by(subject_id) |>
  dplyr::filter(dplyr::n() > 1L) |>
  dplyr::arrange(age_years, state, library_id, .by_group = TRUE) |>
  dplyr::ungroup()
```

## Regenerate frozen `V1_monocytic` healthy-blood support

No-refit audit from the production `pseudobulk_sample` target: healthy blood
(`tissue_groups == "blood"`, `disease_groups___altered == "Normal"`), then
deduplicate by `donor_id`.

```{r}
#| label: panel-a-hca-support

pseudobulk_sample <- targets::tar_read(
  pseudobulk_sample,
  store = file.path(model_path, "_targets")
)

hca_healthy_blood <- SummarizedExperiment::colData(pseudobulk_sample) |>
  as.data.frame() |>
  tibble::as_tibble() |>
  dplyr::filter(
    .data$tissue_groups == "blood",
    as.character(.data$disease_groups___altered) == "Normal"
  ) |>
  dplyr::mutate(
    age_years = as.numeric(.data$age_years),
    sex = as.character(.data$sex),
    assay_groups_altered = as.character(.data$assay_groups___altered),
    dataset_id_altered = as.character(.data$dataset_id___altered)
  )

hca_donors <- hca_healthy_blood |>
  dplyr::group_by(.data$donor_id) |>
  dplyr::summarise(
    sex = dplyr::first(stats::na.omit(.data$sex)),
    age_years = stats::median(.data$age_years, na.rm = TRUE),
    n_samples = dplyr::n(),
    .groups = "drop"
  ) |>
  dplyr::mutate(sex = factor(.data$sex, levels = c("female", "male")))

hca_support_summary <- tibble::tibble(
  n_donors = nrow(hca_donors),
  n_samples = nrow(hca_healthy_blood),
  n_datasets = dplyr::n_distinct(hca_healthy_blood$dataset_id_altered),
  n_assay_groups = dplyr::n_distinct(hca_healthy_blood$assay_groups_altered),
  n_samples_10x3 = sum(
    hca_healthy_blood$assay_groups_altered == "10x Genomics 3",
    na.rm = TRUE
  ),
  age_min = min(hca_donors$age_years, na.rm = TRUE),
  age_max = max(hca_donors$age_years, na.rm = TRUE)
)

overlap_hits <- hca_healthy_blood |>
  dplyr::mutate(
    haystack = paste(
      dplyr::coalesce(as.character(.data$sample_id), ""),
      dplyr::coalesce(as.character(.data$dataset_id), ""),
      dplyr::coalesce(as.character(.data$title), ""),
      dplyr::coalesce(as.character(.data$citation), ""),
      dplyr::coalesce(as.character(.data$url), ""),
      sep = " "
    )
  ) |>
  dplyr::filter(
    stringr::str_detect(
      .data$haystack,
      stringr::regex("GSE226598|GSM70799|de Cevins|SAVI", ignore_case = TRUE)
    )
  )

overlap_status <- if (nrow(overlap_hits) == 0L) {
  "absent (no GSE226598 / SAVI string match)"
} else {
  sprintf("observed (%d rows matched)", nrow(overlap_hits))
}

hca_support_summary
```

Age support for each SAVI library relative to observed HCA donor ages:

```{r}
#| label: panel-a-support-flags

age_support <- range(hca_donors$age_years, na.rm = TRUE)

savi_query_support <- savi_libraries |>
  dplyr::mutate(
    support_flag = dplyr::if_else(
      age_years >= age_support[1] & age_years <= age_support[2],
      "observed",
      "outside support"
    ),
    assay_support = dplyr::if_else(
      hca_support_summary$n_samples_10x3 > 0L,
      "observed (harmonised 10x Genomics 3)",
      "outside support"
    )
  )

savi_query_support |>
  dplyr::count(age_imputed, support_flag, assay_support)
```

## Draw panel A

```{r}
#| label: panel-a-helpers

plot_panel_a_collections <- function(plot_df, connector_df) {
  y_labs <- levels(plot_df$batch_row)

  ggplot2::ggplot() +
    ggplot2::geom_hline(
      yintercept = seq_along(y_labs),
      colour = col_light,
      linewidth = 0.25
    ) +
    ggplot2::geom_path(
      data = connector_df,
      ggplot2::aes(x = age_years, y = y_plot, group = subject_id),
      colour = col_mid,
      linewidth = 0.35,
      alpha = 0.85
    ) +
    ggplot2::geom_point(
      data = plot_df,
      ggplot2::aes(
        x = age_years,
        y = y_plot,
        shape = state,
        colour = state,
        fill = sex
      ),
      size = 1.7,
      stroke = 0.45
    ) +
    ggplot2::geom_segment(
      data = plot_df |> dplyr::filter(!is.na(display_id)),
      ggplot2::aes(
        x = age_years,
        xend = age_years + label_dx,
        y = y_plot,
        yend = y_plot + label_dy
      ),
      colour = col_mid,
      linewidth = 0.2,
      alpha = 0.7
    ) +
    ggplot2::geom_text(
      data = plot_df |> dplyr::filter(!is.na(display_id)),
      ggplot2::aes(
        x = age_years + label_dx,
        y = y_plot + label_dy,
        label = display_id
      ),
      size = 2.5,
      colour = col_dark,
      vjust = 0.5,
      hjust = 0.5
    ) +
    ggplot2::scale_shape_manual(
      values = state_shapes,
      labels = state_labels,
      name = NULL,
      guide = ggplot2::guide_legend(order = 1, override.aes = list(fill = "white"))
    ) +
    ggplot2::scale_colour_manual(
      values = state_colours,
      labels = state_labels,
      name = NULL,
      guide = ggplot2::guide_legend(order = 1)
    ) +
    ggplot2::scale_fill_manual(
      values = colmap_sex,
      breaks = c("female", "male"),
      labels = c("female", "male"),
      name = NULL,
      guide = ggplot2::guide_legend(
        order = 2,
        override.aes = list(shape = 21, colour = col_dark)
      )
    ) +
    ggplot2::scale_y_continuous(
      breaks = seq_along(y_labs),
      labels = y_labs,
      limits = c(0.25, length(y_labs) + 0.75)
    ) +
    ggplot2::scale_x_continuous(
      limits = c(0, 40),
      breaks = seq(0, 40, by = 10),
      expand = ggplot2::expansion(mult = c(0.01, 0.02))
    ) +
    ggplot2::coord_cartesian(clip = "off") +
    ggplot2::labs(x = "Age at collection (years)", y = NULL) +
    theme_multipanel +
    ggplot2::theme(
      legend.position = "bottom",
      legend.box = "horizontal",
      legend.margin = margin(0, 0, 0, 0),
      axis.line.y = element_blank(),
      axis.ticks.y = element_blank()
    )
}

plot_panel_a_hca_continuous <- function(donors, savi_marks) {
  brks <- seq(0, 90, by = 10)

  hist_counts <- donors |>
    dplyr::mutate(
      bin_i = pmin(
        findInterval(.data$age_years, brks, all.inside = TRUE),
        length(brks) - 1L
      )
    ) |>
    dplyr::count(bin_i, sex, name = "n") |>
    tidyr::complete(
      bin_i = seq_len(length(brks) - 1L),
      sex = factor(c("female", "male"), levels = c("female", "male")),
      fill = list(n = 0L)
    ) |>
    dplyr::left_join(
      tibble::tibble(
        bin_i = seq_len(length(brks) - 1L),
        xmid = (brks[-length(brks)] + brks[-1]) / 2
      ),
      by = "bin_i"
    ) |>
    dplyr::mutate(sex = factor(sex, levels = c("male", "female")))

  age_marks <- savi_marks |>
    dplyr::distinct(age_years, sex) |>
    dplyr::mutate(sex = factor(sex, levels = c("female", "male")))

  ggplot2::ggplot(hist_counts, ggplot2::aes(x = xmid, y = n, fill = sex)) +
    ggplot2::geom_col(width = 9.2, colour = NA) +
    ggplot2::geom_vline(
      data = age_marks,
      ggplot2::aes(xintercept = age_years, colour = sex),
      linewidth = 0.35,
      alpha = 0.95
    ) +
    ggplot2::scale_fill_manual(
      values = colmap_sex,
      breaks = c("female", "male"),
      labels = c("female", "male"),
      name = NULL
    ) +
    ggplot2::scale_colour_manual(
      values = colmap_sex,
      breaks = c("female", "male"),
      guide = "none"
    ) +
    ggplot2::scale_x_continuous(
      limits = c(0, 90),
      breaks = c(0, 30, 60, 90),
      expand = c(0, 0)
    ) +
    ggplot2::scale_y_continuous(
      expand = ggplot2::expansion(mult = c(0, 0.08)),
      labels = label_drop_leading_zero()
    ) +
    ggplot2::labs(x = "Unique donor age (years)", y = NULL) +
    theme_multipanel +
    ggplot2::theme(
      legend.position = "bottom",
      legend.margin = margin(0, 0, 0, 0)
    )
}
```

```{r}
#| label: panel-a-draw
#| fig-width: 7.2
#| fig-height: 2.3
#| fig-cap: "Figure 2a. Observed SAVI collections (left; imputed ages for C1/C2/P5) and frozen V1_monocytic healthy-blood donor-age support (right)."

savi_age_marks <- savi_plot |>
  dplyr::distinct(age_years, sex)

p_a_left <- plot_panel_a_collections(savi_plot, savi_connectors)
p_a_right <- plot_panel_a_hca_continuous(hca_donors, savi_age_marks)

panel_a <- (p_a_left | p_a_right) +
  patchwork::plot_layout(widths = c(1.35, 1))

panel_a
```

```{r}
#| label: panel-a-audit-text

cat(
  sprintf(
    paste(
      "Panel a: %d libraries / %d subjects (%d age-imputed: C1, C2, P5);",
      "HCA healthy blood = %d donors / %d samples / %d datasets;",
      "%d samples in harmonised `10x Genomics 3`;",
      "donor age range %.1f–%.1f years;",
      "training-overlap status: %s;",
      "SAVI ages in HCA support: %d observed, %d outside.\n"
    ),
    nrow(savi_libraries),
    dplyr::n_distinct(savi_libraries$subject_id),
    sum(savi_libraries$age_imputed),
    hca_support_summary$n_donors,
    hca_support_summary$n_samples,
    hca_support_summary$n_datasets,
    hca_support_summary$n_samples_10x3,
    hca_support_summary$age_min,
    hca_support_summary$age_max,
    overlap_status,
    sum(savi_query_support$support_flag == "observed"),
    sum(savi_query_support$support_flag == "outside support")
  )
)
```

------------------------------------------------------------------------

# Panel b — Composition vs comprehensive healthy baseline

**Purpose.** Recover the paper’s Figure 1E composition story on a shared L3
vocabulary, and place observed cohort proportions against a comprehensive
healthy blood baseline from the Immune Atlas sccomp fit.

**Approach.**

1. Map the 23 GEO `CellType` labels and 26 HCA `L3` labels onto the same
   **15-part** composition (finest defensible shared vocabulary from the Fig. 2
   manuscript): seven paper T/NK compartments, seven matched B/myeloid/DC
   compartments, and `residual`.
2. Fit the 12 independent untreated/control subjects with `sccomp` using
   `formula_composition = ~ group + sequencing_batch`; treated repeated
   libraries are excluded from this disease-composition estimand.
3. Standardise both SAVI groups equally over STING1–STING3. Retain a matched
   `~ group` fit as a no-batch sensitivity, not the primary analysis.
4. Predict comprehensive and personalised healthy nulls from the atlas fit
   (Normal blood, assay `10x Genomics 3`; other covariates marginalised;
   personalised fills untreated age×sex only).
5. Display the seven paper compartments in plots b1–b2.

## Harmonise SAVI CellType and HCA L3 → 15 shared compartments

Same finest shared composition as the Fig. 2 manuscript report: seven paper
T/NK compartments for display/testing, seven matched B/myeloid/DC compartments
kept separate in the likelihood, and `residual` only for unmatched labels.

```{r}
#| label: panel-b-harmonisation

display_compartments <- c(
  "Naive CD4 T",
  "Eff./memory CD4",
  "Naive CD8 T",
  "Eff./memory CD8",
  "NK",
  "MAIT",
  "gamma-delta T"
)

compartment_levels <- c(
  display_compartments,
  "Naive/transitional B",
  "Memory B",
  "Plasma cells",
  "Classical monocytes",
  "Non-classical monocytes",
  "cDC",
  "pDC",
  "residual"
)

# Alias used by plot helpers / contrasts (paper Figure 1E family).
paper_compartment_levels <- display_compartments

celltype_to_compartment <- tibble::tribble(
  ~cell_type,              ~compartment,
  "0. Naive CD4",          "Naive CD4 T",
  "3. CM CD4",             "Eff./memory CD4",
  "4. EM CD4",             "Eff./memory CD4",
  "8. TE CD4",             "Eff./memory CD4",
  "9. Tregs and TE CD4",   "Eff./memory CD4",
  "2. Naive CD8",          "Naive CD8 T",
  "14. CM CD8",            "Eff./memory CD8",
  "15. EM CD8",            "Eff./memory CD8",
  "7. TE CD8",             "Eff./memory CD8",
  "10. NK",                "NK",
  "26. NK CD56bright",     "NK",
  "22. MAIT",              "MAIT",
  "16. gd T cells",        "gamma-delta T",
  "1. Naive B",            "Naive/transitional B",
  "6. Transitionnal B",    "Naive/transitional B",
  "13. Memory B",          "Memory B",
  "25. Plasmablasts",      "Plasma cells",
  "5. Classical monocytes", "Classical monocytes",
  "12. Non-classical monocytes", "Non-classical monocytes",
  "20. cDCs",              "cDC",
  "27. pDCs",              "pDC",
  "11. CD3+ B cells",      "residual",
  "17. Disease-associated monocytes", "residual"
)

l3_to_compartment <- tibble::tribble(
  ~L3,                ~compartment,
  "b memory",         "Memory B",
  "b naive",          "Naive/transitional B",
  "cd14 mono",        "Classical monocytes",
  "cd16 mono",        "Non-classical monocytes",
  "cdc",              "cDC",
  "pdc",              "pDC",
  "plasma",           "Plasma cells",
  "cd4 naive",        "Naive CD4 T",
  "cd4 tcm",          "Eff./memory CD4",
  "cd4 fh em",        "Eff./memory CD4",
  "cd4 th1 em",       "Eff./memory CD4",
  "cd4 th17 em",      "Eff./memory CD4",
  "cd4 th2 em",       "Eff./memory CD4",
  "cd4 th1/th17 em",  "Eff./memory CD4",
  "treg",             "Eff./memory CD4",
  "cd8 naive",        "Naive CD8 T",
  "cd8 tcm",          "Eff./memory CD8",
  "cd8 tem",          "Eff./memory CD8",
  "nk",               "NK",
  "mait",             "MAIT",
  "tgd",              "gamma-delta T",
  "granulocyte",      "residual",
  "ilc",              "residual",
  "macrophage",       "residual",
  "mast",             "residual",
  "nkt",              "residual"
)

source_direction <- tibble::tibble(
  compartment = factor(display_compartments, levels = display_compartments),
  source_arrow = c("+", "-", "+", "-", "-", "-", "-")
)

stopifnot(nrow(celltype_to_compartment) == 23L)
stopifnot(nrow(l3_to_compartment) == 26L)
stopifnot(!anyDuplicated(celltype_to_compartment$cell_type))
stopifnot(!anyDuplicated(l3_to_compartment$L3))

celltype_to_compartment
l3_to_compartment
```

## SAVI cell-type counts and proportions

```{r}
#| label: panel-b-savi-counts

if (file.exists(params$panel_b_counts_rds)) {
  panel_b_celltype_counts <- readRDS(params$panel_b_counts_rds)
} else {
  seurat_savi <- read_geo_rds_gz(params$savi_seurat_gz)
  md <- seurat_savi@meta.data
  rm(seurat_savi)
  gc()

  panel_b_celltype_counts <- md |>
    dplyr::transmute(
      sample_id = as.character(.data$Sample),
      Category = as.character(.data$Category),
      cell_type = as.character(.data$CellType)
    ) |>
    dplyr::count(.data$sample_id, .data$Category, .data$cell_type, name = "n_cells")

  saveRDS(panel_b_celltype_counts, params$panel_b_counts_rds, compress = "xz")
}

panel_b_sample_meta <- savi_libraries |>
  dplyr::transmute(
    sample_id = library_id,
    subject_id,
    group = dplyr::recode(
      as.character(state),
      control = "control",
      untreated = "untreated",
      treated = "treated"
    ),
    sequencing_batch = sting_batch,
    sex,
    age_years,
    age_imputed
  ) |>
  dplyr::mutate(
    group = factor(group, levels = c("control", "untreated", "treated")),
    sequencing_batch = factor(
      sequencing_batch,
      levels = c("STING1", "STING2", "STING3")
    )
  )

panel_b_counts_mapped <- panel_b_celltype_counts |>
  dplyr::left_join(celltype_to_compartment, by = "cell_type") |>
  dplyr::inner_join(panel_b_sample_meta, by = "sample_id")

stopifnot(!anyNA(panel_b_counts_mapped$compartment))
stopifnot(setequal(
  unique(panel_b_celltype_counts$cell_type),
  celltype_to_compartment$cell_type
))

sample_totals <- panel_b_celltype_counts |>
  dplyr::group_by(.data$sample_id) |>
  dplyr::summarise(n_total = sum(.data$n_cells), .groups = "drop")

# Observed proportions for the seven paper compartments (plot b1).
panel_b_prop_paper <- panel_b_counts_mapped |>
  dplyr::filter(.data$compartment %in% display_compartments) |>
  dplyr::group_by(.data$sample_id, .data$group, .data$compartment) |>
  dplyr::summarise(n_cells = sum(.data$n_cells), .groups = "drop") |>
  tidyr::complete(
    tidyr::nesting(sample_id, group),
    compartment = display_compartments,
    fill = list(n_cells = 0L)
  ) |>
  dplyr::left_join(sample_totals, by = "sample_id") |>
  dplyr::mutate(
    proportion = .data$n_cells / .data$n_total,
    paper_compartment = factor(.data$compartment, levels = display_compartments),
    state = factor(as.character(.data$group), levels = state_levels)
  )

panel_b_prop_paper |>
  dplyr::group_by(.data$state, .data$paper_compartment) |>
  dplyr::summarise(
    n_samples = dplyr::n(),
    mean_prop = mean(.data$proportion),
    .groups = "drop"
  )
```

## SAVI sccomp independent-subject fit (`~ group + sequencing_batch`)

The primary fit uses the full **15-part** closed count composition and the 12
independent untreated/control subjects. The batch term prevents a STING-lane
difference from being attributed to disease. The matched `~ group` model below
is retained only as a formula sensitivity.

```{r}
#| label: panel-b-sccomp-input

panel_b_sccomp_counts <- panel_b_counts_mapped |>
  dplyr::group_by(
    .data$sample_id, .data$subject_id, .data$group,
    .data$sequencing_batch, .data$compartment
  ) |>
  dplyr::summarise(n_cells = sum(.data$n_cells), .groups = "drop") |>
  tidyr::complete(
    tidyr::nesting(sample_id, subject_id, group, sequencing_batch),
    compartment = compartment_levels,
    fill = list(n_cells = 0L)
  ) |>
  dplyr::mutate(
    group = factor(
      as.character(.data$group),
      levels = c("control", "untreated", "treated")
    ),
    sequencing_batch = factor(
      as.character(.data$sequencing_batch),
      levels = c("STING1", "STING2", "STING3")
    ),
    compartment = factor(.data$compartment, levels = compartment_levels)
  )

panel_b_raw_total_check <- panel_b_sccomp_counts |>
  dplyr::group_by(.data$sample_id) |>
  dplyr::summarise(aggregated_n_cells = sum(.data$n_cells), .groups = "drop") |>
  dplyr::left_join(sample_totals, by = "sample_id")

stopifnot(nrow(panel_b_raw_total_check) == 17L)
stopifnot(all(
  panel_b_raw_total_check$aggregated_n_cells == panel_b_raw_total_check$n_total
))

# Panel B2 tests untreated versus control. Repeated treated libraries are not
# independent subjects and do not contribute to this estimand.
panel_b_analysis_counts <- panel_b_sccomp_counts |>
  dplyr::filter(.data$group %in% c("control", "untreated")) |>
  dplyr::mutate(
    group = factor(as.character(.data$group), levels = c("control", "untreated"))
  ) |>
  dplyr::select(
    "sample_id", "subject_id", "group", "sequencing_batch",
    "compartment", "n_cells"
  )

stopifnot(dplyr::n_distinct(panel_b_analysis_counts$sample_id) == 12L)
stopifnot(dplyr::n_distinct(panel_b_analysis_counts$subject_id) == 12L)

panel_b_analysis_counts |>
  dplyr::group_by(.data$sample_id, .data$group) |>
  dplyr::summarise(
    n_cells = sum(.data$n_cells),
    n_compartments = dplyr::n(),
    .groups = "drop"
  )
```

```{r}
#| label: panel-b-sccomp-fit

panel_b_savi_fit <- NULL
if (file.exists(params$panel_b_savi_fit_rds) && !isTRUE(params$run_sccomp_panel_b)) {
  panel_b_candidate <- readRDS(params$panel_b_savi_fit_rds)
  panel_b_terms <- attr(
    stats::terms(attr(panel_b_candidate, "formula_composition")),
    "term.labels"
  )
  if (setequal(panel_b_terms, c("group", "sequencing_batch")) &&
      nrow(attr(panel_b_candidate, "model_input")$X) == 12L) {
    panel_b_savi_fit <- panel_b_candidate
  } else {
    message("Ignoring incompatible Panel-B primary sccomp cache.")
  }
}

if (is.null(panel_b_savi_fit)) {
  panel_b_savi_fit <- sccomp::sccomp_estimate(
    panel_b_analysis_counts,
    formula_composition = ~ group + sequencing_batch,
    formula_variability = ~ 1,
    sample = "sample_id",
    cell_group = "compartment",
    abundance = "n_cells",
    cores = min(6L, max(1L, parallel::detectCores() - 2L)),
    inference_method = "hmc",
    verbose = TRUE
  )
  saveRDS(panel_b_savi_fit, params$panel_b_savi_fit_rds, compress = "xz")
}

stopifnot(nrow(attr(panel_b_savi_fit, "model_input")$X) == 12L)
panel_b_savi_fit
```

### Standardise primary SAVI predictions over sequencing batch

```{r}
#| label: panel-b-savi-predict

panel_b_savi_newdata <- tidyr::expand_grid(
  group = factor(
    c("control", "untreated"),
    levels = c("control", "untreated")
  ),
  sequencing_batch = factor(
    c("STING1", "STING2", "STING3"),
    levels = c("STING1", "STING2", "STING3")
  )
) |>
  dplyr::mutate(
    sample_id = paste(.data$group, .data$sequencing_batch, sep = "__"),
    target_weight = 1 / 3
  )

# sccomp's posterior-row subsampling uses R's global RNG in addition to the
# generated-quantities seed, so set both for render-to-render reproducibility.
set.seed(20260808L)
panel_b_savi_pred_draws <- sccomp::sccomp_predict(
  panel_b_savi_fit,
  new_data = panel_b_savi_newdata,
  number_of_draws = as.integer(params$panel_b_prediction_draws),
  mcmc_seed = 20260808L,
  summary_instead_of_draws = FALSE
)

# Verify closure before response-scale standardisation.
panel_b_savi_prediction_sum_check <- panel_b_savi_pred_draws |>
  dplyr::group_by(.data$.draw, .data$sample_id) |>
  dplyr::summarise(total = sum(.data$proportion), .groups = "drop")

stopifnot(max(abs(panel_b_savi_prediction_sum_check$total - 1)) < 1e-5)

panel_b_savi_group_draws <- panel_b_savi_pred_draws |>
  dplyr::group_by(.data$.draw, .data$group, .data$compartment) |>
  dplyr::summarise(
    proportion = stats::weighted.mean(.data$proportion, .data$target_weight),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    paper_compartment = factor(
      as.character(.data$compartment),
      levels = display_compartments
    ),
    group = factor(as.character(.data$group), levels = c("control", "untreated"))
  )

panel_b_savi_group_sum_check <- panel_b_savi_group_draws |>
  dplyr::group_by(.data$.draw, .data$group) |>
  dplyr::summarise(total = sum(.data$proportion), .groups = "drop")

stopifnot(max(abs(panel_b_savi_group_sum_check$total - 1)) < 1e-5)

panel_b_savi_group_summary <- panel_b_savi_group_draws |>
  dplyr::group_by(.data$group, .data$paper_compartment) |>
  dplyr::summarise(
    median_proportion = stats::median(.data$proportion),
    hpd_lo = stats::quantile(.data$proportion, 0.025),
    hpd_hi = stats::quantile(.data$proportion, 0.975),
    .groups = "drop"
  )

panel_b_savi_group_summary
```

### Matched no-batch formula sensitivity

```{r}
#| label: panel-b-group-only-sensitivity-fit

panel_b_group_only_fit <- NULL
if (file.exists(params$panel_b_savi_group_only_fit_rds) &&
    !isTRUE(params$run_sccomp_panel_b)) {
  panel_b_group_only_candidate <- readRDS(params$panel_b_savi_group_only_fit_rds)
  panel_b_group_only_terms <- attr(
    stats::terms(attr(panel_b_group_only_candidate, "formula_composition")),
    "term.labels"
  )
  if (identical(panel_b_group_only_terms, "group") &&
      nrow(attr(panel_b_group_only_candidate, "model_input")$X) == 12L) {
    panel_b_group_only_fit <- panel_b_group_only_candidate
  } else {
    message("Ignoring incompatible Panel-B group-only sensitivity cache.")
  }
}

if (is.null(panel_b_group_only_fit)) {
  panel_b_group_only_fit <- sccomp::sccomp_estimate(
    panel_b_analysis_counts,
    formula_composition = ~ group,
    formula_variability = ~ 1,
    sample = "sample_id",
    cell_group = "compartment",
    abundance = "n_cells",
    cores = min(6L, max(1L, parallel::detectCores() - 2L)),
    inference_method = "hmc",
    verbose = TRUE
  )
  saveRDS(
    panel_b_group_only_fit,
    params$panel_b_savi_group_only_fit_rds,
    compress = "xz"
  )
}

panel_b_group_only_newdata <- tibble::tibble(
  group = factor(
    c("control", "untreated"),
    levels = c("control", "untreated")
  ),
  sample_id = paste0("group_only__", c("control", "untreated"))
)

set.seed(20260811L)
panel_b_group_only_pred <- sccomp::sccomp_predict(
  panel_b_group_only_fit,
  new_data = panel_b_group_only_newdata,
  number_of_draws = as.integer(params$panel_b_prediction_draws),
  mcmc_seed = 20260811L,
  summary_instead_of_draws = FALSE
)

panel_b_group_only_draws <- panel_b_group_only_pred |>
  dplyr::mutate(
    group = factor(as.character(.data$group), levels = c("control", "untreated")),
    paper_compartment = factor(
      as.character(.data$compartment),
      levels = display_compartments
    )
  )

panel_b_group_only_sum_check <- panel_b_group_only_draws |>
  dplyr::group_by(.data$.draw, .data$group) |>
  dplyr::summarise(total = sum(.data$proportion), .groups = "drop")

stopifnot(max(abs(panel_b_group_only_sum_check$total - 1)) < 1e-5)
```

## Healthy HCA baselines (`sccomp_predict`)

Both baselines use the same marginalised atlas predict: Normal blood, assay
fixed to **`10x Genomics 3`**, ethnicity and dataset left as the sccomp average
effect (`NA`).

- **Comprehensive:** sex and age also `NA` (one population null).
- **Personalised:** sex and age set to each untreated SAVI subject, then
  equal-averaged within each posterior draw.

```{r}
#| label: panel-b-hca-predict

aggregate_hca_to_paper <- function(predict_tbl) {
  predict_tbl |>
    dplyr::inner_join(l3_to_compartment, by = c("cell_type" = "L3")) |>
    dplyr::group_by(.data$sample_id, .data$compartment, .data$.draw) |>
    dplyr::summarise(proportion = sum(.data$proportion), .groups = "drop")
}

# --- Comprehensive (sex/age marginalised) ---
if (file.exists(params$composition_predict_rds)) {
  predict_res <- readRDS(params$composition_predict_rds)
} else {
  sccomp_est <- readRDS(params$sccomp_estimates_rds)
  set.seed(20260809L)
  predict_res <- sccomp_est |>
    sccomp::sccomp_predict(
      formula_composition = NULL,
      new_data = tibble::tibble(
        assay_groups___altered = "10x Genomics 3",
        disease_groups___altered = "Normal",
        tissue_groups = "blood",
        ethnicity_groups_imputed = NA_character_,
        age_decade = NA_character_,
        sex = NA_character_,
        dataset_id___altered = NA_character_,
        sample_id = "comprehensive_baseline"
      ),
      number_of_draws = as.integer(params$panel_b_prediction_draws),
      mcmc_seed = 20260809L,
      summary_instead_of_draws = FALSE
    ) |>
    dplyr::rename(cell_type = L3) |>
    dplyr::select("sample_id", "cell_type", "proportion", ".draw")
  saveRDS(predict_res, params$composition_predict_rds, compress = "xz")
}

healthy_draws_paper <- aggregate_hca_to_paper(predict_res) |>
  dplyr::filter(.data$compartment %in% display_compartments) |>
  dplyr::transmute(
    .draw,
    paper_compartment = factor(.data$compartment, levels = display_compartments),
    proportion,
    reference_target = "comprehensive_hca"
  )

# --- Personalised (untreated age × sex; other covariates still marginalised) ---
panel_b_personalised_meta <- savi_libraries |>
  dplyr::filter(as.character(.data$state) == "untreated") |>
  dplyr::transmute(
    assay_groups___altered = "10x Genomics 3",
    disease_groups___altered = "Normal",
    tissue_groups = "blood",
    ethnicity_groups_imputed = NA_character_,
    age_decade = as.character(ceiling(.data$age_years / 10)),
    sex = as.character(.data$sex),
    dataset_id___altered = NA_character_,
    sample_id = paste0("personalised_", .data$subject_id),
    target_weight = 1 / dplyr::n()
  )

if (file.exists(params$composition_predict_personalised_rds)) {
  predict_res_personalised <- readRDS(params$composition_predict_personalised_rds)
} else {
  if (!exists("sccomp_est", inherits = FALSE)) {
    sccomp_est <- readRDS(params$sccomp_estimates_rds)
  }
  set.seed(20260811L)
  predict_res_personalised <- sccomp_est |>
    sccomp::sccomp_predict(
      formula_composition = NULL,
      new_data = panel_b_personalised_meta |>
        dplyr::select(-"target_weight"),
      number_of_draws = as.integer(params$panel_b_prediction_draws),
      mcmc_seed = 20260811L,
      summary_instead_of_draws = FALSE
    ) |>
    dplyr::rename(cell_type = L3) |>
    dplyr::select("sample_id", "cell_type", "proportion", ".draw")
  saveRDS(
    predict_res_personalised,
    params$composition_predict_personalised_rds,
    compress = "xz"
  )
}

# Equal-average personalised subjects within each draw on the paper scale.
personalised_draws_paper <- aggregate_hca_to_paper(predict_res_personalised) |>
  dplyr::filter(.data$compartment %in% display_compartments) |>
  dplyr::left_join(
    panel_b_personalised_meta |> dplyr::select("sample_id", "target_weight"),
    by = "sample_id"
  ) |>
  dplyr::group_by(.data$compartment, .data$.draw) |>
  dplyr::summarise(
    proportion = stats::weighted.mean(.data$proportion, .data$target_weight),
    .groups = "drop"
  ) |>
  dplyr::transmute(
    .draw,
    paper_compartment = factor(.data$compartment, levels = display_compartments),
    proportion,
    reference_target = "personalised_hca"
  )

hca_draws_paper <- dplyr::bind_rows(healthy_draws_paper, personalised_draws_paper)

healthy_baseline_paper <- healthy_draws_paper |>
  dplyr::group_by(.data$paper_compartment) |>
  dplyr::summarise(
    median_proportion = stats::median(.data$proportion),
    hpd_lo = stats::quantile(.data$proportion, 0.025),
    hpd_hi = stats::quantile(.data$proportion, 0.975),
    .groups = "drop"
  )

healthy_baseline_paper
```

## Plot b1 — Paper compartments vs healthy band

Three cohort boxplots with jitter; grey band = comprehensive healthy 95%
interval; dashed line = healthy median. Y-axis uses the arcsine–square-root
transform.

```{r}
#| label: panel-b-helpers

summarise_panel_b_pp_test <- function(draws, rope_pp) {
  draws |>
    dplyr::filter(.data$compartment %in% display_compartments) |>
    dplyr::group_by(.data$contrast, .data$compartment) |>
    dplyr::summarise(
      estimate = stats::median(.data$delta_pp),
      .lower = unname(stats::quantile(.data$delta_pp, 0.025)),
      .upper = unname(stats::quantile(.data$delta_pp, 0.975)),
      p_increase = mean(.data$delta_pp > rope_pp),
      p_decrease = mean(.data$delta_pp < -rope_pp),
      .groups = "drop"
    ) |>
    dplyr::mutate(
      pp_pH0 = 1 - pmax(.data$p_increase, .data$p_decrease),
      posterior_direction = dplyr::if_else(
        .data$p_increase >= .data$p_decrease,
        "increase",
        "decrease"
      )
    ) |>
    dplyr::group_by(.data$contrast) |>
    dplyr::arrange(.data$pp_pH0, .by_group = TRUE) |>
    dplyr::mutate(
      pp_FDR = cummean(.data$pp_pH0),
      discovery = .data$pp_FDR <= 0.05
    ) |>
    dplyr::ungroup()
}

plot_panel_b1_composition <- function(obs_df, healthy_df) {
  ggplot2::ggplot(
    obs_df,
    ggplot2::aes(x = state, y = proportion, fill = state)
  ) +
    ggplot2::geom_rect(
      data = healthy_df,
      ggplot2::aes(xmin = -Inf, xmax = Inf, ymin = hpd_lo, ymax = hpd_hi),
      fill = col_hca,
      alpha = 0.25,
      colour = NA,
      inherit.aes = FALSE
    ) +
    ggplot2::geom_hline(
      data = healthy_df,
      ggplot2::aes(yintercept = median_proportion),
      colour = col_dark,
      linewidth = 0.45,
      linetype = "dashed",
      inherit.aes = FALSE
    ) +
    ggplot2::geom_boxplot(
      outlier.shape = NA,
      alpha = 0.75,
      colour = col_dark,
      linewidth = 0.35
    ) +
    ggplot2::geom_jitter(
      ggplot2::aes(colour = state, shape = state),
      width = 0.12,
      size = 1.7,
      stroke = 0.45,
      fill = "white",
      alpha = 0.95
    ) +
    ggplot2::facet_wrap(~paper_compartment, scales = "free_y", ncol = 4) +
    ggplot2::scale_fill_manual(
      values = state_colours,
      labels = state_labels,
      name = NULL
    ) +
    ggplot2::scale_colour_manual(
      values = state_colours,
      labels = state_labels,
      name = NULL
    ) +
    ggplot2::scale_shape_manual(
      values = state_shapes,
      labels = state_labels,
      name = NULL
    ) +
    ggplot2::scale_x_discrete(labels = state_labels_short) +
    ggplot2::scale_y_continuous(
      trans = arcsine_sqrt_trans,
      labels = label_drop_leading_zero(accuracy = 0.01)
    ) +
    ggplot2::labs(x = NULL, y = "Proportion (arcsin√ scale)") +
    theme_multipanel +
    ggplot2::theme(legend.position = "bottom")
}

plot_panel_b2_effects <- function(effects_df) {
  y_labels <- effects_df |>
    dplyr::distinct(.data$compartment, .data$source_arrow) |>
    dplyr::mutate(
      y_lab = paste(.data$source_arrow, as.character(.data$compartment))
    ) |>
    dplyr::arrange(.data$compartment)

  lab_map <- setNames(y_labels$y_lab, as.character(y_labels$compartment))

  contrast_breaks <- c(
    "untreated - study control",
    "untreated - comprehensive HCA",
    "untreated - personalised HCA"
  )
  contrast_labels <- c(
    "vs study control",
    "vs comprehensive HCA",
    "vs personalised HCA"
  )
  contrast_colours <- c(
    "untreated - study control" = col_ctrl,
    "untreated - comprehensive HCA" = col_hca,
    "untreated - personalised HCA" = col_meta
  )
  contrast_fills <- c(
    "untreated - study control" = "white",
    "untreated - comprehensive HCA" = col_hca,
    "untreated - personalised HCA" = col_meta
  )
  contrast_shapes <- c(
    "untreated - study control" = 21,
    "untreated - comprehensive HCA" = 23,
    "untreated - personalised HCA" = 24
  )
  # Tight dodge so the three series sit on one compartment row.
  contrast_dodge <- ggplot2::position_dodge(width = 0.55)

  ggplot2::ggplot(
    effects_df,
    ggplot2::aes(
      y = .data$compartment,
      x = .data$estimate,
      colour = .data$contrast,
      fill = .data$contrast,
      shape = .data$contrast,
      group = .data$contrast
    )
  ) +
    ggplot2::geom_vline(
      xintercept = 0,
      colour = col_mid,
      linetype = "22",
      linewidth = 0.35
    ) +
    ggplot2::geom_linerange(
      ggplot2::aes(xmin = .data$.lower, xmax = .data$.upper),
      linewidth = 0.55,
      position = contrast_dodge
    ) +
    ggplot2::geom_point(
      ggplot2::aes(size = .data$discovery),
      stroke = 0.6,
      position = contrast_dodge
    ) +
    ggplot2::scale_colour_manual(
      values = contrast_colours,
      breaks = contrast_breaks,
      labels = contrast_labels,
      name = NULL
    ) +
    ggplot2::scale_fill_manual(
      values = contrast_fills,
      breaks = contrast_breaks,
      labels = contrast_labels,
      name = NULL
    ) +
    ggplot2::scale_shape_manual(
      values = contrast_shapes,
      breaks = contrast_breaks,
      labels = contrast_labels,
      name = NULL
    ) +
    ggplot2::scale_size_manual(
      values = c(`FALSE` = 2.0, `TRUE` = 2.8),
      guide = "none"
    ) +
    ggplot2::scale_y_discrete(
      limits = rev(display_compartments),
      labels = lab_map
    ) +
    ggplot2::scale_x_continuous(
      breaks = c(-30, -20, -10, 0, 10, 20),
      limits = c(-30, 20),
      expand = ggplot2::expansion(mult = 0.03)
    ) +
    ggplot2::coord_cartesian(clip = "off") +
    ggplot2::labs(
      x = "Composition difference (percentage points)",
      y = NULL
    ) +
    theme_multipanel +
    ggplot2::theme(
      legend.position = "bottom",
      legend.direction = "horizontal",
      legend.box = "horizontal",
      legend.margin = margin(0, 0, 0, 0),
      legend.spacing.x = grid::unit(0, "mm"),
      plot.margin = margin(0, 0, 0, 0)
    ) +
    ggplot2::guides(
      colour = ggplot2::guide_legend(override.aes = list(size = 2.4)),
      fill = ggplot2::guide_legend(override.aes = list(size = 2.4)),
      shape = ggplot2::guide_legend(override.aes = list(size = 2.4))
    )
}
```

```{r}
#| label: panel-b1-draw
#| fig-width: 7.2
#| fig-height: 4.2
#| fig-cap: "Figure 2b1. Observed SAVI paper-compartment proportions by cohort versus the comprehensive healthy sccomp baseline. Grey band = healthy 95% interval; dashed line = healthy median. Y-axis arcsine–square-root transformed."

panel_b1 <- plot_panel_b1_composition(panel_b_prop_paper, healthy_baseline_paper)
panel_b1
```

## Plot b2 — Composition differences

Forest plot: untreated SAVI versus study controls, comprehensive HCA, and
personalised HCA (untreated age×sex; assay / ethnicity / dataset still
marginalised as in the comprehensive null). Contrasts are percentage-point
differences with a `r params$panel_b_rope_pp`-pp ROPE and Bayesian FDR across
the seven paper compartments.

```{r}
#| label: panel-b2-contrasts

# Untreated − control (same SAVI posterior draws).
panel_b_uc_draws <- panel_b_savi_group_draws |>
  dplyr::filter(.data$compartment %in% display_compartments) |>
  dplyr::select(".draw", "group", compartment = "paper_compartment", "proportion") |>
  tidyr::pivot_wider(names_from = "group", values_from = "proportion") |>
  dplyr::mutate(
    delta_pp = 100 * (.data$untreated - .data$control),
    contrast = "untreated - study control"
  ) |>
  dplyr::select(".draw", "compartment", "delta_pp", "contrast")

panel_b_uc <- summarise_panel_b_pp_test(
  panel_b_uc_draws,
  rope_pp = params$panel_b_rope_pp
)

# Untreated − HCA references (independent cross-model pairs).
untreated_draws <- panel_b_savi_group_draws |>
  dplyr::filter(
    .data$group == "untreated",
    .data$compartment %in% display_compartments
  ) |>
  dplyr::transmute(
    .draw = .data$.draw,
    compartment = .data$paper_compartment,
    case_proportion = .data$proportion
  )

set.seed(20260810L)
panel_b_cross_fit_pairs <- tibble::tibble(
  .pair = seq_len(as.integer(params$panel_b_pair_draws)),
  .draw_savi = sample(
    unique(untreated_draws$.draw),
    as.integer(params$panel_b_pair_draws),
    replace = TRUE
  ),
  .draw_hca = sample(
    unique(healthy_draws_paper$.draw),
    as.integer(params$panel_b_pair_draws),
    replace = TRUE
  )
)

panel_b_uh_draws <- panel_b_cross_fit_pairs |>
  dplyr::left_join(
    untreated_draws,
    by = c(".draw_savi" = ".draw"),
    relationship = "many-to-many"
  ) |>
  dplyr::left_join(
    hca_draws_paper |>
      dplyr::transmute(
        .draw,
        compartment = .data$paper_compartment,
        hca_proportion = .data$proportion,
        reference_target
      ),
    by = c(".draw_hca" = ".draw", "compartment"),
    relationship = "many-to-many"
  ) |>
  dplyr::mutate(
    delta_pp = 100 * (.data$case_proportion - .data$hca_proportion),
    contrast = dplyr::recode(
      .data$reference_target,
      comprehensive_hca = "untreated - comprehensive HCA",
      personalised_hca = "untreated - personalised HCA"
    )
  )

panel_b_uh <- summarise_panel_b_pp_test(
  panel_b_uh_draws,
  rope_pp = params$panel_b_rope_pp
)

panel_b2_effects <- dplyr::bind_rows(panel_b_uc, panel_b_uh) |>
  dplyr::mutate(
    compartment = factor(
      as.character(.data$compartment),
      levels = display_compartments
    ),
    contrast = factor(
      .data$contrast,
      levels = c(
        "untreated - study control",
        "untreated - comprehensive HCA",
        "untreated - personalised HCA"
      )
    )
  ) |>
  dplyr::left_join(source_direction, by = "compartment")

panel_b2_effects |>
  dplyr::arrange(.data$compartment, .data$contrast)
```

### Formula sensitivity: remove sequencing batch

This sensitivity uses the same 12 subjects, 15-part harmonisation, HCA draws,
ROPE and draw-pairing rule. Consequently, any difference below is attributable
to the SAVI composition formula rather than to a changed sample set or atlas
reference.

```{r}
#| label: panel-b2-formula-sensitivity

panel_b_group_only_uc_draws <- panel_b_group_only_draws |>
  dplyr::filter(.data$compartment %in% display_compartments) |>
  dplyr::select(".draw", "group", "compartment", "proportion") |>
  tidyr::pivot_wider(names_from = "group", values_from = "proportion") |>
  dplyr::mutate(
    delta_pp = 100 * (.data$untreated - .data$control),
    contrast = "untreated - study control"
  ) |>
  dplyr::select(".draw", "compartment", "delta_pp", "contrast")

panel_b_group_only_uc <- summarise_panel_b_pp_test(
  panel_b_group_only_uc_draws,
  rope_pp = params$panel_b_rope_pp
)

panel_b_group_only_untreated <- panel_b_group_only_draws |>
  dplyr::filter(
    .data$group == "untreated",
    .data$compartment %in% display_compartments
  ) |>
  dplyr::transmute(
    .draw = .data$.draw,
    compartment = .data$compartment,
    case_proportion = .data$proportion
  )

set.seed(20260812L)
panel_b_group_only_pairs <- tibble::tibble(
  .pair = seq_len(as.integer(params$panel_b_pair_draws)),
  .draw_savi = sample(
    unique(panel_b_group_only_untreated$.draw),
    as.integer(params$panel_b_pair_draws),
    replace = TRUE
  ),
  .draw_hca = sample(
    unique(healthy_draws_paper$.draw),
    as.integer(params$panel_b_pair_draws),
    replace = TRUE
  )
)

panel_b_group_only_uh_draws <- panel_b_group_only_pairs |>
  dplyr::left_join(
    panel_b_group_only_untreated,
    by = c(".draw_savi" = ".draw"),
    relationship = "many-to-many"
  ) |>
  dplyr::left_join(
    healthy_draws_paper |>
      dplyr::transmute(
        .draw = .data$.draw,
        compartment = as.character(.data$paper_compartment),
        hca_proportion = .data$proportion
      ),
    by = c(".draw_hca" = ".draw", "compartment"),
    relationship = "many-to-many"
  ) |>
  dplyr::mutate(
    delta_pp = 100 * (.data$case_proportion - .data$hca_proportion),
    contrast = "untreated - comprehensive HCA"
  )

panel_b_group_only_uh <- summarise_panel_b_pp_test(
  panel_b_group_only_uh_draws,
  rope_pp = params$panel_b_rope_pp
)

panel_b_formula_sensitivity <- dplyr::bind_rows(
  panel_b2_effects |>
    dplyr::filter(.data$contrast %in% c(
      "untreated - study control",
      "untreated - comprehensive HCA"
    )) |>
    dplyr::transmute(
      contrast = as.character(.data$contrast),
      compartment = as.character(.data$compartment),
      model = "primary",
      estimate,
      discovery
    ),
  dplyr::bind_rows(panel_b_group_only_uc, panel_b_group_only_uh) |>
    dplyr::transmute(
      contrast = as.character(.data$contrast),
      compartment = as.character(.data$compartment),
      model = "no_batch",
      estimate,
      discovery
    )
) |>
  tidyr::pivot_wider(
    names_from = "model",
    values_from = c("estimate", "discovery")
  ) |>
  dplyr::mutate(
    estimate_shift_pp = .data$estimate_no_batch - .data$estimate_primary,
    direction_changed = sign(.data$estimate_primary) != sign(.data$estimate_no_batch),
    discovery_changed = .data$discovery_primary != .data$discovery_no_batch
  ) |>
  dplyr::arrange(.data$contrast, match(.data$compartment, display_compartments))

panel_b_formula_sensitivity |>
  dplyr::select(
    contrast,
    compartment,
    primary_pp = .data$estimate_primary,
    no_batch_pp = .data$estimate_no_batch,
    estimate_shift_pp,
    direction_changed,
    discovery_changed
  ) |>
  knitr::kable(
    digits = 2,
    caption = paste(
      "Sequencing-batch formula sensitivity using an identical 12-subject",
      "analysis set and HCA reference."
    )
  )
```

```{r}
#| label: panel-b2-formula-sensitivity-conclusion
#| results: asis

panel_b_formula_max_shift <- panel_b_formula_sensitivity |>
  dplyr::slice_max(abs(.data$estimate_shift_pp), n = 1, with_ties = FALSE)

panel_b_formula_discovery_changes <- panel_b_formula_sensitivity |>
  dplyr::filter(.data$discovery_changed)

cat(sprintf(
  paste0(
    "**Formula-sensitivity conclusion.** Removing sequencing batch changed no ",
    "effect direction, but shifted the largest posterior median by %.2f percentage ",
    "points (%s; %s) and changed %d discovery decision%s. Therefore the ",
    "batch-adjusted model remains primary.\n\n"
  ),
  abs(panel_b_formula_max_shift$estimate_shift_pp),
  panel_b_formula_max_shift$compartment,
  panel_b_formula_max_shift$contrast,
  nrow(panel_b_formula_discovery_changes),
  ifelse(nrow(panel_b_formula_discovery_changes) == 1L, "", "s")
))
```

```{r}
#| label: panel-b2-draw
#| fig-width: 6.2
#| fig-height: 4.8
#| fig-cap: "Figure 2b2. Posterior mean-composition differences. Green open circles: untreated − study control. Grey diamonds: untreated − comprehensive HCA. Teal triangles: untreated − personalised HCA. Larger symbols meet Bayesian FDR ≤ 0.05 for a 1-percentage-point ROPE."

panel_b2 <- plot_panel_b2_effects(panel_b2_effects)
panel_b2
```

```{r}
#| label: panel-b-audit-text

cat(
  sprintf(
    paste(
      "Panel b: %d observed libraries; primary sccomp uses %d independent",
      "untreated/control subjects with ~ group + sequencing_batch on %d shared compartments;",
      "%d paper compartments shown;",
      "HCA baselines = comprehensive + personalised (%d L3 types × %d draws);",
      "b2 contrasts use %.2f-pp ROPE and %d cross-model pairs.\n"
    ),
    dplyr::n_distinct(panel_b_prop_paper$sample_id),
    dplyr::n_distinct(panel_b_analysis_counts$subject_id),
    length(compartment_levels),
    length(display_compartments),
    dplyr::n_distinct(predict_res$cell_type),
    dplyr::n_distinct(predict_res$.draw),
    params$panel_b_rope_pp,
    nrow(panel_b_cross_fit_pairs)
  )
)
```

------------------------------------------------------------------------

# Panel c — Gene-level abundance vs healthy baselines

Gene panels place STING-signature counts against healthy HCA nulls. **C1**
uses a comprehensive (population) log-μ baseline for all 16 covered genes.
**C2** audits *ADRB2* with a personalised predictive-count null at offset 0.
**C3** shows single-cell *ADRB2* on UMAP (all donors, then without the
high-depth untreated library `P6_STING_ht`).

```{r}
#| label: panel-c-setup

panel_c_gene_order <- c(
  "ADRB2", "PTGS2", "AQP9", "OTUD1", "PTX3", "TGIF1", "TSC22D2", "ZBTB43",
  "ZFP36L2", "JARID2", "KLF4", "TREM1", "ANKRD28", "ANXA1", "PLAUR", "IL1B"
)

panel_c_category_to_state <- c(
  CTRL = "control",
  SAVI = "untreated",
  SAVI_treated = "treated"
)

fit_levels <- function(fit, var) {
  d <- fit$data
  if (is.null(d) || !var %in% names(d)) {
    return(character(0))
  }
  v <- d[[var]]
  lv <- if (is.factor(v)) levels(v) else sort(unique(as.character(v)))
  lv[!is.na(lv) & nzchar(lv)]
}

build_newdata_grid <- function(
  fit,
  newdata,
  marginalize_over = c("sex", "age_decade", "ethnicity_groups"),
  required = c(
    "tissue_groups", "offset", "disease_groups_altered", "assay_groups_altered"
  ),
  new_level_cols = "dataset_id_altered",
  new_level_tag = "__comprehensive_baseline__"
) {
  stopifnot(NROW(newdata) == 1L)
  newdata <- tibble::as_tibble(newdata)

  is_blank <- function(v) {
    length(v) == 0L ||
      is.na(v[[1L]]) ||
      (is.character(v[[1L]]) && !nzchar(v[[1L]]))
  }

  fit_vars <- colnames(fit$data)
  for (col in required) {
    if (!col %in% fit_vars) next
    if (!col %in% names(newdata) || is_blank(newdata[[col]])) {
      stop(sprintf("Required predictor `%s` is missing or empty.", col))
    }
  }

  to_expand <- character(0)
  for (col in marginalize_over) {
    if (!col %in% fit_vars) next
    if (!col %in% names(newdata) || is_blank(newdata[[col]])) {
      to_expand <- c(to_expand, col)
    }
  }

  if (length(to_expand)) {
    levels_list <- lapply(to_expand, function(col) fit_levels(fit, col))
    names(levels_list) <- to_expand
    keep <- vapply(levels_list, length, integer(1L)) > 0L
    levels_list <- levels_list[keep]
    to_expand <- names(levels_list)
  }

  if (length(to_expand)) {
    grid <- do.call(tidyr::expand_grid, levels_list)
    rest <- newdata[, setdiff(names(newdata), to_expand), drop = FALSE]
    out <- dplyr::bind_cols(
      rest[rep(1L, nrow(grid)), , drop = FALSE],
      grid
    )
  } else {
    out <- newdata
  }

  for (col in new_level_cols) {
    if (col %in% fit_vars) out[[col]] <- new_level_tag
  }
  out
}

newdata_comprehensive_gene <- function() {
  tibble::tibble(
    age_decade = NA_character_,
    sex = NA_character_,
    ethnicity_groups = NA_character_,
    assay_groups_altered = "10x Genomics 3",
    disease_groups_altered = "Normal",
    tissue_groups = "blood",
    offset = 0
  )
}

brms_comprehensive_mu_grand_mean_draws <- function(fit, ndraws = NULL) {
  nd_grid <- build_newdata_grid(fit, newdata_comprehensive_gene())
  log_mu_mat <- brms::posterior_linpred(
    fit,
    newdata = nd_grid,
    transform = FALSE,
    re_formula = NULL,
    allow_new_levels = TRUE,
    sample_new_levels = "gaussian",
    ndraws = ndraws
  )
  grand_mean_draws <- as.numeric(rowMeans(log_mu_mat))
  list(
    grand_mean_draws = grand_mean_draws,
    mean_grand_mean = mean(grand_mean_draws),
    sd_grand_mean = stats::sd(grand_mean_draws),
    n_grand = length(grand_mean_draws),
    n_grid_rows = nrow(nd_grid)
  )
}

# Welch t-test from (mean, SE, n) summaries on the log(μ) scale.
# Same definition as SAVI_case_study_universal_baseline.qmd.
welch_t_from_mean_se <- function(mean1, se1, n1, mean2, se2, n2) {
  n1 <- max(n1, 2L)
  n2 <- max(n2, 2L)
  sd1 <- se1 * sqrt(n1)
  sd2 <- se2 * sqrt(n2)
  se_diff <- sqrt(se1^2 + se2^2)
  t_stat <- (mean1 - mean2) / se_diff
  df <- (sd1^2 / n1 + sd2^2 / n2)^2 /
    ((sd1^2 / n1)^2 / (n1 - 1) + (sd2^2 / n2)^2 / (n2 - 1))
  p_value <- 2 * stats::pt(abs(t_stat), df = df, lower.tail = FALSE)
  list(t = t_stat, df = df, p_value = p_value)
}

edger_log_mu_cohort <- function(
  feature_id,
  cohort,
  fit,
  nb_dispersion,
  counts,
  offset_vec
) {
  gi <- match(feature_id, rownames(fit))
  if (is.na(gi) || !cohort %in% colnames(fit$coefficients)) {
    return(list(
      log_mu = NA_real_,
      se = NA_real_,
      df = NA_real_,
      n = 0L
    ))
  }

  disp <- nb_dispersion[gi]
  j <- which(fit$samples$group == cohort)
  if (!length(j)) {
    return(list(
      log_mu = NA_real_,
      se = NA_real_,
      df = NA_real_,
      n = 0L
    ))
  }

  log_mu <- edgeR::mglmOneGroup(
    matrix(counts[gi, j], nrow = 1),
    dispersion = disp,
    offset = offset_vec[j]
  )
  mu_all <- fit$fitted.values[gi, ]
  W <- mu_all / (1 + disp * mu_all)
  cov_un <- solve(t(fit$design) %*% (W * fit$design))
  se <- sqrt(fit$s2.post[gi] * cov_un[cohort, cohort])

  list(
    log_mu = unname(log_mu),
    se = unname(se),
    df = unname(fit$df.residual.adj[gi]),
    n = length(j)
  )
}

summarise_healthy_grand_mean_draws <- function(draws) {
  draws |>
    dplyr::group_by(.data$gene) |>
    dplyr::summarise(
      healthy_mean = mean(.data$log_mu),
      healthy_sd = stats::sd(.data$log_mu),
      healthy_n = dplyr::n(),
      .groups = "drop"
    )
}

welch_p_cohort_vs_healthy <- function(
  cohort_log_mu,
  cohort_se,
  cohort_df,
  healthy_mean,
  healthy_sd,
  healthy_n
) {
  welch_t_from_mean_se(
    mean1 = healthy_mean,
    se1 = healthy_sd,
    n1 = healthy_n,
    mean2 = cohort_log_mu,
    se2 = cohort_se,
    n2 = cohort_df + 1
  )$p_value
}

format_welch_p_label <- function(p) {
  # Match y-axis style: drop the leading zero before the decimal (0.05 → .05).
  drop0 <- function(x) sub("^(-?)0\\.", "\\1.", x)
  dplyr::case_when(
    is.na(p) ~ NA_character_,
    p < 0.001 ~ "p<.001",
    p < 0.01 ~ paste0("p=", drop0(sprintf("%.3f", p))),
    TRUE ~ paste0("p=", drop0(sprintf("%.2f", p)))
  )
}

compute_panel_c1_welch <- function(draws, cohort_lines) {
  healthy_by_gene <- summarise_healthy_grand_mean_draws(draws)

  out <- cohort_lines |>
    dplyr::left_join(healthy_by_gene, by = "gene")

  out$welch_p <- mapply(
    welch_p_cohort_vs_healthy,
    out$cohort_log_mu,
    out$cohort_se,
    out$cohort_df,
    out$healthy_mean,
    out$healthy_sd,
    out$healthy_n
  )

  out$p_label <- format_welch_p_label(out$welch_p)
  out
}
```

## C1. Comprehensive log μ density (16 genes)

**Purpose.** For the 16 covered STING-signature genes, show the healthy
comprehensive posterior of log(μ) at offset 0 (grey density) against cohort
NB MLE log(μ) for study control, untreated SAVI, and JAK-inhibitor–treated
SAVI (coloured vertical lines). Layout matches the case-study density grid
(`SAVI_case_study_universal_baseline.qmd`), as a **4 × 4** facet panel.

**Estimand (aligned with Panel b comprehensive null).**

- Fixed: `tissue = blood`, `disease = Normal`, `assay = 10x Genomics 3`,
  `offset = 0`
- Marginalised (equal average of fitted levels): sex, age decade, ethnicity
- Dataset: unseen study level (`allow_new_levels`, Gaussian new levels)
- Quantity: per-draw mean of `posterior_linpred(..., transform = FALSE)` over
  the covariate grid (grand-mean log μ), not ZINB expected counts

Cohort lines use edgeR `mglmOneGroup` on monocytic pseudobulk counts at the
observed library offsets (same as the universal-baseline case study). Welch
$p$-values compare each cohort NB MLE log(μ) to the comprehensive healthy
grand-mean posterior (`welch_t_from_mean_se` in
`SAVI_case_study_universal_baseline.qmd`).

```{r}
#| label: panel-c-mu-density-data

rebuild_panel_c_mu <- isTRUE(params$run_panel_c_rebuild) ||
  !file.exists(params$panel_c_mu_density_rds)

if (!rebuild_panel_c_mu) {
  panel_c_mu_cache <- readRDS(params$panel_c_mu_density_rds)
  ok_cache <- is.list(panel_c_mu_cache) &&
    all(c("draws", "cohort_log_mu", "gene_order") %in% names(panel_c_mu_cache)) &&
    identical(as.character(panel_c_mu_cache$gene_order), panel_c_gene_order) &&
    "cohort_se" %in% names(panel_c_mu_cache$cohort_log_mu)
  if (!ok_cache) {
    message("Ignoring stale Panel-c mu-density cache.")
    rebuild_panel_c_mu <- TRUE
  } else {
    panel_c_mu_draws <- panel_c_mu_cache$draws
    panel_c_cohort_log_mu <- panel_c_mu_cache$cohort_log_mu
    panel_c_n_grid_rows <- panel_c_mu_cache$n_grid_rows
    message(
      "Loaded Panel-c comprehensive mu density: ",
      nrow(panel_c_mu_draws), " draws; ",
      nrow(panel_c_cohort_log_mu), " cohort lines"
    )
  }
}

if (rebuild_panel_c_mu) {
  se_cache <- readRDS(params$panel_c_se_rds)
  se_monocytic_panel_c <- se_cache$se_monocytic
  panel_c_estimates <- readRDS(params$estimates_rds)
  stopifnot(all(panel_c_gene_order %in% panel_c_estimates$gene))

  counts_mu <- SummarizedExperiment::assay(se_monocytic_panel_c, "counts")
  meta_mu <- SummarizedExperiment::colData(se_monocytic_panel_c) |>
    as.data.frame()
  offset_vec_mu <- if ("offset" %in% names(meta_mu)) {
    meta_mu$offset
  } else {
    log(1 / meta_mu$multiplier)
  }
  group_mu <- as.character(meta_mu$Category)
  group_mu[is.na(group_mu)] <- "reference"
  meta_mu$group <- factor(group_mu) |> forcats::fct_relevel("reference")

  dge_mu <- edgeR::DGEList(
    counts = counts_mu,
    samples = meta_mu,
    genes = as.data.frame(SummarizedExperiment::rowData(se_monocytic_panel_c))
  )
  dge_mu$offset <- matrix(
    offset_vec_mu, nrow = nrow(dge_mu), ncol = ncol(dge_mu), byrow = TRUE
  )
  design_mu <- stats::model.matrix(~ 0 + group, data = meta_mu)
  colnames(design_mu) <- sub("^group", "", colnames(design_mu))
  dge_mu <- edgeR::estimateDisp(dge_mu, design_mu)
  fit_mu <- edgeR::glmQLFit(dge_mu, design_mu)
  nb_dispersion_mu <- if (!is.null(dge_mu$trended.dispersion)) {
    dge_mu$trended.dispersion
  } else if (!is.null(dge_mu$tagwise.dispersion)) {
    dge_mu$tagwise.dispersion
  } else {
    rep_len(dge_mu$common.dispersion, nrow(dge_mu))
  }

  cohort_levels_mu <- names(panel_c_category_to_state)
  draws_list <- vector("list", length(panel_c_gene_order))
  cohort_list <- vector("list", length(panel_c_gene_order))
  names(draws_list) <- names(cohort_list) <- panel_c_gene_order
  panel_c_n_grid_rows <- NA_integer_

  for (g in panel_c_gene_order) {
    row <- panel_c_estimates[match(g, panel_c_estimates$gene), ]
    fit <- row$brms_fit[[1]]
    mu <- brms_comprehensive_mu_grand_mean_draws(fit)
    panel_c_n_grid_rows <- mu$n_grid_rows
    draws_list[[g]] <- tibble::tibble(
      gene = g,
      .feature = row$.feature,
      .draw = seq_along(mu$grand_mean_draws),
      log_mu = mu$grand_mean_draws
    )
    cohort_list[[g]] <- lapply(cohort_levels_mu, function(cat) {
      cm <- edger_log_mu_cohort(
        row$.feature, cat, fit_mu, nb_dispersion_mu, counts_mu, offset_vec_mu
      )
      tibble::tibble(
        gene = g,
        .feature = row$.feature,
        Category = cat,
        state = unname(panel_c_category_to_state[[cat]]),
        cohort_log_mu = cm$log_mu,
        cohort_se = cm$se,
        cohort_df = cm$df,
        n_libraries = cm$n
      )
    }) |>
      dplyr::bind_rows()
  }

  panel_c_mu_draws <- dplyr::bind_rows(draws_list)
  panel_c_cohort_log_mu <- dplyr::bind_rows(cohort_list)

  saveRDS(
    list(
      estimand = paste(
        "comprehensive: blood+Normal+10xGenomics3+offset0;",
        "equal-avg over sex×age×ethnicity; dataset new level"
      ),
      gene_order = panel_c_gene_order,
      n_grid_rows = panel_c_n_grid_rows,
      draws = panel_c_mu_draws,
      cohort_log_mu = panel_c_cohort_log_mu
    ),
    params$panel_c_mu_density_rds,
    compress = "xz"
  )
}

panel_c_cohort_log_mu <- panel_c_cohort_log_mu |>
  dplyr::mutate(
    gene = factor(.data$gene, levels = panel_c_gene_order),
    state = factor(.data$state, levels = state_levels)
  )

panel_c_mu_draws <- panel_c_mu_draws |>
  dplyr::mutate(gene = factor(.data$gene, levels = panel_c_gene_order))

panel_c_cohort_log_mu |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), ~ round(.x, 3)))
```

```{r}
#| label: panel-c1-draw
#| fig-width: 7.2
#| fig-height: 7.2
#| fig-cap: "Figure 2c1. Comprehensive healthy posterior of log(μ) (grey density) versus cohort NB MLE log(μ) for control, untreated SAVI, and JAK-inhibitor–treated SAVI (coloured lines). One facet per covered STING-signature gene (4 × 4)."

plot_panel_c1_mu_density <- function(draws, cohort_lines, welch_annot) {
  # Stack p-labels on the left of each facet (not at the vline x), so close
  # cohort estimates cannot collide.
  welch_left <- welch_annot |>
    dplyr::mutate(
      state = factor(as.character(.data$state), levels = state_levels),
      label_vjust = 1.35 + 1.55 * (as.integer(.data$state) - 1L)
    )

  ggplot2::ggplot() +
    ggplot2::geom_density(
      data = draws,
      ggplot2::aes(x = .data$log_mu),
      fill = col_light,
      colour = col_mid,
      alpha = 0.7,
      linewidth = 0.3
    ) +
    ggplot2::geom_vline(
      data = cohort_lines,
      ggplot2::aes(
        xintercept = .data$cohort_log_mu,
        colour = .data$state
      ),
      linewidth = 0.3,
      alpha = 0.9
    ) +
    ggplot2::geom_text(
      data = welch_left,
      ggplot2::aes(
        x = -Inf,
        y = Inf,
        label = .data$p_label,
        colour = .data$state,
        vjust = .data$label_vjust
      ),
      inherit.aes = FALSE,
      hjust = -0.08,
      size = 2,
      show.legend = FALSE
    ) +
    ggplot2::scale_colour_manual(
      values = state_colours,
      labels = state_labels_short,
      name = NULL
    ) +
    ggplot2::scale_y_continuous(labels = label_drop_leading_zero()) +
    ggplot2::facet_wrap(~ gene, ncol = 4, nrow = 4, scales = "free") +
    ggplot2::labs(
      x = expression(log(mu)~"(offset 0)"),
      y = "Density"
    ) +
    theme_multipanel +
    ggplot2::theme(
      legend.position = "bottom"
    )
}

panel_c1_welch <- compute_panel_c1_welch(panel_c_mu_draws, panel_c_cohort_log_mu)

panel_c1_welch |>
  dplyr::select(gene, state, cohort_log_mu, welch_p, p_label) |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), ~ round(.x, 4)))

panel_c1 <- plot_panel_c1_mu_density(
  panel_c_mu_draws,
  panel_c_cohort_log_mu,
  panel_c1_welch
)
panel_c1
```

```{r}
#| label: panel-c1-audit-text

cat(
  sprintf(
    paste(
      "Panel c1: %d genes; comprehensive HCA grid = %d covariate profiles;",
      "%d posterior draws per gene; cohort lines = edgeR NB MLE log(μ)",
      "for CTRL (n=%d), untreated (n=%d), treated (n=%d).\n"
    ),
    length(panel_c_gene_order),
    as.integer(panel_c_n_grid_rows),
    nrow(panel_c_mu_draws) / length(panel_c_gene_order),
    unique(panel_c_cohort_log_mu$n_libraries[panel_c_cohort_log_mu$state == "control"]),
    unique(panel_c_cohort_log_mu$n_libraries[panel_c_cohort_log_mu$state == "untreated"]),
    unique(panel_c_cohort_log_mu$n_libraries[panel_c_cohort_log_mu$state == "treated"])
  )
)
```

## C2. *ADRB2* personalised predictive counts (offset 0)

Single-gene audit for **ADRB2**: healthy `posterior_predict()` counts under a
**personalised** null (untreated age × sex; assay fixed to `10x Genomics 3`;
ethnicity and dataset marginalised), evaluated at **`offset = 0`** so predictions
sit on the same reference-depth scale as TMM-scaled abundance. The density
shades the **central 95%** of the predictive distribution; vertical lines mark
each untreated patient's observed **`counts_scaled`**.

```{r}
#| label: panel-c2-adrb2-personalised-data

marginalize_predict_draws <- function(y_rep_mat) {
  if (is.null(dim(y_rep_mat)) || ncol(y_rep_mat) <= 1L) {
    return(as.numeric(y_rep_mat))
  }
  as.numeric(y_rep_mat)
}

rebuild_panel_c_adrb2 <- isTRUE(params$run_panel_c_rebuild) ||
  !file.exists(params$panel_c_adrb2_personalised_rds)

if (!rebuild_panel_c_adrb2) {
  panel_c_adrb2_cache <- readRDS(params$panel_c_adrb2_personalised_rds)
  ok_adrb2 <- is.list(panel_c_adrb2_cache) &&
    identical(panel_c_adrb2_cache$gene, "ADRB2") &&
    all(c("draws", "untreated_obs") %in% names(panel_c_adrb2_cache))
  if (!ok_adrb2) {
    message("Ignoring stale ADRB2 personalised cache.")
    rebuild_panel_c_adrb2 <- TRUE
  } else {
    panel_c_adrb2_draws <- panel_c_adrb2_cache$draws
    panel_c_adrb2_obs <- panel_c_adrb2_cache$untreated_obs
    message(
      "Loaded ADRB2 personalised predictive draws: ",
      nrow(panel_c_adrb2_draws), " rows"
    )
  }
}

if (rebuild_panel_c_adrb2) {
  if (!exists("se_monocytic_panel_c", inherits = FALSE)) {
    se_monocytic_panel_c <- readRDS(params$panel_c_se_rds)$se_monocytic
  }
  if (!exists("panel_c_estimates", inherits = FALSE)) {
    panel_c_estimates <- readRDS(params$estimates_rds)
  }

  adrb2_row <- panel_c_estimates[match("ADRB2", panel_c_estimates$gene), ]
  adrb2_fit <- adrb2_row$brms_fit[[1]]
  adrb2_feature <- adrb2_row$.feature

  panel_c_adrb2_obs <- savi_libraries |>
    dplyr::filter(as.character(.data$state) == "untreated") |>
    dplyr::transmute(
      library_id,
      subject_id,
      sex = as.character(.data$sex),
      age_years,
      age_decade = as.character(ceiling(.data$age_years / 10))
    )

  cd_c2 <- SummarizedExperiment::colData(se_monocytic_panel_c) |>
    as.data.frame() |>
    tibble::rownames_to_column("col_id") |>
    dplyr::mutate(library_id = as.character(.data$Sample))

  panel_c_adrb2_obs <- panel_c_adrb2_obs |>
    dplyr::left_join(
      cd_c2 |>
        dplyr::select("col_id", "library_id", "offset"),
      by = "library_id"
    ) |>
    dplyr::mutate(
      observed_count = as.numeric(
        SummarizedExperiment::assay(
          se_monocytic_panel_c, "counts"
        )[adrb2_feature, .data$col_id]
      ),
      observed_scaled = as.numeric(
        SummarizedExperiment::assay(
          se_monocytic_panel_c, "counts_scaled"
        )[adrb2_feature, .data$col_id]
      ),
      offset_obs = as.numeric(.data$offset)
    ) |>
    dplyr::select(-"col_id", -"offset")

  stopifnot(nrow(panel_c_adrb2_obs) == 5L)

  ndraws_c2 <- 400L
  set.seed(20260809L)
  draws_list <- vector("list", nrow(panel_c_adrb2_obs))
  n_eth <- NA_integer_
  for (i in seq_len(nrow(panel_c_adrb2_obs))) {
    nd <- tibble::tibble(
      age_decade = panel_c_adrb2_obs$age_decade[[i]],
      sex = panel_c_adrb2_obs$sex[[i]],
      ethnicity_groups = NA_character_,
      assay_groups_altered = "10x Genomics 3",
      disease_groups_altered = "Normal",
      tissue_groups = "blood",
      offset = 0
    )
    nd_grid <- build_newdata_grid(
      adrb2_fit,
      nd,
      marginalize_over = "ethnicity_groups",
      required = c(
        "tissue_groups", "offset", "disease_groups_altered",
        "assay_groups_altered", "sex", "age_decade"
      ),
      new_level_tag = paste0("__pers__", panel_c_adrb2_obs$subject_id[[i]])
    )
    y_mat <- brms::posterior_predict(
      adrb2_fit,
      newdata = nd_grid,
      summary = FALSE,
      re_formula = NULL,
      allow_new_levels = TRUE,
      sample_new_levels = "gaussian",
      ndraws = ndraws_c2
    )
    n_eth <- ncol(y_mat)
    draws_list[[i]] <- tibble::tibble(
      subject_id = panel_c_adrb2_obs$subject_id[[i]],
      library_id = panel_c_adrb2_obs$library_id[[i]],
      draw = marginalize_predict_draws(y_mat)
    )
  }

  panel_c_adrb2_draws <- dplyr::bind_rows(draws_list)
  saveRDS(
    list(
      gene = "ADRB2",
      .feature = adrb2_feature,
      estimand = paste(
        "personalised untreated age×sex; assay 10x3;",
        "ethnicity marginalised; offset=0; posterior_predict"
      ),
      ndraws_per_profile = ndraws_c2,
      n_grid_ethnicity = n_eth,
      draws = panel_c_adrb2_draws,
      untreated_obs = panel_c_adrb2_obs
    ),
    params$panel_c_adrb2_personalised_rds,
    compress = "xz"
  )
}

panel_c_adrb2_obs <- panel_c_adrb2_obs |>
  dplyr::mutate(
    subject_id = factor(
      .data$subject_id,
      levels = c("P1", "P2", "P4", "P5", "P6")
    )
  )

panel_c_adrb2_obs
```

```{r}
#| label: panel-c2-draw
#| fig-width: 5.5
#| fig-height: 3.2
#| fig-cap: "Figure 2c2. ADRB2 personalised healthy posterior predictive counts at offset 0. Teal fill = central 95% of the predictive density (equal mixture over untreated age×sex); light grey = tails. Coloured lines = TMM-scaled observed counts for untreated SAVI patients."

plot_panel_c2_adrb2 <- function(draws, obs, cred_mass = 0.95) {
  x_draws <- log1p(pmax(draws$draw, 0))
  alpha_tail <- (1 - cred_mass) / 2
  q_ci <- as.numeric(
    stats::quantile(x_draws, probs = c(alpha_tail, 1 - alpha_tail))
  )

  dens <- stats::density(x_draws, n = 512)
  dens_df <- tibble::tibble(x = dens$x, y = pmax(dens$y, 0))
  dens_mid <- dens_df |>
    dplyr::filter(.data$x >= q_ci[[1]], .data$x <= q_ci[[2]])
  dens_lo <- dens_df |>
    dplyr::filter(.data$x <= q_ci[[1]])
  dens_hi <- dens_df |>
    dplyr::filter(.data$x >= q_ci[[2]])

  line_df <- obs |>
    dplyr::mutate(x = log1p(pmax(.data$observed_scaled, 0)))

  ggplot2::ggplot() +
    ggplot2::geom_area(
      data = dens_lo,
      ggplot2::aes(x = .data$x, y = .data$y),
      fill = col_light,
      colour = NA,
      alpha = 0.9
    ) +
    ggplot2::geom_area(
      data = dens_hi,
      ggplot2::aes(x = .data$x, y = .data$y),
      fill = col_light,
      colour = NA,
      alpha = 0.9
    ) +
    ggplot2::geom_area(
      data = dens_mid,
      ggplot2::aes(x = .data$x, y = .data$y, fill = "Central 95%"),
      colour = NA,
      alpha = 0.55
    ) +
    ggplot2::geom_line(
      data = dens_df,
      ggplot2::aes(x = .data$x, y = .data$y),
      colour = col_mid,
      linewidth = 0.35
    ) +
    ggplot2::geom_vline(
      xintercept = q_ci,
      colour = col_meta,
      linetype = "dashed",
      linewidth = 0.35,
      alpha = 0.85
    ) +
    ggplot2::geom_vline(
      data = line_df,
      ggplot2::aes(xintercept = .data$x, colour = .data$subject_id),
      linewidth = 0.5,
      alpha = 0.95
    ) +
    ggplot2::scale_fill_manual(
      values = c("Central 95%" = col_meta),
      name = NULL
    ) +
    ggplot2::scale_colour_manual(
      values = setNames(
        scales::hue_pal()(nrow(obs)),
        levels(obs$subject_id)
      ),
      name = "Untreated"
    ) +
    ggplot2::scale_x_continuous(labels = label_drop_leading_zero()) +
    ggplot2::scale_y_continuous(labels = label_drop_leading_zero()) +
    ggplot2::labs(
      x = expression(log[1 + p]("count @ offset 0")),
      y = "Density"
    ) +
    theme_multipanel +
    ggplot2::theme(legend.position = "bottom") +
    ggplot2::guides(
      fill = ggplot2::guide_legend(
        override.aes = list(alpha = 0.55, colour = NA)
      )
    )
}

panel_c2 <- plot_panel_c2_adrb2(panel_c_adrb2_draws, panel_c_adrb2_obs)
panel_c2
```

```{r}
#| label: panel-c2-audit-text

panel_c_adrb2_x <- log1p(pmax(panel_c_adrb2_draws$draw, 0))
panel_c_adrb2_q <- as.numeric(stats::quantile(panel_c_adrb2_x, c(0.025, 0.975)))

cat(
  sprintf(
    paste(
      "Panel c2 (ADRB2): personalised posterior_predict at offset 0 for %d",
      "untreated age×sex profiles (equal mixture); %d predictive draws;",
      "central 95%% on log1p scale = [%.2f, %.2f];",
      "vertical lines = counts_scaled for P1/P2/P4/P5/P6",
      "(scaled = %.0f, %.0f, %.0f, %.0f, %.0f).\n"
    ),
    dplyr::n_distinct(panel_c_adrb2_draws$subject_id),
    nrow(panel_c_adrb2_draws),
    panel_c_adrb2_q[[1]],
    panel_c_adrb2_q[[2]],
    panel_c_adrb2_obs$observed_scaled[panel_c_adrb2_obs$subject_id == "P1"],
    panel_c_adrb2_obs$observed_scaled[panel_c_adrb2_obs$subject_id == "P2"],
    panel_c_adrb2_obs$observed_scaled[panel_c_adrb2_obs$subject_id == "P4"],
    panel_c_adrb2_obs$observed_scaled[panel_c_adrb2_obs$subject_id == "P5"],
    panel_c_adrb2_obs$observed_scaled[panel_c_adrb2_obs$subject_id == "P6"]
  )
)
```

## C3. *ADRB2* UMAP (all donors vs high-depth donor removed)

Disease-associated monocytes from the published GSE226598 Seurat object
(`params$savi_sc_celltype`), SCT `data`, split by analysis arm. Short facet
labels: **CTRL / Untreated / Treated**. Top row: all donors. Bottom row: same
plot after removing every cell from **`params$adrb2_outlier_sample`**
(`P6_STING_ht`), the high-depth untreated library that can dominate *ADRB2*
pseudobulk and single-cell colouring.

```{r}
#| label: panel-c3-seurat-mono

seurat_reduction_name <- function(se) {
  nms <- names(se@reductions)
  hit <- intersect(c("umap", "UMAP", "tsne", "TSNE"), nms)
  if (length(hit)) {
    return(hit[[1]])
  }
  nms[[1]]
}

category_to_plot_label <- c(
  CTRL = "CTRL",
  SAVI = "Untreated",
  SAVI_treated = "Treated"
)

seurat_savi_c3 <- read_geo_rds_gz(params$savi_seurat_gz)
stopifnot("Category" %in% colnames(seurat_savi_c3@meta.data))
stopifnot("Sample" %in% colnames(seurat_savi_c3@meta.data))
stopifnot("CellType" %in% colnames(seurat_savi_c3@meta.data))

reduction_use <- seurat_reduction_name(seurat_savi_c3)

seurat_savi_mono <- subset(
  seurat_savi_c3,
  subset = CellType == params$savi_sc_celltype
)
rm(seurat_savi_c3)
gc(verbose = FALSE)

if (ncol(seurat_savi_mono) == 0L) {
  stop(
    "No cells for CellType = '", params$savi_sc_celltype,
    "'. Check `CellType` levels in the Seurat object."
  )
}
if (!"SCT" %in% SeuratObject::Assays(seurat_savi_mono)) {
  stop(
    "SCT assay not found. Available: ",
    paste(SeuratObject::Assays(seurat_savi_mono), collapse = ", ")
  )
}
if (!"ADRB2" %in% rownames(seurat_savi_mono[["SCT"]])) {
  stop("ADRB2 not found in SCT assay rownames.")
}

seurat_savi_mono$Category_plot <- factor(
  unname(category_to_plot_label[as.character(seurat_savi_mono$Category)]),
  levels = c("CTRL", "Untreated", "Treated")
)
if (anyNA(seurat_savi_mono$Category_plot)) {
  stop("Unexpected Category levels in Seurat metadata.")
}

outlier_sample <- params$adrb2_outlier_sample
n_outlier_cells <- sum(seurat_savi_mono$Sample == outlier_sample, na.rm = TRUE)
if (n_outlier_cells == 0L) {
  stop("Outlier sample '", outlier_sample, "' not found in monocyte subset.")
}

seurat_savi_mono_no_outlier <- subset(
  seurat_savi_mono,
  subset = Sample != outlier_sample
)

tibble::tibble(
  cell_type = params$savi_sc_celltype,
  reduction = reduction_use,
  n_cells_all = ncol(seurat_savi_mono),
  n_cells_no_outlier = ncol(seurat_savi_mono_no_outlier),
  outlier_sample = outlier_sample,
  n_outlier_cells = n_outlier_cells
)
```

```{r}
#| label: panel-c3-helpers

style_panel_c3_featureplot <- function(p) {
  p &
    ggplot2::theme(
      plot.title = ggplot2::element_text(size = 7),
      plot.subtitle = ggplot2::element_blank(),
      strip.background = ggplot2::element_blank(),
      strip.text = ggplot2::element_text(size = 7),
      axis.title = ggplot2::element_text(size = 7),
      axis.title.y.right = ggplot2::element_text(size = 7),
      axis.text = ggplot2::element_text(size = 6),
      legend.title = ggplot2::element_text(size = 7),
      legend.text = ggplot2::element_text(size = 6),
      legend.key.size = grid::unit(4, "mm"),
      plot.margin = ggplot2::margin(0, 0, 0, 0)
    )
}

panel_c3_row_label <- function(label) {
  ggplot2::ggplot() +
    ggplot2::annotate(
      "text",
      x = 0.5,
      y = 0.5,
      label = label,
      angle = 90,
      size = 2.3,
      colour = col_dark,
      fontface = "bold"
    ) +
    ggplot2::coord_cartesian(xlim = c(0, 1), ylim = c(0, 1), clip = "off") +
    ggplot2::theme_void() +
    ggplot2::theme(plot.margin = ggplot2::margin(0, 0, 0, 0))
}

plot_adrb2_umap_by_category <- function(seurat_obj, reduction) {
  style_panel_c3_featureplot(
    Seurat::FeaturePlot(
      seurat_obj,
      features = "ADRB2",
      assay = "SCT",
      split.by = "Category_plot",
      reduction = reduction,
      slot = "data",
      ncol = 3,
      order = TRUE,
      pt.size = 0.25
    )
  )
}
```

```{r}
#| label: panel-c3-draw
#| fig-width: 7.2
#| fig-height: 4.2
#| fig-cap: "Figure 2c3. ADRB2 on disease-associated-monocyte UMAP, split by CTRL / Untreated / Treated. Top: all donors. Bottom: P6_STING_ht cells removed."

p_adrb2_umap_all <- plot_adrb2_umap_by_category(seurat_savi_mono, reduction_use)
p_adrb2_umap_no_p6 <- plot_adrb2_umap_by_category(
  seurat_savi_mono_no_outlier,
  reduction_use
)

panel_c3 <- 
  (
    ( p_adrb2_umap_all & theme(axis.title.x.bottom = element_blank()) )
  /
    ( p_adrb2_umap_no_p6 & theme(plot.title = element_blank()))
) +
  patchwork::plot_layout(heights = c(1, 1)) 

panel_c3
```

```{r}
#| label: panel-c3-audit-text

cat(
  sprintf(
    paste(
      "Panel c3 (ADRB2 UMAP): cell type = %s; reduction = %s;",
      "%d cells all donors (%s); %d cells after removing %s (%d cells dropped).",
      "Facets: CTRL / Untreated / Treated (from Category CTRL / SAVI / SAVI_treated).\n"
    ),
    params$savi_sc_celltype,
    reduction_use,
    ncol(seurat_savi_mono),
    paste(
      sprintf(
        "%s=%d",
        names(table(seurat_savi_mono$Category_plot)),
        as.integer(table(seurat_savi_mono$Category_plot))
      ),
      collapse = ", "
    ),
    ncol(seurat_savi_mono_no_outlier),
    outlier_sample,
    n_outlier_cells
  )
)
```

------------------------------------------------------------------------

# Next panels

Panels d–f will be added in this v1 report as their analyses land.

```{r}
sessionInfo()
```

