---
title: "SAVI case study — personalised posteriorHCA baseline"
subtitle: "STING signature against donor-matched healthy monocytic references"
date: today
date-format: "DD MMM YYYY"
title-block-banner: true
author:
  - name: Chen Zhan
    orcid: 0000-0002-4794-8339
    email: chen.zhan@adelaide.edu.au
    affiliation:
      - id: sgc
        name: SAiGENCI, The University of Adelaide
format:
  html:
    theme: minty
    df-print: paged
    code-line-numbers: true
    embed-resources: true
    code-fold: true
    code-fold-default: true
    toc: true
    toc-location: left
    toc-title: "Contents"
knitr:
  opts_chunk:
    message: false
    warning: false
editor: source
abstract: |
  This notebook treats the SAVI cohort as a personalised posteriorHCA baseline
  case study. We take a literature-defined STING activation gene set, align it
  to Ensembl gene ids present in the posterior HCA monocytic `targets` store,
  optionally re-fit and compress `brms` models with `poco`, then compare each
  SAVI pseudobulk against a donor-matched healthy monocytic counterfactual. The
  statistical layer follows the HCC personalised baseline report: primary
  posterior predictive count checks are paired with offset-centred log(mu)
  tests from `posterior_linpred(transform = FALSE)`.
params:
  run_fits: false
  n_components: 5
  corr_threshold: 0.15
  target_cell_type: "monocytic"
  estimates_rds: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/dev/SAVI/data/SAVI_estimates_chunk.rds"
  savi_pseudobulk_rds: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/dev/SAVI/data/GSE226598_SAVI_pseudobulk_Sample_CellType.rds"
  savi_seurat_gz: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/dev/SAVI/data/GSE226598_SeuratObject_SAVI.rds.gz"
  savi_sc_celltype: "17. Disease-associated monocytes"
  pseudobulk_h5ad: "/home/a1237163/lab/Mangiola_ImmuneAtlas/zenodo/pseudobulk_se.h5ad"
  composition_predict_rds: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/dev/SAVI/data/SAVI_composition_predict_res.rds"
  sccomp_estimates_rds: "/home/a1237163/lab/Mangiola_ImmuneAtlas/taskforce_shared_folder/sccomp_on_cellNexus_1_0_12/estimates_age_decade___L3___disease_TRUE___immune_only_TRUE.rds"
  model_store_root: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/HPC_posterior/NEW_CELL_TYPE"
---

# Introduction

[STING](https://en.wikipedia.org/wiki/Stimulator_of_interferon_genes)-associated vasculopathy with onset in infancy (**SAVI**) is caused by gain-of-function mutations in *STING1*. De Cevins *et al.* performed single-cell RNA-seq of PBMCs from SAVI patients and healthy controls, and an IFN-β time-course in healthy PBMCs, to dissect IFN-dependent and IFN-independent programmes. They highlight disease-associated monocytes and propose a **21-gene** transcriptomic signature of STING activation that is **enriched for TNF-α / NF-κB–related biology** and **filtered against IFN-β–induced genes** in monocytes and DCs (Figure 6; Table S6).

**Primary publication (open access):** de Cevins, C., *et al.* (2023). Single-cell RNA-sequencing of PBMCs from SAVI patients reveals disease-associated monocytes with elevated integrated stress response. *Cell Reports Medicine* **4**, 101333. <https://doi.org/10.1016/j.xcrm.2023.101333> — PMC: <https://pmc.ncbi.nlm.nih.gov/articles/PMC10772457/>.

**Data:** scRNA-seq are deposited under superseries [GSE226601](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE226601): SAVI cohort [GSE226598](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE226598), IFN-β dataset [GSE226572](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE226572). Pseudobulk for this notebook is prepared in `dev/SAVI/data/prepare_data.R` from the GEO Seurat object.

**Scope of this document:** We do **not** reproduce the paper’s within-study differential analysis or IFN-β filtering steps here. We use their **published 21-gene symbol list** (Table S6 / Figure 6) as biomarkers, fit or load **monocytic-lineage** reference models from the lab `targets` store, and test each **SAVI sample × cell type** pseudobulk against a **personalised healthy atlas baseline**. The primary test asks whether the observed count is surprising under `posterior_predict()` for a matched healthy monocyte; the secondary test asks whether the sample's offset-centred `log(mu)` differs from the matched healthy `posterior_linpred()` distribution.

------------------------------------------------------------------------

# Library setup

```{r}
#| label: setup
# HPC / project R library (adjust if running elsewhere)
.libPaths("/hpcfs/users/a1237163/R/x86_64-pc-linux-gnu-library/4.5")

library(tibble)
library(tidyr)
library(dplyr)
library(tidySummarizedExperiment)
library(SummarizedExperiment)
library(S4Vectors)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(tidybulk)
library(glue)
library(purrr)
library(furrr)
library(future)
library(lobstr)
library(posterior)
library(progressr)
library(brms)
library(poco)
library(tidyverse)
library(ggplot2)
library(forcats)
library(patchwork)
library(pheatmap)
library(RColorBrewer)
library(targets)
library(stringr)
library(Seurat)
library(tidyseurat)
```

------------------------------------------------------------------------

# STING activation signature (21 genes)

The symbols below follow **Table S6** / Figure 6 in de Cevins *et al.* (2023). We map HGNC symbols to Ensembl via `org.Hs.eg.db` for overlap with the reference model feature space; **lncRNAs and some readthrough ids** may return `NA` here even when present in the single-cell count matrix — those rows are dropped from the `targets`-based extraction unless you supply Ensembl ids manually.

```{r}
#| label: sting-signature
sting_signature_21 <- tibble(
  gene = c(
    "AC253572.2",
    "ADRB2",
    "ANKRD28",
    "ANXA1",
    "AQP9",
    "AZIN1-AS1",
    "IL1B",
    "JARID2",
    "KLF4",
    "MIR4435-2HG",
    "OTUD1",
    "PLAUR",
    "PTGS2",
    "PTX3",
    "SELENOK",
    "TGIF1",
    "TREM1",
    "TSC22D2",
    "ZBTB43",
    "ZFP36L2",
    "ZNF503"
  )
) |>
  mutate(
    .feature = AnnotationDbi::mapIds(
      org.Hs.eg.db::org.Hs.eg.db,
      keys = gene,
      column = "ENSEMBL",
      keytype = "SYMBOL",
      multiVals = "first"
    )
  )

sting_signature_21
```

------------------------------------------------------------------------

# Reference model store (`targets`)

Posterior HCA models live under `params$model_store_root`, in a subfolder named after the harmonised cell type (here `V1_monocytic` when `params$target_cell_type` is `"monocytic"`). Branch metadata encodes which Ensembl gene each `estimates_chunk_*` target belongs to (`Gene:___ENSG...___` in the recorded warning field).

```{r}
#| label: model-meta
model_path <- file.path(
  params$model_store_root,
  paste0("V1_", params$target_cell_type)
)

model_meta <-
  targets::tar_meta(
    starts_with("estimates_chunk_"),
    store = file.path(model_path, "_targets")
  ) |>
  dplyr::filter(type == "branch") |>
  mutate(
    .feature = stringr::str_extract(
      warnings,
      "(?<=Gene:___)ENSG\\d+(?=___)"
    )
  )

# How many of the 21 (with successful SYMBOL→ENSEMBL map) appear in the store?
n_sig_in_store <- sum(sting_signature_21$.feature %in% model_meta$.feature, na.rm = TRUE)
n_sig_mapped <- sum(!is.na(sting_signature_21$.feature))

tibble(
  signature_genes = nrow(sting_signature_21),
  mapped_to_ensembl = n_sig_mapped,
  present_in_targets_store = n_sig_in_store
)
```

------------------------------------------------------------------------

# Fit and compress `brms` models (optional, heavy)

Set **`params$run_fits: true`** when rendering (or in the YAML) to re-run extraction, posterior correlation partitioning, `poco::compress_brmsfit`, and `saveRDS`. By default this chunk is **skipped** and the next section loads `params$estimates_rds`.

For each gene: read the precomputed `estimates_chunk` from disk (`tar_read_raw()` from the **targets** package, same usage as `dev/posteriorHCA_brms_fit_compression_test.qmd`), strip large environments from the fit, optionally compress with `poco`, and record layout / fidelity metrics.

```{r}
#| label: fit-sting-biomarkers
#| eval: !expr params$run_fits

# handlers(global = TRUE)
handlers("progress")

options(future.globals.maxSize = 5 * 1024^3)

n_components <- params$n_components
corr_threshold <- params$corr_threshold

extract_one_biomarker <- function(ensg) {
  estimates_chunk <-
    model_meta |>
    dplyr::filter(.feature == ensg) |>
    dplyr::pull(name) |>
    tar_read_raw(store = file.path(model_path, "_targets")) |>
    dplyr::select(-tar_group) |>
    mutate(
      gene = sting_signature_21 |>
        dplyr::filter(!is.na(.feature), .feature == ensg) |>
        dplyr::pull(gene),
      cell_type = params$target_cell_type,
      brms_fit = purrr::map(brms_fit, poco::strip_brmsfit_envs)
    )

  fit_brms <- estimates_chunk$brms_fit[[1]]

  posterior_cor <- poco::posterior_correlation(
    fit_brms,
    type = "correlation",
    method = "pearson"
  )
  original_draws <- posterior::as_draws_matrix(fit_brms)

  partition <- poco::partition_parameters_clusters(
    posterior_cor,
    threshold = corr_threshold,
    min_size = 2L,
    simple_output = FALSE
  )

  layout_tbl <- tibble(
    n_components = n_components,
    corr_threshold = corr_threshold,
    n_blocks = length(partition$blocks),
    block_size_max = if (length(partition$blocks)) max(lengths(partition$blocks)) else 0L,
    block_size_med = if (length(partition$blocks)) {
      as.integer(stats::median(lengths(partition$blocks)))
    } else {
      0L
    },
    n_remainder = length(partition$remainder)
  )

  compress_tic <- Sys.time()
  compressed_blockwise <- tryCatch(
    poco::compress_brmsfit(
      brmsfit = fit_brms,
      method = "mclust",
      n_components = n_components,
      partition = partition,
      cluster_BPPARAM = 1L,
      verbose = FALSE
    ),
    error = function(e) {
      warning(
        "compress_brmsfit failed: ",
        conditionMessage(e),
        call. = FALSE,
        immediate. = TRUE
      )
      NULL
    }
  )
  time_used <- as.numeric(difftime(Sys.time(), compress_tic, units = "secs"))

  fid_blockwise <- if (is.null(compressed_blockwise)) {
    NULL
  } else {
    tryCatch(
      poco::evaluate_compression(
        compressed_blockwise$compressed,
        original_draws,
        seed = 1L
      ),
      error = function(e) {
        warning(
          "evaluate_compression failed: ",
          conditionMessage(e),
          call. = FALSE,
          immediate. = TRUE
        )
        NULL
      }
    )
  }

  if (is.null(compressed_blockwise)) {
    layout_tbl <- layout_tbl |>
      mutate(
        compression_ok = FALSE,
        time_used = NA_real_,
        compression_rate = NA_real_,
        energy_pct = NA_real_,
        c2st_pct = NA_real_,
        reproduction_pct = NA_real_
      )
  } else {
    estimates_chunk$brms_fit_compressed <- list(compressed_blockwise)
    layout_tbl <- layout_tbl |>
      mutate(
        compression_ok = TRUE,
        time_used = time_used,
        compression_rate =
          as.numeric(lobstr::obj_size(compressed_blockwise)) /
            as.numeric(lobstr::obj_size(fit_brms))
      )
  }

  if (is.null(fid_blockwise)) {
    layout_tbl <- layout_tbl |>
      mutate(
        energy_pct = NA_real_,
        c2st_pct = NA_real_,
        reproduction_pct = NA_real_
      )
  } else {
    en <- fid_blockwise$metrics$energy$reproduction_pct
    c2 <- fid_blockwise$metrics$c2st$reproduction_pct
    layout_tbl <- layout_tbl |>
      mutate(
        energy_pct = if (is.null(en)) NA_real_ else en,
        c2st_pct = if (is.null(c2)) NA_real_ else c2,
        reproduction_pct = fid_blockwise$reproduction_pct
      )
  }

  estimates_chunk |> tibble::add_column(layout_tbl)
}

sting_signature_21 |>
  dplyr::filter(.feature %in% model_meta$.feature) |>
  dplyr::pull(.feature) -> biomarkers

with_progress({
  p <- progressr::progressor(along = biomarkers)
  sting_estimates_chunk <- purrr::map_dfr(biomarkers, function(ensg) {
    p(sprintf("Processing %s", ensg))
    extract_one_biomarker(ensg)
  })
})

saveRDS(sting_estimates_chunk, params$estimates_rds)
```

------------------------------------------------------------------------

# Loaded estimates (preview)

```{r}
#| label: load-estimates
sting_estimates_chunk <- readRDS(params$estimates_rds)

sting_estimates_chunk |>
  dplyr::select(.feature, gene, compression_ok, compression_rate, reproduction_pct) |>
  head(10)
```

------------------------------------------------------------------------

# SAVI pseudobulk and healthy reference samples

**SAVI object:** RDS produced by `prepare_data.R` — pseudobulk counts aggregated with Seurat `AggregateExpression` by **`Sample`** and **`CellType`** (see GEO object metadata for treatment and category).

**Reference:** `pseudobulk_sample` from the same monocytic `targets` store is a `SummarizedExperiment` whose count assay is backed by the shared **Zenodo `pseudobulk_se.h5ad`** path. Rows with **`offset == 0`** are interpreted as healthy baseline samples for merge and posterior comparison (analogous to the breast epithelial reference in `pam50_toy_model.qmd`).

**Merge (same idea as `pam50_toy_model.qmd` § merge-se):** `SAVI_pseudobulk_sobj` is a **Seurat** object whose **RNA assay counts are row-named by gene symbol**. We map symbols → **Ensembl** (deduplicated), build a `SummarizedExperiment` with **`sample_role = "savi_pseudobulk"`**, then **column-bind** with reference columns after intersecting **shared Ensembl ids** and **assay names**.

```{r}
#| label: merge-savi-reference


SAVI_pseudobulk_sobj <- 
  readRDS(params$savi_pseudobulk_rds) %>% 
  filter(CellType == '17. Disease-associated monocytes')

pseudobulk_sample <- targets::tar_read(
  pseudobulk_sample,
  store = file.path(model_path, "_targets")
)

# Point DelayedArray/HDF5 backend to the local h5ad-backed matrix (path may differ on HPC)
pseudobulk_sample@assays@data@listData[["counts"]]@seed@seed@seed@filepath <-
  params$pseudobulk_h5ad

ref_sample <-
  pseudobulk_sample |>
  dplyr::filter(offset == 0)

SummarizedExperiment::colData(ref_sample)$sample_role <- "reference"

# --- SAVI: Seurat counts are rownames = gene symbols (see prepare_data.R) ---
counts_sym <- as.matrix(
  tryCatch(
    SeuratObject::LayerData(SAVI_pseudobulk_sobj, assay = "RNA", layer = "counts"),
    error = function(e) {
      Seurat::GetAssayData(SAVI_pseudobulk_sobj, assay = "RNA", slot = "counts")
    }
  )
)

sym <- rownames(counts_sym)
ensg <- AnnotationDbi::mapIds(
  org.Hs.eg.db::org.Hs.eg.db,
  keys = sym,
  column = "ENSEMBL",
  keytype = "SYMBOL",
  multiVals = "first"
)
keep <- !is.na(ensg) & !duplicated(ensg)
ensg <- ensg[keep]

counts_ens <- counts_sym[keep, , drop = FALSE]
rownames(counts_ens) <- ensg

cd_savi <- as.data.frame(SAVI_pseudobulk_sobj@meta.data)
rownames(cd_savi) <- colnames(counts_ens)
cd_savi$sample_role <- "savi_pseudobulk"
cd_savi$offset <- 0L
if (!"cell_type_unified_ensemble" %in% names(cd_savi)) {
  cd_savi$cell_type_unified_ensemble <- cd_savi$CellType
}

se_savi_ensembl <- SummarizedExperiment::SummarizedExperiment(
  assays = list(counts = counts_ens),
  colData = S4Vectors::DataFrame(cd_savi, row.names = colnames(counts_ens)),
  rowData = S4Vectors::DataFrame(
    gene_symbol = sym[keep],
    row.names = ensg
  )
)

# --- rbind-like merge of SE columns (intersect genes, union colData columns) ---
merge_summarized_experiments <- function(...) {
  ses <- list(...)
  assay_names <- Reduce(intersect, lapply(ses, SummarizedExperiment::assayNames))
  if (length(assay_names) == 0L) {
    stop("No shared assays between SummarizedExperiment objects.")
  }

  genes <- Reduce(intersect, lapply(ses, rownames))
  if (length(genes) == 0L) {
    stop("No shared rownames between SummarizedExperiment objects.")
  }

  ses <- lapply(ses, function(se) se[genes, , drop = FALSE])

  cds <- lapply(ses, function(se) as.data.frame(SummarizedExperiment::colData(se)))
  all_nms <- Reduce(union, lapply(cds, names))
  cds <- lapply(cds, function(cd) {
    missing <- setdiff(all_nms, names(cd))
    for (nm in missing) cd[[nm]] <- NA
    cd[, all_nms, drop = FALSE]
  })

  # Dense coercion: SAVI counts are an ordinary matrix; reference counts are often
  # HDF5Array/DelayedArray. Mixed cbind() hits bindCOLS() with no suitable method.
  assays <- stats::setNames(
    lapply(assay_names, function(an) {
      mats <- lapply(ses, function(se) {
        base::as.matrix(SummarizedExperiment::assay(se, an))
      })
      do.call(base::cbind, mats)
    }),
    assay_names
  )

  SummarizedExperiment::SummarizedExperiment(
    assays = assays,
    colData = S4Vectors::DataFrame(
      do.call(rbind, cds),
      row.names = unlist(lapply(ses, colnames), use.names = FALSE)
    ),
    rowData = SummarizedExperiment::rowData(ses[[1]])
  )
}

se_monocytic <- merge_summarized_experiments(
  se_savi_ensembl,
  ref_sample
)

# Copy abundance flag from reference rowData when a single .abundant* column exists
rd_ref <- as.data.frame(SummarizedExperiment::rowData(ref_sample))
abundant_cols <- grep("^\\.abundant", names(rd_ref), value = TRUE)
if (length(abundant_cols) >= 1L) {
  ac <- abundant_cols[[1]]
  SummarizedExperiment::rowData(se_monocytic)$.abundant <-
    SummarizedExperiment::rowData(ref_sample)[rownames(se_monocytic), ac, drop = TRUE]
}

tibble::tibble(
  n_features = nrow(se_monocytic),
  n_samples = ncol(se_monocytic),
  n_savi = sum(SummarizedExperiment::colData(se_monocytic)$sample_role == "savi_pseudobulk"),
  n_reference = sum(SummarizedExperiment::colData(se_monocytic)$sample_role == "reference")
)
```

```{r}
#| label: scale-abundance
ref_sample_name <-
  se_monocytic %>%
  colData() %>%
  as.data.frame() %>%
  filter(sample_role == "reference") %>%
  rownames()

se_monocytic <-
  se_monocytic %>%
  tidybulk::scale_abundance(
    method = "TMMwsp",
    reference_sample = ref_sample_name
  ) %>%
  mutate(offset = log(1 / multiplier))
```

------------------------------------------------------------------------

# SAVI sample annotations (GEO [GSE226598](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE226598))

Per-sample metadata (age, sex, 10x library batch, GSM accession) is taken from the GEO **series matrix** for GSE226598 (same fields as the GEO web “Sample characteristics” table). Each GSM links to its own GEO sample page.

After `scale_abundance`, we add harmonised columns for downstream models:

- **`tissue_groups`**: `"blood"` for SAVI pseudobulk columns only (PBMC in the study).
- **`sex`**, **`age_years`**: GEO values; **C1**, **C2** (“Adult”) and **P5_STING** (“Ped”) get **imputed numeric years** (see below) so `age_decade` / models can run without `NA`.
- **`age_decade`**: `as.character(ceiling(age_years / 10))` for SAVI once **`age_years`** is set (including imputed).
- **`age_group`**: `"pediatric"` if `age_years < 18`, else `"adult"` (SAVI columns only).
- **`age_decade_model`**: same as **`age_decade`** here (decade index from imputed / observed years).
- **`age_imputed`**: `TRUE` when **`age_years`** was filled by the rule below, not present as a number in GEO.

### Interpreting GEO **“Adult”** (C1, C2) and **“Ped”** (P5)

The cohort in de Cevins *et al.* (2023) comprises **five SAVI patients** and **seven healthy controls** (Figure 1A; GEO [GSE226598](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE226598)). GEO lists **“age: Adult”** for **C1** / **C2** and **“age: Ped”** for **P5_STING** without a year.

**Imputation used in this notebook** (rounded integers, documented for reproducibility):

| Sample(s) | GEO label | Imputed `age_years` | Rationale |
|-------------|-----------|---------------------|-----------|
| **C1**, **C2** | Adult | **35** | Mean of adult controls with numeric ages in the same series (**C8** 36 y, **C9** 35 y), aligned with matched adult controls. |
| **P5_STING** | Ped | **8** | Central value among other paediatric SAVI / control ages in GEO for this series (e.g. P1 4 y, P4 7–9 y, P6 8–10 y). |

All other donors keep **GEO-reported** ages. **`age_imputed`** is `TRUE` only for **C1**, **C2**, and **P5_STING**. Replace these constants if you obtain exact ages from authors or supplements.

- **`sex`**, **`assay_groups`**, **`assay_groups___altered`**: after GEO join, values are **harmonised to the Immune Atlas / posteriorHCA vocabulary** (same as `posteriorHCA::expr_predict()` / `Introduction` vignette: **`male`**, **`female`**, **`unknown`**; assay **`10x Genomics 3`**). GEO still uses **M**/**F** and Chromium **3′ v2**/**v3** batches in the series matrix; we do not keep the long Chromium strings in `colData` because the reference model was trained on short **`assay_groups`** labels.
- **`geo_accession`**, **`geo_sample_url`**: GSM id and direct GEO sample link (same pattern as the series page [GSE226598](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE226598)).

```{r}
#| label: coldata-gse226598
# Tibble from GEO series matrix GSE226598_series_matrix.txt.gz (Sample_title,
# Sample_geo_accession, age / Sex / batch rows). Series:
# https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE226598
geo_tbl <- tibble::tribble(
  ~Sample,              ~gse_accession, ~gse_age_years, ~gse_sex, ~gse_sting_batch,
  "C10",                "GSM7079979",            10,  "M",  "STING2",
  "C11",                "GSM7079980",            16,  "M",  "STING2",
  "C1",                 "GSM7079981",            NA,  "M",  "STING3",
  "C2",                 "GSM7079982",            NA,  "M",  "STING3",
  "C7",                 "GSM7079983",             5,  "F",  "STING1",
  "C8_aunt_P1_STING",   "GSM7079984",            36,  "F",  "STING1",
  "C9_mother_P1_STING", "GSM7079985",            35,  "F",  "STING1",
  "P1_STING_ht",        "GSM7079986",             4,  "F",  "STING1",
  "P1_STING_ht_T",      "GSM7079987",             8,  "F",  "STING1",
  "P1_STING_ht_T2",     "GSM7079988",             8,  "F",  "STING1",
  "P2_STING_ht",        "GSM7079989",            31,  "M",  "STING1",
  "P2_STING_ht_T",      "GSM7079990",            34,  "M",  "STING1",
  "P4_STING_ht",        "GSM7079991",             7,  "M",  "STING2",
  "P4_STING_ht_T",      "GSM7079992",             9,  "M",  "STING2",
  "P5_STING",           "GSM7079993",            NA,  "M",  "STING3",
  "P6_STING_ht",        "GSM7079994",             8,  "M",  "STING2",
  "P6_STING_ht_T",      "GSM7079995",            10,  "M",  "STING2"
) |>
  dplyr::mutate(
    # Raw GEO had NA for "Adult" (C1, C2) and "Ped" (P5); impute for modelling (see prose).
    gse_age_imputed = is.na(.data$gse_age_years),
    gse_age_years = dplyr::coalesce(
      .data$gse_age_years,
      dplyr::case_when(
        .data$Sample %in% c("C1", "C2") ~ 35,
        .data$Sample == "P5_STING" ~ 8,
        TRUE ~ NA_real_
      )
    )
  ) |>
  dplyr::mutate(
    gse_geo_sample_url = sprintf(
      "https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=%s",
      .data$gse_accession
    )
  ) |>
  dplyr::select(
    "Sample",
    "gse_accession",
    "gse_geo_sample_url",
    "gse_age_years",
    "gse_age_imputed",
    "gse_sex"
  )

cd <- SummarizedExperiment::colData(se_monocytic) |> as.data.frame()
rn <- rownames(cd)

if (!"Sample" %in% names(cd)) {
  stop("colData must contain `Sample` (from SAVI pseudobulk) to join GEO metadata.")
}

if (!"tissue_groups" %in% names(cd)) {
  cd$tissue_groups <- NA_character_
}
if (!"age_years" %in% names(cd)) {
  cd$age_years <- NA_real_
}
if (!"sex" %in% names(cd)) {
  cd$sex <- NA_character_
}
if (!"assay_groups" %in% names(cd)) {
  cd$assay_groups <- NA_character_
}
if (!"assay_groups___altered" %in% names(cd)) {
  cd$assay_groups___altered <- NA_character_
}

cd <- cd |>
  dplyr::left_join(geo_tbl, by = "Sample")

miss_geo <- is.na(cd$gse_accession) & cd$sample_role == "savi_pseudobulk"
if (any(miss_geo, na.rm = TRUE)) {
  warning(
    "SAVI columns with no GEO match (check `Sample` vs GEO Sample_title): ",
    paste(unique(cd$Sample[miss_geo]), collapse = ", ")
  )
}

sex_ref <- if ("Sex" %in% names(cd)) cd$Sex else rep(NA_character_, nrow(cd))

cd <- cd |>
  dplyr::mutate(
    tissue_groups = dplyr::if_else(
      .data$sample_role == "savi_pseudobulk",
      "blood",
      .data$tissue_groups
    ),
    sex = dplyr::case_when(
      .data$sample_role == "savi_pseudobulk" ~ .data$gse_sex,
      TRUE ~ dplyr::coalesce(.data$sex, sex_ref)
    ),
    age_years = dplyr::case_when(
      .data$sample_role == "savi_pseudobulk" ~ .data$gse_age_years,
      TRUE ~ .data$age_years
    ),
    age_decade = dplyr::if_else(
      is.na(.data$age_years),
      NA_character_,
      as.character(ceiling(.data$age_years / 10))
    ),
    age_group = dplyr::case_when(
      .data$sample_role != "savi_pseudobulk" ~ NA_character_,
      !is.na(.data$age_years) & .data$age_years < 18 ~ "pediatric",
      !is.na(.data$age_years) & .data$age_years >= 18 ~ "adult",
      TRUE ~ NA_character_
    ),
    age_decade_model = .data$age_decade,
    age_imputed = dplyr::if_else(
      .data$sample_role == "savi_pseudobulk",
      .data$gse_age_imputed,
      NA
    ),
    assay_groups = dplyr::if_else(
      .data$sample_role == "savi_pseudobulk",
      "10x Genomics 3",
      .data$assay_groups
    ),
    assay_groups___altered = dplyr::case_when(
      .data$sample_role == "savi_pseudobulk" ~ "10x Genomics 3",
      !is.na(.data$assay_groups___altered) ~ .data$assay_groups___altered,
      !is.na(.data$assay_groups) ~ .data$assay_groups,
      TRUE ~ NA_character_
    ),
    geo_accession = dplyr::if_else(
      .data$sample_role == "savi_pseudobulk",
      .data$gse_accession,
      NA_character_
    ),
    geo_sample_url = dplyr::if_else(
      .data$sample_role == "savi_pseudobulk",
      .data$gse_geo_sample_url,
      NA_character_
    )
  ) |>
  dplyr::select(
    -dplyr::any_of(c(
      "gse_accession",
      "gse_geo_sample_url",
      "gse_age_years",
      "gse_age_imputed",
      "gse_sex",
      "Sex"
    ))
  ) |>
  dplyr::mutate(
    sex = dplyr::case_when(
      is.na(.data$sex) ~ NA_character_,
      .data$sex %in% c("male", "female", "unknown") ~ as.character(.data$sex),
      toupper(trimws(as.character(.data$sex))) %in% c("M", "MALE") ~ "male",
      toupper(trimws(as.character(.data$sex))) %in% c("F", "FEMALE") ~ "female",
      TRUE ~ "unknown"
    )
  )

SummarizedExperiment::colData(se_monocytic) <- S4Vectors::DataFrame(cd, row.names = rn)

cd |>
  dplyr::filter(.data$sample_role == "savi_pseudobulk") |>
  dplyr::distinct(
    .data$Sample,
    .data$tissue_groups,
    .data$sex,
    .data$age_years,
    .data$age_decade,
    .data$age_group,
    .data$age_decade_model,
    .data$age_imputed,
    .data$assay_groups,
    .data$assay_groups___altered,
    .data$geo_accession,
    .data$geo_sample_url
  ) |>
  dplyr::arrange(.data$Sample)
```

GEO sample pages use the same host as the series record above; each **`geo_sample_url`** points to the corresponding GSM (e.g. patient / control library) listed under [GSE226598](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE226598).

------------------------------------------------------------------------

# Compare samples to personalised healthy baseline

For each **STING signature gene** with a loadable compressed `brms` fit, we build
a donor-level healthy counterfactual from the sample's known covariates:
`disease_groups_altered = Normal`, observed `tissue_groups`, observed `offset`,
and available age/sex/assay metadata. Unknown categorical fixed effects are
marginalised over fitted atlas levels, and an unseen SAVI dataset level is drawn
hierarchically.

Following the HCC personalised baseline report, this section computes two
**separate versions** of the result, each reported and discussed in its own
section below:

- **Version 1 — predictive count test:** `posterior_predict()` gives replicated
  healthy counts including model uncertainty, zero-inflation, NB noise, and the
  unseen dataset effect. The statistic is a **two-tailed posterior predictive
  p-value** on the observed count.
- **Version 2 — linear-predictor test:** `posterior_linpred(transform = FALSE)`
  gives the NB count-component location. We subtract the sample offset to compare
  on the shared **offset-zero** log-μ scale, then run a **z-test** between the
  patient log-μ and the matched healthy **grand-mean** log-μ draws. The patient
  side is a single count, so its log-scale SE uses the full delta-method variance
  `sqrt(1 / (y + 0.5) + 1 / shape)`: the `+ 0.5` keeps `y = 0` finite (zeros are
  tested, not dropped) and the overdispersion floor `1 / shape` reflects that one
  draw cannot pin the sample mean more tightly than `sqrt(1 / shape)`. With one
  observation this is a Gaussian companion to Version 1, not an independent,
  more-powerful test.

`posterior_epred()` is retained as an expected-count effect-size summary.
Tail p-values, Welch p-values, and BH adjustments are exploratory because the
tests span many genes and donors.

We use **`re_formula = NULL`** with **`allow_new_levels = TRUE`** and
**`sample_new_levels = "gaussian"`** so that:

- `(1 + age_decade * sex + ethnicity_groups | tissue_groups)` keeps the trained **`blood`** slopes (tissue is observed).
- `(1 | dataset_id_altered)` is marginalised: a fresh dataset intercept is drawn from `N(0, sd_dataset_id_altered)` for every posterior sample — the correct "new SAVI cohort, exchangeable with the atlas" null.
- Unknown **`ethnicity_groups`** (and other NA covariates) are marginalised by expanding `newdata` over training-set levels. For the **predictive count** null we **pool** the full finite mixture (every grid profile × every posterior draw, equal weight, `marginalize_method = "pool"`), so single-donor covariate uncertainty stays in the count distribution; setting ethnicity to a single arbitrary value would bias the null toward that group. The **grand-mean log-μ** reference instead **averages** the grid per draw (`rowMeans`, `method = "mean"`) on purpose, to isolate uncertainty in the healthy *mean* rather than the full covariate spread.

```{r}
#| label: gene-deviation-fn
#| cache: true

# Categorical levels for any predictor pulled from the brms fit's training
# frame. Used to expand newdata over training-set values when a covariate is
# unknown for a test sample.
fit_levels <- function(fit, var) {
  d <- fit$data
  if (is.null(d) || !var %in% names(d)) return(character(0))
  v <- d[[var]]
  lv <- if (is.factor(v)) levels(v) else sort(unique(as.character(v)))
  lv[!is.na(lv) & nzchar(lv)]
}

# First non-NA, non-empty character among the inputs; NA_character_ otherwise.
coalesce_char <- function(...) {
  args <- list(...)
  for (v in args) {
    if (length(v) >= 1L) {
      val <- as.character(v[[1L]])
      if (!is.na(val) && nzchar(val)) return(val)
    }
  }
  NA_character_
}

# Build a single-row newdata template from colData. Columns that may be NA are
# left NA on purpose; build_newdata_grid() will marginalise over training-set
# levels when the model uses them.
newdata_for_sample <- function(se, sample_idx) {
  cd <- SummarizedExperiment::colData(se)[sample_idx, , drop = FALSE] |>
    as.data.frame()

  tibble::tibble(
    # Observed when available (may be NA -> marginalise).
    age_decade = coalesce_char(
      if ("age_decade_model" %in% names(cd)) cd$age_decade_model,
      if ("age_decade" %in% names(cd)) cd$age_decade
    ),
    sex = coalesce_char(if ("sex" %in% names(cd)) cd$sex),
    ethnicity_groups = coalesce_char(
      if ("ethnicity_groups" %in% names(cd)) cd$ethnicity_groups,
      if ("ethnicity_groups_imputed" %in% names(cd)) cd$ethnicity_groups_imputed
    ),

    # Counterfactual / required: forced to the healthy baseline so we test
    # against the atlas null, not the sample's own disease label.
    disease_groups_altered = "Normal",
    # disease_groups_altered = coalesce_char(
    #   if ("disease_groups_altered" %in% names(cd)) cd$disease_groups_altered,
    #   if ("disease_groups___altered" %in% names(cd)) cd$disease_groups___altered,
    #   "Normal"
    # ),


    # Required (atlas-defined / technical):
    assay_groups_altered = coalesce_char(
      if ("assay_groups_altered" %in% names(cd)) cd$assay_groups_altered,
      if ("assay_groups___altered" %in% names(cd)) cd$assay_groups___altered,
      if ("assay_groups" %in% names(cd)) cd$assay_groups,
      "10x Genomics 3"
    ),
    tissue_groups = coalesce_char(if ("tissue_groups" %in% names(cd)) cd$tissue_groups),
    offset = cd$offset[[1]]
  )
}

# Expand a single-row newdata into a prediction grid:
#   - required cols MUST be non-NA (e.g. tissue_groups)
#   - marginalize_over cols that are NA in newdata AND present in fit$data
#     are expanded across the training-set levels.
#   - new_level_cols (e.g. dataset_id_altered) are stamped with a fresh unique
#     placeholder so brms (allow_new_levels = TRUE, sample_new_levels =
#     "gaussian") draws a new group effect from its posterior population.
build_newdata_grid <- function(
  fit,
  newdata,
  marginalize_over = c("sex", "age_decade", "ethnicity_groups", "assay_groups_altered"),
  required = c("tissue_groups", "offset", "disease_groups_altered"),
  new_level_cols = "dataset_id_altered",
  new_level_tag = NULL
) {
  stopifnot(NROW(newdata) == 1L)
  newdata <- tibble::as_tibble(newdata)

  is_blank <- function(v) {
    length(v) == 0L ||
      is.na(v[[1L]]) ||
      (is.character(v[[1L]]) && !nzchar(v[[1L]]))
  }

  fit_vars <- colnames(fit$data)

  # Required columns must be present and non-empty.
  for (col in required) {
    if (!col %in% fit_vars) next
    if (!col %in% names(newdata) || is_blank(newdata[[col]])) {
      stop(sprintf(
        "Required predictor `%s` is missing or empty in newdata.",
        col
      ))
    }
  }

  # Decide which columns to expand.
  to_expand <- character(0)
  for (col in marginalize_over) {
    if (!col %in% fit_vars) next
    if (!col %in% names(newdata) || is_blank(newdata[[col]])) {
      to_expand <- c(to_expand, col)
    }
  }

  if (length(to_expand)) {
    levels_list <- lapply(to_expand, function(col) fit_levels(fit, col))
    names(levels_list) <- to_expand
    keep <- vapply(levels_list, length, integer(1L)) > 0L
    levels_list <- levels_list[keep]
    to_expand <- names(levels_list)
  }

  if (length(to_expand)) {
    grid <- do.call(tidyr::expand_grid, levels_list)
    rest <- newdata[, setdiff(names(newdata), to_expand), drop = FALSE]
    out <- dplyr::bind_cols(
      rest[rep(1L, nrow(grid)), , drop = FALSE],
      grid
    )
  } else {
    out <- newdata
  }

  if (is.null(new_level_tag)) {
    new_level_tag <- sprintf(
      "__new__%d_%d",
      as.integer(Sys.time()),
      sample.int(1e6, 1L)
    )
  }
  for (col in new_level_cols) {
    if (col %in% fit_vars) out[[col]] <- new_level_tag
  }

  tibble::as_tibble(out)
}

# Collapse the covariate grid (columns) of a draws-by-grid matrix to a marginal
# sample. "pool" keeps every grid profile x every draw (the full finite mixture,
# used for the predictive count so single-donor covariate uncertainty is kept);
# "mean" averages the grid per draw (the grand-mean parameter); "sample" thins to
# one grid profile per draw.
marginalize_draw_matrix <- function(draw_mat, method = c("sample", "mean", "pool")) {
  method <- match.arg(method)
  if (is.null(dim(draw_mat))) {
    return(as.numeric(draw_mat))
  }
  if (ncol(draw_mat) <= 1L) {
    return(as.numeric(draw_mat))
  }

  switch(
    method,
    sample = {
      idx <- sample.int(ncol(draw_mat), nrow(draw_mat), replace = TRUE)
      draw_mat[cbind(seq_len(nrow(draw_mat)), idx)]
    },
    mean = rowMeans(draw_mat),
    pool = as.numeric(draw_mat)
  )
}

# Helper used by the later predictive density audit sections. For a predictive
# DENSITY we pool the full finite mixture (every posterior draw x every covariate
# grid profile, equal weight) rather than thinning to one profile per draw. Both
# target the same marginal predictive distribution, but pooling keeps all the mass
# and gives a smoother density to plot.
marginalize_predict_draws <- function(y_rep_mat) {
  marginalize_draw_matrix(y_rep_mat, method = "pool")
}

# Welch t-test from two (mean, SE, n) summaries on the log(mu) scale.
# Same definition as SAVI_case_study_universal_baseline.qmd for cross-report
# consistency. The pooled SE combines the two standard errors, and the degrees
# of freedom follow the Welch-Satterthwaite approximation.
# NOTE: Version 2 below no longer uses this helper. With a single patient
# observation the patient side is a fixed point estimate with a known
# delta-method SE, so the Welch-Satterthwaite df (driven by n1 = 1) collapses to
# ~1 and is meaningless; Version 2 uses an explicit z-test instead. This is kept
# only for parity with the universal-baseline report.
welch_t_from_mean_se <- function(mean1, se1, n1, mean2, se2, n2) {
  n1 <- max(n1, 2)
  n2 <- max(n2, 2)
  sd1 <- se1 * sqrt(n1)
  sd2 <- se2 * sqrt(n2)
  se_diff <- sqrt(se1^2 + se2^2)
  t_stat <- (mean1 - mean2) / se_diff
  df <- (sd1^2 / n1 + sd2^2 / n2)^2 /
    ((sd1^2 / n1)^2 / (n1 - 1) + (sd2^2 / n2)^2 / (n2 - 1))
  p_value <- 2 * stats::pt(abs(t_stat), df = df, lower.tail = FALSE)
  list(
    mean_diff = mean1 - mean2,
    se_diff = se_diff,
    t = t_stat,
    df = df,
    p_value = p_value
  )
}

# Personalised comparison of one observed count against a brms fit.
#   - posterior_predict: calibrated new-observation predictive tails.
#   - posterior_epred: expected-count effect-size summaries.
#   - posterior_linpred(transform = FALSE): offset-centred log(mu) test.
#   - re_formula = NULL: keep tissue_groups random slopes (observed level).
#   - allow_new_levels + sample_new_levels = "gaussian": dataset_id_altered
#     gets a new draw from its random-intercept population per posterior draw.
#   - marginalize_over: unknown covariates are integrated by a finite mixture
#     over the training-set grid.
gene_deviation_from_fit <- function(
  fit,
  newdata,
  obs,
  cred_mass = 0.95,
  alpha = 0.025,
  marginalize_over = c("sex", "age_decade", "ethnicity_groups", "assay_groups_altered"),
  required = c("tissue_groups", "offset", "disease_groups_altered"),
  new_level_cols = "dataset_id_altered",
  new_level_tag = NULL,
  marginalize_method = "pool",
  re_formula = NULL,
  allow_new_levels = TRUE,
  sample_new_levels = "gaussian",
  seed = NULL
) {
  nd_grid <- build_newdata_grid(
    fit,
    newdata,
    marginalize_over = marginalize_over,
    required = required,
    new_level_cols = new_level_cols,
    new_level_tag = new_level_tag
  )

  if (!is.null(seed)) set.seed(seed)
  predictive_mat <- brms::posterior_predict(
    fit,
    newdata = nd_grid,
    summary = FALSE,
    re_formula = re_formula,
    allow_new_levels = allow_new_levels,
    sample_new_levels = sample_new_levels
  )

  if (!is.null(seed)) set.seed(seed + 1L)
  expected_mat <- brms::posterior_epred(
    fit,
    newdata = nd_grid,
    summary = FALSE,
    re_formula = re_formula,
    allow_new_levels = allow_new_levels,
    sample_new_levels = sample_new_levels
  )

  if (!is.null(seed)) set.seed(seed + 2L)
  linpred_mat <- brms::posterior_linpred(
    fit,
    newdata = nd_grid,
    summary = FALSE,
    transform = FALSE,
    re_formula = re_formula,
    allow_new_levels = allow_new_levels,
    sample_new_levels = sample_new_levels
  )

  if (!is.null(seed)) set.seed(seed + 3L)
  shape_mat <- tryCatch(
    brms::posterior_linpred(
      fit,
      newdata = nd_grid,
      summary = FALSE,
      dpar = "shape",
      transform = TRUE,
      re_formula = re_formula,
      allow_new_levels = allow_new_levels,
      sample_new_levels = sample_new_levels
    ),
    error = function(e) matrix(NA_real_, nrow = nrow(linpred_mat), ncol = ncol(linpred_mat))
  )

  # Predictive count and expected count: POOL the full finite mixture
  # (marginalize_method = "pool" -> every grid profile x every draw, equal
  # weight) so the count-scale null carries covariate-marginalisation +
  # observation uncertainty.
  if (!is.null(seed)) set.seed(seed + 4L)
  y_rep <- marginalize_draw_matrix(predictive_mat, method = marginalize_method)
  if (!is.null(seed)) set.seed(seed + 4L)
  epred <- marginalize_draw_matrix(expected_mat, method = marginalize_method)

  # Centre parameter (log mu): collapse the covariate grid with the GRAND MEAN
  # per draw (rowMeans), NOT the sampled mixture. This yields the posterior of
  # the healthy grand mean -- the tight null used in mu_test_report.qmd. Using
  # the sampled mixture here would fold across-grid covariate spread into the SD
  # and make the mean-parameter Welch test artificially insensitive.
  linpred_grand <- marginalize_draw_matrix(linpred_mat, method = "mean")
  shape <- marginalize_draw_matrix(shape_mat, method = "mean")

  hpd <- if (requireNamespace("HDInterval", quietly = TRUE)) {
    HDInterval::hdi(y_rep, credMass = cred_mass)
  } else {
    probs <- c((1 - cred_mass) / 2, 1 - (1 - cred_mass) / 2)
    stats::quantile(y_rep, probs = probs, names = FALSE)
  }

  epred_interval <- stats::quantile(epred, probs = c(0.025, 0.975), names = FALSE)
  offset_value <- newdata$offset[[1]]
  healthy_log_mu <- linpred_grand - offset_value
  healthy_log_mu_mean <- mean(healthy_log_mu, na.rm = TRUE)
  healthy_log_mu_sd <- stats::sd(healthy_log_mu, na.rm = TRUE)
  healthy_log_mu_interval <- stats::quantile(
    healthy_log_mu,
    probs = c(0.025, 0.975),
    names = FALSE,
    na.rm = TRUE
  )
  healthy_shape_median <- stats::median(shape, na.rm = TRUE)

  predictive_p_upper <- (1 + sum(y_rep >= obs, na.rm = TRUE)) / (length(y_rep) + 1)
  predictive_p_lower <- (1 + sum(y_rep <= obs, na.rm = TRUE)) / (length(y_rep) + 1)
  predictive_percentile <- (
    0.5 +
      sum(y_rep < obs, na.rm = TRUE) +
      0.5 * sum(y_rep == obs, na.rm = TRUE)
  ) / (length(y_rep) + 1)
  predictive_tail_p <- pmin(1, 2 * pmin(predictive_p_upper, predictive_p_lower))
  predictive_z <- stats::qnorm(
    pmin(
      1 - 0.5 / (length(y_rep) + 1),
      pmax(0.5 / (length(y_rep) + 1), predictive_percentile)
    )
  )

  if (is.finite(obs)) {
    # Single-observation log(mu) test (honest version).
    #
    # Continuity correction obs + 0.5 keeps log() finite at obs == 0 (so zero
    # counts -- the most "lower-than-healthy" signal -- are tested instead of
    # dropped) and matches the +0.5 convention used elsewhere in this report.
    obs_c <- obs + 0.5
    patient_log_mu <- log(obs_c) - offset_value
    # Delta-method SE of a SINGLE NB observation on the log scale:
    #   Var(log Y) ~= Var(Y) / mu^2 = (mu + mu^2 / shape) / mu^2 = 1/mu + 1/shape.
    # The Poisson/aggregation part is 1 / obs_c; the overdispersion floor
    # 1 / shape is the irreducible relative noise of a single draw and MUST be
    # included. A single count can never pin this sample's mean rate tighter than
    # ~sqrt(1 / shape), no matter how large obs is. (The earlier version dropped
    # 1 / shape -- treating one overdispersed draw as a Poisson-precise estimate
    # of the mean -- which was anti-conservative and made Version 2 spuriously
    # "more sensitive" than the predictive test. With one observation the
    # mean-parameter test cannot legitimately beat Version 1; including 1/shape
    # is what makes it agree.)
    overdisp_term <- if (is.finite(healthy_shape_median) && healthy_shape_median > 0) {
      1 / healthy_shape_median
    } else {
      0
    }
    patient_log_mu_se <- sqrt(1 / obs_c + overdisp_term)
    # The patient side is a fixed point estimate with a known (asymptotic
    # delta-method) SE, so we use a z-test against the healthy grand-mean log(mu)
    # posterior (mean healthy_log_mu_mean, SD healthy_log_mu_sd). welch_df is kept
    # as the healthy-draw count for column continuity, but it is large enough that
    # the t- and z-references coincide.
    delta_log_mu_vs_healthy <- patient_log_mu - healthy_log_mu_mean
    delta_log_mu_se <- sqrt(patient_log_mu_se^2 + healthy_log_mu_sd^2)
    welch_t <- delta_log_mu_vs_healthy / delta_log_mu_se
    welch_df <- max(length(healthy_log_mu) - 1L, 1L)
    welch_p <- 2 * stats::pt(abs(welch_t), df = welch_df, lower.tail = FALSE)
  } else {
    patient_log_mu <- NA_real_
    patient_log_mu_se <- NA_real_
    delta_log_mu_vs_healthy <- NA_real_
    delta_log_mu_se <- NA_real_
    welch_t <- NA_real_
    welch_df <- NA_real_
    welch_p <- NA_real_
  }

  tibble::tibble(
    observed_count = obs,
    n_draws = length(y_rep),
    healthy_median = median(y_rep),
    healthy_mean = mean(y_rep),
    healthy_hpd_lo = hpd[1],
    healthy_hpd_hi = hpd[2],
    healthy_predictive_median = median(y_rep),
    healthy_predictive_lo = hpd[1],
    healthy_predictive_hi = hpd[2],
    healthy_epred_median = median(epred),
    healthy_epred_lo = epred_interval[1],
    healthy_epred_hi = epred_interval[2],
    obs_percentile = predictive_percentile,
    predictive_percentile = predictive_percentile,
    p_above_healthy = predictive_p_upper,
    p_below_healthy = predictive_p_lower,
    predictive_p_upper = predictive_p_upper,
    predictive_p_lower = predictive_p_lower,
    posterior_tail_p = predictive_tail_p,
    predictive_tail_p = predictive_tail_p,
    predictive_z = predictive_z,
    absolute_predictive_z = abs(predictive_z),
    log2fc_vs_median = log2((obs + 1) / (median(y_rep) + 1)),
    log2fc_vs_healthy_epred = log2((obs + 0.5) / (median(epred) + 0.5)),
    patient_log_mu = patient_log_mu,
    patient_log_mu_se = patient_log_mu_se,
    healthy_log_mu_mean = healthy_log_mu_mean,
    healthy_log_mu_sd = healthy_log_mu_sd,
    healthy_log_mu_lo = healthy_log_mu_interval[1],
    healthy_log_mu_hi = healthy_log_mu_interval[2],
    healthy_nb_shape_median = healthy_shape_median,
    delta_log_mu_vs_healthy = delta_log_mu_vs_healthy,
    delta_log_mu_se = delta_log_mu_se,
    welch_t = welch_t,
    welch_df = welch_df,
    welch_p = welch_p,
    n_grid_rows = NROW(nd_grid),
    marginalize_method = marginalize_method,
    direction = dplyr::case_when(
      predictive_p_upper < alpha ~ "higher_than_healthy",
      predictive_p_lower < alpha ~ "lower_than_healthy",
      TRUE ~ "consistent_with_healthy"
    ),
    linpred_direction = dplyr::case_when(
      welch_p < 0.05 & delta_log_mu_vs_healthy > 0 ~ "higher_than_healthy",
      welch_p < 0.05 & delta_log_mu_vs_healthy < 0 ~ "lower_than_healthy",
      TRUE ~ "consistent_with_healthy"
    )
  )
}
```

```{r}
#| label: comparison-samples
#| cache: true
cd_mono <-
  se_monocytic |>
  SummarizedExperiment::colData() |>
  as.data.frame() |>
  tibble::rownames_to_column("sample")

if (!length(ref_sample_name)) {
  stop("No reference columns (offset == 0) found; cannot define comparison baseline.")
}

comparison_samples_tbl <- cd_mono |>
  dplyr::filter(.data$sample_role == "savi_pseudobulk") |>
  dplyr::bind_rows(
    cd_mono |>
      dplyr::filter(.data$sample == !!ref_sample_name[[1]])
  ) |>
  dplyr::distinct(.data$sample, .keep_all = TRUE) |>
  dplyr::mutate(sample_idx = match(.data$sample, colnames(se_monocytic)))

tibble::tibble(
  n_genes_fit = sum(sting_estimates_chunk$compression_ok == TRUE, na.rm = TRUE),
  n_comparison_samples = nrow(comparison_samples_tbl),
  n_savi = sum(comparison_samples_tbl$sample_role == "savi_pseudobulk"),
  n_reference = sum(comparison_samples_tbl$sample_role == "reference")
)
```

```{r}
#| label: predict-sting-vs-healthy
#| cache: true
#| cache.lazy: false
#| dependson: ["gene-deviation-fn", "comparison-samples"]

handlers("progress")

sting_fit_rows <- sting_estimates_chunk |>
  dplyr::filter(.data$compression_ok == TRUE) |>
  dplyr::filter(lengths(.data$brms_fit_compressed) > 0L) |>
  dplyr::filter(.data$.feature %in% rownames(se_monocytic))

if (!nrow(sting_fit_rows)) {
  stop("No compressed biomarker fits available; run with params$run_fits: true or fix estimates_rds.")
}

with_progress({
  p <- progressr::progressor(steps = nrow(sting_fit_rows) * nrow(comparison_samples_tbl))

  gene_deviation_all <- purrr::imap_dfr(
    seq_len(nrow(sting_fit_rows)),
    function(i, idx) {
      chunk_row <- sting_fit_rows[i, ]
      feature_id <- chunk_row$.feature
      gene_sym <- chunk_row$gene
      fit <- poco::reconstruct_brmsfit(chunk_row$brms_fit_compressed[[1]])

      purrr::pmap_dfr(
        comparison_samples_tbl |>
          dplyr::select("sample", "sample_idx", "sample_role"),
        function(sample, sample_idx, sample_role) {
          p()

          newdata <- newdata_for_sample(se_monocytic, sample_idx)
          obs <-
            SummarizedExperiment::assay(
              se_monocytic[feature_id, sample_idx, drop = FALSE],
              "counts"
            ) |>
            as.numeric()

          gene_deviation_from_fit(
            fit,
            newdata,
            obs,
            new_level_tag = paste0("__savi__", as.character(sample)),
            seed = 20250625L + 1000L * idx + sample_idx
          ) |>
            dplyr::mutate(
              .feature = feature_id,
              gene = gene_sym,
              sample = as.character(sample),
              sample_role = as.character(sample_role),
              .before = "observed_count"
            )
        }
      )
    }
  )
})

gene_deviation_all <- gene_deviation_all |>
  tibble::as_tibble() |>
  dplyr::select(tidyselect::where(\(x) !is.list(x)))

if (!"sample_role" %in% colnames(gene_deviation_all)) {
  gene_deviation_all <- gene_deviation_all |>
    dplyr::left_join(
      comparison_samples_tbl |>
        dplyr::select("sample", "sample_role"),
      by = "sample"
    )
}

gene_deviation_all <- gene_deviation_all |>
  dplyr::mutate(
    q_value_global = stats::p.adjust(.data$predictive_tail_p, method = "BH"),
    welch_q_value_global = stats::p.adjust(.data$welch_p, method = "BH")
  ) |>
  dplyr::group_by(.data$sample) |>
  dplyr::mutate(
    q_value_within_sample = stats::p.adjust(.data$predictive_tail_p, method = "BH"),
    welch_q_value_within_sample = stats::p.adjust(.data$welch_p, method = "BH")
  ) |>
  dplyr::ungroup()

linpred_tests <- gene_deviation_all |>
  dplyr::filter(.data$sample_role == "savi_pseudobulk") |>
  dplyr::select(
    gene, .feature, sample, observed_count, patient_log_mu,
    patient_log_mu_se, healthy_log_mu_mean, healthy_log_mu_sd,
    healthy_log_mu_lo, healthy_log_mu_hi, delta_log_mu_vs_healthy,
    delta_log_mu_se, welch_t, welch_df, welch_p,
    welch_q_value_global, welch_q_value_within_sample, linpred_direction,
    n_draws, n_grid_rows, marginalize_method
  )

personalised_sample_scores <- gene_deviation_all |>
  dplyr::filter(.data$sample_role == "savi_pseudobulk") |>
  dplyr::left_join(
    cd_mono |>
      dplyr::select(
        "sample",
        tidyselect::any_of(c("Sample", "Category", "Treatment"))
      ) |>
      dplyr::distinct(),
    by = "sample"
  ) |>
  dplyr::group_by(.data$sample, .data$Sample, .data$Category) |>
  dplyr::summarise(
    n_genes = dplyr::n(),
    mean_absolute_predictive_z = mean(.data$absolute_predictive_z, na.rm = TRUE),
    mean_signed_predictive_z = mean(.data$predictive_z, na.rm = TRUE),
    mean_abs_log2fc_vs_epred = mean(abs(.data$log2fc_vs_healthy_epred), na.rm = TRUE),
    mean_absolute_welch_t = mean(abs(.data$welch_t), na.rm = TRUE),
    mean_signed_welch_t = mean(.data$welch_t, na.rm = TRUE),
    n_predictive_outside_95 = sum(.data$direction != "consistent_with_healthy"),
    n_linpred_welch_p_lt_0_05 = sum(.data$welch_p < 0.05, na.rm = TRUE),
    n_predictive_global_q_lt_0_05 = sum(.data$q_value_global < 0.05, na.rm = TRUE),
    n_welch_global_q_lt_0_05 = sum(.data$welch_q_value_global < 0.05, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::arrange(.data$Category, .data$Sample)

gene_deviation_all
```

```{r}
#| label: deviation-summary-sting
gene_deviation_all |>
  dplyr::count(.data$sample_role, .data$direction) |>
  tidyr::pivot_wider(
    names_from = "direction",
    values_from = "n",
    values_fill = 0L
  )

gene_deviation_savi <- gene_deviation_all |>
  dplyr::filter(.data$sample_role == "savi_pseudobulk")

gene_deviation_savi |>
  dplyr::group_by(.data$gene) |>
  dplyr::summarise(
    n_higher = sum(.data$direction == "higher_than_healthy"),
    n_lower = sum(.data$direction == "lower_than_healthy"),
    n_consistent = sum(.data$direction == "consistent_with_healthy"),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(.data$n_higher + .data$n_lower))
```

The two statistical layers below are reported as **separate versions** because
they answer different questions and use different reference distributions.

------------------------------------------------------------------------

# Version 1 — Predictive count test (two-tailed)

**Primary test.** For each STING gene × SAVI sample we compare the observed
pseudobulk count to the donor-matched healthy `posterior_predict()`
distribution. Evidence is a **two-tailed posterior predictive p-value**
(`predictive_tail_p`) with a finite-draw correction. A small value means the
observed count is hard to generate from the personalised healthy counterfactual
after propagating model, covariate-marginalisation, zero-inflation, NB sampling,
and new-dataset uncertainty. `posterior_epred()` supplies an expected-count
effect size (`log2fc_vs_healthy_epred`). BH q-values are exploratory.

```{r}
#| label: personalised-count-test-preview

gene_deviation_savi |>
  dplyr::select(
    gene, sample, observed_count, healthy_predictive_median,
    healthy_predictive_lo, healthy_predictive_hi, healthy_epred_median,
    log2fc_vs_healthy_epred, predictive_percentile, predictive_tail_p,
    q_value_global, direction
  ) |>
  dplyr::mutate(
    dplyr::across(dplyr::where(is.numeric), ~round(.x, 4))
  ) |>
  dplyr::arrange(.data$sample, .data$gene)
```

```{r}
#| label: plot-version1-predictive
#| fig-width: 8
#| fig-height: 4.5
#| fig-align: center

category_cols <- c(
  CTRL = "#2ca25f",
  SAVI = "#de2d26",
  SAVI_treated = "#756bb1"
)

version1_df <- personalised_sample_scores |>
  dplyr::mutate(
    Category = factor(.data$Category, levels = names(category_cols))
  )

ggplot2::ggplot(
  version1_df,
  ggplot2::aes(
    x = .data$Category,
    y = .data$mean_absolute_predictive_z,
    fill = .data$Category
  )
) +
  ggplot2::geom_boxplot(
    width = 0.55,
    alpha = 0.30,
    outlier.shape = NA,
    colour = "grey35"
  ) +
  ggplot2::geom_jitter(
    ggplot2::aes(colour = .data$Category),
    width = 0.12,
    size = 2.6,
    show.legend = FALSE
  ) +
  ggplot2::scale_fill_manual(values = category_cols, drop = FALSE) +
  ggplot2::scale_colour_manual(values = category_cols, drop = FALSE) +
  ggplot2::labs(
    x = NULL,
    y = "Mean |predictive z| across STING genes",
    title = "Version 1: predictive-count deviation by SAVI category",
    subtitle = "Two-tailed posterior predictive test against the personalised healthy baseline"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    legend.position = "none",
    panel.grid.minor = ggplot2::element_blank()
  )
```

**Reading Version 1.** This test is the conservative, calibrated check for a
single observation: it includes count-level noise, so only counts that fall in
the predictive tails are flagged. We expect **SAVI** samples to carry the largest
mean `|predictive z|`, **CTRL** the smallest, and **SAVI_treated** intermediate.

------------------------------------------------------------------------

# Version 2 — Linear predictor test (single-observation z-test on log μ)

**Secondary test.** Here the estimand is the NB **location parameter** μ on the
log scale, with both sides centred to **offset = 0**.

The **patient side** is a single observed count, summarised as
`patient_log_mu = log(y + 0.5) − offset`. The `+ 0.5` continuity correction keeps
the log finite at `y = 0`, so zero counts (often the strongest
*lower-than-healthy* signal) are tested rather than silently dropped. Its
log-scale standard error is the **full single-observation delta-method SE**

$$\operatorname{SE}(\text{patient log}\,\mu) = \sqrt{\tfrac{1}{y + 0.5} + \tfrac{1}{\text{shape}}},$$

because for a negative-binomial draw
$\operatorname{Var}(\log Y) \approx \operatorname{Var}(Y)/\mu^2 = 1/\mu + 1/\text{shape}$.
The first term is the Poisson/aggregation precision; the second is the
**overdispersion floor**. Including `1 / shape` is essential: a single
overdispersed count can never pin this sample's own mean rate tighter than
`≈ sqrt(1 / shape)`, regardless of how large `y` is. (An earlier version of this
report dropped `1 / shape` and treated `sqrt(1 / y)` as the patient-mean SE. That
made the denominator artificially small and the test *spuriously* more sensitive
than Version 1 — but with a single observation the mean-parameter test **cannot**
legitimately be sharper than the predictive test, so the extra "significance" was
an artifact. Restoring `1 / shape` makes Version 2 a calibrated Gaussian companion
to Version 1 rather than a second, over-confident test.)

The **healthy side** is the posterior of the **grand mean** from
`posterior_linpred(transform = FALSE)`: per draw we average the matched covariate
grid (`rowMeans`), giving `healthy_log_mu_mean` and a tight `healthy_log_mu_sd`
(uncertainty in the healthy *mean*, not the full covariate-marginalised cloud —
see `mu_test_report.qmd`).

Because the patient side is a fixed point estimate with a known (asymptotic) SE,
the two are combined in a **z-test** on the offset-centred log μ with denominator

$$\operatorname{SE}_\text{diff} = \sqrt{\tfrac{1}{y + 0.5} + \tfrac{1}{\text{shape}} + \texttt{healthy\_log\_mu\_sd}^2},$$

`welch_t = delta_log_mu_vs_healthy / SE_diff`, and `welch_p = 2·P(Z > |welch_t|)`.
The `welch_*` column names are retained for continuity with the universal-baseline
report; `welch_df` is reported as the healthy-draw count but is large enough that
the t- and z-references coincide. The p-values remain exploratory.

```{r}
#| label: personalised-logmu-test-preview

linpred_tests |>
  dplyr::select(
    gene, sample, observed_count, patient_log_mu, patient_log_mu_se,
    healthy_log_mu_mean, healthy_log_mu_sd, delta_log_mu_vs_healthy,
    welch_t, welch_df, welch_p, welch_q_value_global, linpred_direction
  ) |>
  dplyr::mutate(
    dplyr::across(dplyr::where(is.numeric), ~round(.x, 4))
  ) |>
  dplyr::arrange(.data$sample, .data$gene)
```

```{r}
#| label: plot-version2-welch
#| fig-width: 8
#| fig-height: 4.5
#| fig-align: center

version2_df <- personalised_sample_scores |>
  dplyr::mutate(
    Category = factor(.data$Category, levels = names(category_cols))
  )

ggplot2::ggplot(
  version2_df,
  ggplot2::aes(
    x = .data$Category,
    y = .data$mean_absolute_welch_t,
    fill = .data$Category
  )
) +
  ggplot2::geom_boxplot(
    width = 0.55,
    alpha = 0.30,
    outlier.shape = NA,
    colour = "grey35"
  ) +
  ggplot2::geom_jitter(
    ggplot2::aes(colour = .data$Category),
    width = 0.12,
    size = 2.6,
    show.legend = FALSE
  ) +
  ggplot2::scale_fill_manual(values = category_cols, drop = FALSE) +
  ggplot2::scale_colour_manual(values = category_cols, drop = FALSE) +
  ggplot2::labs(
    x = NULL,
    y = "Mean |Welch t| across STING genes",
    title = "Version 2: log(mu) z-deviation by SAVI category",
    subtitle = "Offset-centred posterior_linpred mean-parameter test"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    legend.position = "none",
    panel.grid.minor = ggplot2::element_blank()
  )
```

**Reading Version 2.** This test targets the mean parameter on the log scale, but
because the patient side is a single observation its SE carries the full NB
delta-method variance (`1/(y + 0.5) + 1/shape`), including the overdispersion
floor. It is therefore a Gaussian companion to the predictive count test and is
expected to track Version 1 closely rather than out-resolve it; any large
discrepancy between the two points to non-NB tail behaviour for that gene × sample.
As in Version 1, the paper-consistent ordering is **SAVI > SAVI_treated > CTRL** on
the mean `|Welch t|`.

------------------------------------------------------------------------

# Comparing the two versions

The predictive-count test (Version 1) asks whether the **whole observed count**
is surprising, including count-level noise; it is the calibrated reference test
for one observation. The log μ test (Version 2) targets the **mean parameter** on
the log scale, but its patient SE now carries the full single-observation NB
variance (`1/(y + 0.5) + 1/shape`), so it is a Gaussian companion that should
track Version 1 closely rather than out-resolve it. Genes and samples flagged by
**both** versions are the most defensible personalised SAVI signals; a gene
flagged by Version 2 but not Version 1 mostly reflects the Gaussian-vs-exact-NB
tail approximation and should be treated with caution.

```{r}
#| label: personalised-sample-score-preview

personalised_sample_scores |>
  dplyr::mutate(
    dplyr::across(dplyr::where(is.numeric), ~round(.x, 4))
  )
```

```{r}
#| label: plot-personalised-baseline-scores
#| fig-width: 11
#| fig-height: 5
#| fig-align: center

score_plot_df <- personalised_sample_scores |>
  dplyr::mutate(
    Category = factor(.data$Category, levels = names(category_cols))
  ) |>
  dplyr::select(
    sample, Sample, Category,
    `Predictive count (|z|)` = mean_absolute_predictive_z,
    `log(mu) z (|z|)` = mean_absolute_welch_t
  ) |>
  tidyr::pivot_longer(
    cols = c(`Predictive count (|z|)`, `log(mu) z (|z|)`),
    names_to = "test_version",
    values_to = "score"
  )

ggplot2::ggplot(
  score_plot_df,
  ggplot2::aes(
    x = .data$Category,
    y = .data$score,
    fill = .data$Category,
    colour = .data$Category
  )
) +
  ggplot2::geom_boxplot(
    width = 0.55,
    alpha = 0.28,
    outlier.shape = NA,
    colour = "grey35"
  ) +
  ggplot2::geom_jitter(
    width = 0.11,
    height = 0,
    size = 2.6,
    show.legend = FALSE
  ) +
  ggplot2::facet_wrap(~test_version, scales = "free_y") +
  ggplot2::scale_fill_manual(values = category_cols, drop = FALSE) +
  ggplot2::scale_colour_manual(values = category_cols, drop = FALSE) +
  ggplot2::labs(
    x = NULL,
    y = "Mean absolute deviation across STING genes",
    title = "Two personalised-baseline test versions side by side",
    subtitle = "Left: two-tailed predictive count test. Right: offset-centred log(mu) z-test."
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    legend.position = "none",
    panel.grid.minor = ggplot2::element_blank(),
    strip.text = ggplot2::element_text(face = "bold")
  )
```

## E2 / E3 diagnostic: a tighter log μ null does not help a single observation

**E2** claims the **mean-parameter** law (grand-mean log μ) is narrower than the
**predictive count** law. That is true and is shown directly below (blue vs
orange density). **E3** — that this narrowness should make the log μ test more
powerful for a disease sample — does **not** follow, and the same figure shows
why.

A tail/rank p-value is invariant under the monotonic `log` transform, so
"convert the count to log μ and read off its location" is *not* a new test: it is
fully determined by **which null** we rank the single observation against.

- **Predictive null** (orange, `log(posterior_predict)` carrying NB noise,
  `1/shape`, and the unseen-dataset effect): ranking the observation here is
  **identical to Version 1** (same ordering ⇒ same tail p).
- **Mean-parameter null** (blue, grand-mean `posterior_linpred`): the observation
  is a single draw from the *orange* law, so scoring it against the *blue* law is
  **anti-conservative**. Below, `p(mean-null) ≪ p(pred-null)`, and even healthy
  **CTRL** observations are pushed into the blue tails — i.e. specificity
  collapses. This is exactly the artifact removed last session, re-expressed as a
  tail test.

```{r}
#| label: e2e3-diagnostic-draws
#| cache: true
#| cache.lazy: false
#| dependson: ["gene-deviation-fn", "predict-sting-vs-healthy"]

# Overlay the NARROW mean-parameter null (grand-mean log mu) and the WIDE
# predictive null (log of posterior_predict counts) on one offset-centred
# log-rate axis, and show where a single observation lands under each.
sample_category <- cd_mono |>
  dplyr::select("sample", "Category") |>
  dplyr::distinct()

# One representative gene: strongest mean |predictive z| across SAVI samples.
e2e3_example_gene <- gene_deviation_all |>
  dplyr::left_join(sample_category, by = "sample") |>
  dplyr::filter(.data$Category == "SAVI", is.finite(.data$predictive_z)) |>
  dplyr::group_by(.data$gene) |>
  dplyr::summarise(signal = mean(abs(.data$predictive_z), na.rm = TRUE), .groups = "drop") |>
  dplyr::slice_max(.data$signal, n = 1, with_ties = FALSE) |>
  dplyr::pull(.data$gene)

# One example sample per Category. comparison_samples_tbl is derived from cd_mono,
# so it already carries Category -- use it directly (joining sample_category here
# would create Category.x / Category.y).
e2e3_samples <- comparison_samples_tbl |>
  dplyr::filter(.data$sample_role == "savi_pseudobulk") |>
  dplyr::filter(!is.na(.data$Category)) |>
  dplyr::group_by(.data$Category) |>
  dplyr::slice(1) |>
  dplyr::ungroup() |>
  dplyr::select("sample", "sample_idx", "Category")

logmu_vs_predictive_for_gene <- function(gene_sym, samples_tbl, ndraws = 400L) {
  chunk_row <- sting_fit_rows |>
    dplyr::filter(.data$gene == .env$gene_sym) |>
    dplyr::slice(1)
  fit <- poco::reconstruct_brmsfit(chunk_row$brms_fit_compressed[[1]])
  feature_id <- chunk_row$.feature
  # Compressed fits may hold fewer draws than requested; cap to what exists.
  ndraws <- min(ndraws, brms::ndraws(fit))

  purrr::pmap_dfr(
    samples_tbl |> dplyr::select("sample", "sample_idx", "Category"),
    function(sample, sample_idx, Category) {
      newdata <- newdata_for_sample(se_monocytic, sample_idx)
      nd_grid <- build_newdata_grid(
        fit, newdata,
        new_level_tag = paste0("__savi__", as.character(sample))
      )
      offset_value <- newdata$offset[[1]]

      predictive_mat <- brms::posterior_predict(
        fit, newdata = nd_grid, summary = FALSE,
        re_formula = NULL, allow_new_levels = TRUE,
        sample_new_levels = "gaussian", ndraws = ndraws
      )
      linpred_mat <- brms::posterior_linpred(
        fit, newdata = nd_grid, summary = FALSE, transform = FALSE,
        re_formula = NULL, allow_new_levels = TRUE,
        sample_new_levels = "gaussian", ndraws = ndraws
      )
      shape_mat <- tryCatch(
        brms::posterior_linpred(
          fit, newdata = nd_grid, summary = FALSE, dpar = "shape",
          transform = TRUE, re_formula = NULL, allow_new_levels = TRUE,
          sample_new_levels = "gaussian", ndraws = ndraws
        ),
        error = function(e) {
          matrix(NA_real_, nrow = nrow(linpred_mat), ncol = ncol(linpred_mat))
        }
      )

      y_rep <- marginalize_predict_draws(predictive_mat)   # pooled predictive counts
      mean_logmu <- rowMeans(linpred_mat) - offset_value   # narrow mean-parameter null
      pred_logmu <- log(y_rep + 0.5) - offset_value        # wide predictive null (log scale)
      shape_med <- stats::median(
        marginalize_draw_matrix(shape_mat, method = "mean"),
        na.rm = TRUE
      )

      obs <- SummarizedExperiment::assay(
        se_monocytic[feature_id, sample_idx, drop = FALSE], "counts"
      ) |>
        as.numeric()
      patient_log_mu <- log(obs + 0.5) - offset_value
      overdisp <- if (is.finite(shape_med) && shape_med > 0) 1 / shape_med else 0
      patient_se <- sqrt(1 / (obs + 0.5) + overdisp)

      # Same observation, two different nulls.
      p_mean_two <- 2 * min(
        mean(mean_logmu >= patient_log_mu),
        mean(mean_logmu <= patient_log_mu)
      )
      p_pred_two <- 2 * min(mean(y_rep >= obs), mean(y_rep <= obs))

      dplyr::bind_rows(
        tibble::tibble(null_type = "Mean parameter (log mu)", value = mean_logmu),
        tibble::tibble(null_type = "Predictive (log count)", value = pred_logmu)
      ) |>
        dplyr::mutate(
          gene = gene_sym,
          sample = as.character(sample),
          Category = as.character(Category),
          observed_count = obs,
          observed_log_rate = patient_log_mu,
          patient_se = patient_se,
          p_meannull_two = p_mean_two,
          p_prednull_two = p_pred_two
        )
    }
  )
}

e2e3_draws <- logmu_vs_predictive_for_gene(e2e3_example_gene, e2e3_samples)

e2e3_labels <- e2e3_draws |>
  dplyr::distinct(
    gene, sample, Category, observed_count, observed_log_rate,
    patient_se, p_meannull_two, p_prednull_two
  )

e2e3_labels
```

```{r}
#| label: plot-e2e3-diagnostic
#| fig-width: 11
#| fig-height: 4.5
#| fig-align: center

e2e3_null_cols <- c(
  "Mean parameter (log mu)" = "#2C7FB8",
  "Predictive (log count)" = "#D95F0E"
)

e2e3_draws_plot <- e2e3_draws |>
  dplyr::mutate(
    Category = factor(.data$Category, levels = names(category_cols)),
    panel = paste0(.data$Category, " | ", .data$sample)
  )

e2e3_labels_plot <- e2e3_labels |>
  dplyr::mutate(
    Category = factor(.data$Category, levels = names(category_cols)),
    panel = paste0(.data$Category, " | ", .data$sample),
    lab = sprintf(
      "obs = %g\np(mean-null) = %.2g\np(pred-null) = %.2g",
      .data$observed_count, .data$p_meannull_two, .data$p_prednull_two
    )
  )

ggplot2::ggplot(
  e2e3_draws_plot,
  ggplot2::aes(x = .data$value, fill = .data$null_type, colour = .data$null_type)
) +
  ggplot2::geom_density(alpha = 0.30) +
  ggplot2::geom_vline(
    data = e2e3_labels_plot,
    ggplot2::aes(xintercept = .data$observed_log_rate),
    linetype = 2, colour = "black", inherit.aes = FALSE
  ) +
  ggplot2::geom_text(
    data = e2e3_labels_plot,
    ggplot2::aes(x = -Inf, y = Inf, label = .data$lab),
    hjust = -0.05, vjust = 1.2, size = 3, inherit.aes = FALSE
  ) +
  ggplot2::facet_wrap(~panel, scales = "free", ncol = 3) +
  ggplot2::scale_fill_manual(values = e2e3_null_cols) +
  ggplot2::scale_colour_manual(values = e2e3_null_cols) +
  ggplot2::labs(
    x = "Offset-centred log rate (log μ at offset 0)",
    y = "Density",
    fill = NULL, colour = NULL,
    title = sprintf(
      "E2/E3 diagnostic for %s: mean-parameter null vs predictive null",
      e2e3_example_gene
    ),
    subtitle = paste0(
      "Dashed line = observed log rate. The mean-parameter null (blue) is far ",
      "narrower (E2), but the observation is a draw from the predictive null ",
      "(orange). Ranking it against the blue null over-states significance ",
      "(p mean-null ≪ p pred-null = Version 1)."
    )
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    legend.position = "top",
    panel.grid.minor = ggplot2::element_blank()
  )
```

**Reading the diagnostic.** The blue (mean-parameter) density is much tighter than
the orange (predictive) density — **E2 confirmed**. But the dashed observed value
sits in the body of the *orange* law: that is the law a single count is actually
drawn from. `p(pred-null)` (= Version 1) treats it correctly; `p(mean-null)`
divides the same distance by the artificially small blue width and so reports a
much smaller p, including for CTRL. This is **E3 as a test-construction error**,
not evidence that the log μ test should win. The honest options remain: rank
against the predictive law (≡ Version 1) or add the `1/shape` floor back onto the
patient point (the Version 2 z-test) — the two agree by construction.

## Experiment: dataset random effect IN vs OUT (both tests) + CTRL calibration

The CSV audit of `gene_deviation_all` showed the log μ null width
(`healthy_log_mu_sd`, median ≈ 1.2 on the log scale) is **dominated by the
unseen-dataset random intercept** `(1 | dataset_id_altered)`, not by epistemic
uncertainty about the atlas mean — the same component the comprehensive baseline
also injects (`re_formula = NULL`, `sample_new_levels = "gaussian"`). That single
term is why even ~70–300× elevations (IL1B, PTX3) are called *consistent*.

This section runs **both** tests with the dataset RE **in** (current) and **out**
(dropped from `re_formula`, keeping the tissue RE), and reports the **CTRL
false-positive rate** alongside the SAVI sensitivity. The dataset RE is a
**cohort-level** quantity shared by SAVI *and* CTRL samples, so dropping it from
the null only removes a *real* confound if the SAVI cohort's healthy baseline
equals the atlas mean. The CTRL false-positive rate is the direct test of that
assumption. The **CTRL-calibrated** mean test is the symmetric alternative:
centre each SAVI sample on the cohort's own CTRL grand mean (removing the dataset
offset from *both* sides), which is the valid way to "remove the same thing from
the test sample."

```{r}
#| label: dataset-re-experiment
#| cache: true
#| cache.lazy: false
#| dependson: ["gene-deviation-fn", "predict-sting-vs-healthy"]

# re_formula keeping every group-level term EXCEPT the dataset random intercept.
re_formula_drop_group <- function(fit, drop_group = "dataset_id_altered") {
  fstr <- paste(deparse(stats::formula(fit$formula)), collapse = " ")
  bars <- regmatches(fstr, gregexpr("\\([^|()]*\\|[^()]*\\)", fstr))[[1]]
  keep <- bars[!grepl(drop_group, bars, fixed = TRUE)]
  if (!length(keep)) return(NA)
  stats::as.formula(paste("~", paste(keep, collapse = " + ")))
}

# Full deviation pass for a given re_formula. re_formula = "drop_dataset" derives
# the no-dataset formula per fit; otherwise the value is passed through verbatim.
run_deviation_grid <- function(re_formula = NULL) {
  purrr::imap_dfr(
    seq_len(nrow(sting_fit_rows)),
    function(i, idx) {
      chunk_row <- sting_fit_rows[i, ]
      feature_id <- chunk_row$.feature
      gene_sym <- chunk_row$gene
      fit <- poco::reconstruct_brmsfit(chunk_row$brms_fit_compressed[[1]])
      rf <- if (is.character(re_formula) && identical(re_formula, "drop_dataset")) {
        re_formula_drop_group(fit, "dataset_id_altered")
      } else {
        re_formula
      }
      purrr::pmap_dfr(
        comparison_samples_tbl |>
          dplyr::select("sample", "sample_idx", "sample_role"),
        function(sample, sample_idx, sample_role) {
          newdata <- newdata_for_sample(se_monocytic, sample_idx)
          obs <- SummarizedExperiment::assay(
            se_monocytic[feature_id, sample_idx, drop = FALSE], "counts"
          ) |>
            as.numeric()
          gene_deviation_from_fit(
            fit, newdata, obs,
            new_level_tag = paste0("__savi__", as.character(sample)),
            re_formula = rf,
            seed = 20250625L + 1000L * idx + sample_idx
          ) |>
            dplyr::mutate(
              .feature = feature_id, gene = gene_sym,
              sample = as.character(sample),
              sample_role = as.character(sample_role),
              .before = "observed_count"
            )
        }
      )
    }
  ) |>
    tibble::as_tibble() |>
    dplyr::select(tidyselect::where(\(x) !is.list(x)))
}

# Dataset RE OUT: drop (1|dataset_id_altered) from BOTH the predictive and the
# log-mu computations. (RE IN = the existing gene_deviation_all.)
gene_deviation_out <- run_deviation_grid("drop_dataset")
```

```{r}
#| label: dataset-re-experiment-summary

sample_cat <- cd_mono |>
  dplyr::select("sample", "Category") |>
  dplyr::distinct()

summarise_variant <- function(df, variant) {
  df |>
    dplyr::filter(.data$sample_role == "savi_pseudobulk") |>
    dplyr::left_join(sample_cat, by = "sample") |>
    dplyr::filter(!is.na(.data$Category)) |>
    dplyr::group_by(.data$Category) |>
    dplyr::summarise(
      n = dplyr::n(),
      pct_higher_predictive = 100 * mean(.data$direction == "higher_than_healthy"),
      pct_higher_logmu = 100 * mean(.data$linpred_direction == "higher_than_healthy"),
      .groups = "drop"
    ) |>
    dplyr::mutate(dataset_RE = variant, .before = 1)
}

# CTRL rows are expected consistent -> pct_higher there is the FALSE-POSITIVE rate.
# SAVI rows -> pct_higher is sensitivity.
dataset_re_compare <- dplyr::bind_rows(
  summarise_variant(gene_deviation_all, "IN (current)"),
  summarise_variant(gene_deviation_out, "OUT (dropped)")
) |>
  dplyr::arrange(.data$Category, .data$dataset_RE)

dataset_re_compare
```

```{r}
#| label: ctrl-calibrated-logmu

# patient_log_mu / its SE do not depend on re_formula, so reuse the current run.
logmu_tbl <- gene_deviation_all |>
  dplyr::select(
    gene, sample, sample_role, observed_count,
    patient_log_mu, patient_log_mu_se, healthy_nb_shape_median
  ) |>
  dplyr::left_join(sample_cat, by = "sample")

# Cohort CTRL reference per gene (the within-cohort healthy baseline).
ctrl_ref <- logmu_tbl |>
  dplyr::filter(.data$Category == "CTRL", is.finite(.data$patient_log_mu)) |>
  dplyr::group_by(.data$gene) |>
  dplyr::summarise(
    ctrl_mean = mean(.data$patient_log_mu),
    ctrl_sd = stats::sd(.data$patient_log_mu),
    n_ctrl = dplyr::n(),
    .groups = "drop"
  )

# SAVI / treated vs their own cohort CTRL mean: removes the dataset offset from
# BOTH sides. Single SAVI observation (patient SE) vs CTRL mean (sd / sqrt(n)).
ctrl_calibrated <- logmu_tbl |>
  dplyr::filter(
    .data$Category %in% c("SAVI", "SAVI_treated"),
    is.finite(.data$patient_log_mu)
  ) |>
  dplyr::left_join(ctrl_ref, by = "gene") |>
  dplyr::mutate(
    delta_cal = .data$patient_log_mu - .data$ctrl_mean,
    se_cal = sqrt(
      .data$patient_log_mu_se^2 + .data$ctrl_sd^2 / pmax(.data$n_ctrl, 1)
    ),
    t_cal = .data$delta_cal / .data$se_cal,
    df_cal = pmax(.data$n_ctrl - 1, 1),
    p_cal = 2 * stats::pt(abs(.data$t_cal), df = .data$df_cal, lower.tail = FALSE),
    higher_cal = .data$p_cal < 0.05 & .data$delta_cal > 0
  )

ctrl_calibrated |>
  dplyr::group_by(.data$Category) |>
  dplyr::summarise(
    n = dplyr::n(),
    median_n_ctrl = stats::median(.data$n_ctrl, na.rm = TRUE),
    pct_higher_ctrl_calibrated = 100 * mean(.data$higher_cal, na.rm = TRUE),
    .groups = "drop"
  )
```

**Reading the experiment.** In `dataset_re_compare`, compare the `IN` vs `OUT`
rows: dropping the dataset RE should sharply raise SAVI `pct_higher` (more power)
**but also raise the CTRL `pct_higher`** (false positives) if the cohort sits
above the atlas. If CTRL false positives jump, the unconfounded read is the
**CTRL-calibrated** table: SAVI / SAVI_treated elevation measured against the
cohort's own controls, which is the statistically clean "mean test with the
dataset effect removed from both sides." Watch `median_n_ctrl` — with few
controls the calibrated SE is noisy and the test is conservative.

------------------------------------------------------------------------

# Evaluate deviations against the published STING signature

De Cevins *et al.* (2023) define a **21-gene STING-activation module** that is **up in disease-associated monocytes from SAVI patients relative to healthy controls** in the same cohort (Figure 6; Table S6)—after removing IFN-β–inducible genes. Our posterior test asks a **different** question: whether each pseudobulk count lies above the **Immune Atlas healthy predictive** distribution for monocytic cells, given donor metadata. We still expect:

| `Category` (study cohort) | Paper-consistent expectation for signature genes |
|---------------------------|---------------------------------------------------|
| **`SAVI`** | Mostly **`higher_than_healthy`** vs atlas baseline |
| **`CTRL`** | Mostly **`consistent_with_healthy`** (study controls without STING activation) |
| **`SAVI_treated`** | Intermediate—treatment samples often move toward baseline (we score **`consistent_with_healthy`** as the nominal match, same as controls) |

Sample **`Category`** comes from the GEO Seurat object (`prepare_data.R`) and is carried in **`colData(se_monocytic)`** (`CTRL`, `SAVI`, `SAVI_treated`; reference columns are **`NA`**).

**Two versions.** Deviation is evaluated separately for each test method, both at a **0.05 p-value threshold**:

- **Version 1 (predictive count):** a gene × sample is `higher_than_healthy` / `lower_than_healthy` when the two-tailed posterior predictive p-value `predictive_tail_p < 0.05` (sign from the predictive percentile), otherwise `consistent_with_healthy`.
- **Version 2 (log μ z-test):** a gene × sample is `higher_than_healthy` / `lower_than_healthy` when `welch_p < 0.05` (sign from the offset-centred log μ difference), otherwise `consistent_with_healthy`.

```{r}
#| label: sting-literature-eval
sample_annotations <- cd_mono |>
  dplyr::filter(.data$sample_role == "savi_pseudobulk") |>
  dplyr::select(
    "sample",
    "Sample",
    "Category",
    tidyselect::any_of("Treatment")
  ) |>
  dplyr::distinct() |>
  dplyr::mutate(
    Category = factor(
      .data$Category,
      levels = c("CTRL", "SAVI", "SAVI_treated")
    )
  )

category_expected <- tibble::tribble(
  ~Category,        ~expected_deviation,
  "CTRL",           "consistent_with_healthy",
  "SAVI",           "higher_than_healthy",
  "SAVI_treated",   "consistent_with_healthy"
) |>
  dplyr::mutate(Category = factor(.data$Category, levels = levels(sample_annotations$Category)))

# Both deviation directions are already classified at p < 0.05 upstream:
#   `direction`         = predictive count (predictive_tail_p < 0.05)
#   `linpred_direction` = log(mu) z-test (welch_p < 0.05)
evaluate_signature <- function(df, direction_col) {
  df |>
    dplyr::left_join(sample_annotations, by = "sample") |>
    dplyr::left_join(category_expected, by = "Category") |>
    dplyr::mutate(
      eval_direction = .data[[direction_col]],
      deviation_score = dplyr::case_when(
        .data$eval_direction == "higher_than_healthy" ~ 1L,
        .data$eval_direction == "lower_than_healthy" ~ -1L,
        TRUE ~ 0L
      ),
      matches_literature = dplyr::case_when(
        is.na(.data$expected_deviation) ~ NA,
        .data$expected_deviation == .data$eval_direction ~ TRUE,
        TRUE ~ FALSE
      ),
      signature_up = .data$eval_direction == "higher_than_healthy"
    )
}

summarise_by_sample <- function(eval_df) {
  eval_df |>
    dplyr::group_by(.data$sample, .data$Sample, .data$Category) |>
    dplyr::summarise(
      n_genes = dplyr::n(),
      n_higher = sum(.data$signature_up),
      n_lower = sum(.data$eval_direction == "lower_than_healthy"),
      n_consistent = sum(.data$eval_direction == "consistent_with_healthy"),
      pct_higher = 100 * .data$n_higher / .data$n_genes,
      pct_match = 100 * mean(.data$matches_literature, na.rm = TRUE),
      mean_log2fc = mean(.data$log2fc_vs_median, na.rm = TRUE),
      signature_score = mean(.data$deviation_score, na.rm = TRUE),
      .groups = "drop"
    )
}

summarise_agreement <- function(eval_df) {
  eval_df |>
    dplyr::filter(!is.na(.data$Category), !is.na(.data$expected_deviation)) |>
    dplyr::group_by(.data$Category, .data$expected_deviation) |>
    dplyr::summarise(
      n_tests = dplyr::n(),
      n_match = sum(.data$matches_literature, na.rm = TRUE),
      pct_match = 100 * .data$n_match / .data$n_tests,
      pct_higher = 100 * mean(.data$signature_up),
      pct_consistent = 100 * mean(.data$eval_direction == "consistent_with_healthy"),
      .groups = "drop"
    )
}

# Version 1 — predictive count (two-tailed p < 0.05). Kept as the primary object
# consumed by the downstream biomarker, IL1B, and ADRB2 sections.
gene_deviation_eval <- evaluate_signature(gene_deviation_savi, "direction")
signature_by_sample <- summarise_by_sample(gene_deviation_eval)
agreement_by_category <- summarise_agreement(gene_deviation_eval)

# Version 2 — log(mu) z-test (welch_p < 0.05).
gene_deviation_eval_linpred <- evaluate_signature(gene_deviation_savi, "linpred_direction")
signature_by_sample_linpred <- summarise_by_sample(gene_deviation_eval_linpred)
agreement_by_category_linpred <- summarise_agreement(gene_deviation_eval_linpred)

method_levels <- c(
  "Predictive count (p < 0.05)",
  "log(mu) z-test (p < 0.05)"
)

signature_by_sample_both <- dplyr::bind_rows(
  signature_by_sample |> dplyr::mutate(method = method_levels[[1]]),
  signature_by_sample_linpred |> dplyr::mutate(method = method_levels[[2]])
) |>
  dplyr::mutate(method = factor(.data$method, levels = method_levels))

agreement_both <- dplyr::bind_rows(
  agreement_by_category |> dplyr::mutate(method = method_levels[[1]], .before = 1),
  agreement_by_category_linpred |> dplyr::mutate(method = method_levels[[2]], .before = 1)
) |>
  dplyr::mutate(method = factor(.data$method, levels = method_levels))

agreement_both
```

The plots below show both versions side by side (faceted by method). The
downstream biomarker-quality, *IL1B*, and *ADRB2* falsification analyses use
**Version 1 (predictive count)** as the primary deviation call; Version 2 is the
log μ z-test companion for the same 0.05 threshold.

```{r}
#| label: plot-signature-by-category
#| fig-width: 12
#| fig-height: 9
#| fig-align: center

category_cols <- c(
  CTRL = "#2ca25f",
  SAVI = "#de2d26",
  SAVI_treated = "#756bb1"
)

p_pct_higher <- ggplot2::ggplot(
  signature_by_sample_both,
  ggplot2::aes(
    x = .data$Category,
    y = .data$pct_higher,
    fill = .data$Category
  )
) +
  ggplot2::geom_boxplot(
    outlier.shape = NA,
    alpha = 0.35,
    width = 0.55,
    colour = "grey30"
  ) +
  ggplot2::geom_jitter(
    ggplot2::aes(colour = .data$Category),
    width = 0.12,
    size = 2.5,
    show.legend = FALSE
  ) +
  ggplot2::facet_wrap(~method) +
  ggplot2::scale_fill_manual(values = category_cols) +
  ggplot2::scale_colour_manual(values = category_cols) +
  ggplot2::labs(
    x = NULL,
    y = "% signature genes higher than atlas baseline",
    title = "STING signature enrichment by study Category",
    subtitle = "Both versions at p < 0.05. Paper: up in SAVI monocytes vs controls (Figure 6)"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    legend.position = "none",
    strip.text = ggplot2::element_text(face = "bold")
  )

p_signature_score <- ggplot2::ggplot(
  signature_by_sample_both,
  ggplot2::aes(
    x = .data$Category,
    y = .data$signature_score,
    fill = .data$Category
  )
) +
  ggplot2::geom_boxplot(
    outlier.shape = NA,
    alpha = 0.35,
    width = 0.55,
    colour = "grey30"
  ) +
  ggplot2::geom_jitter(
    ggplot2::aes(colour = .data$Category),
    width = 0.12,
    size = 2.5,
    show.legend = FALSE
  ) +
  ggplot2::geom_hline(yintercept = 0, linetype = 2, colour = "grey50") +
  ggplot2::facet_wrap(~method) +
  ggplot2::scale_fill_manual(values = category_cols) +
  ggplot2::scale_colour_manual(values = category_cols) +
  ggplot2::labs(
    x = NULL,
    y = "Mean deviation score\n(−1 lower, 0 consistent, +1 higher)",
    title = "Per-sample composite STING deviation score"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    legend.position = "none",
    strip.text = ggplot2::element_text(face = "bold")
  )

p_pct_higher / p_signature_score
```

```{r}
#| label: plot-literature-agreement-sting
#| fig-width: 11
#| fig-height: 4.5
#| fig-align: center

ggplot2::ggplot(
  agreement_both,
  ggplot2::aes(
    x = .data$Category,
    y = .data$pct_match,
    fill = .data$Category
  )
) +
  ggplot2::geom_col(width = 0.65, colour = "white") +
  ggplot2::geom_text(
    ggplot2::aes(
      label = sprintf(
        "%.0f%% match\n(%.0f%% higher)",
        .data$pct_match,
        .data$pct_higher
      )
    ),
    vjust = -0.15,
    size = 3.0
  ) +
  ggplot2::facet_wrap(~method) +
  ggplot2::scale_fill_manual(values = category_cols) +
  ggplot2::scale_y_continuous(limits = c(0, 115), expand = c(0, 0)) +
  ggplot2::labs(
    x = NULL,
    y = "% gene × sample tests matching category expectation",
    title = "Posterior deviation vs paper-consistent expectation",
    subtitle = "Two versions at p < 0.05. Bars = % of gene × sample tests matching expected direction, pooled within each Category."
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    legend.position = "none",
    strip.text = ggplot2::element_text(face = "bold")
  )
```

```{r}
#| label: heatmap-sting-category
#| fig-width: 15
#| fig-height: 12
#| fig-align: center

heatmap_df <- gene_deviation_eval |>
  dplyr::filter(!is.na(.data$Category)) |>
  dplyr::mutate(
    Category = factor(.data$Category, levels = names(category_cols))
  )

mat_log2fc <- heatmap_df |>
  dplyr::select("gene", "sample", "log2fc_vs_median") |>
  tidyr::pivot_wider(names_from = "sample", values_from = "log2fc_vs_median") |>
  tibble::column_to_rownames("gene") |>
  as.matrix()

score_mat <- heatmap_df |>
  dplyr::select("gene", "sample", "deviation_score") |>
  tidyr::pivot_wider(names_from = "sample", values_from = "deviation_score") |>
  tibble::column_to_rownames("gene") |>
  as.matrix()

# Category only in annotation_col: pheatmap draws one legend per annotated
# column; adding Sample (~17 levels, no ann_colors) stacks a second legend and
# the "Category" title overlaps the colour keys (see pheatmap issue with multi-legends).
annotation_col <- heatmap_df |>
  dplyr::distinct(.data$sample, .data$Category) |>
  tibble::column_to_rownames("sample") |>
  dplyr::select("Category")

sample_labels <- heatmap_df |>
  dplyr::distinct(.data$sample, .data$Sample) |>
  tibble::deframe()

labels_col <- unname(sample_labels[colnames(mat_log2fc)])

ann_colors <- list(Category = category_cols)

lim_log2 <- max(abs(mat_log2fc), na.rm = TRUE)
if (!is.finite(lim_log2) || lim_log2 == 0) lim_log2 <- 1

pheatmap::pheatmap(
  mat_log2fc,
  annotation_col = annotation_col,
  annotation_colors = ann_colors,
  labels_col = labels_col,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  clustering_method = "ward.D2",
  color = grDevices::colorRampPalette(rev(RColorBrewer::brewer.pal(11, "RdBu")))(100),
  breaks = seq(-lim_log2, lim_log2, length.out = 101),
  border_color = NA,
  fontsize_row = 8,
  fontsize_col = 7,
  angle_col = 45,
  main = "log2(obs / healthy predictive median)\nSTING genes × SAVI pseudobulk (Category annotated)"
)

pheatmap::pheatmap(
  score_mat,
  annotation_col = annotation_col,
  annotation_colors = ann_colors,
  labels_col = labels_col,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  clustering_method = "ward.D2",
  color = grDevices::colorRampPalette(c("#2166ac", "#f7f7f7", "#b2182b"))(3),
  breaks = c(-1.5, -0.5, 0.5, 1.5),
  legend_breaks = c(-1, 0, 1),
  legend_labels = c("lower", "consistent", "higher"),
  border_color = NA,
  fontsize_row = 8,
  fontsize_col = 7,
  angle_col = 45,
  main = "Deviation direction vs Immune Atlas baseline\n(Category = CTRL / SAVI / SAVI_treated)"
)
```

```{r}
#| label: plot-gene-match-by-category
#| fig-width: 10
#| fig-height: 6
#| fig-align: center

gene_match_summary <- gene_deviation_eval |>
  dplyr::filter(!is.na(.data$Category)) |>
  dplyr::group_by(.data$gene, .data$Category, .data$expected_deviation) |>
  dplyr::summarise(
    n = dplyr::n(),
    pct_higher = 100 * mean(.data$signature_up),
    pct_match = 100 * mean(.data$matches_literature, na.rm = TRUE),
    mean_log2fc = mean(.data$log2fc_vs_median, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    gene = forcats::fct_reorder(.data$gene, .data$mean_log2fc, .fun = mean, .desc = TRUE)
  )

ggplot2::ggplot(
  gene_match_summary,
  ggplot2::aes(
    x = .data$gene,
    y = .data$Category,
    fill = .data$mean_log2fc
  )
) +
  ggplot2::geom_tile(colour = "white", linewidth = 0.3) +
  ggplot2::scale_fill_gradient2(
    low = "#2166ac",
    mid = "#f7f7f7",
    high = "#b2182b",
    midpoint = 0,
    name = "Mean log2 FC\nvs baseline"
  ) +
  ggplot2::labs(
    x = NULL,
    y = NULL,
    title = "STING signature genes × study Category",
    subtitle = "Tile = mean log2(obs / healthy predictive median) across samples in each Category"
  ) +
  ggplot2::theme_minimal(base_size = 10) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, size = 7),
    panel.grid = ggplot2::element_blank()
  )
```

```{r}
#| label: eval-summary-table
gene_deviation_eval |>
  dplyr::filter(!is.na(.data$Category)) |>
  dplyr::group_by(.data$Category, .data$gene) |>
  dplyr::summarise(
    n_samples = dplyr::n_distinct(.data$sample),
    n_higher = sum(.data$signature_up),
    pct_higher = 100 * mean(.data$signature_up),
    matches_paper = all(.data$matches_literature %in% TRUE),
    .groups = "drop"
  ) |>
  dplyr::group_by(.data$Category) |>
  dplyr::summarise(
    n_genes_tested = dplyr::n(),
    n_genes_all_higher = sum(.data$n_higher == .data$n_samples),
    n_genes_majority_higher = sum(.data$pct_higher > 50),
    n_genes_paper_perfect = sum(.data$matches_paper),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    agreement_by_category |>
      dplyr::select("Category", "pct_match", "pct_higher"),
    by = "Category"
  )
```

------------------------------------------------------------------------

# Biomarker quality check (atlas-based falsification)

De Cevins *et al.* (2023) propose the **21-gene signature** by **within-cohort comparison** (5 SAVI vs 7 in-study controls). With only seven matched donors, a gene that is simply **variable in healthy people** but happened to be low in their controls can pass that filter. We can sanity-check each signature gene against a much broader **healthy reference** (Immune Atlas monocytic posteriors), which is the kind of falsification their paper could not do. The expected pattern under a true SAVI biomarker is:

- **SAVI** samples → consistently **higher** than the atlas healthy distribution
- **SAVI_treated** samples → **intermediate** (attenuated by treatment)
- **CTRL** samples → **consistent with healthy** (atlas baseline matches the in-cohort controls)

A gene fails the check (paper-supported but **not** atlas-supported) when:

- **CTRL drifts up** vs atlas (gene is high in many healthy people, not just SAVI),
- **No SAVI lift** vs CTRL (paper effect is within healthy variability),
- **SAVI_treated ≥ SAVI** (no treatment effect even though paper claims one).

```{r}
#| label: biomarker-per-category
biomarker_per_category <- gene_deviation_eval |>
  dplyr::filter(!is.na(.data$Category)) |>
  dplyr::group_by(.data$gene, .data$Category) |>
  dplyr::summarise(
    n_samples = dplyr::n_distinct(.data$sample),
    mean_log2fc = mean(.data$log2fc_vs_median, na.rm = TRUE),
    median_log2fc = stats::median(.data$log2fc_vs_median, na.rm = TRUE),
    pct_higher = 100 * mean(.data$direction == "higher_than_healthy"),
    pct_lower = 100 * mean(.data$direction == "lower_than_healthy"),
    pct_consistent = 100 * mean(.data$direction == "consistent_with_healthy"),
    mean_p_above_healthy = mean(.data$p_above_healthy, na.rm = TRUE),
    .groups = "drop"
  )

biomarker_per_category |>
  dplyr::arrange(.data$gene, .data$Category)
```

```{r}
#| label: biomarker-quality-score
biomarker_quality <- biomarker_per_category |>
  dplyr::select(
    "gene",
    "Category",
    "mean_log2fc",
    "pct_higher",
    "pct_consistent"
  ) |>
  tidyr::pivot_wider(
    names_from = "Category",
    values_from = c("mean_log2fc", "pct_higher", "pct_consistent"),
    values_fill = NA_real_,
    names_sep = "."
  ) |>
  dplyr::mutate(
    savi_lift = .data$mean_log2fc.SAVI,
    ctrl_drift = .data$mean_log2fc.CTRL,
    treated_lift = .data$mean_log2fc.SAVI_treated,

    # SAVI vs in-cohort controls' atlas-corrected position (paper's main claim).
    savi_vs_ctrl = .data$savi_lift - .data$ctrl_drift,

    # Treatment effect: positive means SAVI_treated drops back toward baseline.
    treatment_attenuation = .data$savi_lift - .data$treated_lift,

    # Atlas baseline agreement of in-study controls (closer to 0 = better).
    abs_ctrl_drift = abs(.data$ctrl_drift),

    # Composite quality: reward SAVI > CTRL, reward treatment drop,
    # penalise CTRL drift from atlas. Weights chosen so all three terms are on
    # the log2 scale; tweak if needed.
    quality_score = .data$savi_vs_ctrl +
      0.5 * .data$treatment_attenuation -
      0.5 * .data$abs_ctrl_drift,

    biomarker_class = dplyr::case_when(
      .data$pct_higher.SAVI >= 60 &
        .data$pct_consistent.CTRL >= 50 &
        .data$treatment_attenuation > 0 ~ "validated",
      .data$pct_higher.SAVI >= 60 &
        .data$treatment_attenuation > 0 &
        .data$pct_consistent.CTRL < 50 ~ "validated_ctrl_drifts",
      .data$pct_higher.SAVI >= 60 &
        .data$treatment_attenuation <= 0 ~ "savi_up_no_treatment_effect",
      .data$pct_higher.SAVI < 30 &
        .data$ctrl_drift > 0.5 ~ "falsified_baseline_high",
      .data$pct_higher.SAVI < 30 ~ "falsified_no_savi_signal",
      TRUE ~ "ambiguous"
    )
  ) |>
  dplyr::arrange(dplyr::desc(.data$quality_score))

biomarker_quality |>
  dplyr::select(
    "gene",
    "biomarker_class",
    "quality_score",
    "savi_vs_ctrl",
    "treatment_attenuation",
    "abs_ctrl_drift",
    dplyr::starts_with("pct_higher."),
    dplyr::starts_with("pct_consistent.")
  )
```

```{r}
#| label: plot-biomarker-ranking
#| fig-width: 9
#| fig-height: 7
#| fig-align: center

gene_order <- biomarker_quality$gene

biomarker_long <- biomarker_per_category |>
  dplyr::filter(.data$gene %in% gene_order) |>
  dplyr::mutate(
    gene = factor(.data$gene, levels = rev(gene_order)),
    Category = factor(
      .data$Category,
      levels = c("CTRL", "SAVI_treated", "SAVI")
    )
  )

ggplot2::ggplot(
  biomarker_long,
  ggplot2::aes(
    x = .data$mean_log2fc,
    y = .data$gene,
    colour = .data$Category
  )
) +
  ggplot2::geom_vline(xintercept = 0, linetype = 2, colour = "grey60") +
  ggplot2::geom_line(
    ggplot2::aes(group = .data$gene),
    colour = "grey80",
    linewidth = 0.4
  ) +
  ggplot2::geom_point(size = 3, alpha = 0.85) +
  ggplot2::scale_colour_manual(values = category_cols) +
  ggplot2::labs(
    x = "Mean log2(obs / healthy predictive median)",
    y = NULL,
    title = "Biomarker ranking vs Immune Atlas baseline",
    subtitle = "Genes ranked by quality_score; ideal pattern: SAVI > SAVI_treated > CTRL ≈ 0"
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

```{r}
#| label: plot-biomarker-ranking2
#| fig-width: 9
#| fig-height: 7
#| fig-align: center


ggplot2::ggplot(
  biomarker_long,
  ggplot2::aes(
    x = .data$median_log2fc,
    y = .data$gene,
    colour = .data$Category
  )
) +
  ggplot2::geom_vline(xintercept = 0, linetype = 2, colour = "grey60") +
  ggplot2::geom_line(
    ggplot2::aes(group = .data$gene),
    colour = "grey80",
    linewidth = 0.4
  ) +
  ggplot2::geom_point(size = 3, alpha = 0.85) +
  ggplot2::scale_colour_manual(values = category_cols) +
  ggplot2::labs(
    x = "Median log2(obs / healthy predictive median)",
    y = NULL,
    title = "Biomarker ranking vs Immune Atlas baseline",
    subtitle = "Genes ranked by quality_score; ideal pattern: SAVI > SAVI_treated > CTRL ≈ 0"
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

```{r}
#| label: plot-biomarker-class
#| fig-width: 9
#| fig-height: 6
#| fig-align: center

class_levels <- c(
  "validated",
  "validated_ctrl_drifts",
  "savi_up_no_treatment_effect",
  "ambiguous",
  "falsified_baseline_high",
  "falsified_no_savi_signal"
)

class_cols <- c(
  validated = "#1a9850",
  validated_ctrl_drifts = "#66bd63",
  savi_up_no_treatment_effect = "#fee08b",
  ambiguous = "#bdbdbd",
  falsified_baseline_high = "#762a83",
  falsified_no_savi_signal = "#d73027"
)

ggplot2::ggplot(
  biomarker_quality |>
    dplyr::mutate(
      gene = factor(.data$gene, levels = rev(gene_order)),
      biomarker_class = factor(
        .data$biomarker_class,
        levels = class_levels
      )
    ),
  ggplot2::aes(
    x = .data$quality_score,
    y = .data$gene,
    fill = .data$biomarker_class
  )
) +
  ggplot2::geom_col() +
  ggplot2::geom_vline(xintercept = 0, linetype = 2, colour = "grey50") +
  ggplot2::scale_fill_manual(values = class_cols, drop = FALSE) +
  ggplot2::labs(
    x = "Quality score = (SAVI − CTRL) + 0.5 × (SAVI − treated) − 0.5 × |CTRL drift|",
    y = NULL,
    title = "STING signature biomarker quality",
    subtitle = "Top: best match to expected pattern; bottom: atlas-falsified candidates",
    fill = NULL
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "right")
```

```{r}
#| label: biomarker-class-summary
biomarker_quality |>
  dplyr::count(.data$biomarker_class, name = "n_genes") |>
  dplyr::arrange(dplyr::desc(.data$n_genes))
```

```{r}
#| label: biomarker-falsified-genes
biomarker_quality |>
  dplyr::filter(.data$biomarker_class %in% c(
    "falsified_no_savi_signal",
    "falsified_baseline_high",
    "savi_up_no_treatment_effect"
  )) |>
  dplyr::select(
    "gene",
    "biomarker_class",
    "savi_lift",
    "ctrl_drift",
    "treated_lift",
    "pct_higher.SAVI",
    "pct_higher.CTRL",
    "pct_consistent.CTRL"
  ) |>
  dplyr::arrange(.data$biomarker_class, .data$gene)
```

**Interpretation:**

- **`validated`** — gene behaves as the paper claims and is **also** consistent with the atlas (high in SAVI, attenuated by treatment, in-cohort controls match atlas baseline). Strong biomarker.
- **`validated_ctrl_drifts`** — gene goes up in SAVI and drops with treatment, but the paper's controls are already above the atlas baseline. The within-cohort effect is real; absolute reference value should be interpreted with care.
- **`savi_up_no_treatment_effect`** — SAVI lift exists but treatment doesn't move it. The gene may be a state marker rather than an activity readout.
- **`ambiguous`** — borderline; needs more samples.
- **`falsified_baseline_high`** — gene reads as high in study controls *and* in the broader atlas, and SAVI doesn't add much on top. The paper's "up in SAVI" is likely a baseline artefact.
- **`falsified_no_savi_signal`** — no SAVI lift vs atlas at all. Strong falsification candidate.

This is the kind of cross-study check the original paper could not perform because of the small healthy-cohort size.

------------------------------------------------------------------------

# Case study: falsifying *IL1B*

**IL1B** is the clearest atlas-falsified gene in this cohort: study **controls sit above** the Immune Atlas healthy predictive distribution, while **SAVI samples do not sit higher still**. That pattern breaks the signature logic (“up in patients vs healthy”) when “healthy” is the atlas rather than seven matched donors.

To make the falsification auditable we show:

1. **Per-sample posterior predictive checks** — observed count vs healthy HPD (already computed).
2. **Predictive densities** — draw-level `posterior_predict` for IL1B vs a **validated** contrast gene (**TSC22D2**).
3. **Three reference frames** — atlas-corrected mean log2 FC vs **within-cohort** category means (raw and TMM-scaled counts averaged over donors in each `Category` only; no atlas comparison on the paper panels).

```{r}
#| label: il1b-falsify-helpers
falsify_contrast_genes <- c("IL1B", "TSC22D2")

scaled_counts_assay_name <- function(se) {
  nms <- SummarizedExperiment::assayNames(se)
  if ("counts_scaled" %in% nms) {
    return("counts_scaled")
  }
  stop(
    "Assay `counts_scaled` not found on `se_monocytic`. ",
    "Run the `scale-abundance` chunk first."
  )
}

il1b_ensembl <- sting_signature_21 |>
  dplyr::filter(.data$gene == "IL1B") |>
  dplyr::pull(.feature)

il1b_scaled_obs <- tibble::tibble(sample = colnames(se_monocytic)) |>
  dplyr::mutate(
    sample_idx = match(.data$sample, colnames(se_monocytic)),
    observed_scaled = as.numeric(
      SummarizedExperiment::assay(
        se_monocytic[il1b_ensembl, .data$sample_idx, drop = FALSE],
        scaled_counts_assay_name(se_monocytic)
      )
    )
  )

il1b_sample_tbl <- gene_deviation_eval |>
  dplyr::filter(.data$gene == "IL1B", !is.na(.data$Category)) |>
  dplyr::left_join(
    sample_annotations |>
      dplyr::select("sample", tidyselect::any_of("Treatment")),
    by = "sample"
  ) |>
  dplyr::left_join(il1b_scaled_obs, by = "sample") |>
  dplyr::mutate(
    Category = factor(.data$Category, levels = names(category_cols)),
    # Map count-scale predictive summaries to TMM-scaled scale (per-sample ratio).
    scale_ratio = .data$observed_scaled / pmax(.data$observed_count, 1e-9),
    healthy_median_scaled = .data$healthy_median * .data$scale_ratio,
    healthy_hpd_lo_scaled = .data$healthy_hpd_lo * .data$scale_ratio,
    healthy_hpd_hi_scaled = .data$healthy_hpd_hi * .data$scale_ratio,
    above_hpd = .data$observed_scaled > .data$healthy_hpd_hi_scaled,
    below_hpd = .data$observed_scaled < .data$healthy_hpd_lo_scaled,
    in_hpd = !.data$above_hpd & !.data$below_hpd
  ) |>
  dplyr::arrange(.data$Category, .data$Sample)

il1b_sample_tbl |>
  dplyr::select(
    "Sample",
    "Category",
    "observed_scaled",
    "healthy_median_scaled",
    "healthy_hpd_lo_scaled",
    "healthy_hpd_hi_scaled",
    "obs_percentile",
    "direction",
    "log2fc_vs_median"
  )
```

```{r}
#| label: plot-il1b-obs-vs-hpd
#| fig-width: 11
#| fig-height: 7
#| fig-align: center

il1b_plot_df <- il1b_sample_tbl |>
  dplyr::mutate(
    sample_label = factor(
      Sample,
      levels = il1b_sample_tbl |>
        dplyr::arrange(Category, Sample) |>
        dplyr::pull(Sample) |>
        unique()
    )
  )

log1p_trans <- scales::trans_new(
  name = "log1p",
  transform = function(x) log1p(pmax(x, 0)),
  inverse = function(x) expm1(x),
  domain = c(0, Inf)
)

ggplot2::ggplot(il1b_plot_df) +
  ggplot2::geom_linerange(
    ggplot2::aes(
      x = sample_label,
      ymin = healthy_hpd_lo,
      ymax = healthy_hpd_hi
    ),
    colour = "grey45",
    linewidth = 0.7
  ) +
  ggplot2::geom_point(
    ggplot2::aes(x = sample_label, y = healthy_median),
    shape = 4,
    size = 2.2,
    colour = "grey55",
    stroke = 0.9
  ) +
  ggplot2::geom_point(
    ggplot2::aes(x = sample_label, y = observed_count, colour = Category),
    size = 3.5
  ) +
  ggplot2::scale_y_continuous(trans = log1p_trans) +
  ggplot2::scale_colour_manual(values = category_cols) +
  ggplot2::labs(
    x = NULL,
    y = expression(log[1 + p]("raw pseudobulk count")),
    title = expression(italic("IL1B") * ": raw counts vs atlas healthy predictive interval"),
    subtitle = "Each point = one donor; x-axis sorted by Category (CTRL, SAVI_treated, SAVI). Grey bar = 95% atlas predictive HPD; + = predictive median."
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```

```{r}
#| label: plot-il1b-percentile
#| fig-width: 14
#| fig-height: 6
#| fig-align: center

il1b_counts_long <- il1b_sample_tbl |>
  dplyr::mutate(
    sample_label = factor(
      Sample,
      levels = il1b_sample_tbl |>
        dplyr::arrange(Category, Sample) |>
        dplyr::pull(Sample) |>
        unique()
    )
  ) |>
  tidyr::pivot_longer(
    cols = c("observed_count", "observed_scaled"),
    names_to = "measure",
    values_to = "abundance"
  ) |>
  dplyr::mutate(
    measure = dplyr::recode(
      .data$measure,
      observed_count = "Raw pseudobulk counts",
      observed_scaled = "TMM-scaled counts"
    ),
    measure = factor(
      .data$measure,
      levels = c("Raw pseudobulk counts", "TMM-scaled counts")
    )
  )

log1p_trans <- scales::trans_new(
  name = "log1p",
  transform = function(x) log1p(pmax(x, 0)),
  inverse = function(x) expm1(x),
  domain = c(0, Inf)
)

ggplot2::ggplot(
  il1b_counts_long,
  ggplot2::aes(
    x = sample_label,
    y = abundance,
    colour = Category
  )
) +
  ggplot2::geom_point(size = 3.5) +
  ggplot2::facet_wrap(~measure, scales = "free_y", ncol = 2) +
  ggplot2::scale_y_continuous(trans = log1p_trans) +
  ggplot2::scale_colour_manual(values = category_cols) +
  ggplot2::labs(
    x = NULL,
    y = expression(log[1 + p]("count")),
    title = expression(italic("IL1B") * ": per-donor abundance (raw vs TMM-scaled)"),
    subtitle = paste0(
      "Each point = one donor pseudobulk in the GSE226598 cohort (not a category mean). ",
      "Atlas percentile is identical on both scales (linear TMM rescaling); ",
      "see obs_percentile in the table above."
    )
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, size = 8),
    legend.position = "top",
    strip.text = ggplot2::element_text(size = 10)
  )
```

```{r}
#| label: plot-il1b-two-nulls
#| fig-width: 14
#| fig-height: 4.5
#| fig-align: center

# Three reference frames for IL1B:
#   Atlas  — mean log2(obs / healthy predictive median) across donors in each Category.
#   Paper  — mean abundance within the GSE226598 cohort only (no atlas); bars = cohort mean
#            per Category (CTRL n≈7, SAVI n≈5, SAVI_treated n varies), not individual donors.
paper_null <- il1b_sample_tbl |>
  dplyr::group_by(Category) |>
  dplyr::summarise(
    mean_obs = mean(observed_count),
    mean_scaled = mean(observed_scaled),
    mean_log1p_scaled = mean(log1p(pmax(observed_scaled, 0))),
    .groups = "drop"
  )

atlas_null <- il1b_sample_tbl |>
  dplyr::group_by(Category) |>
  dplyr::summarise(
    mean_log2fc = mean(log2fc_vs_median),
    .groups = "drop"
  )

two_nulls_long <- dplyr::bind_rows(
  atlas_null |>
    dplyr::mutate(
      metric = "Atlas: mean log2 FC vs healthy baseline",
      value = .data$mean_log2fc,
      y_label = "Mean log2(obs / healthy median)"
    ) |>
    dplyr::select("Category", "metric", "value", "y_label"),
  paper_null |>
    dplyr::mutate(
      metric = "Within-cohort mean (raw pseudobulk counts)",
      value = .data$mean_obs,
      y_label = "Mean raw count"
    ) |>
    dplyr::select("Category", "metric", "value", "y_label"),
  paper_null |>
    dplyr::mutate(
      metric = "Within-cohort mean (log1p TMM-scaled counts)",
      value = .data$mean_log1p_scaled,
      y_label = "log1p(mean scaled count)"
    ) |>
    dplyr::select("Category", "metric", "value", "y_label")
) |>
  dplyr::mutate(
    Category = factor(Category, levels = names(category_cols)),
    metric = factor(
      metric,
      levels = c(
        "Atlas: mean log2 FC vs healthy baseline",
        "Within-cohort mean (raw pseudobulk counts)",
        "Within-cohort mean (log1p TMM-scaled counts)"
      )
    )
  )

ggplot2::ggplot(
  two_nulls_long,
  ggplot2::aes(x = Category, y = value, fill = Category)
) +
  ggplot2::geom_col(width = 0.6, colour = "white") +
  ggplot2::geom_hline(yintercept = 0, linetype = 2, colour = "grey40") +
  ggplot2::facet_wrap(~metric, scales = "free_y", ncol = 3) +
  ggplot2::scale_fill_manual(values = category_cols) +
  ggplot2::labs(
    x = NULL,
    y = NULL,
    title = expression(italic("IL1B") * ": atlas baseline vs within-cohort category means"),
    subtitle = paste0(
      "Bars = mean across donors in each Category (same GSE226598 cohort). ",
      "Left compares to Immune Atlas healthy predictive median; centre/right are paper-style ",
      "within-cohort contrasts only (raw or TMM-scaled pseudobulk counts)."
    )
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    legend.position = "none",
    strip.text = ggplot2::element_text(size = 9)
  )
```

```{r}
#| label: il1b-predictive-draws
#| cache: true
#| cache.lazy: false
#| dependson: ["gene-deviation-fn"]

# Subsampled posterior_predict draws for density overlays (IL1B vs TSC22D2).
# Model scale = raw pseudobulk counts (same as atlas posterior_predict).
predictive_draws_raw_for_gene <- function(
  gene_sym,
  se,
  comparison_tbl,
  sample_ann,
  ndraws = 400L
) {
  chunk_row <- sting_fit_rows |>
    dplyr::filter(.data$gene == .env$gene_sym) |>
    dplyr::slice(1)

  if (!nrow(chunk_row)) {
    stop("No fit found for gene: ", gene_sym)
  }

  fit <- poco::reconstruct_brmsfit(chunk_row$brms_fit_compressed[[1]])
  feature_id <- chunk_row$.feature

  purrr::pmap_dfr(
    comparison_tbl |>
      dplyr::filter(.data$sample_role == "savi_pseudobulk") |>
      dplyr::select("sample", "sample_idx"),
    function(sample, sample_idx) {
      newdata <- newdata_for_sample(se, sample_idx)
      nd_grid <- build_newdata_grid(
        fit,
        newdata,
        new_level_tag = paste0("__savi__", as.character(sample))
      )

      y_rep_mat <- brms::posterior_predict(
        fit,
        newdata = nd_grid,
        summary = FALSE,
        re_formula = NULL,
        allow_new_levels = TRUE,
        sample_new_levels = "gaussian",
        ndraws = ndraws
      )
      y_rep <- marginalize_predict_draws(y_rep_mat)

      obs_raw <- SummarizedExperiment::assay(
        se[feature_id, sample_idx, drop = FALSE],
        "counts"
      ) |>
        as.numeric()

      tibble::tibble(
        gene = gene_sym,
        sample = as.character(sample),
        draw = y_rep,
        observed_count = obs_raw
      )
    }
  ) |>
    dplyr::left_join(
      sample_ann |>
        dplyr::select("sample", "Sample", "Category"),
      by = "sample"
    )
}

contrast_draws <- purrr::map_dfr(
  falsify_contrast_genes,
  predictive_draws_raw_for_gene,
  se = se_monocytic,
  comparison_tbl = comparison_samples_tbl,
  sample_ann = sample_annotations
) |>
  dplyr::mutate(
    Category = factor(.data$Category, levels = names(category_cols)),
    gene = factor(.data$gene, levels = falsify_contrast_genes),
    draw_log1p = log1p(pmax(.data$draw, 0)),
    observed_log1p = log1p(pmax(.data$observed_count, 0))
  )

contrast_draws |>
  dplyr::count(.data$gene, .data$Category, name = "n_draws")
```

```{r}
#| label: plot-il1b-density-contrast
#| fig-width: 11
#| fig-height: 5.5
#| fig-align: center

ggplot2::ggplot(
  contrast_draws,
  ggplot2::aes(x = draw_log1p, colour = Category, fill = Category)
) +
  ggplot2::geom_density(alpha = 0.25, linewidth = 0.9) +
  ggplot2::geom_vline(
    data = contrast_draws |>
      dplyr::distinct(.data$gene, .data$sample, .data$Category, .data$observed_log1p),
    ggplot2::aes(
      xintercept = observed_log1p,
      colour = Category
    ),
    linetype = "solid",
    linewidth = 0.7
  ) +
  ggplot2::facet_wrap(~gene, scales = "free_x") +
  ggplot2::scale_colour_manual(values = category_cols, name = "Category") +
  ggplot2::scale_fill_manual(values = category_cols, guide = "none") +
  ggplot2::labs(
    x = expression(log[1 + p]("raw pseudobulk count")),
    y = "Density",
    title = "Predictive null vs observed (log1p raw counts)",
    subtitle = paste(
      "Curves = atlas posterior_predict draws pooled by Category (model scale).",
      "Vertical lines = each donor's observed raw count, coloured by Category."
    )
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r}
#| label: plot-il1b-density-facet-sample
#| fig-width: 12
#| fig-height: 8
#| fig-align: center

il1b_facet_lookup <- il1b_sample_tbl |>
  dplyr::select(dplyr::any_of(c(
    "sample",
    "Sample",
    "Category",
    "direction",
    "p_above_healthy"
  ))) |>
  dplyr::mutate(
    direction_short = dplyr::case_when(
      .data$direction == "higher_than_healthy" ~ "higher",
      .data$direction == "lower_than_healthy" ~ "lower",
      TRUE ~ "consistent"
    ),
    sample_label = paste0(
      .data$Sample,
      " (",
      .data$Category,
      " · ",
      .data$direction_short,
      ", p=",
      sprintf("%.2f", .data$p_above_healthy),
      ")"
    )
  )

il1b_label_levels <- il1b_facet_lookup |>
  dplyr::arrange(.data$Category, .data$Sample)

il1b_facet_lookup <- il1b_facet_lookup |>
  dplyr::mutate(
    sample_label = factor(
      .data$sample_label,
      levels = il1b_label_levels$sample_label
    )
  )

il1b_draws <- contrast_draws |>
  dplyr::filter(.data$gene == "IL1B") |>
  dplyr::left_join(
    il1b_facet_lookup |>
      dplyr::select(.data$sample, .data$sample_label),
    by = "sample"
  )

ggplot2::ggplot(
  il1b_draws,
  ggplot2::aes(x = draw_log1p, colour = Category)
) +
  ggplot2::geom_density(fill = "grey85", colour = "grey40", linewidth = 0.6, alpha = 0.9) +
  ggplot2::geom_vline(
    ggplot2::aes(xintercept = observed_log1p, colour = Category),
    linewidth = 1
  ) +
  ggplot2::facet_wrap(~sample_label, scales = "free_x", ncol = 4) +
  ggplot2::scale_colour_manual(values = category_cols, name = "Category") +
  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(linewidth = 1.2))) +
  ggplot2::labs(
    x = expression(log[1 + p]("raw pseudobulk count")),
    y = "Density",
    title = expression(italic("IL1B") * ": posterior predictive check per donor"),
    subtitle = paste(
      "Facet = donor (Category · atlas tail, p = P(draw >= obs) on raw counts).",
      "Grey density = atlas healthy null; coloured vertical line = observed raw count."
    )
  ) +
  ggplot2::theme_minimal(base_size = 10) +
  ggplot2::theme(
    legend.position = "top",
    strip.text = ggplot2::element_text(size = 8)
  )
```

```{r}
#| label: il1b-falsify-summary
il1b_ctrl_vs_savi <- il1b_sample_tbl |>
  dplyr::group_by(.data$Category) |>
  dplyr::summarise(
    n = dplyr::n(),
    mean_obs = mean(.data$observed_count),
    mean_log2fc = mean(.data$log2fc_vs_median),
    pct_above_hpd = 100 * mean(.data$above_hpd),
    pct_higher_direction = 100 * mean(.data$direction == "higher_than_healthy"),
    mean_percentile = 100 * mean(.data$obs_percentile),
    .groups = "drop"
  )

ctrl_mean_log2fc <- il1b_ctrl_vs_savi |>
  dplyr::filter(.data$Category == "CTRL") |>
  dplyr::pull(.data$mean_log2fc)

ctrl_pct_above_hpd <- il1b_ctrl_vs_savi |>
  dplyr::filter(.data$Category == "CTRL") |>
  dplyr::pull(.data$pct_above_hpd)

savi_pct_above_hpd <- il1b_ctrl_vs_savi |>
  dplyr::filter(.data$Category == "SAVI") |>
  dplyr::pull(.data$pct_above_hpd)

paper_delta <- il1b_ctrl_vs_savi |>
  dplyr::filter(.data$Category %in% c("SAVI", "CTRL")) |>
  tidyr::pivot_wider(names_from = "Category", values_from = "mean_obs") |>
  dplyr::mutate(delta = .data$SAVI - .data$CTRL) |>
  dplyr::pull(.data$delta)

atlas_delta <- biomarker_quality |>
  dplyr::filter(.data$gene == "IL1B") |>
  dplyr::pull(.data$savi_vs_ctrl)

tibble::tibble(
  gene = "IL1B",
  paper_mean_savi_minus_ctrl_raw_counts = paper_delta,
  atlas_mean_log2fc_savi_minus_ctrl = atlas_delta,
  ctrl_pct_above_healthy_hpd = ctrl_pct_above_hpd,
  savi_pct_above_healthy_hpd = savi_pct_above_hpd,
  falsification_verdict = dplyr::if_else(
    is.finite(atlas_delta) &&
      atlas_delta < 0 &&
      is.finite(ctrl_mean_log2fc) &&
      ctrl_mean_log2fc > 0,
    "Atlas falsified: controls high, SAVI not elevated after correction",
    "Review manually"
  )
)

il1b_ctrl_vs_savi
```

**How to read this as falsification evidence**

- **Per-donor plots** (HPD, raw/scaled abundance, density facets): each point/line is one pseudobulk library; HPD and density overlays use **raw counts** on the model scale (`posterior_predict` draws and observed `counts` assay). The two-panel abundance plot shows the same donors on **raw** and **TMM-scaled** count scales (log1p y-axis); atlas percentile is unchanged under linear scaling.
- **Bar chart (`plot-il1b-two-nulls`)**: height = **mean within the GSE226598 cohort** for donors in that `Category` only (not vs atlas on the two right panels). Left panel = mean log2(obs / atlas predictive median); centre = mean raw pseudobulk count; right = mean log1p(TMM-scaled count).
- If **atlas falsification holds**: per-donor plots show **CTRL** above the atlas HPD and high `obs_percentile` in the table above, while **SAVI** is not systematically above CTRL on the **left bar panel** (mean log2 FC).
- The **centre bar panel** can still show **SAVI > CTRL** on raw within-cohort means — the contrast the original paper could see with seven controls.
- The **summary table** quantifies: `paper_mean_savi_minus_ctrl_raw_counts` can stay **positive** while `atlas_mean_log2fc_savi_minus_ctrl` is **negative**.

------------------------------------------------------------------------

# Case study: falsifying *ADRB2* (one-donor SAVI signal)

**ADRB2** is labelled **`falsified_no_savi_signal`** in the biomarker screen because only **20%** of SAVI donors pass the strict atlas tail call (`higher_than_healthy`), yet the **cohort mean** `log2fc_vs_median` in SAVI looks high. That pattern usually means **one outlier donor** pulls the category mean while the other SAVI pseudobulks sit near the healthy predictive null.

We audit that claim with per-donor tables, leave-one-out SAVI means, and density overlays on **raw pseudobulk counts** (same scale as `posterior_predict`; contrast gene **TSC22D2** = validated atlas behaviour).

```{r}
#| label: adrb2-falsify-helpers

adrb2_ensembl <- sting_signature_21 |>
  dplyr::filter(.data$gene == "ADRB2") |>
  dplyr::pull(.feature)

adrb2_scaled_obs <- tibble::tibble(sample = colnames(se_monocytic)) |>
  dplyr::mutate(
    sample_idx = match(.data$sample, colnames(se_monocytic)),
    observed_scaled = as.numeric(
      SummarizedExperiment::assay(
        se_monocytic[adrb2_ensembl, .data$sample_idx, drop = FALSE],
        scaled_counts_assay_name(se_monocytic)
      )
    )
  )

adrb2_sample_tbl <- gene_deviation_eval |>
  dplyr::filter(.data$gene == "ADRB2", !is.na(.data$Category)) |>
  dplyr::left_join(
    sample_annotations |>
      dplyr::select("sample", tidyselect::any_of("Treatment")),
    by = "sample"
  ) |>
  dplyr::left_join(adrb2_scaled_obs, by = "sample") |>
  dplyr::mutate(
    Category = factor(.data$Category, levels = names(category_cols)),
    scale_ratio = .data$observed_scaled / pmax(.data$observed_count, 1e-9),
    healthy_median_scaled = .data$healthy_median * .data$scale_ratio,
    healthy_hpd_lo_scaled = .data$healthy_hpd_lo * .data$scale_ratio,
    healthy_hpd_hi_scaled = .data$healthy_hpd_hi * .data$scale_ratio,
    above_hpd = .data$observed_count > .data$healthy_hpd_hi,
    below_hpd = .data$observed_count < .data$healthy_hpd_lo,
    in_hpd = !.data$above_hpd & !.data$below_hpd
  ) |>
  dplyr::arrange(.data$Category, .data$Sample)

adrb2_savi_tbl <- adrb2_sample_tbl |>
  dplyr::filter(.data$Category == "SAVI")

adrb2_savi_outlier_pick <- adrb2_savi_tbl |>
  dplyr::filter(.data$direction == "higher_than_healthy") |>
  dplyr::slice(1)

if (nrow(adrb2_savi_outlier_pick) == 0L) {
  adrb2_savi_outlier_pick <- adrb2_savi_tbl |>
    dplyr::slice_max(.data$log2fc_vs_median, n = 1, with_ties = FALSE)
}

adrb2_label_levels <- adrb2_sample_tbl |>
  dplyr::arrange(.data$Category, .data$Sample)

adrb2_sample_tbl <- adrb2_sample_tbl |>
  dplyr::mutate(
    savi_outlier = .data$sample %in% adrb2_savi_outlier_pick$sample,
    direction_short = dplyr::case_when(
      .data$direction == "higher_than_healthy" ~ "higher",
      .data$direction == "lower_than_healthy" ~ "lower",
      TRUE ~ "consistent"
    ),
    sample_facet_label = paste0(
      .data$Sample,
      " (",
      .data$Category,
      " · ",
      .data$direction_short,
      ", p=",
      sprintf("%.2f", .data$p_above_healthy),
      ")"
    ),
    sample_label = factor(
      .data$sample_facet_label,
      levels = adrb2_label_levels |>
        dplyr::mutate(
          lab = paste0(
            .data$Sample,
            " (",
            .data$Category,
            " · ",
            dplyr::case_when(
              .data$direction == "higher_than_healthy" ~ "higher",
              .data$direction == "lower_than_healthy" ~ "lower",
              TRUE ~ "consistent"
            ),
            ", p=",
            sprintf("%.2f", .data$p_above_healthy),
            ")"
          )
        ) |>
        dplyr::pull(.data$lab)
    )
  )

adrb2_savi_outlier <- adrb2_sample_tbl |>
  dplyr::filter(.data$savi_outlier)

adrb2_loo_summary <- tibble::tibble(
  contrast = c(
    "All SAVI donors",
    "Drop atlas-tail SAVI donor",
    "Other SAVI donors (median)"
  ),
  n_savi = c(
    nrow(adrb2_savi_tbl),
    nrow(adrb2_savi_tbl) - 1L,
    nrow(adrb2_savi_tbl) - 1L
  ),
  mean_log2fc = c(
    mean(adrb2_savi_tbl$log2fc_vs_median),
    mean(
      adrb2_savi_tbl$log2fc_vs_median[
        adrb2_savi_tbl$sample != adrb2_savi_outlier$sample
      ]
    ),
    stats::median(
      adrb2_savi_tbl$log2fc_vs_median[
        adrb2_savi_tbl$sample != adrb2_savi_outlier$sample
      ]
    )
  ),
  pct_higher = c(
    100 * mean(adrb2_savi_tbl$direction == "higher_than_healthy"),
    100 * mean(
      adrb2_savi_tbl$direction[
        adrb2_savi_tbl$sample != adrb2_savi_outlier$sample
      ] == "higher_than_healthy"
    ),
    NA_real_
  )
)

adrb2_per_donor <- adrb2_sample_tbl |>
  dplyr::select(
    Sample,
    Category,
    direction,
    direction_short,
    p_above_healthy,
    log2fc_vs_median,
    obs_percentile,
    observed_count,
    observed_scaled,
    savi_outlier,
    sample_facet_label
  )

adrb2_per_donor

adrb2_loo_summary

tibble::tibble(
  outlier_sample = adrb2_savi_outlier$Sample,
  outlier_direction = adrb2_savi_outlier$direction,
  outlier_log2fc = adrb2_savi_outlier$log2fc_vs_median,
  n_savi_higher_than_healthy = sum(adrb2_savi_tbl$direction == "higher_than_healthy"),
  n_savi_consistent = sum(adrb2_savi_tbl$direction == "consistent_with_healthy"),
  biomarker_class = biomarker_quality |>
    dplyr::filter(.data$gene == "ADRB2") |>
    dplyr::pull(.data$biomarker_class)
)
```

```{r}
#| label: plot-adrb2-log2fc-by-sample
#| fig-width: 11
#| fig-height: 7
#| fig-align: center

direction_cols <- c(
  higher_than_healthy = "#E41A1C",
  lower_than_healthy = "#377EB8",
  consistent_with_healthy = "grey65"
)

ggplot2::ggplot(
  adrb2_sample_tbl,
  ggplot2::aes(
    x = sample_label,
    y = log2fc_vs_median,
    colour = direction
  )
) +
  ggplot2::geom_hline(yintercept = 0, linetype = 2, colour = "grey55") +
  ggplot2::geom_point(
    ggplot2::aes(size = savi_outlier),
    alpha = 0.9
  ) +
  ggplot2::geom_point(
    data = adrb2_savi_outlier,
    ggplot2::aes(x = sample_label, y = log2fc_vs_median),
    shape = 21,
    fill = NA,
    colour = "black",
    stroke = 1.2,
    size = 5,
    inherit.aes = FALSE
  ) +
  ggplot2::scale_colour_manual(values = direction_cols, name = "Atlas tail") +
  ggplot2::scale_size_manual(
    values = c(`TRUE` = 5, `FALSE` = 3),
    guide = "none"
  ) +
  ggplot2::labs(
    x = NULL,
    y = expression(log[2]("obs / healthy predictive median")),
    title = expression(italic("ADRB2") * ": per-donor atlas deviation"),
    subtitle = paste(
      "Ring = sole SAVI donor with atlas tail higher_than_healthy",
      paste0("(", adrb2_savi_outlier$Sample, ", p=", sprintf("%.3f", adrb2_savi_outlier$p_above_healthy), ")."),
      "x labels: direction and P(draw >= obs); visual offset ≠ tail call."
    )
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, size = 8),
    legend.position = "top"
  )
```

```{r}
#| label: plot-adrb2-leave-one-out
#| fig-width: 8
#| fig-height: 4.5
#| fig-align: center

adrb2_loo_plot <- adrb2_loo_summary |>
  dplyr::mutate(
    contrast = factor(
      .data$contrast,
      levels = .data$contrast
    )
  )

ggplot2::ggplot(
  adrb2_loo_plot,
  ggplot2::aes(x = contrast, y = mean_log2fc, fill = contrast)
) +
  ggplot2::geom_col(width = 0.65, colour = "white") +
  ggplot2::geom_hline(yintercept = 0, linetype = 2, colour = "grey45") +
  ggplot2::scale_fill_manual(values = c("#E41A1C", "#FC8D59", "#FDBB84"), guide = "none") +
  ggplot2::labs(
    x = NULL,
    y = expression("Mean / median " * log[2] * " FC (SAVI only)"),
    title = expression(italic("ADRB2") * ": SAVI cohort mean driven by one donor"),
    subtitle = paste0(
      "Dropping ",
      adrb2_savi_outlier$Sample,
      " collapses the SAVI mean toward 0; remaining donors do not support a cohort-wide atlas tail."
    )
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 25, hjust = 1))
```

```{r}
#| label: plot-adrb2-obs-vs-hpd
#| fig-width: 14
#| fig-height: 6
#| fig-align: center

log1p_trans <- scales::trans_new(
  name = "log1p",
  transform = function(x) log1p(pmax(x, 0)),
  inverse = function(x) expm1(x),
  domain = c(0, Inf)
)

adrb2_hpd_plot <- adrb2_sample_tbl |>
  dplyr::transmute(
    sample_label,
    Category,
    savi_outlier,
    measure = "Raw pseudobulk counts (atlas null)",
    hpd_lo = .data$healthy_hpd_lo,
    hpd_hi = .data$healthy_hpd_hi,
    pred_median = .data$healthy_median,
    abundance = .data$observed_count
  ) |>
  dplyr::bind_rows(
    adrb2_sample_tbl |>
      dplyr::transmute(
        sample_label,
        Category,
        savi_outlier,
        measure = "TMM-scaled counts (linear rescale of null)",
        hpd_lo = .data$healthy_hpd_lo_scaled,
        hpd_hi = .data$healthy_hpd_hi_scaled,
        pred_median = .data$healthy_median_scaled,
        abundance = .data$observed_scaled
      )
  ) |>
  dplyr::mutate(
    measure = factor(
      .data$measure,
      levels = c(
        "Raw pseudobulk counts (atlas null)",
        "TMM-scaled counts (linear rescale of null)"
      )
    )
  )

ggplot2::ggplot(adrb2_hpd_plot) +
  ggplot2::geom_linerange(
    ggplot2::aes(
      x = sample_label,
      ymin = hpd_lo,
      ymax = hpd_hi
    ),
    colour = "grey45",
    linewidth = 0.7
  ) +
  ggplot2::geom_point(
    ggplot2::aes(x = sample_label, y = pred_median),
    shape = 4,
    size = 2.2,
    colour = "grey55",
    stroke = 0.9
  ) +
  ggplot2::geom_point(
    ggplot2::aes(
      x = sample_label,
      y = abundance,
      colour = Category,
      size = savi_outlier
    ),
    alpha = 0.9
  ) +
  ggplot2::facet_wrap(~measure, scales = "free_y", ncol = 2) +
  ggplot2::scale_y_continuous(trans = log1p_trans) +
  ggplot2::scale_colour_manual(values = category_cols) +
  ggplot2::scale_size_manual(values = c(`TRUE` = 4.5, `FALSE` = 3), guide = "none") +
  ggplot2::labs(
    x = NULL,
    y = expression(log[1 + p]("count")),
    title = expression(italic("ADRB2") * ": observed vs atlas healthy predictive interval"),
    subtitle = paste(
      "Left = raw counts (same scale as tail test and density plots).",
      "Right = TMM-scaled abundance with HPD rescaled by sample library size;",
      "atlas percentile unchanged. Larger point = SAVI atlas-tail outlier."
    )
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, size = 7),
    legend.position = "top",
    strip.text = ggplot2::element_text(size = 9)
  )
```

```{r}
#| label: plot-adrb2-savi-only
#| fig-width: 7
#| fig-height: 4.5
#| fig-align: center

ggplot2::ggplot(
  adrb2_savi_tbl |>
    dplyr::mutate(
      highlight = .data$sample %in% adrb2_savi_outlier$sample,
      Sample = factor(.data$Sample, levels = adrb2_savi_tbl$Sample)
    ),
  ggplot2::aes(x = Sample, y = log2fc_vs_median, colour = direction)
) +
  ggplot2::geom_hline(yintercept = 0, linetype = 2, colour = "grey55") +
  ggplot2::geom_segment(
    ggplot2::aes(x = Sample, xend = Sample, y = 0, yend = log2fc_vs_median),
    colour = "grey75",
    linewidth = 0.6
  ) +
  ggplot2::geom_point(
    ggplot2::aes(size = highlight),
    alpha = 0.95
  ) +
  ggplot2::scale_colour_manual(values = direction_cols, name = "Atlas tail") +
  ggplot2::scale_size_manual(values = c(`TRUE` = 5, `FALSE` = 3), guide = "none") +
  ggplot2::labs(
    x = NULL,
    y = expression(log[2]("obs / healthy predictive median")),
    title = expression(italic("ADRB2") * ": SAVI donors only"),
    subtitle = paste0(
      nrow(adrb2_savi_tbl) - sum(adrb2_savi_tbl$direction == "higher_than_healthy"),
      " of ",
      nrow(adrb2_savi_tbl),
      " SAVI donors are not higher_than_healthy vs atlas."
    )
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r}
#| label: adrb2-predictive-draws
#| cache: true
#| cache.lazy: false
#| dependson: ["gene-deviation-fn", "il1b-predictive-draws"]

# Uses predictive_draws_raw_for_gene() from il1b-predictive-draws (raw model scale).
adrb2_contrast_genes <- c("ADRB2", "TSC22D2")

adrb2_draws <- purrr::map_dfr(
  adrb2_contrast_genes,
  predictive_draws_raw_for_gene,
  se = se_monocytic,
  comparison_tbl = comparison_samples_tbl,
  sample_ann = sample_annotations
) |>
  dplyr::mutate(
    Category = factor(.data$Category, levels = names(category_cols)),
    gene = factor(.data$gene, levels = adrb2_contrast_genes),
    draw_log1p = log1p(pmax(.data$draw, 0)),
    observed_log1p = log1p(pmax(.data$observed_count, 0)),
    savi_outlier = .data$sample %in% adrb2_savi_outlier$sample
  )

adrb2_draws |>
  dplyr::count(.data$gene, .data$Category, name = "n_draws")
```

```{r}
#| label: plot-adrb2-density-contrast
#| fig-width: 11
#| fig-height: 5.5
#| fig-align: center

ggplot2::ggplot(
  adrb2_draws,
  ggplot2::aes(x = draw_log1p, colour = Category, fill = Category)
) +
  ggplot2::geom_density(alpha = 0.25, linewidth = 0.9) +
  ggplot2::geom_vline(
    data = adrb2_draws |>
      dplyr::distinct(.data$gene, .data$sample, .data$Category, .data$observed_log1p, .data$savi_outlier),
    ggplot2::aes(
      xintercept = observed_log1p,
      colour = Category,
      linewidth = savi_outlier
    ),
    linetype = "solid"
  ) +
  ggplot2::scale_linewidth_manual(values = c(`TRUE` = 1.2, `FALSE` = 0.65), guide = "none") +
  ggplot2::facet_wrap(~gene, scales = "free_x") +
  ggplot2::scale_colour_manual(values = category_cols, name = "Category") +
  ggplot2::scale_fill_manual(values = category_cols, guide = "none") +
  ggplot2::labs(
    x = expression(log[1 + p]("raw pseudobulk count")),
    y = "Density",
    title = expression(italic("ADRB2") * " vs validated " * italic("TSC22D2")),
    subtitle = "Grey = atlas posterior_predict (raw counts); vertical lines = observed raw count (log1p)."
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r}
#| label: plot-adrb2-density-facet-sample
#| fig-width: 12
#| fig-height: 8
#| fig-align: center

adrb2_facet_lookup <- adrb2_sample_tbl |>
  dplyr::select(dplyr::any_of(c(
    "sample",
    "sample_label",
    "Sample",
    "Category",
    "direction",
    "p_above_healthy"
  )))

if (!"sample_label" %in% names(adrb2_facet_lookup) ||
    !any(grepl(", p=", adrb2_facet_lookup$sample_label, fixed = TRUE), na.rm = TRUE)) {
  adrb2_facet_lookup <- adrb2_facet_lookup |>
    dplyr::mutate(
      direction_short = dplyr::case_when(
        .data$direction == "higher_than_healthy" ~ "higher",
        .data$direction == "lower_than_healthy" ~ "lower",
        TRUE ~ "consistent"
      ),
      sample_label = paste0(
        .data$Sample,
        " (",
        .data$Category,
        " · ",
        .data$direction_short,
        ", p=",
        sprintf("%.2f", .data$p_above_healthy),
        ")"
      )
    )
}

if ("sample_label" %in% names(adrb2_sample_tbl) && is.factor(adrb2_sample_tbl$sample_label)) {
  adrb2_facet_lookup$sample_label <- factor(
    adrb2_facet_lookup$sample_label,
    levels = levels(adrb2_sample_tbl$sample_label)
  )
}

adrb2_draws_gene <- adrb2_draws |>
  dplyr::filter(.data$gene == "ADRB2") |>
  dplyr::left_join(
    adrb2_facet_lookup |>
      dplyr::select(.data$sample, .data$sample_label),
    by = "sample"
  )

ggplot2::ggplot(
  adrb2_draws_gene,
  ggplot2::aes(x = draw_log1p, colour = Category)
) +
  ggplot2::geom_density(fill = "grey85", colour = "grey40", linewidth = 0.6, alpha = 0.9) +
  ggplot2::geom_vline(
    ggplot2::aes(
      xintercept = observed_log1p,
      colour = Category,
      linewidth = savi_outlier
    ),
    linetype = "solid"
  ) +
  ggplot2::scale_linewidth_manual(values = c(`TRUE` = 1.3, `FALSE` = 0.8), guide = "none") +
  ggplot2::facet_wrap(~sample_label, scales = "free_x", ncol = 4) +
  ggplot2::scale_colour_manual(values = category_cols, name = "Category") +
  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(linewidth = 1.2))) +
  ggplot2::labs(
    x = expression(log[1 + p]("raw pseudobulk count")),
    y = "Density",
    title = expression(italic("ADRB2") * ": posterior predictive check per donor"),
    subtitle = paste(
      "Facet = donor (Category · atlas tail, p = P(draw >= obs) on raw counts).",
      "Thicker line =",
      adrb2_savi_outlier$Sample,
      "(only SAVI donor with p < 0.025)."
    )
  ) +
  ggplot2::theme_minimal(base_size = 10) +
  ggplot2::theme(
    legend.position = "top",
    strip.text = ggplot2::element_text(size = 6.5)
  )
```

```{r}
#| label: adrb2-falsify-summary

adrb2_biomarker_row <- biomarker_quality |>
  dplyr::filter(.data$gene == "ADRB2")

tibble::tibble(
  gene = "ADRB2",
  biomarker_class = adrb2_biomarker_row$biomarker_class,
  pct_higher_SAVI = adrb2_biomarker_row$pct_higher.SAVI,
  mean_log2fc_SAVI = adrb2_biomarker_row$savi_lift,
  savi_vs_ctrl_atlas = adrb2_biomarker_row$savi_vs_ctrl,
  outlier_sample = adrb2_savi_outlier$Sample,
  outlier_log2fc = adrb2_savi_outlier$log2fc_vs_median,
  n_savi_higher = sum(adrb2_savi_tbl$direction == "higher_than_healthy"),
  n_savi_consistent = sum(adrb2_savi_tbl$direction == "consistent_with_healthy"),
  mean_log2fc_SAVI_drop_outlier = adrb2_loo_summary$mean_log2fc[2],
  falsification_verdict = paste(
    "Cohort SAVI mean is not a robust signature signal:",
    "only",
    sum(adrb2_savi_tbl$direction == "higher_than_healthy"),
    "of",
    nrow(adrb2_savi_tbl),
    "SAVI donors exceed the atlas null;",
    "effect driven by",
    adrb2_savi_outlier$Sample
  )
)
```

**How to read**

- **`plot-adrb2-log2fc-by-sample`** and **`plot-adrb2-density-facet-sample`**: same **raw-count** scale as the atlas model; facet labels show **atlas tail** and **`p = P(draw >= obs)`**. Only **P6_STING_ht** has **p < 0.025** (thicker line).
- **`plot-adrb2-leave-one-out`**: removing that donor drops the SAVI mean `log2fc` toward zero (median of the remaining four is lower still).
- **Density panels**: pooled SAVI curve for *ADRB2* can look elevated because draws are stacked across donors; per-donor facets show the mismatch between one extreme line and four lines near the grey null.
- **Single-cell UMAP** (below, with *IL1B*): SCT-normalized *ADRB2* on disease-associated monocytes, split by **`Category`**, to see whether the atlas outlier is visible at cell level or confined to one donor's monocytes.

# Single-cell expression on UMAP (GEO Seurat object)

Published scRNA-seq object [GSE226598](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE226598) (`prepare_data.R`). We subset **`params$savi_sc_celltype`** (disease-associated monocytes — same population as the pseudobulk merge) and plot signature genes on the embedded UMAP, **`split.by = "Category"`** (`CTRL`, `SAVI`, `SAVI_treated`), using the **`SCT`** assay **`data`** slot (SCTransform–normalized expression from the published object).

```{r}
#| label: load-savi-seurat
#| cache: false

read_geo_rds_gz <- function(path) {
  con <- gzcon(gzfile(path))
  on.exit(close(con), add = TRUE)
  readRDS(con)
}

seurat_savi <- read_geo_rds_gz(params$savi_seurat_gz)

if (!"Category" %in% colnames(seurat_savi@meta.data)) {
  stop("`Category` not found in Seurat metadata.")
}

seurat_savi$Category <- factor(
  seurat_savi$Category,
  levels = c("CTRL", "SAVI", "SAVI_treated")
)

seurat_reduction <- function(se) {
  nms <- names(se@reductions)
  hit <- intersect(c("umap", "UMAP", "tsne", "TSNE"), nms)
  if (length(hit)) {
    return(hit[[1]])
  }
  nms[[1]]
}

reduction_use <- seurat_reduction(seurat_savi)

# Subset early so we don't keep the full >2k x 24k object around.
seurat_savi_mono <- seurat_savi |>
  subset(CellType == params$savi_sc_celltype)
# rm(seurat_savi)
# gc(verbose = FALSE)

tibble::tibble(
  n_cells = ncol(seurat_savi_mono),
  reduction = reduction_use,
  categories = paste(levels(seurat_savi_mono$Category), collapse = ", "),
  il1b_present = "IL1B" %in% rownames(seurat_savi_mono),
  adrb2_present = "ADRB2" %in% rownames(seurat_savi_mono)
)
```

```{r}
#| label: subset-savi-monocytes
if (ncol(seurat_savi_mono) == 0L) {
  stop(
    "No cells for CellType = '", params$savi_sc_celltype,
    "'. Check `CellType` levels in the Seurat object."
  )
}

seurat_savi_mono$Category <- factor(
  seurat_savi_mono$Category,
  levels = c("CTRL", "SAVI", "SAVI_treated")
)

tibble::tibble(
  cell_type = params$savi_sc_celltype,
  n_cells = ncol(seurat_savi_mono)
)

table(seurat_savi_mono$Category, useNA = "ifany")
```

```{r}
#| label: plot-il1b-featureplot-category
#| fig-width: 12
#| fig-height: 4
#| fig-align: center

if (!"SCT" %in% SeuratObject::Assays(seurat_savi_mono)) {
  stop(
    "SCT assay not found. Available: ",
    paste(SeuratObject::Assays(seurat_savi_mono), collapse = ", ")
  )
}

if (!"IL1B" %in% rownames(seurat_savi_mono[["SCT"]])) {
  stop("IL1B not found in SCT assay rownames.")
}

Seurat::FeaturePlot(
  seurat_savi_mono,
  features = "IL1B",
  assay = "SCT",
  split.by = "Category",
  reduction = reduction_use,
  slot = "data",
  ncol = 3,
  order = TRUE,
  pt.size = 0.35,
  min.cutoff = "q10",
  max.cutoff = "q90"
) &
  ggplot2::theme(
    plot.title = ggplot2::element_text(size = 11),
    strip.text = ggplot2::element_text(size = 11, face = "bold")
  )
```

Panel titles are **`Category`** (study cohort label from GEO metadata). Expression is from **SCTransform** (`assay = "SCT"`, `slot = "data"`), not raw **`RNA`** counts. High **`IL1B`** in **CTRL** monocytes complements the pseudobulk / atlas falsification checks above.

#### *ADRB2* on UMAP (all donors)

Same layout as *IL1B*: disease-associated monocytes, SCT **`data`**, **`split.by = "Category"`**. Compare the **SAVI** column to **CTRL** and **SAVI_treated** — cells from all donors in each arm are pooled on the UMAP.

```{r}
#| label: plot-adrb2-featureplot-category
#| fig-width: 12
#| fig-height: 4
#| fig-align: center

if (!"ADRB2" %in% rownames(seurat_savi_mono[["SCT"]])) {
  stop("ADRB2 not found in SCT assay rownames.")
}

Seurat::FeaturePlot(
  seurat_savi_mono,
  features = "ADRB2",
  assay = "SCT",
  split.by = "Category",
  reduction = reduction_use,
  slot = "data",
  ncol = 3,
  order = TRUE,
  pt.size = 0.35,
  min.cutoff = "q10",
  max.cutoff = "q90"
) &
  ggplot2::theme(
    plot.title = ggplot2::element_text(size = 11),
    strip.text = ggplot2::element_text(size = 11, face = "bold")
  )
```

The **SAVI** panel above pools monocytes from all five untreated SAVI donors. Pseudobulk checks show atlas deviation is dominated by **`P6_STING_ht`** (4057 raw counts, **`p_above_healthy < 0.025`**); the other four SAVI donors sit near the healthy null (≤65 raw counts). If **`ADRB2`** signal in **SAVI** is sparse overall but driven by cells from that donor, the pooled UMAP can still look elevated.

#### *ADRB2* after removing the atlas-tail donor

Mirroring **`plot-adrb2-leave-one-out`** on pseudobulk, we repeat the UMAP with **every disease-associated monocyte from the atlas-tail SAVI donor removed** (`adrb2_savi_outlier$Sample`, here **`P6_STING_ht`**). Assay, cell type, and **`split.by = "Category"`** are unchanged; only that library’s cells are dropped, so the **SAVI** facet reflects the remaining four untreated SAVI donors. If the **SAVI** panel now resembles **CTRL** / **SAVI_treated**, the cohort-level **`ADRB2`** signal is a **one-donor** artefact at single-cell resolution as well as in pseudobulk.

```{r}
#| label: plot-adrb2-featureplot-category_outlier_removed
#| fig-width: 12
#| fig-height: 4
#| fig-align: center

if (!"ADRB2" %in% rownames(seurat_savi_mono[["SCT"]])) {
  stop("ADRB2 not found in SCT assay rownames.")
}

if (!"Sample" %in% colnames(seurat_savi_mono@meta.data)) {
  stop("`Sample` not found in Seurat metadata.")
}

outlier_sample <- adrb2_savi_outlier$Sample

seurat_savi_mono_no_outlier <- subset(
  seurat_savi_mono,
  subset = Sample != outlier_sample
)

Seurat::FeaturePlot(
  seurat_savi_mono_no_outlier,
  features = "ADRB2",
  assay = "SCT",
  split.by = "Category",
  reduction = reduction_use,
  slot = "data",
  ncol = 3,
  order = TRUE,
  pt.size = 0.35,
  min.cutoff = "q10",
  max.cutoff = "q90"
) &
  ggplot2::theme(
    plot.title = ggplot2::element_text(size = 11),
    strip.text = ggplot2::element_text(size = 11, face = "bold")
  )
```

**Read together:** the all-donor and leave-one-out UMAPs ask the same question as the pseudobulk falsification — is **`ADRB2`** elevated across **SAVI**, or only when **`P6_STING_ht`** is included? Concordance between a muted **SAVI** panel after exclusion and the drop in SAVI mean **`log2fc`** in **`plot-adrb2-leave-one-out`** strengthens the **`falsified_no_savi_signal`** classification (20% of SAVI donors pass the strict atlas tail test).

------------------------------------------------------------------------

# Composition analysis vs original SAVI paper

In the original SAVI scRNA-seq paper (de Cevins *et al.* 2023; GSE226598), after QC, integration, clustering, and annotation, the authors report **23 cell populations** in the SAVI dataset and **29 cell populations** in the IFN-beta dataset. They then compare PBMC composition between SAVI and controls (Figure 1E), with p-values from Kruskal-Wallis tests followed by post hoc Dunn's multiple-comparison testing.

For whole-PBMC composition, the paper reports increased naive CD4 and naive CD8 T cells, decreased effector/memory CD4 and CD8 T-cell compartments, decreased NK cells (especially NK CD56bright), and strong depletion of MAIT and gamma-delta T cells in SAVI. They also report that cell-type proportions did not differ significantly before versus under JAK-inhibitor treatment.

In this notebook, we use those statements as the reference conclusion, then:

1. compute per-sample proportions from GEO **`CellType`** (23 populations),
2. map **`CellType` → Immune Atlas `L3`** labels used by the sccomp composition model,
3. compare observed proportions to healthy **`predict_res`** draws matched on **`sex`** and **`age_decade`**,
4. summarise trends in **`CTRL`**, **`SAVI`**, and **`SAVI_treated`** against the paper’s Figure 1E conclusions.

```{r}
#| label: summarize-paper-composition
paper_composition_summary <- tibble::tribble(
  ~topic, ~paper_finding,
  "Dataset size after integration", "23 populations (SAVI dataset) and 29 populations (IFN-beta dataset).",
  "Naive T-cell compartments", "Naive CD4 and naive CD8 T-cell proportions are increased in SAVI.",
  "Effector/memory T-cell compartments", "Effector/memory CD4 and CD8 compartments are decreased in SAVI.",
  "Innate-like / cytotoxic compartments", "NK cells (notably NK CD56bright), MAIT cells, and gamma-delta T cells are reduced in SAVI.",
  "Treatment comparison", "No significant composition differences were observed before versus under JAK inhibitor treatment.",
  "Statistics in Figure 1E", "Kruskal-Wallis tests with post hoc Dunn's multiple comparisons."
)

paper_composition_summary

paper_ct_alignment <- tibble::tribble(
  ~paper_compartment, ~direction_in_savi, ~celltype_labels_in_this_object, ~mapping_note,
  "Naive CD4 T", "increased", "0. Naive CD4", "Directly aligned.",
  "Effector/memory CD4", "decreased", "3. CM CD4; 4. EM CD4; 8. TE CD4; 9. Tregs and TE CD4", "Approximate effector/memory CD4 pool from available labels.",
  "Naive CD8 T", "increased", "2. Naive CD8", "Directly aligned.",
  "Effector/memory CD8", "decreased", "14. CM CD8; 15. EM CD8; 7. TE CD8", "Approximate effector/memory CD8 pool from available labels.",
  "NK (especially CD56bright)", "decreased", "10. NK; 26. NK CD56bright", "CD56bright is explicitly available.",
  "MAIT", "strongly decreased", "22. MAIT", "Directly aligned.",
  "gamma-delta T", "strongly decreased", "16. gd T cells", "Directly aligned.",
  "Myeloid context (not primary Figure 1E claim)", "context", "5. Classical monocytes; 12. Non-classical monocytes; 17. Disease-associated monocytes; 20. cDCs; 27. pDCs", "Contextual comparison only.",
  "B-cell context (not primary Figure 1E claim)", "context", "1. Naive B; 6. Transitionnal B; 13. Memory B; 11. CD3+ B cells; 25. Plasmablasts", "Contextual comparison only."
)

paper_ct_alignment
```

```{r}
#| label: composition-proportions-build
#| cache: false

if (!exists("read_geo_rds_gz", mode = "function")) {
  read_geo_rds_gz <- function(path) {
    con <- gzcon(gzfile(path))
    on.exit(close(con), add = TRUE)
    readRDS(con)
  }
}

if (!exists("seurat_savi", inherits = FALSE)) {
  seurat_savi <- read_geo_rds_gz(params$savi_seurat_gz)
}

if (!"Category" %in% colnames(seurat_savi@meta.data)) {
  stop("`Category` not found in Seurat metadata.")
}
if (!"CellType" %in% colnames(seurat_savi@meta.data)) {
  stop("`CellType` not found in Seurat metadata.")
}

seurat_savi$Category <- factor(
  seurat_savi$Category,
  levels = c("CTRL", "SAVI", "SAVI_treated")
)

# GEO sample metadata (same table as pseudobulk join above).
geo_tbl_comp <- geo_tbl |>
  dplyr::mutate(
    sex = dplyr::case_when(
      .data$gse_sex == "M" ~ "male",
      .data$gse_sex == "F" ~ "female",
      TRUE ~ NA_character_
    ),
    age_decade = as.character(ceiling(.data$gse_age_years / 10))
  ) |>
  dplyr::select("Sample", "sex", "age_decade", "gse_age_years")

# Fine-grained CellType -> Immune Atlas L3 (posteriorHCA / sccomp) labels.
celltype_to_l3 <- tibble::tribble(
  ~cell_type, ~L3,
  "0. Naive CD4", "cd4 naive",
  "2. Naive CD8", "cd8 naive",
  "3. CM CD4", "cd4 tcm",
  "4. EM CD4", "cd4 th1 em",
  "8. TE CD4", "cd4 th1 em",
  "9. Tregs and TE CD4", "treg",
  "14. CM CD8", "cd8 tcm",
  "15. EM CD8", "cd8 tem",
  "7. TE CD8", "cd8 tem",
  "10. NK", "nk",
  "26. NK CD56bright", "nk",
  "22. MAIT", "mait",
  "16. gd T cells", "tgd",
  "1. Naive B", "b naive",
  "6. Transitionnal B", "b naive",
  "13. Memory B", "b memory",
  "11. CD3+ B cells", "b memory",
  "5. Classical monocytes", "cd14 mono",
  "12. Non-classical monocytes", "cd16 mono",
  "17. Disease-associated monocytes", "cd14 mono",
  "20. cDCs", "cdc",
  "27. pDCs", "pdc",
  "25. Plasmablasts", "plasma"
)

cell_tbl <- seurat_savi@meta.data |>
  tibble::rownames_to_column("cell_id") |>
  dplyr::transmute(
    sample_id = as.character(.data$Sample),
    Category = as.character(.data$Category),
    cell_type = as.character(.data$CellType)
  ) |>
  dplyr::left_join(celltype_to_l3, by = "cell_type") |>
  dplyr::left_join(geo_tbl_comp, by = c("sample_id" = "Sample"))

example_proportions_celltype <- cell_tbl |>
  dplyr::filter(!is.na(.data$cell_type), nzchar(.data$cell_type)) |>
  dplyr::count(.data$sample_id, .data$cell_type, name = "n_cells") |>
  dplyr::group_by(.data$sample_id) |>
  dplyr::mutate(proportion = .data$n_cells / sum(.data$n_cells)) |>
  dplyr::ungroup() |>
  dplyr::select(.data$sample_id, .data$cell_type, .data$proportion)

# L3-level observed proportions (merge CellType labels that map to the same L3).
example_proportions <- cell_tbl |>
  dplyr::filter(!is.na(.data$L3)) |>
  dplyr::count(.data$sample_id, .data$Category, .data$sex, .data$age_decade, .data$L3, name = "n_cells") |>
  dplyr::group_by(.data$sample_id, .data$Category, .data$sex, .data$age_decade, .data$L3) |>
  dplyr::summarise(proportion = sum(.data$n_cells), .groups = "drop_last") |>
  dplyr::group_by(.data$sample_id) |>
  dplyr::mutate(proportion = .data$proportion / sum(.data$proportion)) |>
  dplyr::ungroup() |>
  dplyr::rename(cell_type = L3) |>
  dplyr::arrange(.data$sample_id, .data$cell_type)

unmapped_celltypes <- cell_tbl |>
  dplyr::filter(is.na(.data$L3)) |>
  dplyr::count(.data$cell_type, sort = TRUE)

tibble::tibble(
  label_column_used = "CellType",
  n_samples = dplyr::n_distinct(example_proportions$sample_id),
  n_cells_total = nrow(cell_tbl),
  n_celltypes_raw = dplyr::n_distinct(example_proportions_celltype$cell_type),
  n_l3_types = dplyr::n_distinct(example_proportions$cell_type),
  n_unmapped_celltypes = nrow(unmapped_celltypes)
)

if (nrow(unmapped_celltypes) > 0L) {
  unmapped_celltypes
}
```

### Healthy atlas composition null (`predict_res`)

Posterior draws of blood PBMC composition from the Immune Atlas sccomp model (`disease_groups___altered = Normal`), stratified by **`sex`** (`male` / `female`) and **`age_decade`** (`1`–`4`), matching SAVI donor metadata. The draw table is **not** knitr-cached (too large for R’s cache serializer); it is saved once to **`params$composition_predict_rds`** and trimmed to L3 types present in the SAVI cohort.

```{r}
#| label: composition-predict-healthy
#| cache: false

if (!requireNamespace("sccomp", quietly = TRUE)) {
  stop("Package `sccomp` is required for composition prediction.")
}

l3_for_comparison <- example_proportions |>
  dplyr::distinct(.data$cell_type) |>
  dplyr::pull(.data$cell_type)

trim_predict_res <- function(x, cell_types) {
  x |>
    dplyr::filter(.data$cell_type %in% cell_types) |>
    dplyr::select(dplyr::any_of(c(
      "sample_id",
      "age_decade",
      "sex",
      "cell_type",
      "proportion",
      ".draw"
    )))
}

if (file.exists(params$composition_predict_rds)) {
  predict_res <- readRDS(params$composition_predict_rds)
} else {
  sccomp_est <- readRDS(params$sccomp_estimates_rds)

  sample_metadata <- tidyr::expand_grid(
    age_decade = as.character(1:4),
    sex = c("male", "female")
  ) |>
    dplyr::mutate(
      assay_groups___altered = "10x Genomics 3",
      disease_groups___altered = "Normal",
      tissue_groups = "blood",
      ethnicity_groups_imputed = NA,
      dataset_id___altered = NA,
      sample_id = paste0("S", dplyr::row_number())
    )

  predict_res <- sccomp_est |>
    sccomp::sccomp_predict(
      formula_composition = NULL,
      new_data = sample_metadata,
      summary_instead_of_draws = FALSE
    ) |>
    dplyr::rename(cell_type = L3) |>
    trim_predict_res(l3_for_comparison)

  predict_dir <- dirname(params$composition_predict_rds)
  if (!dir.exists(predict_dir)) {
    dir.create(predict_dir, recursive = TRUE, showWarnings = FALSE)
  }
  saveRDS(predict_res, params$composition_predict_rds, compress = "xz")
}

predict_res <- trim_predict_res(predict_res, l3_for_comparison)

predict_res |>
  dplyr::distinct(.data$age_decade, .data$sex, .data$cell_type) |>
  dplyr::count(.data$age_decade, .data$sex, name = "n_cell_types")
```

### Compare observed SAVI proportions to healthy predictive draws

For each donor we match **`predict_res`** rows with the same **`sex`** and **`age_decade`**, then compute empirical tail probabilities (same logic as `posteriorHCA::composition_posterior_test()`).

```{r}
#| label: composition-vs-healthy

composition_vs_healthy <- example_proportions |>
  dplyr::rename(
    sample_id_observed = sample_id,
    cell_type_observed = cell_type,
    proportion_observed = proportion
  ) |>
  dplyr::inner_join(
    predict_res |>
      dplyr::select(
        age_decade, sex, cell_type, proportion_sampled = proportion, .draw
      ),
    by = c("sex" = "sex", "age_decade" = "age_decade", "cell_type_observed" = "cell_type"),
    relationship = "many-to-many"
  ) |>
  dplyr::group_by(.data$sample_id_observed, .data$Category, .data$cell_type_observed) |>
  dplyr::summarise(
    proportion_observed = dplyr::first(.data$proportion_observed),
    healthy_median = stats::median(.data$proportion_sampled),
    healthy_mean = mean(.data$proportion_sampled),
    healthy_hpd_lo = stats::quantile(.data$proportion_sampled, 0.025),
    healthy_hpd_hi = stats::quantile(.data$proportion_sampled, 0.975),
    # Same tail convention as `gene_deviation_from_fit()`: P(draw >= obs), not P(obs > draw).
    p_above_healthy = mean(.data$proportion_sampled >= .data$proportion_observed),
    p_below_healthy = mean(.data$proportion_sampled <= .data$proportion_observed),
    empirical_confidence = 2 * pmin(.data$p_above_healthy, .data$p_below_healthy),
    direction = dplyr::case_when(
      .data$p_above_healthy < 0.025 ~ "higher_than_healthy",
      .data$p_below_healthy < 0.025 ~ "lower_than_healthy",
      TRUE ~ "consistent_with_healthy"
    ),
    .groups = "drop"
  )

composition_vs_healthy |>
  dplyr::arrange(.data$Category, .data$cell_type_observed, .data$sample_id_observed) |>
  dplyr::slice_head(n = 12)
```

### Paper compartments: trend by `Category` and comparison to published direction

We aggregate mapped L3 labels into the Figure 1E compartments from the paper, then summarise mean proportion per **`Category`** and compare the SAVI–CTRL direction to the published expectation.

```{r}
#| label: composition-paper-compartments

paper_compartment_l3 <- tibble::tribble(
  ~paper_compartment, ~direction_in_savi, ~cell_type,
  "Naive CD4 T", "increased", "cd4 naive",
  "Effector/memory CD4", "decreased", "cd4 tcm",
  "Effector/memory CD4", "decreased", "cd4 th1 em",
  "Effector/memory CD4", "decreased", "cd4 th17 em",
  "Effector/memory CD4", "decreased", "cd4 th2 em",
  "Effector/memory CD4", "decreased", "cd4 fh em",
  "Effector/memory CD4", "decreased", "cd4 th1/th17 em",
  "Effector/memory CD4", "decreased", "treg",
  "Naive CD8 T", "increased", "cd8 naive",
  "Effector/memory CD8", "decreased", "cd8 tcm",
  "Effector/memory CD8", "decreased", "cd8 tem",
  "NK", "decreased", "nk",
  "MAIT", "strongly decreased", "mait",
  "gamma-delta T", "strongly decreased", "tgd"
)

composition_paper_per_donor <- composition_vs_healthy |>
  dplyr::inner_join(paper_compartment_l3, by = c("cell_type_observed" = "cell_type")) |>
  dplyr::group_by(
    .data$sample_id_observed,
    .data$Category,
    .data$paper_compartment,
    .data$direction_in_savi
  ) |>
  dplyr::summarise(
    proportion_observed = sum(.data$proportion_observed),
    healthy_median = sum(.data$healthy_median),
    direction = dplyr::if_else(
      any(.data$direction == "higher_than_healthy"),
      "higher_than_healthy",
      dplyr::if_else(
        any(.data$direction == "lower_than_healthy"),
        "lower_than_healthy",
        "consistent_with_healthy"
      )
    ),
    .groups = "drop"
  )

# Donors with zero cells in a compartment are absent from counts; fill with 0 so
# category means (and CTRL dots) include all donors in each Category.
all_donors_comp <- example_proportions |>
  dplyr::distinct(.data$sample_id, .data$Category) |>
  dplyr::rename(sample_id_observed = sample_id)

compartment_meta <- paper_compartment_l3 |>
  dplyr::distinct(.data$paper_compartment, .data$direction_in_savi)

composition_paper_long <- tidyr::crossing(all_donors_comp, compartment_meta) |>
  dplyr::left_join(
    composition_paper_per_donor,
    by = c("sample_id_observed", "Category", "paper_compartment", "direction_in_savi")
  ) |>
  dplyr::mutate(
    proportion_observed = dplyr::coalesce(.data$proportion_observed, 0),
    healthy_median = dplyr::coalesce(.data$healthy_median, 0),
    direction = dplyr::coalesce(.data$direction, "consistent_with_healthy")
  )

composition_paper_summary <- composition_paper_long |>
  dplyr::group_by(.data$paper_compartment, .data$direction_in_savi, .data$Category) |>
  dplyr::summarise(
    n_samples = dplyr::n(),
    mean_prop = mean(.data$proportion_observed),
    median_prop = stats::median(.data$proportion_observed),
    mean_healthy_median = mean(.data$healthy_median),
    .groups = "drop"
  ) |>
  tidyr::pivot_wider(
    names_from = Category,
    values_from = c(mean_prop, median_prop, n_samples),
    names_sep = "__"
  ) |>
  dplyr::mutate(
    delta_savi_minus_ctrl = .data$mean_prop__SAVI - .data$mean_prop__CTRL,
    delta_treated_minus_ctrl = .data$mean_prop__SAVI_treated - .data$mean_prop__CTRL,
    matches_paper_savi = dplyr::case_when(
      .data$direction_in_savi == "increased" ~ .data$delta_savi_minus_ctrl > 0,
      .data$direction_in_savi %in% c("decreased", "strongly decreased") ~ .data$delta_savi_minus_ctrl < 0,
      TRUE ~ NA
    ),
    matches_paper_treated = dplyr::case_when(
      .data$direction_in_savi == "increased" ~ .data$delta_treated_minus_ctrl > 0,
      .data$direction_in_savi %in% c("decreased", "strongly decreased") ~ .data$delta_treated_minus_ctrl < 0,
      TRUE ~ NA
    )
  ) |>
  dplyr::arrange(.data$paper_compartment)

composition_paper_summary
```

### Healthy baseline vs age decade

Healthy atlas predictive draws (`predict_res`, Normal blood): for each **paper compartment**, sum mapped L3 proportions per draw, then summarise by **`sex`** and **`age_decade`** (1–4).

```{r}
#| label: plot-composition-healthy-baseline-by-age
#| fig-width: 12
#| fig-height: 9
#| fig-align: center

healthy_baseline_compartment_age <- predict_res |>
  dplyr::inner_join(
    paper_compartment_l3 |>
      dplyr::distinct(.data$paper_compartment, .data$cell_type),
    by = "cell_type"
  ) |>
  dplyr::group_by(
    .data$paper_compartment,
    .data$sex,
    .data$age_decade,
    .data$.draw
  ) |>
  dplyr::summarise(
    compartment_proportion = sum(.data$proportion),
    .groups = "drop"
  ) |>
  dplyr::group_by(
    .data$paper_compartment,
    .data$sex,
    .data$age_decade
  ) |>
  dplyr::summarise(
    median_proportion = stats::median(.data$compartment_proportion),
    hpd_lo = stats::quantile(.data$compartment_proportion, 0.025),
    hpd_hi = stats::quantile(.data$compartment_proportion, 0.975),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    age_decade = factor(.data$age_decade, levels = as.character(1:4)),
    sex = factor(.data$sex, levels = c("female", "male")),
    paper_compartment = factor(
      .data$paper_compartment,
      levels = unique(paper_compartment_l3$paper_compartment)
    )
  )

sex_cols <- c(female = "#E41A1C", male = "#377EB8")

ggplot2::ggplot(
  healthy_baseline_compartment_age,
  ggplot2::aes(
    x = age_decade,
    y = median_proportion,
    color = sex,
    group = sex
  )
) +
  ggplot2::geom_ribbon(
    ggplot2::aes(ymin = hpd_lo, ymax = hpd_hi, fill = sex),
    alpha = 0.15,
    color = NA
  ) +
  ggplot2::geom_line(linewidth = 0.9) +
  ggplot2::geom_point(size = 2.2) +
  ggplot2::facet_wrap(~paper_compartment, scales = "free_y", ncol = 2) +
  ggplot2::scale_color_manual(values = sex_cols, name = "Sex") +
  ggplot2::scale_fill_manual(values = sex_cols, guide = "none") +
  ggplot2::labs(
    title = "Healthy atlas compartment proportion by sex and age decade",
    subtitle = paste(
      "Lines = posterior median per decade; ribbon = 2.5–97.5% across predictive draws.",
      "Compartment mass = sum of mapped L3 proportions (sccomp Normal blood)."
    ),
    x = "Age decade",
    y = "Compartment proportion (healthy baseline)"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    legend.position = "bottom",
    strip.text = ggplot2::element_text(face = "bold", size = 9)
  )
```

### Observed composition (within cohort)

Boxplot and bar chart use **observed** compartment proportions only (CTRL / SAVI / SAVI_treated); no healthy atlas reference.

```{r}
#| label: plot-composition-paper-box
#| fig-width: 12
#| fig-height: 12
#| fig-align: center

category_cols <- c(
  CTRL = "#4DAF4A",
  SAVI = "#E41A1C",
  SAVI_treated = "#377EB8"
)

plot_data <- composition_paper_long |>
  dplyr::mutate(
    Category = factor(.data$Category, levels = names(category_cols)),
    paper_compartment = forcats::fct_reorder(
      .data$paper_compartment,
      .data$proportion_observed,
      .fun = median,
      .desc = TRUE
    )
  )

ggplot2::ggplot(
  plot_data,
  ggplot2::aes(
    x = Category,
    y = proportion_observed,
    fill = Category
  )
) +
  ggplot2::geom_boxplot(outlier.shape = NA, alpha = 0.75) +
  ggplot2::geom_jitter(width = 0.12, size = 1.6, alpha = 0.85) +
  ggplot2::facet_wrap(~paper_compartment, scales = "free_y", ncol = 2) +
  ggplot2::scale_fill_manual(values = category_cols) +
  ggplot2::labs(
    title = "SAVI composition vs paper compartments (observed CellType -> L3)",
    subtitle = "Each point = one donor; compartments aggregate mapped L3 labels",
    x = NULL,
    y = "Proportion within sample"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "none")
```

```{r}
#| label: plot-composition-paper-vs-atlas
#| fig-width: 12
#| fig-height: 14
#| fig-align: center

composition_paper_bar <- composition_paper_summary |>
  tidyr::pivot_longer(
    cols = c(
      tidyselect::starts_with("mean_prop__"),
      tidyselect::starts_with("median_prop__")
    ),
    names_to = c("statistic", "Category"),
    names_sep = "__",
    values_to = "prop"
  ) |>
  dplyr::mutate(
    statistic = dplyr::recode(
      .data$statistic,
      mean_prop = "Mean across donors",
      median_prop = "Median across donors"
    ),
    statistic = factor(
      .data$statistic,
      levels = c("Mean across donors", "Median across donors")
    ),
    Category = factor(.data$Category, levels = names(category_cols))
  )

ggplot2::ggplot(
  composition_paper_bar,
  ggplot2::aes(x = Category, y = prop, fill = Category)
) +
  ggplot2::geom_col(position = ggplot2::position_dodge(width = 0.8), width = 0.7) +
  ggplot2::facet_grid(
    paper_compartment ~ statistic,
    scales = "free_y"
  ) +
  ggplot2::scale_fill_manual(values = category_cols) +
  ggplot2::labs(
    title = "Observed compartment proportion by Category",
    subtitle = paste(
      "Left column = mean across donors; right column = median.",
      "Observed data only (no atlas); compare SAVI vs CTRL with paper expectation",
      "(direction_in_savi in composition_paper_summary)."
    ),
    x = NULL,
    y = "Proportion within sample"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    legend.position = "bottom",
    strip.text = ggplot2::element_text(size = 9, face = "bold")
  )
```

### SAVI donors vs healthy baseline (unified compartments)

For **`SAVI`** and **`SAVI_treated`** donors only, we compare each donor’s **compartment proportion** (sum of mapped L3 types) to healthy **`predict_res`** draws matched on **`sex`** and **`age_decade`**. We then ask whether the atlas tail call (`higher` / `lower` / `consistent` vs healthy) agrees with the **paper’s expected direction** in SAVI (increased vs decreased).

```{r}
#| label: composition-savi-vs-healthy-compartments

baseline_compartment_draws <- predict_res |>
  dplyr::inner_join(
    paper_compartment_l3 |>
      dplyr::distinct(.data$paper_compartment, .data$cell_type),
    by = "cell_type"
  ) |>
  dplyr::group_by(
    .data$age_decade,
    .data$sex,
    .data$paper_compartment,
    .data$.draw
  ) |>
  dplyr::summarise(
    compartment_proportion = sum(.data$proportion),
    .groups = "drop"
  )

donor_sex_age_all <- example_proportions |>
  dplyr::distinct(.data$sample_id, .data$sex, .data$age_decade) |>
  dplyr::rename(sample_id_observed = sample_id)

# One reference table for all donors: tail test vs sex/age-matched healthy draws.
composition_matched_reference <- composition_paper_long |>
  dplyr::left_join(donor_sex_age_all, by = "sample_id_observed") |>
  dplyr::inner_join(
    baseline_compartment_draws,
    by = c("sex", "age_decade", "paper_compartment"),
    relationship = "many-to-many"
  ) |>
  dplyr::group_by(
    .data$sample_id_observed,
    .data$Category,
    .data$sex,
    .data$age_decade,
    .data$paper_compartment,
    .data$direction_in_savi,
    .data$proportion_observed
  ) |>
  dplyr::summarise(
    healthy_median = stats::median(.data$compartment_proportion),
    healthy_hpd_lo = stats::quantile(.data$compartment_proportion, 0.025),
    healthy_hpd_hi = stats::quantile(.data$compartment_proportion, 0.975),
    p_above_healthy = mean(.data$compartment_proportion >= .data$proportion_observed),
    p_below_healthy = mean(.data$compartment_proportion <= .data$proportion_observed),
    empirical_confidence = 2 * pmin(.data$p_above_healthy, .data$p_below_healthy),
    direction_vs_healthy = dplyr::case_when(
      .data$p_above_healthy < 0.025 ~ "higher_than_healthy",
      .data$p_below_healthy < 0.025 ~ "lower_than_healthy",
      TRUE ~ "consistent_with_healthy"
    ),
    .groups = "drop"
  )

composition_savi_vs_healthy <- composition_matched_reference |>
  dplyr::filter(.data$Category %in% c("SAVI", "SAVI_treated")) |>
  dplyr::mutate(
    atlas_direction_simple = dplyr::case_when(
      .data$direction_vs_healthy == "higher_than_healthy" ~ "increased",
      .data$direction_vs_healthy == "lower_than_healthy" ~ "decreased",
      TRUE ~ "consistent"
    ),
    agrees_with_paper = dplyr::case_when(
      .data$direction_in_savi == "increased" ~
        .data$atlas_direction_simple == "increased",
      .data$direction_in_savi %in% c("decreased", "strongly decreased") ~
        .data$atlas_direction_simple == "decreased",
      TRUE ~ NA
    ),
    agreement_label = dplyr::case_when(
      .data$agrees_with_paper %in% TRUE ~ "agrees (atlas vs paper)",
      .data$agrees_with_paper %in% FALSE ~ "disagrees (atlas vs paper)",
      TRUE ~ "atlas consistent (no tail)"
    )
  )

composition_savi_vs_healthy |>
  dplyr::arrange(.data$Category, .data$paper_compartment, .data$sample_id_observed)

composition_savi_agreement_summary <- composition_savi_vs_healthy |>
  dplyr::group_by(.data$Category, .data$paper_compartment, .data$direction_in_savi) |>
  dplyr::summarise(
    n_donors = dplyr::n(),
    n_higher = sum(.data$direction_vs_healthy == "higher_than_healthy"),
    n_lower = sum(.data$direction_vs_healthy == "lower_than_healthy"),
    n_consistent = sum(.data$direction_vs_healthy == "consistent_with_healthy"),
    n_agrees_paper = sum(.data$agrees_with_paper %in% TRUE, na.rm = TRUE),
    n_disagrees_paper = sum(.data$agrees_with_paper %in% FALSE, na.rm = TRUE),
    n_testable = sum(!is.na(.data$agrees_with_paper)),
    pct_agrees_paper = dplyr::if_else(
      sum(!is.na(.data$agrees_with_paper)) > 0L,
      100 * mean(.data$agrees_with_paper, na.rm = TRUE),
      NA_real_
    ),
    .groups = "drop"
  ) |>
  dplyr::arrange(.data$Category, .data$paper_compartment)

composition_savi_agreement_summary

# Diagnostic: heatmap tail call vs observed relative to matched healthy median.
composition_naive_cd4_check <- composition_matched_reference |>
  dplyr::filter(
    .data$paper_compartment == "Naive CD4 T",
    .data$Category == "SAVI"
  ) |>
  dplyr::transmute(
    sample_id_observed,
    age_decade,
    sex,
    proportion_observed,
    healthy_median_matched = .data$healthy_median,
    delta_obs_minus_matched = .data$proportion_observed - .data$healthy_median,
    direction_vs_healthy,
    visual_check = dplyr::case_when(
      .data$direction_vs_healthy == "higher_than_healthy" ~
        "solid line should be RIGHT of dashed median",
      .data$direction_vs_healthy == "lower_than_healthy" ~
        "solid line should be LEFT of dashed median",
      TRUE ~ "solid line near dashed median"
    )
  )

composition_naive_cd4_check
```

### How much CTRL samples in data agree with our healthy baseline

In-study **CTRL** donors are the paper’s reference for “healthy”. If they are exchangeable with the Immune Atlas healthy baseline, each **CTRL** pseudobulk compartment proportion should fall inside the **sex/age-matched** predictive null (`direction_vs_healthy == consistent_with_healthy`). Deviations flag compartments where study controls already sit above or below the atlas before comparing SAVI.

```{r}
#| label: composition-ctrl-baseline

composition_ctrl_baseline <- composition_matched_reference |>
  dplyr::filter(.data$Category == "CTRL") |>
  dplyr::mutate(
    agrees_with_healthy_baseline = .data$direction_vs_healthy == "consistent_with_healthy",
    baseline_agreement = dplyr::case_when(
      .data$direction_vs_healthy == "consistent_with_healthy" ~ "consistent with healthy",
      .data$direction_vs_healthy == "higher_than_healthy" ~ "higher than healthy",
      .data$direction_vs_healthy == "lower_than_healthy" ~ "lower than healthy"
    ),
    baseline_agreement = factor(
      .data$baseline_agreement,
      levels = c(
        "consistent with healthy",
        "higher than healthy",
        "lower than healthy"
      )
    ),
    atlas_strength = 1 - .data$empirical_confidence,
    direction_short = dplyr::case_when(
      .data$direction_vs_healthy == "higher_than_healthy" ~ "↑",
      .data$direction_vs_healthy == "lower_than_healthy" ~ "↓",
      TRUE ~ "·"
    ),
    tile_label = sprintf(
      "%s\n%.2f",
      .data$direction_short,
      .data$atlas_strength
    ),
    paper_compartment = factor(
      .data$paper_compartment,
      levels = unique(paper_compartment_l3$paper_compartment)
    )
  )

composition_ctrl_baseline_summary <- composition_ctrl_baseline |>
  dplyr::group_by(.data$sample_id_observed, .data$sex, .data$age_decade) |>
  dplyr::summarise(
    n_compartments = dplyr::n(),
    n_consistent = sum(.data$agrees_with_healthy_baseline),
    n_higher = sum(.data$direction_vs_healthy == "higher_than_healthy"),
    n_lower = sum(.data$direction_vs_healthy == "lower_than_healthy"),
    pct_consistent = 100 * .data$n_consistent / .data$n_compartments,
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(.data$pct_consistent), .data$sample_id_observed)

composition_ctrl_baseline_overall <- composition_ctrl_baseline |>
  dplyr::summarise(
    n_ctrl_donors = dplyr::n_distinct(.data$sample_id_observed),
    n_tests = dplyr::n(),
    n_consistent = sum(.data$agrees_with_healthy_baseline),
    n_higher = sum(.data$direction_vs_healthy == "higher_than_healthy"),
    n_lower = sum(.data$direction_vs_healthy == "lower_than_healthy"),
    pct_consistent = 100 * .data$n_consistent / .data$n_tests,
    mean_pct_consistent_per_donor = mean(composition_ctrl_baseline_summary$pct_consistent)
  )

composition_ctrl_baseline_overall
composition_ctrl_baseline_summary
```

```{r}
#| label: plot-composition-ctrl-baseline-heatmap
#| fig-width: 12
#| fig-height: 5.5
#| fig-align: center

ctrl_baseline_cols <- c(
  "consistent with healthy" = "#4DAF4A",
  "higher than healthy" = "#E41A1C",
  "lower than healthy" = "#377EB8"
)

ctrl_sample_order <- composition_ctrl_baseline_summary |>
  dplyr::arrange(dplyr::desc(.data$pct_consistent), .data$sample_id_observed) |>
  dplyr::pull(.data$sample_id_observed)

ctrl_heatmap_data <- composition_ctrl_baseline |>
  dplyr::mutate(
    sample_id_observed = factor(.data$sample_id_observed, levels = ctrl_sample_order),
    tile_label = dplyr::if_else(
      .data$agrees_with_healthy_baseline,
      "·",
      .data$tile_label
    )
  )

ggplot2::ggplot(
  ctrl_heatmap_data,
  ggplot2::aes(
    x = paper_compartment,
    y = sample_id_observed,
    fill = baseline_agreement
  )
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.45) +
  ggplot2::geom_text(
    ggplot2::aes(label = tile_label),
    size = 2.5,
    lineheight = 0.85,
    color = "grey15"
  ) +
  ggplot2::scale_fill_manual(
    values = ctrl_baseline_cols,
    name = "vs matched\nhealthy null"
  ) +
  ggplot2::labs(
    title = "CTRL donors: agreement with sex/age-matched healthy atlas (paper compartments)",
    subtitle = paste0(
      "Green = consistent_with_healthy (",
      sprintf("%.0f", composition_ctrl_baseline_overall$pct_consistent),
      "% of donor × compartment tests cohort-wide). ",
      "Non-green tiles show direction (↑/↓) and tail strength (1 - empirical confidence)."
    ),
    x = NULL,
    y = "CTRL donor"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 35, hjust = 1, size = 8),
    axis.text.y = ggplot2::element_text(size = 9),
    legend.position = "bottom"
  )
```

Each panel below is one **CTRL donor × paper compartment**: grey density = healthy **`predict_res`** draws for that donor’s **`sex`** and **`age_decade`**; coloured vertical line = observed proportion. Line colour matches the heatmap (green = consistent, red/blue = tail call). Text at each line reports **`p_above_healthy`** = P(draw ≥ obs) and **`p_below_healthy`** = P(draw ≤ obs) (same tail convention as gene-level tests).

```{r}
#| label: plot-composition-ctrl-baseline-density
#| fig-width: 20
#| fig-height: 15
#| fig-align: center

compartment_levels <- paper_compartment_l3 |>
  dplyr::distinct(.data$paper_compartment) |>
  dplyr::pull(.data$paper_compartment)

if (!exists("baseline_compartment_draws")) {
  baseline_compartment_draws <- predict_res |>
    dplyr::inner_join(
      paper_compartment_l3 |>
        dplyr::distinct(.data$paper_compartment, .data$cell_type),
      by = "cell_type"
    ) |>
    dplyr::group_by(
      .data$paper_compartment,
      .data$age_decade,
      .data$sex,
      .data$.draw
    ) |>
    dplyr::summarise(
      compartment_proportion = sum(.data$proportion),
      .groups = "drop"
    )
}

ctrl_panel_meta <- composition_ctrl_baseline |>
  dplyr::distinct(
    .data$sample_id_observed,
    .data$sex,
    .data$age_decade,
    .data$paper_compartment
  ) |>
  dplyr::mutate(
    sample_panel = paste0(
      .data$sample_id_observed,
      " (",
      .data$sex,
      " | decade ",
      .data$age_decade,
      ")"
    )
  )

ctrl_panel_order <- composition_ctrl_baseline_summary |>
  dplyr::mutate(
    sample_panel = paste0(
      .data$sample_id_observed,
      " (",
      .data$sex,
      " | decade ",
      .data$age_decade,
      ")"
    )
  ) |>
  dplyr::arrange(dplyr::desc(.data$pct_consistent), .data$sample_id_observed)

ctrl_density_draws <- ctrl_panel_meta |>
  dplyr::inner_join(
    baseline_compartment_draws,
    by = c("sex", "age_decade", "paper_compartment")
  ) |>
  dplyr::mutate(
    sample_panel = factor(
      .data$sample_panel,
      levels = ctrl_panel_order$sample_panel
    ),
    paper_compartment = factor(.data$paper_compartment, levels = compartment_levels)
  ) |>
  dplyr::filter(
    is.finite(.data$compartment_proportion),
    .data$compartment_proportion >= 0
  )

ctrl_density_vlines <- composition_ctrl_baseline |>
  dplyr::mutate(
    sample_panel = paste0(
      .data$sample_id_observed,
      " (",
      .data$sex,
      " | decade ",
      .data$age_decade,
      ")"
    ),
    sample_panel = factor(
      .data$sample_panel,
      levels = ctrl_panel_order$sample_panel
    ),
    paper_compartment = factor(.data$paper_compartment, levels = compartment_levels),
    p_label = sprintf(
      "p_above=%.2f\np_below=%.2f",
      .data$p_above_healthy,
      .data$p_below_healthy
    )
  ) |>
  dplyr::filter(is.finite(.data$proportion_observed)) |>
  dplyr::left_join(
    ctrl_density_draws |>
      dplyr::group_by(.data$sample_panel, .data$paper_compartment) |>
      dplyr::summarise(
        x_mid = stats::median(.data$compartment_proportion, na.rm = TRUE),
        .groups = "drop"
      ),
    by = c("sample_panel", "paper_compartment")
  ) |>
  dplyr::mutate(
    text_hjust = dplyr::if_else(
      .data$proportion_observed > .data$x_mid,
      1.05,
      -0.05
    )
  )

ggplot2::ggplot(
  ctrl_density_draws,
  ggplot2::aes(x = compartment_proportion)
) +
  ggplot2::geom_density(
    fill = "grey80",
    color = "grey35",
    alpha = 0.55,
    linewidth = 0.35,
    na.rm = TRUE
  ) +
  ggplot2::geom_vline(
    data = ctrl_density_vlines,
    ggplot2::aes(
      xintercept = proportion_observed,
      color = baseline_agreement
    ),
    inherit.aes = FALSE,
    linewidth = 0.75
  ) +
  ggplot2::geom_text(
    data = ctrl_density_vlines,
    ggplot2::aes(
      x = proportion_observed,
      y = Inf,
      label = p_label,
      hjust = text_hjust
    ),
    inherit.aes = FALSE,
    vjust = 1.05,
    size = 2.1,
    lineheight = 0.85,
    color = "grey15"
  ) +
  ggplot2::facet_grid(
    sample_panel ~ paper_compartment,
    scales = "free",
    labeller = ggplot2::labeller(
      .cols = ggplot2::label_wrap_gen(10),
      .rows = ggplot2::label_wrap_gen(16)
    )
  ) +
  ggplot2::scale_color_manual(
    values = ctrl_baseline_cols,
    name = "CTRL vs\nmatched null"
  ) +
  ggplot2::scale_x_continuous(expand = ggplot2::expansion(mult = c(0.04, 0.06))) +
  ggplot2::labs(
    title = "CTRL donors vs sex/age-matched healthy atlas (density grid)",
    subtitle = paste(
      "Rows = CTRL donor (sex | age decade); columns = paper compartments.",
      "Grey = healthy predictive draws for that donor's metadata;",
      "vertical line = observed CTRL proportion;",
      "labels = p_above_healthy / p_below_healthy."
    ),
    x = "Compartment proportion",
    y = "Density"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    legend.position = "bottom",
    strip.text.x = ggplot2::element_text(size = 7, face = "bold"),
    strip.text.y = ggplot2::element_text(size = 7),
    panel.spacing = grid::unit(0.25, "lines")
  )
```

**Read:** donors sorted by **% compartments consistent** (top = closest to atlas). Widespread non-green tiles mean in-study controls are **not** drawn from the same reference as Immune Atlas healthy PBMCs for those compartments — relevant when interpreting SAVI vs atlas (and gene-level CTRL drift). The **density grid** shows the same comparison panel-by-panel: a line inside the grey band supports consistency; a line in the tail matches the heatmap colour.

### Healthy atlas comparison (plots)

Tail tests and visualizations below use **sex/age-matched** healthy **`predict_res`** draws (`composition_matched_reference`).

```{r}
#| label: plot-savi-atlas-direction-heatmap
#| fig-width: 12
#| fig-height: 7
#| fig-align: center

# empirical_confidence = 2 * min(P(draw>=obs), P(draw<=obs)): near 0 = strong tail;
# near 1 = central. Signed score: direction x (1 - empirical_confidence).
heatmap_data <- composition_savi_vs_healthy |>
  dplyr::mutate(
    Category = factor(.data$Category, levels = c("SAVI", "SAVI_treated")),
    atlas_strength = 1 - .data$empirical_confidence,
    atlas_signed_score = dplyr::case_when(
      .data$direction_vs_healthy == "higher_than_healthy" ~ .data$atlas_strength,
      .data$direction_vs_healthy == "lower_than_healthy" ~ -.data$atlas_strength,
      TRUE ~ 0
    ),
    direction_short = dplyr::case_when(
      .data$direction_vs_healthy == "higher_than_healthy" ~ "↑",
      .data$direction_vs_healthy == "lower_than_healthy" ~ "↓",
      TRUE ~ "·"
    ),
    tile_label = sprintf(
      "%s\n%.2f",
      .data$direction_short,
      .data$atlas_strength
    ),
    sample_label = paste0(.data$sample_id_observed, " (", .data$Category, ")"),
    paper_compartment = factor(
      .data$paper_compartment,
      levels = unique(paper_compartment_l3$paper_compartment)
    )
  )

score_lim <- max(abs(heatmap_data$atlas_signed_score), na.rm = TRUE)
score_lim <- if (is.finite(score_lim) && score_lim > 0) score_lim else 1

ggplot2::ggplot(
  heatmap_data,
  ggplot2::aes(
    x = paper_compartment,
    y = sample_label,
    fill = atlas_signed_score
  )
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.4) +
  ggplot2::geom_text(
    ggplot2::aes(label = tile_label),
    size = 2.4,
    lineheight = 0.85,
    color = "grey15"
  ) +
  ggplot2::scale_fill_gradient2(
    low = "#377EB8",
    mid = "grey92",
    high = "#E41A1C",
    midpoint = 0,
    limits = c(-score_lim, score_lim),
    oob = scales::squish,
    name = "Signed evidence\nvs healthy"
  ) +
  ggplot2::facet_grid(.data$Category ~ ., scales = "free_y", space = "free_y") +
  ggplot2::labs(
    title = "SAVI donors: strength of deviation from healthy atlas (unified compartments)",
    subtitle = paste(
      "Tail test vs sex/age-matched healthy draws (not pooled across all decades).",
      "Red = higher, blue = lower; intensity = 1 - empirical confidence."
    ),
    x = NULL,
    y = NULL
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 35, hjust = 1, size = 8),
    axis.text.y = ggplot2::element_text(size = 7),
    strip.text = ggplot2::element_text(face = "bold"),
    legend.position = "bottom"
  )
```

```{r}
#| label: plot-savi-atlas-direction-counts
#| fig-width: 12
#| fig-height: 6
#| fig-align: center

direction_cols <- c(
  higher_than_healthy = "#E41A1C",
  lower_than_healthy = "#377EB8",
  consistent_with_healthy = "grey75"
)

count_long <- composition_savi_agreement_summary |>
  tidyr::pivot_longer(
    cols = c(n_higher, n_lower, n_consistent),
    names_to = "atlas_call",
    values_to = "n"
  ) |>
  dplyr::mutate(
    Category = factor(.data$Category, levels = c("SAVI", "SAVI_treated")),
    atlas_call = dplyr::recode(
      .data$atlas_call,
      n_higher = "higher_than_healthy",
      n_lower = "lower_than_healthy",
      n_consistent = "consistent_with_healthy"
    ),
    atlas_call = factor(.data$atlas_call, levels = names(direction_cols))
  )

ggplot2::ggplot(
  count_long,
  ggplot2::aes(x = paper_compartment, y = n, fill = atlas_call)
) +
  ggplot2::geom_col(position = "stack") +
  ggplot2::facet_wrap(~Category, ncol = 1) +
  ggplot2::scale_fill_manual(values = direction_cols, name = "Atlas call") +
  ggplot2::labs(
    title = "Count of SAVI donors by atlas direction per compartment",
    subtitle = "Stacked bars: how many donors are higher / lower / consistent vs healthy null",
    x = NULL,
    y = "Number of donors"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 35, hjust = 1, size = 8),
    legend.position = "bottom"
  )
```

```{r}
#| label: plot-savi-paper-agreement
#| fig-width: 12
#| fig-height: 6
#| fig-align: center

agreement_cols <- c(
  "agrees (atlas vs paper)" = "#4DAF4A",
  "disagrees (atlas vs paper)" = "#FF7F00",
  "atlas consistent (no tail)" = "grey70"
)

ggplot2::ggplot(
  composition_savi_vs_healthy |>
    dplyr::mutate(
      Category = factor(.data$Category, levels = c("SAVI", "SAVI_treated")),
      agreement_label = factor(.data$agreement_label, levels = names(agreement_cols))
    ),
  ggplot2::aes(x = paper_compartment, fill = agreement_label)
) +
  ggplot2::geom_bar(position = "fill") +
  ggplot2::facet_wrap(~Category, ncol = 1) +
  ggplot2::scale_fill_manual(values = agreement_cols, name = "Agreement") +
  ggplot2::scale_y_continuous(labels = scales::percent_format()) +
  ggplot2::labs(
    title = "Atlas vs paper agreement (SAVI donors, unified compartments)",
    subtitle = paste(
      "Paper expects increased (naive CD4/CD8) or decreased (effector/memory, NK, MAIT, gdT).",
      "Green = atlas tail matches paper direction; orange = opposite tail."
    ),
    x = NULL,
    y = "Fraction of donors"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 35, hjust = 1, size = 8),
    legend.position = "bottom"
  )
```

```{r}
#| label: plot-savi-paper-agreement-rate
#| fig-width: 10
#| fig-height: 5
#| fig-align: center

ggplot2::ggplot(
  composition_savi_agreement_summary |>
    dplyr::mutate(
      Category = factor(.data$Category, levels = c("SAVI", "SAVI_treated")),
      paper_compartment = factor(
        .data$paper_compartment,
        levels = unique(paper_compartment_l3$paper_compartment)
      )
    ),
  ggplot2::aes(
    x = paper_compartment,
    y = pct_agrees_paper,
    fill = Category
  )
) +
  ggplot2::geom_col(position = ggplot2::position_dodge(width = 0.75), width = 0.65) +
  ggplot2::geom_hline(yintercept = 50, linetype = "dashed", color = "grey50") +
  ggplot2::scale_fill_manual(values = c(SAVI = "#E41A1C", SAVI_treated = "#377EB8")) +
  ggplot2::scale_y_continuous(limits = c(0, 100), expand = ggplot2::expansion(mult = c(0, 0.02))) +
  ggplot2::labs(
    title = "Share of SAVI donors where atlas tail agrees with paper",
    subtitle = "Among donors with a clear atlas tail (excludes 'consistent' only when testing agreement)",
    x = NULL,
    y = "% donors agreeing with paper direction"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 35, hjust = 1, size = 8))
```

### Compartment density vs matched healthy null

Same layout as the **CTRL** density grid (`plot-composition-ctrl-baseline-density`): one panel per **donor × paper compartment**. Grey density = healthy **`predict_res`** draws for that donor’s **`sex`** and **`age_decade`**; vertical line = observed proportion with tail colours and **`p_above_healthy`** / **`p_below_healthy`** labels. Below: **SAVI** and **SAVI_treated** donors only.

```{r}
#| label: plot-composition-compartment-density
#| fig-width: 20
#| fig-height: 18
#| fig-align: center

compartment_levels <- paper_compartment_l3 |>
  dplyr::distinct(.data$paper_compartment) |>
  dplyr::pull(.data$paper_compartment)

atlas_null_cols <- c(
  "consistent with healthy" = "#4DAF4A",
  "higher than healthy" = "#E41A1C",
  "lower than healthy" = "#377EB8"
)

if (!exists("baseline_compartment_draws")) {
  baseline_compartment_draws <- predict_res |>
    dplyr::inner_join(
      paper_compartment_l3 |>
        dplyr::distinct(.data$paper_compartment, .data$cell_type),
      by = "cell_type"
    ) |>
    dplyr::group_by(
      .data$paper_compartment,
      .data$age_decade,
      .data$sex,
      .data$.draw
    ) |>
    dplyr::summarise(
      compartment_proportion = sum(.data$proportion),
      .groups = "drop"
    )
}

savi_cohort_atlas <- composition_matched_reference |>
  dplyr::filter(.data$Category %in% c("SAVI", "SAVI_treated")) |>
  dplyr::mutate(
    baseline_agreement = dplyr::case_when(
      .data$direction_vs_healthy == "consistent_with_healthy" ~ "consistent with healthy",
      .data$direction_vs_healthy == "higher_than_healthy" ~ "higher than healthy",
      .data$direction_vs_healthy == "lower_than_healthy" ~ "lower than healthy"
    ),
    baseline_agreement = factor(
      .data$baseline_agreement,
      levels = names(atlas_null_cols)
    ),
    sample_panel = paste0(
      .data$sample_id_observed,
      " (",
      .data$Category,
      " · ",
      .data$sex,
      " | decade ",
      .data$age_decade,
      ")"
    )
  )

savi_panel_order <- savi_cohort_atlas |>
  dplyr::distinct(.data$sample_panel, .data$Category, .data$sample_id_observed) |>
  dplyr::mutate(
    Category = factor(.data$Category, levels = c("SAVI", "SAVI_treated"))
  ) |>
  dplyr::arrange(.data$Category, .data$sample_id_observed)

savi_panel_meta <- savi_cohort_atlas |>
  dplyr::distinct(
    .data$sample_id_observed,
    .data$Category,
    .data$sex,
    .data$age_decade,
    .data$paper_compartment,
    .data$sample_panel
  )

savi_density_draws <- savi_panel_meta |>
  dplyr::inner_join(
    baseline_compartment_draws,
    by = c("sex", "age_decade", "paper_compartment")
  ) |>
  dplyr::mutate(
    sample_panel = factor(
      .data$sample_panel,
      levels = savi_panel_order$sample_panel
    ),
    paper_compartment = factor(.data$paper_compartment, levels = compartment_levels)
  ) |>
  dplyr::filter(
    is.finite(.data$compartment_proportion),
    .data$compartment_proportion >= 0
  )

savi_density_vlines <- savi_cohort_atlas |>
  dplyr::mutate(
    sample_panel = factor(
      .data$sample_panel,
      levels = savi_panel_order$sample_panel
    ),
    paper_compartment = factor(.data$paper_compartment, levels = compartment_levels),
    p_label = sprintf(
      "p_above=%.2f\np_below=%.2f",
      .data$p_above_healthy,
      .data$p_below_healthy
    )
  ) |>
  dplyr::filter(is.finite(.data$proportion_observed)) |>
  dplyr::left_join(
    savi_density_draws |>
      dplyr::group_by(.data$sample_panel, .data$paper_compartment) |>
      dplyr::summarise(
        x_mid = stats::median(.data$compartment_proportion, na.rm = TRUE),
        .groups = "drop"
      ),
    by = c("sample_panel", "paper_compartment")
  ) |>
  dplyr::mutate(
    text_hjust = dplyr::if_else(
      .data$proportion_observed > .data$x_mid,
      1.05,
      -0.05
    )
  )

ggplot2::ggplot(
  savi_density_draws,
  ggplot2::aes(x = compartment_proportion)
) +
  ggplot2::geom_density(
    fill = "grey80",
    color = "grey35",
    alpha = 0.55,
    linewidth = 0.35,
    na.rm = TRUE
  ) +
  ggplot2::geom_vline(
    data = savi_density_vlines,
    ggplot2::aes(
      xintercept = proportion_observed,
      color = baseline_agreement
    ),
    inherit.aes = FALSE,
    linewidth = 0.75
  ) +
  ggplot2::geom_text(
    data = savi_density_vlines,
    ggplot2::aes(
      x = proportion_observed,
      y = Inf,
      label = p_label,
      hjust = text_hjust
    ),
    inherit.aes = FALSE,
    vjust = 1.05,
    size = 2.1,
    lineheight = 0.85,
    color = "grey15"
  ) +
  ggplot2::facet_grid(
    sample_panel ~ paper_compartment,
    scales = "free",
    labeller = ggplot2::labeller(
      .cols = ggplot2::label_wrap_gen(10),
      .rows = ggplot2::label_wrap_gen(16)
    )
  ) +
  ggplot2::scale_color_manual(
    values = atlas_null_cols,
    name = "Donor vs\nmatched null"
  ) +
  ggplot2::scale_x_continuous(expand = ggplot2::expansion(mult = c(0.04, 0.06))) +
  ggplot2::labs(
    title = "SAVI and SAVI_treated donors vs sex/age-matched healthy atlas (density grid)",
    subtitle = paste(
      "Rows = donor (Category · sex | age decade); columns = paper compartments.",
      "Grey = healthy predictive draws for that donor's metadata;",
      "vertical line = observed proportion;",
      "labels = p_above_healthy / p_below_healthy."
    ),
    x = "Compartment proportion",
    y = "Density"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    legend.position = "bottom",
    strip.text.x = ggplot2::element_text(size = 7, face = "bold"),
    strip.text.y = ggplot2::element_text(size = 7),
    panel.spacing = grid::unit(0.25, "lines")
  )
```

**How to read vs the paper**

1. **Observed cohort trends** (boxplot / bar chart): within-study CTRL vs SAVI vs SAVI_treated using observed proportions only (paper Figure 1E style). **`composition_paper_summary`**: `delta_savi_minus_ctrl` and `matches_paper_savi` flag agreement with published SAVI vs control direction.
2. **Healthy baseline by age** (line chart): how female vs male compartment proportions change across age decades 1–4 in **`predict_res`** (context for sex/age matching).
3. **CTRL vs atlas** (`plot-composition-ctrl-baseline-heatmap`, `plot-composition-ctrl-baseline-density`): sample × compartment heatmap plus per-donor density grid — grey curve = sex/age-matched healthy draws; line = observed CTRL proportion.
4. **Healthy atlas comparison** (heatmap, count/agreement bars): per donor × compartment tail test vs sex/age-matched **`predict_res`** draws (`composition_matched_reference`, `composition_savi_vs_healthy`).
5. **Compartment density plot** (`plot-composition-compartment-density`): same donor × compartment grid as CTRL — matched healthy density, tail-coloured line, **p_above** / **p_below** labels (**SAVI** and **SAVI_treated**).
6. **Fine-grained L3 table** (`composition_vs_healthy`): per donor × L3, `direction` and `empirical_confidence` with the same tail convention as gene-level deviation tests.


