---
title: "SAVI case study — STING signature on monocytic reference"
subtitle: "Posterior models, compression, and merge with SAVI pseudobulk"
date: today
date-format: "DD MMM YYYY"
title-block-banner: true
author:
  - name: Chen Zhan
    orcid: 0000-0002-4794-8339
    email: chen.zhan@adelaide.edu.au
    affiliation:
      - id: sgc
        name: SAiGENCI, The University of Adelaide
format:
  html:
    theme: minty
    df-print: paged
    code-line-numbers: true
    embed-resources: true
    code-fold: true
    code-fold-default: true
    toc: true
    toc-location: left
    toc-title: "Contents"
knitr:
  opts_chunk:
    message: false
    warning: false
editor: source
abstract: |
  This notebook applies the same pattern as the Wu breast cancer PAM50 toy model:
  we take a literature-defined gene set from a published SAVI PBMC study, align it
  to Ensembl gene ids present in the posterior HCA monocytic `targets` store,
  optionally re-fit and compress `brms` models with `poco`, then load SAVI
  pseudobulk (GEO) and healthy reference samples for downstream comparison of
  observed counts to reference posteriors.
params:
  run_fits: false
  n_components: 5
  corr_threshold: 0.15
  target_cell_type: "monocytic"
  estimates_rds: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/dev/SAVI/data/SAVI_estimates_chunk.rds"
  savi_pseudobulk_rds: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/dev/SAVI/data/GSE226598_SAVI_pseudobulk_Sample_CellType.rds"
  savi_seurat_gz: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/dev/SAVI/data/GSE226598_SeuratObject_SAVI.rds.gz"
  savi_sc_celltype: "17. Disease-associated monocytes"
  pseudobulk_h5ad: "/home/a1237163/lab/Mangiola_ImmuneAtlas/zenodo/pseudobulk_se.h5ad"
  composition_predict_rds: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/dev/SAVI/data/SAVI_composition_predict_res.rds"
  sccomp_estimates_rds: "/home/a1237163/lab/Mangiola_ImmuneAtlas/taskforce_shared_folder/sccomp_on_cellNexus_1_0_12/estimates_age_decade___L3___disease_TRUE___immune_only_TRUE.rds"
  model_store_root: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/HPC_posterior/NEW_CELL_TYPE"
---

# Introduction

[STING](https://en.wikipedia.org/wiki/Stimulator_of_interferon_genes)-associated vasculopathy with onset in infancy (**SAVI**) is caused by gain-of-function mutations in *STING1*. De Cevins *et al.* performed single-cell RNA-seq of PBMCs from SAVI patients and healthy controls, and an IFN-β time-course in healthy PBMCs, to dissect IFN-dependent and IFN-independent programmes. They highlight disease-associated monocytes and propose a **21-gene** transcriptomic signature of STING activation that is **enriched for TNF-α / NF-κB–related biology** and **filtered against IFN-β–induced genes** in monocytes and DCs (Figure 6; Table S6).

**Primary publication (open access):** de Cevins, C., *et al.* (2023). Single-cell RNA-sequencing of PBMCs from SAVI patients reveals disease-associated monocytes with elevated integrated stress response. *Cell Reports Medicine* **4**, 101333. <https://doi.org/10.1016/j.xcrm.2023.101333> — PMC: <https://pmc.ncbi.nlm.nih.gov/articles/PMC10772457/>.

**Data:** scRNA-seq are deposited under superseries [GSE226601](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE226601): SAVI cohort [GSE226598](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE226598), IFN-β dataset [GSE226572](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE226572). Pseudobulk for this notebook is prepared in `dev/SAVI/data/prepare_data.R` from the GEO Seurat object.

**Scope of this document:** We do **not** reproduce the paper’s within-study differential analysis or IFN-β filtering steps here. We use their **published 21-gene symbol list** (Table S6 / Figure 6) as biomarkers, fit or load **monocytic-lineage** reference models from the lab `targets` store, and attach **SAVI sample × cell type** pseudobulk plus **`offset == 0`** reference samples for planned comparisons to healthy baselines (same structural role as `pam50_toy_model.qmd` for breast cancer).

------------------------------------------------------------------------

# Library setup

```{r}
#| label: setup
# HPC / project R library (adjust if running elsewhere)
.libPaths("/hpcfs/users/a1237163/R/x86_64-pc-linux-gnu-library/4.5")

library(tibble)
library(tidyr)
library(dplyr)
library(tidySummarizedExperiment)
library(SummarizedExperiment)
library(S4Vectors)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(tidybulk)
library(glue)
library(purrr)
library(furrr)
library(future)
library(lobstr)
library(posterior)
library(progressr)
library(brms)
library(poco)
library(tidyverse)
library(ggplot2)
library(forcats)
library(patchwork)
library(pheatmap)
library(RColorBrewer)
library(targets)
library(stringr)
library(Seurat)
library(tidyseurat)
```

------------------------------------------------------------------------

# STING activation signature (21 genes)

The symbols below follow **Table S6** / Figure 6 in de Cevins *et al.* (2023). We map HGNC symbols to Ensembl via `org.Hs.eg.db` for overlap with the reference model feature space; **lncRNAs and some readthrough ids** may return `NA` here even when present in the single-cell count matrix — those rows are dropped from the `targets`-based extraction unless you supply Ensembl ids manually.

```{r}
#| label: sting-signature
sting_signature_21 <- tibble(
  gene = c(
    "AC253572.2",
    "ADRB2",
    "ANKRD28",
    "ANXA1",
    "AQP9",
    "AZIN1-AS1",
    "IL1B",
    "JARID2",
    "KLF4",
    "MIR4435-2HG",
    "OTUD1",
    "PLAUR",
    "PTGS2",
    "PTX3",
    "SELENOK",
    "TGIF1",
    "TREM1",
    "TSC22D2",
    "ZBTB43",
    "ZFP36L2",
    "ZNF503"
  )
) |>
  mutate(
    .feature = AnnotationDbi::mapIds(
      org.Hs.eg.db::org.Hs.eg.db,
      keys = gene,
      column = "ENSEMBL",
      keytype = "SYMBOL",
      multiVals = "first"
    )
  )

sting_signature_21
```

------------------------------------------------------------------------

# Reference model store (`targets`)

Posterior HCA models live under `params$model_store_root`, in a subfolder named after the harmonised cell type (here `V1_monocytic` when `params$target_cell_type` is `"monocytic"`). Branch metadata encodes which Ensembl gene each `estimates_chunk_*` target belongs to (`Gene:___ENSG...___` in the recorded warning field).

```{r}
#| label: model-meta
model_path <- file.path(
  params$model_store_root,
  paste0("V1_", params$target_cell_type)
)

model_meta <-
  targets::tar_meta(
    starts_with("estimates_chunk_"),
    store = file.path(model_path, "_targets")
  ) |>
  dplyr::filter(type == "branch") |>
  mutate(
    .feature = stringr::str_extract(
      warnings,
      "(?<=Gene:___)ENSG\\d+(?=___)"
    )
  )

# How many of the 21 (with successful SYMBOL→ENSEMBL map) appear in the store?
n_sig_in_store <- sum(sting_signature_21$.feature %in% model_meta$.feature, na.rm = TRUE)
n_sig_mapped <- sum(!is.na(sting_signature_21$.feature))

tibble(
  signature_genes = nrow(sting_signature_21),
  mapped_to_ensembl = n_sig_mapped,
  present_in_targets_store = n_sig_in_store
)
```

------------------------------------------------------------------------

# Fit and compress `brms` models (optional, heavy)

Set **`params$run_fits: true`** when rendering (or in the YAML) to re-run extraction, posterior correlation partitioning, `poco::compress_brmsfit`, and `saveRDS`. By default this chunk is **skipped** and the next section loads `params$estimates_rds`.

For each gene: read the precomputed `estimates_chunk` from disk (`tar_read_raw()` from the **targets** package, same usage as `dev/posteriorHCA_brms_fit_compression_test.qmd`), strip large environments from the fit, optionally compress with `poco`, and record layout / fidelity metrics.

```{r}
#| label: fit-sting-biomarkers
#| eval: !expr params$run_fits

handlers(global = TRUE)
handlers("progress")

options(future.globals.maxSize = 5 * 1024^3)

n_components <- params$n_components
corr_threshold <- params$corr_threshold

extract_one_biomarker <- function(ensg) {
  estimates_chunk <-
    model_meta |>
    dplyr::filter(.feature == ensg) |>
    dplyr::pull(name) |>
    tar_read_raw(store = file.path(model_path, "_targets")) |>
    dplyr::select(-tar_group) |>
    mutate(
      gene = sting_signature_21 |>
        dplyr::filter(!is.na(.feature), .feature == ensg) |>
        dplyr::pull(gene),
      cell_type = params$target_cell_type,
      brms_fit = purrr::map(brms_fit, poco::strip_brmsfit_envs)
    )

  fit_brms <- estimates_chunk$brms_fit[[1]]

  posterior_cor <- poco::posterior_correlation(
    fit_brms,
    type = "correlation",
    method = "pearson"
  )
  original_draws <- posterior::as_draws_matrix(fit_brms)

  partition <- poco::partition_parameters_clusters(
    posterior_cor,
    threshold = corr_threshold,
    min_size = 2L,
    simple_output = FALSE
  )

  layout_tbl <- tibble(
    n_components = n_components,
    corr_threshold = corr_threshold,
    n_blocks = length(partition$blocks),
    block_size_max = if (length(partition$blocks)) max(lengths(partition$blocks)) else 0L,
    block_size_med = if (length(partition$blocks)) {
      as.integer(stats::median(lengths(partition$blocks)))
    } else {
      0L
    },
    n_remainder = length(partition$remainder)
  )

  compress_tic <- Sys.time()
  compressed_blockwise <- tryCatch(
    poco::compress_brmsfit(
      brmsfit = fit_brms,
      method = "mclust",
      n_components = n_components,
      partition = partition,
      cluster_BPPARAM = 1L,
      verbose = FALSE
    ),
    error = function(e) {
      warning(
        "compress_brmsfit failed: ",
        conditionMessage(e),
        call. = FALSE,
        immediate. = TRUE
      )
      NULL
    }
  )
  time_used <- as.numeric(difftime(Sys.time(), compress_tic, units = "secs"))

  fid_blockwise <- if (is.null(compressed_blockwise)) {
    NULL
  } else {
    tryCatch(
      poco::evaluate_compression(
        compressed_blockwise$compressed,
        original_draws,
        seed = 1L
      ),
      error = function(e) {
        warning(
          "evaluate_compression failed: ",
          conditionMessage(e),
          call. = FALSE,
          immediate. = TRUE
        )
        NULL
      }
    )
  }

  if (is.null(compressed_blockwise)) {
    layout_tbl <- layout_tbl |>
      mutate(
        compression_ok = FALSE,
        time_used = NA_real_,
        compression_rate = NA_real_,
        energy_pct = NA_real_,
        c2st_pct = NA_real_,
        reproduction_pct = NA_real_
      )
  } else {
    estimates_chunk$brms_fit_compressed <- list(compressed_blockwise)
    layout_tbl <- layout_tbl |>
      mutate(
        compression_ok = TRUE,
        time_used = time_used,
        compression_rate =
          as.numeric(lobstr::obj_size(compressed_blockwise)) /
            as.numeric(lobstr::obj_size(fit_brms))
      )
  }

  if (is.null(fid_blockwise)) {
    layout_tbl <- layout_tbl |>
      mutate(
        energy_pct = NA_real_,
        c2st_pct = NA_real_,
        reproduction_pct = NA_real_
      )
  } else {
    en <- fid_blockwise$metrics$energy$reproduction_pct
    c2 <- fid_blockwise$metrics$c2st$reproduction_pct
    layout_tbl <- layout_tbl |>
      mutate(
        energy_pct = if (is.null(en)) NA_real_ else en,
        c2st_pct = if (is.null(c2)) NA_real_ else c2,
        reproduction_pct = fid_blockwise$reproduction_pct
      )
  }

  estimates_chunk |> tibble::add_column(layout_tbl)
}

sting_signature_21 |>
  dplyr::filter(.feature %in% model_meta$.feature) |>
  dplyr::pull(.feature) -> biomarkers

with_progress({
  p <- progressr::progressor(along = biomarkers)
  sting_estimates_chunk <- purrr::map_dfr(biomarkers, function(ensg) {
    p(sprintf("Processing %s", ensg))
    extract_one_biomarker(ensg)
  })
})

saveRDS(sting_estimates_chunk, params$estimates_rds)
```

------------------------------------------------------------------------

# Loaded estimates (preview)

```{r}
#| label: load-estimates
sting_estimates_chunk <- readRDS(params$estimates_rds)

sting_estimates_chunk |>
  dplyr::select(.feature, gene, compression_ok, compression_rate, reproduction_pct) |>
  head(10)
```

------------------------------------------------------------------------

# SAVI pseudobulk and healthy reference samples

**SAVI object:** RDS produced by `prepare_data.R` — pseudobulk counts aggregated with Seurat `AggregateExpression` by **`Sample`** and **`CellType`** (see GEO object metadata for treatment and category).

**Reference:** `pseudobulk_sample` from the same monocytic `targets` store is a `SummarizedExperiment` whose count assay is backed by the shared **Zenodo `pseudobulk_se.h5ad`** path. Rows with **`offset == 0`** are interpreted as healthy baseline samples for merge and posterior comparison (analogous to the breast epithelial reference in `pam50_toy_model.qmd`).

**Merge (same idea as `pam50_toy_model.qmd` § merge-se):** `SAVI_pseudobulk_sobj` is a **Seurat** object whose **RNA assay counts are row-named by gene symbol**. We map symbols → **Ensembl** (deduplicated), build a `SummarizedExperiment` with **`sample_role = "savi_pseudobulk"`**, then **column-bind** with reference columns after intersecting **shared Ensembl ids** and **assay names**.

```{r}
#| label: merge-savi-reference


SAVI_pseudobulk_sobj <- 
  readRDS(params$savi_pseudobulk_rds) %>% 
  filter(CellType == '17. Disease-associated monocytes')

pseudobulk_sample <- targets::tar_read(
  pseudobulk_sample,
  store = file.path(model_path, "_targets")
)

# Point DelayedArray/HDF5 backend to the local h5ad-backed matrix (path may differ on HPC)
pseudobulk_sample@assays@data@listData[["counts"]]@seed@seed@seed@filepath <-
  params$pseudobulk_h5ad

ref_sample <-
  pseudobulk_sample |>
  dplyr::filter(offset == 0)

SummarizedExperiment::colData(ref_sample)$sample_role <- "reference"

# --- SAVI: Seurat counts are rownames = gene symbols (see prepare_data.R) ---
counts_sym <- as.matrix(
  tryCatch(
    SeuratObject::LayerData(SAVI_pseudobulk_sobj, assay = "RNA", layer = "counts"),
    error = function(e) {
      Seurat::GetAssayData(SAVI_pseudobulk_sobj, assay = "RNA", slot = "counts")
    }
  )
)

sym <- rownames(counts_sym)
ensg <- AnnotationDbi::mapIds(
  org.Hs.eg.db::org.Hs.eg.db,
  keys = sym,
  column = "ENSEMBL",
  keytype = "SYMBOL",
  multiVals = "first"
)
keep <- !is.na(ensg) & !duplicated(ensg)
ensg <- ensg[keep]

counts_ens <- counts_sym[keep, , drop = FALSE]
rownames(counts_ens) <- ensg

cd_savi <- as.data.frame(SAVI_pseudobulk_sobj@meta.data)
rownames(cd_savi) <- colnames(counts_ens)
cd_savi$sample_role <- "savi_pseudobulk"
cd_savi$offset <- 0L
if (!"cell_type_unified_ensemble" %in% names(cd_savi)) {
  cd_savi$cell_type_unified_ensemble <- cd_savi$CellType
}

se_savi_ensembl <- SummarizedExperiment::SummarizedExperiment(
  assays = list(counts = counts_ens),
  colData = S4Vectors::DataFrame(cd_savi, row.names = colnames(counts_ens)),
  rowData = S4Vectors::DataFrame(
    gene_symbol = sym[keep],
    row.names = ensg
  )
)

# --- rbind-like merge of SE columns (intersect genes, union colData columns) ---
merge_summarized_experiments <- function(...) {
  ses <- list(...)
  assay_names <- Reduce(intersect, lapply(ses, SummarizedExperiment::assayNames))
  if (length(assay_names) == 0L) {
    stop("No shared assays between SummarizedExperiment objects.")
  }

  genes <- Reduce(intersect, lapply(ses, rownames))
  if (length(genes) == 0L) {
    stop("No shared rownames between SummarizedExperiment objects.")
  }

  ses <- lapply(ses, function(se) se[genes, , drop = FALSE])

  cds <- lapply(ses, function(se) as.data.frame(SummarizedExperiment::colData(se)))
  all_nms <- Reduce(union, lapply(cds, names))
  cds <- lapply(cds, function(cd) {
    missing <- setdiff(all_nms, names(cd))
    for (nm in missing) cd[[nm]] <- NA
    cd[, all_nms, drop = FALSE]
  })

  # Dense coercion: SAVI counts are an ordinary matrix; reference counts are often
  # HDF5Array/DelayedArray. Mixed cbind() hits bindCOLS() with no suitable method.
  assays <- stats::setNames(
    lapply(assay_names, function(an) {
      mats <- lapply(ses, function(se) {
        base::as.matrix(SummarizedExperiment::assay(se, an))
      })
      do.call(base::cbind, mats)
    }),
    assay_names
  )

  SummarizedExperiment::SummarizedExperiment(
    assays = assays,
    colData = S4Vectors::DataFrame(
      do.call(rbind, cds),
      row.names = unlist(lapply(ses, colnames), use.names = FALSE)
    ),
    rowData = SummarizedExperiment::rowData(ses[[1]])
  )
}

se_monocytic <- merge_summarized_experiments(
  se_savi_ensembl,
  ref_sample
)

# Copy abundance flag from reference rowData when a single .abundant* column exists
rd_ref <- as.data.frame(SummarizedExperiment::rowData(ref_sample))
abundant_cols <- grep("^\\.abundant", names(rd_ref), value = TRUE)
if (length(abundant_cols) >= 1L) {
  ac <- abundant_cols[[1]]
  SummarizedExperiment::rowData(se_monocytic)$.abundant <-
    SummarizedExperiment::rowData(ref_sample)[rownames(se_monocytic), ac, drop = TRUE]
}

tibble::tibble(
  n_features = nrow(se_monocytic),
  n_samples = ncol(se_monocytic),
  n_savi = sum(SummarizedExperiment::colData(se_monocytic)$sample_role == "savi_pseudobulk"),
  n_reference = sum(SummarizedExperiment::colData(se_monocytic)$sample_role == "reference")
)
```

```{r}
#| label: scale-abundance
ref_sample_name <-
  se_monocytic %>%
  colData() %>%
  as.data.frame() %>%
  filter(sample_role == "reference") %>%
  rownames()

se_monocytic <-
  se_monocytic %>%
  tidybulk::scale_abundance(
    method = "TMMwsp",
    reference_sample = ref_sample_name
  ) %>%
  mutate(offset = log(1 / multiplier))
```

------------------------------------------------------------------------

# SAVI sample annotations (GEO [GSE226598](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE226598))

Per-sample metadata (age, sex, 10x library batch, GSM accession) is taken from the GEO **series matrix** for GSE226598 (same fields as the GEO web “Sample characteristics” table). Each GSM links to its own GEO sample page.

After `scale_abundance`, we add harmonised columns for downstream models:

- **`tissue_groups`**: `"blood"` for SAVI pseudobulk columns only (PBMC in the study).
- **`sex`**, **`age_years`**: GEO values; **C1**, **C2** (“Adult”) and **P5_STING** (“Ped”) get **imputed numeric years** (see below) so `age_decade` / models can run without `NA`.
- **`age_decade`**: `as.character(ceiling(age_years / 10))` for SAVI once **`age_years`** is set (including imputed).
- **`age_group`**: `"pediatric"` if `age_years < 18`, else `"adult"` (SAVI columns only).
- **`age_decade_model`**: same as **`age_decade`** here (decade index from imputed / observed years).
- **`age_imputed`**: `TRUE` when **`age_years`** was filled by the rule below, not present as a number in GEO.

### Interpreting GEO **“Adult”** (C1, C2) and **“Ped”** (P5)

The cohort in de Cevins *et al.* (2023) comprises **five SAVI patients** and **seven healthy controls** (Figure 1A; GEO [GSE226598](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE226598)). GEO lists **“age: Adult”** for **C1** / **C2** and **“age: Ped”** for **P5_STING** without a year.

**Imputation used in this notebook** (rounded integers, documented for reproducibility):

| Sample(s) | GEO label | Imputed `age_years` | Rationale |
|-------------|-----------|---------------------|-----------|
| **C1**, **C2** | Adult | **35** | Mean of adult controls with numeric ages in the same series (**C8** 36 y, **C9** 35 y), aligned with matched adult controls. |
| **P5_STING** | Ped | **8** | Central value among other paediatric SAVI / control ages in GEO for this series (e.g. P1 4 y, P4 7–9 y, P6 8–10 y). |

All other donors keep **GEO-reported** ages. **`age_imputed`** is `TRUE` only for **C1**, **C2**, and **P5_STING**. Replace these constants if you obtain exact ages from authors or supplements.

- **`sex`**, **`assay_groups`**, **`assay_groups___altered`**: after GEO join, values are **harmonised to the Immune Atlas / posteriorHCA vocabulary** (same as `posteriorHCA::expr_predict()` / `Introduction` vignette: **`male`**, **`female`**, **`unknown`**; assay **`10x Genomics 3`**). GEO still uses **M**/**F** and Chromium **3′ v2**/**v3** batches in the series matrix; we do not keep the long Chromium strings in `colData` because the reference model was trained on short **`assay_groups`** labels.
- **`geo_accession`**, **`geo_sample_url`**: GSM id and direct GEO sample link (same pattern as the series page [GSE226598](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE226598)).

```{r}
#| label: coldata-gse226598
# Tibble from GEO series matrix GSE226598_series_matrix.txt.gz (Sample_title,
# Sample_geo_accession, age / Sex / batch rows). Series:
# https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE226598
geo_tbl <- tibble::tribble(
  ~Sample,              ~gse_accession, ~gse_age_years, ~gse_sex, ~gse_sting_batch,
  "C10",                "GSM7079979",            10,  "M",  "STING2",
  "C11",                "GSM7079980",            16,  "M",  "STING2",
  "C1",                 "GSM7079981",            NA,  "M",  "STING3",
  "C2",                 "GSM7079982",            NA,  "M",  "STING3",
  "C7",                 "GSM7079983",             5,  "F",  "STING1",
  "C8_aunt_P1_STING",   "GSM7079984",            36,  "F",  "STING1",
  "C9_mother_P1_STING", "GSM7079985",            35,  "F",  "STING1",
  "P1_STING_ht",        "GSM7079986",             4,  "F",  "STING1",
  "P1_STING_ht_T",      "GSM7079987",             8,  "F",  "STING1",
  "P1_STING_ht_T2",     "GSM7079988",             8,  "F",  "STING1",
  "P2_STING_ht",        "GSM7079989",            31,  "M",  "STING1",
  "P2_STING_ht_T",      "GSM7079990",            34,  "M",  "STING1",
  "P4_STING_ht",        "GSM7079991",             7,  "M",  "STING2",
  "P4_STING_ht_T",      "GSM7079992",             9,  "M",  "STING2",
  "P5_STING",           "GSM7079993",            NA,  "M",  "STING3",
  "P6_STING_ht",        "GSM7079994",             8,  "M",  "STING2",
  "P6_STING_ht_T",      "GSM7079995",            10,  "M",  "STING2"
) |>
  dplyr::mutate(
    # Raw GEO had NA for "Adult" (C1, C2) and "Ped" (P5); impute for modelling (see prose).
    gse_age_imputed = is.na(.data$gse_age_years),
    gse_age_years = dplyr::coalesce(
      .data$gse_age_years,
      dplyr::case_when(
        .data$Sample %in% c("C1", "C2") ~ 35,
        .data$Sample == "P5_STING" ~ 8,
        TRUE ~ NA_real_
      )
    )
  ) |>
  dplyr::mutate(
    gse_geo_sample_url = sprintf(
      "https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=%s",
      .data$gse_accession
    )
  ) |>
  dplyr::select(
    "Sample",
    "gse_accession",
    "gse_geo_sample_url",
    "gse_age_years",
    "gse_age_imputed",
    "gse_sex"
  )

cd <- SummarizedExperiment::colData(se_monocytic) |> as.data.frame()
rn <- rownames(cd)

if (!"Sample" %in% names(cd)) {
  stop("colData must contain `Sample` (from SAVI pseudobulk) to join GEO metadata.")
}

if (!"tissue_groups" %in% names(cd)) {
  cd$tissue_groups <- NA_character_
}
if (!"age_years" %in% names(cd)) {
  cd$age_years <- NA_real_
}
if (!"sex" %in% names(cd)) {
  cd$sex <- NA_character_
}
if (!"assay_groups" %in% names(cd)) {
  cd$assay_groups <- NA_character_
}
if (!"assay_groups___altered" %in% names(cd)) {
  cd$assay_groups___altered <- NA_character_
}

cd <- cd |>
  dplyr::left_join(geo_tbl, by = "Sample")

miss_geo <- is.na(cd$gse_accession) & cd$sample_role == "savi_pseudobulk"
if (any(miss_geo, na.rm = TRUE)) {
  warning(
    "SAVI columns with no GEO match (check `Sample` vs GEO Sample_title): ",
    paste(unique(cd$Sample[miss_geo]), collapse = ", ")
  )
}

sex_ref <- if ("Sex" %in% names(cd)) cd$Sex else rep(NA_character_, nrow(cd))

cd <- cd |>
  dplyr::mutate(
    tissue_groups = dplyr::if_else(
      .data$sample_role == "savi_pseudobulk",
      "blood",
      .data$tissue_groups
    ),
    sex = dplyr::case_when(
      .data$sample_role == "savi_pseudobulk" ~ .data$gse_sex,
      TRUE ~ dplyr::coalesce(.data$sex, sex_ref)
    ),
    age_years = dplyr::case_when(
      .data$sample_role == "savi_pseudobulk" ~ .data$gse_age_years,
      TRUE ~ .data$age_years
    ),
    age_decade = dplyr::if_else(
      is.na(.data$age_years),
      NA_character_,
      as.character(ceiling(.data$age_years / 10))
    ),
    age_group = dplyr::case_when(
      .data$sample_role != "savi_pseudobulk" ~ NA_character_,
      !is.na(.data$age_years) & .data$age_years < 18 ~ "pediatric",
      !is.na(.data$age_years) & .data$age_years >= 18 ~ "adult",
      TRUE ~ NA_character_
    ),
    age_decade_model = .data$age_decade,
    age_imputed = dplyr::if_else(
      .data$sample_role == "savi_pseudobulk",
      .data$gse_age_imputed,
      NA
    ),
    assay_groups = dplyr::if_else(
      .data$sample_role == "savi_pseudobulk",
      "10x Genomics 3",
      .data$assay_groups
    ),
    assay_groups___altered = dplyr::case_when(
      .data$sample_role == "savi_pseudobulk" ~ "10x Genomics 3",
      !is.na(.data$assay_groups___altered) ~ .data$assay_groups___altered,
      !is.na(.data$assay_groups) ~ .data$assay_groups,
      TRUE ~ NA_character_
    ),
    geo_accession = dplyr::if_else(
      .data$sample_role == "savi_pseudobulk",
      .data$gse_accession,
      NA_character_
    ),
    geo_sample_url = dplyr::if_else(
      .data$sample_role == "savi_pseudobulk",
      .data$gse_geo_sample_url,
      NA_character_
    )
  ) |>
  dplyr::select(
    -dplyr::any_of(c(
      "gse_accession",
      "gse_geo_sample_url",
      "gse_age_years",
      "gse_age_imputed",
      "gse_sex",
      "Sex"
    ))
  ) |>
  dplyr::mutate(
    sex = dplyr::case_when(
      is.na(.data$sex) ~ NA_character_,
      .data$sex %in% c("male", "female", "unknown") ~ as.character(.data$sex),
      toupper(trimws(as.character(.data$sex))) %in% c("M", "MALE") ~ "male",
      toupper(trimws(as.character(.data$sex))) %in% c("F", "FEMALE") ~ "female",
      TRUE ~ "unknown"
    )
  )

SummarizedExperiment::colData(se_monocytic) <- S4Vectors::DataFrame(cd, row.names = rn)

cd |>
  dplyr::filter(.data$sample_role == "savi_pseudobulk") |>
  dplyr::distinct(
    .data$Sample,
    .data$tissue_groups,
    .data$sex,
    .data$age_years,
    .data$age_decade,
    .data$age_group,
    .data$age_decade_model,
    .data$age_imputed,
    .data$assay_groups,
    .data$assay_groups___altered,
    .data$geo_accession,
    .data$geo_sample_url
  ) |>
  dplyr::arrange(.data$Sample)
```

GEO sample pages use the same host as the series record above; each **`geo_sample_url`** points to the corresponding GSM (e.g. patient / control library) listed under [GSE226598](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE226598).

------------------------------------------------------------------------

# Compare samples to healthy posterior predictive baseline

For each **STING signature gene** with a loadable compressed `brms` fit, we draw replicated counts from **`brms::posterior_predict()`** at the sample’s **`offset`** and covariates in **`colData`** (same pattern as `pam50_toy_model.qmd` § *Compare samples*). **`healthy_median`** / **`healthy_hpd_*`** summarise the predictive distribution; **`direction`** flags counts outside the central **95%** predictive mass (**two-sided** `posterior_tail_p` is for exploration only—no multiple-testing correction across genes × samples).

We use **`posterior_predict`** (not **`posterior_epred`**): we want the predictive distribution of a **new observation** including likelihood noise, because each test asks whether the observed count is in the tail of a healthy sample, not whether the latent mean differs. We also use **`re_formula = NULL`** with **`allow_new_levels = TRUE`** and **`sample_new_levels = "uncertainty"`** so that:

- `(1 + age_decade * sex + ethnicity_groups | tissue_groups)` keeps the trained **`blood`** slopes (tissue is observed).
- `(1 | dataset_id_altered)` is marginalised: a fresh dataset intercept is drawn from `N(0, sd_dataset_id_altered)` for every posterior sample — the correct "new SAVI cohort, exchangeable with the atlas" null.
- Unknown **`ethnicity_groups`** is marginalised explicitly by expanding `newdata` over all ethnicities seen in the model's training data and pooling draws (uniform mixture). Setting ethnicity to a single arbitrary value would bias the null toward that group; zeroing the term in the design matrix is not possible for a categorical fixed effect.

```{r}
#| label: gene-deviation-fn

# Categorical levels for any predictor pulled from the brms fit's training
# frame. Used to expand newdata over training-set values when a covariate is
# unknown for a test sample.
fit_levels <- function(fit, var) {
  d <- fit$data
  if (is.null(d) || !var %in% names(d)) return(character(0))
  v <- d[[var]]
  lv <- if (is.factor(v)) levels(v) else sort(unique(as.character(v)))
  lv[!is.na(lv) & nzchar(lv)]
}

# First non-NA, non-empty character among the inputs; NA_character_ otherwise.
coalesce_char <- function(...) {
  args <- list(...)
  for (v in args) {
    if (length(v) >= 1L) {
      val <- as.character(v[[1L]])
      if (!is.na(val) && nzchar(val)) return(val)
    }
  }
  NA_character_
}

# Build a single-row newdata template from colData. Columns that may be NA are
# left NA on purpose; build_newdata_grid() will marginalise over training-set
# levels when the model uses them.
newdata_for_sample <- function(se, sample_idx) {
  cd <- SummarizedExperiment::colData(se)[sample_idx, , drop = FALSE] |>
    as.data.frame()

  tibble::tibble(
    # Observed when available (may be NA -> marginalise).
    age_decade = coalesce_char(
      if ("age_decade_model" %in% names(cd)) cd$age_decade_model,
      if ("age_decade" %in% names(cd)) cd$age_decade
    ),
    sex = coalesce_char(if ("sex" %in% names(cd)) cd$sex),
    ethnicity_groups = coalesce_char(
      if ("ethnicity_groups" %in% names(cd)) cd$ethnicity_groups,
      if ("ethnicity_groups_imputed" %in% names(cd)) cd$ethnicity_groups_imputed
    ),

    # Counterfactual / required: forced to the healthy baseline so we test
    # against the atlas null, not the sample's own disease label.
    disease_groups_altered = "Normal",
    # disease_groups_altered = coalesce_char(
    #   if ("disease_groups_altered" %in% names(cd)) cd$disease_groups_altered,
    #   if ("disease_groups___altered" %in% names(cd)) cd$disease_groups___altered,
    #   "Normal"
    # ),


    # Required (atlas-defined / technical):
    assay_groups_altered = coalesce_char(
      if ("assay_groups_altered" %in% names(cd)) cd$assay_groups_altered,
      if ("assay_groups___altered" %in% names(cd)) cd$assay_groups___altered,
      if ("assay_groups" %in% names(cd)) cd$assay_groups,
      "10x Genomics 3"
    ),
    tissue_groups = coalesce_char(if ("tissue_groups" %in% names(cd)) cd$tissue_groups),
    offset = cd$offset[[1]]
  )
}

# Expand a single-row newdata into a prediction grid:
#   - required cols MUST be non-NA (e.g. tissue_groups)
#   - marginalize_over cols that are NA in newdata AND present in fit$data
#     are expanded across the training-set levels (uniform mixture)
#   - new_level_cols (e.g. dataset_id_altered) are stamped with a fresh unique
#     placeholder so brms (allow_new_levels = TRUE, sample_new_levels =
#     "uncertainty") draws a new group effect from its posterior population.
build_newdata_grid <- function(
  fit,
  newdata,
  marginalize_over = c("sex", "age_decade", "ethnicity_groups", "assay_groups_altered"),
  required = c("tissue_groups", "offset", "disease_groups_altered"),
  new_level_cols = "dataset_id_altered",
  new_level_tag = NULL
) {
  stopifnot(NROW(newdata) == 1L)
  newdata <- tibble::as_tibble(newdata)

  is_blank <- function(v) {
    length(v) == 0L ||
      is.na(v[[1L]]) ||
      (is.character(v[[1L]]) && !nzchar(v[[1L]]))
  }

  fit_vars <- colnames(fit$data)

  # Required columns must be present and non-empty.
  for (col in required) {
    if (!col %in% fit_vars) next
    if (!col %in% names(newdata) || is_blank(newdata[[col]])) {
      stop(sprintf(
        "Required predictor `%s` is missing or empty in newdata.",
        col
      ))
    }
  }

  # Decide which columns to expand.
  to_expand <- character(0)
  for (col in marginalize_over) {
    if (!col %in% fit_vars) next
    if (!col %in% names(newdata) || is_blank(newdata[[col]])) {
      to_expand <- c(to_expand, col)
    }
  }

  if (length(to_expand)) {
    levels_list <- lapply(to_expand, function(col) fit_levels(fit, col))
    names(levels_list) <- to_expand
    keep <- vapply(levels_list, length, integer(1L)) > 0L
    levels_list <- levels_list[keep]
    to_expand <- names(levels_list)
  }

  if (length(to_expand)) {
    grid <- do.call(tidyr::expand_grid, levels_list)
    rest <- newdata[, setdiff(names(newdata), to_expand), drop = FALSE]
    out <- dplyr::bind_cols(
      rest[rep(1L, nrow(grid)), , drop = FALSE],
      grid
    )
  } else {
    out <- newdata
  }

  if (is.null(new_level_tag)) {
    new_level_tag <- sprintf(
      "__new__%d_%d",
      as.integer(Sys.time()),
      sample.int(1e6, 1L)
    )
  }
  for (col in new_level_cols) {
    if (col %in% fit_vars) out[[col]] <- new_level_tag
  }

  tibble::as_tibble(out)
}

# Posterior predictive comparison of obs against a brms fit.
#   - posterior_predict (NOT epred): include likelihood noise so tail tests
#     are calibrated against new-observation predictive intervals.
#   - re_formula = NULL: keep tissue_groups random slopes (observed level).
#   - allow_new_levels + sample_new_levels = "uncertainty": dataset_id_altered
#     gets a new draw from its random-intercept population per posterior draw.
#   - marginalize_over: unknown covariates are integrated out by uniform
#     mixture over training-set levels.
gene_deviation_from_fit <- function(
  fit,
  newdata,
  obs,
  cred_mass = 0.95,
  alpha = 0.025,
  marginalize_over = c("sex", "age_decade", "ethnicity_groups", "assay_groups_altered"),
  required = c("tissue_groups", "offset", "disease_groups_altered"),
  new_level_cols = "dataset_id_altered",
  new_level_tag = NULL,
  re_formula = NULL,
  allow_new_levels = TRUE,
  sample_new_levels = "uncertainty"
) {
  nd_grid <- build_newdata_grid(
    fit,
    newdata,
    marginalize_over = marginalize_over,
    required = required,
    new_level_cols = new_level_cols,
    new_level_tag = new_level_tag
  )

  y_rep_mat <- brms::posterior_predict(
    fit,
    newdata = nd_grid,
    summary = FALSE,
    re_formula = re_formula,
    allow_new_levels = allow_new_levels,
    sample_new_levels = sample_new_levels
  )

  y_rep <- as.numeric(y_rep_mat)

  hpd <- if (requireNamespace("HDInterval", quietly = TRUE)) {
    HDInterval::hdi(y_rep, credMass = cred_mass)
  } else {
    probs <- c((1 - cred_mass) / 2, 1 - (1 - cred_mass) / 2)
    stats::quantile(y_rep, probs = probs, names = FALSE)
  }

  p_above <- mean(y_rep >= obs)
  p_below <- mean(y_rep <= obs)

  tibble::tibble(
    observed_count = obs,
    healthy_median = median(y_rep),
    healthy_mean = mean(y_rep),
    healthy_hpd_lo = hpd[1],
    healthy_hpd_hi = hpd[2],
    obs_percentile = mean(y_rep <= obs),
    p_above_healthy = p_above,
    p_below_healthy = p_below,
    posterior_tail_p = 2 * min(p_above, p_below),
    log2fc_vs_median = log2((obs + 1) / (median(y_rep) + 1)),
    n_grid_rows = NROW(nd_grid),
    direction = dplyr::case_when(
      p_above < alpha ~ "higher_than_healthy",
      p_below < alpha ~ "lower_than_healthy",
      TRUE ~ "consistent_with_healthy"
    )
  )
}
```

```{r}
#| label: comparison-samples
cd_mono <-
  se_monocytic |>
  SummarizedExperiment::colData() |>
  as.data.frame() |>
  tibble::rownames_to_column("sample")

if (!length(ref_sample_name)) {
  stop("No reference columns (offset == 0) found; cannot define comparison baseline.")
}

comparison_samples_tbl <- cd_mono |>
  dplyr::filter(.data$sample_role == "savi_pseudobulk") |>
  dplyr::bind_rows(
    cd_mono |>
      dplyr::filter(.data$sample == !!ref_sample_name[[1]])
  ) |>
  dplyr::distinct(.data$sample, .keep_all = TRUE) |>
  dplyr::mutate(sample_idx = match(.data$sample, colnames(se_monocytic)))

tibble::tibble(
  n_genes_fit = sum(sting_estimates_chunk$compression_ok == TRUE, na.rm = TRUE),
  n_comparison_samples = nrow(comparison_samples_tbl),
  n_savi = sum(comparison_samples_tbl$sample_role == "savi_pseudobulk"),
  n_reference = sum(comparison_samples_tbl$sample_role == "reference")
)
```

```{r}
#| label: predict-sting-vs-healthy
#| cache: false

handlers("progress")

sting_fit_rows <- sting_estimates_chunk |>
  dplyr::filter(.data$compression_ok == TRUE) |>
  dplyr::filter(lengths(.data$brms_fit_compressed) > 0L) |>
  dplyr::filter(.data$.feature %in% rownames(se_monocytic))

if (!nrow(sting_fit_rows)) {
  stop("No compressed biomarker fits available; run with params$run_fits: true or fix estimates_rds.")
}

with_progress({
  p <- progressr::progressor(steps = nrow(sting_fit_rows) * nrow(comparison_samples_tbl))

  gene_deviation_all <- purrr::imap_dfr(
    seq_len(nrow(sting_fit_rows)),
    function(i, idx) {
      chunk_row <- sting_fit_rows[i, ]
      feature_id <- chunk_row$.feature
      gene_sym <- chunk_row$gene
      fit <- poco::reconstruct_brmsfit(chunk_row$brms_fit_compressed[[1]])

      purrr::pmap_dfr(
        comparison_samples_tbl |>
          dplyr::select("sample", "sample_idx", "sample_role"),
        function(sample, sample_idx, sample_role) {
          p()

          newdata <- newdata_for_sample(se_monocytic, sample_idx)
          obs <-
            SummarizedExperiment::assay(
              se_monocytic[feature_id, sample_idx, drop = FALSE],
              "counts"
            ) |>
            as.numeric()

          gene_deviation_from_fit(
            fit,
            newdata,
            obs,
            new_level_tag = paste0("__savi__", as.character(sample))
            # sample_new_levels = 'gaussian'
          ) |>
            dplyr::mutate(
              .feature = feature_id,
              gene = gene_sym,
              sample = as.character(sample),
              sample_role = as.character(sample_role),
              .before = "observed_count"
            )
        }
      )
    }
  )
})

gene_deviation_all <- gene_deviation_all |>
  tibble::as_tibble() |>
  dplyr::select(tidyselect::where(\(x) !is.list(x)))

if (!"sample_role" %in% colnames(gene_deviation_all)) {
  gene_deviation_all <- gene_deviation_all |>
    dplyr::left_join(
      comparison_samples_tbl |>
        dplyr::select("sample", "sample_role"),
      by = "sample"
    )
}

gene_deviation_all
```

```{r}
#| label: deviation-summary-sting
gene_deviation_all |>
  dplyr::count(.data$sample_role, .data$direction) |>
  tidyr::pivot_wider(
    names_from = "direction",
    values_from = "n",
    values_fill = 0L
  )

gene_deviation_savi <- gene_deviation_all |>
  dplyr::filter(.data$sample_role == "savi_pseudobulk")

gene_deviation_savi |>
  dplyr::group_by(.data$gene) |>
  dplyr::summarise(
    n_higher = sum(.data$direction == "higher_than_healthy"),
    n_lower = sum(.data$direction == "lower_than_healthy"),
    n_consistent = sum(.data$direction == "consistent_with_healthy"),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(.data$n_higher + .data$n_lower))
```

------------------------------------------------------------------------

# Evaluate deviations against the published STING signature

De Cevins *et al.* (2023) define a **21-gene STING-activation module** that is **up in disease-associated monocytes from SAVI patients relative to healthy controls** in the same cohort (Figure 6; Table S6)—after removing IFN-β–inducible genes. Our posterior test asks a **different** question: whether each pseudobulk count lies above the **Immune Atlas healthy predictive** distribution for monocytic cells, given donor metadata. We still expect:

| `Category` (study cohort) | Paper-consistent expectation for signature genes |
|---------------------------|---------------------------------------------------|
| **`SAVI`** | Mostly **`higher_than_healthy`** vs atlas baseline |
| **`CTRL`** | Mostly **`consistent_with_healthy`** (study controls without STING activation) |
| **`SAVI_treated`** | Intermediate—treatment samples often move toward baseline (we score **`consistent_with_healthy`** as the nominal match, same as controls) |

Sample **`Category`** comes from the GEO Seurat object (`prepare_data.R`) and is carried in **`colData(se_monocytic)`** (`CTRL`, `SAVI`, `SAVI_treated`; reference columns are **`NA`**).

```{r}
#| label: sting-literature-eval
sample_annotations <- cd_mono |>
  dplyr::filter(.data$sample_role == "savi_pseudobulk") |>
  dplyr::select(
    "sample",
    "Sample",
    "Category",
    tidyselect::any_of("Treatment")
  ) |>
  dplyr::distinct() |>
  dplyr::mutate(
    Category = factor(
      .data$Category,
      levels = c("CTRL", "SAVI", "SAVI_treated")
    )
  )

category_expected <- tibble::tribble(
  ~Category,        ~expected_deviation,
  "CTRL",           "consistent_with_healthy",
  "SAVI",           "higher_than_healthy",
  "SAVI_treated",   "consistent_with_healthy"
) |>
  dplyr::mutate(Category = factor(.data$Category, levels = levels(sample_annotations$Category)))

gene_deviation_eval <- gene_deviation_savi |>
  dplyr::left_join(sample_annotations, by = "sample") |>
  dplyr::left_join(category_expected, by = "Category") |>
  dplyr::mutate(
    deviation_score = dplyr::case_when(
      .data$direction == "higher_than_healthy" ~ 1L,
      .data$direction == "lower_than_healthy" ~ -1L,
      TRUE ~ 0L
    ),
    matches_literature = dplyr::case_when(
      is.na(.data$expected_deviation) ~ NA,
      .data$expected_deviation == .data$direction ~ TRUE,
      TRUE ~ FALSE
    ),
    signature_up = .data$direction == "higher_than_healthy"
  )

signature_by_sample <- gene_deviation_eval |>
  dplyr::group_by(.data$sample, .data$Sample, .data$Category) |>
  dplyr::summarise(
    n_genes = dplyr::n(),
    n_higher = sum(.data$signature_up),
    n_lower = sum(.data$direction == "lower_than_healthy"),
    n_consistent = sum(.data$direction == "consistent_with_healthy"),
    pct_higher = 100 * .data$n_higher / .data$n_genes,
    pct_match = 100 * mean(.data$matches_literature, na.rm = TRUE),
    mean_log2fc = mean(.data$log2fc_vs_median, na.rm = TRUE),
    signature_score = mean(.data$deviation_score, na.rm = TRUE),
    .groups = "drop"
  )

agreement_by_category <- gene_deviation_eval |>
  dplyr::filter(!is.na(.data$Category), !is.na(.data$expected_deviation)) |>
  dplyr::group_by(.data$Category, .data$expected_deviation) |>
  dplyr::summarise(
    n_tests = dplyr::n(),
    n_match = sum(.data$matches_literature, na.rm = TRUE),
    pct_match = 100 * .data$n_match / .data$n_tests,
    pct_higher = 100 * mean(.data$signature_up),
    pct_consistent = 100 * mean(.data$direction == "consistent_with_healthy"),
    .groups = "drop"
  )

agreement_by_category
```

```{r}
#| label: plot-signature-by-category
#| fig-width: 12
#| fig-height: 5
#| fig-align: center

category_cols <- c(
  CTRL = "#2ca25f",
  SAVI = "#de2d26",
  SAVI_treated = "#756bb1"
)

p_pct_higher <- ggplot2::ggplot(
  signature_by_sample,
  ggplot2::aes(
    x = .data$Category,
    y = .data$pct_higher,
    fill = .data$Category
  )
) +
  ggplot2::geom_boxplot(
    outlier.shape = NA,
    alpha = 0.35,
    width = 0.55,
    colour = "grey30"
  ) +
  ggplot2::geom_jitter(
    ggplot2::aes(colour = .data$Category),
    width = 0.12,
    size = 2.5,
    show.legend = FALSE
  ) +
  ggplot2::scale_fill_manual(values = category_cols) +
  ggplot2::scale_colour_manual(values = category_cols) +
  ggplot2::labs(
    x = NULL,
    y = "% signature genes higher than atlas baseline",
    title = "STING signature enrichment by study Category",
    subtitle = "Paper: up in SAVI monocytes vs controls (Figure 6)"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "none")

p_signature_score <- ggplot2::ggplot(
  signature_by_sample,
  ggplot2::aes(
    x = .data$Category,
    y = .data$signature_score,
    fill = .data$Category
  )
) +
  ggplot2::geom_boxplot(
    outlier.shape = NA,
    alpha = 0.35,
    width = 0.55,
    colour = "grey30"
  ) +
  ggplot2::geom_jitter(
    ggplot2::aes(colour = .data$Category),
    width = 0.12,
    size = 2.5,
    show.legend = FALSE
  ) +
  ggplot2::geom_hline(yintercept = 0, linetype = 2, colour = "grey50") +
  ggplot2::scale_fill_manual(values = category_cols) +
  ggplot2::scale_colour_manual(values = category_cols) +
  ggplot2::labs(
    x = NULL,
    y = "Mean deviation score\n(−1 lower, 0 consistent, +1 higher)",
    title = "Per-sample composite STING deviation score"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "none")

p_pct_higher / p_signature_score +
  patchwork::plot_layout(ncol = 2)
```

```{r}
#| label: plot-literature-agreement-sting
#| fig-width: 8
#| fig-height: 4
#| fig-align: center

ggplot2::ggplot(
  agreement_by_category,
  ggplot2::aes(
    x = .data$Category,
    y = .data$pct_match,
    fill = .data$Category
  )
) +
  ggplot2::geom_col(width = 0.65, colour = "white") +
  ggplot2::geom_text(
    ggplot2::aes(
      label = sprintf(
        "%.0f%% match\n(%.0f%% higher)",
        .data$pct_match,
        .data$pct_higher
      )
    ),
    vjust = -0.15,
    size = 3.2
  ) +
  ggplot2::scale_fill_manual(values = category_cols) +
  ggplot2::scale_y_continuous(limits = c(0, 115), expand = c(0, 0)) +
  ggplot2::labs(
    x = NULL,
    y = "% gene × sample tests matching category expectation",
    title = "Posterior deviation vs paper-consistent expectation",
    subtitle = "Bars = % of gene × sample tests matching expected direction, pooled within each Category (not mean expression)."
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "none")
```

```{r}
#| label: heatmap-sting-category
#| fig-width: 15
#| fig-height: 12
#| fig-align: center

heatmap_df <- gene_deviation_eval |>
  dplyr::filter(!is.na(.data$Category)) |>
  dplyr::mutate(
    Category = factor(.data$Category, levels = names(category_cols))
  )

mat_log2fc <- heatmap_df |>
  dplyr::select("gene", "sample", "log2fc_vs_median") |>
  tidyr::pivot_wider(names_from = "sample", values_from = "log2fc_vs_median") |>
  tibble::column_to_rownames("gene") |>
  as.matrix()

score_mat <- heatmap_df |>
  dplyr::select("gene", "sample", "deviation_score") |>
  tidyr::pivot_wider(names_from = "sample", values_from = "deviation_score") |>
  tibble::column_to_rownames("gene") |>
  as.matrix()

# Category only in annotation_col: pheatmap draws one legend per annotated
# column; adding Sample (~17 levels, no ann_colors) stacks a second legend and
# the "Category" title overlaps the colour keys (see pheatmap issue with multi-legends).
annotation_col <- heatmap_df |>
  dplyr::distinct(.data$sample, .data$Category) |>
  tibble::column_to_rownames("sample") |>
  dplyr::select("Category")

sample_labels <- heatmap_df |>
  dplyr::distinct(.data$sample, .data$Sample) |>
  tibble::deframe()

labels_col <- unname(sample_labels[colnames(mat_log2fc)])

ann_colors <- list(Category = category_cols)

lim_log2 <- max(abs(mat_log2fc), na.rm = TRUE)
if (!is.finite(lim_log2) || lim_log2 == 0) lim_log2 <- 1

pheatmap::pheatmap(
  mat_log2fc,
  annotation_col = annotation_col,
  annotation_colors = ann_colors,
  labels_col = labels_col,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  clustering_method = "ward.D2",
  color = grDevices::colorRampPalette(rev(RColorBrewer::brewer.pal(11, "RdBu")))(100),
  breaks = seq(-lim_log2, lim_log2, length.out = 101),
  border_color = NA,
  fontsize_row = 8,
  fontsize_col = 7,
  angle_col = 45,
  main = "log2(obs / healthy predictive median)\nSTING genes × SAVI pseudobulk (Category annotated)"
)

pheatmap::pheatmap(
  score_mat,
  annotation_col = annotation_col,
  annotation_colors = ann_colors,
  labels_col = labels_col,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  clustering_method = "ward.D2",
  color = grDevices::colorRampPalette(c("#2166ac", "#f7f7f7", "#b2182b"))(3),
  breaks = c(-1.5, -0.5, 0.5, 1.5),
  legend_breaks = c(-1, 0, 1),
  legend_labels = c("lower", "consistent", "higher"),
  border_color = NA,
  fontsize_row = 8,
  fontsize_col = 7,
  angle_col = 45,
  main = "Deviation direction vs Immune Atlas baseline\n(Category = CTRL / SAVI / SAVI_treated)"
)
```

```{r}
#| label: plot-gene-match-by-category
#| fig-width: 10
#| fig-height: 6
#| fig-align: center

gene_match_summary <- gene_deviation_eval |>
  dplyr::filter(!is.na(.data$Category)) |>
  dplyr::group_by(.data$gene, .data$Category, .data$expected_deviation) |>
  dplyr::summarise(
    n = dplyr::n(),
    pct_higher = 100 * mean(.data$signature_up),
    pct_match = 100 * mean(.data$matches_literature, na.rm = TRUE),
    mean_log2fc = mean(.data$log2fc_vs_median, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    gene = forcats::fct_reorder(.data$gene, .data$mean_log2fc, .fun = mean, .desc = TRUE)
  )

ggplot2::ggplot(
  gene_match_summary,
  ggplot2::aes(
    x = .data$gene,
    y = .data$Category,
    fill = .data$mean_log2fc
  )
) +
  ggplot2::geom_tile(colour = "white", linewidth = 0.3) +
  ggplot2::scale_fill_gradient2(
    low = "#2166ac",
    mid = "#f7f7f7",
    high = "#b2182b",
    midpoint = 0,
    name = "Mean log2 FC\nvs baseline"
  ) +
  ggplot2::labs(
    x = NULL,
    y = NULL,
    title = "STING signature genes × study Category",
    subtitle = "Tile = mean log2(obs / healthy predictive median) across samples in each Category"
  ) +
  ggplot2::theme_minimal(base_size = 10) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, size = 7),
    panel.grid = ggplot2::element_blank()
  )
```

```{r}
#| label: eval-summary-table
gene_deviation_eval |>
  dplyr::filter(!is.na(.data$Category)) |>
  dplyr::group_by(.data$Category, .data$gene) |>
  dplyr::summarise(
    n_samples = dplyr::n_distinct(.data$sample),
    n_higher = sum(.data$signature_up),
    pct_higher = 100 * mean(.data$signature_up),
    matches_paper = all(.data$matches_literature %in% TRUE),
    .groups = "drop"
  ) |>
  dplyr::group_by(.data$Category) |>
  dplyr::summarise(
    n_genes_tested = dplyr::n(),
    n_genes_all_higher = sum(.data$n_higher == .data$n_samples),
    n_genes_majority_higher = sum(.data$pct_higher > 50),
    n_genes_paper_perfect = sum(.data$matches_paper),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    agreement_by_category |>
      dplyr::select("Category", "pct_match", "pct_higher"),
    by = "Category"
  )
```

------------------------------------------------------------------------

# Biomarker quality check (atlas-based falsification)

De Cevins *et al.* (2023) propose the **21-gene signature** by **within-cohort comparison** (5 SAVI vs 7 in-study controls). With only seven matched donors, a gene that is simply **variable in healthy people** but happened to be low in their controls can pass that filter. We can sanity-check each signature gene against a much broader **healthy reference** (Immune Atlas monocytic posteriors), which is the kind of falsification their paper could not do. The expected pattern under a true SAVI biomarker is:

- **SAVI** samples → consistently **higher** than the atlas healthy distribution
- **SAVI_treated** samples → **intermediate** (attenuated by treatment)
- **CTRL** samples → **consistent with healthy** (atlas baseline matches the in-cohort controls)

A gene fails the check (paper-supported but **not** atlas-supported) when:

- **CTRL drifts up** vs atlas (gene is high in many healthy people, not just SAVI),
- **No SAVI lift** vs CTRL (paper effect is within healthy variability),
- **SAVI_treated ≥ SAVI** (no treatment effect even though paper claims one).

```{r}
#| label: biomarker-per-category
biomarker_per_category <- gene_deviation_eval |>
  dplyr::filter(!is.na(.data$Category)) |>
  dplyr::group_by(.data$gene, .data$Category) |>
  dplyr::summarise(
    n_samples = dplyr::n_distinct(.data$sample),
    mean_log2fc = mean(.data$log2fc_vs_median, na.rm = TRUE),
    median_log2fc = stats::median(.data$log2fc_vs_median, na.rm = TRUE),
    pct_higher = 100 * mean(.data$direction == "higher_than_healthy"),
    pct_lower = 100 * mean(.data$direction == "lower_than_healthy"),
    pct_consistent = 100 * mean(.data$direction == "consistent_with_healthy"),
    mean_p_above_healthy = mean(.data$p_above_healthy, na.rm = TRUE),
    .groups = "drop"
  )

biomarker_per_category |>
  dplyr::arrange(.data$gene, .data$Category)
```

```{r}
#| label: biomarker-quality-score
biomarker_quality <- biomarker_per_category |>
  dplyr::select(
    "gene",
    "Category",
    "mean_log2fc",
    "pct_higher",
    "pct_consistent"
  ) |>
  tidyr::pivot_wider(
    names_from = "Category",
    values_from = c("mean_log2fc", "pct_higher", "pct_consistent"),
    values_fill = NA_real_,
    names_sep = "."
  ) |>
  dplyr::mutate(
    savi_lift = .data$mean_log2fc.SAVI,
    ctrl_drift = .data$mean_log2fc.CTRL,
    treated_lift = .data$mean_log2fc.SAVI_treated,

    # SAVI vs in-cohort controls' atlas-corrected position (paper's main claim).
    savi_vs_ctrl = .data$savi_lift - .data$ctrl_drift,

    # Treatment effect: positive means SAVI_treated drops back toward baseline.
    treatment_attenuation = .data$savi_lift - .data$treated_lift,

    # Atlas baseline agreement of in-study controls (closer to 0 = better).
    abs_ctrl_drift = abs(.data$ctrl_drift),

    # Composite quality: reward SAVI > CTRL, reward treatment drop,
    # penalise CTRL drift from atlas. Weights chosen so all three terms are on
    # the log2 scale; tweak if needed.
    quality_score = .data$savi_vs_ctrl +
      0.5 * .data$treatment_attenuation -
      0.5 * .data$abs_ctrl_drift,

    biomarker_class = dplyr::case_when(
      .data$pct_higher.SAVI >= 60 &
        .data$pct_consistent.CTRL >= 50 &
        .data$treatment_attenuation > 0 ~ "validated",
      .data$pct_higher.SAVI >= 60 &
        .data$treatment_attenuation > 0 &
        .data$pct_consistent.CTRL < 50 ~ "validated_ctrl_drifts",
      .data$pct_higher.SAVI >= 60 &
        .data$treatment_attenuation <= 0 ~ "savi_up_no_treatment_effect",
      .data$pct_higher.SAVI < 30 &
        .data$ctrl_drift > 0.5 ~ "falsified_baseline_high",
      .data$pct_higher.SAVI < 30 ~ "falsified_no_savi_signal",
      TRUE ~ "ambiguous"
    )
  ) |>
  dplyr::arrange(dplyr::desc(.data$quality_score))

biomarker_quality |>
  dplyr::select(
    "gene",
    "biomarker_class",
    "quality_score",
    "savi_vs_ctrl",
    "treatment_attenuation",
    "abs_ctrl_drift",
    dplyr::starts_with("pct_higher."),
    dplyr::starts_with("pct_consistent.")
  )
```

```{r}
#| label: plot-biomarker-ranking
#| fig-width: 9
#| fig-height: 7
#| fig-align: center

gene_order <- biomarker_quality$gene

biomarker_long <- biomarker_per_category |>
  dplyr::filter(.data$gene %in% gene_order) |>
  dplyr::mutate(
    gene = factor(.data$gene, levels = rev(gene_order)),
    Category = factor(
      .data$Category,
      levels = c("CTRL", "SAVI_treated", "SAVI")
    )
  )

ggplot2::ggplot(
  biomarker_long,
  ggplot2::aes(
    x = .data$mean_log2fc,
    y = .data$gene,
    colour = .data$Category
  )
) +
  ggplot2::geom_vline(xintercept = 0, linetype = 2, colour = "grey60") +
  ggplot2::geom_line(
    ggplot2::aes(group = .data$gene),
    colour = "grey80",
    linewidth = 0.4
  ) +
  ggplot2::geom_point(size = 3, alpha = 0.85) +
  ggplot2::scale_colour_manual(values = category_cols) +
  ggplot2::labs(
    x = "Mean log2(obs / healthy predictive median)",
    y = NULL,
    title = "Biomarker ranking vs Immune Atlas baseline",
    subtitle = "Genes ranked by quality_score; ideal pattern: SAVI > SAVI_treated > CTRL ≈ 0"
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

```{r}
#| label: plot-biomarker-ranking2
#| fig-width: 9
#| fig-height: 7
#| fig-align: center


ggplot2::ggplot(
  biomarker_long,
  ggplot2::aes(
    x = .data$median_log2fc,
    y = .data$gene,
    colour = .data$Category
  )
) +
  ggplot2::geom_vline(xintercept = 0, linetype = 2, colour = "grey60") +
  ggplot2::geom_line(
    ggplot2::aes(group = .data$gene),
    colour = "grey80",
    linewidth = 0.4
  ) +
  ggplot2::geom_point(size = 3, alpha = 0.85) +
  ggplot2::scale_colour_manual(values = category_cols) +
  ggplot2::labs(
    x = "Median log2(obs / healthy predictive median)",
    y = NULL,
    title = "Biomarker ranking vs Immune Atlas baseline",
    subtitle = "Genes ranked by quality_score; ideal pattern: SAVI > SAVI_treated > CTRL ≈ 0"
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

```{r}
#| label: plot-biomarker-class
#| fig-width: 9
#| fig-height: 6
#| fig-align: center

class_levels <- c(
  "validated",
  "validated_ctrl_drifts",
  "savi_up_no_treatment_effect",
  "ambiguous",
  "falsified_baseline_high",
  "falsified_no_savi_signal"
)

class_cols <- c(
  validated = "#1a9850",
  validated_ctrl_drifts = "#66bd63",
  savi_up_no_treatment_effect = "#fee08b",
  ambiguous = "#bdbdbd",
  falsified_baseline_high = "#762a83",
  falsified_no_savi_signal = "#d73027"
)

ggplot2::ggplot(
  biomarker_quality |>
    dplyr::mutate(
      gene = factor(.data$gene, levels = rev(gene_order)),
      biomarker_class = factor(
        .data$biomarker_class,
        levels = class_levels
      )
    ),
  ggplot2::aes(
    x = .data$quality_score,
    y = .data$gene,
    fill = .data$biomarker_class
  )
) +
  ggplot2::geom_col() +
  ggplot2::geom_vline(xintercept = 0, linetype = 2, colour = "grey50") +
  ggplot2::scale_fill_manual(values = class_cols, drop = FALSE) +
  ggplot2::labs(
    x = "Quality score = (SAVI − CTRL) + 0.5 × (SAVI − treated) − 0.5 × |CTRL drift|",
    y = NULL,
    title = "STING signature biomarker quality",
    subtitle = "Top: best match to expected pattern; bottom: atlas-falsified candidates",
    fill = NULL
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "right")
```

```{r}
#| label: biomarker-class-summary
biomarker_quality |>
  dplyr::count(.data$biomarker_class, name = "n_genes") |>
  dplyr::arrange(dplyr::desc(.data$n_genes))
```

```{r}
#| label: biomarker-falsified-genes
biomarker_quality |>
  dplyr::filter(.data$biomarker_class %in% c(
    "falsified_no_savi_signal",
    "falsified_baseline_high",
    "savi_up_no_treatment_effect"
  )) |>
  dplyr::select(
    "gene",
    "biomarker_class",
    "savi_lift",
    "ctrl_drift",
    "treated_lift",
    "pct_higher.SAVI",
    "pct_higher.CTRL",
    "pct_consistent.CTRL"
  ) |>
  dplyr::arrange(.data$biomarker_class, .data$gene)
```

**Interpretation:**

- **`validated`** — gene behaves as the paper claims and is **also** consistent with the atlas (high in SAVI, attenuated by treatment, in-cohort controls match atlas baseline). Strong biomarker.
- **`validated_ctrl_drifts`** — gene goes up in SAVI and drops with treatment, but the paper's controls are already above the atlas baseline. The within-cohort effect is real; absolute reference value should be interpreted with care.
- **`savi_up_no_treatment_effect`** — SAVI lift exists but treatment doesn't move it. The gene may be a state marker rather than an activity readout.
- **`ambiguous`** — borderline; needs more samples.
- **`falsified_baseline_high`** — gene reads as high in study controls *and* in the broader atlas, and SAVI doesn't add much on top. The paper's "up in SAVI" is likely a baseline artefact.
- **`falsified_no_savi_signal`** — no SAVI lift vs atlas at all. Strong falsification candidate.

This is the kind of cross-study check the original paper could not perform because of the small healthy-cohort size.

------------------------------------------------------------------------

# Case study: falsifying *IL1B*

**IL1B** is the clearest atlas-falsified gene in this cohort: study **controls sit above** the Immune Atlas healthy predictive distribution, while **SAVI samples do not sit higher still**. That pattern breaks the signature logic (“up in patients vs healthy”) when “healthy” is the atlas rather than seven matched donors.

To make the falsification auditable we show:

1. **Per-sample posterior predictive checks** — observed count vs healthy HPD (already computed).
2. **Predictive densities** — draw-level `posterior_predict` for IL1B vs a **validated** contrast gene (**TSC22D2**).
3. **Three reference frames** — atlas-corrected mean log2 FC vs **within-cohort** category means (raw and TMM-scaled counts averaged over donors in each `Category` only; no atlas comparison on the paper panels).

```{r}
#| label: il1b-falsify-helpers
falsify_contrast_genes <- c("IL1B", "TSC22D2")

scaled_counts_assay_name <- function(se) {
  nms <- SummarizedExperiment::assayNames(se)
  if ("counts_scaled" %in% nms) {
    return("counts_scaled")
  }
  stop(
    "Assay `counts_scaled` not found on `se_monocytic`. ",
    "Run the `scale-abundance` chunk first."
  )
}

il1b_ensembl <- sting_signature_21 |>
  dplyr::filter(.data$gene == "IL1B") |>
  dplyr::pull(.feature)

il1b_scaled_obs <- tibble::tibble(sample = colnames(se_monocytic)) |>
  dplyr::mutate(
    sample_idx = match(.data$sample, colnames(se_monocytic)),
    observed_scaled = as.numeric(
      SummarizedExperiment::assay(
        se_monocytic[il1b_ensembl, .data$sample_idx, drop = FALSE],
        scaled_counts_assay_name(se_monocytic)
      )
    )
  )

il1b_sample_tbl <- gene_deviation_eval |>
  dplyr::filter(.data$gene == "IL1B", !is.na(.data$Category)) |>
  dplyr::left_join(
    sample_annotations |>
      dplyr::select("sample", tidyselect::any_of("Treatment")),
    by = "sample"
  ) |>
  dplyr::left_join(il1b_scaled_obs, by = "sample") |>
  dplyr::mutate(
    Category = factor(.data$Category, levels = names(category_cols)),
    # Map count-scale predictive summaries to TMM-scaled scale (per-sample ratio).
    scale_ratio = .data$observed_scaled / pmax(.data$observed_count, 1e-9),
    healthy_median_scaled = .data$healthy_median * .data$scale_ratio,
    healthy_hpd_lo_scaled = .data$healthy_hpd_lo * .data$scale_ratio,
    healthy_hpd_hi_scaled = .data$healthy_hpd_hi * .data$scale_ratio,
    above_hpd = .data$observed_scaled > .data$healthy_hpd_hi_scaled,
    below_hpd = .data$observed_scaled < .data$healthy_hpd_lo_scaled,
    in_hpd = !.data$above_hpd & !.data$below_hpd
  ) |>
  dplyr::arrange(.data$Category, .data$Sample)

il1b_sample_tbl |>
  dplyr::select(
    "Sample",
    "Category",
    "observed_scaled",
    "healthy_median_scaled",
    "healthy_hpd_lo_scaled",
    "healthy_hpd_hi_scaled",
    "obs_percentile",
    "direction",
    "log2fc_vs_median"
  )
```

```{r}
#| label: plot-il1b-obs-vs-hpd
#| fig-width: 11
#| fig-height: 7
#| fig-align: center

il1b_plot_df <- il1b_sample_tbl |>
  dplyr::mutate(
    sample_label = factor(
      Sample,
      levels = il1b_sample_tbl |>
        dplyr::arrange(Category, Sample) |>
        dplyr::pull(Sample) |>
        unique()
    )
  )

log1p_trans <- scales::trans_new(
  name = "log1p",
  transform = function(x) log1p(pmax(x, 0)),
  inverse = function(x) expm1(x),
  domain = c(0, Inf)
)

ggplot2::ggplot(il1b_plot_df) +
  ggplot2::geom_linerange(
    ggplot2::aes(
      x = sample_label,
      ymin = healthy_hpd_lo,
      ymax = healthy_hpd_hi
    ),
    colour = "grey45",
    linewidth = 0.7
  ) +
  ggplot2::geom_point(
    ggplot2::aes(x = sample_label, y = healthy_median),
    shape = 4,
    size = 2.2,
    colour = "grey55",
    stroke = 0.9
  ) +
  ggplot2::geom_point(
    ggplot2::aes(x = sample_label, y = observed_count, colour = Category),
    size = 3.5
  ) +
  ggplot2::scale_y_continuous(trans = log1p_trans) +
  ggplot2::scale_colour_manual(values = category_cols) +
  ggplot2::labs(
    x = NULL,
    y = expression(log[1 + p]("raw pseudobulk count")),
    title = expression(italic("IL1B") * ": raw counts vs atlas healthy predictive interval"),
    subtitle = "Each point = one donor; x-axis sorted by Category (CTRL, SAVI_treated, SAVI). Grey bar = 95% atlas predictive HPD; + = predictive median."
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```

```{r}
#| label: plot-il1b-percentile
#| fig-width: 14
#| fig-height: 6
#| fig-align: center

il1b_counts_long <- il1b_sample_tbl |>
  dplyr::mutate(
    sample_label = factor(
      Sample,
      levels = il1b_sample_tbl |>
        dplyr::arrange(Category, Sample) |>
        dplyr::pull(Sample) |>
        unique()
    )
  ) |>
  tidyr::pivot_longer(
    cols = c("observed_count", "observed_scaled"),
    names_to = "measure",
    values_to = "abundance"
  ) |>
  dplyr::mutate(
    measure = dplyr::recode(
      .data$measure,
      observed_count = "Raw pseudobulk counts",
      observed_scaled = "TMM-scaled counts"
    ),
    measure = factor(
      .data$measure,
      levels = c("Raw pseudobulk counts", "TMM-scaled counts")
    )
  )

log1p_trans <- scales::trans_new(
  name = "log1p",
  transform = function(x) log1p(pmax(x, 0)),
  inverse = function(x) expm1(x),
  domain = c(0, Inf)
)

ggplot2::ggplot(
  il1b_counts_long,
  ggplot2::aes(
    x = sample_label,
    y = abundance,
    colour = Category
  )
) +
  ggplot2::geom_point(size = 3.5) +
  ggplot2::facet_wrap(~measure, scales = "free_y", ncol = 2) +
  ggplot2::scale_y_continuous(trans = log1p_trans) +
  ggplot2::scale_colour_manual(values = category_cols) +
  ggplot2::labs(
    x = NULL,
    y = expression(log[1 + p]("count")),
    title = expression(italic("IL1B") * ": per-donor abundance (raw vs TMM-scaled)"),
    subtitle = paste0(
      "Each point = one donor pseudobulk in the GSE226598 cohort (not a category mean). ",
      "Atlas percentile is identical on both scales (linear TMM rescaling); ",
      "see obs_percentile in the table above."
    )
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, size = 8),
    legend.position = "top",
    strip.text = ggplot2::element_text(size = 10)
  )
```

```{r}
#| label: plot-il1b-two-nulls
#| fig-width: 14
#| fig-height: 4.5
#| fig-align: center

# Three reference frames for IL1B:
#   Atlas  — mean log2(obs / healthy predictive median) across donors in each Category.
#   Paper  — mean abundance within the GSE226598 cohort only (no atlas); bars = cohort mean
#            per Category (CTRL n≈7, SAVI n≈5, SAVI_treated n varies), not individual donors.
paper_null <- il1b_sample_tbl |>
  dplyr::group_by(Category) |>
  dplyr::summarise(
    mean_obs = mean(observed_count),
    mean_scaled = mean(observed_scaled),
    mean_log1p_scaled = mean(log1p(pmax(observed_scaled, 0))),
    .groups = "drop"
  )

atlas_null <- il1b_sample_tbl |>
  dplyr::group_by(Category) |>
  dplyr::summarise(
    mean_log2fc = mean(log2fc_vs_median),
    .groups = "drop"
  )

two_nulls_long <- dplyr::bind_rows(
  atlas_null |>
    dplyr::mutate(
      metric = "Atlas: mean log2 FC vs healthy baseline",
      value = .data$mean_log2fc,
      y_label = "Mean log2(obs / healthy median)"
    ) |>
    dplyr::select("Category", "metric", "value", "y_label"),
  paper_null |>
    dplyr::mutate(
      metric = "Within-cohort mean (raw pseudobulk counts)",
      value = .data$mean_obs,
      y_label = "Mean raw count"
    ) |>
    dplyr::select("Category", "metric", "value", "y_label"),
  paper_null |>
    dplyr::mutate(
      metric = "Within-cohort mean (log1p TMM-scaled counts)",
      value = .data$mean_log1p_scaled,
      y_label = "log1p(mean scaled count)"
    ) |>
    dplyr::select("Category", "metric", "value", "y_label")
) |>
  dplyr::mutate(
    Category = factor(Category, levels = names(category_cols)),
    metric = factor(
      metric,
      levels = c(
        "Atlas: mean log2 FC vs healthy baseline",
        "Within-cohort mean (raw pseudobulk counts)",
        "Within-cohort mean (log1p TMM-scaled counts)"
      )
    )
  )

ggplot2::ggplot(
  two_nulls_long,
  ggplot2::aes(x = Category, y = value, fill = Category)
) +
  ggplot2::geom_col(width = 0.6, colour = "white") +
  ggplot2::geom_hline(yintercept = 0, linetype = 2, colour = "grey40") +
  ggplot2::facet_wrap(~metric, scales = "free_y", ncol = 3) +
  ggplot2::scale_fill_manual(values = category_cols) +
  ggplot2::labs(
    x = NULL,
    y = NULL,
    title = expression(italic("IL1B") * ": atlas baseline vs within-cohort category means"),
    subtitle = paste0(
      "Bars = mean across donors in each Category (same GSE226598 cohort). ",
      "Left compares to Immune Atlas healthy predictive median; centre/right are paper-style ",
      "within-cohort contrasts only (raw or TMM-scaled pseudobulk counts)."
    )
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    legend.position = "none",
    strip.text = ggplot2::element_text(size = 9)
  )
```

```{r}
#| label: il1b-predictive-draws
#| cache: true

# Subsampled posterior_predict draws for density overlays (IL1B vs TSC22D2).
# Model scale = raw pseudobulk counts (same as atlas posterior_predict).
predictive_draws_raw_for_gene <- function(
  gene_sym,
  se,
  comparison_tbl,
  sample_ann,
  ndraws = 400L
) {
  chunk_row <- sting_fit_rows |>
    dplyr::filter(.data$gene == .env$gene_sym) |>
    dplyr::slice(1)

  if (!nrow(chunk_row)) {
    stop("No fit found for gene: ", gene_sym)
  }

  fit <- poco::reconstruct_brmsfit(chunk_row$brms_fit_compressed[[1]])
  feature_id <- chunk_row$.feature

  purrr::pmap_dfr(
    comparison_tbl |>
      dplyr::filter(.data$sample_role == "savi_pseudobulk") |>
      dplyr::select("sample", "sample_idx"),
    function(sample, sample_idx) {
      newdata <- newdata_for_sample(se, sample_idx)
      nd_grid <- build_newdata_grid(
        fit,
        newdata,
        new_level_tag = paste0("__savi__", as.character(sample))
      )

      y_rep <- brms::posterior_predict(
        fit,
        newdata = nd_grid,
        summary = FALSE,
        re_formula = NULL,
        allow_new_levels = TRUE,
        sample_new_levels = "uncertainty",
        ndraws = ndraws
      ) |>
        as.numeric()

      obs_raw <- SummarizedExperiment::assay(
        se[feature_id, sample_idx, drop = FALSE],
        "counts"
      ) |>
        as.numeric()

      tibble::tibble(
        gene = gene_sym,
        sample = as.character(sample),
        draw = y_rep,
        observed_count = obs_raw
      )
    }
  ) |>
    dplyr::left_join(
      sample_ann |>
        dplyr::select("sample", "Sample", "Category"),
      by = "sample"
    )
}

contrast_draws <- purrr::map_dfr(
  falsify_contrast_genes,
  predictive_draws_raw_for_gene,
  se = se_monocytic,
  comparison_tbl = comparison_samples_tbl,
  sample_ann = sample_annotations
) |>
  dplyr::mutate(
    Category = factor(.data$Category, levels = names(category_cols)),
    gene = factor(.data$gene, levels = falsify_contrast_genes),
    draw_log1p = log1p(pmax(.data$draw, 0)),
    observed_log1p = log1p(pmax(.data$observed_count, 0))
  )

contrast_draws |>
  dplyr::count(.data$gene, .data$Category, name = "n_draws")
```

```{r}
#| label: plot-il1b-density-contrast
#| fig-width: 11
#| fig-height: 5.5
#| fig-align: center

ggplot2::ggplot(
  contrast_draws,
  ggplot2::aes(x = draw_log1p, colour = Category, fill = Category)
) +
  ggplot2::geom_density(alpha = 0.25, linewidth = 0.9) +
  ggplot2::geom_vline(
    data = contrast_draws |>
      dplyr::distinct(.data$gene, .data$sample, .data$Category, .data$observed_log1p),
    ggplot2::aes(
      xintercept = observed_log1p,
      colour = Category
    ),
    linetype = "solid",
    linewidth = 0.7
  ) +
  ggplot2::facet_wrap(~gene, scales = "free_x") +
  ggplot2::scale_colour_manual(values = category_cols, name = "Category") +
  ggplot2::scale_fill_manual(values = category_cols, guide = "none") +
  ggplot2::labs(
    x = expression(log[1 + p]("raw pseudobulk count")),
    y = "Density",
    title = "Predictive null vs observed (log1p raw counts)",
    subtitle = paste(
      "Curves = atlas posterior_predict draws pooled by Category (model scale).",
      "Vertical lines = each donor's observed raw count, coloured by Category."
    )
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r}
#| label: plot-il1b-density-facet-sample
#| fig-width: 12
#| fig-height: 8
#| fig-align: center

il1b_facet_lookup <- il1b_sample_tbl |>
  dplyr::select(dplyr::any_of(c(
    "sample",
    "Sample",
    "Category",
    "direction",
    "p_above_healthy"
  ))) |>
  dplyr::mutate(
    direction_short = dplyr::case_when(
      .data$direction == "higher_than_healthy" ~ "higher",
      .data$direction == "lower_than_healthy" ~ "lower",
      TRUE ~ "consistent"
    ),
    sample_label = paste0(
      .data$Sample,
      " (",
      .data$Category,
      " · ",
      .data$direction_short,
      ", p=",
      sprintf("%.2f", .data$p_above_healthy),
      ")"
    )
  )

il1b_label_levels <- il1b_facet_lookup |>
  dplyr::arrange(.data$Category, .data$Sample)

il1b_facet_lookup <- il1b_facet_lookup |>
  dplyr::mutate(
    sample_label = factor(
      .data$sample_label,
      levels = il1b_label_levels$sample_label
    )
  )

il1b_draws <- contrast_draws |>
  dplyr::filter(.data$gene == "IL1B") |>
  dplyr::left_join(
    il1b_facet_lookup |>
      dplyr::select(.data$sample, .data$sample_label),
    by = "sample"
  )

ggplot2::ggplot(
  il1b_draws,
  ggplot2::aes(x = draw_log1p, colour = Category)
) +
  ggplot2::geom_density(fill = "grey85", colour = "grey40", linewidth = 0.6, alpha = 0.9) +
  ggplot2::geom_vline(
    ggplot2::aes(xintercept = observed_log1p, colour = Category),
    linewidth = 1
  ) +
  ggplot2::facet_wrap(~sample_label, scales = "free_x", ncol = 4) +
  ggplot2::scale_colour_manual(values = category_cols, name = "Category") +
  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(linewidth = 1.2))) +
  ggplot2::labs(
    x = expression(log[1 + p]("raw pseudobulk count")),
    y = "Density",
    title = expression(italic("IL1B") * ": posterior predictive check per donor"),
    subtitle = paste(
      "Facet = donor (Category · atlas tail, p = P(draw >= obs) on raw counts).",
      "Grey density = atlas healthy null; coloured vertical line = observed raw count."
    )
  ) +
  ggplot2::theme_minimal(base_size = 10) +
  ggplot2::theme(
    legend.position = "top",
    strip.text = ggplot2::element_text(size = 8)
  )
```

```{r}
#| label: il1b-falsify-summary
il1b_ctrl_vs_savi <- il1b_sample_tbl |>
  dplyr::group_by(.data$Category) |>
  dplyr::summarise(
    n = dplyr::n(),
    mean_obs = mean(.data$observed_count),
    mean_log2fc = mean(.data$log2fc_vs_median),
    pct_above_hpd = 100 * mean(.data$above_hpd),
    pct_higher_direction = 100 * mean(.data$direction == "higher_than_healthy"),
    mean_percentile = 100 * mean(.data$obs_percentile),
    .groups = "drop"
  )

ctrl_mean_log2fc <- il1b_ctrl_vs_savi |>
  dplyr::filter(.data$Category == "CTRL") |>
  dplyr::pull(.data$mean_log2fc)

ctrl_pct_above_hpd <- il1b_ctrl_vs_savi |>
  dplyr::filter(.data$Category == "CTRL") |>
  dplyr::pull(.data$pct_above_hpd)

savi_pct_above_hpd <- il1b_ctrl_vs_savi |>
  dplyr::filter(.data$Category == "SAVI") |>
  dplyr::pull(.data$pct_above_hpd)

paper_delta <- il1b_ctrl_vs_savi |>
  dplyr::filter(.data$Category %in% c("SAVI", "CTRL")) |>
  tidyr::pivot_wider(names_from = "Category", values_from = "mean_obs") |>
  dplyr::mutate(delta = .data$SAVI - .data$CTRL) |>
  dplyr::pull(.data$delta)

atlas_delta <- biomarker_quality |>
  dplyr::filter(.data$gene == "IL1B") |>
  dplyr::pull(.data$savi_vs_ctrl)

tibble::tibble(
  gene = "IL1B",
  paper_mean_savi_minus_ctrl_raw_counts = paper_delta,
  atlas_mean_log2fc_savi_minus_ctrl = atlas_delta,
  ctrl_pct_above_healthy_hpd = ctrl_pct_above_hpd,
  savi_pct_above_healthy_hpd = savi_pct_above_hpd,
  falsification_verdict = dplyr::if_else(
    is.finite(atlas_delta) &&
      atlas_delta < 0 &&
      is.finite(ctrl_mean_log2fc) &&
      ctrl_mean_log2fc > 0,
    "Atlas falsified: controls high, SAVI not elevated after correction",
    "Review manually"
  )
)

il1b_ctrl_vs_savi
```

**How to read this as falsification evidence**

- **Per-donor plots** (HPD, raw/scaled abundance, density facets): each point/line is one pseudobulk library; HPD and density overlays use **raw counts** on the model scale (`posterior_predict` draws and observed `counts` assay). The two-panel abundance plot shows the same donors on **raw** and **TMM-scaled** count scales (log1p y-axis); atlas percentile is unchanged under linear scaling.
- **Bar chart (`plot-il1b-two-nulls`)**: height = **mean within the GSE226598 cohort** for donors in that `Category` only (not vs atlas on the two right panels). Left panel = mean log2(obs / atlas predictive median); centre = mean raw pseudobulk count; right = mean log1p(TMM-scaled count).
- If **atlas falsification holds**: per-donor plots show **CTRL** above the atlas HPD and high `obs_percentile` in the table above, while **SAVI** is not systematically above CTRL on the **left bar panel** (mean log2 FC).
- The **centre bar panel** can still show **SAVI > CTRL** on raw within-cohort means — the contrast the original paper could see with seven controls.
- The **summary table** quantifies: `paper_mean_savi_minus_ctrl_raw_counts` can stay **positive** while `atlas_mean_log2fc_savi_minus_ctrl` is **negative**.

------------------------------------------------------------------------

# Case study: falsifying *ADRB2* (one-donor SAVI signal)

**ADRB2** is labelled **`falsified_no_savi_signal`** in the biomarker screen because only **20%** of SAVI donors pass the strict atlas tail call (`higher_than_healthy`), yet the **cohort mean** `log2fc_vs_median` in SAVI looks high. That pattern usually means **one outlier donor** pulls the category mean while the other SAVI pseudobulks sit near the healthy predictive null.

We audit that claim with per-donor tables, leave-one-out SAVI means, and density overlays on **raw pseudobulk counts** (same scale as `posterior_predict`; contrast gene **TSC22D2** = validated atlas behaviour).

```{r}
#| label: adrb2-falsify-helpers

adrb2_ensembl <- sting_signature_21 |>
  dplyr::filter(.data$gene == "ADRB2") |>
  dplyr::pull(.feature)

adrb2_scaled_obs <- tibble::tibble(sample = colnames(se_monocytic)) |>
  dplyr::mutate(
    sample_idx = match(.data$sample, colnames(se_monocytic)),
    observed_scaled = as.numeric(
      SummarizedExperiment::assay(
        se_monocytic[adrb2_ensembl, .data$sample_idx, drop = FALSE],
        scaled_counts_assay_name(se_monocytic)
      )
    )
  )

adrb2_sample_tbl <- gene_deviation_eval |>
  dplyr::filter(.data$gene == "ADRB2", !is.na(.data$Category)) |>
  dplyr::left_join(
    sample_annotations |>
      dplyr::select("sample", tidyselect::any_of("Treatment")),
    by = "sample"
  ) |>
  dplyr::left_join(adrb2_scaled_obs, by = "sample") |>
  dplyr::mutate(
    Category = factor(.data$Category, levels = names(category_cols)),
    scale_ratio = .data$observed_scaled / pmax(.data$observed_count, 1e-9),
    healthy_median_scaled = .data$healthy_median * .data$scale_ratio,
    healthy_hpd_lo_scaled = .data$healthy_hpd_lo * .data$scale_ratio,
    healthy_hpd_hi_scaled = .data$healthy_hpd_hi * .data$scale_ratio,
    above_hpd = .data$observed_count > .data$healthy_hpd_hi,
    below_hpd = .data$observed_count < .data$healthy_hpd_lo,
    in_hpd = !.data$above_hpd & !.data$below_hpd
  ) |>
  dplyr::arrange(.data$Category, .data$Sample)

adrb2_savi_tbl <- adrb2_sample_tbl |>
  dplyr::filter(.data$Category == "SAVI")

adrb2_savi_outlier_pick <- adrb2_savi_tbl |>
  dplyr::filter(.data$direction == "higher_than_healthy") |>
  dplyr::slice(1)

if (nrow(adrb2_savi_outlier_pick) == 0L) {
  adrb2_savi_outlier_pick <- adrb2_savi_tbl |>
    dplyr::slice_max(.data$log2fc_vs_median, n = 1, with_ties = FALSE)
}

adrb2_label_levels <- adrb2_sample_tbl |>
  dplyr::arrange(.data$Category, .data$Sample)

adrb2_sample_tbl <- adrb2_sample_tbl |>
  dplyr::mutate(
    savi_outlier = .data$sample %in% adrb2_savi_outlier_pick$sample,
    direction_short = dplyr::case_when(
      .data$direction == "higher_than_healthy" ~ "higher",
      .data$direction == "lower_than_healthy" ~ "lower",
      TRUE ~ "consistent"
    ),
    sample_facet_label = paste0(
      .data$Sample,
      " (",
      .data$Category,
      " · ",
      .data$direction_short,
      ", p=",
      sprintf("%.2f", .data$p_above_healthy),
      ")"
    ),
    sample_label = factor(
      .data$sample_facet_label,
      levels = adrb2_label_levels |>
        dplyr::mutate(
          lab = paste0(
            .data$Sample,
            " (",
            .data$Category,
            " · ",
            dplyr::case_when(
              .data$direction == "higher_than_healthy" ~ "higher",
              .data$direction == "lower_than_healthy" ~ "lower",
              TRUE ~ "consistent"
            ),
            ", p=",
            sprintf("%.2f", .data$p_above_healthy),
            ")"
          )
        ) |>
        dplyr::pull(.data$lab)
    )
  )

adrb2_savi_outlier <- adrb2_sample_tbl |>
  dplyr::filter(.data$savi_outlier)

adrb2_loo_summary <- tibble::tibble(
  contrast = c(
    "All SAVI donors",
    "Drop atlas-tail SAVI donor",
    "Other SAVI donors (median)"
  ),
  n_savi = c(
    nrow(adrb2_savi_tbl),
    nrow(adrb2_savi_tbl) - 1L,
    nrow(adrb2_savi_tbl) - 1L
  ),
  mean_log2fc = c(
    mean(adrb2_savi_tbl$log2fc_vs_median),
    mean(
      adrb2_savi_tbl$log2fc_vs_median[
        adrb2_savi_tbl$sample != adrb2_savi_outlier$sample
      ]
    ),
    stats::median(
      adrb2_savi_tbl$log2fc_vs_median[
        adrb2_savi_tbl$sample != adrb2_savi_outlier$sample
      ]
    )
  ),
  pct_higher = c(
    100 * mean(adrb2_savi_tbl$direction == "higher_than_healthy"),
    100 * mean(
      adrb2_savi_tbl$direction[
        adrb2_savi_tbl$sample != adrb2_savi_outlier$sample
      ] == "higher_than_healthy"
    ),
    NA_real_
  )
)

adrb2_per_donor <- adrb2_sample_tbl |>
  dplyr::select(
    Sample,
    Category,
    direction,
    direction_short,
    p_above_healthy,
    log2fc_vs_median,
    obs_percentile,
    observed_count,
    observed_scaled,
    savi_outlier,
    sample_facet_label
  )

adrb2_per_donor

adrb2_loo_summary

tibble::tibble(
  outlier_sample = adrb2_savi_outlier$Sample,
  outlier_direction = adrb2_savi_outlier$direction,
  outlier_log2fc = adrb2_savi_outlier$log2fc_vs_median,
  n_savi_higher_than_healthy = sum(adrb2_savi_tbl$direction == "higher_than_healthy"),
  n_savi_consistent = sum(adrb2_savi_tbl$direction == "consistent_with_healthy"),
  biomarker_class = biomarker_quality |>
    dplyr::filter(.data$gene == "ADRB2") |>
    dplyr::pull(.data$biomarker_class)
)
```

```{r}
#| label: plot-adrb2-log2fc-by-sample
#| fig-width: 11
#| fig-height: 7
#| fig-align: center

direction_cols <- c(
  higher_than_healthy = "#E41A1C",
  lower_than_healthy = "#377EB8",
  consistent_with_healthy = "grey65"
)

ggplot2::ggplot(
  adrb2_sample_tbl,
  ggplot2::aes(
    x = sample_label,
    y = log2fc_vs_median,
    colour = direction
  )
) +
  ggplot2::geom_hline(yintercept = 0, linetype = 2, colour = "grey55") +
  ggplot2::geom_point(
    ggplot2::aes(size = savi_outlier),
    alpha = 0.9
  ) +
  ggplot2::geom_point(
    data = adrb2_savi_outlier,
    ggplot2::aes(x = sample_label, y = log2fc_vs_median),
    shape = 21,
    fill = NA,
    colour = "black",
    stroke = 1.2,
    size = 5,
    inherit.aes = FALSE
  ) +
  ggplot2::scale_colour_manual(values = direction_cols, name = "Atlas tail") +
  ggplot2::scale_size_manual(
    values = c(`TRUE` = 5, `FALSE` = 3),
    guide = "none"
  ) +
  ggplot2::labs(
    x = NULL,
    y = expression(log[2]("obs / healthy predictive median")),
    title = expression(italic("ADRB2") * ": per-donor atlas deviation"),
    subtitle = paste(
      "Ring = sole SAVI donor with atlas tail higher_than_healthy",
      paste0("(", adrb2_savi_outlier$Sample, ", p=", sprintf("%.3f", adrb2_savi_outlier$p_above_healthy), ")."),
      "x labels: direction and P(draw >= obs); visual offset ≠ tail call."
    )
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, size = 8),
    legend.position = "top"
  )
```

```{r}
#| label: plot-adrb2-leave-one-out
#| fig-width: 8
#| fig-height: 4.5
#| fig-align: center

adrb2_loo_plot <- adrb2_loo_summary |>
  dplyr::mutate(
    contrast = factor(
      .data$contrast,
      levels = .data$contrast
    )
  )

ggplot2::ggplot(
  adrb2_loo_plot,
  ggplot2::aes(x = contrast, y = mean_log2fc, fill = contrast)
) +
  ggplot2::geom_col(width = 0.65, colour = "white") +
  ggplot2::geom_hline(yintercept = 0, linetype = 2, colour = "grey45") +
  ggplot2::scale_fill_manual(values = c("#E41A1C", "#FC8D59", "#FDBB84"), guide = "none") +
  ggplot2::labs(
    x = NULL,
    y = expression("Mean / median " * log[2] * " FC (SAVI only)"),
    title = expression(italic("ADRB2") * ": SAVI cohort mean driven by one donor"),
    subtitle = paste0(
      "Dropping ",
      adrb2_savi_outlier$Sample,
      " collapses the SAVI mean toward 0; remaining donors do not support a cohort-wide atlas tail."
    )
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 25, hjust = 1))
```

```{r}
#| label: plot-adrb2-obs-vs-hpd
#| fig-width: 14
#| fig-height: 6
#| fig-align: center

log1p_trans <- scales::trans_new(
  name = "log1p",
  transform = function(x) log1p(pmax(x, 0)),
  inverse = function(x) expm1(x),
  domain = c(0, Inf)
)

adrb2_hpd_plot <- adrb2_sample_tbl |>
  dplyr::transmute(
    sample_label,
    Category,
    savi_outlier,
    measure = "Raw pseudobulk counts (atlas null)",
    hpd_lo = .data$healthy_hpd_lo,
    hpd_hi = .data$healthy_hpd_hi,
    pred_median = .data$healthy_median,
    abundance = .data$observed_count
  ) |>
  dplyr::bind_rows(
    adrb2_sample_tbl |>
      dplyr::transmute(
        sample_label,
        Category,
        savi_outlier,
        measure = "TMM-scaled counts (linear rescale of null)",
        hpd_lo = .data$healthy_hpd_lo_scaled,
        hpd_hi = .data$healthy_hpd_hi_scaled,
        pred_median = .data$healthy_median_scaled,
        abundance = .data$observed_scaled
      )
  ) |>
  dplyr::mutate(
    measure = factor(
      .data$measure,
      levels = c(
        "Raw pseudobulk counts (atlas null)",
        "TMM-scaled counts (linear rescale of null)"
      )
    )
  )

ggplot2::ggplot(adrb2_hpd_plot) +
  ggplot2::geom_linerange(
    ggplot2::aes(
      x = sample_label,
      ymin = hpd_lo,
      ymax = hpd_hi
    ),
    colour = "grey45",
    linewidth = 0.7
  ) +
  ggplot2::geom_point(
    ggplot2::aes(x = sample_label, y = pred_median),
    shape = 4,
    size = 2.2,
    colour = "grey55",
    stroke = 0.9
  ) +
  ggplot2::geom_point(
    ggplot2::aes(
      x = sample_label,
      y = abundance,
      colour = Category,
      size = savi_outlier
    ),
    alpha = 0.9
  ) +
  ggplot2::facet_wrap(~measure, scales = "free_y", ncol = 2) +
  ggplot2::scale_y_continuous(trans = log1p_trans) +
  ggplot2::scale_colour_manual(values = category_cols) +
  ggplot2::scale_size_manual(values = c(`TRUE` = 4.5, `FALSE` = 3), guide = "none") +
  ggplot2::labs(
    x = NULL,
    y = expression(log[1 + p]("count")),
    title = expression(italic("ADRB2") * ": observed vs atlas healthy predictive interval"),
    subtitle = paste(
      "Left = raw counts (same scale as tail test and density plots).",
      "Right = TMM-scaled abundance with HPD rescaled by sample library size;",
      "atlas percentile unchanged. Larger point = SAVI atlas-tail outlier."
    )
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, size = 7),
    legend.position = "top",
    strip.text = ggplot2::element_text(size = 9)
  )
```

```{r}
#| label: plot-adrb2-savi-only
#| fig-width: 7
#| fig-height: 4.5
#| fig-align: center

ggplot2::ggplot(
  adrb2_savi_tbl |>
    dplyr::mutate(
      highlight = .data$sample %in% adrb2_savi_outlier$sample,
      Sample = factor(.data$Sample, levels = adrb2_savi_tbl$Sample)
    ),
  ggplot2::aes(x = Sample, y = log2fc_vs_median, colour = direction)
) +
  ggplot2::geom_hline(yintercept = 0, linetype = 2, colour = "grey55") +
  ggplot2::geom_segment(
    ggplot2::aes(x = Sample, xend = Sample, y = 0, yend = log2fc_vs_median),
    colour = "grey75",
    linewidth = 0.6
  ) +
  ggplot2::geom_point(
    ggplot2::aes(size = highlight),
    alpha = 0.95
  ) +
  ggplot2::scale_colour_manual(values = direction_cols, name = "Atlas tail") +
  ggplot2::scale_size_manual(values = c(`TRUE` = 5, `FALSE` = 3), guide = "none") +
  ggplot2::labs(
    x = NULL,
    y = expression(log[2]("obs / healthy predictive median")),
    title = expression(italic("ADRB2") * ": SAVI donors only"),
    subtitle = paste0(
      nrow(adrb2_savi_tbl) - sum(adrb2_savi_tbl$direction == "higher_than_healthy"),
      " of ",
      nrow(adrb2_savi_tbl),
      " SAVI donors are not higher_than_healthy vs atlas."
    )
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r}
#| label: adrb2-predictive-draws
#| cache: true

# Uses predictive_draws_raw_for_gene() from il1b-predictive-draws (raw model scale).
adrb2_contrast_genes <- c("ADRB2", "TSC22D2")

adrb2_draws <- purrr::map_dfr(
  adrb2_contrast_genes,
  predictive_draws_raw_for_gene,
  se = se_monocytic,
  comparison_tbl = comparison_samples_tbl,
  sample_ann = sample_annotations
) |>
  dplyr::mutate(
    Category = factor(.data$Category, levels = names(category_cols)),
    gene = factor(.data$gene, levels = adrb2_contrast_genes),
    draw_log1p = log1p(pmax(.data$draw, 0)),
    observed_log1p = log1p(pmax(.data$observed_count, 0)),
    savi_outlier = .data$sample %in% adrb2_savi_outlier$sample
  )

adrb2_draws |>
  dplyr::count(.data$gene, .data$Category, name = "n_draws")
```

```{r}
#| label: plot-adrb2-density-contrast
#| fig-width: 11
#| fig-height: 5.5
#| fig-align: center

ggplot2::ggplot(
  adrb2_draws,
  ggplot2::aes(x = draw_log1p, colour = Category, fill = Category)
) +
  ggplot2::geom_density(alpha = 0.25, linewidth = 0.9) +
  ggplot2::geom_vline(
    data = adrb2_draws |>
      dplyr::distinct(.data$gene, .data$sample, .data$Category, .data$observed_log1p, .data$savi_outlier),
    ggplot2::aes(
      xintercept = observed_log1p,
      colour = Category,
      linewidth = savi_outlier
    ),
    linetype = "solid"
  ) +
  ggplot2::scale_linewidth_manual(values = c(`TRUE` = 1.2, `FALSE` = 0.65), guide = "none") +
  ggplot2::facet_wrap(~gene, scales = "free_x") +
  ggplot2::scale_colour_manual(values = category_cols, name = "Category") +
  ggplot2::scale_fill_manual(values = category_cols, guide = "none") +
  ggplot2::labs(
    x = expression(log[1 + p]("raw pseudobulk count")),
    y = "Density",
    title = expression(italic("ADRB2") * " vs validated " * italic("TSC22D2")),
    subtitle = "Grey = atlas posterior_predict (raw counts); vertical lines = observed raw count (log1p)."
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

```{r}
#| label: plot-adrb2-density-facet-sample
#| fig-width: 12
#| fig-height: 8
#| fig-align: center

adrb2_facet_lookup <- adrb2_sample_tbl |>
  dplyr::select(dplyr::any_of(c(
    "sample",
    "sample_label",
    "Sample",
    "Category",
    "direction",
    "p_above_healthy"
  )))

if (!"sample_label" %in% names(adrb2_facet_lookup) ||
    !any(grepl(", p=", adrb2_facet_lookup$sample_label, fixed = TRUE), na.rm = TRUE)) {
  adrb2_facet_lookup <- adrb2_facet_lookup |>
    dplyr::mutate(
      direction_short = dplyr::case_when(
        .data$direction == "higher_than_healthy" ~ "higher",
        .data$direction == "lower_than_healthy" ~ "lower",
        TRUE ~ "consistent"
      ),
      sample_label = paste0(
        .data$Sample,
        " (",
        .data$Category,
        " · ",
        .data$direction_short,
        ", p=",
        sprintf("%.2f", .data$p_above_healthy),
        ")"
      )
    )
}

if ("sample_label" %in% names(adrb2_sample_tbl) && is.factor(adrb2_sample_tbl$sample_label)) {
  adrb2_facet_lookup$sample_label <- factor(
    adrb2_facet_lookup$sample_label,
    levels = levels(adrb2_sample_tbl$sample_label)
  )
}

adrb2_draws_gene <- adrb2_draws |>
  dplyr::filter(.data$gene == "ADRB2") |>
  dplyr::left_join(
    adrb2_facet_lookup |>
      dplyr::select(.data$sample, .data$sample_label),
    by = "sample"
  )

ggplot2::ggplot(
  adrb2_draws_gene,
  ggplot2::aes(x = draw_log1p, colour = Category)
) +
  ggplot2::geom_density(fill = "grey85", colour = "grey40", linewidth = 0.6, alpha = 0.9) +
  ggplot2::geom_vline(
    ggplot2::aes(
      xintercept = observed_log1p,
      colour = Category,
      linewidth = savi_outlier
    ),
    linetype = "solid"
  ) +
  ggplot2::scale_linewidth_manual(values = c(`TRUE` = 1.3, `FALSE` = 0.8), guide = "none") +
  ggplot2::facet_wrap(~sample_label, scales = "free_x", ncol = 4) +
  ggplot2::scale_colour_manual(values = category_cols, name = "Category") +
  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(linewidth = 1.2))) +
  ggplot2::labs(
    x = expression(log[1 + p]("raw pseudobulk count")),
    y = "Density",
    title = expression(italic("ADRB2") * ": posterior predictive check per donor"),
    subtitle = paste(
      "Facet = donor (Category · atlas tail, p = P(draw >= obs) on raw counts).",
      "Thicker line =",
      adrb2_savi_outlier$Sample,
      "(only SAVI donor with p < 0.025)."
    )
  ) +
  ggplot2::theme_minimal(base_size = 10) +
  ggplot2::theme(
    legend.position = "top",
    strip.text = ggplot2::element_text(size = 6.5)
  )
```

```{r}
#| label: adrb2-falsify-summary

adrb2_biomarker_row <- biomarker_quality |>
  dplyr::filter(.data$gene == "ADRB2")

tibble::tibble(
  gene = "ADRB2",
  biomarker_class = adrb2_biomarker_row$biomarker_class,
  pct_higher_SAVI = adrb2_biomarker_row$pct_higher.SAVI,
  mean_log2fc_SAVI = adrb2_biomarker_row$savi_lift,
  savi_vs_ctrl_atlas = adrb2_biomarker_row$savi_vs_ctrl,
  outlier_sample = adrb2_savi_outlier$Sample,
  outlier_log2fc = adrb2_savi_outlier$log2fc_vs_median,
  n_savi_higher = sum(adrb2_savi_tbl$direction == "higher_than_healthy"),
  n_savi_consistent = sum(adrb2_savi_tbl$direction == "consistent_with_healthy"),
  mean_log2fc_SAVI_drop_outlier = adrb2_loo_summary$mean_log2fc[2],
  falsification_verdict = paste(
    "Cohort SAVI mean is not a robust signature signal:",
    "only",
    sum(adrb2_savi_tbl$direction == "higher_than_healthy"),
    "of",
    nrow(adrb2_savi_tbl),
    "SAVI donors exceed the atlas null;",
    "effect driven by",
    adrb2_savi_outlier$Sample
  )
)
```

**How to read**

- **`plot-adrb2-log2fc-by-sample`** and **`plot-adrb2-density-facet-sample`**: same **raw-count** scale as the atlas model; facet labels show **atlas tail** and **`p = P(draw >= obs)`**. Only **P6_STING_ht** has **p < 0.025** (thicker line).
- **`plot-adrb2-leave-one-out`**: removing that donor drops the SAVI mean `log2fc` toward zero (median of the remaining four is lower still).
- **Density panels**: pooled SAVI curve for *ADRB2* can look elevated because draws are stacked across donors; per-donor facets show the mismatch between one extreme line and four lines near the grey null.
- **Single-cell UMAP** (below, with *IL1B*): SCT-normalized *ADRB2* on disease-associated monocytes, split by **`Category`**, to see whether the atlas outlier is visible at cell level or confined to one donor's monocytes.

# Single-cell expression on UMAP (GEO Seurat object)

Published scRNA-seq object [GSE226598](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE226598) (`prepare_data.R`). We subset **`params$savi_sc_celltype`** (disease-associated monocytes — same population as the pseudobulk merge) and plot signature genes on the embedded UMAP, **`split.by = "Category"`** (`CTRL`, `SAVI`, `SAVI_treated`), using the **`SCT`** assay **`data`** slot (SCTransform–normalized expression from the published object).

```{r}
#| label: load-savi-seurat
#| cache: false

read_geo_rds_gz <- function(path) {
  con <- gzcon(gzfile(path))
  on.exit(close(con), add = TRUE)
  readRDS(con)
}

seurat_savi <- read_geo_rds_gz(params$savi_seurat_gz)

if (!"Category" %in% colnames(seurat_savi@meta.data)) {
  stop("`Category` not found in Seurat metadata.")
}

seurat_savi$Category <- factor(
  seurat_savi$Category,
  levels = c("CTRL", "SAVI", "SAVI_treated")
)

seurat_reduction <- function(se) {
  nms <- names(se@reductions)
  hit <- intersect(c("umap", "UMAP", "tsne", "TSNE"), nms)
  if (length(hit)) {
    return(hit[[1]])
  }
  nms[[1]]
}

reduction_use <- seurat_reduction(seurat_savi)

# Subset early so we don't keep the full >2k x 24k object around.
seurat_savi_mono <- seurat_savi |>
  subset(CellType == params$savi_sc_celltype)
# rm(seurat_savi)
# gc(verbose = FALSE)

tibble::tibble(
  n_cells = ncol(seurat_savi_mono),
  reduction = reduction_use,
  categories = paste(levels(seurat_savi_mono$Category), collapse = ", "),
  il1b_present = "IL1B" %in% rownames(seurat_savi_mono),
  adrb2_present = "ADRB2" %in% rownames(seurat_savi_mono)
)
```

```{r}
#| label: subset-savi-monocytes
if (ncol(seurat_savi_mono) == 0L) {
  stop(
    "No cells for CellType = '", params$savi_sc_celltype,
    "'. Check `CellType` levels in the Seurat object."
  )
}

seurat_savi_mono$Category <- factor(
  seurat_savi_mono$Category,
  levels = c("CTRL", "SAVI", "SAVI_treated")
)

tibble::tibble(
  cell_type = params$savi_sc_celltype,
  n_cells = ncol(seurat_savi_mono)
)

table(seurat_savi_mono$Category, useNA = "ifany")
```

```{r}
#| label: plot-il1b-featureplot-category
#| fig-width: 12
#| fig-height: 4
#| fig-align: center

if (!"SCT" %in% SeuratObject::Assays(seurat_savi_mono)) {
  stop(
    "SCT assay not found. Available: ",
    paste(SeuratObject::Assays(seurat_savi_mono), collapse = ", ")
  )
}

if (!"IL1B" %in% rownames(seurat_savi_mono[["SCT"]])) {
  stop("IL1B not found in SCT assay rownames.")
}

Seurat::FeaturePlot(
  seurat_savi_mono,
  features = "IL1B",
  assay = "SCT",
  split.by = "Category",
  reduction = reduction_use,
  slot = "data",
  ncol = 3,
  order = TRUE,
  pt.size = 0.35,
  min.cutoff = "q10",
  max.cutoff = "q90"
) &
  ggplot2::theme(
    plot.title = ggplot2::element_text(size = 11),
    strip.text = ggplot2::element_text(size = 11, face = "bold")
  )
```

Panel titles are **`Category`** (study cohort label from GEO metadata). Expression is from **SCTransform** (`assay = "SCT"`, `slot = "data"`), not raw **`RNA`** counts. High **`IL1B`** in **CTRL** monocytes complements the pseudobulk / atlas falsification checks above.

#### *ADRB2* on UMAP (all donors)

Same layout as *IL1B*: disease-associated monocytes, SCT **`data`**, **`split.by = "Category"`**. Compare the **SAVI** column to **CTRL** and **SAVI_treated** — cells from all donors in each arm are pooled on the UMAP.

```{r}
#| label: plot-adrb2-featureplot-category
#| fig-width: 12
#| fig-height: 4
#| fig-align: center

if (!"ADRB2" %in% rownames(seurat_savi_mono[["SCT"]])) {
  stop("ADRB2 not found in SCT assay rownames.")
}

Seurat::FeaturePlot(
  seurat_savi_mono,
  features = "ADRB2",
  assay = "SCT",
  split.by = "Category",
  reduction = reduction_use,
  slot = "data",
  ncol = 3,
  order = TRUE,
  pt.size = 0.35,
  min.cutoff = "q10",
  max.cutoff = "q90"
) &
  ggplot2::theme(
    plot.title = ggplot2::element_text(size = 11),
    strip.text = ggplot2::element_text(size = 11, face = "bold")
  )
```

The **SAVI** panel above pools monocytes from all five untreated SAVI donors. Pseudobulk checks show atlas deviation is dominated by **`P6_STING_ht`** (4057 raw counts, **`p_above_healthy < 0.025`**); the other four SAVI donors sit near the healthy null (≤65 raw counts). If **`ADRB2`** signal in **SAVI** is sparse overall but driven by cells from that donor, the pooled UMAP can still look elevated.

#### *ADRB2* after removing the atlas-tail donor

Mirroring **`plot-adrb2-leave-one-out`** on pseudobulk, we repeat the UMAP with **every disease-associated monocyte from the atlas-tail SAVI donor removed** (`adrb2_savi_outlier$Sample`, here **`P6_STING_ht`**). Assay, cell type, and **`split.by = "Category"`** are unchanged; only that library’s cells are dropped, so the **SAVI** facet reflects the remaining four untreated SAVI donors. If the **SAVI** panel now resembles **CTRL** / **SAVI_treated**, the cohort-level **`ADRB2`** signal is a **one-donor** artefact at single-cell resolution as well as in pseudobulk.

```{r}
#| label: plot-adrb2-featureplot-category_outlier_removed
#| fig-width: 12
#| fig-height: 4
#| fig-align: center

if (!"ADRB2" %in% rownames(seurat_savi_mono[["SCT"]])) {
  stop("ADRB2 not found in SCT assay rownames.")
}

if (!"Sample" %in% colnames(seurat_savi_mono@meta.data)) {
  stop("`Sample` not found in Seurat metadata.")
}

outlier_sample <- adrb2_savi_outlier$Sample

seurat_savi_mono_no_outlier <- subset(
  seurat_savi_mono,
  subset = Sample != outlier_sample
)

Seurat::FeaturePlot(
  seurat_savi_mono_no_outlier,
  features = "ADRB2",
  assay = "SCT",
  split.by = "Category",
  reduction = reduction_use,
  slot = "data",
  ncol = 3,
  order = TRUE,
  pt.size = 0.35,
  min.cutoff = "q10",
  max.cutoff = "q90"
) &
  ggplot2::theme(
    plot.title = ggplot2::element_text(size = 11),
    strip.text = ggplot2::element_text(size = 11, face = "bold")
  )
```

**Read together:** the all-donor and leave-one-out UMAPs ask the same question as the pseudobulk falsification — is **`ADRB2`** elevated across **SAVI**, or only when **`P6_STING_ht`** is included? Concordance between a muted **SAVI** panel after exclusion and the drop in SAVI mean **`log2fc`** in **`plot-adrb2-leave-one-out`** strengthens the **`falsified_no_savi_signal`** classification (20% of SAVI donors pass the strict atlas tail test).

------------------------------------------------------------------------

# Composition analysis vs original SAVI paper

In the original SAVI scRNA-seq paper (de Cevins *et al.* 2023; GSE226598), after QC, integration, clustering, and annotation, the authors report **23 cell populations** in the SAVI dataset and **29 cell populations** in the IFN-beta dataset. They then compare PBMC composition between SAVI and controls (Figure 1E), with p-values from Kruskal-Wallis tests followed by post hoc Dunn's multiple-comparison testing.

For whole-PBMC composition, the paper reports increased naive CD4 and naive CD8 T cells, decreased effector/memory CD4 and CD8 T-cell compartments, decreased NK cells (especially NK CD56bright), and strong depletion of MAIT and gamma-delta T cells in SAVI. They also report that cell-type proportions did not differ significantly before versus under JAK-inhibitor treatment.

In this notebook, we use those statements as the reference conclusion, then:

1. compute per-sample proportions from GEO **`CellType`** (23 populations),
2. map **`CellType` → Immune Atlas `L3`** labels used by the sccomp composition model,
3. compare observed proportions to healthy **`predict_res`** draws matched on **`sex`** and **`age_decade`**,
4. summarise trends in **`CTRL`**, **`SAVI`**, and **`SAVI_treated`** against the paper’s Figure 1E conclusions.

```{r}
#| label: summarize-paper-composition
paper_composition_summary <- tibble::tribble(
  ~topic, ~paper_finding,
  "Dataset size after integration", "23 populations (SAVI dataset) and 29 populations (IFN-beta dataset).",
  "Naive T-cell compartments", "Naive CD4 and naive CD8 T-cell proportions are increased in SAVI.",
  "Effector/memory T-cell compartments", "Effector/memory CD4 and CD8 compartments are decreased in SAVI.",
  "Innate-like / cytotoxic compartments", "NK cells (notably NK CD56bright), MAIT cells, and gamma-delta T cells are reduced in SAVI.",
  "Treatment comparison", "No significant composition differences were observed before versus under JAK inhibitor treatment.",
  "Statistics in Figure 1E", "Kruskal-Wallis tests with post hoc Dunn's multiple comparisons."
)

paper_composition_summary

paper_ct_alignment <- tibble::tribble(
  ~paper_compartment, ~direction_in_savi, ~celltype_labels_in_this_object, ~mapping_note,
  "Naive CD4 T", "increased", "0. Naive CD4", "Directly aligned.",
  "Effector/memory CD4", "decreased", "3. CM CD4; 4. EM CD4; 8. TE CD4; 9. Tregs and TE CD4", "Approximate effector/memory CD4 pool from available labels.",
  "Naive CD8 T", "increased", "2. Naive CD8", "Directly aligned.",
  "Effector/memory CD8", "decreased", "14. CM CD8; 15. EM CD8; 7. TE CD8", "Approximate effector/memory CD8 pool from available labels.",
  "NK (especially CD56bright)", "decreased", "10. NK; 26. NK CD56bright", "CD56bright is explicitly available.",
  "MAIT", "strongly decreased", "22. MAIT", "Directly aligned.",
  "gamma-delta T", "strongly decreased", "16. gd T cells", "Directly aligned.",
  "Myeloid context (not primary Figure 1E claim)", "context", "5. Classical monocytes; 12. Non-classical monocytes; 17. Disease-associated monocytes; 20. cDCs; 27. pDCs", "Contextual comparison only.",
  "B-cell context (not primary Figure 1E claim)", "context", "1. Naive B; 6. Transitionnal B; 13. Memory B; 11. CD3+ B cells; 25. Plasmablasts", "Contextual comparison only."
)

paper_ct_alignment
```

```{r}
#| label: composition-proportions-build
#| cache: false

if (!exists("read_geo_rds_gz", mode = "function")) {
  read_geo_rds_gz <- function(path) {
    con <- gzcon(gzfile(path))
    on.exit(close(con), add = TRUE)
    readRDS(con)
  }
}

if (!exists("seurat_savi", inherits = FALSE)) {
  seurat_savi <- read_geo_rds_gz(params$savi_seurat_gz)
}

if (!"Category" %in% colnames(seurat_savi@meta.data)) {
  stop("`Category` not found in Seurat metadata.")
}
if (!"CellType" %in% colnames(seurat_savi@meta.data)) {
  stop("`CellType` not found in Seurat metadata.")
}

seurat_savi$Category <- factor(
  seurat_savi$Category,
  levels = c("CTRL", "SAVI", "SAVI_treated")
)

# GEO sample metadata (same table as pseudobulk join above).
geo_tbl_comp <- geo_tbl |>
  dplyr::mutate(
    sex = dplyr::case_when(
      .data$gse_sex == "M" ~ "male",
      .data$gse_sex == "F" ~ "female",
      TRUE ~ NA_character_
    ),
    age_decade = as.character(ceiling(.data$gse_age_years / 10))
  ) |>
  dplyr::select("Sample", "sex", "age_decade", "gse_age_years")

# Fine-grained CellType -> Immune Atlas L3 (posteriorHCA / sccomp) labels.
celltype_to_l3 <- tibble::tribble(
  ~cell_type, ~L3,
  "0. Naive CD4", "cd4 naive",
  "2. Naive CD8", "cd8 naive",
  "3. CM CD4", "cd4 tcm",
  "4. EM CD4", "cd4 th1 em",
  "8. TE CD4", "cd4 th1 em",
  "9. Tregs and TE CD4", "treg",
  "14. CM CD8", "cd8 tcm",
  "15. EM CD8", "cd8 tem",
  "7. TE CD8", "cd8 tem",
  "10. NK", "nk",
  "26. NK CD56bright", "nk",
  "22. MAIT", "mait",
  "16. gd T cells", "tgd",
  "1. Naive B", "b naive",
  "6. Transitionnal B", "b naive",
  "13. Memory B", "b memory",
  "11. CD3+ B cells", "b memory",
  "5. Classical monocytes", "cd14 mono",
  "12. Non-classical monocytes", "cd16 mono",
  "17. Disease-associated monocytes", "cd14 mono",
  "20. cDCs", "cdc",
  "27. pDCs", "pdc",
  "25. Plasmablasts", "plasma"
)

cell_tbl <- seurat_savi@meta.data |>
  tibble::rownames_to_column("cell_id") |>
  dplyr::transmute(
    sample_id = as.character(.data$Sample),
    Category = as.character(.data$Category),
    cell_type = as.character(.data$CellType)
  ) |>
  dplyr::left_join(celltype_to_l3, by = "cell_type") |>
  dplyr::left_join(geo_tbl_comp, by = c("sample_id" = "Sample"))

example_proportions_celltype <- cell_tbl |>
  dplyr::filter(!is.na(.data$cell_type), nzchar(.data$cell_type)) |>
  dplyr::count(.data$sample_id, .data$cell_type, name = "n_cells") |>
  dplyr::group_by(.data$sample_id) |>
  dplyr::mutate(proportion = .data$n_cells / sum(.data$n_cells)) |>
  dplyr::ungroup() |>
  dplyr::select(.data$sample_id, .data$cell_type, .data$proportion)

# L3-level observed proportions (merge CellType labels that map to the same L3).
example_proportions <- cell_tbl |>
  dplyr::filter(!is.na(.data$L3)) |>
  dplyr::count(.data$sample_id, .data$Category, .data$sex, .data$age_decade, .data$L3, name = "n_cells") |>
  dplyr::group_by(.data$sample_id, .data$Category, .data$sex, .data$age_decade, .data$L3) |>
  dplyr::summarise(proportion = sum(.data$n_cells), .groups = "drop_last") |>
  dplyr::group_by(.data$sample_id) |>
  dplyr::mutate(proportion = .data$proportion / sum(.data$proportion)) |>
  dplyr::ungroup() |>
  dplyr::rename(cell_type = L3) |>
  dplyr::arrange(.data$sample_id, .data$cell_type)

unmapped_celltypes <- cell_tbl |>
  dplyr::filter(is.na(.data$L3)) |>
  dplyr::count(.data$cell_type, sort = TRUE)

tibble::tibble(
  label_column_used = "CellType",
  n_samples = dplyr::n_distinct(example_proportions$sample_id),
  n_cells_total = nrow(cell_tbl),
  n_celltypes_raw = dplyr::n_distinct(example_proportions_celltype$cell_type),
  n_l3_types = dplyr::n_distinct(example_proportions$cell_type),
  n_unmapped_celltypes = nrow(unmapped_celltypes)
)

if (nrow(unmapped_celltypes) > 0L) {
  unmapped_celltypes
}
```

### Healthy atlas composition null (`predict_res`)

Posterior draws of blood PBMC composition from the Immune Atlas sccomp model (`disease_groups___altered = Normal`), stratified by **`sex`** (`male` / `female`) and **`age_decade`** (`1`–`4`), matching SAVI donor metadata. The draw table is **not** knitr-cached (too large for R’s cache serializer); it is saved once to **`params$composition_predict_rds`** and trimmed to L3 types present in the SAVI cohort.

```{r}
#| label: composition-predict-healthy
#| cache: false

if (!requireNamespace("sccomp", quietly = TRUE)) {
  stop("Package `sccomp` is required for composition prediction.")
}

l3_for_comparison <- example_proportions |>
  dplyr::distinct(.data$cell_type) |>
  dplyr::pull(.data$cell_type)

trim_predict_res <- function(x, cell_types) {
  x |>
    dplyr::filter(.data$cell_type %in% cell_types) |>
    dplyr::select(dplyr::any_of(c(
      "sample_id",
      "age_decade",
      "sex",
      "cell_type",
      "proportion",
      ".draw"
    )))
}

if (file.exists(params$composition_predict_rds)) {
  predict_res <- readRDS(params$composition_predict_rds)
} else {
  sccomp_est <- readRDS(params$sccomp_estimates_rds)

  sample_metadata <- tidyr::expand_grid(
    age_decade = as.character(1:4),
    sex = c("male", "female")
  ) |>
    dplyr::mutate(
      assay_groups___altered = "10x Genomics 3",
      disease_groups___altered = "Normal",
      tissue_groups = "blood",
      ethnicity_groups_imputed = NA,
      dataset_id___altered = NA,
      sample_id = paste0("S", dplyr::row_number())
    )

  predict_res <- sccomp_est |>
    sccomp::sccomp_predict(
      formula_composition = NULL,
      new_data = sample_metadata,
      summary_instead_of_draws = FALSE
    ) |>
    dplyr::rename(cell_type = L3) |>
    trim_predict_res(l3_for_comparison)

  predict_dir <- dirname(params$composition_predict_rds)
  if (!dir.exists(predict_dir)) {
    dir.create(predict_dir, recursive = TRUE, showWarnings = FALSE)
  }
  saveRDS(predict_res, params$composition_predict_rds, compress = "xz")
}

predict_res <- trim_predict_res(predict_res, l3_for_comparison)

predict_res |>
  dplyr::distinct(.data$age_decade, .data$sex, .data$cell_type) |>
  dplyr::count(.data$age_decade, .data$sex, name = "n_cell_types")
```

### Compare observed SAVI proportions to healthy predictive draws

For each donor we match **`predict_res`** rows with the same **`sex`** and **`age_decade`**, then compute empirical tail probabilities (same logic as `posteriorHCA::composition_posterior_test()`).

```{r}
#| label: composition-vs-healthy

composition_vs_healthy <- example_proportions |>
  dplyr::rename(
    sample_id_observed = sample_id,
    cell_type_observed = cell_type,
    proportion_observed = proportion
  ) |>
  dplyr::inner_join(
    predict_res |>
      dplyr::select(
        age_decade, sex, cell_type, proportion_sampled = proportion, .draw
      ),
    by = c("sex" = "sex", "age_decade" = "age_decade", "cell_type_observed" = "cell_type"),
    relationship = "many-to-many"
  ) |>
  dplyr::group_by(.data$sample_id_observed, .data$Category, .data$cell_type_observed) |>
  dplyr::summarise(
    proportion_observed = dplyr::first(.data$proportion_observed),
    healthy_median = stats::median(.data$proportion_sampled),
    healthy_mean = mean(.data$proportion_sampled),
    healthy_hpd_lo = stats::quantile(.data$proportion_sampled, 0.025),
    healthy_hpd_hi = stats::quantile(.data$proportion_sampled, 0.975),
    # Same tail convention as `gene_deviation_from_fit()`: P(draw >= obs), not P(obs > draw).
    p_above_healthy = mean(.data$proportion_sampled >= .data$proportion_observed),
    p_below_healthy = mean(.data$proportion_sampled <= .data$proportion_observed),
    empirical_confidence = 2 * pmin(.data$p_above_healthy, .data$p_below_healthy),
    direction = dplyr::case_when(
      .data$p_above_healthy < 0.025 ~ "higher_than_healthy",
      .data$p_below_healthy < 0.025 ~ "lower_than_healthy",
      TRUE ~ "consistent_with_healthy"
    ),
    .groups = "drop"
  )

composition_vs_healthy |>
  dplyr::arrange(.data$Category, .data$cell_type_observed, .data$sample_id_observed) |>
  dplyr::slice_head(n = 12)
```

### Paper compartments: trend by `Category` and comparison to published direction

We aggregate mapped L3 labels into the Figure 1E compartments from the paper, then summarise mean proportion per **`Category`** and compare the SAVI–CTRL direction to the published expectation.

```{r}
#| label: composition-paper-compartments

paper_compartment_l3 <- tibble::tribble(
  ~paper_compartment, ~direction_in_savi, ~cell_type,
  "Naive CD4 T", "increased", "cd4 naive",
  "Effector/memory CD4", "decreased", "cd4 tcm",
  "Effector/memory CD4", "decreased", "cd4 th1 em",
  "Effector/memory CD4", "decreased", "cd4 th17 em",
  "Effector/memory CD4", "decreased", "cd4 th2 em",
  "Effector/memory CD4", "decreased", "cd4 fh em",
  "Effector/memory CD4", "decreased", "cd4 th1/th17 em",
  "Effector/memory CD4", "decreased", "treg",
  "Naive CD8 T", "increased", "cd8 naive",
  "Effector/memory CD8", "decreased", "cd8 tcm",
  "Effector/memory CD8", "decreased", "cd8 tem",
  "NK", "decreased", "nk",
  "MAIT", "strongly decreased", "mait",
  "gamma-delta T", "strongly decreased", "tgd"
)

composition_paper_per_donor <- composition_vs_healthy |>
  dplyr::inner_join(paper_compartment_l3, by = c("cell_type_observed" = "cell_type")) |>
  dplyr::group_by(
    .data$sample_id_observed,
    .data$Category,
    .data$paper_compartment,
    .data$direction_in_savi
  ) |>
  dplyr::summarise(
    proportion_observed = sum(.data$proportion_observed),
    healthy_median = sum(.data$healthy_median),
    direction = dplyr::if_else(
      any(.data$direction == "higher_than_healthy"),
      "higher_than_healthy",
      dplyr::if_else(
        any(.data$direction == "lower_than_healthy"),
        "lower_than_healthy",
        "consistent_with_healthy"
      )
    ),
    .groups = "drop"
  )

# Donors with zero cells in a compartment are absent from counts; fill with 0 so
# category means (and CTRL dots) include all donors in each Category.
all_donors_comp <- example_proportions |>
  dplyr::distinct(.data$sample_id, .data$Category) |>
  dplyr::rename(sample_id_observed = sample_id)

compartment_meta <- paper_compartment_l3 |>
  dplyr::distinct(.data$paper_compartment, .data$direction_in_savi)

composition_paper_long <- tidyr::crossing(all_donors_comp, compartment_meta) |>
  dplyr::left_join(
    composition_paper_per_donor,
    by = c("sample_id_observed", "Category", "paper_compartment", "direction_in_savi")
  ) |>
  dplyr::mutate(
    proportion_observed = dplyr::coalesce(.data$proportion_observed, 0),
    healthy_median = dplyr::coalesce(.data$healthy_median, 0),
    direction = dplyr::coalesce(.data$direction, "consistent_with_healthy")
  )

composition_paper_summary <- composition_paper_long |>
  dplyr::group_by(.data$paper_compartment, .data$direction_in_savi, .data$Category) |>
  dplyr::summarise(
    n_samples = dplyr::n(),
    mean_prop = mean(.data$proportion_observed),
    median_prop = stats::median(.data$proportion_observed),
    mean_healthy_median = mean(.data$healthy_median),
    .groups = "drop"
  ) |>
  tidyr::pivot_wider(
    names_from = Category,
    values_from = c(mean_prop, median_prop, n_samples),
    names_sep = "__"
  ) |>
  dplyr::mutate(
    delta_savi_minus_ctrl = .data$mean_prop__SAVI - .data$mean_prop__CTRL,
    delta_treated_minus_ctrl = .data$mean_prop__SAVI_treated - .data$mean_prop__CTRL,
    matches_paper_savi = dplyr::case_when(
      .data$direction_in_savi == "increased" ~ .data$delta_savi_minus_ctrl > 0,
      .data$direction_in_savi %in% c("decreased", "strongly decreased") ~ .data$delta_savi_minus_ctrl < 0,
      TRUE ~ NA
    ),
    matches_paper_treated = dplyr::case_when(
      .data$direction_in_savi == "increased" ~ .data$delta_treated_minus_ctrl > 0,
      .data$direction_in_savi %in% c("decreased", "strongly decreased") ~ .data$delta_treated_minus_ctrl < 0,
      TRUE ~ NA
    )
  ) |>
  dplyr::arrange(.data$paper_compartment)

composition_paper_summary
```

### Healthy baseline vs age decade

Healthy atlas predictive draws (`predict_res`, Normal blood): for each **paper compartment**, sum mapped L3 proportions per draw, then summarise by **`sex`** and **`age_decade`** (1–4).

```{r}
#| label: plot-composition-healthy-baseline-by-age
#| fig-width: 12
#| fig-height: 9
#| fig-align: center

healthy_baseline_compartment_age <- predict_res |>
  dplyr::inner_join(
    paper_compartment_l3 |>
      dplyr::distinct(.data$paper_compartment, .data$cell_type),
    by = "cell_type"
  ) |>
  dplyr::group_by(
    .data$paper_compartment,
    .data$sex,
    .data$age_decade,
    .data$.draw
  ) |>
  dplyr::summarise(
    compartment_proportion = sum(.data$proportion),
    .groups = "drop"
  ) |>
  dplyr::group_by(
    .data$paper_compartment,
    .data$sex,
    .data$age_decade
  ) |>
  dplyr::summarise(
    median_proportion = stats::median(.data$compartment_proportion),
    hpd_lo = stats::quantile(.data$compartment_proportion, 0.025),
    hpd_hi = stats::quantile(.data$compartment_proportion, 0.975),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    age_decade = factor(.data$age_decade, levels = as.character(1:4)),
    sex = factor(.data$sex, levels = c("female", "male")),
    paper_compartment = factor(
      .data$paper_compartment,
      levels = unique(paper_compartment_l3$paper_compartment)
    )
  )

sex_cols <- c(female = "#E41A1C", male = "#377EB8")

ggplot2::ggplot(
  healthy_baseline_compartment_age,
  ggplot2::aes(
    x = age_decade,
    y = median_proportion,
    color = sex,
    group = sex
  )
) +
  ggplot2::geom_ribbon(
    ggplot2::aes(ymin = hpd_lo, ymax = hpd_hi, fill = sex),
    alpha = 0.15,
    color = NA
  ) +
  ggplot2::geom_line(linewidth = 0.9) +
  ggplot2::geom_point(size = 2.2) +
  ggplot2::facet_wrap(~paper_compartment, scales = "free_y", ncol = 2) +
  ggplot2::scale_color_manual(values = sex_cols, name = "Sex") +
  ggplot2::scale_fill_manual(values = sex_cols, guide = "none") +
  ggplot2::labs(
    title = "Healthy atlas compartment proportion by sex and age decade",
    subtitle = paste(
      "Lines = posterior median per decade; ribbon = 2.5–97.5% across predictive draws.",
      "Compartment mass = sum of mapped L3 proportions (sccomp Normal blood)."
    ),
    x = "Age decade",
    y = "Compartment proportion (healthy baseline)"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    legend.position = "bottom",
    strip.text = ggplot2::element_text(face = "bold", size = 9)
  )
```

### Observed composition (within cohort)

Boxplot and bar chart use **observed** compartment proportions only (CTRL / SAVI / SAVI_treated); no healthy atlas reference.

```{r}
#| label: plot-composition-paper-box
#| fig-width: 12
#| fig-height: 12
#| fig-align: center

category_cols <- c(
  CTRL = "#4DAF4A",
  SAVI = "#E41A1C",
  SAVI_treated = "#377EB8"
)

plot_data <- composition_paper_long |>
  dplyr::mutate(
    Category = factor(.data$Category, levels = names(category_cols)),
    paper_compartment = forcats::fct_reorder(
      .data$paper_compartment,
      .data$proportion_observed,
      .fun = median,
      .desc = TRUE
    )
  )

ggplot2::ggplot(
  plot_data,
  ggplot2::aes(
    x = Category,
    y = proportion_observed,
    fill = Category
  )
) +
  ggplot2::geom_boxplot(outlier.shape = NA, alpha = 0.75) +
  ggplot2::geom_jitter(width = 0.12, size = 1.6, alpha = 0.85) +
  ggplot2::facet_wrap(~paper_compartment, scales = "free_y", ncol = 2) +
  ggplot2::scale_fill_manual(values = category_cols) +
  ggplot2::labs(
    title = "SAVI composition vs paper compartments (observed CellType -> L3)",
    subtitle = "Each point = one donor; compartments aggregate mapped L3 labels",
    x = NULL,
    y = "Proportion within sample"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "none")
```

```{r}
#| label: plot-composition-paper-vs-atlas
#| fig-width: 12
#| fig-height: 14
#| fig-align: center

composition_paper_bar <- composition_paper_summary |>
  tidyr::pivot_longer(
    cols = c(
      tidyselect::starts_with("mean_prop__"),
      tidyselect::starts_with("median_prop__")
    ),
    names_to = c("statistic", "Category"),
    names_sep = "__",
    values_to = "prop"
  ) |>
  dplyr::mutate(
    statistic = dplyr::recode(
      .data$statistic,
      mean_prop = "Mean across donors",
      median_prop = "Median across donors"
    ),
    statistic = factor(
      .data$statistic,
      levels = c("Mean across donors", "Median across donors")
    ),
    Category = factor(.data$Category, levels = names(category_cols))
  )

ggplot2::ggplot(
  composition_paper_bar,
  ggplot2::aes(x = Category, y = prop, fill = Category)
) +
  ggplot2::geom_col(position = ggplot2::position_dodge(width = 0.8), width = 0.7) +
  ggplot2::facet_grid(
    paper_compartment ~ statistic,
    scales = "free_y"
  ) +
  ggplot2::scale_fill_manual(values = category_cols) +
  ggplot2::labs(
    title = "Observed compartment proportion by Category",
    subtitle = paste(
      "Left column = mean across donors; right column = median.",
      "Observed data only (no atlas); compare SAVI vs CTRL with paper expectation",
      "(direction_in_savi in composition_paper_summary)."
    ),
    x = NULL,
    y = "Proportion within sample"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    legend.position = "bottom",
    strip.text = ggplot2::element_text(size = 9, face = "bold")
  )
```

### SAVI donors vs healthy baseline (unified compartments)

For **`SAVI`** and **`SAVI_treated`** donors only, we compare each donor’s **compartment proportion** (sum of mapped L3 types) to healthy **`predict_res`** draws matched on **`sex`** and **`age_decade`**. We then ask whether the atlas tail call (`higher` / `lower` / `consistent` vs healthy) agrees with the **paper’s expected direction** in SAVI (increased vs decreased).

```{r}
#| label: composition-savi-vs-healthy-compartments

baseline_compartment_draws <- predict_res |>
  dplyr::inner_join(
    paper_compartment_l3 |>
      dplyr::distinct(.data$paper_compartment, .data$cell_type),
    by = "cell_type"
  ) |>
  dplyr::group_by(
    .data$age_decade,
    .data$sex,
    .data$paper_compartment,
    .data$.draw
  ) |>
  dplyr::summarise(
    compartment_proportion = sum(.data$proportion),
    .groups = "drop"
  )

donor_sex_age_all <- example_proportions |>
  dplyr::distinct(.data$sample_id, .data$sex, .data$age_decade) |>
  dplyr::rename(sample_id_observed = sample_id)

# One reference table for all donors: tail test vs sex/age-matched healthy draws.
composition_matched_reference <- composition_paper_long |>
  dplyr::left_join(donor_sex_age_all, by = "sample_id_observed") |>
  dplyr::inner_join(
    baseline_compartment_draws,
    by = c("sex", "age_decade", "paper_compartment"),
    relationship = "many-to-many"
  ) |>
  dplyr::group_by(
    .data$sample_id_observed,
    .data$Category,
    .data$sex,
    .data$age_decade,
    .data$paper_compartment,
    .data$direction_in_savi,
    .data$proportion_observed
  ) |>
  dplyr::summarise(
    healthy_median = stats::median(.data$compartment_proportion),
    healthy_hpd_lo = stats::quantile(.data$compartment_proportion, 0.025),
    healthy_hpd_hi = stats::quantile(.data$compartment_proportion, 0.975),
    p_above_healthy = mean(.data$compartment_proportion >= .data$proportion_observed),
    p_below_healthy = mean(.data$compartment_proportion <= .data$proportion_observed),
    empirical_confidence = 2 * pmin(.data$p_above_healthy, .data$p_below_healthy),
    direction_vs_healthy = dplyr::case_when(
      .data$p_above_healthy < 0.025 ~ "higher_than_healthy",
      .data$p_below_healthy < 0.025 ~ "lower_than_healthy",
      TRUE ~ "consistent_with_healthy"
    ),
    .groups = "drop"
  )

composition_savi_vs_healthy <- composition_matched_reference |>
  dplyr::filter(.data$Category %in% c("SAVI", "SAVI_treated")) |>
  dplyr::mutate(
    atlas_direction_simple = dplyr::case_when(
      .data$direction_vs_healthy == "higher_than_healthy" ~ "increased",
      .data$direction_vs_healthy == "lower_than_healthy" ~ "decreased",
      TRUE ~ "consistent"
    ),
    agrees_with_paper = dplyr::case_when(
      .data$direction_in_savi == "increased" ~
        .data$atlas_direction_simple == "increased",
      .data$direction_in_savi %in% c("decreased", "strongly decreased") ~
        .data$atlas_direction_simple == "decreased",
      TRUE ~ NA
    ),
    agreement_label = dplyr::case_when(
      .data$agrees_with_paper %in% TRUE ~ "agrees (atlas vs paper)",
      .data$agrees_with_paper %in% FALSE ~ "disagrees (atlas vs paper)",
      TRUE ~ "atlas consistent (no tail)"
    )
  )

composition_savi_vs_healthy |>
  dplyr::arrange(.data$Category, .data$paper_compartment, .data$sample_id_observed)

composition_savi_agreement_summary <- composition_savi_vs_healthy |>
  dplyr::group_by(.data$Category, .data$paper_compartment, .data$direction_in_savi) |>
  dplyr::summarise(
    n_donors = dplyr::n(),
    n_higher = sum(.data$direction_vs_healthy == "higher_than_healthy"),
    n_lower = sum(.data$direction_vs_healthy == "lower_than_healthy"),
    n_consistent = sum(.data$direction_vs_healthy == "consistent_with_healthy"),
    n_agrees_paper = sum(.data$agrees_with_paper %in% TRUE, na.rm = TRUE),
    n_disagrees_paper = sum(.data$agrees_with_paper %in% FALSE, na.rm = TRUE),
    n_testable = sum(!is.na(.data$agrees_with_paper)),
    pct_agrees_paper = dplyr::if_else(
      sum(!is.na(.data$agrees_with_paper)) > 0L,
      100 * mean(.data$agrees_with_paper, na.rm = TRUE),
      NA_real_
    ),
    .groups = "drop"
  ) |>
  dplyr::arrange(.data$Category, .data$paper_compartment)

composition_savi_agreement_summary

# Diagnostic: heatmap tail call vs observed relative to matched healthy median.
composition_naive_cd4_check <- composition_matched_reference |>
  dplyr::filter(
    .data$paper_compartment == "Naive CD4 T",
    .data$Category == "SAVI"
  ) |>
  dplyr::transmute(
    sample_id_observed,
    age_decade,
    sex,
    proportion_observed,
    healthy_median_matched = .data$healthy_median,
    delta_obs_minus_matched = .data$proportion_observed - .data$healthy_median,
    direction_vs_healthy,
    visual_check = dplyr::case_when(
      .data$direction_vs_healthy == "higher_than_healthy" ~
        "solid line should be RIGHT of dashed median",
      .data$direction_vs_healthy == "lower_than_healthy" ~
        "solid line should be LEFT of dashed median",
      TRUE ~ "solid line near dashed median"
    )
  )

composition_naive_cd4_check
```

### How much CTRL samples in data agree with our healthy baseline

In-study **CTRL** donors are the paper’s reference for “healthy”. If they are exchangeable with the Immune Atlas healthy baseline, each **CTRL** pseudobulk compartment proportion should fall inside the **sex/age-matched** predictive null (`direction_vs_healthy == consistent_with_healthy`). Deviations flag compartments where study controls already sit above or below the atlas before comparing SAVI.

```{r}
#| label: composition-ctrl-baseline

composition_ctrl_baseline <- composition_matched_reference |>
  dplyr::filter(.data$Category == "CTRL") |>
  dplyr::mutate(
    agrees_with_healthy_baseline = .data$direction_vs_healthy == "consistent_with_healthy",
    baseline_agreement = dplyr::case_when(
      .data$direction_vs_healthy == "consistent_with_healthy" ~ "consistent with healthy",
      .data$direction_vs_healthy == "higher_than_healthy" ~ "higher than healthy",
      .data$direction_vs_healthy == "lower_than_healthy" ~ "lower than healthy"
    ),
    baseline_agreement = factor(
      .data$baseline_agreement,
      levels = c(
        "consistent with healthy",
        "higher than healthy",
        "lower than healthy"
      )
    ),
    atlas_strength = 1 - .data$empirical_confidence,
    direction_short = dplyr::case_when(
      .data$direction_vs_healthy == "higher_than_healthy" ~ "↑",
      .data$direction_vs_healthy == "lower_than_healthy" ~ "↓",
      TRUE ~ "·"
    ),
    tile_label = sprintf(
      "%s\n%.2f",
      .data$direction_short,
      .data$atlas_strength
    ),
    paper_compartment = factor(
      .data$paper_compartment,
      levels = unique(paper_compartment_l3$paper_compartment)
    )
  )

composition_ctrl_baseline_summary <- composition_ctrl_baseline |>
  dplyr::group_by(.data$sample_id_observed, .data$sex, .data$age_decade) |>
  dplyr::summarise(
    n_compartments = dplyr::n(),
    n_consistent = sum(.data$agrees_with_healthy_baseline),
    n_higher = sum(.data$direction_vs_healthy == "higher_than_healthy"),
    n_lower = sum(.data$direction_vs_healthy == "lower_than_healthy"),
    pct_consistent = 100 * .data$n_consistent / .data$n_compartments,
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(.data$pct_consistent), .data$sample_id_observed)

composition_ctrl_baseline_overall <- composition_ctrl_baseline |>
  dplyr::summarise(
    n_ctrl_donors = dplyr::n_distinct(.data$sample_id_observed),
    n_tests = dplyr::n(),
    n_consistent = sum(.data$agrees_with_healthy_baseline),
    n_higher = sum(.data$direction_vs_healthy == "higher_than_healthy"),
    n_lower = sum(.data$direction_vs_healthy == "lower_than_healthy"),
    pct_consistent = 100 * .data$n_consistent / .data$n_tests,
    mean_pct_consistent_per_donor = mean(composition_ctrl_baseline_summary$pct_consistent)
  )

composition_ctrl_baseline_overall
composition_ctrl_baseline_summary
```

```{r}
#| label: plot-composition-ctrl-baseline-heatmap
#| fig-width: 12
#| fig-height: 5.5
#| fig-align: center

ctrl_baseline_cols <- c(
  "consistent with healthy" = "#4DAF4A",
  "higher than healthy" = "#E41A1C",
  "lower than healthy" = "#377EB8"
)

ctrl_sample_order <- composition_ctrl_baseline_summary |>
  dplyr::arrange(dplyr::desc(.data$pct_consistent), .data$sample_id_observed) |>
  dplyr::pull(.data$sample_id_observed)

ctrl_heatmap_data <- composition_ctrl_baseline |>
  dplyr::mutate(
    sample_id_observed = factor(.data$sample_id_observed, levels = ctrl_sample_order),
    tile_label = dplyr::if_else(
      .data$agrees_with_healthy_baseline,
      "·",
      .data$tile_label
    )
  )

ggplot2::ggplot(
  ctrl_heatmap_data,
  ggplot2::aes(
    x = paper_compartment,
    y = sample_id_observed,
    fill = baseline_agreement
  )
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.45) +
  ggplot2::geom_text(
    ggplot2::aes(label = tile_label),
    size = 2.5,
    lineheight = 0.85,
    color = "grey15"
  ) +
  ggplot2::scale_fill_manual(
    values = ctrl_baseline_cols,
    name = "vs matched\nhealthy null"
  ) +
  ggplot2::labs(
    title = "CTRL donors: agreement with sex/age-matched healthy atlas (paper compartments)",
    subtitle = paste0(
      "Green = consistent_with_healthy (",
      sprintf("%.0f", composition_ctrl_baseline_overall$pct_consistent),
      "% of donor × compartment tests cohort-wide). ",
      "Non-green tiles show direction (↑/↓) and tail strength (1 - empirical confidence)."
    ),
    x = NULL,
    y = "CTRL donor"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 35, hjust = 1, size = 8),
    axis.text.y = ggplot2::element_text(size = 9),
    legend.position = "bottom"
  )
```

Each panel below is one **CTRL donor × paper compartment**: grey density = healthy **`predict_res`** draws for that donor’s **`sex`** and **`age_decade`**; coloured vertical line = observed proportion. Line colour matches the heatmap (green = consistent, red/blue = tail call). Text at each line reports **`p_above_healthy`** = P(draw ≥ obs) and **`p_below_healthy`** = P(draw ≤ obs) (same tail convention as gene-level tests).

```{r}
#| label: plot-composition-ctrl-baseline-density
#| fig-width: 20
#| fig-height: 15
#| fig-align: center

compartment_levels <- paper_compartment_l3 |>
  dplyr::distinct(.data$paper_compartment) |>
  dplyr::pull(.data$paper_compartment)

if (!exists("baseline_compartment_draws")) {
  baseline_compartment_draws <- predict_res |>
    dplyr::inner_join(
      paper_compartment_l3 |>
        dplyr::distinct(.data$paper_compartment, .data$cell_type),
      by = "cell_type"
    ) |>
    dplyr::group_by(
      .data$paper_compartment,
      .data$age_decade,
      .data$sex,
      .data$.draw
    ) |>
    dplyr::summarise(
      compartment_proportion = sum(.data$proportion),
      .groups = "drop"
    )
}

ctrl_panel_meta <- composition_ctrl_baseline |>
  dplyr::distinct(
    .data$sample_id_observed,
    .data$sex,
    .data$age_decade,
    .data$paper_compartment
  ) |>
  dplyr::mutate(
    sample_panel = paste0(
      .data$sample_id_observed,
      " (",
      .data$sex,
      " | decade ",
      .data$age_decade,
      ")"
    )
  )

ctrl_panel_order <- composition_ctrl_baseline_summary |>
  dplyr::mutate(
    sample_panel = paste0(
      .data$sample_id_observed,
      " (",
      .data$sex,
      " | decade ",
      .data$age_decade,
      ")"
    )
  ) |>
  dplyr::arrange(dplyr::desc(.data$pct_consistent), .data$sample_id_observed)

ctrl_density_draws <- ctrl_panel_meta |>
  dplyr::inner_join(
    baseline_compartment_draws,
    by = c("sex", "age_decade", "paper_compartment")
  ) |>
  dplyr::mutate(
    sample_panel = factor(
      .data$sample_panel,
      levels = ctrl_panel_order$sample_panel
    ),
    paper_compartment = factor(.data$paper_compartment, levels = compartment_levels)
  ) |>
  dplyr::filter(
    is.finite(.data$compartment_proportion),
    .data$compartment_proportion >= 0
  )

ctrl_density_vlines <- composition_ctrl_baseline |>
  dplyr::mutate(
    sample_panel = paste0(
      .data$sample_id_observed,
      " (",
      .data$sex,
      " | decade ",
      .data$age_decade,
      ")"
    ),
    sample_panel = factor(
      .data$sample_panel,
      levels = ctrl_panel_order$sample_panel
    ),
    paper_compartment = factor(.data$paper_compartment, levels = compartment_levels),
    p_label = sprintf(
      "p_above=%.2f\np_below=%.2f",
      .data$p_above_healthy,
      .data$p_below_healthy
    )
  ) |>
  dplyr::filter(is.finite(.data$proportion_observed)) |>
  dplyr::left_join(
    ctrl_density_draws |>
      dplyr::group_by(.data$sample_panel, .data$paper_compartment) |>
      dplyr::summarise(
        x_mid = stats::median(.data$compartment_proportion, na.rm = TRUE),
        .groups = "drop"
      ),
    by = c("sample_panel", "paper_compartment")
  ) |>
  dplyr::mutate(
    text_hjust = dplyr::if_else(
      .data$proportion_observed > .data$x_mid,
      1.05,
      -0.05
    )
  )

ggplot2::ggplot(
  ctrl_density_draws,
  ggplot2::aes(x = compartment_proportion)
) +
  ggplot2::geom_density(
    fill = "grey80",
    color = "grey35",
    alpha = 0.55,
    linewidth = 0.35,
    na.rm = TRUE
  ) +
  ggplot2::geom_vline(
    data = ctrl_density_vlines,
    ggplot2::aes(
      xintercept = proportion_observed,
      color = baseline_agreement
    ),
    inherit.aes = FALSE,
    linewidth = 0.75
  ) +
  ggplot2::geom_text(
    data = ctrl_density_vlines,
    ggplot2::aes(
      x = proportion_observed,
      y = Inf,
      label = p_label,
      hjust = text_hjust
    ),
    inherit.aes = FALSE,
    vjust = 1.05,
    size = 2.1,
    lineheight = 0.85,
    color = "grey15"
  ) +
  ggplot2::facet_grid(
    sample_panel ~ paper_compartment,
    scales = "free",
    labeller = ggplot2::labeller(
      .cols = ggplot2::label_wrap_gen(10),
      .rows = ggplot2::label_wrap_gen(16)
    )
  ) +
  ggplot2::scale_color_manual(
    values = ctrl_baseline_cols,
    name = "CTRL vs\nmatched null"
  ) +
  ggplot2::scale_x_continuous(expand = ggplot2::expansion(mult = c(0.04, 0.06))) +
  ggplot2::labs(
    title = "CTRL donors vs sex/age-matched healthy atlas (density grid)",
    subtitle = paste(
      "Rows = CTRL donor (sex | age decade); columns = paper compartments.",
      "Grey = healthy predictive draws for that donor's metadata;",
      "vertical line = observed CTRL proportion;",
      "labels = p_above_healthy / p_below_healthy."
    ),
    x = "Compartment proportion",
    y = "Density"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    legend.position = "bottom",
    strip.text.x = ggplot2::element_text(size = 7, face = "bold"),
    strip.text.y = ggplot2::element_text(size = 7),
    panel.spacing = grid::unit(0.25, "lines")
  )
```

**Read:** donors sorted by **% compartments consistent** (top = closest to atlas). Widespread non-green tiles mean in-study controls are **not** drawn from the same reference as Immune Atlas healthy PBMCs for those compartments — relevant when interpreting SAVI vs atlas (and gene-level CTRL drift). The **density grid** shows the same comparison panel-by-panel: a line inside the grey band supports consistency; a line in the tail matches the heatmap colour.

### Healthy atlas comparison (plots)

Tail tests and visualizations below use **sex/age-matched** healthy **`predict_res`** draws (`composition_matched_reference`).

```{r}
#| label: plot-savi-atlas-direction-heatmap
#| fig-width: 12
#| fig-height: 7
#| fig-align: center

# empirical_confidence = 2 * min(P(draw>=obs), P(draw<=obs)): near 0 = strong tail;
# near 1 = central. Signed score: direction x (1 - empirical_confidence).
heatmap_data <- composition_savi_vs_healthy |>
  dplyr::mutate(
    Category = factor(.data$Category, levels = c("SAVI", "SAVI_treated")),
    atlas_strength = 1 - .data$empirical_confidence,
    atlas_signed_score = dplyr::case_when(
      .data$direction_vs_healthy == "higher_than_healthy" ~ .data$atlas_strength,
      .data$direction_vs_healthy == "lower_than_healthy" ~ -.data$atlas_strength,
      TRUE ~ 0
    ),
    direction_short = dplyr::case_when(
      .data$direction_vs_healthy == "higher_than_healthy" ~ "↑",
      .data$direction_vs_healthy == "lower_than_healthy" ~ "↓",
      TRUE ~ "·"
    ),
    tile_label = sprintf(
      "%s\n%.2f",
      .data$direction_short,
      .data$atlas_strength
    ),
    sample_label = paste0(.data$sample_id_observed, " (", .data$Category, ")"),
    paper_compartment = factor(
      .data$paper_compartment,
      levels = unique(paper_compartment_l3$paper_compartment)
    )
  )

score_lim <- max(abs(heatmap_data$atlas_signed_score), na.rm = TRUE)
score_lim <- if (is.finite(score_lim) && score_lim > 0) score_lim else 1

ggplot2::ggplot(
  heatmap_data,
  ggplot2::aes(
    x = paper_compartment,
    y = sample_label,
    fill = atlas_signed_score
  )
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.4) +
  ggplot2::geom_text(
    ggplot2::aes(label = tile_label),
    size = 2.4,
    lineheight = 0.85,
    color = "grey15"
  ) +
  ggplot2::scale_fill_gradient2(
    low = "#377EB8",
    mid = "grey92",
    high = "#E41A1C",
    midpoint = 0,
    limits = c(-score_lim, score_lim),
    oob = scales::squish,
    name = "Signed evidence\nvs healthy"
  ) +
  ggplot2::facet_grid(.data$Category ~ ., scales = "free_y", space = "free_y") +
  ggplot2::labs(
    title = "SAVI donors: strength of deviation from healthy atlas (unified compartments)",
    subtitle = paste(
      "Tail test vs sex/age-matched healthy draws (not pooled across all decades).",
      "Red = higher, blue = lower; intensity = 1 - empirical confidence."
    ),
    x = NULL,
    y = NULL
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 35, hjust = 1, size = 8),
    axis.text.y = ggplot2::element_text(size = 7),
    strip.text = ggplot2::element_text(face = "bold"),
    legend.position = "bottom"
  )
```

```{r}
#| label: plot-savi-atlas-direction-counts
#| fig-width: 12
#| fig-height: 6
#| fig-align: center

direction_cols <- c(
  higher_than_healthy = "#E41A1C",
  lower_than_healthy = "#377EB8",
  consistent_with_healthy = "grey75"
)

count_long <- composition_savi_agreement_summary |>
  tidyr::pivot_longer(
    cols = c(n_higher, n_lower, n_consistent),
    names_to = "atlas_call",
    values_to = "n"
  ) |>
  dplyr::mutate(
    Category = factor(.data$Category, levels = c("SAVI", "SAVI_treated")),
    atlas_call = dplyr::recode(
      .data$atlas_call,
      n_higher = "higher_than_healthy",
      n_lower = "lower_than_healthy",
      n_consistent = "consistent_with_healthy"
    ),
    atlas_call = factor(.data$atlas_call, levels = names(direction_cols))
  )

ggplot2::ggplot(
  count_long,
  ggplot2::aes(x = paper_compartment, y = n, fill = atlas_call)
) +
  ggplot2::geom_col(position = "stack") +
  ggplot2::facet_wrap(~Category, ncol = 1) +
  ggplot2::scale_fill_manual(values = direction_cols, name = "Atlas call") +
  ggplot2::labs(
    title = "Count of SAVI donors by atlas direction per compartment",
    subtitle = "Stacked bars: how many donors are higher / lower / consistent vs healthy null",
    x = NULL,
    y = "Number of donors"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 35, hjust = 1, size = 8),
    legend.position = "bottom"
  )
```

```{r}
#| label: plot-savi-paper-agreement
#| fig-width: 12
#| fig-height: 6
#| fig-align: center

agreement_cols <- c(
  "agrees (atlas vs paper)" = "#4DAF4A",
  "disagrees (atlas vs paper)" = "#FF7F00",
  "atlas consistent (no tail)" = "grey70"
)

ggplot2::ggplot(
  composition_savi_vs_healthy |>
    dplyr::mutate(
      Category = factor(.data$Category, levels = c("SAVI", "SAVI_treated")),
      agreement_label = factor(.data$agreement_label, levels = names(agreement_cols))
    ),
  ggplot2::aes(x = paper_compartment, fill = agreement_label)
) +
  ggplot2::geom_bar(position = "fill") +
  ggplot2::facet_wrap(~Category, ncol = 1) +
  ggplot2::scale_fill_manual(values = agreement_cols, name = "Agreement") +
  ggplot2::scale_y_continuous(labels = scales::percent_format()) +
  ggplot2::labs(
    title = "Atlas vs paper agreement (SAVI donors, unified compartments)",
    subtitle = paste(
      "Paper expects increased (naive CD4/CD8) or decreased (effector/memory, NK, MAIT, gdT).",
      "Green = atlas tail matches paper direction; orange = opposite tail."
    ),
    x = NULL,
    y = "Fraction of donors"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 35, hjust = 1, size = 8),
    legend.position = "bottom"
  )
```

```{r}
#| label: plot-savi-paper-agreement-rate
#| fig-width: 10
#| fig-height: 5
#| fig-align: center

ggplot2::ggplot(
  composition_savi_agreement_summary |>
    dplyr::mutate(
      Category = factor(.data$Category, levels = c("SAVI", "SAVI_treated")),
      paper_compartment = factor(
        .data$paper_compartment,
        levels = unique(paper_compartment_l3$paper_compartment)
      )
    ),
  ggplot2::aes(
    x = paper_compartment,
    y = pct_agrees_paper,
    fill = Category
  )
) +
  ggplot2::geom_col(position = ggplot2::position_dodge(width = 0.75), width = 0.65) +
  ggplot2::geom_hline(yintercept = 50, linetype = "dashed", color = "grey50") +
  ggplot2::scale_fill_manual(values = c(SAVI = "#E41A1C", SAVI_treated = "#377EB8")) +
  ggplot2::scale_y_continuous(limits = c(0, 100), expand = ggplot2::expansion(mult = c(0, 0.02))) +
  ggplot2::labs(
    title = "Share of SAVI donors where atlas tail agrees with paper",
    subtitle = "Among donors with a clear atlas tail (excludes 'consistent' only when testing agreement)",
    x = NULL,
    y = "% donors agreeing with paper direction"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 35, hjust = 1, size = 8))
```

### Compartment density vs matched healthy null

Same layout as the **CTRL** density grid (`plot-composition-ctrl-baseline-density`): one panel per **donor × paper compartment**. Grey density = healthy **`predict_res`** draws for that donor’s **`sex`** and **`age_decade`**; vertical line = observed proportion with tail colours and **`p_above_healthy`** / **`p_below_healthy`** labels. Below: **SAVI** and **SAVI_treated** donors only.

```{r}
#| label: plot-composition-compartment-density
#| fig-width: 20
#| fig-height: 18
#| fig-align: center

compartment_levels <- paper_compartment_l3 |>
  dplyr::distinct(.data$paper_compartment) |>
  dplyr::pull(.data$paper_compartment)

atlas_null_cols <- c(
  "consistent with healthy" = "#4DAF4A",
  "higher than healthy" = "#E41A1C",
  "lower than healthy" = "#377EB8"
)

if (!exists("baseline_compartment_draws")) {
  baseline_compartment_draws <- predict_res |>
    dplyr::inner_join(
      paper_compartment_l3 |>
        dplyr::distinct(.data$paper_compartment, .data$cell_type),
      by = "cell_type"
    ) |>
    dplyr::group_by(
      .data$paper_compartment,
      .data$age_decade,
      .data$sex,
      .data$.draw
    ) |>
    dplyr::summarise(
      compartment_proportion = sum(.data$proportion),
      .groups = "drop"
    )
}

savi_cohort_atlas <- composition_matched_reference |>
  dplyr::filter(.data$Category %in% c("SAVI", "SAVI_treated")) |>
  dplyr::mutate(
    baseline_agreement = dplyr::case_when(
      .data$direction_vs_healthy == "consistent_with_healthy" ~ "consistent with healthy",
      .data$direction_vs_healthy == "higher_than_healthy" ~ "higher than healthy",
      .data$direction_vs_healthy == "lower_than_healthy" ~ "lower than healthy"
    ),
    baseline_agreement = factor(
      .data$baseline_agreement,
      levels = names(atlas_null_cols)
    ),
    sample_panel = paste0(
      .data$sample_id_observed,
      " (",
      .data$Category,
      " · ",
      .data$sex,
      " | decade ",
      .data$age_decade,
      ")"
    )
  )

savi_panel_order <- savi_cohort_atlas |>
  dplyr::distinct(.data$sample_panel, .data$Category, .data$sample_id_observed) |>
  dplyr::mutate(
    Category = factor(.data$Category, levels = c("SAVI", "SAVI_treated"))
  ) |>
  dplyr::arrange(.data$Category, .data$sample_id_observed)

savi_panel_meta <- savi_cohort_atlas |>
  dplyr::distinct(
    .data$sample_id_observed,
    .data$Category,
    .data$sex,
    .data$age_decade,
    .data$paper_compartment,
    .data$sample_panel
  )

savi_density_draws <- savi_panel_meta |>
  dplyr::inner_join(
    baseline_compartment_draws,
    by = c("sex", "age_decade", "paper_compartment")
  ) |>
  dplyr::mutate(
    sample_panel = factor(
      .data$sample_panel,
      levels = savi_panel_order$sample_panel
    ),
    paper_compartment = factor(.data$paper_compartment, levels = compartment_levels)
  ) |>
  dplyr::filter(
    is.finite(.data$compartment_proportion),
    .data$compartment_proportion >= 0
  )

savi_density_vlines <- savi_cohort_atlas |>
  dplyr::mutate(
    sample_panel = factor(
      .data$sample_panel,
      levels = savi_panel_order$sample_panel
    ),
    paper_compartment = factor(.data$paper_compartment, levels = compartment_levels),
    p_label = sprintf(
      "p_above=%.2f\np_below=%.2f",
      .data$p_above_healthy,
      .data$p_below_healthy
    )
  ) |>
  dplyr::filter(is.finite(.data$proportion_observed)) |>
  dplyr::left_join(
    savi_density_draws |>
      dplyr::group_by(.data$sample_panel, .data$paper_compartment) |>
      dplyr::summarise(
        x_mid = stats::median(.data$compartment_proportion, na.rm = TRUE),
        .groups = "drop"
      ),
    by = c("sample_panel", "paper_compartment")
  ) |>
  dplyr::mutate(
    text_hjust = dplyr::if_else(
      .data$proportion_observed > .data$x_mid,
      1.05,
      -0.05
    )
  )

ggplot2::ggplot(
  savi_density_draws,
  ggplot2::aes(x = compartment_proportion)
) +
  ggplot2::geom_density(
    fill = "grey80",
    color = "grey35",
    alpha = 0.55,
    linewidth = 0.35,
    na.rm = TRUE
  ) +
  ggplot2::geom_vline(
    data = savi_density_vlines,
    ggplot2::aes(
      xintercept = proportion_observed,
      color = baseline_agreement
    ),
    inherit.aes = FALSE,
    linewidth = 0.75
  ) +
  ggplot2::geom_text(
    data = savi_density_vlines,
    ggplot2::aes(
      x = proportion_observed,
      y = Inf,
      label = p_label,
      hjust = text_hjust
    ),
    inherit.aes = FALSE,
    vjust = 1.05,
    size = 2.1,
    lineheight = 0.85,
    color = "grey15"
  ) +
  ggplot2::facet_grid(
    sample_panel ~ paper_compartment,
    scales = "free",
    labeller = ggplot2::labeller(
      .cols = ggplot2::label_wrap_gen(10),
      .rows = ggplot2::label_wrap_gen(16)
    )
  ) +
  ggplot2::scale_color_manual(
    values = atlas_null_cols,
    name = "Donor vs\nmatched null"
  ) +
  ggplot2::scale_x_continuous(expand = ggplot2::expansion(mult = c(0.04, 0.06))) +
  ggplot2::labs(
    title = "SAVI and SAVI_treated donors vs sex/age-matched healthy atlas (density grid)",
    subtitle = paste(
      "Rows = donor (Category · sex | age decade); columns = paper compartments.",
      "Grey = healthy predictive draws for that donor's metadata;",
      "vertical line = observed proportion;",
      "labels = p_above_healthy / p_below_healthy."
    ),
    x = "Compartment proportion",
    y = "Density"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    legend.position = "bottom",
    strip.text.x = ggplot2::element_text(size = 7, face = "bold"),
    strip.text.y = ggplot2::element_text(size = 7),
    panel.spacing = grid::unit(0.25, "lines")
  )
```

**How to read vs the paper**

1. **Observed cohort trends** (boxplot / bar chart): within-study CTRL vs SAVI vs SAVI_treated using observed proportions only (paper Figure 1E style). **`composition_paper_summary`**: `delta_savi_minus_ctrl` and `matches_paper_savi` flag agreement with published SAVI vs control direction.
2. **Healthy baseline by age** (line chart): how female vs male compartment proportions change across age decades 1–4 in **`predict_res`** (context for sex/age matching).
3. **CTRL vs atlas** (`plot-composition-ctrl-baseline-heatmap`, `plot-composition-ctrl-baseline-density`): sample × compartment heatmap plus per-donor density grid — grey curve = sex/age-matched healthy draws; line = observed CTRL proportion.
4. **Healthy atlas comparison** (heatmap, count/agreement bars): per donor × compartment tail test vs sex/age-matched **`predict_res`** draws (`composition_matched_reference`, `composition_savi_vs_healthy`).
5. **Compartment density plot** (`plot-composition-compartment-density`): same donor × compartment grid as CTRL — matched healthy density, tail-coloured line, **p_above** / **p_below** labels (**SAVI** and **SAVI_treated**).
6. **Fine-grained L3 table** (`composition_vs_healthy`): per donor × L3, `direction` and `empirical_confidence` with the same tail convention as gene-level deviation tests.


