---
title: "posteriorHCA Figure 2 — SAVI manuscript figure"
subtitle: "External population-reference audit of a published STING-activation signature"
date: today
date-format: "DD MMM YYYY"
title-block-banner: true
author:
  - name: Chen Zhan
    orcid: 0000-0002-4794-8339
    email: chen.zhan@adelaide.edu.au
    affiliation:
      - id: sgc
        name: SAiGENCI, The University of Adelaide
format:
  html:
    theme: minty
    df-print: paged
    code-line-numbers: true
    embed-resources: true
    code-fold: true
    code-fold-default: true
    toc: true
    toc-location: left
    toc-title: "Contents"
knitr:
  opts_chunk:
    message: false
    warning: false
    fig.retina: 2
editor: source
abstract: |
  Dedicated manuscript-figure report for the SAVI case (Nature Methods
  double-column canvas, max 183 mm × 220 mm; no panel titles/subtitles).
  Panel a freezes the observed cohort design and regenerates healthy-blood
  support from the production `V1_monocytic` model frame. Panel b tests
  published composition directions with a patient-level 15-compartment
  `sccomp` model, calibrates equal-average and metadata-standardised HCA
  references against study controls, and gates external-reference inference.
  Panel c audits the 16-gene covered STING signature with subject-coupled NB
  log-μ contrasts and an HCA-standardised reference. Panel f tests
  subject-deletion influence on the two main signature estimands. Panels d–e
  remain scaffolded until their analyses pass gates. The Nature blueprint PDF is
  layout reference only; no schematic numbers are treated as results.
params:
  case_dir: "/home/a1237163/lab/chen/posteriorHCA/dev/SAVI"
  fig_dir: "/home/a1237163/lab/chen/posteriorHCA/dev/SAVI/figures"
  data_dir: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/dev/SAVI/data"
  model_store_root: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/HPC_posterior/NEW_CELL_TYPE"
  target_cell_type: "monocytic"
  fig_width_mm: 183
  fig_height_mm: 220
  panel_a_height_mm: 58
  panel_b_height_mm: 70
  run_sccomp_panel_b: false
  run_hca_panel_b: false
  panel_b_prediction_draws: 4000
  panel_b_pair_draws: 20000
  panel_b_rope_pp: 1
  run_panel_c_rebuild: false
  savi_seurat_gz: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/dev/SAVI/data/GSE226598_SeuratObject_SAVI.rds.gz"
  savi_pseudobulk_rds: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/dev/SAVI/data/GSE226598_SAVI_pseudobulk_Sample_CellType.rds"
  estimates_rds: "/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/posteriorHCA/dev/SAVI/data/SAVI_estimates_chunk.rds"
  pseudobulk_h5ad: "/home/a1237163/lab/Mangiola_ImmuneAtlas/zenodo/pseudobulk_se.h5ad"
  sccomp_estimates_rds: "/home/a1237163/lab/Mangiola_ImmuneAtlas/taskforce_shared_folder/sccomp_on_cellNexus_1_0_12/estimates_age_decade___L3___disease_TRUE___immune_only_TRUE.rds"
---

# Scope

**Defensible claim (Fig. 2).** posteriorHCA adds an external population-scale
healthy reference to a small, internally controlled rare-disease study. After
matching target and atlas cell populations and treating biological subjects as
inferential units, the comprehensive query tests whether the published
STING-activation signature (and covered constituents) remains shifted relative
to a broader healthy mean; the metadata-conditioned query tests whether
demographic standardisation resolves reference-domain shift before an external
contrast is interpreted as disease associated.

**This report.** Build the manuscript multipanel from audited tables. **Panels
a–c and f are complete in this draft.** Panel c uses subject-coupled NB cohort
locations and evaluates HCA contrasts after study-control calibration. Panel f
tests whether those signature conclusions depend on individual subjects.
Panels d–e remain placeholders until their analyses are finished. Helper
functions and `theme_multipanel` live in this file (no external theme source
required at render time).

**Colour / mark grammar (locked).**

| Element | Legend (full) | Plot text (short) | Encoding |
|---|---|---|---|
| Study healthy control | Control | CTRL | green `#009E73`, circle |
| Pre-therapy SAVI | Untreated SAVI | Untreated | vermilion `#D55E00`, diamond |
| Post–JAK-inhibitor SAVI | JAK-inhibitor treated | Treated | purple `#7A5195`, square |
| Sex | female / male | — | `RColorBrewer` PRGn (`colmap_sex`) |
| Comprehensive HCA | — | — | charcoal grey |
| Metadata-conditioned HCA | — | — | teal |

Figure panels follow Nature Methods style: axis labels and legends only
(no plot titles or subtitles); canvas ≤ 183 mm wide × 220 mm high.

------------------------------------------------------------------------

# Setup

```{r}
#| label: setup

suppressPackageStartupMessages({
  library(tibble)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(ggplot2)
  library(patchwork)
  library(targets)
  library(SummarizedExperiment)
  library(S4Vectors)
  library(scales)
  library(RColorBrewer)
  library(sccomp)
})

# Use headless-safe raster devices for Quarto/knitr figure previews on HPC.
options(bitmapType = "cairo")

dir.create(params$fig_dir, recursive = TRUE, showWarnings = FALSE)

model_path <- file.path(
  params$model_store_root,
  paste0("V1_", params$target_cell_type)
)

# Nature Methods double-column: max 183 mm wide × 220 mm high.
mm_to_in <- function(mm) mm / 25.4
fig_w_in <- mm_to_in(params$fig_width_mm)
fig_h_in <- mm_to_in(params$fig_height_mm)
panel_a_h_in <- mm_to_in(params$panel_a_height_mm)
panel_b_h_in <- mm_to_in(params$panel_b_height_mm)

panel_b_counts_rds <- file.path(params$data_dir, "SAVI_panel_b_celltype_counts.rds")
panel_b_fit_rds <- file.path(
  params$data_dir,
  "SAVI_panel_b_sccomp_shared15_untreated_control_hmc.rds"
)
panel_b_hca_predict_rds <- file.path(
  params$data_dir,
  "SAVI_panel_b_hca_predict_donor_marginal_shared15_v3.rds"
)
panel_c_effects_rds <- file.path(params$data_dir, "SAVI_panel_c_effects.rds")
panel_c_sample_logmu_rds <- file.path(
  params$data_dir,
  "SAVI_panel_c_sample_logmu.rds"
)
panel_c_hca_logmu_rds <- file.path(
  params$data_dir,
  "SAVI_panel_c_hca_logmu_draws.rds"
)
panel_c_se_rds <- file.path(params$data_dir, "SAVI_panel_c_se_monocytic.rds")
panel_c_gene_order <- c(
  "ADRB2", "PTGS2", "AQP9", "OTUD1", "PTX3", "TGIF1", "TSC22D2", "ZBTB43",
  "ZFP36L2", "JARID2", "KLF4", "TREM1", "ANKRD28", "ANXA1", "PLAUR", "IL1B"
)
panel_c_contrast_levels <- c(
  "untreated - study control",
  "untreated - HCA",
  "paired untreated - treated",
  "study control - HCA"
)
panel_c_contrast_labs <- c(
  "untreated - study control" = "Untreated - CTRL",
  "untreated - HCA" = "Untreated - HCA",
  "paired untreated - treated" = "Untreated - treated",
  "study control - HCA" = "CTRL - HCA"
)
panel_f_effects_rds <- file.path(params$data_dir, "SAVI_panel_f_subject_influence.rds")

# Figure grammar colours.
col_ctrl <- "#009E73"
col_untreated <- "#D55E00"
col_treated <- "#7A5195"
col_hca <- "#8C8C8C"
col_meta <- "#2A9D8F"
col_dark <- "#3F3F3F"
col_mid <- "#777777"
col_light <- "#D9D9D9"

colmap_sex <- setNames(
  RColorBrewer::brewer.pal(3, "PRGn")[c(1, 3)],
  c("female", "male")
)

state_levels <- c("control", "untreated", "treated")
# Legend: full names. In-plot annotations: short forms (CTRL / Untreated / Treated).
state_labels <- c(
  control = "Control",
  untreated = "Untreated SAVI",
  treated = "JAK-inhibitor treated"
)
state_labels_short <- c(
  control = "CTRL",
  untreated = "Untreated",
  treated = "Treated"
)
state_colours <- c(
  control = col_ctrl,
  untreated = col_untreated,
  treated = col_treated
)
# shape 21 circle, 23 diamond, 22 square; fill = sex (PRGn).
state_shapes <- c(control = 21, untreated = 23, treated = 22)

theme_multipanel <-
  theme_bw() +
  theme(
    panel.border = element_blank(),
    axis.line = element_line(linewidth = 0.1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    strip.background = element_blank(),
    axis.title.y = element_text(margin = margin(0, 0, 0, 0), size = 7),
    axis.title.x = element_text(margin = margin(0, 0, 0, 0), size = 7),
    panel.spacing.x = unit(0.1, "lines"),
    axis.text.x = element_text(size = 6),
    axis.text.y = element_text(size = 6),
    strip.text.x = element_text(size = 7),
    strip.text.y = element_text(size = 7),
    legend.key.size = unit(4, "mm"),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 6),
    strip.clip = "off",
    # Nature Methods: caption carries titles; panels stay label-free.
    plot.title = element_blank(),
    plot.subtitle = element_blank(),
    plot.tag = element_text(size = 8, face = "bold"),
    axis.line.x = element_line(linewidth = 0.2),
    axis.line.y = element_line(linewidth = 0.2),
    axis.ticks.x = element_line(linewidth = 0.2),
    axis.ticks.y = element_line(linewidth = 0.2),
    plot.margin = margin(2, 4, 2, 2)
  )

save_panel_pdf <- function(plot, filename, width_mm, height_mm) {
  path <- file.path(params$fig_dir, filename)
  ggsave(
    filename = path,
    plot = plot,
    width = width_mm,
    height = height_mm,
    units = "mm",
    dpi = 600,
    device = cairo_pdf
  )
  path
}
```

------------------------------------------------------------------------

# Panel a — Observed cohort design and HCA training support

**Purpose.** Factual evidence of what posteriorHCA can condition on: source
library ages, sequencing batches / chemistry, repeated measures, family links,
and unique healthy-blood donor age support in the frozen `V1_monocytic` frame.

**Rules enforced here.**

- One point = one library; subjects (not cells) are biological units.
- C1 / C2 are two **healthy study controls** whose GEO age is only labelled
  `Adult` (no year). They stay in a categorical `Adult; exact age unavailable`
  box — no imputed age on the axis.
- P5 uses **10.5 years** (source Fig. 1a), not the legacy eight-year imputation.
- HCA counts are regenerated from the production model frame (not copied from
  the blueprint).

## Freeze the 17-library source metadata table

Ages, sex and STING batch come from GEO GSE226598. Subject / family / treatment
annotations follow de Cevins *et al.* (2023) Fig. 1a and the case-study
specification. Chemistry: STING1 = 10x 3′ v2; STING2 / STING3 = 10x 3′ v3.

```{r}
#| label: panel-a-savi-libraries

savi_libraries <- tibble::tribble(
  ~library_id,          ~subject_id, ~state,      ~sex,     ~age_years, ~age_status, ~sting_batch, ~chemistry,   ~family_role,              ~tx_label,
  "C7",                 "C7",        "control",   "female",         5, "exact",     "STING1",     "10x 3′ v2",  NA_character_,             NA_character_,
  "C8_aunt_P1_STING",   "C8",        "control",   "female",        36, "exact",     "STING1",     "10x 3′ v2",  "paternal aunt of P1",     NA_character_,
  "C9_mother_P1_STING", "C9",        "control",   "female",        35, "exact",     "STING1",     "10x 3′ v2",  "mother of P1",            NA_character_,
  "P1_STING_ht",        "P1",        "untreated", "female",         4, "exact",     "STING1",     "10x 3′ v2",  "proband",                 NA_character_,
  "P1_STING_ht_T",      "P1",        "treated",   "female",         8, "exact",     "STING1",     "10x 3′ v2",  "proband",                 "R52",
  "P1_STING_ht_T2",     "P1",        "treated",   "female",         8, "exact",     "STING1",     "10x 3′ v2",  "proband",                 "T2",
  "P2_STING_ht",        "P2",        "untreated", "male",          31, "exact",     "STING1",     "10x 3′ v2",  "uncle of P1",             NA_character_,
  "P2_STING_ht_T",      "P2",        "treated",   "male",          34, "exact",     "STING1",     "10x 3′ v2",  "uncle of P1",             "R1",
  "C10",                "C10",       "control",   "male",          10, "exact",     "STING2",     "10x 3′ v3",  NA_character_,             NA_character_,
  "C11",                "C11",       "control",   "male",          16, "exact",     "STING2",     "10x 3′ v3",  NA_character_,             NA_character_,
  "P4_STING_ht",        "P4",        "untreated", "male",           7, "exact",     "STING2",     "10x 3′ v3",  NA_character_,             NA_character_,
  "P4_STING_ht_T",      "P4",        "treated",   "male",           9, "exact",     "STING2",     "10x 3′ v3",  NA_character_,             "R13",
  "P6_STING_ht",        "P6",        "untreated", "male",           8, "exact",     "STING2",     "10x 3′ v3",  NA_character_,             NA_character_,
  "P6_STING_ht_T",      "P6",        "treated",   "male",          10, "exact",     "STING2",     "10x 3′ v3",  NA_character_,             "R13",
  "C1",                 "C1",        "control",   "male",          NA, "adult_unavailable", "STING3", "10x 3′ v3", NA_character_,          NA_character_,
  "C2",                 "C2",        "control",   "male",          NA, "adult_unavailable", "STING3", "10x 3′ v3", NA_character_,          NA_character_,
  "P5_STING",           "P5",        "untreated", "male",        10.5, "source_fig1a", "STING3",  "10x 3′ v3",  NA_character_,             NA_character_
) |>
  dplyr::mutate(
    state = factor(state, levels = state_levels),
    sex = factor(sex, levels = c("female", "male")),
    batch_row = factor(
      paste0(sting_batch, " · ", dplyr::recode(chemistry, "10x 3′ v2" = "v2", "10x 3′ v3" = "v3")),
      levels = c("STING1 · v2", "STING2 · v3", "STING3 · v3")
    ),
    display_id = dplyr::case_when(
      subject_id == "P5" ~ "P5",
      !is.na(tx_label) ~ tx_label,
      state == "untreated" ~ subject_id,
      TRUE ~ NA_character_
    )
  )

stopifnot(nrow(savi_libraries) == 17L)
stopifnot(dplyr::n_distinct(savi_libraries$subject_id) == 12L)

savi_libraries |>
  dplyr::select(
    library_id, subject_id, state, sex, age_years, age_status,
    sting_batch, chemistry, family_role, tx_label
  )
```

```{r}
#| label: panel-a-lane-offsets
# Vertical jitter within a batch row so co-located libraries remain readable.
# Label offsets are hand-tuned: ggrepel / fixed nudges fail on the STING2 7–10
# cluster and the P1 R52/T2 same-age stack.

assign_lanes <- function(df, lane_step = 0.32) {
  df |>
    dplyr::group_by(batch_row, age_years, age_status) |>
    dplyr::arrange(state, subject_id, library_id, .by_group = TRUE) |>
    dplyr::mutate(
      n_here = dplyr::n(),
      lane = if (dplyr::n() == 1L) {
        0
      } else {
        (dplyr::row_number() - (dplyr::n() + 1) / 2) * lane_step
      },
      y_plot = as.numeric(batch_row) + lane
    ) |>
    dplyr::ungroup()
}

# Explicit label anchors (years on x, batch-lane units on y).
label_offset_tbl <- tibble::tribble(
  ~library_id,       ~label_dx, ~label_dy,
  "P1_STING_ht",           0.0,      0.40, # P1
  "P1_STING_ht_T",        -2.0,      0.48, # R52 left / above
  "P1_STING_ht_T2",        2.0,     -0.48, # T2 right / below
  "P2_STING_ht",           0.0,      0.40, # P2
  "P2_STING_ht_T",         0.0,      0.40, # R1
  "P4_STING_ht",          -0.9,      0.48, # P4 above-left
  "P6_STING_ht",          -0.9,     -0.48, # P6 below-left
  "P4_STING_ht_T",         0.9,      0.48, # R13 (P4) above-right
  "P6_STING_ht_T",         0.9,     -0.48, # R13 (P6) below-right
  "P5_STING",              0.0,      0.40  # P5
)

savi_exact <- savi_libraries |>
  dplyr::filter(age_status != "adult_unavailable") |>
  assign_lanes() |>
  dplyr::left_join(label_offset_tbl, by = "library_id") |>
  dplyr::mutate(
    label_dx = dplyr::coalesce(label_dx, 0),
    label_dy = dplyr::coalesce(label_dy, 0.40)
  )

savi_adult <- savi_libraries |>
  dplyr::filter(age_status == "adult_unavailable")

savi_connectors <- savi_exact |>
  dplyr::group_by(subject_id) |>
  dplyr::filter(dplyr::n() > 1L) |>
  dplyr::arrange(age_years, state, library_id, .by_group = TRUE) |>
  dplyr::ungroup()
```

## Regenerate frozen `V1_monocytic` healthy-blood support

No-refit audit from the production `pseudobulk_sample` target: healthy blood
(`tissue_groups == "blood"`, `disease_groups___altered == "Normal"`), then
deduplicate by `donor_id`.

```{r}
#| label: panel-a-hca-support

pseudobulk_sample <- targets::tar_read(
  pseudobulk_sample,
  store = file.path(model_path, "_targets")
)

hca_cd <- SummarizedExperiment::colData(pseudobulk_sample) |>
  as.data.frame() |>
  tibble::as_tibble()

hca_healthy_blood <- hca_cd |>
  dplyr::filter(
    .data$tissue_groups == "blood",
    as.character(.data$disease_groups___altered) == "Normal"
  ) |>
  dplyr::mutate(
    age_years = as.numeric(.data$age_years),
    sex = as.character(.data$sex),
    assay_groups_altered = as.character(.data$assay_groups___altered),
    dataset_id_altered = as.character(.data$dataset_id___altered)
  )

stopifnot(nrow(hca_healthy_blood) > 0L)

hca_donors <- hca_healthy_blood |>
  dplyr::group_by(.data$donor_id) |>
  dplyr::summarise(
    sex = dplyr::first(stats::na.omit(.data$sex)),
    age_years = stats::median(.data$age_years, na.rm = TRUE),
    age_decade = dplyr::first(stats::na.omit(as.character(.data$age_decade))),
    n_samples = dplyr::n(),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    sex = factor(.data$sex, levels = c("female", "male")),
    age_bin = cut(
      .data$age_years,
      breaks = seq(0, 90, by = 10),
      right = FALSE,
      include.lowest = TRUE
    )
  )

hca_support_summary <- tibble::tibble(
  n_donors = nrow(hca_donors),
  n_samples = nrow(hca_healthy_blood),
  n_datasets = dplyr::n_distinct(hca_healthy_blood$dataset_id_altered),
  n_assay_groups = dplyr::n_distinct(hca_healthy_blood$assay_groups_altered),
  n_samples_10x3 = sum(
    hca_healthy_blood$assay_groups_altered == "10x Genomics 3",
    na.rm = TRUE
  ),
  age_min = min(hca_donors$age_years, na.rm = TRUE),
  age_max = max(hca_donors$age_years, na.rm = TRUE)
)

# Provisional overlap audit against GSE226598 accessions / SAVI citation strings.
overlap_hits <- hca_healthy_blood |>
  dplyr::mutate(
    haystack = paste(
      dplyr::coalesce(as.character(.data$sample_id), ""),
      dplyr::coalesce(as.character(.data$dataset_id), ""),
      dplyr::coalesce(as.character(.data$title), ""),
      dplyr::coalesce(as.character(.data$citation), ""),
      dplyr::coalesce(as.character(.data$url), ""),
      sep = " "
    )
  ) |>
  dplyr::filter(
    stringr::str_detect(
      .data$haystack,
      stringr::regex("GSE226598|GSM70799|de Cevins|SAVI", ignore_case = TRUE)
    )
  )

overlap_status <- if (nrow(overlap_hits) == 0L) {
  "absent (no GSE226598 / SAVI string match in current healthy-blood frame)"
} else {
  sprintf("observed (%d rows matched accession/citation strings)", nrow(overlap_hits))
}

hca_support_summary
```

```{r}
#| label: panel-a-support-flags

# Per exact-age SAVI library: is the collection age inside observed HCA donor ages?
age_support <- range(hca_donors$age_years, na.rm = TRUE)

savi_query_support <- savi_libraries |>
  dplyr::mutate(
    support_flag = dplyr::case_when(
      age_status == "adult_unavailable" ~ "marginalised",
      is.na(age_years) ~ "marginalised",
      age_years >= age_support[1] & age_years <= age_support[2] ~ "observed",
      TRUE ~ "outside support"
    ),
    assay_support = dplyr::if_else(
      hca_support_summary$n_samples_10x3 > 0L,
      "observed (harmonised 10x Genomics 3)",
      "outside support"
    )
  )

savi_query_support |>
  dplyr::count(support_flag, assay_support)
```

## Draw panel A

```{r}
#| label: panel-a-helpers

plot_panel_a_collections <- function(exact_df, connector_df, adult_df) {
  y_labs <- levels(exact_df$batch_row)

  p_main <- ggplot2::ggplot() +
    ggplot2::geom_hline(
      yintercept = seq_along(y_labs),
      colour = col_light,
      linewidth = 0.25
    ) +
    ggplot2::geom_path(
      data = connector_df,
      ggplot2::aes(x = age_years, y = y_plot, group = subject_id),
      colour = col_mid,
      linewidth = 0.35,
      alpha = 0.85
    ) +
    ggplot2::geom_point(
      data = exact_df,
      ggplot2::aes(
        x = age_years,
        y = y_plot,
        shape = state,
        colour = state,
        fill = sex
      ),
      size = 1.7,
      stroke = 0.45
    ) +
    ggplot2::geom_segment(
      data = exact_df |> dplyr::filter(!is.na(display_id)),
      ggplot2::aes(
        x = age_years,
        xend = age_years + label_dx,
        y = y_plot,
        yend = y_plot + label_dy
      ),
      colour = col_mid,
      linewidth = 0.2,
      alpha = 0.7
    ) +
    ggplot2::geom_text(
      data = exact_df |> dplyr::filter(!is.na(display_id)),
      ggplot2::aes(
        x = age_years + label_dx,
        y = y_plot + label_dy,
        label = display_id
      ),
      size = 2.5,
      colour = col_dark,
      vjust = 0.5,
      hjust = 0.5
    ) +
    ggplot2::scale_shape_manual(
      values = state_shapes,
      labels = state_labels,
      name = NULL,
      guide = ggplot2::guide_legend(order = 1, override.aes = list(fill = "white"))
    ) +
    ggplot2::scale_colour_manual(
      values = state_colours,
      labels = state_labels,
      name = NULL,
      guide = ggplot2::guide_legend(order = 1)
    ) +
    ggplot2::scale_fill_manual(
      values = colmap_sex,
      breaks = c("female", "male"),
      labels = c("female", "male"),
      name = NULL,
      guide = ggplot2::guide_legend(
        order = 2,
        override.aes = list(shape = 21, colour = col_dark)
      )
    ) +
    ggplot2::scale_y_continuous(
      breaks = seq_along(y_labs),
      labels = y_labs,
      limits = c(0.25, length(y_labs) + 0.75)
    ) +
    ggplot2::scale_x_continuous(
      limits = c(0, 40),
      breaks = seq(0, 40, by = 10),
      expand = ggplot2::expansion(mult = c(0.01, 0.02))
    ) +
    ggplot2::coord_cartesian(clip = "off") +
    ggplot2::labs(x = "Age at collection (years)", y = NULL) +
    theme_multipanel +
    ggplot2::theme(
      legend.position = "bottom",
      legend.box = "horizontal",
      legend.margin = margin(0, 0, 0, 0),
      axis.line.y = element_blank(),
      axis.ticks.y = element_blank()
    )

  # Separate non-numeric Adult category (C1/C2); do not invent an age interval.
  p_adult <- ggplot2::ggplot() +
    ggplot2::annotate(
      "text",
      x = 0.5, y = 3.15,
      label = "C7/8/9 · P1/2",
      size = 1.7,
      colour = col_mid
    ) +
    ggplot2::annotate(
      "text",
      x = 0.5, y = 2.15,
      label = "C10/11 · P4/6",
      size = 1.7,
      colour = col_mid
    ) +
    ggplot2::annotate(
      "rect",
      xmin = 0.06, xmax = 0.94, ymin = 0.85, ymax = 1.55,
      fill = "white",
      colour = col_ctrl,
      linewidth = 0.4
    ) +
    ggplot2::annotate(
      "text",
      x = 0.5, y = 1.32,
      label = "C1/C2 · Adult",
      size = 2.1,
      fontface = "bold",
      colour = col_dark
    ) +
    ggplot2::annotate(
      "text",
      x = 0.5, y = 1.05,
      label = "exact age unavailable",
      size = 1.8,
      colour = col_mid
    ) +
    ggplot2::scale_x_continuous(limits = c(0, 1), expand = c(0, 0)) +
    ggplot2::scale_y_continuous(limits = c(0.55, 3.55)) +
    ggplot2::labs(x = NULL, y = NULL) +
    ggplot2::theme_void() +
    ggplot2::theme(plot.margin = margin(8, 4, 18, 2))

  # p_main + p_adult + patchwork::plot_layout(widths = c(4.2, 1.15))
  p_main
}

plot_panel_a_hca_continuous <- function(donors, support_summary, savi_marks, overlap_status) {
  brks <- seq(0, 90, by = 10)

  hist_counts <- donors |>
    dplyr::mutate(
      bin_i = pmin(
        findInterval(.data$age_years, brks, all.inside = TRUE),
        length(brks) - 1L
      )
    ) |>
    dplyr::count(bin_i, sex, name = "n") |>
    tidyr::complete(
      bin_i = seq_len(length(brks) - 1L),
      sex = factor(c("female", "male"), levels = c("female", "male")),
      fill = list(n = 0L)
    ) |>
    dplyr::left_join(
      tibble::tibble(
        bin_i = seq_len(length(brks) - 1L),
        xmid = (brks[-length(brks)] + brks[-1]) / 2
      ),
      by = "bin_i"
    ) |>
    dplyr::mutate(sex = factor(sex, levels = c("male", "female")))

  age_marks <- savi_marks |>
    dplyr::distinct(age_years, sex) |>
    dplyr::mutate(sex = factor(sex, levels = c("female", "male")))

  ggplot2::ggplot(hist_counts, ggplot2::aes(x = xmid, y = n, fill = sex)) +
    ggplot2::geom_col(width = 9.2, colour = NA) +
    ggplot2::geom_vline(
      data = age_marks,
      ggplot2::aes(xintercept = age_years, colour = sex),
      linewidth = 0.35,
      alpha = 0.95
    ) +
    ggplot2::scale_fill_manual(
      values = colmap_sex,
      breaks = c("female", "male"),
      labels = c("female", "male"),
      name = NULL
    ) +
    ggplot2::scale_colour_manual(
      values = colmap_sex,
      breaks = c("female", "male"),
      guide = "none"
    ) +
    ggplot2::scale_x_continuous(
      limits = c(0, 90),
      breaks = c(0, 30, 60, 90),
      expand = c(0, 0)
    ) +
    ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0, 0.08))) +
    ggplot2::labs(
      x = "Unique donor age (years)",
      y = NULL
    ) +
    theme_multipanel +
    ggplot2::theme(
      legend.position = "bottom",
      legend.margin = margin(0, 0, 0, 0)
    )
}
```
```{r}
#| label: panel-a-draw
#| fig-width: 7.2
#| fig-height: 2.3
#| fig-cap: "Figure 2a. Observed SAVI collections (left) and frozen V1_monocytic healthy-blood donor-age support (right)."

savi_age_marks <- savi_exact |>
  dplyr::distinct(age_years, sex)

p_a_left <- plot_panel_a_collections(savi_exact, savi_connectors, savi_adult)
p_a_right <- plot_panel_a_hca_continuous(
  donors = hca_donors,
  support_summary = hca_support_summary,
  savi_marks = savi_age_marks,
  overlap_status = overlap_status
)

panel_a <- (p_a_left | p_a_right) +
  patchwork::plot_layout(widths = c(1.35, 1))

panel_a
```
```{r}
#| label: panel-a-audit-text

cat(
  sprintf(
    paste(
      "Panel a audit: %d libraries from %d subjects;",
      "HCA healthy blood = %d donors / %d samples / %d datasets / %d assay groups;",
      "%d samples in harmonised `10x Genomics 3`;",
      "donor age range %.1f–%.1f years;",
      "training-overlap status: %s.",
      "Support flags — observed: %d; marginalised (C1/C2 Adult): %d; outside: %d.\n"
    ),
    nrow(savi_libraries),
    dplyr::n_distinct(savi_libraries$subject_id),
    hca_support_summary$n_donors,
    hca_support_summary$n_samples,
    hca_support_summary$n_datasets,
    hca_support_summary$n_assay_groups,
    hca_support_summary$n_samples_10x3,
    hca_support_summary$age_min,
    hca_support_summary$age_max,
    overlap_status,
    sum(savi_query_support$support_flag == "observed"),
    sum(savi_query_support$support_flag == "marginalised"),
    sum(savi_query_support$support_flag == "outside support")
  )
)
```

------------------------------------------------------------------------

# Panel b — Composition directions

Panel b asks whether seven published PBMC composition directions (de Cevins
*et al.* Fig. 1E) are recovered relative to study controls and two HCA healthy
references. Calibration of the healthy references is reported before the
disease contrasts so that HCA differences are not interpreted as disease
effects when study controls already reveal reference-domain shift.

## B1. Finest defensible SAVI–HCA harmonisation

The 23 SAVI labels and 26 HCA L3 labels are mapped to the same 15-part
composition. Seven prespecified T/NK compartments form the testing family;
seven matched B/myeloid/DC compartments retain nuisance-lineage information;
only genuinely unmatched or ambiguous populations enter `residual`.

```{r}
#| label: panel-b-harmonisation

# The seven primary compartments follow the order in the source figure.
display_compartments <- c(
  "Naive CD4 T",
  "Eff./memory CD4",
  "Naive CD8 T",
  "Eff./memory CD8",
  "NK",
  "MAIT",
  "gamma-delta T"
)

# Additional matched lineages remain separate in the likelihood. This avoids
# asking one heterogeneous residual component to absorb B, myeloid and DC mass.
compartment_levels <- c(
  display_compartments,
  "Naive/transitional B",
  "Memory B",
  "Plasma cells",
  "Classical monocytes",
  "Non-classical monocytes",
  "cDC",
  "pDC",
  "residual"
)

# Every SAVI label is assigned explicitly; no new label can silently enter the
# residual category without failing the coverage check in the next chunk.
celltype_to_compartment <- tibble::tribble(
  ~cell_type,              ~compartment,
  "0. Naive CD4",          "Naive CD4 T",
  "3. CM CD4",             "Eff./memory CD4",
  "4. EM CD4",             "Eff./memory CD4",
  "8. TE CD4",             "Eff./memory CD4",
  "9. Tregs and TE CD4",   "Eff./memory CD4",
  "2. Naive CD8",          "Naive CD8 T",
  "14. CM CD8",            "Eff./memory CD8",
  "15. EM CD8",            "Eff./memory CD8",
  "7. TE CD8",             "Eff./memory CD8",
  "10. NK",                "NK",
  "26. NK CD56bright",     "NK",
  "22. MAIT",              "MAIT",
  "16. gd T cells",        "gamma-delta T",
  "1. Naive B",            "Naive/transitional B",
  "6. Transitionnal B",    "Naive/transitional B",
  "13. Memory B",          "Memory B",
  "25. Plasmablasts",      "Plasma cells",
  "5. Classical monocytes", "Classical monocytes",
  "12. Non-classical monocytes", "Non-classical monocytes",
  "20. cDCs",              "cDC",
  "27. pDCs",              "pDC",
  "11. CD3+ B cells",      "residual",
  "17. Disease-associated monocytes", "residual"
)

# Source-reported direction (context arrows only).
source_direction <- tibble::tibble(
  compartment = factor(display_compartments, levels = display_compartments),
  source_arrow = c("+", "-", "+", "-", "-", "-", "-")
)

# Library metadata reused from panel a (group + sequencing batch).
panel_b_sample_meta <- savi_libraries |>
  dplyr::transmute(
    sample_id = library_id,
    subject_id,
    group = dplyr::recode(
      as.character(state),
      control = "control",
      untreated = "untreated",
      treated = "treated"
    ),
    sequencing_batch = sting_batch
  ) |>
  dplyr::mutate(
    group = factor(group, levels = c("control", "untreated", "treated")),
    sequencing_batch = factor(
      sequencing_batch,
      levels = c("STING1", "STING2", "STING3")
    )
  )

# All 26 HCA L3 types are also assigned explicitly. L3 types without a SAVI
# counterpart remain in `residual` so posterior compositions retain unit mass.
l3_to_compartment <- tibble::tribble(
  ~cell_type,         ~compartment,
  "b memory",         "Memory B",
  "b naive",          "Naive/transitional B",
  "cd14 mono",        "Classical monocytes",
  "cd16 mono",        "Non-classical monocytes",
  "cdc",              "cDC",
  "pdc",              "pDC",
  "plasma",           "Plasma cells",
  "cd4 naive",        "Naive CD4 T",
  "cd4 tcm",          "Eff./memory CD4",
  "cd4 fh em",        "Eff./memory CD4",
  "cd4 th1 em",       "Eff./memory CD4",
  "cd4 th17 em",      "Eff./memory CD4",
  "cd4 th2 em",       "Eff./memory CD4",
  "cd4 th1/th17 em",  "Eff./memory CD4",
  "treg",             "Eff./memory CD4",
  "cd8 naive",        "Naive CD8 T",
  "cd8 tcm",          "Eff./memory CD8",
  "cd8 tem",          "Eff./memory CD8",
  "nk",               "NK",
  "mait",             "MAIT",
  "tgd",              "gamma-delta T",
  "granulocyte",      "residual",
  "ilc",              "residual",
  "macrophage",       "residual",
  "mast",             "residual",
  "nkt",              "residual"
)

stopifnot(nrow(celltype_to_compartment) == 23L)
stopifnot(nrow(l3_to_compartment) == 26L)
stopifnot(!anyDuplicated(celltype_to_compartment$cell_type))
stopifnot(!anyDuplicated(l3_to_compartment$cell_type))

celltype_to_compartment
l3_to_compartment
```

## B2. SAVI counts and independent-subject model

### Construct the 15-part count composition

```{r}
#| label: panel-b-counts
# CellType counts from the GEO Seurat object (cached when present).

if (file.exists(panel_b_counts_rds)) {
  panel_b_celltype_counts <- readRDS(panel_b_counts_rds)
} else {
  stopifnot(file.exists(params$savi_seurat_gz))
  message("Building panel_b_counts_rds from Seurat: ", params$savi_seurat_gz)

  read_geo_rds_gz <- function(path) {
    con <- gzcon(gzfile(path))
    on.exit(close(con), add = TRUE)
    readRDS(con)
  }

  seurat_savi <- read_geo_rds_gz(params$savi_seurat_gz)
  md <- seurat_savi@meta.data
  rm(seurat_savi)
  gc()

  stopifnot(all(c("Sample", "Category", "CellType") %in% colnames(md)))

  panel_b_celltype_counts <- md |>
    tibble::rownames_to_column("cell_id") |>
    dplyr::transmute(
      sample_id = as.character(.data$Sample),
      Category = as.character(.data$Category),
      cell_type = as.character(.data$CellType)
    ) |>
    dplyr::count(.data$sample_id, .data$Category, .data$cell_type, name = "n_cells")

  dir.create(dirname(panel_b_counts_rds), recursive = TRUE, showWarnings = FALSE)
  saveRDS(panel_b_celltype_counts, panel_b_counts_rds, compress = "xz")
}

# Fail loudly if the source annotation changes. The mapping is intended to be
# an explicit biological decision, not an automatic string-matching step.
stopifnot(setequal(
  unique(panel_b_celltype_counts$cell_type),
  celltype_to_compartment$cell_type
))

panel_b_counts <- panel_b_celltype_counts |>
  dplyr::left_join(celltype_to_compartment, by = "cell_type") |>
  dplyr::group_by(.data$sample_id, .data$compartment) |>
  dplyr::summarise(n_cells = sum(.data$n_cells), .groups = "drop") |>
  tidyr::complete(
    sample_id = panel_b_sample_meta$sample_id,
    compartment = compartment_levels,
    fill = list(n_cells = 0L)
  ) |>
  dplyr::inner_join(panel_b_sample_meta, by = "sample_id") |>
  dplyr::mutate(
    compartment = factor(.data$compartment, levels = compartment_levels)
  )

# Each library's 15 compartments must sum to its original total cell count.
panel_b_count_check <- panel_b_counts |>
  dplyr::group_by(.data$sample_id, .data$group, .data$sequencing_batch) |>
  dplyr::summarise(
    n_cells = sum(.data$n_cells),
    n_compartments = dplyr::n(),
    .groups = "drop"
  )

stopifnot(nrow(panel_b_counts) == 17L * length(compartment_levels))
stopifnot(all(panel_b_count_check$n_compartments == length(compartment_levels)))

panel_b_raw_total_check <- panel_b_celltype_counts |>
  dplyr::group_by(.data$sample_id) |>
  dplyr::summarise(raw_n_cells = sum(.data$n_cells), .groups = "drop") |>
  dplyr::inner_join(
    panel_b_count_check |>
      dplyr::transmute(
        sample_id = .data$sample_id,
        aggregated_n_cells = .data$n_cells
      ),
    by = "sample_id"
  )

stopifnot(nrow(panel_b_raw_total_check) == 17L)
stopifnot(all(panel_b_raw_total_check$raw_n_cells ==
              panel_b_raw_total_check$aggregated_n_cells))

# The inferential comparison uses the 12 independent untreated/control subjects.
# Treated repeated measures are not needed for this hypothesis and are excluded
# rather than being treated as independent libraries.
panel_b_analysis_counts <- panel_b_counts |>
  dplyr::filter(.data$group %in% c("control", "untreated")) |>
  dplyr::mutate(
    group = factor(as.character(.data$group), levels = c("control", "untreated"))
  )

stopifnot(dplyr::n_distinct(panel_b_analysis_counts$sample_id) == 12L)
stopifnot(dplyr::n_distinct(panel_b_analysis_counts$subject_id) == 12L)

panel_b_count_check
```

### Fit the untreated/control composition

```{r}
#| label: panel-b-sccomp-fit
# Fifteen-compartment multi-beta-binomial on untreated/control subjects:
# ~ group + sequencing_batch (HMC).

if (file.exists(panel_b_fit_rds) && !isTRUE(params$run_sccomp_panel_b)) {
  panel_b_fit <- readRDS(panel_b_fit_rds)
} else {
  message("Fitting panel_b sccomp HMC model...")
  panel_b_fit <- sccomp::sccomp_estimate(
    panel_b_analysis_counts,
    formula_composition = ~ group + sequencing_batch,
    formula_variability = ~ 1,
    sample = "sample_id",
    cell_group = "compartment",
    abundance = "n_cells",
    cores = min(6L, max(1L, parallel::detectCores() - 2L)),
    inference_method = "hmc",
    verbose = TRUE
  )
  dir.create(dirname(panel_b_fit_rds), recursive = TRUE, showWarnings = FALSE)
  saveRDS(panel_b_fit, panel_b_fit_rds, compress = "xz")
}

stopifnot(identical(attr(panel_b_fit, "inference_method"), "hmc"))
stopifnot(nrow(attr(panel_b_fit, "model_input")$X) == 12L)
panel_b_fit
```

## B3. SAVI posterior on the common probability scale

The native coefficient-scale `sccomp_test()` is retained only as a sensitivity
analysis. Every displayed comparison is instead calculated from posterior
latent mean proportions and tested with the same percentage-point ROPE.

```{r}
#| label: panel-b-native-sensitivity

panel_b_native_tested <- sccomp::sccomp_test(
  panel_b_fit,
  contrasts = "groupuntreated"
)
```

### Standardise SAVI predictions over sequencing batch

```{r}
#| label: panel-b-savi-prediction

panel_b_newdata <- tidyr::expand_grid(
  group = factor(c("control", "untreated"),
                 levels = c("control", "untreated")),
  sequencing_batch = factor(c("STING1", "STING2", "STING3"),
                            levels = c("STING1", "STING2", "STING3"))
) |>
  dplyr::mutate(
    sample_id = paste(.data$group, .data$sequencing_batch, sep = "__"),
    target_weight = 1 / 3
  )

panel_b_pred_draws <- sccomp::sccomp_predict(
  panel_b_fit,
  new_data = panel_b_newdata,
  number_of_draws = as.integer(params$panel_b_prediction_draws),
  mcmc_seed = 20260808L,
  summary_instead_of_draws = FALSE
)

# sccomp_predict returns the latent softmax mean. Verify closure before
# standardising each group equally over the three sequencing batches.
panel_b_prediction_sum_check <- panel_b_pred_draws |>
  dplyr::group_by(.data$.draw, .data$sample_id) |>
  dplyr::summarise(total = sum(.data$proportion), .groups = "drop")

stopifnot(max(abs(panel_b_prediction_sum_check$total - 1)) < 1e-5)

panel_b_group_draws <- panel_b_pred_draws |>
  dplyr::group_by(.data$.draw, .data$group, .data$compartment) |>
  dplyr::summarise(
    proportion = stats::weighted.mean(.data$proportion, .data$target_weight),
    .groups = "drop"
  )

panel_b_group_sum_check <- panel_b_group_draws |>
  dplyr::group_by(.data$.draw, .data$group) |>
  dplyr::summarise(total = sum(.data$proportion), .groups = "drop")

stopifnot(max(abs(panel_b_group_sum_check$total - 1)) < 1e-5)
```

### Define the common probability-point test

```{r}
#| label: panel-b-probability-test

summarise_panel_b_pp_test <- function(draws, rope_pp) {
  stopifnot(all(c("contrast", "compartment", "delta_pp") %in% names(draws)))

  draws |>
    dplyr::filter(.data$compartment %in% display_compartments) |>
    dplyr::group_by(.data$contrast, .data$compartment) |>
    dplyr::summarise(
      estimate = stats::median(.data$delta_pp),
      .lower = unname(stats::quantile(.data$delta_pp, 0.025)),
      .upper = unname(stats::quantile(.data$delta_pp, 0.975)),
      p_increase = mean(.data$delta_pp > rope_pp),
      p_decrease = mean(.data$delta_pp < -rope_pp),
      .groups = "drop"
    ) |>
    dplyr::mutate(
      pp_pH0 = 1 - pmax(.data$p_increase, .data$p_decrease),
      posterior_direction = dplyr::if_else(
        .data$p_increase >= .data$p_decrease,
        "increase",
        "decrease"
      )
    ) |>
    dplyr::group_by(.data$contrast) |>
    dplyr::arrange(.data$pp_pH0, .by_group = TRUE) |>
    dplyr::mutate(
      pp_FDR = cummean(.data$pp_pH0),
      discovery = .data$pp_FDR <= 0.05
    ) |>
    dplyr::ungroup()
}

# Internal baseline: untreated minus study control, with both group means
# evaluated over the same sequencing-batch target distribution.
panel_b_uc_draws <- panel_b_group_draws |>
  dplyr::filter(.data$group %in% c("control", "untreated")) |>
  tidyr::pivot_wider(names_from = "group", values_from = "proportion") |>
  dplyr::mutate(
    delta_pp = 100 * (.data$untreated - .data$control),
    contrast = "untreated - study control"
  )

panel_b_uc <- summarise_panel_b_pp_test(
  panel_b_uc_draws,
  rope_pp = params$panel_b_rope_pp
)
```

## B4. Healthy HCA reference populations

The comprehensive baseline equal-averages age and sex over the observed healthy
HCA age×sex profiles (unweighted g-computation on the response scale). The
personalised baselines use the SAVI subjects' age and sex; unknown control ages
are integrated over the HCA age distribution conditional on known sex. Both
targets fix healthy blood and the same `10x Genomics 3` assay, while dataset and
ethnicity effects are handled identically so that their difference isolates
demographic personalisation.

### Build equal-average and SAVI-standardised metadata profiles

```{r}
#| label: panel-b-hca-targets

# Comprehensive target: average probabilities within each posterior draw over
# unique observed age-decade × sex profiles with equal weight. That is
# unweighted g-computation on the response scale; donor-frequency weights would
# instead emphasise common demographic cells, and setting every covariate to NA
# would predict at an average coefficient. Donors remain the units used to
# enumerate which age/sex profiles exist (repeat samples from one donor do not
# create extra profiles), but they do not weight the average.
panel_b_hca_comprehensive_meta <- hca_donors |>
  dplyr::transmute(
    age_decade = as.character(.data$age_decade),
    sex = as.character(.data$sex)
  ) |>
  dplyr::filter(
    !is.na(.data$age_decade),
    .data$sex %in% c("female", "male")
  ) |>
  dplyr::count(.data$age_decade, .data$sex, name = "n_donors") |>
  dplyr::mutate(
    target_weight = 1 / dplyr::n(),
    reference_target = "comprehensive_hca",
    source_subject_id = NA_character_,
    sample_id = paste0(
      "hca_comprehensive_age", .data$age_decade, "_", .data$sex
    )
  )

panel_b_hca_comprehensive_n_donors <- sum(
  panel_b_hca_comprehensive_meta$n_donors
)
panel_b_hca_comprehensive_n_excluded <-
  nrow(hca_donors) - panel_b_hca_comprehensive_n_donors

panel_b_hca_personalised_untreated_meta <- savi_libraries |>
  dplyr::filter(as.character(.data$state) == "untreated") |>
  dplyr::transmute(
    age_decade = dplyr::if_else(
      is.na(.data$age_years),
      NA_character_,
      as.character(ceiling(.data$age_years / 10))
    ),
    sex = as.character(.data$sex),
    reference_target = "personalised_untreated_hca",
    source_subject_id = as.character(.data$subject_id),
    sample_id = paste0("hca_untreated_", .data$subject_id),
    target_weight = 1 / dplyr::n()
  )

# Known-age study controls contribute one seventh each. C1 and C2 have known
# sex but unknown exact age, so each subject's one-seventh mass is distributed
# over the donor-derived HCA age distribution conditional on male sex.
panel_b_control_target_source <- savi_libraries |>
  dplyr::filter(as.character(.data$state) == "control") |>
  dplyr::transmute(
    age_decade = dplyr::if_else(
      is.na(.data$age_years),
      NA_character_,
      as.character(ceiling(.data$age_years / 10))
    ),
    sex = as.character(.data$sex),
    source_subject_id = as.character(.data$subject_id)
  )

panel_b_n_controls <- dplyr::n_distinct(
  panel_b_control_target_source$source_subject_id
)

panel_b_hca_age_given_sex <- panel_b_hca_comprehensive_meta |>
  dplyr::group_by(.data$sex) |>
  dplyr::mutate(
    conditional_age_weight = .data$n_donors / sum(.data$n_donors)
  ) |>
  dplyr::ungroup() |>
  dplyr::select("age_decade", "sex", "conditional_age_weight")

panel_b_hca_personalised_control_known <- panel_b_control_target_source |>
  dplyr::filter(!is.na(.data$age_decade)) |>
  dplyr::mutate(
    reference_target = "personalised_control_hca",
    sample_id = paste0("hca_control_", .data$source_subject_id),
    target_weight = 1 / panel_b_n_controls
  )

panel_b_hca_personalised_control_missing <- panel_b_control_target_source |>
  dplyr::filter(is.na(.data$age_decade)) |>
  dplyr::select(-"age_decade") |>
  dplyr::inner_join(
    panel_b_hca_age_given_sex,
    by = "sex",
    relationship = "many-to-many"
  ) |>
  dplyr::mutate(
    reference_target = "personalised_control_hca",
    sample_id = paste0(
      "hca_control_", .data$source_subject_id,
      "_age", .data$age_decade
    ),
    target_weight = .data$conditional_age_weight / panel_b_n_controls
  )

panel_b_hca_personalised_control_meta <- dplyr::bind_rows(
  panel_b_hca_personalised_control_known,
  panel_b_hca_personalised_control_missing
) |>
  dplyr::select(
    "age_decade", "sex", "reference_target", "source_subject_id",
    "sample_id", "target_weight"
  )

panel_b_hca_metadata <- dplyr::bind_rows(
  panel_b_hca_comprehensive_meta,
  panel_b_hca_personalised_untreated_meta,
  panel_b_hca_personalised_control_meta
) |>
  dplyr::mutate(
    # Hold technical context and tissue/disease eligibility constant for both
    # comprehensive and personalised targets. Dataset and ethnicity are treated
    # identically through the model's average-effect convention.
    assay_groups___altered = "10x Genomics 3",
    disease_groups___altered = "Normal",
    tissue_groups = "blood",
    ethnicity_groups_imputed = NA_character_,
    dataset_id___altered = NA_character_,
    profile_signature = paste(
      .data$sample_id,
      dplyr::coalesce(.data$age_decade, "NA"),
      dplyr::coalesce(.data$sex, "NA"),
      .data$reference_target,
      .data$assay_groups___altered,
      .data$disease_groups___altered,
      .data$tissue_groups,
      format(.data$target_weight, digits = 17, scientific = FALSE),
      sep = "|"
    )
  )

panel_b_hca_weight_check <- panel_b_hca_metadata |>
  dplyr::group_by(.data$reference_target) |>
  dplyr::summarise(total_weight = sum(.data$target_weight), .groups = "drop")

stopifnot(all(abs(panel_b_hca_weight_check$total_weight - 1) < 1e-12))
stopifnot(dplyr::n_distinct(panel_b_hca_personalised_untreated_meta$sample_id) == 5L)
stopifnot(dplyr::n_distinct(
  panel_b_hca_personalised_control_meta$source_subject_id
) == 7L)

panel_b_hca_weight_check
```

### Generate or validate the full-L3 HCA prediction cache

```{r}
#| label: panel-b-hca-prediction

# A valid Panel-b cache retains every HCA L3 component. Filtering before broad
# aggregation would lose compositional mass and under-count broad compartments.
hca_predict <- NULL
if (file.exists(panel_b_hca_predict_rds) && !isTRUE(params$run_hca_panel_b)) {
  hca_candidate <- readRDS(panel_b_hca_predict_rds)
  hca_required_columns <- c(
    "sample_id", "profile_signature", "reference_target", "target_weight", "cell_type",
    "proportion", ".draw"
  )

  if (all(hca_required_columns %in% names(hca_candidate)) &&
      setequal(
        unique(hca_candidate$profile_signature),
        panel_b_hca_metadata$profile_signature
      )) {
    hca_candidate_sum <- hca_candidate |>
      dplyr::group_by(.data$.draw, .data$sample_id) |>
      dplyr::summarise(total = sum(.data$proportion), .groups = "drop")

    if (max(abs(hca_candidate_sum$total - 1)) < 1e-5) {
      hca_predict <- hca_candidate
    } else {
      message("Ignoring incomplete Panel-b HCA cache: retained L3 mass does not sum to one.")
    }
  } else {
    message("Ignoring incompatible Panel-b HCA cache: target profiles or columns changed.")
  }
}

if (is.null(hca_predict)) {
  stopifnot(file.exists(params$sccomp_estimates_rds))
  message("Building full harmonised Panel-b HCA prediction cache...")

  sccomp_est <- readRDS(params$sccomp_estimates_rds)

  hca_predict <- sccomp_est |>
    sccomp::sccomp_predict(
      formula_composition = NULL,
      new_data = panel_b_hca_metadata,
      number_of_draws = as.integer(params$panel_b_prediction_draws),
      mcmc_seed = 20260809L,
      summary_instead_of_draws = FALSE
    ) |>
    dplyr::rename(cell_type = "L3") |>
    dplyr::select(
      dplyr::any_of(c(
        "sample_id", "profile_signature", "reference_target", "target_weight",
        "source_subject_id", "age_decade", "sex", "cell_type", "proportion", ".draw"
      ))
    )

  dir.create(dirname(panel_b_hca_predict_rds), recursive = TRUE, showWarnings = FALSE)
  saveRDS(hca_predict, panel_b_hca_predict_rds, compress = "xz")
  rm(sccomp_est)
  gc()
}

stopifnot(setequal(unique(hca_predict$sample_id), panel_b_hca_metadata$sample_id))
stopifnot(setequal(unique(hca_predict$cell_type), l3_to_compartment$cell_type))

hca_profile_sum_check <- hca_predict |>
  dplyr::group_by(.data$.draw, .data$sample_id) |>
  dplyr::summarise(total = sum(.data$proportion), .groups = "drop")

stopifnot(max(abs(hca_profile_sum_check$total - 1)) < 1e-5)
```

### Aggregate HCA L3 draws to the shared 15-part composition

```{r}
#| label: panel-b-hca-harmonised-draws

hca_profile_comp_draws <- hca_predict |>
  dplyr::left_join(l3_to_compartment, by = "cell_type") |>
  dplyr::group_by(
    .data$.draw, .data$sample_id, .data$reference_target,
    .data$target_weight, .data$compartment
  ) |>
  dplyr::summarise(proportion = sum(.data$proportion), .groups = "drop")

hca_profile_comp_sum_check <- hca_profile_comp_draws |>
  dplyr::group_by(.data$.draw, .data$sample_id) |>
  dplyr::summarise(total = sum(.data$proportion), .groups = "drop")

stopifnot(max(abs(hca_profile_comp_sum_check$total - 1)) < 1e-5)

hca_comp_draws <- hca_profile_comp_draws |>
  dplyr::group_by(.data$.draw, .data$reference_target, .data$compartment) |>
  dplyr::summarise(
    proportion = sum(.data$target_weight * .data$proportion),
    .groups = "drop"
  )

hca_comp_sum_check <- hca_comp_draws |>
  dplyr::group_by(.data$.draw, .data$reference_target) |>
  dplyr::summarise(total = sum(.data$proportion), .groups = "drop")

stopifnot(max(abs(hca_comp_sum_check$total - 1)) < 1e-5)
```

## B5. Independent cross-model posterior pairing

SAVI and HCA are separately fitted Bayesian models, so draw indices have no
cross-model pairing. We sample their indices independently and apply each pair
to the full composition, preserving within-draw dependence among cell types.

```{r}
#| label: panel-b-cross-model-pairs

untreated_draws <- panel_b_group_draws |>
  dplyr::filter(.data$group == "untreated") |>
  dplyr::transmute(
    .draw = .data$.draw,
    compartment = .data$compartment,
    case_proportion = .data$proportion
  )

control_draws <- panel_b_group_draws |>
  dplyr::filter(.data$group == "control") |>
  dplyr::transmute(
    .draw = .data$.draw,
    compartment = .data$compartment,
    case_proportion = .data$proportion
  )

set.seed(20260810L)
panel_b_cross_fit_pairs <- tibble::tibble(
  .pair = seq_len(as.integer(params$panel_b_pair_draws)),
  .draw_savi = sample(
    unique(untreated_draws$.draw),
    as.integer(params$panel_b_pair_draws),
    replace = TRUE
  ),
  .draw_hca = sample(
    unique(hca_comp_draws$.draw),
    as.integer(params$panel_b_pair_draws),
    replace = TRUE
  )
)
```

## B6. Healthy-reference calibration — conclusion first

Calibration compares the study controls with each HCA reference under the same
probability-point estimand and seven-compartment testing family. This is an
interpretation gate: failure indicates that an untreated–HCA contrast contains
residual reference-domain shift as well as disease-associated composition.

```{r}
#| label: panel-b-hca-calibration

# Study controls should be compatible with HCA after population marginalisation
# or control-metadata standardisation. The same cross-model draw pairs are used
# for all compartments, preserving compositional dependence.
panel_b_hca_calibration_draws <- panel_b_cross_fit_pairs |>
  dplyr::left_join(
    control_draws,
    by = c(".draw_savi" = ".draw"),
    relationship = "many-to-many"
  ) |>
  dplyr::left_join(
    hca_comp_draws |>
      dplyr::filter(.data$reference_target %in% c(
        "comprehensive_hca", "personalised_control_hca"
      )) |>
      dplyr::rename(hca_proportion = "proportion"),
    by = c(".draw_hca" = ".draw", "compartment"),
    relationship = "many-to-many"
  ) |>
  dplyr::mutate(
    delta_pp = 100 * (.data$case_proportion - .data$hca_proportion),
    contrast = dplyr::recode(
      .data$reference_target,
      comprehensive_hca = "study control - comprehensive HCA",
      personalised_control_hca = "study control - personalised HCA"
    )
  )

panel_b_hca_calibration <- summarise_panel_b_pp_test(
  panel_b_hca_calibration_draws,
  rope_pp = params$panel_b_rope_pp
)

panel_b_calibration_discoveries <- panel_b_hca_calibration |>
  dplyr::group_by(.data$contrast) |>
  dplyr::summarise(
    n_discoveries = sum(.data$discovery),
    discovered_compartments = paste(
      as.character(.data$compartment[.data$discovery]),
      collapse = ", "
    ),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    discovered_compartments = dplyr::if_else(
      .data$n_discoveries == 0L,
      "None",
      .data$discovered_compartments
    )
  )

panel_b_comprehensive_calibration_n <- panel_b_calibration_discoveries |>
  dplyr::filter(.data$contrast == "study control - comprehensive HCA") |>
  dplyr::pull(.data$n_discoveries)

panel_b_personalised_calibration_n <- panel_b_calibration_discoveries |>
  dplyr::filter(.data$contrast == "study control - personalised HCA") |>
  dplyr::pull(.data$n_discoveries)

panel_b_personalised_calibration_types <- panel_b_calibration_discoveries |>
  dplyr::filter(.data$contrast == "study control - personalised HCA") |>
  dplyr::pull(.data$discovered_compartments)

panel_b_comprehensive_discovery_term <- ifelse(
  panel_b_comprehensive_calibration_n == 1L,
  "discovery",
  "discoveries"
)

panel_b_calibration_conclusion <- dplyr::case_when(
  panel_b_personalised_calibration_n == 0L &&
    panel_b_comprehensive_calibration_n == 0L ~ paste(
      "Neither HCA reference produced a calibration discovery among the seven",
      "prespecified compartments. This supports interpreting the subsequent",
      "untreated–HCA contrasts as disease-associated shifts within the tested",
      "composition and model assumptions."
    ),
  panel_b_personalised_calibration_n == 0L ~ paste0(
    "The comprehensive reference produced ",
    panel_b_comprehensive_calibration_n,
    " calibration ", panel_b_comprehensive_discovery_term,
    ", whereas the personalised reference",
    " produced none. Metadata standardisation therefore improved transportability;",
    " the personalised HCA contrast is the primary external-reference result and",
    " the comprehensive contrast quantifies population-reference shift."
  ),
  TRUE ~ paste0(
    "The comprehensive and personalised references produced ",
    panel_b_comprehensive_calibration_n, " and ",
    panel_b_personalised_calibration_n,
    " calibration discoveries, respectively. Personalisation did not resolve",
    " the healthy-reference mismatch; the personalised discoveries were ",
    "(",
    panel_b_personalised_calibration_types,
    "). The HCA disease contrasts therefore retain residual reference-domain",
    " shift and are reported as sensitivity analyses; the study-control contrast",
    " remains the primary disease comparison."
  )
)
```

```{r}
#| label: panel-b-calibration-conclusion
#| results: asis

cat("**Calibration conclusion.** ", panel_b_calibration_conclusion, "\n\n", sep = "")
```

```{r}
#| label: panel-b-calibration-summary

panel_b_calibration_discoveries |>
  dplyr::mutate(
    baseline = dplyr::recode(
      .data$contrast,
      `study control - comprehensive HCA` = "Comprehensive HCA",
      `study control - personalised HCA` = "Personalised HCA"
    )
  ) |>
  dplyr::select(
    baseline,
    `FDR discoveries (of 7)` = .data$n_discoveries,
    `Discovered compartments` = .data$discovered_compartments
  ) |>
  knitr::kable(
    caption = "Healthy-reference calibration is evaluated before disease contrasts."
  )
```

## B7. Disease contrasts conditional on calibration

Open circles compare untreated SAVI with study controls. Grey diamonds compare
untreated SAVI with the equal-average comprehensive HCA population; teal
triangles use HCA predictions standardised to the untreated subjects' age and
sex. All use a 1-percentage-point ROPE and Bayesian FDR across the same seven
prespecified compartments.

```{r}
#| label: panel-b-disease-contrasts

panel_b_uh_draws <- panel_b_cross_fit_pairs |>
  dplyr::left_join(
    untreated_draws,
    by = c(".draw_savi" = ".draw"),
    relationship = "many-to-many"
  ) |>
  dplyr::left_join(
    hca_comp_draws |>
      dplyr::filter(.data$reference_target %in% c(
        "comprehensive_hca", "personalised_untreated_hca"
      )) |>
      dplyr::rename(hca_proportion = "proportion"),
    by = c(".draw_hca" = ".draw", "compartment"),
    relationship = "many-to-many"
  ) |>
  dplyr::mutate(
    delta_pp = 100 * (.data$case_proportion - .data$hca_proportion),
    contrast = dplyr::recode(
      .data$reference_target,
      comprehensive_hca = "untreated - comprehensive HCA",
      personalised_untreated_hca = "untreated - personalised HCA"
    )
  )

panel_b_uh <- summarise_panel_b_pp_test(
  panel_b_uh_draws,
  rope_pp = params$panel_b_rope_pp
)

panel_b_effects <- dplyr::bind_rows(panel_b_uc, panel_b_uh) |>
  dplyr::mutate(
    compartment = factor(
      as.character(.data$compartment),
      levels = display_compartments
    ),
    contrast = factor(
      .data$contrast,
      levels = c(
        "untreated - study control",
        "untreated - comprehensive HCA",
        "untreated - personalised HCA"
      )
    )
  ) |>
  dplyr::filter(!is.na(.data$compartment)) |>
  dplyr::left_join(source_direction, by = "compartment")

panel_b_effects |>
  dplyr::arrange(.data$compartment, .data$contrast)
```

### Plot the three calibrated comparison targets

```{r}
#| label: panel-b-helpers

plot_panel_b_composition <- function(effects_df) {
  y_labels <- effects_df |>
    dplyr::distinct(.data$compartment, .data$source_arrow) |>
    dplyr::mutate(
      y_lab = paste(.data$source_arrow, as.character(.data$compartment))
    ) |>
    dplyr::arrange(.data$compartment)

  lab_map <- setNames(
    y_labels$y_lab,
    as.character(y_labels$compartment)
  )

  contrast_breaks <- c(
    "untreated - study control",
    "untreated - comprehensive HCA",
    "untreated - personalised HCA"
  )
  contrast_labels <- c(
    "Study control",
    "Comprehensive HCA",
    "Personalised HCA"
  )
  contrast_colours <- c(
    "untreated - study control" = col_dark,
    "untreated - comprehensive HCA" = col_hca,
    "untreated - personalised HCA" = col_meta
  )
  contrast_fills <- c(
    "untreated - study control" = "white",
    "untreated - comprehensive HCA" = col_hca,
    "untreated - personalised HCA" = col_meta
  )
  contrast_shapes <- c(
    "untreated - study control" = 21,
    "untreated - comprehensive HCA" = 23,
    "untreated - personalised HCA" = 24
  )
  contrast_dodge <- ggplot2::position_dodge(width = 0.55)

  ggplot2::ggplot(
    effects_df,
    ggplot2::aes(y = .data$compartment, group = .data$contrast)
  ) +
    ggplot2::geom_vline(
      xintercept = 0,
      colour = col_mid,
      linetype = "22",
      linewidth = 0.35
    ) +
    ggplot2::geom_linerange(
      ggplot2::aes(
        xmin = .data$.lower,
        xmax = .data$.upper,
        x = .data$estimate,
        colour = .data$contrast
      ),
      linewidth = 0.45,
      position = contrast_dodge
    ) +
    ggplot2::geom_point(
      ggplot2::aes(
        x = .data$estimate,
        colour = .data$contrast,
        fill = .data$contrast,
        shape = .data$contrast,
        size = .data$discovery
      ),
      stroke = 0.55,
      position = contrast_dodge
    ) +
    ggplot2::scale_colour_manual(
      values = contrast_colours,
      breaks = contrast_breaks,
      labels = contrast_labels,
      name = NULL
    ) +
    ggplot2::scale_fill_manual(
      values = contrast_fills,
      breaks = contrast_breaks,
      labels = contrast_labels,
      name = NULL
    ) +
    ggplot2::scale_shape_manual(
      values = contrast_shapes,
      breaks = contrast_breaks,
      labels = contrast_labels,
      name = NULL
    ) +
    ggplot2::scale_size_manual(values = c(`FALSE` = 1.7, `TRUE` = 2.4), guide = "none") +
    ggplot2::scale_y_discrete(
      limits = rev(display_compartments),
      labels = lab_map
    ) +
    ggplot2::scale_x_continuous(
      breaks = c(-20, -10, 0, 10, 20)
    ) +
    ggplot2::coord_cartesian(xlim = c(-22, 22), clip = "off") +
    ggplot2::labs(
      x = "Composition difference (percentage points)",
      y = NULL
    ) +
    theme_multipanel +
    ggplot2::theme(
      legend.position = "bottom",
      legend.direction = "horizontal",
      legend.key.width = grid::unit(4, "mm"),
      legend.spacing.x = grid::unit(1.5, "mm"),
      plot.margin = margin(4, 8, 4, 4)
    )
}
```

```{r}
#| label: panel-b-draw
#| fig-width: 6
#| fig-height: 8
#| fig-cap: "Figure 2b. Posterior mean-composition tests versus study controls, the equal-average comprehensive HCA population and SAVI-standardised personalised HCA. Calibration is reported before these disease contrasts. Larger symbols meet Bayesian FDR ≤ 0.05 for a 1-percentage-point ROPE."

panel_b <- plot_panel_b_composition(panel_b_effects)

panel_b
```

### Reproducibility and sensitivity audit

```{r}
#| label: panel-b-audit-text

panel_b_native_sensitivity <- panel_b_native_tested |>
  dplyr::filter(.data$compartment %in% display_compartments) |>
  dplyr::transmute(
    compartment = .data$compartment,
    native_c_effect = .data$c_effect,
    native_c_lower = .data$c_lower,
    native_c_upper = .data$c_upper,
    native_c_pH0 = .data$c_pH0,
    native_c_FDR = .data$c_FDR
  )

panel_b_test_audit <- panel_b_effects |>
  dplyr::mutate(
    matches_source = dplyr::case_when(
      .data$source_arrow == "+" ~ .data$estimate > 0,
      .data$source_arrow == "-" ~ .data$estimate < 0,
      TRUE ~ NA
    )
  )

panel_b_internal_audit <- panel_b_test_audit |>
  dplyr::filter(.data$contrast == "untreated - study control") |>
  dplyr::left_join(panel_b_native_sensitivity, by = "compartment") |>
  dplyr::mutate(
    native_matches_source = dplyr::case_when(
      .data$source_arrow == "+" ~ .data$native_c_effect > 0,
      .data$source_arrow == "-" ~ .data$native_c_effect < 0,
      TRUE ~ NA
    )
  )

hca_missing_mapped_types <- setdiff(
  l3_to_compartment$cell_type,
  unique(hca_predict$cell_type)
)

cat(
  sprintf(
    paste(
      "Panel b audit: %d raw GEO labels -> %d shared compartments;",
      "HMC fit on %d independent untreated/control subjects;",
      "formula ~ group + sequencing_batch;",
      "comprehensive HCA target equal-averages %d age/sex profiles",
      "(from %d healthy donors; %d missing age or sex excluded);",
      "all displayed contrasts use a %.2f-pp ROPE and Bayesian FDR across 7 compartments;",
      "%d case posterior draws, %d HCA posterior draws and %d independent cross-fit pairs;",
      "full HCA cache contains %d L3 types and %d expected mapping types are absent;",
      "maximum HCA closure error = %.3g.\n"
    ),
    dplyr::n_distinct(panel_b_celltype_counts$cell_type),
    length(compartment_levels),
    dplyr::n_distinct(panel_b_analysis_counts$subject_id),
    nrow(panel_b_hca_comprehensive_meta),
    panel_b_hca_comprehensive_n_donors,
    panel_b_hca_comprehensive_n_excluded,
    params$panel_b_rope_pp,
    dplyr::n_distinct(panel_b_group_draws$.draw),
    dplyr::n_distinct(hca_comp_draws$.draw),
    nrow(panel_b_cross_fit_pairs),
    dplyr::n_distinct(hca_predict$cell_type),
    length(hca_missing_mapped_types),
    max(abs(hca_profile_sum_check$total - 1))
  )
)

panel_b_test_audit |>
  dplyr::arrange(.data$compartment, .data$contrast) |>
  dplyr::select(
    compartment, source_arrow, estimate, .lower, .upper,
    p_increase, p_decrease, pp_pH0, pp_FDR, discovery, matches_source, contrast
  )

panel_b_hca_calibration |>
  dplyr::arrange(.data$compartment, .data$contrast) |>
  dplyr::select(
    compartment, estimate, .lower, .upper,
    p_increase, p_decrease, pp_pH0, pp_FDR, discovery, contrast
  )

panel_b_internal_audit |>
  dplyr::select(
    compartment, estimate, pp_FDR, native_c_effect, native_c_lower,
    native_c_upper, native_c_FDR, matches_source, native_matches_source
  )

panel_b_calibration_discoveries
```

------------------------------------------------------------------------

# Panel c — Covered signature log-μ audit

**Method (cohort log-μ analysis aligned with Panel b).** Both sides target the
NB location parameter on the shared **offset-zero** log-μ scale used by the
monocytic atlas models:

- **SAVI cohorts:** counts from repeated libraries within one subject×state are
  first added, with their exposures added on the natural scale. Gene-wise cohort
  log-μ is then re-estimated by `edgeR::mglmOneGroup()` using trended NB
  dispersions estimated from all shared genes. A subject-level Bayesian
  bootstrap supplies the same Dirichlet weight vector to all 16 genes, so gene
  correlations induced by the study subjects are retained in the signature
  summary without zero-only discrete resamples producing infinite log-μ limits.
- **HCA comprehensive reference:** `posterior_linpred(transform = FALSE)` is
  standardized to the same observed healthy-HCA age×sex profiles and profile
  weights used by Panel b. Ethnicity is equal-standardized over fitted levels;
  healthy blood, `10x Genomics 3`, and `offset = 0` are fixed. An unseen-study
  dataset intercept (`NEW_STUDY_SAVI`) is drawn from the hierarchical population
  (`allow_new_levels = TRUE`, `sample_new_levels = "gaussian"`).

**Population.** Cluster **17. Disease-associated monocytes** (case-study
population). Blueprint monocyte 5+12+17 alignment is a declared sensitivity, not
this panel.

**Four aligned log2 contrasts.** (1) untreated − study control; (2) untreated −
HCA; (3) paired untreated − treated (P1/P2/P4/P6; P1's two treated libraries
are collapsed before modelling); (4) study control − HCA. Every interval is
formed from a complete replicate of the relevant contrast. The top strip is the
equal-weight mean across the 16 genes *inside each replicate*, rather than a
bootstrap over 16 fixed gene estimates. Study-only intervals are subject-level
Bayesian-bootstrap intervals; HCA intervals independently pair a Bayesian-
bootstrap replicate with draws from the separately fitted HCA gene posteriors
and are therefore modular hybrid, not joint-model, intervals. Because the HCA
genes were fitted separately, their posterior cross-gene covariance is not
estimable and is not manufactured by matching nominal draw indices.

Set `params$run_panel_c_rebuild: true` to force rebuild of all panel-c caches.

## C1. Covered gene set and helpers

```{r}
#| label: panel-c-setup

suppressPackageStartupMessages({
  library(SeuratObject)
  library(AnnotationDbi)
  library(org.Hs.eg.db)
  library(tidybulk)
  library(tidySummarizedExperiment)
  library(brms)
  library(edgeR)
})

merge_summarized_experiments <- function(...) {
  ses <- list(...)
  assay_names <- Reduce(intersect, lapply(ses, SummarizedExperiment::assayNames))
  genes <- Reduce(intersect, lapply(ses, rownames))
  if (!length(assay_names) || !length(genes)) {
    stop("No shared assays/genes between SummarizedExperiment objects.")
  }
  ses <- lapply(ses, function(se) se[genes, , drop = FALSE])
  cds <- lapply(ses, function(se) as.data.frame(SummarizedExperiment::colData(se)))
  all_nms <- Reduce(union, lapply(cds, names))
  cds <- lapply(cds, function(cd) {
    for (nm in setdiff(all_nms, names(cd))) cd[[nm]] <- NA
    cd[, all_nms, drop = FALSE]
  })
  assays <- stats::setNames(
    lapply(assay_names, function(an) {
      do.call(base::cbind, lapply(ses, function(se) {
        base::as.matrix(SummarizedExperiment::assay(se, an))
      }))
    }),
    assay_names
  )
  SummarizedExperiment::SummarizedExperiment(
    assays = assays,
    colData = S4Vectors::DataFrame(
      do.call(rbind, cds),
      row.names = unlist(lapply(ses, colnames), use.names = FALSE)
    ),
    rowData = SummarizedExperiment::rowData(ses[[1]])
  )
}

build_hca_covariate_grid <- function(fit, age_sex_target) {
  # Comprehensive target:
  #   fixed       — blood, Normal, 10x Genomics 3, offset = 0
  #   standardize — observed HCA age×sex profiles at Panel-b target weights
  #   ethnicity   — equal-standardize fitted levels within each profile
  #   dataset     — unseen study level; drawn hierarchically under
  #                 allow_new_levels = TRUE, sample_new_levels = "gaussian"
  factor_levels <- function(v) {
    lv <- if (is.factor(v)) {
      levels(v)
    } else {
      sort(unique(as.character(v)))
    }
    lv[!is.na(lv) & nzchar(lv)]
  }

  fd <- fit$data
  age_lv <- factor_levels(fd$age_decade)
  sex_lv <- factor_levels(fd$sex)
  eth_lv <- factor_levels(fd$ethnicity_groups)
  assay_lv <- factor_levels(fd$assay_groups_altered)
  if (!"10x Genomics 3" %in% assay_lv) {
    stop("`10x Genomics 3` is absent from the fitted assay support.")
  }
  if (!length(eth_lv)) {
    stop("Ethnicity levels are absent from the fitted model data.")
  }

  target <- age_sex_target |>
    dplyr::transmute(
      age_decade = as.character(.data$age_decade),
      sex = as.character(.data$sex),
      age_sex_weight = as.numeric(.data$target_weight)
    ) |>
    dplyr::distinct()

  stopifnot(abs(sum(target$age_sex_weight) - 1) < 1e-12)
  if (!all(target$age_decade %in% age_lv) || !all(target$sex %in% sex_lv)) {
    stop("Panel-b age×sex target contains levels absent from a Panel-c fit.")
  }

  tidyr::crossing(target, ethnicity_groups = eth_lv) |>
    dplyr::mutate(
      disease_groups_altered = "Normal",
      tissue_groups = "blood",
      assay_groups_altered = "10x Genomics 3",
      dataset_id_altered = "NEW_STUDY_SAVI",
      offset = 0,
      counts = 1,
      profile_weight = .data$age_sex_weight / length(eth_lv)
    )
}

# Cache key for the comprehensive linpred target (bump when the estimand changes).
panel_c_hca_profile_id <- panel_b_hca_comprehensive_meta |>
  dplyr::transmute(
    id = paste(
      as.character(.data$age_decade), as.character(.data$sex),
      format(.data$target_weight, digits = 17, scientific = FALSE),
      sep = ":"
    )
  ) |>
  dplyr::pull(.data$id) |>
  sort() |>
  paste(collapse = ";")

panel_c_estimates_info <- file.info(params$estimates_rds)
panel_c_hca_grid_id <- paste(
  "v3_fixed=blood+Normal+10xGenomics3+offset0",
  paste0("age_sex_target=", panel_c_hca_profile_id),
  "ethnicity=equal_fitted_levels",
  "dataset=NEW_STUDY_SAVI_gaussian_new_level",
  paste0("fit_size=", panel_c_estimates_info$size),
  paste0("fit_mtime=", as.numeric(panel_c_estimates_info$mtime)),
  sep = "|"
)

panel_c_tail_p <- function(x) {
  stopifnot(length(x) > 0L, all(is.finite(x)))
  p_lo <- (1 + sum(x <= 0)) / (length(x) + 1)
  p_hi <- (1 + sum(x >= 0)) / (length(x) + 1)
  min(1, 2 * min(p_lo, p_hi))
}

summarise_panel_c_draw_matrix <- function(
  delta,
  genes,
  contrast_name,
  uncertainty_method,
  n_case,
  n_reference,
  is_aggregate = FALSE
) {
  delta <- as.matrix(delta)
  stopifnot(ncol(delta) == length(genes))

  tibble::tibble(
    gene = genes,
    estimate = apply(delta, 2, stats::median, na.rm = TRUE),
    .lower = apply(delta, 2, stats::quantile, probs = 0.025, na.rm = TRUE),
    .upper = apply(delta, 2, stats::quantile, probs = 0.975, na.rm = TRUE),
    tail_p = apply(delta, 2, panel_c_tail_p),
    contrast = contrast_name,
    n_resamples = nrow(delta),
    n_case = n_case,
    n_reference = n_reference,
    uncertainty_method = uncertainty_method,
    is_aggregate = is_aggregate
  )
}

tibble::tibble(
  gene = panel_c_gene_order,
  rank = seq_along(panel_c_gene_order)
)
```

## C2. Merge SAVI cluster 17 with atlas reference and set TMM offsets

```{r}
#| label: panel-c-merge-se

rebuild_panel_c_se <- isTRUE(params$run_panel_c_rebuild) || !file.exists(panel_c_se_rds)

if (!rebuild_panel_c_se) {
  .panel_c_se_cache <- readRDS(panel_c_se_rds)
  se_monocytic_panel_c <- .panel_c_se_cache$se_monocytic
  panel_c_ref_sample_name <- .panel_c_se_cache$ref_sample_name
  message(
    "Loaded panel-c SE: ",
    nrow(se_monocytic_panel_c), " genes x ",
    ncol(se_monocytic_panel_c), " samples"
  )
} else {
  message("Building panel-c SE merge + TMMwsp ...")
  stopifnot(file.exists(params$savi_pseudobulk_rds))
  stopifnot(file.exists(params$pseudobulk_h5ad))

  SAVI_pseudobulk_sobj <-
    readRDS(params$savi_pseudobulk_rds) |>
    subset(subset = CellType == "17. Disease-associated monocytes")

  counts_sym <- as.matrix(
    SeuratObject::LayerData(SAVI_pseudobulk_sobj, assay = "RNA", layer = "counts")
  )
  sym <- rownames(counts_sym)
  ensg <- AnnotationDbi::mapIds(
    org.Hs.eg.db::org.Hs.eg.db,
    keys = sym,
    column = "ENSEMBL",
    keytype = "SYMBOL",
    multiVals = "first"
  )
  keep <- !is.na(ensg) & !duplicated(ensg)
  ensg <- ensg[keep]
  counts_ens <- counts_sym[keep, , drop = FALSE]
  rownames(counts_ens) <- ensg

  cd_savi <- as.data.frame(SAVI_pseudobulk_sobj@meta.data)
  rownames(cd_savi) <- colnames(counts_ens)
  cd_savi$sample_role <- "savi_pseudobulk"
  cd_savi$offset <- 0L
  if (!"cell_type_unified_ensemble" %in% names(cd_savi)) {
    cd_savi$cell_type_unified_ensemble <- cd_savi$CellType
  }

  se_savi_ensembl <- SummarizedExperiment::SummarizedExperiment(
    assays = list(counts = counts_ens),
    colData = S4Vectors::DataFrame(cd_savi, row.names = colnames(counts_ens)),
    rowData = S4Vectors::DataFrame(gene_symbol = sym[keep], row.names = ensg)
  )

  pseudobulk_sample <- targets::tar_read(
    pseudobulk_sample,
    store = file.path(model_path, "_targets")
  )
  # Point DelayedArray/HDF5 backend to the local h5ad-backed matrix.
  pseudobulk_sample@assays@data@listData[["counts"]]@seed@seed@seed@filepath <-
    params$pseudobulk_h5ad
  ref_sample <-
    pseudobulk_sample[, SummarizedExperiment::colData(pseudobulk_sample)$offset == 0]
  SummarizedExperiment::colData(ref_sample)$sample_role <- "reference"

  se_monocytic_panel_c <- merge_summarized_experiments(se_savi_ensembl, ref_sample)
  panel_c_ref_sample_name <-
    se_monocytic_panel_c |>
    SummarizedExperiment::colData() |>
    as.data.frame() |>
    dplyr::filter(.data$sample_role == "reference") |>
    rownames()

  se_monocytic_panel_c <-
    se_monocytic_panel_c |>
    tidybulk::scale_abundance(
      method = "TMMwsp",
      reference_sample = panel_c_ref_sample_name
    ) |>
    dplyr::mutate(offset = log(1 / .data$multiplier))

  dir.create(dirname(panel_c_se_rds), recursive = TRUE, showWarnings = FALSE)
  saveRDS(
    list(
      se_monocytic = se_monocytic_panel_c,
      ref_sample_name = panel_c_ref_sample_name
    ),
    panel_c_se_rds,
    compress = "xz"
  )
}

tibble::tibble(
  n_features = nrow(se_monocytic_panel_c),
  n_samples = ncol(se_monocytic_panel_c),
  n_savi = sum(
    SummarizedExperiment::colData(se_monocytic_panel_c)$sample_role ==
      "savi_pseudobulk"
  ),
  n_reference = sum(
    SummarizedExperiment::colData(se_monocytic_panel_c)$sample_role ==
      "reference"
  ),
  ref_sample = panel_c_ref_sample_name[[1]]
)
```

## C3. Library metadata and offset-centred log-rate audit

```{r}
#| label: panel-c-sample-logmu

cd_panel_c <- as.data.frame(SummarizedExperiment::colData(se_monocytic_panel_c)) |>
  tibble::rownames_to_column("col_id")
savi_cd_panel_c <- cd_panel_c |>
  dplyr::filter(.data$sample_role == "savi_pseudobulk")
stopifnot(nrow(savi_cd_panel_c) > 0L)
stopifnot(all(c("offset", "multiplier") %in% names(savi_cd_panel_c)))

rd_panel_c <- as.data.frame(SummarizedExperiment::rowData(se_monocytic_panel_c))
sym_col <- if ("gene_symbol" %in% names(rd_panel_c)) {
  "gene_symbol"
} else if ("gene_name" %in% names(rd_panel_c)) {
  "gene_name"
} else {
  stop("No gene symbol column in panel-c rowData.")
}

panel_c_feature_map <- tibble::tibble(
  .feature = rownames(rd_panel_c),
  gene = as.character(rd_panel_c[[sym_col]])
) |>
  dplyr::filter(.data$gene %in% panel_c_gene_order)
stopifnot(nrow(panel_c_feature_map) == length(panel_c_gene_order))

panel_c_counts <- as.matrix(
  SummarizedExperiment::assay(se_monocytic_panel_c, "counts")
)

panel_c_sample_meta <- savi_cd_panel_c |>
  dplyr::mutate(
    sample_id = as.character(dplyr::coalesce(
      as.character(.data$Sample),
      .data$col_id
    )),
    Category = as.character(.data$Category),
    group = dplyr::recode(
      .data$Category,
      CTRL = "control",
      SAVI = "untreated",
      SAVI_treated = "treated"
    ),
    subject_id = dplyr::case_when(
      grepl("^P1", .data$sample_id) ~ "P1",
      grepl("^P2", .data$sample_id) ~ "P2",
      grepl("^P4", .data$sample_id) ~ "P4",
      grepl("^P5", .data$sample_id) ~ "P5",
      grepl("^P6", .data$sample_id) ~ "P6",
      grepl("^C", .data$sample_id) ~ .data$sample_id,
      TRUE ~ .data$sample_id
    ),
    offset = as.numeric(.data$offset)
  )

panel_c_sample_cache_id <- paste(
  "v2_offset_log_rate_audit",
  paste(panel_c_gene_order, collapse = ","),
  paste(
    panel_c_sample_meta$col_id,
    panel_c_sample_meta$subject_id,
    panel_c_sample_meta$group,
    format(panel_c_sample_meta$offset, digits = 17, scientific = FALSE),
    sep = ":",
    collapse = ";"
  ),
  paste(
    as.numeric(panel_c_counts[
      panel_c_feature_map$.feature[match(panel_c_gene_order, panel_c_feature_map$gene)],
      panel_c_sample_meta$col_id,
      drop = FALSE
    ]),
    collapse = ","
  ),
  sep = "|"
)

rebuild_panel_c_sample <- isTRUE(params$run_panel_c_rebuild) ||
  !file.exists(panel_c_sample_logmu_rds)

if (!rebuild_panel_c_sample) {
  panel_c_sample_cache <- readRDS(panel_c_sample_logmu_rds)
  if (is.list(panel_c_sample_cache) &&
      identical(panel_c_sample_cache$cache_id, panel_c_sample_cache_id)) {
    panel_c_sample_logmu <- panel_c_sample_cache$data
    message("Loaded panel-c offset-log-rate audit: ", nrow(panel_c_sample_logmu), " rows")
  } else {
    message("Ignoring stale Panel-c sample cache: inputs or policy changed.")
    rebuild_panel_c_sample <- TRUE
  }
}

if (rebuild_panel_c_sample) {
  panel_c_sample_logmu <- lapply(panel_c_gene_order, function(g) {
    fid <- panel_c_feature_map$.feature[match(g, panel_c_feature_map$gene)]
    y <- as.numeric(panel_c_counts[fid, panel_c_sample_meta$col_id])
    panel_c_sample_meta |>
      dplyr::mutate(
        gene = g,
        .feature = fid,
        observed_count = y,
        # Diagnostic only: this transformed observation is not an estimate of
        # latent NB log-mu. All inferential contrasts below use mglmOneGroup().
        offset_log_rate = log(y + 0.5) - .data$offset
      )
  }) |>
    dplyr::bind_rows()

  saveRDS(
    list(data = panel_c_sample_logmu, cache_id = panel_c_sample_cache_id),
    panel_c_sample_logmu_rds,
    compress = "xz"
  )
}

stopifnot(nrow(panel_c_sample_logmu) == length(panel_c_gene_order) * nrow(panel_c_sample_meta))

panel_c_sample_logmu |>
  dplyr::group_by(.data$group) |>
  dplyr::summarise(
    n_rows = dplyr::n(),
    offset_min = min(.data$offset),
    offset_max = max(.data$offset),
    log_rate_min = min(.data$offset_log_rate),
    log_rate_max = max(.data$offset_log_rate),
    .groups = "drop"
  )
```

## C4. HCA grand-mean log μ draws at offset 0

```{r}
#| label: panel-c-hca-logmu

rebuild_panel_c_hca <- isTRUE(params$run_panel_c_rebuild) ||
  !file.exists(panel_c_hca_logmu_rds)

if (!rebuild_panel_c_hca) {
  panel_c_hca_cache <- readRDS(panel_c_hca_logmu_rds)
  if (is.list(panel_c_hca_cache) &&
      all(c("draws", "grid_id") %in% names(panel_c_hca_cache))) {
    panel_c_hca_draws <- panel_c_hca_cache$draws
    cached_grid_id <- panel_c_hca_cache$grid_id
  } else {
    panel_c_hca_draws <- NULL
    cached_grid_id <- NULL
  }
  if (is.null(panel_c_hca_draws) ||
      !all(c("gene", ".draw", "hca_log_mu") %in% names(panel_c_hca_draws)) ||
      !setequal(unique(panel_c_hca_draws$gene), panel_c_gene_order) ||
      !identical(cached_grid_id, panel_c_hca_grid_id)) {
    message("Ignoring Panel-c HCA cache: comprehensive grid policy changed.")
    rebuild_panel_c_hca <- TRUE
  } else if (abs(
    stats::median(panel_c_hca_draws$hca_log_mu) -
      stats::median(panel_c_sample_logmu$offset_log_rate)
  ) >= 8) {
    message(
      "HCA/sample scale mismatch; rebuilding HCA draws."
    )
    rebuild_panel_c_hca <- TRUE
  } else {
    message("Loaded panel-c HCA draws: ", nrow(panel_c_hca_draws), " rows")
  }
}

if (rebuild_panel_c_hca) {
  stopifnot(file.exists(params$estimates_rds))
  panel_c_estimates <- readRDS(params$estimates_rds)
  stopifnot(all(panel_c_gene_order %in% panel_c_estimates$gene))

  hca_list <- vector("list", length(panel_c_gene_order))
  names(hca_list) <- panel_c_gene_order
  panel_c_hca_grid_n <- NA_integer_
  for (g in panel_c_gene_order) {
    message("HCA linpred ", g, " ...")
    fit <- panel_c_estimates$brms_fit[[match(g, panel_c_estimates$gene)]]
    nd <- build_hca_covariate_grid(
      fit,
      age_sex_target = panel_b_hca_comprehensive_meta
    )
    panel_c_hca_grid_n <- nrow(nd)
    stopifnot(abs(sum(nd$profile_weight) - 1) < 1e-12)
    lp <- brms::posterior_linpred(
      fit,
      newdata = nd,
      transform = FALSE,
      re_formula = NULL,
      allow_new_levels = TRUE,
      sample_new_levels = "gaussian"
    )
    # Weighted g-computation on E_grid[log(mu)]. This is a mean log-rate
    # (geometric-mean mu) estimand, not log(E_grid[mu]).
    gm <- as.numeric(lp %*% nd$profile_weight)
    hca_list[[g]] <- tibble::tibble(
      gene = g,
      .draw = seq_along(gm),
      hca_log_mu = gm
    )
  }
  panel_c_hca_draws <- dplyr::bind_rows(hca_list)
  saveRDS(
    list(
      draws = panel_c_hca_draws,
      grid_id = panel_c_hca_grid_id,
      n_grid_profiles = panel_c_hca_grid_n,
      estimand = "weighted mean log(mu); Panel-b age-sex target"
    ),
    panel_c_hca_logmu_rds,
    compress = "xz"
  )
}

tibble::tibble(
  grid_id = panel_c_hca_grid_id,
  n_draws_total = nrow(panel_c_hca_draws)
) |>
  print()

panel_c_hca_draws |>
  dplyr::group_by(.data$gene) |>
  dplyr::summarise(
    n_draws = dplyr::n(),
    median_log_mu = stats::median(.data$hca_log_mu),
    .groups = "drop"
  ) |>
  dplyr::mutate(gene = factor(.data$gene, levels = panel_c_gene_order)) |>
  dplyr::arrange(.data$gene)
```

## C5. Subject-coupled NB contrasts and 16-gene aggregate

```{r}
#| label: panel-c-effects

panel_c_n_boot <- 2000L
panel_c_effects_cache_id <- paste(
  "v5_subject_coupled_nb_bayesian_bootstrap_independent_hca_gene_draws",
  paste0("n_boot=", panel_c_n_boot),
  paste0("sample=", panel_c_sample_cache_id),
  paste0("hca=", panel_c_hca_grid_id),
  "dispersion=edgeR_trended_all_shared_genes",
  "repeat_libraries=sum_counts_and_exposure",
  sep = "|"
)

rebuild_panel_c_effects <- isTRUE(params$run_panel_c_rebuild) ||
  isTRUE(rebuild_panel_c_sample) ||
  isTRUE(rebuild_panel_c_hca) ||
  !file.exists(panel_c_effects_rds)

if (!rebuild_panel_c_effects) {
  panel_c_effects_cache <- readRDS(panel_c_effects_rds)
  if (is.list(panel_c_effects_cache) &&
      identical(panel_c_effects_cache$cache_id, panel_c_effects_cache_id) &&
      is.data.frame(panel_c_effects_cache$data)) {
    panel_c_effects <- panel_c_effects_cache$data
    message("Loaded panel-c effects: ", nrow(panel_c_effects), " rows")
  } else {
    message("Ignoring stale Panel-c effects cache: inputs or policy changed.")
    rebuild_panel_c_effects <- TRUE
  }
}

if (rebuild_panel_c_effects) {
  panel_c_signature_ids <- panel_c_feature_map$.feature[
    match(panel_c_gene_order, panel_c_feature_map$gene)
  ]
  stopifnot(!anyNA(panel_c_signature_ids), !anyDuplicated(panel_c_signature_ids))

  log_sum_exp <- function(x) {
    m <- max(x)
    m + log(sum(exp(x - m)))
  }

  # One independent column per subject×state. For repeated libraries, add raw
  # counts and add their model exposures exp(offset) before returning to log scale.
  collapse_subject_state <- function(counts, meta) {
    keys <- meta |>
      dplyr::distinct(.data$subject_id, .data$group) |>
      dplyr::arrange(.data$group, .data$subject_id) |>
      dplyr::mutate(sample_key = paste(.data$subject_id, .data$group, sep = "__"))

    collapsed_counts <- do.call(cbind, lapply(seq_len(nrow(keys)), function(i) {
      idx <- which(
        meta$subject_id == keys$subject_id[[i]] &
          meta$group == keys$group[[i]]
      )
      rowSums(counts[, meta$col_id[idx], drop = FALSE])
    }))
    colnames(collapsed_counts) <- keys$sample_key
    rownames(collapsed_counts) <- rownames(counts)

    keys$offset <- vapply(seq_len(nrow(keys)), function(i) {
      idx <- which(
        meta$subject_id == keys$subject_id[[i]] &
          meta$group == keys$group[[i]]
      )
      log_sum_exp(meta$offset[idx])
    }, numeric(1))

    list(counts = collapsed_counts, meta = keys)
  }

  panel_c_subject_data <- collapse_subject_state(
    panel_c_counts,
    panel_c_sample_meta
  )

  # Estimate NB dispersions with all shared genes, not just the selected 16.
  fit_panel_c_dispersion <- function(counts, meta, design) {
    dge <- edgeR::DGEList(counts = counts, samples = meta)
    dge$offset <- matrix(
      meta$offset,
      nrow = nrow(dge),
      ncol = ncol(dge),
      byrow = TRUE
    )
    keep <- edgeR::filterByExpr(dge, design = design)
    signature_idx <- match(panel_c_signature_ids, rownames(dge))
    stopifnot(!anyNA(signature_idx))
    keep[signature_idx] <- TRUE
    dge <- dge[keep, , keep.lib.sizes = FALSE]
    dge <- edgeR::estimateDisp(dge, design, robust = TRUE)
    fit <- edgeR::glmQLFit(dge, design, robust = TRUE)
    dispersion <- if (!is.null(dge$trended.dispersion)) {
      dge$trended.dispersion
    } else if (!is.null(dge$tagwise.dispersion)) {
      dge$tagwise.dispersion
    } else {
      rep_len(dge$common.dispersion, nrow(dge))
    }
    names(dispersion) <- rownames(dge)
    list(dge = dge, fit = fit, dispersion = dispersion)
  }

  subject_counts <- panel_c_subject_data$counts
  subject_meta <- panel_c_subject_data$meta

  uc_meta <- subject_meta |>
    dplyr::filter(.data$group %in% c("control", "untreated")) |>
    dplyr::mutate(group = factor(.data$group, levels = c("control", "untreated")))
  uc_counts <- subject_counts[, uc_meta$sample_key, drop = FALSE]
  uc_design <- stats::model.matrix(~ 0 + group, data = uc_meta)
  colnames(uc_design) <- sub("^group", "", colnames(uc_design))
  uc_fit <- fit_panel_c_dispersion(uc_counts, uc_meta, uc_design)
  uc_dispersion <- unname(uc_fit$dispersion[panel_c_signature_ids])

  paired_subjects <- intersect(
    subject_meta$subject_id[subject_meta$group == "untreated"],
    subject_meta$subject_id[subject_meta$group == "treated"]
  )
  paired_meta <- subject_meta |>
    dplyr::filter(
      .data$subject_id %in% paired_subjects,
      .data$group %in% c("untreated", "treated")
    ) |>
    dplyr::mutate(
      subject_id = factor(.data$subject_id, levels = sort(paired_subjects)),
      state = factor(.data$group, levels = c("treated", "untreated"))
    ) |>
    dplyr::arrange(.data$subject_id, .data$state)
  paired_counts <- subject_counts[, paired_meta$sample_key, drop = FALSE]
  paired_design <- stats::model.matrix(~ subject_id + state, data = paired_meta)
  paired_fit <- fit_panel_c_dispersion(paired_counts, paired_meta, paired_design)
  paired_dispersion <- unname(paired_fit$dispersion[panel_c_signature_ids])

  signature_counts_uc <- uc_counts[panel_c_signature_ids, , drop = FALSE]
  signature_counts_paired <- paired_counts[panel_c_signature_ids, , drop = FALSE]

  mglm_signature <- function(counts, offsets, dispersion, weights = NULL) {
    as.numeric(edgeR::mglmOneGroup(
      counts,
      offset = offsets,
      dispersion = dispersion,
      weights = weights
    ))
  }

  draw_dirichlet_weights <- function(n) {
    w <- stats::rexp(n)
    n * w / sum(w)
  }

  # A common subject-weight vector is used for every gene in a replicate. This
  # is the key coupling that the previous gene-bootstrap aggregate did not
  # preserve. Continuous positive weights also avoid infinite MLEs caused by a
  # discrete bootstrap replicate containing only zero-count subjects.
  bayes_bootstrap_unpaired <- function(n_boot, seed) {
    idx_u <- which(uc_meta$group == "untreated")
    idx_c <- which(uc_meta$group == "control")
    out_u <- matrix(NA_real_, nrow = n_boot, ncol = length(panel_c_gene_order))
    out_c <- out_u
    set.seed(seed)
    for (b in seq_len(n_boot)) {
      wu <- draw_dirichlet_weights(length(idx_u))
      wc <- draw_dirichlet_weights(length(idx_c))
      out_u[b, ] <- mglm_signature(
        signature_counts_uc[, idx_u, drop = FALSE],
        uc_meta$offset[idx_u],
        uc_dispersion,
        weights = wu
      )
      out_c[b, ] <- mglm_signature(
        signature_counts_uc[, idx_c, drop = FALSE],
        uc_meta$offset[idx_c],
        uc_dispersion,
        weights = wc
      )
    }
    colnames(out_u) <- panel_c_gene_order
    colnames(out_c) <- panel_c_gene_order
    list(untreated = out_u, control = out_c)
  }

  bayes_bootstrap_paired <- function(n_boot, seed) {
    key <- paste(as.character(paired_meta$subject_id), paired_meta$group, sep = "__")
    out <- matrix(NA_real_, nrow = n_boot, ncol = length(panel_c_gene_order))
    set.seed(seed)
    for (b in seq_len(n_boot)) {
      w <- draw_dirichlet_weights(length(paired_subjects))
      iu <- match(paste(paired_subjects, "untreated", sep = "__"), key)
      it <- match(paste(paired_subjects, "treated", sep = "__"), key)
      u <- mglm_signature(
        signature_counts_paired[, iu, drop = FALSE],
        paired_meta$offset[iu],
        paired_dispersion,
        weights = w
      )
      t <- mglm_signature(
        signature_counts_paired[, it, drop = FALSE],
        paired_meta$offset[it],
        paired_dispersion,
        weights = w
      )
      out[b, ] <- u - t
    }
    colnames(out) <- panel_c_gene_order
    out
  }

  uc_boot <- bayes_bootstrap_unpaired(panel_c_n_boot, seed = 20260809L)
  paired_boot <- bayes_bootstrap_paired(panel_c_n_boot, seed = 20260810L)

  hca_matrix <- panel_c_hca_draws |>
    dplyr::select(.data$.draw, .data$gene, .data$hca_log_mu) |>
    tidyr::pivot_wider(names_from = .data$gene, values_from = .data$hca_log_mu) |>
    dplyr::arrange(.data$.draw)
  stopifnot(all(panel_c_gene_order %in% names(hca_matrix)))
  hca_draw_id <- hca_matrix$.draw
  hca_matrix <- as.matrix(hca_matrix[, panel_c_gene_order, drop = FALSE])

  # Each gene was fitted in a separate brms model. Equal numeric draw indices
  # therefore are not joint posterior draws and must not be used to manufacture
  # cross-gene dependence. Sample one posterior draw independently per gene,
  # while retaining the genuine across-gene coupling in each SAVI Bayesian-
  # bootstrap draw.
  set.seed(20260811L)
  hca_idx <- matrix(
    sample.int(
      nrow(hca_matrix),
      panel_c_n_boot * length(panel_c_gene_order),
      replace = TRUE
    ),
    nrow = panel_c_n_boot,
    ncol = length(panel_c_gene_order),
    dimnames = list(NULL, panel_c_gene_order)
  )
  hca_pair <- vapply(seq_along(panel_c_gene_order), function(j) {
    hca_matrix[hca_idx[, j], j]
  }, numeric(panel_c_n_boot))
  colnames(hca_pair) <- panel_c_gene_order

  delta_matrices <- list(
    "untreated - study control" =
      (uc_boot$untreated - uc_boot$control) / log(2),
    "untreated - HCA" =
      (uc_boot$untreated - hca_pair) / log(2),
    "paired untreated - treated" = paired_boot / log(2),
    "study control - HCA" =
      (uc_boot$control - hca_pair) / log(2)
  )

  n_untreated <- sum(uc_meta$group == "untreated")
  n_control <- sum(uc_meta$group == "control")
  n_paired <- length(paired_subjects)
  n_hca_donors <- panel_b_hca_comprehensive_n_donors

  effects <- dplyr::bind_rows(
    summarise_panel_c_draw_matrix(
      delta_matrices[["untreated - study control"]],
      panel_c_gene_order,
      "untreated - study control",
      "subject-level Bayesian bootstrap of NB cohort locations",
      n_untreated,
      n_control
    ),
    summarise_panel_c_draw_matrix(
      delta_matrices[["untreated - HCA"]],
      panel_c_gene_order,
      "untreated - HCA",
      "subject-level Bayesian bootstrap independently paired with HCA posterior",
      n_untreated,
      n_hca_donors
    ),
    summarise_panel_c_draw_matrix(
      delta_matrices[["paired untreated - treated"]],
      panel_c_gene_order,
      "paired untreated - treated",
      "paired-subject Bayesian bootstrap of NB cohort locations",
      n_paired,
      n_paired
    ),
    summarise_panel_c_draw_matrix(
      delta_matrices[["study control - HCA"]],
      panel_c_gene_order,
      "study control - HCA",
      "subject-level Bayesian bootstrap independently paired with HCA posterior",
      n_control,
      n_hca_donors
    )
  ) |>
    dplyr::group_by(.data$contrast) |>
    dplyr::mutate(q_value = stats::p.adjust(.data$tail_p, method = "BH")) |>
    dplyr::ungroup()

  aggregate_matrices <- lapply(delta_matrices, function(x) {
    matrix(rowMeans(x), ncol = 1L)
  })
  agg <- dplyr::bind_rows(
    summarise_panel_c_draw_matrix(
      aggregate_matrices[["untreated - study control"]],
      "16-gene covered subset",
      "untreated - study control",
      "subject-coupled 16-gene Bayesian bootstrap",
      n_untreated,
      n_control,
      is_aggregate = TRUE
    ),
    summarise_panel_c_draw_matrix(
      aggregate_matrices[["untreated - HCA"]],
      "16-gene covered subset",
      "untreated - HCA",
      "subject-coupled Bayesian bootstrap independently paired with HCA posterior",
      n_untreated,
      n_hca_donors,
      is_aggregate = TRUE
    ),
    summarise_panel_c_draw_matrix(
      aggregate_matrices[["paired untreated - treated"]],
      "16-gene covered subset",
      "paired untreated - treated",
      "paired-subject-coupled 16-gene Bayesian bootstrap",
      n_paired,
      n_paired,
      is_aggregate = TRUE
    ),
    summarise_panel_c_draw_matrix(
      aggregate_matrices[["study control - HCA"]],
      "16-gene covered subset",
      "study control - HCA",
      "subject-coupled Bayesian bootstrap independently paired with HCA posterior",
      n_control,
      n_hca_donors,
      is_aggregate = TRUE
    )
  ) |>
    dplyr::mutate(q_value = stats::p.adjust(.data$tail_p, method = "BH"))

  panel_c_effects <- dplyr::bind_rows(agg, effects) |>
    dplyr::mutate(
      gene = factor(
        .data$gene,
        levels = c("16-gene covered subset", panel_c_gene_order)
      ),
      contrast = factor(.data$contrast, levels = panel_c_contrast_levels)
    )

  panel_c_effects_cache <- list(
    data = panel_c_effects,
    cache_id = panel_c_effects_cache_id,
    paired_subjects = paired_subjects,
    hca_draw_ids_sampled = matrix(
      hca_draw_id[hca_idx],
      nrow = panel_c_n_boot,
      ncol = length(panel_c_gene_order),
      dimnames = list(NULL, panel_c_gene_order)
    )
  )
  saveRDS(panel_c_effects_cache, panel_c_effects_rds, compress = "xz")
}

panel_c_effects <- panel_c_effects |>
  dplyr::mutate(
    gene = factor(
      as.character(.data$gene),
      levels = c("16-gene covered subset", panel_c_gene_order)
    ),
    contrast = factor(
      as.character(.data$contrast),
      levels = panel_c_contrast_levels
    ),
    contrast_lab = factor(
      unname(panel_c_contrast_labs[as.character(.data$contrast)]),
      levels = unname(panel_c_contrast_labs)
    )
  )

stopifnot(nrow(panel_c_effects) == 4L * (1L + length(panel_c_gene_order)))
stopifnot(all(panel_c_gene_order %in% as.character(panel_c_effects$gene)))
stopifnot(all(is.finite(unlist(panel_c_effects[c(
  "estimate", ".lower", ".upper", "tail_p", "q_value"
)]))))

panel_c_effects |>
  dplyr::filter(.data$is_aggregate) |>
  dplyr::select(
    .data$contrast, .data$estimate, .data$.lower, .data$.upper,
    .data$tail_p, .data$q_value, .data$n_case, .data$n_reference
  )
```

## C6. Plot the four aligned contrasts

```{r}
#| label: panel-c-helpers

plot_panel_c_logmu <- function(effects_df) {
  gene_levels <- levels(effects_df$gene)
  agg_df <- effects_df |> dplyr::filter(.data$is_aggregate)
  gene_df <- effects_df |> dplyr::filter(!.data$is_aggregate)
  x_rng <- range(c(effects_df$.lower, effects_df$.upper), finite = TRUE)
  x_pad <- diff(x_rng) * 0.05

  ggplot2::ggplot(effects_df, ggplot2::aes(y = .data$gene)) +
    ggplot2::geom_vline(
      xintercept = 0,
      colour = col_mid,
      linetype = "22",
      linewidth = 0.35
    ) +
    ggplot2::geom_linerange(
      data = gene_df,
      ggplot2::aes(xmin = .data$.lower, xmax = .data$.upper, x = .data$estimate),
      colour = col_dark,
      linewidth = 0.35
    ) +
    ggplot2::geom_point(
      data = gene_df,
      ggplot2::aes(x = .data$estimate),
      shape = 16,
      size = 1.35,
      colour = col_dark
    ) +
    ggplot2::geom_linerange(
      data = agg_df,
      ggplot2::aes(xmin = .data$.lower, xmax = .data$.upper, x = .data$estimate),
      colour = col_untreated,
      linewidth = 0.55
    ) +
    ggplot2::geom_point(
      data = agg_df,
      ggplot2::aes(x = .data$estimate),
      shape = 23,
      size = 2.2,
      fill = col_untreated,
      colour = col_dark,
      stroke = 0.4
    ) +
    ggplot2::facet_grid(cols = ggplot2::vars(.data$contrast_lab)) +
    ggplot2::scale_y_discrete(limits = rev(gene_levels)) +
    ggplot2::coord_cartesian(
      xlim = c(x_rng[1] - x_pad, x_rng[2] + x_pad),
      clip = "off"
    ) +
    ggplot2::labs(
      x = expression(log[2] ~ "fold change (offset-centred" ~ mu * ")"),
      y = NULL
    ) +
    theme_multipanel +
    ggplot2::theme(
      strip.text.x = ggplot2::element_text(size = 6),
      axis.text.y = ggplot2::element_text(size = 5.5),
      panel.spacing.x = grid::unit(0.4, "lines"),
      plot.margin = margin(4, 6, 4, 2)
    )
}
```

```{r}
#| label: panel-c-draw
#| fig-width: 7.2
#| fig-height: 4.8
#| fig-cap: "Figure 2c. Covered 16-gene STING signature on the offset-zero NB log-μ scale. Top diamond: equal-weight signature effect calculated inside each subject-coupled replicate. Lines are 95% subject-level Bayesian-bootstrap intervals for study contrasts and Bayesian-bootstrap × HCA-posterior hybrid intervals for HCA contrasts. Columns: untreated versus study control, untreated versus comprehensive HCA, paired untreated versus treated, and study control versus HCA. Population: disease-associated monocytes (cluster 17)."

panel_c <- plot_panel_c_logmu(panel_c_effects)
panel_c


```

## C7. Calibration-gated result

```{r}
#| label: panel-c-result-text
#| include: false

panel_c_aggregate_row <- function(contrast_name) {
  panel_c_effects |>
    dplyr::filter(
      .data$is_aggregate,
      as.character(.data$contrast) == contrast_name
    ) |>
    dplyr::slice(1)
}

panel_c_format_result <- function(x) {
  sprintf(
    "%.2f (95%% interval %.2f to %.2f; q = %s)",
    x$estimate,
    x$.lower,
    x$.upper,
    format.pval(x$q_value, digits = 2, eps = 0.001)
  )
}

panel_c_result_calibration <- panel_c_aggregate_row("study control - HCA")
panel_c_result_primary <- panel_c_aggregate_row("untreated - study control")
panel_c_result_hca <- panel_c_aggregate_row("untreated - HCA")
panel_c_result_treatment <- panel_c_aggregate_row("paired untreated - treated")
```

**Reading Panel c.** Interpret study control − HCA first. If that calibration
contrast remains shifted, untreated − HCA contains residual reference-domain
shift and is a sensitivity analysis; untreated − study control remains the
primary disease contrast. The paired treatment contrast is supported only when
its subject-coupled aggregate interval excludes zero. Gene-wise `q_value` is
an exploratory BH adjustment of empirical two-sided sign-tail probabilities
across the 16 genes within each contrast; aggregate values use the four
prespecified signature contrasts. These are modular posterior-tail summaries,
not likelihood-ratio p-values with a frequentist FDR guarantee.

**Current result.** Calibration does not pass on the expression scale: the
study-control − HCA aggregate is
`r panel_c_format_result(panel_c_result_calibration)`, showing that the
comprehensive HCA reference is systematically lower than the SAVI study controls
for this signature. The primary untreated − study-control contrast is
`r panel_c_format_result(panel_c_result_primary)`; the larger untreated − HCA
contrast, `r panel_c_format_result(panel_c_result_hca)`, therefore contains both
disease-associated signal and reference-domain shift. The paired untreated −
treated contrast is `r panel_c_format_result(panel_c_result_treatment)`, consistent
with attenuation after treatment in these four paired subjects, but it is not a
population-wide treatment-efficacy claim.

## C8. Audit and artifact paths

```{r}
#| label: panel-c-audit

panel_c_agg <- panel_c_effects |> dplyr::filter(.data$is_aggregate)

panel_c_direction <- panel_c_effects |>
  dplyr::filter(!.data$is_aggregate) |>
  dplyr::mutate(
    direction = dplyr::case_when(
      .data$q_value <= 0.05 & .data$estimate > 0 ~ "positive",
      .data$q_value <= 0.05 & .data$estimate < 0 ~ "negative",
      TRUE ~ "unresolved"
    )
  ) |>
  dplyr::count(.data$contrast, .data$direction) |>
  tidyr::pivot_wider(
    names_from = .data$direction,
    values_from = .data$n,
    values_fill = 0L
  )

list(
  aggregate = panel_c_agg |>
    dplyr::select(
      .data$contrast, .data$estimate, .data$.lower, .data$.upper,
      .data$tail_p, .data$q_value, .data$n_case, .data$n_reference,
      .data$uncertainty_method
    ),
  constituent_directions = panel_c_direction,
  artifacts = tibble::tibble(
    artifact = c("se_monocytic", "sample_log_rate_audit", "hca_draws", "effects"),
    path = c(
      panel_c_se_rds,
      panel_c_sample_logmu_rds,
      panel_c_hca_logmu_rds,
      panel_c_effects_rds
    ),
    exists = file.exists(c(
      panel_c_se_rds,
      panel_c_sample_logmu_rds,
      panel_c_hca_logmu_rds,
      panel_c_effects_rds
    ))
  )
)
```

------------------------------------------------------------------------

# Panel f — Subject and family influence

**Purpose.** Ask whether Panel c's two main signature estimands
(untreated − comprehensive HCA; untreated − study control) depend on one or a
few of the 12 biological subjects. HCA draws stay fixed; only the SAVI target
side is refit after each deletion.

**Method.** Collapse libraries to one subject×state column (sum counts; sum
exposures). For each deletion, recompute gene-wise untreated and control
locations with `edgeR::mglmOneGroup`, then form log2 contrasts. Deleted
medians are the leave-one-out point estimates. Full-data diamonds and intervals
come from Panel c. Gene rows on the main plot are the 16-gene aggregate plus up
to five constituents selected by deletion range \(R_g\), preferring genes whose
direction category flips.

## F1. Subjects and deletion sets

```{r}
#| label: panel-f-subjects

# Harmonise control IDs (C8/C9) to short display names used in the cohort table.
panel_f_meta <- panel_c_sample_meta |>
  dplyr::mutate(
    person = dplyr::case_when(
      grepl("^P1", .data$sample_id) ~ "P1",
      grepl("^P2", .data$sample_id) ~ "P2",
      grepl("^P4", .data$sample_id) ~ "P4",
      grepl("^P5", .data$sample_id) ~ "P5",
      grepl("^P6", .data$sample_id) ~ "P6",
      grepl("^C8", .data$sample_id) ~ "C8",
      grepl("^C9", .data$sample_id) ~ "C9",
      TRUE ~ .data$sample_id
    )
  )

panel_f_people <- sort(unique(panel_f_meta$person))

# Named deletion sets: 12 single-person deletions + two family sets (audit).
panel_f_deletions <- c(
  stats::setNames(lapply(panel_f_people, identity), panel_f_people),
  list(
    `C8+C9` = c("C8", "C9"),
    `P1/P2/C8/C9` = c("P1", "P2", "C8", "C9")
  )
)

tibble::tibble(
  n_people = length(panel_f_people),
  people = paste(panel_f_people, collapse = ", "),
  n_deletion_sets = length(panel_f_deletions)
)
```

## F2. Subject×state counts for the 16 genes

```{r}
#| label: panel-f-subject-counts

panel_f_feature_ids <- panel_c_feature_map$.feature[
  match(panel_c_gene_order, panel_c_feature_map$gene)
]

# One column per person × group; repeated libraries are added on the count and
# exposure scales (same rule as Panel c).
panel_f_keys <- panel_f_meta |>
  dplyr::distinct(.data$person, .data$group) |>
  dplyr::arrange(.data$group, .data$person)

panel_f_counts <- sapply(seq_len(nrow(panel_f_keys)), function(i) {
  idx <- which(
    panel_f_meta$person == panel_f_keys$person[[i]] &
      panel_f_meta$group == panel_f_keys$group[[i]]
  )
  rowSums(panel_c_counts[panel_f_feature_ids, panel_f_meta$col_id[idx], drop = FALSE])
})
colnames(panel_f_counts) <- paste(
  panel_f_keys$person, panel_f_keys$group, sep = "__"
)
rownames(panel_f_counts) <- panel_c_gene_order

panel_f_keys$offset <- sapply(seq_len(nrow(panel_f_keys)), function(i) {
  idx <- which(
    panel_f_meta$person == panel_f_keys$person[[i]] &
      panel_f_meta$group == panel_f_keys$group[[i]]
  )
  off <- panel_f_meta$offset[idx]
  max(off) + log(sum(exp(off - max(off))))
})

panel_f_hca_median <- panel_c_hca_draws |>
  dplyr::group_by(.data$gene) |>
  dplyr::summarise(hca_log_mu = stats::median(.data$hca_log_mu), .groups = "drop")

tibble::tibble(
  n_genes = nrow(panel_f_counts),
  n_subject_state = ncol(panel_f_counts),
  n_untreated = sum(panel_f_keys$group == "untreated"),
  n_control = sum(panel_f_keys$group == "control")
)
```

## F3. Leave-one-out log2 effects (HCA fixed)

```{r}
#| label: panel-f-leave-one-out

panel_f_group_logmu <- function(keep_people, group_name) {
  hit <- panel_f_keys$person %in% keep_people & panel_f_keys$group == group_name
  y <- panel_f_counts[, hit, drop = FALSE]
  off <- panel_f_keys$offset[hit]
  vapply(panel_c_gene_order, function(g) {
    as.numeric(edgeR::mglmOneGroup(
      matrix(y[g, ], nrow = 1L),
      offset = off,
      dispersion = 0.05
    ))
  }, numeric(1))
}

panel_f_effects_one <- function(deleted, label) {
  keep <- setdiff(panel_f_people, deleted)
  log_u <- panel_f_group_logmu(keep, "untreated")
  log_c <- panel_f_group_logmu(keep, "control")
  hca <- panel_f_hca_median$hca_log_mu[
    match(panel_c_gene_order, panel_f_hca_median$gene)
  ]

  dplyr::bind_rows(
    tibble::tibble(
      gene = panel_c_gene_order,
      contrast = "untreated - HCA",
      estimate = (log_u - hca) / log(2),
      deletion = label
    ),
    tibble::tibble(
      gene = panel_c_gene_order,
      contrast = "untreated - study control",
      estimate = (log_u - log_c) / log(2),
      deletion = label
    )
  )
}

# Full-data point estimates use the same mglm rule (for deletion comparisons).
panel_f_full_points <- panel_f_effects_one(character(0), "full")

# Single-person deletions for the main panel.
panel_f_loo <- lapply(panel_f_people, function(p) {
  panel_f_effects_one(p, p)
}) |>
  dplyr::bind_rows()

# Family deletions for the audit table only.
panel_f_family <- dplyr::bind_rows(
  panel_f_effects_one(c("C8", "C9"), "C8+C9"),
  panel_f_effects_one(c("P1", "P2", "C8", "C9"), "P1/P2/C8/C9")
)

panel_f_loo |>
  dplyr::group_by(.data$contrast) |>
  dplyr::summarise(
    n_rows = dplyr::n(),
    n_deletions = dplyr::n_distinct(.data$deletion),
    .groups = "drop"
  )
```

## F4. Influence scores and gene selection

```{r}
#| label: panel-f-select-genes

panel_f_direction <- function(estimate, .lower = NA_real_, .upper = NA_real_) {
  dplyr::case_when(
    is.finite(.lower) & is.finite(.upper) & .lower > 0 ~ "positive",
    is.finite(.lower) & is.finite(.upper) & .upper < 0 ~ "negative",
    is.finite(estimate) & estimate > 0 ~ "positive",
    is.finite(estimate) & estimate < 0 ~ "negative",
    TRUE ~ "unresolved"
  )
}

# Full-data intervals from Panel c (same two estimands).
panel_f_full <- panel_c_effects |>
  dplyr::filter(
    !.data$is_aggregate,
    .data$contrast %in% c("untreated - HCA", "untreated - study control")
  ) |>
  dplyr::transmute(
    gene = as.character(.data$gene),
    contrast = as.character(.data$contrast),
    full_estimate = .data$estimate,
    .lower = .data$.lower,
    .upper = .data$.upper,
    full_direction = panel_f_direction(.data$estimate, .data$.lower, .data$.upper)
  )

panel_f_agg_full <- panel_c_effects |>
  dplyr::filter(
    .data$is_aggregate,
    .data$contrast %in% c("untreated - HCA", "untreated - study control")
  ) |>
  dplyr::transmute(
    gene = "16-gene covered subset",
    contrast = as.character(.data$contrast),
    full_estimate = .data$estimate,
    .lower = .data$.lower,
    .upper = .data$.upper,
    full_direction = panel_f_direction(.data$estimate, .data$.lower, .data$.upper)
  )

panel_f_loo_summary <- panel_f_loo |>
  dplyr::left_join(panel_f_full, by = c("gene", "contrast")) |>
  dplyr::mutate(
    deleted_direction = panel_f_direction(.data$estimate),
    direction_changed = .data$deleted_direction != .data$full_direction
  )

panel_f_gene_scores <- panel_f_loo_summary |>
  dplyr::group_by(.data$gene, .data$contrast) |>
  dplyr::summarise(
    R = max(.data$estimate) - min(.data$estimate),
    any_flip = any(.data$direction_changed),
    .groups = "drop"
  ) |>
  tidyr::pivot_wider(
    names_from = .data$contrast,
    values_from = c(.data$R, .data$any_flip)
  ) |>
  dplyr::mutate(
    R_g = pmax(
      .data$`R_untreated - HCA`,
      .data$`R_untreated - study control`,
      na.rm = TRUE
    ),
    any_flip = .data$`any_flip_untreated - HCA` |
      .data$`any_flip_untreated - study control`
  ) |>
  dplyr::arrange(dplyr::desc(.data$any_flip), dplyr::desc(.data$R_g), .data$gene)

panel_f_selected_genes <- panel_f_gene_scores |>
  dplyr::slice_head(n = 5L) |>
  dplyr::pull(.data$gene)

# Influential deletion label: person whose deleted estimate is farthest from full.
panel_f_influencer <- dplyr::bind_rows(
  panel_f_loo_summary,
  panel_f_loo |>
    dplyr::group_by(.data$contrast, .data$deletion) |>
    dplyr::summarise(
      gene = "16-gene covered subset",
      estimate = mean(.data$estimate),
      .groups = "drop"
    ) |>
    dplyr::left_join(panel_f_agg_full, by = c("gene", "contrast"))
) |>
  dplyr::group_by(.data$gene, .data$contrast) |>
  dplyr::mutate(abs_shift = abs(.data$estimate - .data$full_estimate)) |>
  dplyr::slice_max(.data$abs_shift, n = 1L, with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::transmute(
    .data$gene, .data$contrast,
    influential_deletion = .data$deletion
  )

panel_f_gene_scores |>
  dplyr::select(.data$gene, .data$R_g, .data$any_flip) |>
  dplyr::slice_head(n = 8L)
```

## F5. Plot table (aggregate + selected genes)

```{r}
#| label: panel-f-plot-data

panel_f_plot_genes <- c("16-gene covered subset", panel_f_selected_genes)

# Aggregate leave-one-out: equal-weight mean of the 16 gene effects.
panel_f_loo_agg <- panel_f_loo |>
  dplyr::group_by(.data$contrast, .data$deletion) |>
  dplyr::summarise(
    gene = "16-gene covered subset",
    estimate = mean(.data$estimate),
    .groups = "drop"
  ) |>
  dplyr::left_join(panel_f_agg_full, by = c("gene", "contrast")) |>
  dplyr::mutate(
    deleted_direction = panel_f_direction(.data$estimate),
    direction_changed = .data$deleted_direction != .data$full_direction
  )

panel_f_loo_sel <- panel_f_loo_summary |>
  dplyr::filter(.data$gene %in% panel_f_selected_genes)

panel_f_range <- dplyr::bind_rows(panel_f_loo_agg, panel_f_loo_sel) |>
  dplyr::group_by(.data$gene, .data$contrast) |>
  dplyr::summarise(
    loo_min = min(.data$estimate),
    loo_max = max(.data$estimate),
    n_flip = sum(.data$direction_changed),
    .groups = "drop"
  )

panel_f_full_plot <- dplyr::bind_rows(panel_f_agg_full, panel_f_full) |>
  dplyr::filter(.data$gene %in% panel_f_plot_genes) |>
  dplyr::left_join(panel_f_range, by = c("gene", "contrast")) |>
  dplyr::left_join(panel_f_influencer, by = c("gene", "contrast")) |>
  dplyr::mutate(
    gene = factor(.data$gene, levels = rev(panel_f_plot_genes)),
    contrast_lab = factor(
      dplyr::case_when(
        .data$contrast == "untreated - HCA" ~ "Untreated - HCA",
        .data$contrast == "untreated - study control" ~ "Untreated - CTRL"
      ),
      levels = c("Untreated - HCA", "Untreated - CTRL")
    ),
    row_flip = .data$n_flip > 0
  )

panel_f_points <- dplyr::bind_rows(panel_f_loo_agg, panel_f_loo_sel) |>
  dplyr::filter(.data$gene %in% panel_f_plot_genes) |>
  dplyr::mutate(
    gene = factor(.data$gene, levels = rev(panel_f_plot_genes)),
    contrast_lab = factor(
      dplyr::case_when(
        .data$contrast == "untreated - HCA" ~ "Untreated - HCA",
        .data$contrast == "untreated - study control" ~ "Untreated - CTRL"
      ),
      levels = c("Untreated - HCA", "Untreated - CTRL")
    )
  )

saveRDS(
  list(
    gene_scores = panel_f_gene_scores,
    selected_genes = panel_f_selected_genes,
    full_plot = panel_f_full_plot,
    loo_points = panel_f_points,
    family = panel_f_family
  ),
  panel_f_effects_rds,
  compress = "xz"
)

panel_f_full_plot |>
  dplyr::select(
    .data$gene, .data$contrast_lab, .data$full_estimate,
    .data$loo_min, .data$loo_max, .data$n_flip, .data$influential_deletion
  )
```

## F6. Plot

```{r}
#| label: panel-f-helpers

plot_panel_f_influence <- function(full_df, points_df) {
  ggplot2::ggplot(full_df, ggplot2::aes(y = .data$gene)) +
    ggplot2::geom_vline(
      xintercept = 0,
      colour = col_mid,
      linetype = "22",
      linewidth = 0.35
    ) +
    ggplot2::geom_linerange(
      ggplot2::aes(xmin = .data$loo_min, xmax = .data$loo_max),
      colour = col_light,
      linewidth = 1.1
    ) +
    ggplot2::geom_linerange(
      ggplot2::aes(xmin = .data$.lower, xmax = .data$.upper),
      colour = col_dark,
      linewidth = 0.45
    ) +
    ggplot2::geom_point(
      data = points_df,
      ggplot2::aes(x = .data$estimate),
      shape = 1,
      size = 1.4,
      colour = col_mid,
      stroke = 0.35
    ) +
    ggplot2::geom_point(
      ggplot2::aes(
        x = .data$full_estimate,
        fill = .data$row_flip
      ),
      shape = 23,
      size = 2.2,
      colour = col_dark,
      stroke = 0.4
    ) +
    ggplot2::geom_text(
      ggplot2::aes(
        x = .data$loo_max,
        label = .data$influential_deletion
      ),
      hjust = -0.15,
      size = 1.8,
      colour = col_mid
    ) +
    ggplot2::scale_fill_manual(
      values = c(`FALSE` = col_dark, `TRUE` = col_treated),
      guide = "none"
    ) +
    ggplot2::facet_grid(cols = ggplot2::vars(.data$contrast_lab)) +
    ggplot2::coord_cartesian(clip = "off") +
    ggplot2::labs(
      x = expression(log[2] ~ "fold change"),
      y = NULL
    ) +
    theme_multipanel +
    ggplot2::theme(
      strip.text.x = ggplot2::element_text(size = 6),
      axis.text.y = ggplot2::element_text(size = 5.5),
      plot.margin = margin(4, 18, 4, 2)
    )
}
```

```{r}
#| label: panel-f-draw
#| fig-width: 7.2
#| fig-height: 3.6
#| fig-cap: "Figure 2f. Subject influence on the 16-gene signature. Diamonds and dark lines: full-data Panel-c estimates and intervals. Grey bars: range of leave-one-subject medians. Open points: each single-subject deletion. Purple fill marks at least one direction-category change. Right labels name the most shifting deletion."

panel_f <- plot_panel_f_influence(panel_f_full_plot, panel_f_points)
panel_f


```

## F7. Family-deletion audit

```{r}
#| label: panel-f-audit

panel_f_family_agg <- panel_f_family |>
  dplyr::group_by(.data$contrast, .data$deletion) |>
  dplyr::summarise(
    estimate = mean(.data$estimate),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    panel_f_agg_full |>
      dplyr::select(.data$contrast, full_estimate = .data$full_estimate, .data$full_direction),
    by = "contrast"
  ) |>
  dplyr::mutate(
    deleted_direction = panel_f_direction(.data$estimate),
    direction_changed = .data$deleted_direction != .data$full_direction
  )

list(
  selected_genes = panel_f_selected_genes,
  gene_scores = panel_f_gene_scores |>
    dplyr::select(.data$gene, .data$R_g, .data$any_flip),
  family_aggregate = panel_f_family_agg,
  artifact = panel_f_effects_rds
)
```

------------------------------------------------------------------------

# Full-figure canvas (scaffold)

Panels d–e are blank frames at production size. They will be replaced in place
as analyses land; the canvas size stays **≤ 183 × 220 mm**.

```{r}
#| label: fig2-scaffold-helpers

blank_panel <- function(title, note) {
  ggplot2::ggplot() +
    ggplot2::annotate(
      "rect",
      xmin = 0, xmax = 1, ymin = 0, ymax = 1,
      fill = "#F7F7F7",
      colour = col_light,
      linewidth = 0.4,
      linetype = "22"
    ) +
    ggplot2::annotate(
      "text",
      x = 0.5, y = 0.58,
      label = title,
      size = 2.4,
      fontface = "bold",
      colour = col_dark
    ) +
    ggplot2::annotate(
      "text",
      x = 0.5, y = 0.40,
      label = note,
      size = 2.0,
      colour = col_mid
    ) +
    ggplot2::coord_cartesian(xlim = c(0, 1), ylim = c(0, 1), expand = FALSE) +
    ggplot2::theme_void() +
    ggplot2::theme(plot.margin = margin(4, 4, 4, 4))
}
```

```{r}
#| label: fig2-canvas
#| fig-width: 7.2
#| fig-height: 8.7
#| fig-cap: "Figure 2 canvas (183 × 220 mm). Panels a–c and f populated; d–e awaiting gated analyses."

panel_d <- blank_panel(
  "d  density archetypes",
  "VECTOR DENSITIES AFTER c + f"
)
panel_e <- blank_panel(
  "e  healthy-prediction percentiles",
  "AGE + CELL-ALIGNMENT RERUN NEEDED"
)

fig2 <- (
  patchwork::wrap_elements(full = panel_a)
) / (
  patchwork::wrap_elements(full = panel_b) | panel_c
) / (
  panel_d
) / (
  panel_e | patchwork::wrap_elements(full = panel_f)
) +
  patchwork::plot_layout(heights = c(1.05, 1.55, 1.05, 1.45)) &
  ggplot2::theme(plot.margin = margin(1, 1, 1, 1))

fig2

fig2_pdf <- save_panel_pdf(
  fig2,
  "Fig2_SAVI_manuscript_canvas.pdf",
  width_mm = params$fig_width_mm,
  height_mm = params$fig_height_mm
)
```

```{r}
#| label: output-paths
tibble::tibble(
  artifact = c("fig2_canvas", "fig2f_panel"),
  path = c(fig2_pdf, panel_f_pdf)
)
```

------------------------------------------------------------------------

# Next panels (not drawn yet)

| Panel | Gate |
|---|---|
| d | Gene selection after c + f; native-vector comprehensive / predictive dens |
| e | Corrected ages; metadata-conditioned percentiles for all 17 libraries |

------------------------------------------------------------------------

# Reviewer-facing self-review

| Dimension | Review question | Status and evidence |
|---|---|---|
| Contribution | Does Panel b add information beyond the original study control? | **Pass with boundary.** It evaluates two external HCA targets, but calibration correctly prevents an unqualified disease claim. |
| Writing clarity | Can the analyses be reproduced from the report? | **Pass.** Panel b records the 23- and 26-label maps, 15-part likelihood, target weights, draw pairing, ROPE and FDR family. Panel c records its cohort estimand, repeat-library rule, subject resampling, HCA standardisation and multiplicity families. |
| Experimental strength | Is the external baseline validated against an internal healthy benchmark? | **Pass.** Study-control calibration is reported before disease contrasts; 4/7 comprehensive and 5/7 personalised calibration discoveries are retained as a failure mode. |
| Evaluation completeness | Are important design choices stress-tested? | **Needs evidence.** The 15-part harmonisation is biologically explicit, but the previous coarse residual model should remain a supplementary resolution-sensitivity analysis. Panel c still has only four paired subjects and no joint HCA gene model. |
| Method soundness | Are compositional mass, biological units and cross-model uncertainty handled correctly? | **Pass within scope.** Panel b closes counts over 15 parts and uses 12 independent subjects. Panel c collapses repeat libraries, couples all genes to the same subject resample, and independently samples separate HCA gene posteriors without pretending they form a joint posterior. |

## Panel-b claim–evidence map

| Claim | Evidence | Status |
|---|---|---|
| SAVI and HCA are compared on a shared biological estimand. | Explicit coverage checks map all 23 SAVI labels and all 26 HCA L3 labels to the same 15-part composition. | Supported |
| The comprehensive baseline equal-averages observed age/sex profiles rather than donor-weighting the HCA population. | Each observed age-decade × sex profile receives weight `1 / n_profiles` before response-scale averaging within draws. | Supported |
| Demographic personalisation does not remove reference-domain shift in this case. | Calibration yields 4/7 discoveries for comprehensive HCA and 5/7 for personalised HCA. | Supported |
| HCA contrasts alone establish SAVI disease effects. | Calibration fails; HCA contrasts are therefore labelled sensitivity analyses and the study-control contrast remains primary. | Not supported; claim explicitly rejected |

## Panel-c claim–evidence map

| Claim | Evidence | Status |
|---|---|---|
| The SAVI uncertainty unit is the biological subject. | Repeat libraries are collapsed within subject×state, and every Bayesian-bootstrap replicate uses the same subject weights for all 16 genes. | Supported |
| The signature interval represents uncertainty across subjects, not variability across a fixed gene list. | The 16-gene mean is recalculated inside each subject-coupled Bayesian-bootstrap replicate; genes are not treated as exchangeable bootstrap units. | Supported |
| Panel c uses the same comprehensive demographic target as Panel b. | HCA log-μ draws use Panel b's observed healthy-HCA age×sex profiles and exact target weights, with ethnicity equal-standardised and dataset random effects excluded. | Supported |
| The untreated − HCA interval is a fully Bayesian joint-model interval. | The SAVI side is a Bayesian bootstrap and the 16 HCA genes come from separate fits; the code independently pairs these sources and labels the interval modular hybrid. | Not supported; claim explicitly rejected |
| A treatment-associated attenuation is present in the paired observations. | The four-patient paired aggregate interval excludes zero, while the report explicitly limits the claim to these four subjects. | Supported with a severe sample-size boundary |
