---
title: "TM4SF1 Fig. 3 figure development"
subtitle: "Publication-panel drafts from the corrected v2.2.1 case-study outputs"
date: today
date-format: "DD MMM YYYY"
format:
  html:
    theme: minty
    toc: true
    toc-location: left
    code-fold: true
    code-fold-default: true
    embed-resources: true
    fig-path: results/TM4SF1_Fig3_figure_development/figures/
    html-math-method: mathjax
knitr:
  opts_chunk:
    message: false
    warning: false
    dpi: 300
    fig.retina: 2
    fig.align: center
    fig.width: 7
    fig.height: 5.5
execute:
  freeze: false
editor: source
---

This report develops candidate panels for a provisional full-width manuscript
Fig. 3. It is a plotting layer over the corrected v2.2.1 TM4SF1 report: no
posterior prediction, marginalisation, threshold construction, admissibility
rule, donor-support rule, candidate definition, or tumour reconstruction is
rerun or changed. Panels are deliberately shown separately rather than
assembled.

::: {.callout-important collapse="false"}
The current epithelial and endothelial fits do not pass the registered model-
readiness gate. Overlap is a prioritisation signal, not evidence of toxicity.
Prediction visibility and empirical support remain separate throughout.
:::

```{r}
#| label: setup
#| include: false

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(readr)
  library(readxl)
  library(stringr)
  library(tibble)
  library(ggplot2)
  library(scales)
  library(RColorBrewer)
  library(ggridges)
})

options(stringsAsFactors = FALSE, scipen = 6, bitmapType = "cairo")
set.seed(293189)

case_dir <- normalizePath(".", mustWork = TRUE)
source_results <- file.path(
  case_dir, "results", "tm4sf1_toxicity_report_v2.2.1_audit_cleanup"
)
figure_results <- file.path(case_dir, "results", "TM4SF1_Fig3_figure_development")
figure_dir <- file.path(figure_results, "figures")
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
knitr::opts_chunk$set(
  fig.path = "results/TM4SF1_Fig3_figure_development/figures/"
)

required_files <- c(
  "posteriorHCA_TM4SF1_profile_summary.tsv",
  "TM4SF1_study_logmu_draws.rds",
  "TM4SF1_threshold_registry.tsv",
  "TM4SF1_model_diagnostics.tsv",
  "TM4SF1_paper_DE_replication.tsv",
  "TM4SF1_paper_source_patient_summary.tsv"
)
stopifnot(all(file.exists(file.path(source_results, required_files))))

profile_summary <- read_tsv(
  file.path(source_results, "posteriorHCA_TM4SF1_profile_summary.tsv"),
  show_col_types = FALSE
)
study_draws <- readRDS(file.path(source_results, "TM4SF1_study_logmu_draws.rds"))
threshold_registry <- read_tsv(
  file.path(source_results, "TM4SF1_threshold_registry.tsv"),
  show_col_types = FALSE
)
model_diagnostics <- read_tsv(
  file.path(source_results, "TM4SF1_model_diagnostics.tsv"),
  show_col_types = FALSE
)
replication_de <- read_tsv(
  file.path(source_results, "TM4SF1_paper_DE_replication.tsv"),
  show_col_types = FALSE
)
paper_patient_summary <- read_tsv(
  file.path(source_results, "TM4SF1_paper_source_patient_summary.tsv"),
  show_col_types = FALSE
)

stopifnot(
  all(c("epithelial", "endothelial") %in% profile_summary$cell_type),
  all(!model_diagnostics$model_ready),
  all(profile_summary$baseline_method == "design_standardised_v2.2.1"),
  all(profile_summary$expression_overlap_alert == FALSE)
)

paper_workbook <- file.path(
  case_dir, "data", "raw", "paper", "41467_2025_59888_MOESM4_ESM.xlsx"
)
paper_cells <- read_excel(paper_workbook, sheet = "Figure5B") |>
  transmute(
    barcode = as.character(Barcode),
    paper_id = as.character(Name),
    source_expression = as.numeric(TM4SF1),
    histology_group = case_when(
      str_starts(paper_id, "VAR") ~ "HV",
      str_starts(paper_id, "UC") ~ "UC",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(histology_group))

tm4_patient <- read_tsv(
  file.path(case_dir, "data", "metadata", "TM4SF1_patient_summary.tsv"),
  show_col_types = FALSE
)

paper_cell_patient <- paper_cells |>
  group_by(paper_id, histology_group) |>
  summarise(
    mean_source_expression = mean(source_expression),
    n_cells = n(),
    .groups = "drop"
  )

# Fig. 3 palette: biological state, model identity, and support are separate.
fig3_col <- c(
  HV = "#D55E00",
  UC = "#0072B2",
  Normal = "#009E73",
  epithelial = "#CC79A7",
  endothelial = "#56B4E9",
  neutral = "#6F6F6F",
  faint = "#D4D4D4"
)
sex_col <- c(
  female = brewer.pal(3, "PRGn")[[1]],
  male = brewer.pal(3, "PRGn")[[3]],
  `female (structural)` = brewer.pal(3, "PRGn")[[1]]
)

# Retained verbatim as the visual base used by the corrected analysis report.
theme_multipanel <-
  theme_bw() +
  theme(
    panel.border = element_blank(),
    axis.line = element_line(linewidth = 0.1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    strip.background = element_blank(),
    axis.title.y = element_text(margin = margin(0, 0, 0, 0), size = 12),
    axis.title.x = element_text(margin = margin(0, 0, 0, 0), size = 12),
    panel.spacing.x = grid::unit(0.1, "lines"),
    axis.text.x = element_text(size = 11),
    axis.text.y = element_text(size = 11),
    strip.text.x = element_text(size = 12),
    strip.text.y = element_text(size = 12),
    legend.key.size = grid::unit(5, "mm"),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10),
    strip.clip = "off",
    plot.title = element_blank(),
    plot.subtitle = element_blank(),
    plot.tag = element_text(size = 13, face = "bold"),
    axis.line.x = element_line(linewidth = 0.2),
    axis.line.y = element_line(linewidth = 0.2),
    axis.ticks.x = element_line(linewidth = 0.2),
    axis.ticks.y = element_line(linewidth = 0.2),
    plot.margin = margin(2, 4, 2, 2)
  )

theme_fig3_base <- function() {
  theme_multipanel +
    theme(
      text = element_text(colour = "#2F2F2F"),
      axis.line = element_line(colour = "#4A4A4A", linewidth = 0.22),
      axis.ticks = element_line(colour = "#4A4A4A", linewidth = 0.22),
      axis.ticks.length = grid::unit(1.2, "mm"),
      legend.spacing.x = grid::unit(1, "mm"),
      legend.margin = margin(0, 0, 0, 0)
    )
}

theme_fig3_facet <- function() {
  theme_fig3_base() +
    theme(
      strip.text = element_text(size = 10, lineheight = 0.95),
      panel.spacing = grid::unit(1.8, "mm"),
      axis.text = element_text(size = 10),
      axis.title = element_text(size = 11),
      legend.text = element_text(size = 10)
    )
}

theme_fig3_compact <- function() {
  theme_fig3_base() +
    theme(
      axis.text = element_text(size = 10),
      axis.title = element_text(size = 11),
      legend.text = element_text(size = 10),
      legend.key.height = grid::unit(2.5, "mm"),
      plot.margin = margin(1.5, 2.5, 1.5, 1.5)
    )
}

support_level <- function(n) {
  factor(
    case_when(
      n >= 10 ~ "adequate",
      n >= 1 ~ "limited",
      TRUE ~ "extrapolated"
    ),
    levels = c("adequate", "limited", "extrapolated")
  )
}

support_marker <- function(n) {
  case_when(n >= 10 ~ "●", n >= 1 ~ "○", TRUE ~ "×")
}

publication_tissue <- function(x) {
  recode(
    x,
    "female reproductive system" = "Female reproductive",
    "integumentary system (skin)" = "Skin",
    "brainstem and cerebellar structures" = "Brainstem/cerebellar",
    "cerebral lobes and cortical areas" = "Cerebral/cortical",
    "general brain and major structures" = "General brain",
    "limbic and basal systems" = "Limbic/basal",
    "nasal, oral, and pharyngeal regions" = "Nasal/oral/pharyngeal",
    "muscular system (skeletal muscles)" = "Skeletal muscle",
    "sensory-related structures" = "Sensory-related",
    "digestive system (general)" = "Digestive (general)",
    .default = str_to_sentence(x)
  )
}

publication_ethnicity <- function(x) {
  recode(
    x,
    "Hispanic/Latin American" = "Hispanic/Latin Am.",
    "Middle Eastern & North African" = "Middle Eastern/N. African",
    "Native American & Pacific Islander" = "Native Am./Pacific Isl.",
    .default = x
  )
}

# The cache stores exact medians and 95% intervals, but not the full tissue-level
# draw matrices. This continuous split-Gaussian envelope is a plotting-only
# interpolation through those exact summaries; it is never used to recompute P.
summary_density <- function(median, lower, upper, x) {
  sigma_lower <- max((median - lower) / 1.96, 0.05)
  sigma_upper <- max((upper - median) / 1.96, 0.05)
  z <- ifelse(x <= median, (x - median) / sigma_lower, (x - median) / sigma_upper)
  y <- exp(-0.5 * z^2)
  y / max(y)
}

make_tissue_density_data <- function(cell_type, tissues = NULL) {
  summary <- profile_summary |>
    filter(dimension == "tissue", .data$cell_type == .env$cell_type)
  if (!is.null(tissues)) {
    summary <- filter(summary, tissue_groups %in% tissues)
  }
  summary <- summary |>
    arrange(desc(probability_above_threshold), desc(median_log_mu)) |>
    mutate(
      support_level = support_level(n_unique_donor_keys),
      tissue_display = publication_tissue(tissue_groups),
      facet_label = paste0(
        tissue_display, "\nP=", sprintf("%.2f", probability_above_threshold),
        " · n=", n_unique_donor_keys, " ", support_marker(n_unique_donor_keys)
      ),
      facet_label = factor(facet_label, levels = facet_label)
    )

  hv <- study_draws |>
    filter(
      .data$cell_type == .env$cell_type,
      group == "HV tumour epithelial"
    ) |>
    pull(log_mu)
  x_limits <- range(c(summary$lower_log_mu, summary$upper_log_mu, hv), na.rm = TRUE)
  x_pad <- diff(x_limits) * 0.025
  x_grid <- seq(x_limits[[1]] - x_pad, x_limits[[2]] + x_pad, length.out = 181)

  normal_density <- map_dfr(seq_len(nrow(summary)), function(i) {
    row <- summary[i, ]
    tibble(
      facet_label = row$facet_label,
      x = x_grid,
      density = summary_density(
        row$median_log_mu, row$lower_log_mu, row$upper_log_mu, x_grid
      ),
      support_level = row$support_level
    )
  })

  hv_density_base <- density(
    hv, from = min(x_grid), to = max(x_grid), n = length(x_grid)
  )
  hv_density_base <- tibble(
    x = hv_density_base$x,
    density = hv_density_base$y / max(hv_density_base$y)
  )
  hv_density <- tidyr::crossing(
    facet_label = summary$facet_label,
    hv_density_base
  )

  list(
    summary = summary,
    normal_density = normal_density,
    hv_density = hv_density,
    x_limits = range(x_grid)
  )
}

plot_tissue_density <- function(cell_type, tissues = NULL, ncol = 4) {
  dat <- make_tissue_density_data(cell_type, tissues)
  model_colour <- unname(fig3_col[[cell_type]])

  ggplot() +
    geom_area(
      data = dat$hv_density,
      aes(x, density, fill = "HV tumour anchor"),
      alpha = 0.14, linewidth = 0
    ) +
    geom_line(
      data = dat$hv_density,
      aes(x, density), colour = fig3_col[["HV"]], linewidth = 0.28
    ) +
    geom_area(
      data = dat$normal_density,
      aes(x, density, fill = "Normal posterior", alpha = support_level),
      linewidth = 0
    ) +
    geom_line(
      data = dat$normal_density,
      aes(x, density, linetype = support_level),
      colour = model_colour, linewidth = 0.34
    ) +
    geom_segment(
      data = dat$summary,
      aes(
        x = lower_log_mu, xend = upper_log_mu,
        y = -0.055, yend = -0.055
      ),
      colour = fig3_col[["neutral"]], linewidth = 0.35
    ) +
    geom_point(
      data = dat$summary,
      aes(median_log_mu, -0.055),
      colour = model_colour, fill = "white", shape = 21,
      size = 0.9, stroke = 0.3
    ) +
    geom_vline(
      data = dat$summary,
      aes(xintercept = threshold_log_mu),
      colour = fig3_col[["HV"]], linetype = 3, linewidth = 0.25
    ) +
    facet_wrap(~facet_label, ncol = ncol) +
    coord_cartesian(xlim = dat$x_limits, ylim = c(-0.09, 1.03), clip = "off") +
    scale_fill_manual(
      values = c(
        "Normal posterior" = fig3_col[["Normal"]],
        "HV tumour anchor" = fig3_col[["HV"]]
      ),
      name = NULL
    ) +
    scale_alpha_manual(
      values = c(adequate = 0.42, limited = 0.25, extrapolated = 0.13),
      guide = "none"
    ) +
    scale_linetype_manual(
      values = c(adequate = 1, limited = 2, extrapolated = 3),
      guide = "none"
    ) +
    scale_x_continuous(n.breaks = 3, labels = label_number(accuracy = 1)) +
    labs(x = "Model-aligned latent log(μ)", y = NULL) +
    theme_fig3_facet() +
    theme(
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      axis.line.y = element_blank(),
      legend.position = "bottom",
      legend.key.width = grid::unit(3.3, "mm"),
      plot.margin = margin(4, 2, 2, 1.5)
    )
}

prepare_story_sex <- function(comparator) {
  female_repro <- profile_summary |>
    filter(
      cell_type == "endothelial",
      dimension == "tissue",
      tissue_groups == "female reproductive system"
    ) |>
    mutate(sex = "female (structural)")

  comparator_sex <- profile_summary |>
    filter(
      cell_type == "endothelial",
      dimension == "tissue_sex",
      tissue_groups == comparator
    )

  bind_rows(female_repro, comparator_sex) |>
    mutate(
      support_level = support_level(n_unique_donor_keys),
      tissue_display = factor(
        publication_tissue(tissue_groups),
        levels = rev(publication_tissue(c("female reproductive system", comparator)))
      )
    )
}

plot_story_sex <- function(comparator) {
  dat <- prepare_story_sex(comparator)
  threshold <- unique(dat$threshold_log_mu)
  ggplot(dat, aes(median_log_mu, tissue_display, colour = sex, shape = support_level)) +
    geom_errorbar(
      aes(xmin = lower_log_mu, xmax = upper_log_mu),
      orientation = "y", width = 0.12, linewidth = 0.32,
      position = position_dodge(width = 0.42), alpha = 0.72
    ) +
    geom_point(position = position_dodge(width = 0.42), size = 1.45, stroke = 0.35) +
    geom_vline(
      xintercept = threshold, colour = fig3_col[["HV"]],
      linetype = 3, linewidth = 0.28
    ) +
    scale_colour_manual(
      values = sex_col,
      breaks = c("female", "male", "female (structural)"),
      labels = c("Female", "Male", "Female-only tissue")
    ) +
    scale_shape_manual(values = c(adequate = 16, limited = 1, extrapolated = 4)) +
    labs(x = "Endothelial Normal log(μ)", y = NULL, colour = NULL, shape = NULL) +
    theme_fig3_compact() +
    theme(legend.position = "none")
}

prepare_story_dimension <- function(dimension, comparator) {
  profile_summary |>
    filter(
      cell_type == "endothelial",
      .data$dimension == .env$dimension,
      tissue_groups %in% c("female reproductive system", comparator)
    ) |>
    mutate(
      support_level = support_level(n_unique_donor_keys),
      tissue_display = factor(
        publication_tissue(tissue_groups),
        levels = publication_tissue(c("female reproductive system", comparator))
      )
    )
}

plot_story_age <- function(comparator) {
  dat <- prepare_story_dimension("tissue_age", comparator) |>
    arrange(tissue_display, age_decade)
  ggplot(dat, aes(age_decade, probability_above_threshold, group = tissue_display)) +
    geom_hline(yintercept = 0.8, colour = fig3_col[["neutral"]], linetype = 3, linewidth = 0.24) +
    geom_line(colour = fig3_col[["Normal"]], linewidth = 0.35, alpha = 0.52) +
    geom_point(
      aes(shape = support_level, alpha = support_level),
      colour = fig3_col[["Normal"]], size = 1.25, stroke = 0.32
    ) +
    facet_wrap(~tissue_display, ncol = 1) +
    scale_shape_manual(values = c(adequate = 16, limited = 1, extrapolated = 4)) +
    scale_alpha_manual(values = c(adequate = 1, limited = 0.68, extrapolated = 0.38)) +
    scale_x_continuous(breaks = c(2, 6, 10)) +
    scale_y_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1)) +
    labs(
      x = "Age decade",
      y = "Overlap probability", shape = "Support", alpha = "Support"
    ) +
    theme_fig3_facet() +
    theme(legend.position = "none", strip.text = element_text(hjust = 0))
}

plot_story_ethnicity <- function(comparator) {
  dat <- prepare_story_dimension("tissue_ethnicity", comparator) |>
    group_by(ethnicity_groups) |>
    mutate(order_metric = mean(probability_above_threshold)) |>
    ungroup() |>
    mutate(
      ethnicity_display = publication_ethnicity(ethnicity_groups),
      ethnicity_display = reorder(ethnicity_display, order_metric)
    )
  ggplot(dat, aes(probability_above_threshold, ethnicity_display)) +
    geom_vline(xintercept = 0.8, colour = fig3_col[["neutral"]], linetype = 3, linewidth = 0.24) +
    geom_point(
      aes(shape = support_level, alpha = support_level),
      colour = fig3_col[["Normal"]], size = 1.25, stroke = 0.32
    ) +
    facet_wrap(~tissue_display, ncol = 1, scales = "free_y") +
    scale_shape_manual(values = c(adequate = 16, limited = 1, extrapolated = 4)) +
    scale_alpha_manual(values = c(adequate = 1, limited = 0.68, extrapolated = 0.38)) +
    scale_x_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1)) +
    labs(x = "Overlap probability", y = NULL, shape = "Support", alpha = "Support") +
    theme_fig3_facet() +
    theme(
      legend.position = "none",
      strip.text = element_text(hjust = 0),
      axis.text.y = element_text(size = 8.5)
    )
}
```

# Proposed Fig. 3 layout

The planning canvas is 183 mm × approximately 203 mm: Row 1, 48 mm; Row 2,
92 mm; Row 3, 55 mm; with two approximately 4 mm gutters. HTML chunks are
enlarged for on-screen review; intended print sizes remain in the Size/change
notes.

## Row 1 — establish the case and compartments

### Panel A — released study result

#### Draft A1 — released cell-level source with patient summaries

```{r}
#| label: panel-a-released
#| fig-width: 6.2
#| fig-height: 5.2
#| out-width: 85%
#| fig-cap: "Released Figure 5B TM4SF1 values. Violins show released cell-level values; filled points show patient means."

panel_a_released <- ggplot(
  paper_cells,
  aes(histology_group, source_expression, fill = histology_group)
) +
  geom_violin(
    scale = "width", trim = TRUE, colour = fig3_col[["neutral"]],
    linewidth = 0.22, alpha = 0.58
  ) +
  geom_boxplot(
    width = 0.13, outlier.shape = NA, fill = "white",
    colour = fig3_col[["neutral"]], linewidth = 0.24
  ) +
  geom_point(
    data = paper_cell_patient,
    aes(y = mean_source_expression),
    position = position_jitter(width = 0.055, height = 0),
    shape = 21, fill = "white", colour = fig3_col[["neutral"]],
    size = 1.2, stroke = 0.3
  ) +
  scale_fill_manual(values = fig3_col[c("HV", "UC")]) +
  labs(x = NULL, y = "Released TM4SF1 expression") +
  guides(fill = "none") +
  theme_fig3_compact()

panel_a_released
```

- **Message:** the released study source shows TM4SF1 enrichment in HV tumour epithelial cells.
- **Source:** released Figure 5B values and patient summaries; the released Figure 5A result ranks TM4SF1 first among HV-enriched genes (log2FC 2.36).
- **Size/change:** 57 × 48 mm; replaces a large volcano with a compact source-faithful distribution and patient-level overlay.
- **Weakness/next:** cell values are descriptive and patients are the independent units; use this as Panel A, with the reconstruction limitation in the caption.

#### Draft A2 — public-DGE patient reconstruction

```{r}
#| label: panel-a-reconstruction
#| fig-width: 6.2
#| fig-height: 5.2
#| out-width: 85%
#| fig-cap: "Patient-level TM4SF1 CPM reconstructed from the public DGE matrices. Open points fail the reconstructed 150-cell rule."

panel_a_reconstruction_data <- tm4_patient |>
  filter(
    sample_type == "tumour",
    broad_cell_type == "epithelial",
    paper_id %in% c(sprintf("VAR%02d", 1:9), sprintf("UC%02d", 1:3))
  ) |>
  mutate(
    reconstruction_support = if_else(
      reconstructed_n150_threshold_met, ">=150 cells", "<150 cells"
    )
  )

panel_a_reconstruction <- ggplot(
  panel_a_reconstruction_data,
  aes(histology_group, CPM, colour = histology_group, shape = reconstruction_support)
) +
  geom_point(
    position = position_jitter(width = 0.08, height = 0),
    size = 1.45, stroke = 0.38
  ) +
  stat_summary(
    fun = median, geom = "crossbar", width = 0.38,
    colour = fig3_col[["neutral"]], linewidth = 0.28
  ) +
  scale_colour_manual(values = fig3_col[c("HV", "UC")]) +
  scale_shape_manual(values = c(">=150 cells" = 16, "<150 cells" = 1)) +
  scale_y_continuous(
    trans = scales::transform_pseudo_log(sigma = 0.1, base = 10),
    breaks = c(0, 0.1, 1, 10, 100, 1000),
    labels = label_number()
  ) +
  labs(x = NULL, y = "Patient TM4SF1 CPM", colour = NULL, shape = NULL) +
  theme_fig3_compact() +
  theme(legend.position = "bottom", legend.box = "vertical")

panel_a_reconstruction
```

- **Message:** the patient-level public reconstruction supports HV enrichment but also exposes strong tumour heterogeneity.
- **Source:** reconstructed raw-count patient pseudobulks; closest fixed-seed comparison is 9 HV versus 2 UC at the 150-cell rule.
- **Size/change:** 57 × 48 mm; retains UC01 as a visibly low-cell, open point rather than hiding it.
- **Weakness/next:** it is not an exact 9 HV + 3 UC reproduction; retain as an alternate/audit panel, not the primary Panel A.

### Panel B — tumour anchor to Normal reference

```{r}
#| label: panel-b-anchor-bridge
#| fig-width: 6.2
#| fig-height: 5.2
#| out-width: 85%
#| fig-cap: "Tumour-cohort bootstrap distributions and the comprehensive epithelial posteriorHCA Normal reference on the epithelial model-aligned scale. The dashed line is the registered 5% HV anchor."

panel_b_draws <- study_draws |>
  filter(
    cell_type == "epithelial",
    group %in% c(
      "HV tumour epithelial",
      "UC tumour epithelial (2 public-DGE >=150)"
    )
  ) |>
  mutate(
    display_group = recode(
      group,
      "HV tumour epithelial" = "HV (n=9)",
      "UC tumour epithelial (2 public-DGE >=150)" = "UC (n=2)"
    ),
    display_group = factor(
      display_group,
      levels = c("Normal HCA", "UC (n=2)", "HV (n=9)")
    )
  )

panel_b_hca <- profile_summary |>
  filter(cell_type == "epithelial", dimension == "comprehensive") |>
  mutate(
    display_group = factor(
      "Normal HCA",
      levels = c("Normal HCA", "UC (n=2)", "HV (n=9)")
    )
  )

panel_b <- ggplot() +
  ggridges::geom_density_ridges(
    data = panel_b_draws,
    aes(log_mu, display_group, fill = display_group),
    scale = 0.82, rel_min_height = 0.015,
    colour = "white", linewidth = 0.24, alpha = 0.72
  ) +
  geom_segment(
    data = panel_b_hca,
    aes(
      x = lower_log_mu, xend = upper_log_mu,
      y = display_group, yend = display_group
    ),
    colour = fig3_col[["Normal"]], linewidth = 1.2, lineend = "round"
  ) +
  geom_point(
    data = panel_b_hca,
    aes(median_log_mu, display_group),
    shape = 21, fill = "white", colour = fig3_col[["Normal"]],
    size = 1.35, stroke = 0.42
  ) +
  geom_vline(
    xintercept = panel_b_hca$threshold_log_mu,
    colour = fig3_col[["HV"]], linetype = 3, linewidth = 0.3
  ) +
  scale_fill_manual(values = c(
    "HV (n=9)" = fig3_col[["HV"]],
    "UC (n=2)" = fig3_col[["UC"]]
  )) +
  labs(x = "Model-aligned log(μ)", y = NULL) +
  guides(fill = "none") +
  theme_fig3_compact() +
  theme(
    axis.line.y = element_blank(), axis.ticks.y = element_blank(),
    panel.grid.major.y = element_line(colour = "#ECECEC", linewidth = 0.2)
  )

panel_b
```

- **Message:** posteriorHCA evaluates Normal expression on an HV-derived tumour scale; it is not another tumour DEG test.
- **Source:** exact patient Bayesian-bootstrap study draws plus the exact comprehensive epithelial posterior median and 95% interval.
- **Size/change:** 57 × 48 mm; aligns tumour distributions, the HV anchor, and Normal reference in one compact bridge.
- **Weakness/next:** the panel is epithelial-model specific; state that cell-type model scales are separately aligned and not interchangeable.

### Panel C — broad Normal compartments

```{r}
#| label: panel-c-broad-compartments
#| fig-width: 6.2
#| fig-height: 5.2
#| out-width: 85%
#| fig-cap: "Patient-level TM4SF1 CPM in paired/adjacent normal bladder broad populations. Large diamonds are medians and bars are interquartile ranges."

panel_c_points <- tm4_patient |>
  filter(
    sample_type == "paired_normal",
    broad_cell_type %in% c("epithelial", "endothelial", "stromal", "immune")
  )

panel_c_summary <- panel_c_points |>
  group_by(broad_cell_type) |>
  summarise(
    median_CPM = median(CPM),
    lower_CPM = quantile(CPM, 0.25),
    upper_CPM = quantile(CPM, 0.75),
    .groups = "drop"
  ) |>
  arrange(median_CPM) |>
  mutate(broad_cell_type = factor(broad_cell_type, levels = broad_cell_type))

panel_c_points <- panel_c_points |>
  mutate(
    broad_cell_type = factor(
      broad_cell_type, levels = levels(panel_c_summary$broad_cell_type)
    )
  )

population_col <- c(
  epithelial = fig3_col[["epithelial"]],
  endothelial = fig3_col[["endothelial"]],
  stromal = fig3_col[["neutral"]],
  immune = "#A8A8A8"
)

panel_c <- ggplot(panel_c_points, aes(CPM, broad_cell_type, colour = broad_cell_type)) +
  geom_point(
    position = position_jitter(width = 0, height = 0.055),
    size = 0.8, alpha = 0.55
  ) +
  geom_segment(
    data = panel_c_summary,
    aes(x = lower_CPM, xend = upper_CPM, yend = broad_cell_type),
    linewidth = 0.62, lineend = "round"
  ) +
  geom_point(
    data = panel_c_summary, aes(median_CPM),
    shape = 18, size = 1.65
  ) +
  scale_colour_manual(values = population_col) +
  scale_x_continuous(
    trans = scales::transform_log1p(),
    breaks = c(0, 1, 10, 100, 1000), labels = label_number()
  ) +
  labs(x = "Paired-normal TM4SF1 CPM", y = NULL) +
  guides(colour = "none") +
  theme_fig3_compact() +
  theme(axis.line.y = element_blank(), axis.ticks.y = element_blank())

panel_c
```

- **Message:** paired/adjacent normal broad-population data motivate separate epithelial and endothelial follow-up.
- **Source:** reconstructed patient pseudobulks; points remain descriptive because these samples are adjacent/paired normal, not fully healthy tissue.
- **Size/change:** 57 × 48 mm; ranks populations by median expression and highlights the two modelled compartments without implying a broader posteriorHCA model set.
- **Weakness/next:** posteriorHCA has only epithelial and endothelial models here; keep that limitation explicit in the caption.

**Intended Row 1 message:** TM4SF1 is an HV-enriched tumour target in the
released source, its patient-level tumour abundance is heterogeneous, and
epithelial plus endothelial Normal compartments are the defensible focus for
model-aligned screening. A1–C make that transition successfully if A1 remains
the main source-faithful Panel A and A2 is retained as an audit alternative.

## Row 2 — main tissue-level posteriorHCA result

The smooth Normal slabs below are quantile-matched visual envelopes through the
cached posterior median and 95% interval. The interval bar, median point,
reported overlap probability (`P`), donor count (`n`), HV bootstrap density,
and registered anchor are exact cached results. Filled circle = at least 10
donor keys; open circle = 1–9; cross = zero. Support changes opacity and line
style only; it does not control visibility.

### Panel D — endothelial

#### Draft D1 — comprehensive tissue landscape

```{r}
#| label: panel-d-comprehensive
#| fig-width: 10
#| fig-height: 12
#| out-width: 100%
#| fig-cap: "All endothelial tissue baselines, ordered by P(Normal ≥ HV anchor). Normal slabs are quantile-matched renderings; exact medians and 95% intervals appear at each facet baseline."

panel_d_comprehensive <- plot_tissue_density("endothelial", ncol = 4)
panel_d_comprehensive
```

- **Message/source:** complete endothelial tissue landscape from the corrected profile summary and exact endothelial HV bootstrap draws.
- **Size/change:** 89 × 92 mm; replaces the report forest plot with repeated Normal-versus-HV distribution grammar and direct `P`, `n`, and support marks.
- **Weakness:** 29 facets are too dense for confident main-text reading at final size, and unsupported tissues broaden the common scale.
- **Recommendation:** retain as a Supplementary/exhaustive design reference, not the main Panel D.

#### Draft D2 — main-text prioritised

```{r}
#| label: panel-d-prioritised
#| fig-width: 8.5
#| fig-height: 9.0
#| out-width: 100%
#| fig-cap: "Prioritised endothelial tissue set. Female reproductive, skin, and respiratory tissues are retained for story evaluation; breast, liver, and renal provide supported comparators."

endo_main_tissues <- c(
  "female reproductive system",
  "integumentary system (skin)",
  "breast",
  "liver",
  "renal system",
  "respiratory system"
)
panel_d_prioritised <- plot_tissue_density(
  "endothelial", endo_main_tissues, ncol = 2
)
panel_d_prioritised
```

- **Message/source:** the high-overlap female-reproductive estimate is supported; the similarly high skin estimate is visibly low-support; respiratory is a supported negative comparator.
- **Size/change:** 89 × 92 mm; six facets preserve density shape and annotations at final size.
- **Weakness:** skin remains 98.6% extrapolation-weighted and should not be presented as a robust signal.
- **Recommendation:** use this grammar for Panel D, but consider moving skin to Supplementary once Row 3 adopts respiratory.

### Panel E — epithelial

#### Draft E1 — comprehensive tissue landscape

```{r}
#| label: panel-e-comprehensive
#| fig-width: 10
#| fig-height: 12
#| out-width: 100%
#| fig-cap: "All epithelial tissue baselines, ordered by P(Normal ≥ HV anchor). Normal slabs are quantile-matched renderings; exact medians and 95% intervals appear at each facet baseline."

panel_e_comprehensive <- plot_tissue_density("epithelial", ncol = 4)
panel_e_comprehensive
```

- **Message/source:** complete epithelial tissue landscape from the corrected profile summary and exact epithelial HV bootstrap draws.
- **Size/change:** 89 × 92 mm; uses the same visual grammar as D without forcing a shared, biologically non-equivalent model scale.
- **Weakness:** 21 facets remain crowded and the wide uncertainty makes small ranking differences visually unimportant.
- **Recommendation:** use as the exhaustive/Supplementary version.

#### Draft E2 — main-text prioritised

```{r}
#| label: panel-e-prioritised
#| fig-width: 8.5
#| fig-height: 9.0
#| out-width: 100%
#| fig-cap: "Prioritised epithelial tissue set, retaining supported high-expression tissues and respiratory/renal comparators."

epi_main_tissues <- c(
  "breast",
  "female reproductive system",
  "nasal, oral, and pharyngeal regions",
  "sensory-related structures",
  "liver",
  "renal system",
  "respiratory system"
)
panel_e_prioritised <- plot_tissue_density(
  "epithelial", epi_main_tissues, ncol = 2
)
panel_e_prioritised
```

- **Message/source:** epithelial tissue estimates show weaker overlap than the endothelial lead signal, with broad uncertainty despite adequate donor counts in the selected set.
- **Size/change:** 89 × 92 mm; seven supported facets remain legible and preserve the D-panel grammar.
- **Weakness:** none pass the registered 0.80 prototype rule, and the model itself is not readiness-qualified.
- **Recommendation:** use E2; keep the comprehensive E1 landscape in Supplementary.

**Intended Row 2 message:** tissue-specific Normal relationships to the HV
anchor differ sharply between endothelial and epithelial models. The overlaid
density/interval style is the most interpretable at 89 × 92 mm when restricted
to approximately six or seven facets. The comprehensive drafts are valuable
for inclusion review but are too dense for the main text.

## Row 3 — demographic refinement of selected tissues

Row 3 focuses on the endothelial model because the female-reproductive and
skin lead signals arise there. Each panel shows both proposed stories:
female reproductive + respiratory (supported negative comparator) and female
reproductive + skin (high-overlap but extrapolation-sensitive). Structurally
impossible male estimates for female reproductive tissue are absent.

### Panel F — sex

#### Draft F1 — female reproductive + respiratory

```{r}
#| label: panel-f-respiratory
#| fig-width: 6.4
#| fig-height: 6.0
#| out-width: 85%
#| fig-cap: "Endothelial sex marginals for the respiratory comparator. Purple is female and green is male; point shape denotes support. Female reproductive tissue has one structurally female tissue baseline rather than an invented male estimate."

panel_f_respiratory <- plot_story_sex("respiratory system")
panel_f_respiratory
```

- **Message/source:** respiratory female and male marginals are directly comparable and supported; female reproductive remains a female-only tissue baseline.
- **Size/change:** 58 × 55 mm; paired point-ranges replace the report-wide sex forest plot.
- **Weakness:** intervals are broad and do not establish a robust sex effect.
- **Recommendation:** preferred Panel F draft because the comparator is adequately supported.

#### Draft F2 — female reproductive + skin

```{r}
#| label: panel-f-skin
#| fig-width: 6.4
#| fig-height: 6.0
#| out-width: 85%
#| fig-cap: "Endothelial sex marginals for skin. Purple is female and green is male; point shape denotes support. Both skin sex estimates have limited or absent direct support."

panel_f_skin <- plot_story_sex("integumentary system (skin)")
panel_f_skin
```

- **Message/source:** the apparent skin sex contrast is prediction-visible but empirically weak.
- **Size/change:** 58 × 55 mm; open/cross marks keep low-support estimates visible without giving them equal visual authority.
- **Weakness:** female skin has zero donor keys and male skin has two; neither can sustain a main-text demographic claim.
- **Recommendation:** retain for development/Supplementary only.

### Panel G — age decade

#### Draft G1 — female reproductive + respiratory

```{r}
#| label: panel-g-respiratory
#| fig-width: 6.4
#| fig-height: 6.0
#| out-width: 85%
#| fig-cap: "Endothelial overlap probability by age decade for female reproductive and respiratory tissues. Lines are visual guides between discrete marginal estimates."

panel_g_respiratory <- plot_story_age("respiratory system")
panel_g_respiratory
```

- **Message/source:** female-reproductive overlap is consistently high but sparsely supported; respiratory remains lower with several supported decades.
- **Size/change:** 58 × 55 mm; ordered points and guide-lines replace the age heatmap.
- **Weakness:** decade-specific marginals are not a fitted continuous biological age trend.
- **Recommendation:** preferred Panel G draft, with the guide-line caveat in the caption.

#### Draft G2 — female reproductive + skin

```{r}
#| label: panel-g-skin
#| fig-width: 6.4
#| fig-height: 6.0
#| out-width: 85%
#| fig-cap: "Endothelial overlap probability by age decade for female reproductive and skin tissues. Nearly all skin strata are extrapolated."

panel_g_skin <- plot_story_age("integumentary system (skin)")
panel_g_skin
```

- **Message/source:** high skin overlap across decades is dominated by extrapolated predictions rather than repeated empirical support.
- **Size/change:** 58 × 55 mm; support-aware point/line design makes that limitation immediately visible.
- **Weakness:** the apparent pattern is unsuitable for a demographic biology claim.
- **Recommendation:** move to Supplementary or retire after story selection.

### Panel H — ethnicity

#### Draft H1 — female reproductive + respiratory

```{r}
#| label: panel-h-respiratory
#| fig-width: 6.4
#| fig-height: 6.0
#| out-width: 85%
#| fig-cap: "Endothelial overlap probability across unordered ethnicity groups for female reproductive and respiratory tissues. No lines connect categories."

panel_h_respiratory <- plot_story_ethnicity("respiratory system")
panel_h_respiratory
```

- **Message/source:** ethnicity estimates are heterogeneous in uncertainty and support; respiratory includes supported African and European strata.
- **Size/change:** 58 × 55 mm; a horizontal dot plot replaces the ethnicity heatmap and avoids implying ordinal structure.
- **Weakness:** sparse categories, especially in female reproductive tissue, prevent strong demographic interpretation.
- **Recommendation:** preferred Panel H draft if Row 3 remains in the main figure; caption as exploratory marginal structure.

#### Draft H2 — female reproductive + skin

```{r}
#| label: panel-h-skin
#| fig-width: 6.4
#| fig-height: 6.0
#| out-width: 85%
#| fig-cap: "Endothelial overlap probability across unordered ethnicity groups for female reproductive and skin tissues. Skin strata are almost entirely extrapolated."

panel_h_skin <- plot_story_ethnicity("integumentary system (skin)")
panel_h_skin
```

- **Message/source:** skin ethnicity estimates remain visible but their extrapolation-heavy basis is unmistakable.
- **Size/change:** 58 × 55 mm; restrained opacity and open/cross marks keep support secondary to the biological metric.
- **Weakness:** no skin ethnicity category has adequate direct donor support.
- **Recommendation:** do not use for the main-text story.

**Intended Row 3 message:** demographic marginals can refine a supported lead
and a motivated negative comparator without turning sparse predictions into
demographic claims. Female reproductive + respiratory is the cleaner main-text
story; female reproductive + skin is extrapolation-sensitive and belongs in
Supplementary development material.

# Recommended Fig. 3 direction

1. **Panels A–C:** use A1 (released source distribution with patient overlays), B (tumour bootstrap → HV anchor → epithelial Normal interval), and C (ranked paired-normal broad compartments). Keep A2 as an audit alternative.
2. **Row allocation:** retain the proposed 48 / 92 / 55 mm row heights with approximately 4 mm gutters; Row 2 remains the dominant visual row.
3. **Panels D–E:** use the prioritised two-column density-facet drafts. For D retain female reproductive, breast, liver, renal, respiratory, and initially skin for transparent story evaluation; move the other endothelial tissues to Supplementary. For E retain the seven supported tissues shown in E2 and move the full landscape to Supplementary.
4. **Panels F–H:** use female reproductive + respiratory. Respiratory is a motivated, adequately supported negative comparator; skin should not carry the main-text narrative while its tissue result is 98.6% extrapolation-weighted.
5. **Retire from the main figure:** the large TM4SF1 volcano, pooled broad-population bars, all-tissue forest plot, sex-wide forest plot, and age/ethnicity heatmaps. They remain useful audit or Supplementary outputs.
6. **Still requiring refinement:** replace the quantile-matched Row 2 Normal envelopes with exact tissue posterior draws if a lightweight draw cache is exported; decide whether skin remains in D after Row 3 is fixed; test final caption density and strip abbreviation after assembly.
7. **Estimated assembled size:** 183 mm × approximately 203 mm, within the 183 mm × 220 mm Nature-style cap, before caption.

## Figure-development self-review

- **Contribution — pass:** the row sequence carries one coherent story from target selection to tissue screening to demographic refinement.
- **Clarity — pass with one caption requirement:** model-specific scales and the plotting-only Row 2 density interpolation must be stated explicitly.
- **Experimental strength — needs model refit:** the released and reconstructed study evidence is transparent, but neither posteriorHCA fit is model-ready.
- **Evaluation completeness — needs one export:** exact tissue-level posterior draw matrices are absent from the cache; exporting them would remove the only visual approximation.
- **Design soundness — pass for development:** overlap, support, extrapolation, candidate status, and readiness remain separate; no panel implies toxicity or demographic causality.

## Claim–evidence check

| Draft claim | Evidence used | Status |
|---|---|---|
| TM4SF1 is HV-enriched in the released tumour study | Released Figure 5A rank/effect and Figure 5B expression values | Supported |
| Endothelial and epithelial compartments warrant focused Normal screening | Patient broad-population summaries plus the two available posteriorHCA model classes | Supported with scope stated |
| Female reproductive is a lead endothelial overlap signal | Corrected tissue summary: P = 0.87, 10 donor keys; model readiness false | Supported as an ungated prioritisation signal only |
| Skin shows robust demographic structure | Two donor keys at tissue level and extrapolation-heavy strata | Not supported; excluded from recommended main-text story |

## Session info

```{r}
sessionInfo()
```

